---
title: "`r local({ .c <- yaml::read_yaml(file.path(normalizePath(dirname(knitr::current_input())), 'execution_info', 'config_used.yaml')); pname <- .c$project$name %||% ''; analyst <- gsub('_', ' ', .c$project$analyst %||% ''); parts <- c('RNA-seq Analysis Report'); if (nzchar(pname)) parts <- c(parts, pname); if (nzchar(analyst)) parts <- c(parts, analyst); paste(parts, collapse = ' — ') })`"
subtitle: "multiomics-core Pipeline"
author: "Bioinformatician: `r yaml::read_yaml(file.path(normalizePath(dirname(knitr::current_input())), 'execution_info', 'config_used.yaml'))$project$analyst`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    fig_width: 10
    fig_height: 7
    self_contained: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%"
)

library(yaml)
library(knitr)
library(DT)
library(plotly)
library(ggplot2)
library(htmltools)
library(jsonlite)

# Source interactive cutoff panel helpers
tryCatch({
  .cp_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
  .cp_found <- FALSE
  for (.cp_i in 1:6) {
    .cp_try <- file.path(.cp_dir, "R", "core", "17_cutoff_panel.R")
    if (file.exists(.cp_try)) { source(.cp_try); .cp_found <- TRUE; break }
    .cp_dir <- dirname(.cp_dir)
  }
  if (!.cp_found) message("cutoff_panel helpers not found (walked 6 levels up)")
}, error = function(e) message("cutoff_panel helpers not found: ", e$message))

# Load ggVennDiagram if available
has_ggVennDiagram <- FALSE
tryCatch({
    library(ggVennDiagram)
    has_ggVennDiagram <- TRUE
}, error = function(e) {
    message("ggVennDiagram not available: ", e$message)
})

# Helper function for downloadable DT tables
make_dt <- function(df, caption = NULL, pageLength = 15) {
    DT::datatable(
        df,
        caption = caption,
        extensions = 'Buttons',
        options = list(
            scrollX = TRUE,
            pageLength = pageLength,
            dom = 'Bfrtip',
            buttons = list(
                list(extend = 'csv', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data")),
                list(extend = 'excel', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data"))
            )
        ),
        rownames = FALSE
    )
}

# Paths
run_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
config_path <- file.path(run_dir, "execution_info", "config_used.yaml")
diag_dir <- file.path(run_dir, "rna", "Diagnostic_plots")
qc_post_dir <- file.path(diag_dir, "QC_post")
datasets_dir <- file.path(run_dir, "rna", "Datasets")
enrich_dir <- file.path(run_dir, "rna", "Enrichment")
enrich_plot_dir <- file.path(enrich_dir, "plots")
filter_plot <- file.path(dirname(run_dir), "rnaseq", "filtering_threshold_qc.png")
commentary_dir <- file.path(run_dir, "rna", "commentary")

# Load commentary if available
commentary <- NULL
has_commentary <- FALSE
commentary_json <- file.path(commentary_dir, "commentary_all.json")
if (file.exists(commentary_json)) {
    commentary <- tryCatch(
        jsonlite::read_json(commentary_json),
        error = function(e) NULL
    )
    has_commentary <- !is.null(commentary) && length(commentary) > 0
}

# Helper: render figure legend (paper-style caption below figure)
figure_legend <- function(text) {
    cat(sprintf("<p class='figure-legend' style='text-align: center; font-size: 0.9em; color: #555; margin: 4px auto 12px auto; max-width: 88%%; line-height: 1.45; font-style: italic;'>%s</p>\n", text))
}

# Helper: render commentary block for a figure_id (collapsible)
render_commentary <- function(figure_id) {
    if (!has_commentary) return(invisible(NULL))
    c <- commentary[[figure_id]]
    if (is.null(c)) return(invisible(NULL))

    cat("<details class='commentary-details' style='margin: 4px 0 18px 0;'>\n")
    cat("<summary style='cursor: pointer; color: #4a90d9; font-weight: 500; font-size: 0.93em; padding: 3px 0;'>Show AI Interpretation</summary>\n")

    html <- "<div class='commentary-block' style='background: #f8f9fa; border-left: 4px solid #4a90d9; padding: 12px 16px; margin: 6px 0 4px 0; border-radius: 0 4px 4px 0; font-size: 0.92em;'>\n"
    html <- paste0(html, "<p style='margin: 0 0 8px 0;'><strong>What this shows:</strong> ",
                   htmltools::htmlEscape(c$what_is_this %||% ""), "</p>\n")

    # Biological interpretation (prose paragraph)
    bio_interp <- c$biological_interpretation %||% ""
    if (nzchar(bio_interp)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Interpretation:</strong> ",
                       htmltools::htmlEscape(bio_interp), "</p>\n")
    }

    obs <- c$observations
    if (length(obs) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Key observations:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (o in obs) html <- paste0(html, "<li>", htmltools::htmlEscape(o), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    # Statistical assessment
    stat_assess <- c$statistical_assessment %||% ""
    if (nzchar(stat_assess)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Statistical note:</strong> ",
                       htmltools::htmlEscape(stat_assess), "</p>\n")
    }

    issues <- c$issues_checks
    if (length(issues) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Checks:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (issue in issues) html <- paste0(html, "<li>", htmltools::htmlEscape(issue), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    steps <- c$next_steps
    if (length(steps) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Next steps:</strong></p>\n<ul style='margin: 2px 0 4px 0; padding-left: 20px;'>\n")
        for (s in steps) html <- paste0(html, "<li>", htmltools::htmlEscape(s), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    backend <- c$backend %||% "unknown"
    confidence <- c$confidence %||% ""
    footer <- backend
    if (nzchar(confidence) && confidence != "none") footer <- paste0(footer, " | confidence: ", confidence)
    # Attribution: Claude (Anthropic) + K-Dense AI scientific-skills
    backend_label <- if (grepl("claude", backend, ignore.case = TRUE)) {
        "Claude (Anthropic) + K-Dense AI scientific-skills"
    } else if (grepl("openai", backend, ignore.case = TRUE)) {
        "OpenAI + K-Dense AI scientific-skills"
    } else if (backend == "data_driven") {
        "data-driven + K-Dense AI scientific-skills"
    } else {
        footer
    }
    if (nzchar(confidence) && confidence != "none") {
        backend_label <- paste0(backend_label, " | confidence: ", confidence)
    }
    html <- paste0(html, "<p style='margin: 4px 0 0 0; color: #888; font-size: 0.85em;'><em>Generated by: ",
                   htmltools::htmlEscape(backend_label), "</em></p>\n</div>\n")

    cat(html)
    cat("</details>\n")
}

# Load config
cfg <- yaml::read_yaml(config_path)
rna_cfg <- cfg$modes$rna

# Load DE data (used in multiple sections)
final_results_file <- file.path(datasets_dir, "final_results.tsv")
de_data <- NULL
if (file.exists(final_results_file)) {
    de_data <- read.delim(final_results_file, stringsAsFactors = FALSE, check.names = FALSE)
    # Add gene names from annotation if available
    anno_file_setup <- file.path(enrich_dir, "gene_annotation.csv")
    if (file.exists(anno_file_setup)) {
        anno_setup <- read.csv(anno_file_setup, stringsAsFactors = FALSE)
        id_col_setup <- if ("Gene" %in% names(de_data)) "Gene" else if ("FeatureID" %in% names(de_data)) "FeatureID" else NULL
        if (!is.null(id_col_setup)) {
            de_data$gene_id_clean <- gsub("^Gene:", "", de_data[[id_col_setup]])
            anno_match_setup <- anno_setup[match(de_data$gene_id_clean, anno_setup$gene_id), ]
            de_data$GeneName <- ifelse(
                !is.na(anno_match_setup$symbol) & anno_match_setup$symbol != anno_match_setup$gene_id,
                anno_match_setup$symbol,
                de_data$gene_id_clean
            )
        }
    }
}

# Get contrast names from column patterns
contrast_names <- character(0)
if (!is.null(de_data)) {
    padj_cols <- grep("^padj\\.", names(de_data), value = TRUE)
    contrast_names <- gsub("^padj\\.", "", padj_cols)
}

# Target genes from config (if specified)
target_genes <- rna_cfg$target_genes %||% character(0)
```

```{r force-dependencies, include=FALSE}
# Force DT and Plotly dependencies to be included in the HTML
# This ensures the JavaScript/CSS is loaded before any dynamic content
DT::datatable(data.frame(x = 1))
plotly::plot_ly(data.frame(x = 1, y = 1), x = ~x, y = ~y, type = "scatter", mode = "markers")
```

```{r report-mode-note, results='asis', echo=FALSE}
cat('<div style="background: #eef6ff; border: 1px solid #b8d4f0; border-radius: 6px; padding: 10px 16px; margin: 0 0 20px 0; font-size: 0.9em; color: #2c5282;">
<strong>Report mode:</strong> This report is in <em>clean mode</em> by default (code hidden). To view the R code behind any section, click the <code style="background: #fff; padding: 1px 5px; border-radius: 3px; border: 1px solid #ccc;">Code</code> button that appears at the top-right of each section. Use the global <code style="background: #fff; padding: 1px 5px; border-radius: 3px; border: 1px solid #ccc;">Code</code> dropdown at the top of the page to show/hide all code at once.
</div>\n')
```

# Technical Summary

```{r technical-summary, results='asis', echo=FALSE}
tech <- rna_cfg$technical_report

# --- Sample & Sequencing Overview ---
# Always render this subsection using computed stats + config
meta_ts <- tryCatch(
    read.csv(file.path(cfg$project$dir, cfg$paths$raw, rna_cfg$files$metadata),
             stringsAsFactors = FALSE),
    error = function(e) NULL
)
grp_col <- rna_cfg$filtering$group_col

n_samples_ts <- if (!is.null(meta_ts)) nrow(meta_ts) else NA
n_groups_ts <- if (!is.null(meta_ts) && grp_col %in% names(meta_ts)) length(unique(meta_ts[[grp_col]])) else NA

cat("## Sample & Sequencing Overview\n\n")
overview_parts <- character(0)
if (!is.na(n_samples_ts) && !is.na(n_groups_ts)) {
    overview_parts <- c(overview_parts,
        sprintf("This analysis comprises **%d samples** across **%d experimental groups** (grouping variable: *%s*).",
                n_samples_ts, n_groups_ts, grp_col))
}
if (!is.null(tech$facility)) {
    overview_parts <- c(overview_parts,
        sprintf("Sequencing was performed at the **%s**.", tech$facility))
}
if (length(overview_parts) > 0) cat(paste(overview_parts, collapse = " "), "\n\n")

# --- Upstream sections (conditional on technical_report block) ---
if (!is.null(tech)) {

    if (!is.null(tech$library_prep)) {
        cat("## Library Preparation\n\n")
        cat(tech$library_prep, "\n\n")
    }

    if (!is.null(tech$sequencing)) {
        cat("## Sequencing\n\n")
        cat(tech$sequencing, "\n\n")
    }

    if (!is.null(tech$upstream_bioinformatics)) {
        cat("## Upstream Bioinformatics\n\n")
        cat(tech$upstream_bioinformatics, "\n\n")
    }
}

# --- Downstream Analysis (always rendered from pipeline config) ---
cat("## Downstream Analysis (multiomics-core)\n\n")

norm_method <- rna_cfg$normalization$method %||% "TMM"
prior_ct <- rna_cfg$normalization$prior.count %||% 1
filt_min <- rna_cfg$filtering$auto_filter$min %||% 0.5
filt_max <- rna_cfg$filtering$auto_filter$max %||% 2.0
de_method <- rna_cfg$de$method %||% "deseq2"
deseq_mode <- rna_cfg$de$deseq_mode %||% "default"
p_cut <- rna_cfg$de$p_cutoff %||% 0.05
fc_cut <- rna_cfg$de$linear_fc_cutoff %||% 1.5
use_adj <- isTRUE(rna_cfg$de$use_adj)
p_label <- if (use_adj) "adjusted p-value" else "p-value"
p_corr <- if (use_adj) " P-values were corrected using Benjamini-Hochberg FDR." else ""

pathway_enabled <- isTRUE(rna_cfg$pathway$enabled)
pathway_method <- rna_cfg$pathway$method %||% "fgsea"
pathway_dbs <- paste(rna_cfg$pathway$databases %||% "GO", collapse = " and ")

cat(sprintf(
    paste0(
        "Gene filtering was performed using an adaptive density-based threshold ",
        "(KDE on log2-CPM distribution), selecting an optimal CPM threshold within the ",
        "[%.1f\u2013%.1f] range. Genes passing the threshold in at least one complete ",
        "biological group were retained. Normalization was performed using the %s method ",
        "with log2-CPM transformation (prior count = %d). ",
        "Differential expression was tested using %s (%s test, %s mode). ",
        "Genes were called significant at %s \u2264 %s and absolute linear fold-change \u2265 %.1f.%s"
    ),
    filt_min, filt_max,
    norm_method, prior_ct,
    toupper(de_method), "Wald", deseq_mode,
    p_label, format(p_cut, nsmall = 2), fc_cut,
    p_corr
))

if (pathway_enabled) {
    cat(sprintf(" Pathway enrichment analysis was performed using %s against %s databases.",
                pathway_method, pathway_dbs))
}
cat("\n\n")

# --- Methodological Note (conditional on upstream_de_note) ---
if (!is.null(tech$upstream_de_note)) {
    cat("## Methodological Note\n\n")

    # Compute our pipeline DE count if available
    de_counts_ts_file <- file.path(datasets_dir, "de_summary_counts.tsv")
    our_de_n <- NA
    if (file.exists(de_counts_ts_file)) {
        de_cnt_ts <- read.delim(de_counts_ts_file, stringsAsFactors = FALSE)
        pass_row <- de_cnt_ts[grepl("pass_any", de_cnt_ts[[1]], ignore.case = TRUE), ]
        if (nrow(pass_row) > 0 && ncol(de_cnt_ts) >= 4) {
            our_de_n <- pass_row[1, 4]
        }
    }
    # Fallback: count from final_results
    if (is.na(our_de_n) && !is.null(de_data)) {
        pac <- grep("pass_any", names(de_data), value = TRUE)[1]
        if (!is.na(pac)) our_de_n <- sum(de_data[[pac]] == TRUE, na.rm = TRUE)
    }

    cat(tech$upstream_de_note, " ")
    cat(sprintf(
        paste0(
            "This report applies additional filtering: adaptive CPM-based gene filtering, ",
            "%s normalization, and a fold-change cutoff of \u2265 %.1f"
        ),
        norm_method, fc_cut
    ))
    if (!is.na(our_de_n)) {
        cat(sprintf(", resulting in **%s DEGs**", format(as.integer(our_de_n), big.mark = ",")))
    }
    cat(". The stricter criteria focus on genes with both statistical significance and biological effect size.\n\n")
}

# --- Acknowledgment (conditional) ---
if (!is.null(tech$acknowledgment)) {
    cat("## Acknowledgment\n\n")
    cat(sprintf("*%s*\n\n", tech$acknowledgment))
}
```

```{r project-overview, results='asis'}
# Project Overview — user-provided notes (optionally AI-enhanced)
user_notes <- cfg$project$user_notes
if (!is.null(user_notes) && nzchar(trimws(user_notes))) {
    cat("# Project Overview\n\n")

    # Try to load pre-generated summary if available
    summary_file <- file.path(run_dir, "rna", "project_summary.html")
    if (file.exists(summary_file)) {
        cat(paste(readLines(summary_file, warn = FALSE), collapse = "\n"))
    } else {
        # Fallback: render user notes as-is
        notes_html <- gsub("\n", "<br>\n", htmltools::htmlEscape(trimws(user_notes)))
        cat(sprintf("<div style='background:#f8f9fa; border-left:4px solid #4a90d9; padding:16px 20px; border-radius:6px; margin-bottom:20px; font-size:14px; line-height:1.7; color:#333;'>%s</div>\n", notes_html))
    }
    cat("\n")
}
```

# Analysis Parameters {.tabset}

## Project

```{r params-project}
proj <- data.frame(
  Parameter = c("Project Name", "Analysis Round", "Project Directory",
                 "Raw Data Path", "Output Path"),
  Value = c(cfg$project$name, cfg$project$analysis_round, cfg$project$dir,
            cfg$paths$raw, cfg$paths$out),
  stringsAsFactors = FALSE
)
make_dt(proj, "Analysis Parameters")
```

## Experimental Design

```{r params-design}
meta <- read.csv(
  file.path(cfg$project$dir, cfg$paths$raw, rna_cfg$files$metadata),
  stringsAsFactors = FALSE
)
group_col <- rna_cfg$filtering$group_col

design_tbl <- as.data.frame(table(meta[[group_col]]))
colnames(design_tbl) <- c("Group", "n (replicates)")
make_dt(design_tbl, "Experimental Design")
```

**Grouping variable:** `r group_col`
**Total samples:** `r nrow(meta)`

## Pipeline Settings

```{r params-pipeline}
prior_count <- rna_cfg$normalization$prior.count %||% 1
settings <- data.frame(
  Parameter = c(
    "Normalization Method",
    paste0("Prior Count (added before log2 to avoid log(0): log2(CPM + ", prior_count, "))"),
    "Filtering Group Column", "Auto-filter CPM Threshold Range",
    "DE Method", "DESeq2 Mode",
    "Adjusted p-value Cutoff", "Linear FC Cutoff",
    "Clustering Enabled"
  ),
  Value = c(
    rna_cfg$normalization$method,
    prior_count,
    rna_cfg$filtering$group_col,
    paste0(rna_cfg$filtering$auto_filter$min, " – ", rna_cfg$filtering$auto_filter$max, " CPM (pipeline selects optimal within this range)"),
    rna_cfg$de$method, rna_cfg$de$deseq_mode,
    rna_cfg$de$p_cutoff, rna_cfg$de$linear_fc_cutoff,
    ifelse(isTRUE(rna_cfg$clustering$enabled), "Yes", "No")
  ),
  stringsAsFactors = FALSE
)
make_dt(settings, "Pipeline Settings")
```

## Contrasts

```{r params-contrasts}
contrasts <- read.csv(
  file.path(cfg$project$dir, cfg$paths$raw, rna_cfg$files$contrasts),
  stringsAsFactors = FALSE
)
make_dt(contrasts, "Contrasts")
```

## Filtering Strategy

The pipeline applies **CPM-based auto-filtering** to remove lowly expressed genes before differential expression analysis. Genes that do **not** reach a minimum expression level (CPM threshold) are removed. This is a standard and necessary practice in RNA-seq to:

1. **Reduce noise**: Genes with very low counts have unreliable fold-change estimates
2. **Improve statistical power**: Fewer tests means less severe multiple testing correction
3. **Remove technical artifacts**: Many zero/low counts represent sequencing noise, not biology

**How auto-filtering works:**

```{r filtering-strategy}
filter_params <- data.frame(
  Parameter = c(
    "Grouping Column",
    "CPM Threshold Search Range",
    "Fallback Threshold",
    "Filtering Rule"
  ),
  Value = c(
    rna_cfg$filtering$group_col,
    paste0(rna_cfg$filtering$auto_filter$min, " – ", rna_cfg$filtering$auto_filter$max, " CPM (pipeline automatically selects the optimal threshold within this range)"),
    paste0(rna_cfg$filtering$auto_filter$fallback %||% 1.0, " CPM"),
    "Gene is KEPT only if it has CPM ≥ threshold in ALL samples of at least ONE biological group"
  ),
  stringsAsFactors = FALSE
)
make_dt(filter_params, "Filtering Parameters")
```

**Algorithm:**

1. Calculate CPM (counts per million) for each gene in each sample
2. For each candidate threshold (from min to max range):
   - For each biological group, check if the gene has CPM ≥ threshold in ALL samples of that group
   - A gene is **kept** if it meets the threshold in at least one complete group; otherwise it is **removed**
3. The pipeline automatically selects the optimal threshold that balances gene retention with noise reduction
4. The selected threshold is shown in the QC plot below

This group-aware filtering ensures that genes specific to certain conditions are retained, even if they have low expression in other conditions. Genes that are lowly expressed across **all** conditions are removed as uninformative.

---

# Executive Summary

```{r executive-summary, results='asis', echo=FALSE}
# Compute gene counts early (also used by filter-stats section later)
raw_counts_file_exec <- file.path(cfg$project$dir, cfg$paths$raw, rna_cfg$files$counts)
filt_counts_file_exec <- file.path(datasets_dir, "rna_counts_filtered.tsv")
n_genes_raw <- NA
n_genes_filt <- NA
if (file.exists(raw_counts_file_exec)) {
    n_genes_raw <- nrow(read.csv(raw_counts_file_exec, stringsAsFactors = FALSE, check.names = FALSE))
}
if (file.exists(filt_counts_file_exec)) {
    n_genes_filt <- nrow(read.delim(filt_counts_file_exec, stringsAsFactors = FALSE, check.names = FALSE))
}

exec_summary_path <- file.path(run_dir, "rna", "executive_summary.md")
exec_available <- file.exists(exec_summary_path)
exec_md_raw <- ""
if (exec_available) {
    exec_md_raw <- paste(readLines(exec_summary_path, warn = FALSE), collapse = "\n")
    # Remove duplicate heading if present (the section heading above is enough)
    exec_md_raw <- gsub("^##?\\s*Executive\\s*Summary\\s*\n+", "", exec_md_raw)
}

# Build a rich executive summary from available data
cat("<div style='background: linear-gradient(135deg, #f8f9fa 0%, #e9ecef 100%); border: 1px solid #dee2e6; border-radius: 8px; padding: 20px 24px; margin: 10px 0 20px 0;'>\n")

# Key numbers panel
cat("<div style='display: flex; flex-wrap: wrap; gap: 16px; margin-bottom: 16px;'>\n")

# Samples
n_samp <- if (exists("meta") && is.data.frame(meta)) nrow(meta) else NA
if (!is.na(n_samp)) {
    cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #2c3e50;'>%d</div><div style='font-size: 0.85em; color: #666;'>Samples</div></div>\n", n_samp))
}

# Groups
n_groups <- if (exists("meta") && is.data.frame(meta) && exists("group_col") && group_col %in% names(meta)) length(unique(meta[[group_col]])) else NA
if (!is.na(n_groups)) {
    cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #2c3e50;'>%d</div><div style='font-size: 0.85em; color: #666;'>Groups</div></div>\n", n_groups))
}

# Genes after filtering
if (!is.na(n_genes_filt)) {
    cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #2c3e50;'>%s</div><div style='font-size: 0.85em; color: #666;'>Genes (filtered)</div></div>\n", format(n_genes_filt, big.mark = ",")))
}

# Contrasts
n_contrasts <- length(contrast_names)
if (n_contrasts > 0) {
    cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #2c3e50;'>%d</div><div style='font-size: 0.85em; color: #666;'>Contrasts</div></div>\n", n_contrasts))
}

# Total DE genes
if (!is.null(de_data)) {
    pass_any_col <- grep("pass_any", names(de_data), value = TRUE)[1]
    if (!is.na(pass_any_col)) {
        n_de_total <- sum(de_data[[pass_any_col]] == TRUE, na.rm = TRUE)
    } else {
        padj_cols_exec <- grep("^padj\\.", names(de_data), value = TRUE)
        if (length(padj_cols_exec) > 0) {
            any_sig <- apply(de_data[, padj_cols_exec, drop = FALSE], 1, function(x) any(x <= 0.05, na.rm = TRUE))
            n_de_total <- sum(any_sig, na.rm = TRUE)
        } else {
            n_de_total <- NA
        }
    }
    if (!is.na(n_de_total)) {
        cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #c0392b;'>%s</div><div style='font-size: 0.85em; color: #666;'>DE Genes</div></div>\n", format(n_de_total, big.mark = ",")))
    }
}

cat("</div>\n")

# DE breakdown per contrast (skip pass_any rows)
de_counts_exec <- file.path(datasets_dir, "de_summary_counts.tsv")
if (file.exists(de_counts_exec)) {
    de_cnt <- read.delim(de_counts_exec, stringsAsFactors = FALSE)
    # Filter out pass_any rows
    de_cnt <- de_cnt[!grepl("pass_any", de_cnt[[1]], ignore.case = TRUE), , drop = FALSE]
    if (nrow(de_cnt) > 0) {
        cat("<div style='margin: 12px 0;'>\n")
        cat("<strong>Differential Expression per Contrast:</strong>\n<ul style='margin: 6px 0; padding-left: 20px;'>\n")
        for (i in seq_len(nrow(de_cnt))) {
            cn_name <- de_cnt[i, 1]
            up_val <- if (ncol(de_cnt) >= 2) de_cnt[i, 2] else "?"
            dn_val <- if (ncol(de_cnt) >= 3) de_cnt[i, 3] else "?"
            total_val <- if (ncol(de_cnt) >= 4) de_cnt[i, 4] else "?"
            cat(sprintf("<li><strong>%s</strong>: %s DE genes (%s up, %s down)</li>\n",
                        cn_name, total_val, up_val, dn_val))
        }
        cat("</ul>\n</div>\n")
    }
}

# Pipeline-generated summary (from exec summary module)
if (nzchar(trimws(exec_md_raw)) && trimws(exec_md_raw) != "No summary available.") {
    cat("<div style='margin-top: 12px; padding-top: 12px; border-top: 1px solid #dee2e6;'>\n")
    cat(exec_md_raw, "\n")
    cat("</div>\n")
}

cat("</div>\n")
```

---

# Quality Control — Pre-DE

## Gene Filtering

Lowly expressed genes were removed using the auto-filtering strategy described above.
The pipeline evaluates CPM distributions across biological groups and selects an optimal threshold.

Low-abundance features with insufficient signal are removed to reduce noise and improve statistical power. The filtering threshold balances retaining biologically relevant features against eliminating measurement noise.

```{r filter-stats}
# Read raw and filtered counts to get gene numbers
raw_counts_file <- file.path(cfg$project$dir, cfg$paths$raw, rna_cfg$files$counts)
filt_counts_file <- file.path(datasets_dir, "rna_counts_filtered.tsv")

n_genes_raw <- NA
n_genes_filt <- NA
raw_counts <- NULL
filt_counts <- NULL

if (file.exists(raw_counts_file)) {
    raw_counts <- read.csv(raw_counts_file, stringsAsFactors = FALSE, check.names = FALSE)
    n_genes_raw <- nrow(raw_counts)
}
if (file.exists(filt_counts_file)) {
    filt_counts <- read.delim(filt_counts_file, stringsAsFactors = FALSE, check.names = FALSE)
    n_genes_filt <- nrow(filt_counts)
}

if (!is.na(n_genes_raw) && !is.na(n_genes_filt)) {
    n_removed <- n_genes_raw - n_genes_filt
    pct_kept <- round(100 * n_genes_filt / n_genes_raw, 1)
    cat(sprintf("**Genes before filtering:** %s\n\n", format(n_genes_raw, big.mark = ",")))
    cat(sprintf("**Genes after filtering:** %s\n\n", format(n_genes_filt, big.mark = ",")))
    cat(sprintf("**Genes removed:** %s (%.1f%% retained)\n\n",
                format(n_removed, big.mark = ","), pct_kept))
}
```

```{r filter-plot, fig.cap="CPM filtering threshold selection"}
if (file.exists(filter_plot)) knitr::include_graphics(filter_plot)
```

```{r legend-filter-plot, results='asis', echo=FALSE}
if (file.exists(filter_plot)) {
    figure_legend("CPM filtering threshold selection. The pipeline evaluates the number of genes retained at each candidate CPM threshold within the specified range and selects the optimal value that balances noise reduction with gene retention.")
}
```

### Expression Distribution Before vs After Filtering {.tabset}

These histograms demonstrate why filtering is necessary: the "Before" histogram shows a large spike of lowly/non-expressed genes (noise), which is removed after filtering.

#### Before Filtering

```{r filter-hist-before, fig.height=5, fig.width=8}
if (!is.null(raw_counts)) {
    # Compute log2 CPM for raw counts (skip first column if it's gene IDs)
    num_cols_raw <- sapply(raw_counts, is.numeric)
    if (any(num_cols_raw)) {
        raw_mat <- as.matrix(raw_counts[, num_cols_raw, drop = FALSE])
        raw_mat[is.na(raw_mat)] <- 0
        lib_sizes <- colSums(raw_mat, na.rm = TRUE)
        raw_cpm <- t(t(raw_mat) / lib_sizes * 1e6)
        raw_log2cpm <- log2(raw_cpm + 1)
        avg_expr_raw <- rowMeans(raw_log2cpm, na.rm = TRUE)

        hist(avg_expr_raw, breaks = 80, col = "#E41A1C88", border = "white",
             main = paste0("Before Filtering (", format(length(avg_expr_raw), big.mark = ","), " genes)"),
             xlab = "log2(CPM + 1)", ylab = "Number of genes", cex.main = 1.2)
        abline(v = median(avg_expr_raw), col = "black", lty = 2, lwd = 2)
        legend("topright", legend = paste("Median:", round(median(avg_expr_raw), 2)),
               lty = 2, lwd = 2, col = "black", bty = "n")
    }
} else {
    cat("Raw counts not available for histogram.\n")
}
```

```{r legend-hist-before, results='asis', echo=FALSE}
figure_legend("Distribution of mean log2(CPM + 1) expression per gene before low-expression filtering. The prominent left-hand peak represents lowly/non-expressed genes (noise) that will be removed by filtering. Dashed line indicates the median expression.")
```

#### After Filtering

```{r filter-hist-after, fig.height=5, fig.width=8}
if (!is.null(filt_counts)) {
    num_cols_filt <- sapply(filt_counts, is.numeric)
    if (any(num_cols_filt)) {
        filt_mat <- as.matrix(filt_counts[, num_cols_filt, drop = FALSE])
        filt_mat[is.na(filt_mat)] <- 0
        lib_sizes_f <- colSums(filt_mat, na.rm = TRUE)
        filt_cpm <- t(t(filt_mat) / lib_sizes_f * 1e6)
        filt_log2cpm <- log2(filt_cpm + 1)
        avg_expr_filt <- rowMeans(filt_log2cpm, na.rm = TRUE)

        hist(avg_expr_filt, breaks = 80, col = "#377EB888", border = "white",
             main = paste0("After Filtering (", format(length(avg_expr_filt), big.mark = ","), " genes)"),
             xlab = "log2(CPM + 1)", ylab = "Number of genes", cex.main = 1.2)
        abline(v = median(avg_expr_filt), col = "black", lty = 2, lwd = 2)
        legend("topright", legend = paste("Median:", round(median(avg_expr_filt), 2)),
               lty = 2, lwd = 2, col = "black", bty = "n")
    }
} else {
    cat("Filtered counts not available for histogram.\n")
}
```

```{r legend-hist-after, results='asis', echo=FALSE}
figure_legend("Distribution of mean log2(CPM + 1) expression per gene after low-expression filtering. The left-hand noise peak has been removed, leaving a more symmetric distribution of expressed genes suitable for differential expression analysis.")
```

## Expression Distributions {.tabset}

**Normalization method: TMMlogCPM**

TMMlogCPM is a two-step normalization approach commonly used in RNA-seq analysis:

1. **TMM (Trimmed Mean of M-values)**: A normalization method from the edgeR package that calculates scaling factors to account for differences in library composition between samples. TMM trims the most extreme log-fold-changes and absolute expression values before computing a weighted mean, making it robust to outliers and genes with very high/low expression.

2. **logCPM (log2 Counts Per Million)**: After TMM scaling, counts are converted to CPM (counts per million) to normalize for sequencing depth, then log2-transformed to stabilize variance and make the data more normally distributed.

The resulting **TMMlogCPM** values are suitable for visualization, clustering, and downstream analysis, providing comparable expression levels across samples regardless of library size or composition differences.

### Before Normalization

```{r boxplot-raw, fig.cap="Per-sample expression boxplot — before normalization (log2 raw counts + 1)"}
bp_raw <- file.path(diag_dir, "raw_boxplot.png")
if (file.exists(bp_raw)) {
    knitr::include_graphics(bp_raw)
} else {
    cat("Raw counts boxplot not available.")
}
```

```{r legend-raw-boxplot, results='asis', echo=FALSE}
figure_legend("Per-sample distribution of raw expression values (log2 counts + 1). Differences in box heights reflect variation in library sizes and composition before normalization. Boxes show the interquartile range (IQR), with whiskers extending to 1.5x IQR.")
```

```{r commentary-raw-boxplot, results='asis'}
render_commentary("raw_boxplot")
```

### After Normalization

```{r boxplot, fig.cap="Per-sample normalized expression boxplot (TMMlogCPM)"}
bp <- file.path(diag_dir, "norm_boxplot.png")
if (file.exists(bp)) knitr::include_graphics(bp)
```

```{r legend-norm-boxplot, results='asis', echo=FALSE}
figure_legend("Per-sample distribution of TMMlogCPM-normalized expression values. Alignment of median lines and box heights indicates successful normalization for library size and composition differences.")
```

```{r commentary-boxplot, results='asis'}
render_commentary("norm_boxplot")
```

### Density Plot

```{r histograms, fig.cap="Expression distribution histograms"}
hp <- file.path(diag_dir, "rna_histograms_summary.png")
if (file.exists(hp)) knitr::include_graphics(hp)
```

```{r legend-density, results='asis', echo=FALSE}
figure_legend("Kernel density estimates of per-sample expression distributions. Overlapping curves indicate consistent expression profiles across samples after normalization. Deviating samples may warrant further investigation.")
```

```{r commentary-density, results='asis'}
render_commentary("density_plot")
```

## Principal Component Analysis (PCA) {.tabset}

PCA was performed on normalized log-CPM values to assess overall sample
grouping and detect potential outliers or batch effects.

Principal Component Analysis (PCA) reduces high-dimensional expression data to a low-dimensional representation that captures the largest sources of variation. Samples that cluster together share similar expression profiles; separation along PC1 indicates the dominant source of biological or technical variation.

### By Top Variable Genes

Select the number of top variable genes used for PCA calculation:

```{r pca-var-genes-setup, echo=FALSE}
# Find all PCA files with top N variable genes pattern
pca_files <- list.files(diag_dir, pattern = "^PCA_top[0-9]+\\.png$", full.names = TRUE)

# Build a map of n_top -> image data URI
pca_data_map <- list()
available_n_top <- c()

if (length(pca_files) > 0) {
    for (f in pca_files) {
        n_val <- as.numeric(sub(".*PCA_top([0-9]+)\\.png$", "\\1", basename(f)))
        if (file.exists(f)) {
            pca_data_map[[as.character(n_val)]] <- knitr::image_uri(f)
            available_n_top <- c(available_n_top, n_val)
        }
    }
}

# Sort available options
if (length(available_n_top) > 0) {
    available_n_top <- sort(unique(available_n_top))
    default_n_top <- available_n_top[1]
    default_img_uri <- pca_data_map[[as.character(default_n_top)]]
} else {
    # Fallback: use standard PCA plot
    default_pca_path <- file.path(diag_dir, "PCA_PC1.vs.PC2.png")
    if (file.exists(default_pca_path)) {
        default_n_top <- 0
        default_img_uri <- knitr::image_uri(default_pca_path)
        pca_data_map[["0"]] <- default_img_uri
        available_n_top <- 0
    } else {
        default_n_top <- NULL
        default_img_uri <- NULL
    }
}
```

<div style="margin-bottom: 20px;">
<label for="pca_selector"><strong>Number of top variable genes:</strong></label>
<select id="pca_selector" onchange="updatePCAPlot(this.value)" style="padding: 8px; border-radius: 4px; border: 1px solid #ccc; font-size: 14px;">
```{r pca-dropdown, echo=FALSE, results='asis'}
if (length(available_n_top) > 0 && available_n_top[1] != 0) {
    for (n_val in available_n_top) {
        selected <- if (n_val == default_n_top) " selected" else ""
        cat(sprintf('<option value="%d"%s>%s genes</option>\n', n_val, selected, format(n_val, big.mark = ",")))
    }
} else if (!is.null(default_img_uri)) {
    cat('<option value="0" selected>All genes (default)</option>\n')
} else {
    cat('<option disabled>No PCA plots found</option>\n')
}
```
</select>
</div>

<div id="pca_plot_container" style="text-align: center;">
```{r pca-render, echo=FALSE, results='asis'}
if (!is.null(default_img_uri)) {
    cat(sprintf('<img id="pca_main_image" src="%s" style="width:90%%; max-width: 900px; height: auto;" alt="PCA Plot" />', default_img_uri))
} else {
    cat('<p><em>PCA plot not available.</em></p>')
}
```
</div>

<script>
var embeddedPcaImages = `r if (length(pca_data_map) > 0) jsonlite::toJSON(pca_data_map, auto_unbox = TRUE) else "{}"`;

function updatePCAPlot(n_top) {
    var img = document.getElementById("pca_main_image");
    if (embeddedPcaImages && embeddedPcaImages[n_top]) {
        img.src = embeddedPcaImages[n_top];
    }
}
</script>

### PC1 vs PC2

```{r pca-12, fig.cap="PCA — PC1 vs PC2"}
p <- file.path(diag_dir, "PCA_PC1.vs.PC2.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-pca-12, results='asis', echo=FALSE}
figure_legend("Principal Component Analysis (PCA) of normalized expression data, showing sample separation along PC1 and PC2. Each point represents one sample, colored by experimental group. The percentage of variance explained by each component is indicated on the axes. Samples clustering by group suggest a strong biological signal relative to technical noise.")
```

```{r commentary-pca, results='asis'}
render_commentary("pca_plot")
```

### PC1 vs PC3

```{r pca-13, fig.cap="PCA — PC1 vs PC3"}
p <- file.path(diag_dir, "PCA_PC1.vs.PC3.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-pca-13, results='asis', echo=FALSE}
figure_legend("PCA plot showing sample separation along PC1 and PC3. PC3 may capture additional biological variation or batch effects not visible in the PC1-PC2 projection.")
```

```{r commentary-pca-13, results='asis'}
render_commentary("pca_plot_13")
```

### 3D Interactive

An interactive 3D PCA plot is available:

```{r pca-3d-link, results='asis'}
pca3d_file <- file.path(diag_dir, "PCA_3D.html")
if (file.exists(pca3d_file)) {
    cat(sprintf('<p><a href="rna/Diagnostic_plots/PCA_3D.html" target="_blank" style="font-size: 1.1em;">🔗 Open 3D PCA (new tab)</a></p>'))
} else {
    cat("*3D PCA plot not available.*")
}
```

```{r legend-pca-3d, results='asis', echo=FALSE}
if (file.exists(file.path(diag_dir, "PCA_3D.html"))) {
    figure_legend("Interactive 3D PCA plot showing sample separation along PC1, PC2, and PC3. Rotate the plot to explore sample clustering in three dimensions.")
}
```

## Sample Similarity {.tabset}

Pairwise sample correlation and distance metrics assess overall reproducibility within and between experimental groups. High within-group correlation and clear between-group separation indicate good data quality and a detectable biological signal.

### Correlation Heatmap

```{r corr-heatmap, fig.cap="Sample-to-sample correlation heatmap"}
p <- file.path(diag_dir, "sample_correlation_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-corr-heatmap, results='asis', echo=FALSE}
figure_legend("Pairwise Pearson correlation heatmap of normalized expression profiles. Higher correlation values (warmer colors) indicate greater transcriptomic similarity between samples. Within-group correlations should be higher than between-group correlations.")
```

```{r commentary-correlation, results='asis'}
render_commentary("sample_correlation")
```

### Distance Heatmap

```{r dist-heatmap, fig.cap="Sample-to-sample distance heatmap"}
p <- file.path(diag_dir, "sample_distance_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-dist-heatmap, results='asis', echo=FALSE}
figure_legend("Euclidean distance matrix of normalized expression values. Darker colors indicate smaller distances (higher similarity). Samples from the same experimental group are expected to cluster together with smaller inter-sample distances.")
```

```{r commentary-distance, results='asis'}
render_commentary("sample_distance")
```

## Expression Heatmaps {.tabset}

Expression heatmaps display the relative abundance of features across samples, with hierarchical clustering grouping features and/or samples by expression similarity. They reveal global expression patterns, condition-specific signatures, and potential outlier samples.

### With Clustering

```{r expr-heatmap, fig.cap="Sample expression heatmap (subsampled to top variable genes)"}
p <- file.path(diag_dir, "samples_rna_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-expr-heatmap, results='asis', echo=FALSE}
figure_legend("Unsupervised hierarchical clustering heatmap based on the top most variable genes. Both rows (genes) and columns (samples) are clustered by expression similarity. Color scale represents z-score normalized expression (row-scaled).")
```

```{r commentary-expr-heatmap, results='asis'}
render_commentary("expr_heatmap")
```

### Without Column Clustering

```{r expr-heatmap-nocol, fig.cap="Expression heatmap — without column clustering"}
p <- file.path(diag_dir, "samples_rna_heatmap_wo_col.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-expr-heatmap-nocol, results='asis', echo=FALSE}
figure_legend("Expression heatmap with hierarchical row clustering but without column reordering, preserving the original sample order for comparison with the clustered version above.")
```

---

# Differential Expression Analysis

```{r cutoff-panel, results='asis', echo=FALSE}
if (exists("cutoff_panel_html")) {
    .lfc_default <- log2(rna_cfg$de$linear_fc_cutoff %||% 1.5)
    .p_default   <- rna_cfg$de$p_cutoff %||% 0.05
    cat(cutoff_panel_html(.lfc_default, .p_default, default_ptype = "padj"))
}
```

Differential expression was performed using **DESeq2** (`r rna_cfg$de$deseq_mode` mode).
Genes were considered significant at adjusted p-value $\leq$ `r rna_cfg$de$p_cutoff`
and |linear fold change| $\geq$ `r rna_cfg$de$linear_fc_cutoff`.

## DE Summary

Differential expression analysis identifies features with statistically significant abundance changes between experimental conditions. Results are corrected for multiple testing to control the false discovery rate.

```{r de-summary, results='asis'}
de_counts_file <- file.path(datasets_dir, "de_summary_counts.tsv")
if (file.exists(de_counts_file)) {
    de_counts <- read.delim(de_counts_file, stringsAsFactors = FALSE)
    if (exists("cutoff_summary_span") && nrow(de_counts) > 0) {
        # Detect column names for Up/Down counts
        up_col   <- intersect(c("Up", "up", "n_up"), names(de_counts))[1]
        down_col <- intersect(c("Down", "down", "n_down"), names(de_counts))[1]
        ctr_col  <- intersect(c("Contrast", "contrast"), names(de_counts))[1]
        if (!is.na(up_col) && !is.na(down_col) && !is.na(ctr_col)) {
            cat("<div class='cutoff-de-summary' style='margin-bottom:12px;'>\n")
            for (row_i in seq_len(nrow(de_counts))) {
                cat(sprintf("<div style='margin:3px 0;'><strong>%s:</strong> %s</div>\n",
                    de_counts[[ctr_col]][row_i],
                    cutoff_summary_span(de_counts[[ctr_col]][row_i],
                                        de_counts[[up_col]][row_i],
                                        de_counts[[down_col]][row_i], "genes")))
            }
            cat("</div>\n")
        }
    }
    make_dt(de_counts, "DE Gene Counts per Contrast")
} else {
    cat("DE summary counts file not found.")
}
```

**Column definitions in DE results tables:**

- **`[contrast]_pass`**: Whether the gene is significantly DE in that specific contrast (TRUE/FALSE), based on the p-value and fold-change cutoffs defined above.

## DE Gene Overlap {.tabset}

### All DE Genes

```{r de-overlap-venn-all, fig.height=6, fig.width=10, results='asis'}
# Create Venn diagram of DE genes across contrasts
if (!is.null(de_data) && length(contrast_names) >= 2) {
    # Get DE genes for each contrast
    padj_cut <- rna_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- log2(rna_cfg$de$linear_fc_cutoff %||% 1.5)

    de_gene_lists <- list()
    for (cn in contrast_names) {
        padj_col <- grep(paste0("padj\\.", cn), names(de_data), value = TRUE)[1]
        lfc_col <- grep(paste0("(log2FoldChange|linearFC)\\.", cn), names(de_data), value = TRUE)[1]
        if (!is.na(padj_col) && !is.na(lfc_col)) {
            padj_vals <- as.numeric(de_data[[padj_col]])
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            if (grepl("^linearFC", lfc_col)) lfc_vals <- log2(abs(lfc_vals)) * sign(lfc_vals)
            is_de <- !is.na(padj_vals) & padj_vals <= padj_cut & abs(lfc_vals) >= lfc_cut
            gene_col <- if ("GeneName" %in% names(de_data)) "GeneName" else names(de_data)[1]
            de_gene_lists[[gsub("_", " ", cn)]] <- de_data[[gene_col]][is_de]
        }
    }

    if (length(de_gene_lists) >= 2 && length(de_gene_lists) <= 7) {
        # Use ggVennDiagram if available
        if (has_ggVennDiagram) {
            total_unique <- length(unique(unlist(de_gene_lists)))
            # Shorten category names for cleaner display
            short_names <- gsub(" vs ", "\nvs ", names(de_gene_lists))
            names(de_gene_lists) <- short_names
            venn_plot <- ggVennDiagram(de_gene_lists, label_alpha = 0,
                                       label = "both", label_percent_digit = 1,
                                       set_size = 2.8) +
                scale_fill_gradient(low = "white", high = "steelblue") +
                scale_x_continuous(expand = expansion(mult = 0.4)) +
                labs(title = "DE Gene Overlap (All)",
                     subtitle = paste0("Total unique DE genes: ", total_unique)) +
                theme(legend.position = "none",
                      plot.title = element_text(hjust = 0),
                      plot.subtitle = element_text(hjust = 0),
                      plot.margin = margin(10, 60, 10, 60))
            print(venn_plot)
        } else {
            # Fallback: simple overlap table
            cat("**DE gene counts per contrast:**\n\n")
            for (nm in names(de_gene_lists)) {
                cat(sprintf("- **%s**: %d genes\n", nm, length(de_gene_lists[[nm]])))
            }
            # Calculate pairwise overlaps
            if (length(de_gene_lists) == 2) {
                overlap <- length(intersect(de_gene_lists[[1]], de_gene_lists[[2]]))
                cat(sprintf("\n- **Shared**: %d genes\n", overlap))
            }
        }
    } else if (length(de_gene_lists) > 7) {
        cat("Too many contrasts for Venn diagram (max 7). Showing summary:\n\n")
        for (nm in names(de_gene_lists)) {
            cat(sprintf("- **%s**: %d genes\n", nm, length(de_gene_lists[[nm]])))
        }
    }
} else if (length(contrast_names) < 2) {
    cat("Venn diagram requires at least 2 contrasts.\n")
} else {
    cat("DE data not available for Venn diagram.\n")
}
```

### Up-regulated

```{r de-overlap-venn-up, fig.height=6, fig.width=10, results='asis'}
if (!is.null(de_data) && length(contrast_names) >= 2) {
    padj_cut <- rna_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- log2(rna_cfg$de$linear_fc_cutoff %||% 1.5)

    up_gene_lists <- list()
    for (cn in contrast_names) {
        padj_col <- grep(paste0("padj\\.", cn), names(de_data), value = TRUE)[1]
        lfc_col <- grep(paste0("(log2FoldChange|linearFC)\\.", cn), names(de_data), value = TRUE)[1]
        if (!is.na(padj_col) && !is.na(lfc_col)) {
            padj_vals <- as.numeric(de_data[[padj_col]])
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            if (grepl("^linearFC", lfc_col)) lfc_vals <- log2(abs(lfc_vals)) * sign(lfc_vals)
            is_up <- !is.na(padj_vals) & padj_vals <= padj_cut & lfc_vals >= lfc_cut
            gene_col <- if ("GeneName" %in% names(de_data)) "GeneName" else names(de_data)[1]
            up_gene_lists[[gsub("_", " ", cn)]] <- de_data[[gene_col]][is_up]
        }
    }

    if (length(up_gene_lists) >= 2 && length(up_gene_lists) <= 7) {
        if (has_ggVennDiagram) {
            total_up <- length(unique(unlist(up_gene_lists)))
            short_names_up <- gsub(" vs ", "\nvs ", names(up_gene_lists))
            names(up_gene_lists) <- short_names_up
            venn_plot <- ggVennDiagram(up_gene_lists, label_alpha = 0,
                                       label = "both", label_percent_digit = 1,
                                       set_size = 2.8) +
                scale_fill_gradient(low = "white", high = "#E41A1C") +
                scale_x_continuous(expand = expansion(mult = 0.4)) +
                labs(title = "Up-regulated Gene Overlap",
                     subtitle = paste0("Total unique up-regulated genes: ", total_up)) +
                theme(legend.position = "none",
                      plot.title = element_text(hjust = 0),
                      plot.subtitle = element_text(hjust = 0),
                      plot.margin = margin(10, 60, 10, 60))
            print(venn_plot)
        } else {
            cat("**Up-regulated gene counts:**\n\n")
            for (nm in names(up_gene_lists)) {
                cat(sprintf("- **%s**: %d genes\n", nm, length(up_gene_lists[[nm]])))
            }
        }
    }
}
```

### Down-regulated

```{r de-overlap-venn-down, fig.height=6, fig.width=10, results='asis'}
if (!is.null(de_data) && length(contrast_names) >= 2) {
    padj_cut <- rna_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- log2(rna_cfg$de$linear_fc_cutoff %||% 1.5)

    down_gene_lists <- list()
    for (cn in contrast_names) {
        padj_col <- grep(paste0("padj\\.", cn), names(de_data), value = TRUE)[1]
        lfc_col <- grep(paste0("(log2FoldChange|linearFC)\\.", cn), names(de_data), value = TRUE)[1]
        if (!is.na(padj_col) && !is.na(lfc_col)) {
            padj_vals <- as.numeric(de_data[[padj_col]])
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            if (grepl("^linearFC", lfc_col)) lfc_vals <- log2(abs(lfc_vals)) * sign(lfc_vals)
            is_down <- !is.na(padj_vals) & padj_vals <= padj_cut & lfc_vals <= -lfc_cut
            gene_col <- if ("GeneName" %in% names(de_data)) "GeneName" else names(de_data)[1]
            down_gene_lists[[gsub("_", " ", cn)]] <- de_data[[gene_col]][is_down]
        }
    }

    if (length(down_gene_lists) >= 2 && length(down_gene_lists) <= 7) {
        if (has_ggVennDiagram) {
            total_down <- length(unique(unlist(down_gene_lists)))
            short_names_dn <- gsub(" vs ", "\nvs ", names(down_gene_lists))
            names(down_gene_lists) <- short_names_dn
            venn_plot <- ggVennDiagram(down_gene_lists, label_alpha = 0,
                                       label = "both", label_percent_digit = 1,
                                       set_size = 2.8) +
                scale_fill_gradient(low = "white", high = "#377EB8") +
                scale_x_continuous(expand = expansion(mult = 0.4)) +
                labs(title = "Down-regulated Gene Overlap",
                     subtitle = paste0("Total unique down-regulated genes: ", total_down)) +
                theme(legend.position = "none",
                      plot.title = element_text(hjust = 0),
                      plot.subtitle = element_text(hjust = 0),
                      plot.margin = margin(10, 60, 10, 60))
            print(venn_plot)
        } else {
            cat("**Down-regulated gene counts:**\n\n")
            for (nm in names(down_gene_lists)) {
                cat(sprintf("- **%s**: %d genes\n", nm, length(down_gene_lists[[nm]])))
            }
        }
    }
}
```

## Volcano Plots {.tabset}

Volcano plots display the relationship between statistical significance (-log10 adjusted p-value) and biological effect size (log2 fold-change) for each feature. Points in the upper corners represent features that are both statistically significant and have large fold changes.

```{r volcano-data-prep, include=FALSE}
# Load DE data for interactive plots
final_results_file <- file.path(datasets_dir, "final_results.tsv")
de_data <- NULL
if (file.exists(final_results_file)) {
    de_data <- read.delim(final_results_file, stringsAsFactors = FALSE, check.names = FALSE)
    # Add gene names from annotation if available
    anno_file_vol <- file.path(enrich_dir, "gene_annotation.csv")
    if (file.exists(anno_file_vol)) {
        anno_vol <- read.csv(anno_file_vol, stringsAsFactors = FALSE)
        id_col_vol <- if ("Gene" %in% names(de_data)) "Gene" else if ("FeatureID" %in% names(de_data)) "FeatureID" else NULL
        if (!is.null(id_col_vol)) {
            de_data$gene_id_clean <- gsub("^Gene:", "", de_data[[id_col_vol]])
            anno_match_vol <- anno_vol[match(de_data$gene_id_clean, anno_vol$gene_id), ]
            de_data$GeneName <- ifelse(
                !is.na(anno_match_vol$symbol) & anno_match_vol$symbol != anno_match_vol$gene_id,
                anno_match_vol$symbol,
                de_data$gene_id_clean
            )
        }
    }
}

# Create interactive volcano plot
make_volcano_plotly <- function(df, contrast, padj_cut = 0.05, lfc_cut = log2(1.5), title = NULL) {
    lfc_col <- grep(paste0("(log2FoldChange|linearFC)\\.", contrast), names(df), value = TRUE)[1]
    padj_col <- grep(paste0("padj\\.", contrast), names(df), value = TRUE)[1]
    if (is.na(lfc_col) || is.na(padj_col)) return(NULL)

    plot_df <- data.frame(
        gene = if ("GeneName" %in% names(df)) df$GeneName else df[[1]],
        logFC = as.numeric(df[[lfc_col]]),
        padj = as.numeric(df[[padj_col]]),
        stringsAsFactors = FALSE
    )
    if (grepl("^linearFC", lfc_col)) {
        plot_df$logFC <- log2(abs(plot_df$logFC)) * sign(plot_df$logFC)
    }
    raw_nlogp <- -log10(pmax(plot_df$padj, 1e-300))
    ycap_v <- max(quantile(raw_nlogp[is.finite(raw_nlogp)], 0.995, na.rm = TRUE), 10)
    plot_df$neglog10padj <- pmin(raw_nlogp, ycap_v)
    plot_df$Direction <- ifelse(
        !is.na(plot_df$padj) & plot_df$padj <= padj_cut & abs(plot_df$logFC) >= lfc_cut,
        ifelse(plot_df$logFC > 0, "Up", "Down"), "NS"
    )
    plot_df$Direction <- factor(plot_df$Direction, levels = c("NS", "Down", "Up"))

    p <- ggplot(plot_df, aes(x = logFC, y = neglog10padj, color = Direction,
                              text = paste0("Gene: ", gene, "\nlog2FC: ", round(logFC, 3),
                                           "\npadj: ", signif(padj, 3)))) +
        geom_point(alpha = 0.6, size = 1.5) +
        scale_color_manual(values = c("NS" = "grey70", "Down" = "blue", "Up" = "red")) +
        geom_vline(xintercept = c(-lfc_cut, lfc_cut), linetype = "dashed") +
        geom_hline(yintercept = -log10(padj_cut), linetype = "dashed") +
        labs(title = title, x = "log2 Fold Change", y = "-log10(adj.P.Val)") +
        theme_minimal()
    widget <- ggplotly(p, tooltip = "text")
    # Include both raw and adjusted p-values for cutoff panel
    pval_col <- grep(paste0("pvalue\\.", contrast), names(df), value = TRUE)[1]
    raw_pval <- if (!is.na(pval_col)) as.numeric(df[[pval_col]]) else plot_df$padj
    reg_df <- data.frame(name = as.character(plot_df$gene),
                         logFC = plot_df$logFC,
                         pval = raw_pval,
                         adjPval = plot_df$padj,
                         stringsAsFactors = FALSE)
    list(widget = widget, plot_df = reg_df, ycap = ycap_v)
}

volcano_files <- list.files(qc_post_dir, pattern = "^volcano_.*\\.png$", full.names = TRUE)
contrast_names <- gsub("^volcano_|\\.png$", "", basename(volcano_files))
```

### Static {.tabset}

```{r volcano-static, results='asis'}
padj_cut_vol <- rna_cfg$de$p_cutoff %||% 0.05
lfc_cut_vol <- rna_cfg$de$linear_fc_cutoff %||% 1.5
if (length(volcano_files) > 0) {
    for (i in seq_along(volcano_files)) {
        vf <- volcano_files[i]
        cn <- contrast_names[i]
        contrast_clean <- gsub("_", " ", cn)
        cat(sprintf("\n#### %s\n\n", contrast_clean))
        cat(sprintf("![Volcano — %s](%s)\n\n", contrast_clean, vf))
        figure_legend(sprintf("Volcano plot for the %s contrast. Each point represents a gene; the x-axis shows log2 fold change and the y-axis shows -log10 adjusted p-value. Red: significantly up-regulated; blue: significantly down-regulated (adj. p &le; %s, |FC| &ge; %s). Dashed lines indicate significance thresholds.", contrast_clean, padj_cut_vol, lfc_cut_vol))
        render_commentary(paste0("volcano_", cn))
    }
} else {
    cat("No volcano plots found.\n")
}
```

### Interactive {.tabset}

Hover over points to see gene names and statistics. Enter gene names to highlight them on the plot.

```{r volcano-interactive, results='asis', fig.height=6}
if (!is.null(de_data) && length(contrast_names) > 0) {
    for (i in seq_along(contrast_names)) {
        cn <- contrast_names[i]
        contrast_clean <- gsub("_", " ", cn)
        clean_id <- gsub("[^a-zA-Z0-9]", "_", cn)

        cat(sprintf("\n#### %s\n\n", contrast_clean))

        # Add gene highlighting controls with autocomplete
        cat(sprintf('
<div style="margin-bottom: 15px; position: relative;">
<label for="volcano_genes_%s"><strong>Highlight genes:</strong></label>
<div style="display: inline-block; position: relative; margin-left: 10px;">
<input type="text" id="volcano_genes_%s" placeholder="Start typing a gene name..." autocomplete="off"
  style="padding: 8px; border-radius: 4px; border: 1px solid #ccc; width: 350px;"
  oninput="volcanoAutocomplete_%s(this)" onkeydown="volcanoAutoKey_%s(event)">
<div id="volcano_ac_%s" style="position: absolute; top: 100%%; left: 0; width: 350px; max-height: 200px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 4px 4px; z-index: 1000; display: none; box-shadow: 0 4px 6px rgba(0,0,0,0.1);"></div>
</div>
<button onclick="highlightVolcanoGenes_%s()" style="padding: 8px 15px; margin-left: 5px; background: #007bff; color: white; border: none; border-radius: 4px; cursor: pointer;">Highlight</button>
<button onclick="clearVolcanoHighlight_%s()" style="padding: 8px 15px; margin-left: 5px; background: #6c757d; color: white; border: none; border-radius: 4px; cursor: pointer;">Clear</button>
<div id="volcano_tags_%s" style="margin-top: 5px; display: flex; flex-wrap: wrap; gap: 4px;"></div>
</div>
', clean_id, clean_id, clean_id, clean_id, clean_id, clean_id, clean_id, clean_id))

        vol_res <- make_volcano_plotly(de_data, cn, rna_cfg$de$p_cutoff %||% 0.05,
                                  log2(rna_cfg$de$linear_fc_cutoff %||% 1.5), contrast_clean)
        if (!is.null(vol_res)) {
            p <- vol_res$widget
            p$elementId <- paste0("volcano_plot_", clean_id)
            print(htmltools::tagList(p))
            if (exists("cutoff_register_plot")) {
                cat(cutoff_register_plot(
                    paste0("volcano_plot_", clean_id),
                    vol_res$plot_df, "volcano", "Gene", cn,
                    ycap = vol_res$ycap
                ))
            }

            # Add JavaScript for autocomplete + gene highlighting
            cat(sprintf('
<script>
(function() {
    var cid = "%s";
    var selectedGenes = [];
    var acIndex = -1;

    // Extract all gene names from the plotly data
    function getGeneNames() {
        var plotDiv = document.getElementById("volcano_plot_" + cid);
        if (!plotDiv || !plotDiv.data) return [];
        var names = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var data = plotDiv.data[t];
            if (!data.text) continue;
            for (var i = 0; i < data.text.length; i++) {
                var parts = data.text[i].split(/<br\\s*\\/?>/i);
                if (parts.length > 0) {
                    var g = parts[0].replace(/Gene:\\s*/i, "").trim();
                    if (g && names.indexOf(g) === -1) names.push(g);
                }
            }
        }
        return names.sort();
    }

    // Autocomplete
    window.volcanoAutocomplete_%s = function(el) {
        var acDiv = document.getElementById("volcano_ac_" + cid);
        var val = el.value.trim().toLowerCase();
        acDiv.innerHTML = "";
        acIndex = -1;
        if (val.length < 1) { acDiv.style.display = "none"; return; }
        // Get text after last comma
        var parts = val.split(",");
        var current = parts[parts.length - 1].trim();
        if (current.length < 1) { acDiv.style.display = "none"; return; }
        var allGenes = getGeneNames();
        var matches = allGenes.filter(function(g) {
            return g.toLowerCase().indexOf(current) !== -1 && selectedGenes.indexOf(g) === -1;
        }).slice(0, 15);
        if (matches.length === 0) { acDiv.style.display = "none"; return; }
        matches.forEach(function(g, idx) {
            var item = document.createElement("div");
            item.textContent = g;
            item.style.cssText = "padding: 6px 10px; cursor: pointer; font-size: 13px;";
            item.onmouseenter = function() { this.style.background = "#e9ecef"; };
            item.onmouseleave = function() { this.style.background = "white"; };
            item.onclick = function() { selectGene(g); };
            acDiv.appendChild(item);
        });
        acDiv.style.display = "block";
    };

    // Keyboard nav for autocomplete
    window.volcanoAutoKey_%s = function(e) {
        var acDiv = document.getElementById("volcano_ac_" + cid);
        var items = acDiv.children;
        if (items.length === 0) return;
        if (e.key === "ArrowDown") {
            e.preventDefault();
            acIndex = Math.min(acIndex + 1, items.length - 1);
            updateAcHighlight(items);
        } else if (e.key === "ArrowUp") {
            e.preventDefault();
            acIndex = Math.max(acIndex - 1, 0);
            updateAcHighlight(items);
        } else if (e.key === "Enter" && acIndex >= 0) {
            e.preventDefault();
            items[acIndex].click();
        }
    };

    function updateAcHighlight(items) {
        for (var i = 0; i < items.length; i++) {
            items[i].style.background = i === acIndex ? "#e9ecef" : "white";
        }
    }

    function selectGene(gene) {
        if (selectedGenes.indexOf(gene) === -1) selectedGenes.push(gene);
        document.getElementById("volcano_genes_" + cid).value = "";
        document.getElementById("volcano_ac_" + cid).style.display = "none";
        renderTags();
        applyHighlight();
    }

    function renderTags() {
        var container = document.getElementById("volcano_tags_" + cid);
        container.innerHTML = "";
        selectedGenes.forEach(function(g) {
            var tag = document.createElement("span");
            tag.style.cssText = "display: inline-flex; align-items: center; background: #007bff; color: white; padding: 2px 8px; border-radius: 12px; font-size: 12px; gap: 4px;";
            tag.innerHTML = g + " <span style=\\"cursor:pointer; font-weight:bold; margin-left:3px;\\" onclick=\\"removeVolcanoGene_" + cid + "(\'" + g + "\')\\">×</span>";
            container.appendChild(tag);
        });
    }

    window["removeVolcanoGene_" + cid] = function(gene) {
        selectedGenes = selectedGenes.filter(function(g) { return g !== gene; });
        renderTags();
        applyHighlight();
    };

    function applyHighlight() {
        var plotDiv = document.getElementById("volcano_plot_" + cid);
        if (!plotDiv || !plotDiv.data) return;
        if (selectedGenes.length === 0) {
            Plotly.relayout(plotDiv, { annotations: [] });
            return;
        }
        var genesLower = selectedGenes.map(function(g) { return g.toLowerCase(); });
        var annotations = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var data = plotDiv.data[t];
            if (!data.text) continue;
            for (var i = 0; i < data.text.length; i++) {
                var parts = data.text[i].split(/<br\\s*\\/?>/i);
                if (parts.length > 0) {
                    var geneName = parts[0].replace(/Gene:\\s*/i, "").trim();
                    if (genesLower.indexOf(geneName.toLowerCase()) !== -1) {
                        annotations.push({
                            x: data.x[i], y: data.y[i], text: geneName,
                            showarrow: true, arrowhead: 2, ax: 20, ay: -30,
                            font: { size: 12, color: "black" },
                            bgcolor: "rgba(255,255,255,0.9)",
                            bordercolor: "black", borderwidth: 1
                        });
                    }
                }
            }
        }
        Plotly.relayout(plotDiv, { annotations: annotations });
    }

    // Highlight button: also accepts comma-separated manual input
    window.highlightVolcanoGenes_%s = function() {
        var input = document.getElementById("volcano_genes_" + cid).value;
        var manual = input.split(",").map(function(s) { return s.trim(); }).filter(function(s) { return s.length > 0; });
        // Match manual entries to actual gene names (case-insensitive)
        var allGenes = getGeneNames();
        manual.forEach(function(m) {
            var mLower = m.toLowerCase();
            var match = allGenes.find(function(g) { return g.toLowerCase() === mLower; });
            if (match && selectedGenes.indexOf(match) === -1) selectedGenes.push(match);
        });
        document.getElementById("volcano_genes_" + cid).value = "";
        document.getElementById("volcano_ac_" + cid).style.display = "none";
        renderTags();
        applyHighlight();
    };

    // Clear button
    window.clearVolcanoHighlight_%s = function() {
        selectedGenes = [];
        document.getElementById("volcano_genes_" + cid).value = "";
        renderTags();
        var plotDiv = document.getElementById("volcano_plot_" + cid);
        if (plotDiv) Plotly.relayout(plotDiv, { annotations: [] });
    };

    // Close autocomplete on outside click
    document.addEventListener("click", function(e) {
        var acDiv = document.getElementById("volcano_ac_" + cid);
        var input = document.getElementById("volcano_genes_" + cid);
        if (acDiv && e.target !== input && !acDiv.contains(e.target)) {
            acDiv.style.display = "none";
        }
    });
})();
</script>
', clean_id, clean_id, clean_id, clean_id, clean_id))
        }
        figure_legend(sprintf("Interactive volcano plot for the %s contrast. Hover over points to see gene names, fold changes, and adjusted p-values. Use the search box to highlight specific genes of interest.", contrast_clean))
        cat("\n\n")
    }
} else {
    cat("Interactive plots require DE data.\n")
}
```

## MA Plots {.tabset}

MA plots show the relationship between average expression (A, x-axis) and log-fold-change (M, y-axis). They help assess whether fold-change estimates are independent of expression level and whether the data is properly normalized.

### Static {.tabset}

```{r ma-plots-static, results='asis'}
ma_files <- list.files(qc_post_dir, pattern = "^ma_.*\\.png$", full.names = TRUE)
if (length(ma_files) > 0) {
    for (mf in ma_files) {
        cn <- gsub("^ma_|\\.png$", "", basename(mf))
        contrast_label <- gsub("_", " ", cn)
        cat(sprintf("\n#### %s\n\n", contrast_label))
        cat(sprintf("![MA — %s](%s)\n\n", contrast_label, mf))
        figure_legend(sprintf("MA plot for the %s contrast. The x-axis shows mean expression (log10 base mean) and the y-axis shows log2 fold change. Red: significantly up-regulated; blue: significantly down-regulated. Dashed lines indicate fold-change cutoffs. Genes with higher expression have more reliable fold-change estimates.", contrast_label))
        render_commentary(paste0("ma_", cn))
    }
} else {
    cat("No MA plots found.\n")
}
```

### Interactive {.tabset}

Enter gene names to highlight them on the plot (comma-separated).

```{r ma-interactive-setup, include=FALSE}
# Function to create interactive MA plot with gene highlighting
make_ma_plotly <- function(df, contrast, padj_cut = 0.05, lfc_cut = log2(1.5), title = NULL) {
    lfc_col <- grep(paste0("(log2FoldChange|linearFC)\\.", contrast), names(df), value = TRUE)[1]
    padj_col <- grep(paste0("padj\\.", contrast), names(df), value = TRUE)[1]

    if (is.na(lfc_col) || is.na(padj_col)) return(NULL)

    # Calculate baseMean from expression columns (numeric columns that aren't stats)
    # These are sample columns with expression values
    stat_patterns <- "^(Gene|linearFC|log2FoldChange|pvalue|padj|upDown|manual_cutoffs|pass|FeatureID|GeneName)"
    expr_cols <- names(df)[!grepl(stat_patterns, names(df), ignore.case = TRUE)]
    expr_cols <- expr_cols[sapply(df[expr_cols], is.numeric)]

    if (length(expr_cols) == 0) return(NULL)

    # Calculate mean expression per gene
    baseMean <- rowMeans(as.matrix(df[, expr_cols, drop = FALSE]), na.rm = TRUE)

    plot_df <- data.frame(
        gene = if ("GeneName" %in% names(df)) df$GeneName else df[[1]],
        logFC = as.numeric(df[[lfc_col]]),
        padj = as.numeric(df[[padj_col]]),
        baseMean = baseMean,
        stringsAsFactors = FALSE
    )

    if (grepl("^linearFC", lfc_col)) {
        plot_df$logFC <- log2(abs(plot_df$logFC)) * sign(plot_df$logFC)
    }

    # Filter valid rows
    plot_df <- plot_df[!is.na(plot_df$baseMean) & plot_df$baseMean > 0, ]
    plot_df$log10baseMean <- log10(plot_df$baseMean)

    plot_df$Direction <- ifelse(
        !is.na(plot_df$padj) & plot_df$padj <= padj_cut & abs(plot_df$logFC) >= lfc_cut,
        ifelse(plot_df$logFC > 0, "Up", "Down"), "NS"
    )
    plot_df$Direction <- factor(plot_df$Direction, levels = c("NS", "Down", "Up"))

    p <- ggplot(plot_df, aes(x = log10baseMean, y = logFC, color = Direction,
                              text = paste0("Gene: ", gene, "\nlog2FC: ", round(logFC, 3),
                                           "\nbaseMean: ", round(baseMean, 1),
                                           "\npadj: ", signif(padj, 3)))) +
        geom_point(alpha = 0.6, size = 1.5) +
        scale_color_manual(values = c("NS" = "grey70", "Down" = "blue", "Up" = "red")) +
        geom_hline(yintercept = c(-lfc_cut, lfc_cut), linetype = "dashed", alpha = 0.5) +
        geom_hline(yintercept = 0, color = "black", size = 0.3) +
        labs(title = title, x = "log10(baseMean)", y = "log2 Fold Change") +
        theme_minimal()

    widget <- ggplotly(p, tooltip = "text")
    pval_col <- grep(paste0("pvalue\\.", contrast), names(df), value = TRUE)[1]
    raw_pval_all <- if (!is.na(pval_col)) as.numeric(df[[pval_col]]) else as.numeric(df[[padj_col]])
    # Filter to same rows as plot_df
    valid_rows <- !is.na(baseMean) & baseMean > 0
    raw_pval_filt <- raw_pval_all[valid_rows]
    reg_df <- data.frame(name = as.character(plot_df$gene),
                         logFC = plot_df$logFC,
                         pval = raw_pval_filt,
                         adjPval = plot_df$padj,
                         avgExpr = plot_df$log10baseMean,
                         stringsAsFactors = FALSE)
    list(widget = widget, plot_df = reg_df)
}
```

```{r ma-interactive, results='asis', fig.height=6}
if (!is.null(de_data) && length(contrast_names) > 0) {
    for (i in seq_along(contrast_names)) {
        cn <- contrast_names[i]
        contrast_clean <- gsub("_", " ", cn)
        clean_id <- gsub("[^a-zA-Z0-9]", "_", cn)

        cat(sprintf("\n#### %s\n\n", contrast_clean))

        # Create unique ID for this contrast's MA plot with autocomplete
        cat(sprintf('
<div style="margin-bottom: 15px; position: relative;">
<label for="ma_genes_%s"><strong>Highlight genes:</strong></label>
<div style="display: inline-block; position: relative; margin-left: 10px;">
<input type="text" id="ma_genes_%s" placeholder="Start typing a gene name..." autocomplete="off"
  style="padding: 8px; border-radius: 4px; border: 1px solid #ccc; width: 350px;"
  oninput="maAutocomplete_%s(this)" onkeydown="maAutoKey_%s(event)">
<div id="ma_ac_%s" style="position: absolute; top: 100%%; left: 0; width: 350px; max-height: 200px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 4px 4px; z-index: 1000; display: none; box-shadow: 0 4px 6px rgba(0,0,0,0.1);"></div>
</div>
<button onclick="highlightMAGenes_%s()" style="padding: 8px 15px; margin-left: 5px; background: #007bff; color: white; border: none; border-radius: 4px; cursor: pointer;">Highlight</button>
<button onclick="clearMAHighlight_%s()" style="padding: 8px 15px; margin-left: 5px; background: #6c757d; color: white; border: none; border-radius: 4px; cursor: pointer;">Clear</button>
<div id="ma_tags_%s" style="margin-top: 5px; display: flex; flex-wrap: wrap; gap: 4px;"></div>
</div>
', clean_id, clean_id, clean_id, clean_id, clean_id, clean_id, clean_id, clean_id))

        ma_res <- make_ma_plotly(de_data, cn, rna_cfg$de$p_cutoff %||% 0.05,
                            log2(rna_cfg$de$linear_fc_cutoff %||% 1.5), contrast_clean)
        if (!is.null(ma_res)) {
            p <- ma_res$widget
            p$elementId <- paste0("ma_plot_", clean_id)
            print(htmltools::tagList(p))
            if (exists("cutoff_register_plot")) {
                cat(cutoff_register_plot(
                    paste0("ma_plot_", clean_id),
                    ma_res$plot_df, "ma", "Gene", cn
                ))
            }

            # Add JavaScript for autocomplete + gene highlighting (MA)
            cat(sprintf('
<script>
(function() {
    var cid = "%s";
    var selectedGenes = [];
    var acIndex = -1;

    function getGeneNames() {
        var plotDiv = document.getElementById("ma_plot_" + cid);
        if (!plotDiv || !plotDiv.data) return [];
        var names = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var data = plotDiv.data[t];
            if (!data.text) continue;
            for (var i = 0; i < data.text.length; i++) {
                var parts = data.text[i].split(/<br\\s*\\/?>/i);
                if (parts.length > 0) {
                    var g = parts[0].replace(/Gene:\\s*/i, "").trim();
                    if (g && names.indexOf(g) === -1) names.push(g);
                }
            }
        }
        return names.sort();
    }

    window.maAutocomplete_%s = function(el) {
        var acDiv = document.getElementById("ma_ac_" + cid);
        var val = el.value.trim().toLowerCase();
        acDiv.innerHTML = "";
        acIndex = -1;
        if (val.length < 1) { acDiv.style.display = "none"; return; }
        var parts = val.split(",");
        var current = parts[parts.length - 1].trim();
        if (current.length < 1) { acDiv.style.display = "none"; return; }
        var allGenes = getGeneNames();
        var matches = allGenes.filter(function(g) {
            return g.toLowerCase().indexOf(current) !== -1 && selectedGenes.indexOf(g) === -1;
        }).slice(0, 15);
        if (matches.length === 0) { acDiv.style.display = "none"; return; }
        matches.forEach(function(g, idx) {
            var item = document.createElement("div");
            item.textContent = g;
            item.style.cssText = "padding: 6px 10px; cursor: pointer; font-size: 13px;";
            item.onmouseenter = function() { this.style.background = "#e9ecef"; };
            item.onmouseleave = function() { this.style.background = "white"; };
            item.onclick = function() { selectGene(g); };
            acDiv.appendChild(item);
        });
        acDiv.style.display = "block";
    };

    window.maAutoKey_%s = function(e) {
        var acDiv = document.getElementById("ma_ac_" + cid);
        var items = acDiv.children;
        if (items.length === 0) return;
        if (e.key === "ArrowDown") {
            e.preventDefault();
            acIndex = Math.min(acIndex + 1, items.length - 1);
            updateAcHighlight(items);
        } else if (e.key === "ArrowUp") {
            e.preventDefault();
            acIndex = Math.max(acIndex - 1, 0);
            updateAcHighlight(items);
        } else if (e.key === "Enter" && acIndex >= 0) {
            e.preventDefault();
            items[acIndex].click();
        }
    };

    function updateAcHighlight(items) {
        for (var i = 0; i < items.length; i++) {
            items[i].style.background = i === acIndex ? "#e9ecef" : "white";
        }
    }

    function selectGene(gene) {
        if (selectedGenes.indexOf(gene) === -1) selectedGenes.push(gene);
        document.getElementById("ma_genes_" + cid).value = "";
        document.getElementById("ma_ac_" + cid).style.display = "none";
        renderTags();
        applyHighlight();
    }

    function renderTags() {
        var container = document.getElementById("ma_tags_" + cid);
        container.innerHTML = "";
        selectedGenes.forEach(function(g) {
            var tag = document.createElement("span");
            tag.style.cssText = "display: inline-flex; align-items: center; background: #007bff; color: white; padding: 2px 8px; border-radius: 12px; font-size: 12px; gap: 4px;";
            tag.innerHTML = g + " <span style=\\"cursor:pointer; font-weight:bold; margin-left:3px;\\" onclick=\\"removeMAGene_" + cid + "(\'" + g + "\')\\">×</span>";
            container.appendChild(tag);
        });
    }

    window["removeMAGene_" + cid] = function(gene) {
        selectedGenes = selectedGenes.filter(function(g) { return g !== gene; });
        renderTags();
        applyHighlight();
    };

    function applyHighlight() {
        var plotDiv = document.getElementById("ma_plot_" + cid);
        if (!plotDiv || !plotDiv.data) return;
        if (selectedGenes.length === 0) {
            Plotly.relayout(plotDiv, { annotations: [] });
            return;
        }
        var genesLower = selectedGenes.map(function(g) { return g.toLowerCase(); });
        var annotations = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var data = plotDiv.data[t];
            if (!data.text) continue;
            for (var i = 0; i < data.text.length; i++) {
                var parts = data.text[i].split(/<br\\s*\\/?>/i);
                if (parts.length > 0) {
                    var geneName = parts[0].replace(/Gene:\\s*/i, "").trim();
                    if (genesLower.indexOf(geneName.toLowerCase()) !== -1) {
                        annotations.push({
                            x: data.x[i], y: data.y[i], text: geneName,
                            showarrow: true, arrowhead: 2, ax: 20, ay: -30,
                            font: { size: 12, color: "black" },
                            bgcolor: "rgba(255,255,255,0.9)",
                            bordercolor: "black", borderwidth: 1
                        });
                    }
                }
            }
        }
        Plotly.relayout(plotDiv, { annotations: annotations });
    }

    window.highlightMAGenes_%s = function() {
        var input = document.getElementById("ma_genes_" + cid).value;
        var manual = input.split(",").map(function(s) { return s.trim(); }).filter(function(s) { return s.length > 0; });
        var allGenes = getGeneNames();
        manual.forEach(function(m) {
            var mLower = m.toLowerCase();
            var match = allGenes.find(function(g) { return g.toLowerCase() === mLower; });
            if (match && selectedGenes.indexOf(match) === -1) selectedGenes.push(match);
        });
        document.getElementById("ma_genes_" + cid).value = "";
        document.getElementById("ma_ac_" + cid).style.display = "none";
        renderTags();
        applyHighlight();
    };

    window.clearMAHighlight_%s = function() {
        selectedGenes = [];
        document.getElementById("ma_genes_" + cid).value = "";
        renderTags();
        var plotDiv = document.getElementById("ma_plot_" + cid);
        if (plotDiv) Plotly.relayout(plotDiv, { annotations: [] });
    };

    document.addEventListener("click", function(e) {
        var acDiv = document.getElementById("ma_ac_" + cid);
        var input = document.getElementById("ma_genes_" + cid);
        if (acDiv && e.target !== input && !acDiv.contains(e.target)) {
            acDiv.style.display = "none";
        }
    });
})();
</script>
', clean_id, clean_id, clean_id, clean_id, clean_id))
        }
        figure_legend(sprintf("Interactive MA plot for the %s contrast. Hover over points to see gene names, fold changes, and mean expression. Use the search box to highlight specific genes of interest.", contrast_clean))
        cat("\n\n")
    }
} else {
    cat("Interactive MA plots require DE data.\n")
}
```

## Top DE Genes

```{r de-table-prep, include=FALSE}
de_summary_file <- file.path(datasets_dir, "deseq2_summary_p0.05.tsv")
anno_file <- file.path(enrich_dir, "gene_annotation.csv")
de_summary_full <- NULL

if (file.exists(de_summary_file)) {
    de_summary_full <- read.delim(de_summary_file, stringsAsFactors = FALSE, check.names = FALSE)

    # Try to add gene names from annotation file
    if (file.exists(anno_file)) {
        anno <- read.csv(anno_file, stringsAsFactors = FALSE)
        id_col <- if ("Gene" %in% names(de_summary_full)) "Gene" else if ("FeatureID" %in% names(de_summary_full)) "FeatureID" else NULL

        if (!is.null(id_col) && "gene_id" %in% names(anno)) {
            de_summary_full$gene_id_clean <- gsub("^Gene:", "", de_summary_full[[id_col]])
            anno_match <- anno[match(de_summary_full$gene_id_clean, anno$gene_id), ]
            de_summary_full$GeneName <- ifelse(
                !is.na(anno_match$symbol) & anno_match$symbol != anno_match$gene_id,
                anno_match$symbol,
                de_summary_full$gene_id_clean
            )
            de_summary_full$gene_id_clean <- NULL
            cols <- names(de_summary_full)
            id_idx <- which(cols == id_col)
            new_order <- c(cols[1:id_idx], "GeneName",
                           setdiff(cols[(id_idx + 1):length(cols)], "GeneName"))
            de_summary_full <- de_summary_full[, new_order]
        }
    }

    padj_cols <- grep("^padj\\.", names(de_summary_full), value = TRUE)
    if (length(padj_cols) > 0) {
        de_summary_full$min_padj <- apply(de_summary_full[, padj_cols, drop = FALSE], 1, min, na.rm = TRUE)
        de_summary_full <- de_summary_full[order(de_summary_full$min_padj), ]
    }
}
```

```{r de-table}
if (!is.null(de_summary_full)) {
    de_top <- head(de_summary_full, 50)
    de_top$min_padj <- NULL
    # Remove pass_any_contrast column from display
    pass_any_cols <- grep("pass_any", names(de_top), value = TRUE)
    if (length(pass_any_cols) > 0) de_top <- de_top[, !names(de_top) %in% pass_any_cols, drop = FALSE]

    DT::datatable(
      de_top,
      caption = "Top 50 differentially expressed genes (sorted by smallest adjusted p-value)",
      extensions = 'Buttons',
      options = list(
          scrollX = TRUE,
          pageLength = 15,
          dom = 'Bfrtip',
          buttons = list(
              list(extend = 'csv', filename = 'top_DE_genes'),
              list(extend = 'excel', filename = 'top_DE_genes')
          )
      ),
      rownames = FALSE
    ) |> DT::formatSignif(columns = which(sapply(de_top, is.numeric)), digits = 4)
} else {
    cat("DE summary file not found.\n")
}
```

```{r caption-de-table, results='asis', echo=FALSE}
if (!is.null(de_summary_full)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Top differentially expressed genes ranked by adjusted p-value. Fold changes and significance values are shown for each contrast. Use the search box to filter and the download buttons to export.</em></p>\n")
}
```

## Gene Lookup

Search for any gene by name or ID to see its differential expression statistics and normalized expression across all contrasts and samples.

```{r gene-lookup}
if (!is.null(de_summary_full)) {
    # Remove min_padj and pass_any_contrast columns from display
    display_cols <- setdiff(names(de_summary_full), c("min_padj", grep("pass_any", names(de_summary_full), value = TRUE)))
    lookup_df <- de_summary_full[, display_cols]

    # Merge normalized counts into DE results
    norm_counts_file_lookup <- file.path(datasets_dir, "rna_norm_TMMlogCPM.tsv")
    if (file.exists(norm_counts_file_lookup)) {
        nc <- read.delim(norm_counts_file_lookup, stringsAsFactors = FALSE, check.names = FALSE)
        nc_id_col <- names(nc)[1]
        nc[[nc_id_col]] <- sub("^Gene:", "", nc[[nc_id_col]])

        # Try merging on GeneName first, then on the first column (FeatureID)
        merged <- NULL
        if ("GeneName" %in% names(lookup_df)) {
            merged <- merge(lookup_df, nc, by.x = "GeneName", by.y = nc_id_col, all.x = TRUE)
            # If gene symbols didn't match well, try FeatureID
            sample_cols <- setdiff(names(nc), nc_id_col)
            if (sum(!is.na(merged[[sample_cols[1]]])) < nrow(merged) * 0.3) {
                merged <- merge(lookup_df, nc, by.x = names(lookup_df)[1], by.y = nc_id_col, all.x = TRUE)
            }
        } else {
            merged <- merge(lookup_df, nc, by.x = names(lookup_df)[1], by.y = nc_id_col, all.x = TRUE)
        }
        if (!is.null(merged) && nrow(merged) > 0) lookup_df <- merged
    }

    DT::datatable(
      lookup_df,
      caption = "Full DE results + normalized expression — search for any gene",
      extensions = 'Buttons',
      options = list(
          scrollX = TRUE,
          pageLength = 10,
          dom = 'Bfrtip',
          buttons = list(
              list(extend = 'csv', filename = 'DE_results_with_expression'),
              list(extend = 'excel', filename = 'DE_results_with_expression')
          ),
          search = list(regex = TRUE, caseInsensitive = TRUE)
      ),
      rownames = FALSE,
      filter = 'top'
    ) |> DT::formatSignif(columns = which(sapply(lookup_df, is.numeric)), digits = 4)
}
```

## Gene Expression Explorer

The expression explorer provides an interactive interface for examining individual gene expression patterns across experimental groups. Box plots show the distribution of expression values for each condition, enabling visual assessment of effect sizes and within-group variability.

Type a gene name and click **Add Gene** to visualize its expression across experimental groups. Add multiple genes to create a combined comparison plot. Export figures as SVG or PNG.

```{r gene-explorer-data, include=FALSE}
# Prepare data for interactive gene boxplot explorer
explorer_ready <- FALSE
explorer_json_str <- "{}"

# Load normalized expression matrix
norm_expr_file_exp <- file.path(datasets_dir, "rna_norm_TMMlogCPM.tsv")
norm_expr_exp <- NULL
if (file.exists(norm_expr_file_exp)) {
    norm_expr_exp <- read.delim(norm_expr_file_exp, stringsAsFactors = FALSE,
                                 row.names = 1, check.names = FALSE)
}

# Load metadata for group assignment
meta_file_exp <- file.path(cfg$project$dir, cfg$paths$raw, rna_cfg$files$metadata)
meta_df_exp <- if (file.exists(meta_file_exp)) read.csv(meta_file_exp, stringsAsFactors = FALSE) else NULL
group_col_exp <- rna_cfg$filtering$group_col %||% rna_cfg$effects$color

if (!is.null(norm_expr_exp) && !is.null(meta_df_exp)) {
    # Build gene display names
    gene_ids_exp <- rownames(norm_expr_exp)
    gene_display_exp <- gsub("^Gene:", "", gene_ids_exp)

    # Try annotation for gene symbols
    anno_file_exp <- file.path(enrich_dir, "gene_annotation.csv")
    if (file.exists(anno_file_exp)) {
        anno_exp <- read.csv(anno_file_exp, stringsAsFactors = FALSE)
        symbols_exp <- anno_exp$symbol[match(gene_display_exp, anno_exp$gene_id)]
        has_sym <- !is.na(symbols_exp) & symbols_exp != gene_display_exp
        gene_display_exp[has_sym] <- symbols_exp[has_sym]
    }

    # Build sample -> group mapping
    sample_col_exp <- rna_cfg$effects$samples %||% "SampleID"
    if (sample_col_exp %in% names(meta_df_exp) && !is.null(group_col_exp) && group_col_exp %in% names(meta_df_exp)) {
        group_map_exp <- setNames(as.character(meta_df_exp[[group_col_exp]]), meta_df_exp[[sample_col_exp]])
        sample_groups_exp <- as.character(group_map_exp[colnames(norm_expr_exp)])
        sample_groups_exp[is.na(sample_groups_exp)] <- "Unknown"
    } else {
        sample_groups_exp <- rep("All", ncol(norm_expr_exp))
    }

    # Build matrix as list of numeric vectors (one per gene), rounded to 3 decimals
    expr_matrix_list <- lapply(seq_len(nrow(norm_expr_exp)), function(i) {
        round(as.numeric(norm_expr_exp[i, ]), 3)
    })

    explorer_data <- list(
        genes = gene_display_exp,
        samples = colnames(norm_expr_exp),
        groups = sample_groups_exp,
        expr = expr_matrix_list
    )

    explorer_json_str <- jsonlite::toJSON(explorer_data, auto_unbox = FALSE)
    explorer_ready <- TRUE
}
```

```{r gene-explorer-ui, eval=explorer_ready, results='asis', echo=FALSE}
cat('
<div style="margin: 15px 0 20px 0;">
<div style="display: flex; align-items: center; gap: 10px; flex-wrap: wrap;">
<label for="boxplot_gene_input" style="font-weight: bold; white-space: nowrap;">Search gene:</label>
<div style="position: relative; flex: 1; max-width: 400px;">
<input type="text" id="boxplot_gene_input" placeholder="Start typing a gene name..."
  autocomplete="off"
  style="padding: 10px 12px; border-radius: 6px; border: 1px solid #ccc; width: 100%; font-size: 14px; box-sizing: border-box;"
  oninput="boxplotAutocomplete(this)" onkeydown="boxplotAutoKey(event)">
<div id="boxplot_ac" style="position: absolute; top: 100%; left: 0; width: 100%; max-height: 220px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 6px 6px; z-index: 1000; display: none; box-shadow: 0 4px 8px rgba(0,0,0,0.12);"></div>
</div>
<button onclick="addGeneToPlot()" style="padding: 10px 20px; background: #007bff; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Add Gene</button>
<button onclick="clearBoxplot()" style="padding: 10px 20px; background: #6c757d; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Clear All</button>
</div>
<div id="boxplot_gene_tags" style="margin-top: 8px; display: flex; flex-wrap: wrap; gap: 6px;"></div>
<div id="boxplot_gene_info" style="margin-top: 8px; font-size: 0.9em; color: #555;"></div>
</div>
<div id="boxplot_container" style="width: 100%; min-height: 450px; margin-bottom: 10px;"></div>
<div id="boxplot_export_bar" style="display: none; margin-bottom: 20px; gap: 8px; align-items: center;">
  <span style="font-size: 0.85em; color: #555; margin-right: 4px;">Export:</span>
  <button onclick="exportBoxplot(\'svg\')" style="padding: 5px 14px; background: #28a745; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">SVG</button>
  <button onclick="exportBoxplot(\'png\')" style="padding: 5px 14px; background: #17a2b8; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">PNG</button>
  <button onclick="exportBoxplot(\'tiff\')" style="padding: 5px 14px; background: #6f42c1; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">TIFF</button>
</div>
')
```

```{r gene-explorer-js, eval=explorer_ready, results='asis', echo=FALSE}
# Emit JSON data directly (avoid sprintf which can choke on large strings or % in data)
cat('<script>\nvar geneExprData = ')
cat(explorer_json_str)
cat(';\n')
cat(sprintf('var bpGroupLabel = "%s";\n', group_col_exp))
cat('
var boxplotAcIndex = -1;
var selectedGenes = [];  // array of {name, idx}

function boxplotAutocomplete(el) {
    var acDiv = document.getElementById("boxplot_ac");
    var val = el.value.trim().toLowerCase();
    acDiv.innerHTML = "";
    boxplotAcIndex = -1;
    if (val.length < 1) { acDiv.style.display = "none"; return; }
    var matches = [];
    for (var i = 0; i < geneExprData.genes.length; i++) {
        if (geneExprData.genes[i].toLowerCase().indexOf(val) !== -1) {
            // Skip if already selected
            if (selectedGenes.some(function(g) { return g.idx === i; })) continue;
            matches.push({name: geneExprData.genes[i], idx: i});
            if (matches.length >= 20) break;
        }
    }
    if (matches.length === 0) { acDiv.style.display = "none"; return; }
    matches.forEach(function(m) {
        var item = document.createElement("div");
        item.textContent = m.name;
        item.style.cssText = "padding: 8px 12px; cursor: pointer; font-size: 13px; border-bottom: 1px solid #f0f0f0;";
        item.onmouseenter = function() { this.style.background = "#e9ecef"; };
        item.onmouseleave = function() { this.style.background = "white"; };
        item.onclick = function() {
            document.getElementById("boxplot_gene_input").value = "";
            acDiv.style.display = "none";
            addGeneByIdx(m.idx, m.name);
        };
        acDiv.appendChild(item);
    });
    acDiv.style.display = "block";
}

function boxplotAutoKey(e) {
    var acDiv = document.getElementById("boxplot_ac");
    var items = acDiv.children;
    if (e.key === "Escape") { acDiv.style.display = "none"; return; }
    if (items.length === 0) {
        if (e.key === "Enter") { addGeneToPlot(); }
        return;
    }
    if (e.key === "ArrowDown") {
        e.preventDefault();
        boxplotAcIndex = Math.min(boxplotAcIndex + 1, items.length - 1);
        for (var i = 0; i < items.length; i++) items[i].style.background = i === boxplotAcIndex ? "#e9ecef" : "white";
    } else if (e.key === "ArrowUp") {
        e.preventDefault();
        boxplotAcIndex = Math.max(boxplotAcIndex - 1, 0);
        for (var i = 0; i < items.length; i++) items[i].style.background = i === boxplotAcIndex ? "#e9ecef" : "white";
    } else if (e.key === "Enter" && boxplotAcIndex >= 0) {
        e.preventDefault();
        items[boxplotAcIndex].click();
    }
}

function addGeneByIdx(idx, name) {
    if (selectedGenes.some(function(g) { return g.idx === idx; })) return;
    selectedGenes.push({name: name, idx: idx});
    renderGeneTags();
    renderMultiBoxplot();
}

function addGeneToPlot() {
    var val = document.getElementById("boxplot_gene_input").value.trim().toLowerCase();
    if (!val) return;
    // Exact match
    for (var i = 0; i < geneExprData.genes.length; i++) {
        if (geneExprData.genes[i].toLowerCase() === val) {
            document.getElementById("boxplot_gene_input").value = "";
            addGeneByIdx(i, geneExprData.genes[i]);
            return;
        }
    }
    // Partial match fallback
    for (var i = 0; i < geneExprData.genes.length; i++) {
        if (geneExprData.genes[i].toLowerCase().indexOf(val) !== -1) {
            document.getElementById("boxplot_gene_input").value = "";
            addGeneByIdx(i, geneExprData.genes[i]);
            return;
        }
    }
    document.getElementById("boxplot_gene_info").innerHTML = "<span style=\\"color: #c0392b;\\">Gene not found. Try a different name.</span>";
}

function removeGene(idx) {
    selectedGenes = selectedGenes.filter(function(g) { return g.idx !== idx; });
    renderGeneTags();
    if (selectedGenes.length > 0) {
        renderMultiBoxplot();
    } else {
        Plotly.purge("boxplot_container");
        document.getElementById("boxplot_gene_info").innerHTML = "";
        document.getElementById("boxplot_export_bar").style.display = "none";
    }
}

function renderGeneTags() {
    var tagsDiv = document.getElementById("boxplot_gene_tags");
    tagsDiv.innerHTML = "";
    selectedGenes.forEach(function(g) {
        var tag = document.createElement("span");
        tag.style.cssText = "display: inline-flex; align-items: center; gap: 4px; background: #e3f2fd; color: #1565c0; padding: 4px 10px; border-radius: 16px; font-size: 13px; font-weight: 500;";
        tag.innerHTML = g.name + " <span onclick=\\"removeGene(" + g.idx + ")\\" style=\\"cursor: pointer; color: #c62828; font-weight: bold; margin-left: 2px; font-size: 15px;\\">&times;</span>";
        tagsDiv.appendChild(tag);
    });
}

function renderMultiBoxplot() {
    if (selectedGenes.length === 0) return;

    var groups = geneExprData.groups;
    var samples = geneExprData.samples;
    var uniqueGroups = [];
    groups.forEach(function(g) { if (uniqueGroups.indexOf(g) === -1) uniqueGroups.push(g); });

    var colors = ["#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00", "#a65628", "#f781bf", "#999999"];
    var nGenes = selectedGenes.length;
    var traces = [];

    if (nGenes === 1) {
        // Single gene: one box per group on the x-axis
        var geneIdx = selectedGenes[0].idx;
        var values = geneExprData.expr[geneIdx];
        uniqueGroups.forEach(function(g, gi) {
            var gValues = [], gSamples = [];
            for (var i = 0; i < values.length; i++) {
                if (groups[i] === g) { gValues.push(values[i]); gSamples.push(samples[i]); }
            }
            traces.push({
                y: gValues, type: "box", name: g,
                boxpoints: "all", jitter: 0.4, pointpos: 0,
                text: gSamples,
                hovertemplate: "Sample: %{text}<br>Expression: %{y:.3f}<extra>" + g + "</extra>",
                marker: { color: colors[gi % colors.length], size: 6, opacity: 0.7 },
                line: { color: colors[gi % colors.length] },
                fillcolor: colors[gi % colors.length] + "33"
            });
        });
        var layout = {
            title: { text: selectedGenes[0].name + " Expression", font: { size: 16 } },
            yaxis: { title: "Expression (TMMlogCPM)" },
            xaxis: { title: bpGroupLabel },
            showlegend: false,
            margin: { t: 50, b: 80, l: 70, r: 30 },
            boxgap: 0.3
        };
        var containerH = 450;
    } else {
        // Multiple genes: single combined plot, genes on x-axis, grouped by condition
        var containerH = Math.max(450, 350 + nGenes * 30);

        // One trace per group — each trace has x = gene names for that group
        uniqueGroups.forEach(function(g, gi) {
            var xVals = [], yVals = [], sampleTexts = [];
            selectedGenes.forEach(function(sg) {
                var values = geneExprData.expr[sg.idx];
                for (var i = 0; i < values.length; i++) {
                    if (groups[i] === g) {
                        xVals.push(sg.name);
                        yVals.push(values[i]);
                        sampleTexts.push(samples[i]);
                    }
                }
            });
            traces.push({
                x: xVals, y: yVals, type: "box", name: g,
                boxpoints: "all", jitter: 0.4, pointpos: 0,
                text: sampleTexts,
                hovertemplate: "%{x}<br>Sample: %{text}<br>Expression: %{y:.3f}<extra>" + g + "</extra>",
                marker: { color: colors[gi % colors.length], size: 5, opacity: 0.7 },
                line: { color: colors[gi % colors.length] },
                fillcolor: colors[gi % colors.length] + "33"
            });
        });

        var geneNames = selectedGenes.map(function(sg) { return sg.name; });
        var layout = {
            title: { text: "Gene Expression Comparison", font: { size: 16 } },
            yaxis: { title: "Expression (TMMlogCPM)" },
            xaxis: { title: "Gene", categoryorder: "array", categoryarray: geneNames },
            showlegend: true,
            legend: { title: { text: bpGroupLabel } },
            boxmode: "group",
            margin: { t: 50, b: 100, l: 70, r: 30 },
            boxgap: 0.15,
            boxgroupgap: 0.1
        };
    }

    document.getElementById("boxplot_container").style.height = containerH + "px";
    Plotly.newPlot("boxplot_container", traces, layout, { responsive: true });

    // Show info
    var infoLines = selectedGenes.map(function(sg) {
        var vals = geneExprData.expr[sg.idx].filter(function(v) { return !isNaN(v); });
        var mean = vals.reduce(function(a, b) { return a + b; }, 0) / vals.length;
        return "<strong>" + sg.name + "</strong>: mean " + mean.toFixed(3) + " TMMlogCPM (" + vals.length + " samples)";
    });
    document.getElementById("boxplot_gene_info").innerHTML = infoLines.join(" &nbsp;|&nbsp; ");
    document.getElementById("boxplot_export_bar").style.display = "flex";
}

function clearBoxplot() {
    selectedGenes = [];
    document.getElementById("boxplot_gene_input").value = "";
    document.getElementById("boxplot_gene_info").innerHTML = "";
    document.getElementById("boxplot_gene_tags").innerHTML = "";
    document.getElementById("boxplot_export_bar").style.display = "none";
    Plotly.purge("boxplot_container");
}

function exportBoxplot(fmt) {
    var container = document.getElementById("boxplot_container");
    if (!container || !container.data) { alert("No plot to export."); return; }
    var geneNames = selectedGenes.map(function(g) { return g.name; }).join("_");
    if (fmt === "tiff") {
        // Plotly does not natively support TIFF. Export as high-res PNG instead
        // and inform the user.
        Plotly.downloadImage(container, {
            format: "png", width: 2400, height: 1600, scale: 3,
            filename: "gene_expression_" + geneNames
        });
        alert("TIFF is not natively supported in browsers. A high-resolution PNG (300 DPI equivalent) has been saved instead. You can convert it to TIFF using ImageMagick or any image editor.");
    } else {
        var w = fmt === "svg" ? 1200 : 2400;
        var h = fmt === "svg" ? 800 : 1600;
        var s = fmt === "svg" ? 1 : 3;
        Plotly.downloadImage(container, {
            format: fmt, width: w, height: h, scale: s,
            filename: "gene_expression_" + geneNames
        });
    }
}

// Close autocomplete on outside click
document.addEventListener("click", function(e) {
    var acDiv = document.getElementById("boxplot_ac");
    var input = document.getElementById("boxplot_gene_input");
    if (acDiv && e.target !== input && !acDiv.contains(e.target)) {
        acDiv.style.display = "none";
    }
});
</script>')
```

```{r gene-explorer-fallback, eval=!explorer_ready, results='asis', echo=FALSE}
cat("*Gene expression explorer requires normalized expression data and metadata.*\n")
```

## Top DE Heatmap {.tabset}

Heatmaps of the most significantly differentially expressed features provide a focused view of the strongest expression changes. Hierarchical clustering of both features and samples reveals co-regulation patterns and group-level expression signatures.

Expression heatmap of top differentially expressed genes. Select the number of genes to display.

```{r top-de-heatmap-prep, include=FALSE}
# Load normalized expression data
norm_expr_file <- file.path(datasets_dir, "rna_norm_TMMlogCPM.tsv")
norm_expr <- NULL
if (file.exists(norm_expr_file) && !is.null(de_summary_full)) {
    norm_expr <- read.delim(norm_expr_file, stringsAsFactors = FALSE, row.names = 1, check.names = FALSE)
}

# Function to create heatmap for top N genes
make_top_de_heatmap <- function(n_genes, expr_mat, de_df, meta_df, group_col) {
    if (is.null(expr_mat) || is.null(de_df)) return(NULL)

    # Get top N gene IDs
    top_genes <- head(de_df, n_genes)
    id_col <- if ("Gene" %in% names(top_genes)) "Gene" else names(top_genes)[1]
    gene_ids <- as.character(top_genes[[id_col]])

    # Match to expression matrix
    gene_ids_clean <- gsub("^Gene:", "", gene_ids)
    matched_idx <- which(rownames(expr_mat) %in% gene_ids | rownames(expr_mat) %in% gene_ids_clean)

    if (length(matched_idx) < 2) return(NULL)

    mat <- as.matrix(expr_mat[matched_idx, , drop = FALSE])

    # Use gene names if available
    if ("GeneName" %in% names(top_genes)) {
        name_map <- setNames(top_genes$GeneName, gsub("^Gene:", "", top_genes[[id_col]]))
        new_names <- name_map[rownames(mat)]
        new_names[is.na(new_names)] <- rownames(mat)[is.na(new_names)]
        rownames(mat) <- new_names
    }

    # Scale rows
    mat_scaled <- t(scale(t(mat)))
    mat_scaled[is.na(mat_scaled)] <- 0

    # Create annotation
    if (!is.null(meta_df) && !is.null(group_col) && group_col %in% names(meta_df)) {
        sample_col <- rna_cfg$effects$samples %||% "SampleID"
        if (sample_col %in% names(meta_df)) {
            anno_df <- data.frame(Group = meta_df[[group_col]], row.names = meta_df[[sample_col]])
            # Match columns
            common_samples <- intersect(colnames(mat_scaled), rownames(anno_df))
            if (length(common_samples) > 0) {
                mat_scaled <- mat_scaled[, common_samples, drop = FALSE]
                anno_df <- anno_df[common_samples, , drop = FALSE]
            }
        } else {
            anno_df <- NULL
        }
    } else {
        anno_df <- NULL
    }

    # Create heatmap
    if (requireNamespace("pheatmap", quietly = TRUE)) {
        pheatmap::pheatmap(
            mat_scaled,
            annotation_col = anno_df,
            show_rownames = nrow(mat_scaled) <= 50,
            show_colnames = ncol(mat_scaled) <= 30,
            cluster_rows = TRUE,
            cluster_cols = TRUE,
            main = sprintf("Top %d DE Genes (z-score)", n_genes),
            fontsize_row = if (nrow(mat_scaled) <= 30) 8 else 6,
            silent = TRUE
        )
    }
}

# Load metadata
meta_file <- file.path(cfg$project$dir, cfg$paths$raw, rna_cfg$files$metadata)
meta_df <- if (file.exists(meta_file)) read.csv(meta_file, stringsAsFactors = FALSE) else NULL
group_col <- rna_cfg$filtering$group_col %||% rna_cfg$effects$color
```

### Top 10

```{r heatmap-top10, fig.height=6, fig.width=10}
if (!is.null(norm_expr)) {
    hm <- make_top_de_heatmap(10, norm_expr, de_summary_full, meta_df, group_col)
    if (!is.null(hm)) hm
} else {
    cat("Expression matrix not available.")
}
```

```{r legend-heatmap-top10, results='asis', echo=FALSE}
if (!is.null(norm_expr) && !is.null(de_summary_full)) {
    figure_legend("Heatmap of the top 10 differentially expressed genes (z-score normalized) across samples. Rows are clustered by expression similarity; columns are annotated by experimental group.")
}
```

### Top 50

```{r heatmap-top50, fig.height=8, fig.width=10}
if (!is.null(norm_expr)) {
    hm <- make_top_de_heatmap(50, norm_expr, de_summary_full, meta_df, group_col)
    if (!is.null(hm)) hm
}
```

```{r legend-heatmap-top50, results='asis', echo=FALSE}
if (!is.null(norm_expr) && !is.null(de_summary_full)) {
    figure_legend("Heatmap of the top 50 differentially expressed genes (z-score normalized) across samples. Rows are clustered by expression similarity; columns are annotated by experimental group.")
}
```

### Top 100

```{r heatmap-top100, fig.height=10, fig.width=10}
if (!is.null(norm_expr)) {
    hm <- make_top_de_heatmap(100, norm_expr, de_summary_full, meta_df, group_col)
    if (!is.null(hm)) hm
}
```

```{r legend-heatmap-top100, results='asis', echo=FALSE}
if (!is.null(norm_expr) && !is.null(de_summary_full)) {
    figure_legend("Heatmap of the top 100 differentially expressed genes (z-score normalized) across samples. Rows are clustered by expression similarity; columns are annotated by experimental group.")
}
```

### Top 200

```{r heatmap-top200, fig.height=12, fig.width=10}
if (!is.null(norm_expr)) {
    hm <- make_top_de_heatmap(200, norm_expr, de_summary_full, meta_df, group_col)
    if (!is.null(hm)) hm
}
```

```{r legend-heatmap-top200, results='asis', echo=FALSE}
if (!is.null(norm_expr) && !is.null(de_summary_full)) {
    figure_legend("Heatmap of the top 200 differentially expressed genes (z-score normalized) across samples. Rows are clustered by expression similarity; columns are annotated by experimental group.")
}
```

### All DE

```{r heatmap-all, fig.height=14, fig.width=10}
if (!is.null(norm_expr) && !is.null(de_summary_full)) {
    n_all <- nrow(de_summary_full)
    if (n_all > 500) {
        cat(sprintf("Showing top 500 of %d DE genes (full heatmap too large).\n\n", n_all))
        n_all <- 500
    }
    hm <- make_top_de_heatmap(n_all, norm_expr, de_summary_full, meta_df, group_col)
    if (!is.null(hm)) hm
}
```

```{r legend-heatmap-all, results='asis', echo=FALSE}
if (!is.null(norm_expr) && !is.null(de_summary_full)) {
    figure_legend("Heatmap of all significantly differentially expressed genes (z-score normalized) across samples. Rows are clustered by expression similarity; columns are annotated by experimental group.")
}
```

---

# Cell Type Deconvolution

Cell type deconvolution estimates the relative proportions of distinct cell populations within bulk RNA-seq samples. This computational approach leverages reference expression signatures to infer cellular composition without physical cell separation.

```{r deconv-setup}
deconv_dir <- file.path(run_dir, "rna", "Deconvolution")
has_deconvolution <- dir.exists(deconv_dir) &&
    (file.exists(file.path(deconv_dir, "xcell2_scores.csv")) ||
     file.exists(file.path(deconv_dir, "deconvolution_results.rds")))
```

```{r deconv-skip, eval=!has_deconvolution, results='asis'}
cat("Cell type deconvolution was not run. This analysis is only available for **human** and **mouse** datasets.\n\n")
cat("To enable, set `deconvolution.enabled: true` in the configuration file.\n")
```

```{r deconv-heatmap, eval=has_deconvolution, fig.cap="Cell type enrichment scores (xCell2)"}
heatmap_path <- file.path(deconv_dir, "plots", "xcell2_heatmap.png")
if (file.exists(heatmap_path)) knitr::include_graphics(heatmap_path)
```

```{r legend-deconv-heatmap, eval=has_deconvolution, results='asis', echo=FALSE}
figure_legend("Cell type enrichment score heatmap (xCell2). Rows represent inferred cell types and columns represent samples. Higher scores indicate greater predicted abundance of a given cell type in the sample.")
```

```{r deconv-heatmap-commentary, eval=has_deconvolution, results='asis'}
render_commentary("deconv_heatmap")
```

```{r deconv-barplot, eval=has_deconvolution, fig.cap="Cell type composition by condition"}
bar_path <- file.path(deconv_dir, "plots", "celltype_stacked_bar.png")
if (file.exists(bar_path)) knitr::include_graphics(bar_path)
```

```{r legend-deconv-barplot, eval=has_deconvolution, results='asis', echo=FALSE}
figure_legend("Stacked bar chart of estimated cell type proportions across experimental conditions. Each bar represents a sample or condition, with segments showing the relative abundance of different cell types.")
```

```{r deconv-barplot-commentary, eval=has_deconvolution, results='asis'}
render_commentary("deconv_stacked_bar")
```

```{r deconv-diff-table, eval=has_deconvolution}
diff_comp_file <- file.path(deconv_dir, "differential_composition.csv")
if (file.exists(diff_comp_file)) {
    diff_comp <- read.csv(diff_comp_file, stringsAsFactors = FALSE)
    make_dt(diff_comp, caption = "Differential Cell Type Composition")
}
```

---

# Enrichment / Pathway Analysis

Gene set enrichment analysis tests whether predefined biological pathways show coordinated expression changes beyond what is expected by chance. fGSEA uses a ranked list of all genes (rank-based), while ORA tests only the significant gene list against a background (threshold-based).

```{r pathway-setup}
# Check if enrichment results exist
has_enrichment <- dir.exists(enrich_dir) && length(list.files(enrich_dir, pattern = "\\.csv$")) > 0
has_pathway_plots <- dir.exists(enrich_plot_dir) && length(list.files(enrich_plot_dir, pattern = "\\.png$")) > 0
```

```{r pathway-skip, eval=!has_enrichment, results='asis'}
cat("Pathway analysis was not run or produced no results. To enable it, set ",
    "`pathway.enabled: true` in the configuration file and ensure the required ",
    "R packages are installed (fgsea, org.*.eg.db, clusterProfiler).\n")
```

```{r pathway-annotation, eval=has_enrichment}
anno_file <- file.path(enrich_dir, "gene_annotation.csv")
if (file.exists(anno_file)) {
    anno <- read.csv(anno_file, stringsAsFactors = FALSE)
    n_annotated <- sum(!is.na(anno$symbol) & anno$symbol != anno$gene_id)
    n_total <- nrow(anno)
    cat(sprintf("**Gene annotation:** %d / %d genes annotated (%.1f%%)\n\n",
                n_annotated, n_total, 100 * n_annotated / n_total))
}
```

**Method:** Fast Gene Set Enrichment Analysis (fGSEA) ranks all genes by their differential expression statistic and tests whether genes in each pathway are enriched at the top or bottom of the ranked list.

**Key metrics:**

- **NES (Normalized Enrichment Score)**: The primary statistic from fGSEA. Positive NES indicates the pathway is enriched in **up-regulated** genes; negative NES indicates enrichment in **down-regulated** genes. The magnitude reflects the degree of enrichment.
- **padj**: Benjamini-Hochberg adjusted p-value (FDR). Pathways with padj < 0.05 are considered significant.
- **size**: Number of genes in the pathway that were present in the dataset.

## fGSEA Results {.tabset}

```{r fgsea-tables-prep, eval=has_enrichment, include=FALSE}
# Pre-load fGSEA data for rendering in regular (non-asis) chunks
fgsea_files <- list.files(enrich_dir, pattern = "pathway_.*_fgsea\\.csv$", full.names = TRUE)
fgsea_tables <- list()
if (length(fgsea_files) > 0) {
    for (i in seq_along(fgsea_files)) {
        ff <- fgsea_files[i]
        label <- gsub("^pathway_|_fgsea\\.csv$", "", basename(ff))
        label_clean <- gsub("_", " ", label)
        fgsea_df <- read.csv(ff, stringsAsFactors = FALSE)
        sig <- fgsea_df[!is.na(fgsea_df$padj) & fgsea_df$padj < 0.05, ]
        if (nrow(sig) > 0) {
            sig <- sig[order(sig$padj), ]
            if ("pathway_name" %in% names(sig)) {
                show_cols <- intersect(c("pathway_name", "pval", "padj", "NES", "size"), names(sig))
            } else {
                show_cols <- intersect(c("pathway", "pval", "padj", "NES", "size"), names(sig))
            }
            sig_show <- sig[, show_cols]
            sig_show$Direction <- ifelse(sig_show$NES > 0, "Up-regulated", "Down-regulated")
            fgsea_tables[[i]] <- list(label = label_clean, data = sig_show)
        } else {
            fgsea_tables[[i]] <- list(label = label_clean, data = NULL)
        }
    }
}
```

```{r fgsea-tables, eval=has_enrichment, results='asis'}
if (length(fgsea_tables) > 0) {
    for (i in seq_along(fgsea_tables)) {
        tbl <- fgsea_tables[[i]]
        cat(sprintf("\n### %s {.tabset}\n\n", tbl$label))
        if (is.null(tbl$data)) {
            cat("No significant pathways (padj < 0.05) for this contrast/database.\n\n")
            next
        }
        sig_show <- tbl$data
        n_up <- sum(sig_show$Direction == "Up-regulated")
        n_down <- sum(sig_show$Direction == "Down-regulated")

        cat(sprintf("\n#### All (%d pathways)\n\n", nrow(sig_show)))
        cat(sprintf("**Up-regulated: %d** | **Down-regulated: %d**\n\n", n_up, n_down))
        cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Fast Gene Set Enrichment Analysis (fGSEA) results showing significantly enriched pathways (adjusted p-value &lt; 0.05). NES indicates direction and magnitude of enrichment.</em></p>\n")
        sig_all <- sig_show[order(sig_show$padj), ]
        print(htmltools::tagList(make_dt(sig_all, paste("All significant pathways —", tbl$label))))
        cat("\n\n")

        if (n_up > 0) {
            cat(sprintf("\n#### Up-regulated (%d)\n\n", n_up))
            cat("Pathways enriched in **up-regulated** genes (positive NES).\n\n")
            sig_up <- sig_show[sig_show$Direction == "Up-regulated", ]
            sig_up <- sig_up[order(sig_up$padj), ]
            print(htmltools::tagList(make_dt(sig_up, paste("Up-regulated pathways —", tbl$label))))
            cat("\n\n")
        }

        if (n_down > 0) {
            cat(sprintf("\n#### Down-regulated (%d)\n\n", n_down))
            cat("Pathways enriched in **down-regulated** genes (negative NES).\n\n")
            sig_down <- sig_show[sig_show$Direction == "Down-regulated", ]
            sig_down <- sig_down[order(sig_down$padj), ]
            print(htmltools::tagList(make_dt(sig_down, paste("Down-regulated pathways —", tbl$label))))
            cat("\n\n")
        }
    }
} else {
    cat("No fGSEA result files found.\n")
}
```

## Pathway Dotplots {.tabset}

```{r pathway-dotplots, eval=has_pathway_plots, results='asis'}
plot_pngs <- list.files(enrich_plot_dir, pattern = "\\.png$", full.names = TRUE)
if (length(plot_pngs) > 0) {
    for (pp in plot_pngs) {
        label_raw <- gsub("^pathway_|\\.png$", "", basename(pp))
        label <- gsub("_", " ", label_raw)
        cat(sprintf("\n### %s\n\n", label))
        cat(sprintf("![%s](%s)\n\n", label, pp))
        figure_legend(sprintf("Dotplot of top enriched pathways for %s. Dot size represents the number of pathway members; color intensity reflects statistical significance (adjusted p-value). Positive NES values indicate up-regulated pathways; negative values indicate down-regulated pathways.", label))
        render_commentary(paste0("pathway_", label_raw))
    }
}
```

## Leading Edge Genes {.tabset}

Leading edge genes are the core subset of pathway members that drive the enrichment signal. They represent the genes most responsible for a pathway's statistical significance and are candidates for validation experiments.

The **leading edge** genes are those that contribute most to the enrichment signal of each pathway. These are the genes that appear before the peak of the running enrichment score.

```{r leading-edge, eval=has_enrichment, results='asis'}
# Load annotation for gene name lookup
anno_le <- NULL
anno_file_le <- file.path(enrich_dir, "gene_annotation.csv")
if (file.exists(anno_file_le)) {
    anno_le <- read.csv(anno_file_le, stringsAsFactors = FALSE)
}

fgsea_files <- list.files(enrich_dir, pattern = "pathway_.*_fgsea\\.csv$", full.names = TRUE)
if (length(fgsea_files) > 0) {
    for (ff in fgsea_files) {
        label <- gsub("^pathway_|_fgsea\\.csv$", "", basename(ff))
        label_clean <- gsub("_", " ", label)
        cat(sprintf("\n### %s\n\n", label_clean))

        fgsea_df <- read.csv(ff, stringsAsFactors = FALSE)
        sig <- fgsea_df[!is.na(fgsea_df$padj) & fgsea_df$padj < 0.05, ]

        if (nrow(sig) > 0 && "leadingEdge" %in% names(sig)) {
            sig <- head(sig[order(sig$padj), ], 10)  # Top 10 pathways

            le_list <- list()
            for (i in seq_len(nrow(sig))) {
                pathway_name <- if ("pathway_name" %in% names(sig)) sig$pathway_name[i] else sig$pathway[i]
                le_genes <- strsplit(sig$leadingEdge[i], ",")[[1]]

                # Convert to gene symbols if annotation available
                if (!is.null(anno_le)) {
                    le_symbols <- anno_le$symbol[match(le_genes, anno_le$gene_id)]
                    le_symbols[is.na(le_symbols)] <- le_genes[is.na(le_symbols)]
                    le_genes <- le_symbols
                }

                le_list[[i]] <- data.frame(
                    Pathway = pathway_name,
                    NES = round(sig$NES[i], 2),
                    padj = signif(sig$padj[i], 3),
                    `Leading Edge Genes` = paste(head(le_genes, 15), collapse = ", "),
                    check.names = FALSE,
                    stringsAsFactors = FALSE
                )
            }

            le_df <- do.call(rbind, le_list)
            cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Leading edge genes for the top enriched pathways. These core genes drive the enrichment signal and represent candidates for experimental validation.</em></p>\n")
            print(htmltools::tagList(make_dt(le_df, paste("Leading Edge Genes —", label_clean))))
            cat("\n\n")
        } else {
            cat("No significant pathways with leading edge information.\n\n")
        }
    }
}
```

## Pathway Volcano {.tabset}

```{r pathway-volcano-setup-rna, include=FALSE}
pw_volcano_files_rna <- list.files(enrich_dir, pattern = "^pathway_volcano_data_.*\\.csv$", full.names = TRUE)
has_pathway_volcano_rna <- length(pw_volcano_files_rna) > 0
if (has_pathway_volcano_rna) {
    has_pathway_volcano_rna <- any(vapply(pw_volcano_files_rna, function(f) {
        df <- read.csv(f, stringsAsFactors = FALSE, nrows = 5)
        "pathways" %in% colnames(df) && any(nzchar(df$pathways), na.rm = TRUE)
    }, logical(1)))
}
```

```{r pathway-volcano-skip-rna, eval=!has_pathway_volcano_rna, results='asis'}
cat("No pathway volcano data available. Enable pathway analysis with significant results to generate pathway-colored volcano plots.\n")
```

```{r pathway-volcano-legend-rna, eval=has_pathway_volcano_rna, results='asis'}
figure_legend("Pathway volcano plots overlay enriched pathway membership onto the standard DE volcano plot. Select a pathway from the dropdown to highlight its member genes in orange. Non-member genes are dimmed for contrast.")
```

```{r pathway-volcano-plots-rna, eval=has_pathway_volcano_rna, results='asis'}
for (vf in pw_volcano_files_rna) {
    contrast_name <- gsub("^pathway_volcano_data_|\\.csv$", "", basename(vf))
    contrast_clean <- gsub("_", " ", contrast_name)
    clean_id <- gsub("[^a-zA-Z0-9]", "_", contrast_name)

    volcano_df <- read.csv(vf, stringsAsFactors = FALSE)
    if (!"pathways" %in% colnames(volcano_df)) next

    # Cap -log10(p) at 99.5th percentile to keep plot readable
    yvals <- volcano_df$neg_log10_pvalue[is.finite(volcano_df$neg_log10_pvalue)]
    ycap <- max(quantile(yvals, 0.995, na.rm = TRUE), 10)
    volcano_df$neg_log10_pvalue <- pmin(volcano_df$neg_log10_pvalue, ycap)

    cat(sprintf("\n### %s\n\n", contrast_clean))

    # Collect fGSEA info for NES / padj display
    fgsea_files_for_contrast <- list.files(enrich_dir,
        pattern = paste0("pathway_", contrast_name, "_.*_fgsea\\.csv$"),
        full.names = TRUE)
    pathway_info <- list()
    for (ff in fgsea_files_for_contrast) {
        fdf <- read.csv(ff, stringsAsFactors = FALSE)
        sig <- fdf[!is.na(fdf$padj) & fdf$padj < 0.05, , drop = FALSE]
        for (r in seq_len(nrow(sig))) {
            # Use pathway_name (human-readable) to match the volcano data keys
            pname <- if ("pathway_name" %in% names(sig) && nzchar(sig$pathway_name[r] %||% "")) {
                sig$pathway_name[r]
            } else {
                sig$pathway[r]
            }
            pathway_info[[pname]] <- list(
                NES = round(sig$NES[r], 3),
                padj = signif(sig$padj[r], 3),
                size = sig$size[r]
            )
        }
    }

    # Assign direction colors
    padj_cut_v <- rna_cfg$de$p_cutoff %||% 0.05
    lfc_cut_v <- log2(rna_cfg$de$linear_fc_cutoff %||% 1.5)
    volcano_df$Direction <- ifelse(
        is.na(volcano_df$padj) | volcano_df$padj >= padj_cut_v, "NS",
        ifelse(volcano_df$log2FC > lfc_cut_v, "Up",
               ifelse(volcano_df$log2FC < -lfc_cut_v, "Down", "NS")))

    dir_colors <- c("Up" = "#e74c3c", "Down" = "#3498db", "NS" = "#bdc3c7")

    p <- plotly::plot_ly(
        data = volcano_df,
        x = ~log2FC,
        y = ~neg_log10_pvalue,
        type = "scatter",
        mode = "markers",
        color = ~Direction,
        colors = dir_colors,
        marker = list(size = 5, opacity = 0.7),
        text = ~paste0("Gene: ", FeatureID,
                        "<br>log2FC: ", round(log2FC, 3),
                        "<br>padj: ", signif(padj, 3),
                        "<br>Pathways: ", ifelse(nzchar(pathways), pathways, "none")),
        hoverinfo = "text",
        customdata = ~FeatureID,
        source = paste0("pw_volcano_rna_", clean_id)
    ) |>
        plotly::layout(
            title = list(text = paste("Pathway Volcano -", contrast_clean), font = list(size = 14)),
            xaxis = list(title = "log2 Fold Change"),
            yaxis = list(title = "-log10(p-value)", rangemode = "tozero"),
            shapes = list(
                list(type = "line", x0 = -lfc_cut_v, x1 = -lfc_cut_v, y0 = 0, y1 = 1,
                     yref = "paper", line = list(dash = "dash", color = "grey50")),
                list(type = "line", x0 = lfc_cut_v, x1 = lfc_cut_v, y0 = 0, y1 = 1,
                     yref = "paper", line = list(dash = "dash", color = "grey50")),
                list(type = "line", x0 = 0, x1 = 1, xref = "paper",
                     y0 = -log10(padj_cut_v), y1 = -log10(padj_cut_v),
                     line = list(dash = "dash", color = "grey50"))
            )
        ) |>
        plotly::config(
            toImageButtonOptions = list(format = "svg", filename = paste0("pathway_volcano_", clean_id),
                                         width = 1200, height = 900, scale = 3),
            modeBarButtonsToAdd = list(
                list(name = "Download PNG (300 dpi)", icon = list(
                    path = "M12 2C6.48 2 2 6.48 2 12s4.48 10 10 10 10-4.48 10-10S17.52 2 12 2zm-1 14H8l4-4 4 4h-3v3h-2v-3z",
                    width = 24, height = 24
                ), click = htmlwidgets::JS(sprintf(
                    "function(gd) { Plotly.downloadImage(gd, {format: 'png', width: 2400, height: 1800, scale: 3, filename: 'pathway_volcano_%s'}); }",
                    clean_id
                )))
            )
        )

    p$elementId <- paste0("pw_volcano_rna_plot_", clean_id)

    # Build pathway->genes mapping from the pathways column
    all_pathways <- unique(unlist(strsplit(volcano_df$pathways[nzchar(volcano_df$pathways)], "; ")))
    all_pathways <- sort(all_pathways[nzchar(all_pathways)])

    pw_genes_map <- list()
    for (pw in all_pathways) {
        members <- as.character(volcano_df$FeatureID[grepl(pw, volcano_df$pathways, fixed = TRUE)])
        pw_genes_map[[pw]] <- I(members)
    }

    pw_genes_json <- jsonlite::toJSON(pw_genes_map, auto_unbox = TRUE)
    pw_info_json <- jsonlite::toJSON(pathway_info, auto_unbox = TRUE)

    # Build dropdown options
    options_html <- '<option value="">None (show all)</option>'
    for (pw in all_pathways) {
        pw_escaped <- gsub("\"", "&quot;", pw)
        info <- pathway_info[[pw]]
        label <- if (!is.null(info)) {
            sprintf("%s (NES=%.2f, padj=%.3g)", pw, info$NES, info$padj)
        } else { pw }
        label_escaped <- gsub("\"", "&quot;", label)
        options_html <- paste0(options_html,
            sprintf('<option value="%s">%s</option>', pw_escaped, label_escaped))
    }

    cat(sprintf('
<div style="margin-bottom: 10px;">
<label for="pw_select_rna_%s"><strong>Highlight pathway:</strong></label>
<select id="pw_select_rna_%s" style="padding: 6px; border-radius: 4px; border: 1px solid #ccc; margin-left: 8px; max-width: 600px;">
%s
</select>
<div id="pw_info_rna_%s" style="margin-top: 6px; font-size: 13px; color: #555;"></div>
<div id="pw_legend_rna_%s" style="margin-top: 4px; font-size: 13px;"></div>
</div>
', clean_id, clean_id, options_html, clean_id, clean_id))

    print(htmltools::tagList(p))

    cat(sprintf('
<script>
(function() {
    var cid = "%s";
    var plotId = "pw_volcano_rna_plot_" + cid;
    var pwGenes = %s;
    var pwInfo = %s;
    var origColors = null;
    var origOpacities = null;

    function captureOriginal() {
        var plotDiv = document.getElementById(plotId);
        if (!plotDiv || !plotDiv.data) return;
        origColors = [];
        origOpacities = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var mc = plotDiv.data[t].marker;
            origColors.push(mc ? (mc.color || null) : null);
            origOpacities.push(mc ? (mc.opacity || 0.7) : 0.7);
        }
    }

    function onSelect() {
        var plotDiv = document.getElementById(plotId);
        if (!plotDiv || !plotDiv.data) return;
        if (!origColors) captureOriginal();

        var pw = document.getElementById("pw_select_rna_" + cid).value;
        var infoDiv = document.getElementById("pw_info_rna_" + cid);
        var legendDiv = document.getElementById("pw_legend_rna_" + cid);

        // Freeze axes so restyle cannot trigger auto-range
        var xRange = plotDiv.layout.xaxis.range ? plotDiv.layout.xaxis.range.slice() : null;
        var yRange = plotDiv.layout.yaxis.range ? plotDiv.layout.yaxis.range.slice() : null;
        if (xRange && yRange) {
            Plotly.relayout(plotDiv, {
                "xaxis.autorange": false, "xaxis.range": xRange,
                "yaxis.autorange": false, "yaxis.range": yRange
            });
        }

        if (!pw) {
            for (var t = 0; t < plotDiv.data.length; t++) {
                Plotly.restyle(plotDiv, {
                    "marker.color": [origColors[t]],
                    "marker.opacity": [origOpacities[t]],
                    "marker.size": [5]
                }, [t]);
            }
            infoDiv.innerHTML = "";
            if (legendDiv) legendDiv.innerHTML = "";
            return;
        }

        var genes = pwGenes[pw] || [];
        if (typeof genes === "string") genes = [genes];
        var genesSet = {};
        for (var g = 0; g < genes.length; g++) genesSet[genes[g]] = true;

        var info = pwInfo[pw];
        if (info) {
            infoDiv.innerHTML = "<strong>" + pw + "</strong> &mdash; NES: " +
                info.NES + " | padj: " + info.padj + " | Members in data: " + genes.length;
        } else {
            infoDiv.innerHTML = "<strong>" + pw + "</strong> &mdash; Members in data: " + genes.length;
        }
        if (legendDiv) {
            legendDiv.innerHTML = "<font color=#e67e22>&#9679;</font> Pathway member &nbsp;" +
                "<font color=#d5d8dc>&#9679;</font> Non-member";
        }

        for (var t = 0; t < plotDiv.data.length; t++) {
            var cd = plotDiv.data[t].customdata;
            if (!cd) { continue; }
            var colors = [];
            var opacities = [];
            var sizes = [];
            for (var i = 0; i < cd.length; i++) {
                var fid = Array.isArray(cd[i]) ? cd[i][0] : cd[i];
                if (genesSet[fid]) {
                    colors.push("#e67e22");
                    opacities.push(0.9);
                    sizes.push(8);
                } else {
                    colors.push("#d5d8dc");
                    opacities.push(0.15);
                    sizes.push(4);
                }
            }
            Plotly.restyle(plotDiv, {
                "marker.color": [colors],
                "marker.opacity": [opacities],
                "marker.size": [sizes]
            }, [t]);
        }
    }

    var attempts = 0;
    var timer = setInterval(function() {
        var plotDiv = document.getElementById(plotId);
        if ((plotDiv && plotDiv.data) || attempts > 50) {
            clearInterval(timer);
            if (plotDiv && plotDiv.data) {
                captureOriginal();
                document.getElementById("pw_select_rna_" + cid).addEventListener("change", onSelect);
            }
        }
        attempts++;
    }, 200);
})();
</script>
', clean_id, pw_genes_json, pw_info_json))

    cat("\n\n")
}
```

---

# Output Files

```{r output-files}
files_tbl <- data.frame(
  File = c(
    "Final_results_ALL_P_0.05.xlsx",
    "Final_results_DE_P_0.05.xlsx",
    "Datasets/final_results.tsv",
    "Datasets/deseq2_summary_p0.05.tsv",
    "Datasets/de_summary_counts.tsv",
    "Datasets/rna_counts_filtered.tsv",
    "Datasets/rna_norm_TMMlogCPM.tsv"
  ),
  Description = c(
    "Excel workbook with all genes and DE statistics (p < 0.05)",
    "Excel workbook with DE genes only (p < 0.05)",
    "Full results table (TSV) — all genes with fold changes, p-values, and pass flags",
    "DESeq2 summary for significant genes",
    "DE gene counts per contrast (up/down/total)",
    "Filtered raw counts matrix (genes that passed CPM threshold)",
    "TMMlogCPM-normalized expression matrix"
  ),
  stringsAsFactors = FALSE
)

# Add enrichment files if they exist
if (has_enrichment) {
    enrich_csvs <- list.files(enrich_dir, pattern = "\\.csv$")
    if (length(enrich_csvs) > 0) {
        enrich_tbl <- data.frame(
            File = paste0("Enrichment/", enrich_csvs),
            Description = paste("Pathway analysis results —",
                                gsub("^pathway_|\\.csv$", "", enrich_csvs)),
            stringsAsFactors = FALSE
        )
        files_tbl <- rbind(files_tbl, enrich_tbl)
    }
}

make_dt(files_tbl, "Key Output Files")
```

---

# Methods

This section provides a summary suitable for publication methods sections.

```{r methods-text, results='asis'}
# Auto-generate methods paragraph based on config
norm_method <- rna_cfg$normalization$method %||% "TMMlogCPM"
de_method <- rna_cfg$de$method %||% "DESeq2"
padj_cut <- rna_cfg$de$p_cutoff %||% 0.05
lfc_cut <- rna_cfg$de$linear_fc_cutoff %||% 1.5
filter_min <- rna_cfg$filtering$auto_filter$min %||% 0.5
filter_max <- rna_cfg$filtering$auto_filter$max %||% 2.0

# Get sample and gene counts
n_samples <- if (exists("meta") && is.data.frame(meta)) nrow(meta) else "N"
n_genes_final <- if (!is.null(de_summary_full)) nrow(de_summary_full) else "N"
n_de <- if (!is.null(de_summary_full)) {
    pass_col <- grep("pass_any", names(de_summary_full), value = TRUE)[1]
    if (!is.na(pass_col)) sum(de_summary_full[[pass_col]] == TRUE, na.rm = TRUE) else "N"
} else "N"

# Pathway info
pathway_method <- rna_cfg$pathway$method %||% "fGSEA"
pathway_dbs <- paste(rna_cfg$pathway$databases %||% c("GO", "KEGG"), collapse = " and ")

cat("**RNA-seq Analysis Methods**\n\n")

cat(sprintf(
"RNA-seq count data from %s samples were processed using the multiomics-core pipeline. ",
n_samples
))

cat(sprintf(
"Lowly expressed genes were filtered using an adaptive CPM threshold approach (range: %.1f–%.1f CPM), ",
filter_min, filter_max
))

cat(sprintf(
"retaining genes with sufficient expression in at least one biological group. "
))

cat(sprintf(
"Expression values were normalized using the %s method, which applies TMM (Trimmed Mean of M-values) scaling factors followed by log2 CPM transformation. ",
norm_method
))

cat(sprintf(
"Differential expression analysis was performed using %s with default parameters. ",
de_method
))

cat(sprintf(
"Genes were considered differentially expressed at adjusted p-value ≤ %.2f and |fold change| ≥ %.1f. ",
padj_cut, lfc_cut
))

cat(sprintf(
"A total of %s genes were tested, with %s identified as differentially expressed in at least one contrast. ",
format(as.numeric(n_genes_final), big.mark = ","),
format(as.numeric(n_de), big.mark = ",")
))

cat(sprintf(
"Pathway enrichment analysis was conducted using %s on %s gene sets. ",
pathway_method, pathway_dbs
))

cat("Pathways with adjusted p-value < 0.05 were considered significantly enriched.\n\n")

cat("**Copy-ready version:**\n\n")
cat("<div style='background-color: #f5f5f5; padding: 15px; border-radius: 5px; font-size: 0.9em;'>\n")

cat(sprintf(
"RNA-seq analysis was performed using the multiomics-core pipeline (v1.0). Lowly expressed genes were filtered using adaptive CPM thresholds (%.1f–%.1f CPM range). Expression values were normalized using %s. Differential expression was assessed with %s (adjusted p ≤ %.2f, |FC| ≥ %.1f). Pathway analysis was performed using %s on %s databases.",
filter_min, filter_max, norm_method, de_method, padj_cut, lfc_cut, pathway_method, pathway_dbs
))

cat("\n</div>\n")
```

---

# Session Info

```{r session-info, comment=""}
si_file <- file.path(run_dir, "execution_info", "sessionInfo.txt")
if (file.exists(si_file)) {
    si <- readLines(si_file)
    cat(paste(si, collapse = "\n"))
} else {
    cat(capture.output(sessionInfo()), sep = "\n")
}
```

---

<p style="text-align:center; color:grey; font-size:0.9em;">
Report generated by multiomics-core pipeline | `r format(Sys.time(), '%Y-%m-%d %H:%M')`
</p>
