---
title: "Lipidomics Analysis Report"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
    df_print: paged
params:
  pre: NULL
  qc_res: NULL
  de_res: NULL
  rf_res: NULL
  plsda_res: NULL
  class_res: NULL
  biomarker_res: NULL
  pathway_res: NULL
  qc_enhanced: NULL
  config: NULL
  commentary_file: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
library(ggplot2)
library(DT)

safe_print_plot <- function(p) {
  if (inherits(p, "gg") || inherits(p, "ggplot")) {
    print(p)
  } else {
    invisible(NULL)
  }
}

embed_png <- function(files, pattern, fig_width = "100%") {
  if (is.null(files) || length(files) == 0) return(invisible(NULL))
  hits <- files[grepl(pattern, basename(files), ignore.case = TRUE)]
  hits <- hits[file.exists(hits)]
  if (length(hits) == 0) return(invisible(NULL))
  knitr::include_graphics(hits[1])
}

render_dt <- function(df, caption = NULL, pageLength = 10) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  tbl <- DT::datatable(df, caption = caption,
                rownames = FALSE,
                extensions = "Buttons",
                options = list(pageLength = pageLength,
                               scrollX = TRUE,
                               dom = "Bfrtip",
                               buttons = list("csv", "excel")),
                class = "compact stripe hover")
  num_cols <- names(df)[vapply(df, is.numeric, logical(1))]
  pval_pat <- "^(p[._]?val|padj|adj[._]?p|fdr|q[._]?val|BH|bonferroni|p\\.value|adj\\.P\\.Val|PValue|FDR)"
  pval_cols <- grep(pval_pat, num_cols, ignore.case = TRUE, value = TRUE)
  other_cols <- setdiff(num_cols, pval_cols)
  if (length(other_cols) > 0) tbl <- DT::formatRound(tbl, columns = other_cols, digits = 2)
  if (length(pval_cols) > 0) {
    pval_idx <- which(names(df) %in% pval_cols) - 1L
    sci_render <- DT::JS(
      "function(data, type, row) {",
      "  if (type === 'display' && data !== null && data !== '') {",
      "    return parseFloat(data).toExponential(2);",
      "  }",
      "  return data;",
      "}"
    )
    col_defs <- lapply(pval_idx, function(i) list(targets = i, render = sci_render))
    tbl$x$options$columnDefs <- c(tbl$x$options$columnDefs, col_defs)
  }
  htmltools::tagList(tbl)
}

cfg <- params$config
lipid_cfg <- cfg$modes$lipidomics

# Load AI commentary if available
commentary <- NULL
has_commentary <- FALSE
if (!is.null(params$commentary_file) && file.exists(params$commentary_file)) {
    commentary <- tryCatch(
        jsonlite::read_json(params$commentary_file),
        error = function(e) NULL
    )
    has_commentary <- !is.null(commentary) && length(commentary) > 0
}

render_commentary <- function(figure_id) {
    if (!has_commentary) return(invisible(NULL))
    c <- commentary[[figure_id]]
    if (is.null(c)) return(invisible(NULL))

    cat("<details style='margin: 4px 0 18px 0;'>\n")
    cat("<summary style='cursor: pointer; color: #4a90d9; font-weight: 500; font-size: 0.93em;'>Show AI Interpretation</summary>\n")
    html <- "<div style='background: #f8f9fa; border-left: 4px solid #4a90d9; padding: 12px 16px; margin: 6px 0 4px 0; border-radius: 0 4px 4px 0; font-size: 0.92em;'>\n"
    html <- paste0(html, "<p style='margin: 0 0 8px 0;'><strong>What this shows:</strong> ",
                   htmltools::htmlEscape(c$what_is_this %||% ""), "</p>\n")
    bio_interp <- c$biological_interpretation %||% ""
    if (nzchar(bio_interp)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Interpretation:</strong> ",
                       htmltools::htmlEscape(bio_interp), "</p>\n")
    }
    obs <- c$key_observations %||% c$observations
    if (!is.null(obs) && length(obs) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0 4px 0;'><strong>Key observations:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (ob in obs) html <- paste0(html, "<li>", htmltools::htmlEscape(ob), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }
    stat_note <- c$statistical_note %||% c$statistical_assessment %||% ""
    if (nzchar(stat_note)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Statistical note:</strong> ",
                       htmltools::htmlEscape(stat_note), "</p>\n")
    }
    issues <- c$issues_checks
    if (!is.null(issues) && length(issues) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0 4px 0;'><strong>Checks:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (issue in issues) html <- paste0(html, "<li>", htmltools::htmlEscape(issue), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }
    steps <- c$next_steps
    if (!is.null(steps) && length(steps) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0 4px 0;'><strong>Next steps:</strong></p>\n<ul style='margin: 2px 0 4px 0; padding-left: 20px;'>\n")
        for (s in steps) html <- paste0(html, "<li>", htmltools::htmlEscape(s), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }
    # Attribution: Claude (Anthropic) + K-Dense AI scientific-skills
    backend <- c$backend %||% "unknown"
    confidence <- c$confidence %||% ""
    backend_label <- if (grepl("claude", backend, ignore.case = TRUE)) {
        "Claude (Anthropic) + K-Dense AI scientific-skills"
    } else if (grepl("openai", backend, ignore.case = TRUE)) {
        "OpenAI + K-Dense AI scientific-skills"
    } else if (backend == "data_driven") {
        "data-driven + K-Dense AI scientific-skills"
    } else {
        backend
    }
    if (nzchar(confidence) && confidence != "none") {
        backend_label <- paste0(backend_label, " | confidence: ", confidence)
    }
    html <- paste0(html, "<p style='margin: 4px 0 0 0; color: #888; font-size: 0.85em;'><em>Generated by: ",
                   htmltools::htmlEscape(backend_label), "</em></p>\n</div>\n")
    cat(html)
    cat("</details>\n")
}
```

# Summary {.tabset}

```{r summary}
pre <- params$pre
qc_res <- params$qc_res
de_res <- params$de_res
rf_res <- params$rf_res
plsda_res <- params$plsda_res
class_res <- params$class_res

n_features <- if (!is.null(pre)) nrow(pre$expr_work) else NA
n_samples  <- if (!is.null(pre)) ncol(pre$expr_work) else NA
n_classes  <- if (!is.null(pre$info$n_lipid_classes)) pre$info$n_lipid_classes else NA

norm_method <- "N/A"
if (!is.null(pre$info$normalization)) {
  ni <- pre$info$normalization
  parts <- character(0)
  if (!is.null(ni$sample_norm) && tolower(ni$sample_norm) != "none")
    parts <- c(parts, toupper(ni$sample_norm))
  if (!is.null(ni$transform) && tolower(ni$transform) != "none")
    parts <- c(parts, ni$transform)
  if (!is.null(ni$scaling) && tolower(ni$scaling) != "none")
    parts <- c(parts, ni$scaling)
  if (length(parts) > 0) norm_method <- paste(parts, collapse = " + ")
}

de_method <- if (!is.null(de_res$method)) de_res$method else "N/A"

module_status <- data.frame(
  Module = c("Preprocessing", "QC Diagnostics", "Differential Expression",
             "Random Forest", "PLS-DA", "Lipid Class Analysis"),
  Status = c(
    if (!is.null(pre)) "Completed" else "Skipped",
    if (!is.null(qc_res)) "Completed" else "Skipped",
    if (!is.null(de_res) && !is.null(de_res$summary_df)) "Completed" else "Skipped",
    if (!is.null(rf_res)) "Completed" else "Skipped",
    if (!is.null(plsda_res)) "Completed" else "Skipped",
    if (!is.null(class_res)) "Completed" else "Skipped"
  ),
  stringsAsFactors = FALSE
)
```

| Parameter | Value |
|-----------|-------|
| Lipid Features | `r n_features` |
| Samples   | `r n_samples` |
| Lipid Classes | `r n_classes` |
| Normalization | `r norm_method` |
| DE method | `r de_method` |

```{r summary-table}
knitr::kable(module_status, align = "ll")
```

```{r class-count-table}
if (!is.null(pre$info$lipid_class_counts)) {
  cc <- pre$info$lipid_class_counts
  cc_df <- data.frame(
    Lipid_Class = names(cc),
    Count = as.integer(unlist(cc)),
    stringsAsFactors = FALSE
  )
  cc_df <- cc_df[order(-cc_df$Count), ]
  knitr::kable(cc_df, caption = "Lipid species per class", row.names = FALSE)
}
```

---

`r if (is.null(qc_res)) "<!--"`

# QC Diagnostics {.tabset}

## PCA

```{r qc-pca, fig.width=9, fig.height=7, results='asis'}
if (!is.null(qc_res$plots)) {
  pca_keys <- grep("^pca_1_2", names(qc_res$plots), value = TRUE)
  pca_main <- pca_keys[!grepl("_noQC$", pca_keys)][1]
  if (!is.na(pca_main)) { safe_print_plot(qc_res$plots[[pca_main]]); render_commentary("pca_plot") }

  pca13_keys <- grep("^pca_1_3", names(qc_res$plots), value = TRUE)
  pca13_main <- pca13_keys[!grepl("_noQC$", pca13_keys)][1]
  if (!is.na(pca13_main)) { safe_print_plot(qc_res$plots[[pca13_main]]); render_commentary("pca_plot_13") }
}
```

## Density

```{r qc-density, fig.width=9, fig.height=6, results='asis'}
if (!is.null(qc_res$plots)) {
  safe_print_plot(qc_res$plots[["density_raw"]])
  render_commentary("density_raw")
  safe_print_plot(qc_res$plots[["density_norm"]])
  render_commentary("density_normalized")
}
```

## Boxplots

```{r qc-boxplots, fig.width=10, fig.height=6, results='asis'}
if (!is.null(qc_res$plots)) {
  safe_print_plot(qc_res$plots[["boxplot_raw"]])
  render_commentary("boxplot_raw")
  safe_print_plot(qc_res$plots[["boxplot"]])
  render_commentary("boxplot_normalized")
}
```

## Heatmaps

```{r qc-heatmaps, out.width="100%", results='asis'}
if (!is.null(qc_res$files)) {
  embed_png(qc_res$files, "sample_distance_heatmap\\.png$")
  render_commentary("sample_distance")
  embed_png(qc_res$files, "sample_correlation_heatmap\\.png$")
  render_commentary("sample_correlation")
}
```

`r if (is.null(qc_res)) "-->"`

---

`r if (is.null(params$qc_enhanced)) "<!--"`

# Enhanced QC {.tabset}

## PCA Scree

```{r eqc-scree, fig.width=8, fig.height=6, results='asis'}
if (!is.null(params$qc_enhanced$plots$pca_scree)) {
  safe_print_plot(params$qc_enhanced$plots$pca_scree)
  render_commentary("pca_scree")
}
```

## PCA Loadings

```{r eqc-loading, fig.width=9, fig.height=7, results='asis'}
if (!is.null(params$qc_enhanced$plots$pca_loading)) {
  safe_print_plot(params$qc_enhanced$plots$pca_loading)
  render_commentary("pca_loading")
}
```

## Category Donut

```{r eqc-donut, fig.width=8, fig.height=8, results='asis'}
if (!is.null(params$qc_enhanced$plots$category_donut)) {
  safe_print_plot(params$qc_enhanced$plots$category_donut)
  render_commentary("category_donut")
}
```

## Class Correlation

```{r eqc-class-cor, fig.width=10, fig.height=8, results='asis'}
if (!is.null(params$qc_enhanced$plots$class_correlation)) {
  safe_print_plot(params$qc_enhanced$plots$class_correlation)
  render_commentary("class_correlation")
}
```

## PERMANOVA

```{r eqc-permanova, results='asis'}
perm <- params$qc_enhanced$permanova
if (!is.null(perm)) {
  perm_df <- data.frame(
    Statistic = c("F-statistic", "R-squared", "P-value"),
    Value = c(
      round(perm$F_statistic %||% NA, 2),
      round(perm$R2 %||% NA, 2),
      sprintf("%.2e", perm$p_value %||% NA)
    ),
    stringsAsFactors = FALSE
  )
  knitr::kable(perm_df, caption = "PERMANOVA Results")
} else {
  cat("*PERMANOVA not available — the vegan package is not installed.*\n\n")
}
```

`r if (is.null(params$qc_enhanced)) "-->"`

---

`r if (is.null(de_res) || is.null(de_res$summary_df)) "<!--"`

# Differential Expression {.tabset}

```{r de-setup}
de_contrasts <- NULL
if (!is.null(de_res) && !is.null(de_res$summary_df)) {
  de_cfg <- lipid_cfg$de %||% list()
  de_contrasts <- de_cfg$contrasts
  if (is.list(de_contrasts)) de_contrasts <- unlist(de_contrasts)
}
```

```{r de-note, results='asis'}
if (!is.null(de_res) && !is.null(de_res$summary_df)) {
  n_sig_any <- sum(de_res$summary_df$pass_any_contrast == 1, na.rm = TRUE)
  if (n_sig_any == 0) {
    n_samp <- ncol(pre$expr_work)
    cat("\n> **Note:** No lipids reached significance (adjusted p < ",
        lipid_cfg$de$p_cutoff %||% 0.05, " and |log2FC| >= ",
        round(log2(lipid_cfg$de$linear_fc_cutoff %||% 1.5), 2),
        "). With only **", n_samp, " samples**, consider relaxing thresholds ",
        "or using a rank-based method.\n\n", sep = "")
  }
}
```

```{r de-tabs, results='asis'}
if (!is.null(de_contrasts)) {
  for (ctr in de_contrasts) {
    ctr_label <- make_contrast_label(ctr)

    cat("\n## ", ctr, " {.tabset}\n\n", sep = "")

    p_volcano <- de_res$plots[[paste0("volcano_", ctr_label)]]
    if (!is.null(p_volcano)) {
      cat("### Volcano\n\n")
      print(p_volcano)
      render_commentary(paste0("volcano_", ctr_label))
      cat("\n\n")
    }

    p_ma <- de_res$plots[[paste0("ma_", ctr_label)]]
    if (!is.null(p_ma)) {
      cat("### MA Plot\n\n")
      print(p_ma)
      render_commentary(paste0("ma_", ctr_label))
      cat("\n\n")
    }

    p_phist <- de_res$plots[[paste0("pval_hist_", ctr_label)]]
    if (!is.null(p_phist)) {
      cat("### P-value Distribution\n\n")
      print(p_phist)
      render_commentary(paste0("pvalue_hist_", ctr_label))
      cat("\n\n")
    }

    # Interactive volcano
    ctr_tbl <- tryCatch(
      extract_contrast_table(de_res$summary_df, ctr_label),
      error = function(e) NULL
    )
    if (!is.null(ctr_tbl) &&
        requireNamespace("plotly", quietly = TRUE) &&
        requireNamespace("crosstalk", quietly = TRUE)) {

      name_map_v <- NULL
      if (!is.null(pre$row_data)) {
        nm_col <- if ("display_name" %in% colnames(pre$row_data)) "display_name" else "Name"
        if (nm_col %in% colnames(pre$row_data)) {
          name_map_v <- stats::setNames(
            as.character(pre$row_data[[nm_col]]),
            as.character(pre$row_data$feature_id)
          )
        }
      }

      vdf <- ctr_tbl
      vdf$Name <- if (!is.null(name_map_v)) {
        ifelse(is.na(name_map_v[vdf$feature_id]),
               vdf$feature_id, name_map_v[vdf$feature_id])
      } else {
        vdf$feature_id
      }
      vdf$neglog10_FDR <- -log10(pmax(
        ifelse(is.na(vdf$adj.P.Val), 1, vdf$adj.P.Val), 1e-300
      ))
      vdf$logFC_display <- round(vdf$logFC, 2)
      vdf$FDR_display   <- sprintf("%.2e", vdf$adj.P.Val)

      cat("### Interactive Volcano\n\n")
      cat("*Type lipid names in the search box to highlight them.*\n\n")

      sd <- crosstalk::SharedData$new(vdf, key = ~Name,
                                       group = paste0("volcano_", ctr_label))

      vol_fig <- plotly::plot_ly(
        sd, x = ~logFC, y = ~neglog10_FDR,
        type = "scatter", mode = "markers",
        text = ~paste0(Name, "\nlog2FC: ", logFC_display, "\nFDR: ", FDR_display),
        hoverinfo = "text",
        marker = list(size = 7, opacity = 0.5, color = "grey30")
      )
      vol_fig <- plotly::layout(
        vol_fig,
        title = list(text = paste0("Volcano: ", ctr), font = list(size = 14)),
        xaxis = list(title = "log2 Fold Change"),
        yaxis = list(title = "-log10(FDR)"),
        dragmode = "lasso"
      )
      vol_fig <- plotly::highlight(
        vol_fig,
        on = "plotly_selected", off = "plotly_deselect",
        selectize = TRUE, persistent = TRUE,
        color = "dodgerblue", opacityDim = 0.15,
        selected = plotly::attrs_selected(
          marker = list(size = 10, opacity = 1), showlegend = FALSE
        )
      )

      print(htmltools::tagList(vol_fig))
      cat("\n\n")
    }

    # Top results table
    if (!is.null(ctr_tbl)) {
      ctr_tbl <- ctr_tbl[order(ctr_tbl$P.Value), ]
      top_tbl <- head(ctr_tbl, 50)
      cat("### Top Results\n\n")
      print(render_dt(top_tbl, caption = paste0("Top DE lipids: ", ctr)))
      cat("\n\n")
    }
  }
}
```

`r if (is.null(de_res) || is.null(de_res$summary_df)) "-->"`

---

`r if (is.null(class_res)) "<!--"`

# Lipid Class Analysis {.tabset}

## Class Composition

```{r class-comp, fig.width=10, fig.height=7, results='asis'}
if (!is.null(class_res$plots$class_normalized)) {
  safe_print_plot(class_res$plots$class_normalized)
  render_commentary("class_normalized")
}
if (!is.null(class_res$plots$class_absolute)) {
  safe_print_plot(class_res$plots$class_absolute)
  render_commentary("class_absolute")
}
```

## Chain Saturation

Lipids are grouped by the number of carbon-carbon double bonds (DB) in their fatty acyl chains:

- **SFA — Saturated Fatty Acids (0 DB):** no double bonds; chains pack tightly, increasing membrane rigidity and contributing to lipid storage.
- **MUFA — Monounsaturated Fatty Acids (1 DB):** a single double bond introduces a kink that increases membrane fluidity; common in dietary oils (e.g. oleic acid).
- **PUFA — Polyunsaturated Fatty Acids (2+ DB):** multiple double bonds further increase fluidity and serve as precursors for signaling lipids (eicosanoids, resolvins). Includes essential omega-3 and omega-6 fatty acids.

**Biological interpretation of shifts between groups:**

| Category | ↑ Increase typically reflects | ↓ Decrease typically reflects |
|----------|-------------------------------|-------------------------------|
| **SFA**  | Greater membrane rigidity; active de novo lipogenesis; pro-inflammatory signaling (e.g. palmitate → TLR4); commonly seen with high-fat diet, lipotoxicity, insulin resistance. | Increased membrane fluidity; reduced lipogenic flux; generally a metabolically healthier profile. |
| **MUFA** | Elevated SCD1 (Δ9-desaturase) activity converting SFA → MUFA; adaptive / cytoprotective response to lipid stress; oleic-acid-rich (Mediterranean-style) diets; anti-inflammatory tendency. | Reduced desaturation capacity; possible accumulation of toxic SFA species. |
| **PUFA** | Larger pool of essential fatty acids and signaling-lipid precursors (eicosanoids, resolvins); higher membrane fluidity; ω-3-driven shifts often anti-inflammatory; greater susceptibility to lipid peroxidation. | Reduced essential FA availability or dietary deficiency; smaller signaling-lipid pool; reduced membrane fluidity; possible oxidative losses. |

```{r chain-sat, fig.width=8, fig.height=6, results='asis'}
if (!is.null(class_res$plots$chain_saturation)) {
  safe_print_plot(class_res$plots$chain_saturation)
  render_commentary("chain_saturation")
}
```

## Chain Length

```{r chain-len, fig.width=8, fig.height=6, results='asis'}
if (!is.null(class_res$plots$chain_length)) {
  safe_print_plot(class_res$plots$chain_length)
  render_commentary("chain_length")
}
```

## Per-Class Chain Profiles {.tabset}

```{r chain-profiles, fig.width=10, fig.height=6, results='asis'}
if (!is.null(class_res$chain_profiles)) {
  classes <- names(class_res$chain_profiles)
  for (cls in classes) {
    cat(paste0("\n### ", cls, "\n\n"))
    p <- class_res$plots[[paste0("chain_profile_", cls)]]
    if (!is.null(p)) {
      safe_print_plot(p)
      render_commentary(paste0("chain_profile_", cls))
    }
    cat("\n\n")
  }
}
```

## Class Enrichment (ORA)

```{r class-ora, fig.width=10, fig.height=7, results='asis'}
if (!is.null(class_res$plots$class_ora)) {
  safe_print_plot(class_res$plots$class_ora)
  render_commentary("class_ora")
}
```

```{r class-ora-table}
if (!is.null(class_res$class_ora)) {
  render_dt(class_res$class_ora, caption = "Lipid Class Over-Representation Analysis")
}
```

`r if (is.null(class_res)) "-->"`

---

`r if (is.null(rf_res) && is.null(plsda_res)) "<!--"`

# Feature Selection {.tabset}

`r if (is.null(rf_res)) "<!--"`

## Random Forest {.tabset}

### Importance Plot

```{r rf-plot, fig.width=9, fig.height=7, results='asis'}
if (!is.null(rf_res$plots$rf_importance)) {
  safe_print_plot(rf_res$plots$rf_importance)
  render_commentary("rf_importance")
}
```

### Importance Table

```{r rf-table}
if (!is.null(rf_res$importance_df)) {
  imp_df <- rf_res$importance_df
  sort_col <- if ("MeanDecreaseAccuracy" %in% names(imp_df)) "MeanDecreaseAccuracy" else "importance"
  imp_df <- imp_df[order(-imp_df[[sort_col]]), ]
  render_dt(head(imp_df, 50), caption = "RF Feature Importance (top 50)")
}
```

`r if (is.null(rf_res)) "-->"`

`r if (is.null(plsda_res)) "<!--"`

## PLS-DA {.tabset}

### Scores Plot

```{r plsda-scores, fig.width=9, fig.height=7, results='asis'}
if (!is.null(plsda_res$plots$plsda_scores)) {
  safe_print_plot(plsda_res$plots$plsda_scores)
  render_commentary("plsda_scores")
}
```

### VIP Plot

```{r plsda-vip-plot, fig.width=9, fig.height=7, results='asis'}
if (!is.null(plsda_res$plots$plsda_vip)) {
  safe_print_plot(plsda_res$plots$plsda_vip)
  render_commentary("plsda_vip")
}
```

### VIP Table

```{r plsda-vip-table}
if (!is.null(plsda_res$vip_df)) {
  vip_df <- plsda_res$vip_df
  vip_col <- if ("VIP" %in% names(vip_df)) "VIP" else grep("^VIP", names(vip_df), value = TRUE)[1]
  vip_df <- vip_df[order(-vip_df[[vip_col]]), ]
  render_dt(head(vip_df, 50), caption = "PLS-DA VIP Scores (top 50)")
}
```

`r if (is.null(plsda_res)) "-->"`

`r if (is.null(rf_res) && is.null(plsda_res)) "-->"`

---

`r if (is.null(params$biomarker_res)) "<!--"`

# Biomarker Discovery {.tabset}

## Biomarker Table

```{r biomarker-table-plot, fig.width=10, fig.height=8, results='asis'}
if (!is.null(params$biomarker_res$plots$biomarker_table)) {
  safe_print_plot(params$biomarker_res$plots$biomarker_table)
  render_commentary("biomarker_table")
}
```

## ROC Curves

```{r biomarker-roc, fig.width=9, fig.height=7, results='asis'}
if (!is.null(params$biomarker_res$plots$roc_curves)) {
  safe_print_plot(params$biomarker_res$plots$roc_curves)
  render_commentary("roc_curves")
}
```

## Multivariate ROC

```{r biomarker-roc-multi, fig.width=9, fig.height=7, results='asis'}
if (!is.null(params$biomarker_res$plots$roc_multivariate)) {
  safe_print_plot(params$biomarker_res$plots$roc_multivariate)
  render_commentary("roc_multivariate")
}
```

## Top Candidates

```{r biomarker-candidates}
if (!is.null(params$biomarker_res$biomarker_df)) {
  bm_df <- params$biomarker_res$biomarker_df
  bm_df <- head(bm_df, 20)
  render_dt(bm_df, caption = "Top 20 Biomarker Candidates")
}
```

`r if (is.null(params$biomarker_res)) "-->"`

---

`r if (is.null(params$pathway_res)) "<!--"`

# Lipid Pathway Analysis {.tabset}

## Pathway Network

```{r pathway-network, fig.width=10, fig.height=8, results='asis'}
if (!is.null(params$pathway_res$plots$pathway_network)) {
  safe_print_plot(params$pathway_res$plots$pathway_network)
  render_commentary("pathway_network")
}
```

## Pathway Barplot

```{r pathway-barplot, fig.width=9, fig.height=7, results='asis'}
if (!is.null(params$pathway_res$plots$pathway_barplot)) {
  safe_print_plot(params$pathway_res$plots$pathway_barplot)
  render_commentary("pathway_barplot")
}
```

## Pathway Scores

```{r pathway-scores}
if (!is.null(params$pathway_res$pathway_scores)) {
  render_dt(params$pathway_res$pathway_scores, caption = "Lipid Pathway Scores")
}
```

## Predicted Enzymes

```{r pathway-enzymes}
if (!is.null(params$pathway_res$enzyme_predictions)) {
  render_dt(params$pathway_res$enzyme_predictions, caption = "Predicted Enzymes")
}
```

## Enzyme Network

```{r pathway-enzyme-network, fig.width=10, fig.height=8, results='asis'}
if (!is.null(params$pathway_res$plots$enzyme_network)) {
  safe_print_plot(params$pathway_res$plots$enzyme_network)
  render_commentary("enzyme_network")
}
```

`r if (is.null(params$pathway_res)) "-->"`

---

# Session Info

<details>
<summary>Click to expand</summary>

```{r session-info}
sessionInfo()
```

</details>
