---
title: "`r local({ cfg <- params$config; pname <- cfg$project$name %||% ''; analyst <- gsub('_', ' ', cfg$project$analyst %||% ''); parts <- c('Lipidomics Analysis Report'); if (nzchar(pname)) parts <- c(parts, pname); if (nzchar(analyst)) parts <- c(parts, analyst); paste(parts, collapse = ' - ') })`"
subtitle: "multiomics-core Pipeline"
author: "`r paste0('Bioinformatician: ', gsub('_', ' ', params$config$project$analyst %||% ''))`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    fig_width: 10
    fig_height: 7
    self_contained: true
    code_folding: hide
    df_print: paged
params:
  pre: NULL
  qc_res: NULL
  de_res: NULL
  rf_res: NULL
  plsda_res: NULL
  class_res: NULL
  clustering_res: NULL
  biomarker_res: NULL
  pathway_res: NULL
  qc_enhanced: NULL
  config: NULL
  commentary_file: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%"
)

library(ggplot2)
library(DT)
library(htmltools)

if (requireNamespace("plotly", quietly = TRUE))   library(plotly)
if (requireNamespace("crosstalk", quietly = TRUE)) library(crosstalk)
if (requireNamespace("jsonlite", quietly = TRUE))  library(jsonlite)

# Source domain helpers needed for report
make_contrast_label <- function(contrast_str) {
  parts <- strsplit(contrast_str, " - ", fixed = TRUE)[[1]]
  paste(trimws(parts), collapse = "_vs_")
}

extract_contrast_table <- function(summary_df, contrast) {
  linear_fc <- summary_df[[paste0("linearFC.", contrast)]]
  logfc <- ifelse(linear_fc >= 0, log2(linear_fc), -log2(abs(linear_fc)))
  data.frame(
    feature_id = summary_df$feature_id,
    logFC      = logfc,
    AveExpr    = summary_df[[paste0("AveExpr.", contrast)]],
    P.Value    = summary_df[[paste0("P.Value.", contrast)]],
    adj.P.Val  = summary_df[[paste0("adj.P.Val.", contrast)]],
    stringsAsFactors = FALSE
  )
}

# Source interactive cutoff panel helpers
tryCatch({
  .cp_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
  .cp_found <- FALSE
  for (.cp_i in 1:6) {
    .cp_try <- file.path(.cp_dir, "R", "core", "17_cutoff_panel.R")
    if (file.exists(.cp_try)) { source(.cp_try); .cp_found <- TRUE; break }
    .cp_dir <- dirname(.cp_dir)
  }
  if (!.cp_found) message("cutoff_panel helpers not found (walked 6 levels up)")
}, error = function(e) message("cutoff_panel helpers not found: ", e$message))

# ---------------------------------------------------------------------------
# Helper: downloadable DT table with CSV/Excel export (proteomics style)
# ---------------------------------------------------------------------------
make_dt <- function(df, caption = NULL, pageLength = 15) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  safe_caption <- gsub("[^a-zA-Z0-9]", "_", caption %||% "data")
  tbl <- DT::datatable(
    df,
    caption  = caption,
    extensions = "Buttons",
    options = list(
      scrollX    = TRUE,
      pageLength = pageLength,
      dom        = "Bfrtip",
      buttons    = list(
        list(extend = "csv",   filename = safe_caption),
        list(extend = "excel", filename = safe_caption)
      )
    ),
    rownames = FALSE,
    class    = "compact stripe hover"
  )
  # Format numeric columns: scientific notation for p-values, 2 decimals for the rest
  num_cols <- names(df)[vapply(df, is.numeric, logical(1))]
  pval_pattern <- "^(p[._]?val|p[._]?adj|padj|adj[._]?p|fdr|q[._]?val|BH|bonferroni|p\\.value|adj\\.P\\.Val|PValue|FDR)"
  pval_cols <- grep(pval_pattern, num_cols, ignore.case = TRUE, value = TRUE)
  other_cols <- setdiff(num_cols, pval_cols)
  if (length(other_cols) > 0) {
    tbl <- DT::formatRound(tbl, columns = other_cols, digits = 2)
  }
  if (length(pval_cols) > 0) {
    pval_idx <- which(names(df) %in% pval_cols) - 1L
    sci_render <- DT::JS(
      "function(data, type, row) {",
      "  if (type === 'display' && data !== null && data !== '') {",
      "    return parseFloat(data).toExponential(2);",
      "  }",
      "  return data;",
      "}"
    )
    col_defs <- lapply(pval_idx, function(i) list(targets = i, render = sci_render))
    tbl$x$options$columnDefs <- c(tbl$x$options$columnDefs, col_defs)
  }
  htmltools::tagList(tbl)
}

# ---------------------------------------------------------------------------
# Helper: paper-style figure legend below plots (matching proteomics)
# ---------------------------------------------------------------------------
figure_legend <- function(text) {
  cat(sprintf("<p class='figure-legend' style='text-align: center; font-size: 0.9em; color: #555; margin: 4px auto 12px auto; max-width: 88%%; line-height: 1.45; font-style: italic;'>%s</p>\n", text))
}

# ---------------------------------------------------------------------------
# Helper: safely print a ggplot or return invisible(NULL)
# ---------------------------------------------------------------------------
safe_print_plot <- function(p) {
  if (inherits(p, "gg") || inherits(p, "ggplot")) {
    print(p)
  } else {
    invisible(NULL)
  }
}

# ---------------------------------------------------------------------------
# Helper: embed a PNG from a file path vector by basename pattern
# ---------------------------------------------------------------------------
embed_png <- function(files, pattern, fig_width = "100%") {
  if (is.null(files) || length(files) == 0) return(invisible(NULL))
  hits <- files[grepl(pattern, basename(files), ignore.case = TRUE)]
  hits <- hits[file.exists(hits)]
  if (length(hits) == 0) return(invisible(NULL))
  knitr::include_graphics(hits[1])
}

# ---------------------------------------------------------------------------
# Load AI commentary if available
# ---------------------------------------------------------------------------
commentary <- NULL
has_commentary <- FALSE
if (!is.null(params$commentary_file) && file.exists(params$commentary_file)) {
    commentary <- tryCatch(
        jsonlite::read_json(params$commentary_file),
        error = function(e) NULL
    )
    has_commentary <- !is.null(commentary) && length(commentary) > 0
}

# Helper: render commentary block for a figure_id (collapsible)
render_commentary <- function(figure_id) {
    if (!has_commentary) return(invisible(NULL))
    c <- commentary[[figure_id]]
    if (is.null(c)) return(invisible(NULL))

    cat("<details class='commentary-details' style='margin: 4px 0 18px 0;'>\n")
    cat("<summary style='cursor: pointer; color: #4a90d9; font-weight: 500; font-size: 0.93em; padding: 3px 0;'>Show AI Interpretation</summary>\n")

    html <- "<div class='commentary-block' style='background: #f8f9fa; border-left: 4px solid #4a90d9; padding: 12px 16px; margin: 6px 0 4px 0; border-radius: 0 4px 4px 0; font-size: 0.92em;'>\n"
    html <- paste0(html, "<p style='margin: 0 0 8px 0;'><strong>What this shows:</strong> ",
                   htmltools::htmlEscape(c$what_is_this %||% ""), "</p>\n")

    bio_interp <- c$biological_interpretation %||% ""
    if (nzchar(bio_interp)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Interpretation:</strong> ",
                       htmltools::htmlEscape(bio_interp), "</p>\n")
    }

    # Support both field names: key_observations (lipidomics) and observations (rnaseq)
    obs <- c$key_observations %||% c$observations
    if (!is.null(obs) && length(obs) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0 4px 0;'><strong>Key observations:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (ob in obs) {
            html <- paste0(html, "<li>", htmltools::htmlEscape(ob), "</li>\n")
        }
        html <- paste0(html, "</ul>\n")
    }

    stat_note <- c$statistical_note %||% c$statistical_assessment %||% ""
    if (nzchar(stat_note)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Statistical note:</strong> ",
                       htmltools::htmlEscape(stat_note), "</p>\n")
    }

    # Issues / checks
    issues <- c$issues_checks
    if (!is.null(issues) && length(issues) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0 4px 0;'><strong>Checks:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (issue in issues) html <- paste0(html, "<li>", htmltools::htmlEscape(issue), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    checks <- c$checks
    if (!is.null(checks) && is.list(checks) && length(checks) > 0 && !is.null(names(checks))) {
        html <- paste0(html, "<p style='margin: 4px 0 4px 0; color: #888; font-size: 0.88em;'>Checks: ")
        check_parts <- vapply(names(checks), function(k) {
            val <- checks[[k]]
            icon <- if (isTRUE(val)) "&#9989;" else "&#10060;"
            paste0(icon, " ", gsub("_", " ", k))
        }, character(1))
        html <- paste0(html, paste(check_parts, collapse = " &nbsp; "), "</p>\n")
    }

    # Next steps
    steps <- c$next_steps
    if (!is.null(steps) && length(steps) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0 4px 0;'><strong>Next steps:</strong></p>\n<ul style='margin: 2px 0 4px 0; padding-left: 20px;'>\n")
        for (s in steps) html <- paste0(html, "<li>", htmltools::htmlEscape(s), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    # Attribution footer: Claude (Anthropic) + K-Dense AI scientific-skills
    backend <- c$backend %||% "unknown"
    confidence <- c$confidence %||% ""
    backend_label <- if (grepl("claude", backend, ignore.case = TRUE)) {
        "Claude (Anthropic) + K-Dense AI scientific-skills"
    } else if (grepl("openai", backend, ignore.case = TRUE)) {
        "OpenAI + K-Dense AI scientific-skills"
    } else if (backend == "data_driven") {
        "data-driven + K-Dense AI scientific-skills"
    } else {
        backend
    }
    if (nzchar(confidence) && confidence != "none") {
        backend_label <- paste0(backend_label, " | confidence: ", confidence)
    }
    html <- paste0(html, "<p style='margin: 4px 0 0 0; color: #888; font-size: 0.85em;'><em>Generated by: ",
                   htmltools::htmlEscape(backend_label), "</em></p>\n</div>\n")

    cat(html)
    cat("</details>\n")
}

# ---------------------------------------------------------------------------
# Unpack config and module results
# ---------------------------------------------------------------------------
cfg      <- params$config
lipid_cfg <- cfg$modes$lipidomics

# Section visibility flags (default-off for MA + p-value distribution after
# Apr 2026 reviewer feedback; toggle on via report_sections in config).
.lip_rs        <- lipid_cfg$report_sections %||% list()
.lip_de_subs   <- .lip_rs$de %||% list()
show_de_ma_lip       <- isTRUE(.lip_de_subs$ma %||% FALSE)
show_de_pvalhist_lip <- isTRUE(.lip_de_subs$pval_hist %||% FALSE)

# Top-level section visibility: each defaults to shown unless the user
# explicitly sets `<key>: false` in report_sections. Sub-keyed entries
# (e.g. `de: {ma: true}`) keep the section shown — fine-grained toggles
# live below where sub-keys are inspected.
.show_section_lip <- function(key, default = TRUE) {
    v <- .lip_rs[[key]]
    if (is.null(v)) return(isTRUE(default))
    !identical(v, FALSE)
}
show_executive_lip      <- .show_section_lip("executive_summary")
show_module_status_lip  <- .show_section_lip("module_status")
show_technical_summary_lip <- .show_section_lip("technical_summary")
show_qc_lip             <- .show_section_lip("qc")
show_de_lip             <- .show_section_lip("de")
show_de_heatmaps_lip    <- .show_section_lip("de_heatmaps")
show_class_lip          <- .show_section_lip("class_analysis")
show_fs_lip             <- .show_section_lip("feature_selection")
show_biomarker_lip      <- .show_section_lip("biomarker")
show_pathway_lip        <- .show_section_lip("pathway")
show_output_files_lip   <- .show_section_lip("output_files")
show_methods_lip        <- .show_section_lip("methods")
show_session_lip        <- .show_section_lip("session_info")
```

```{r force-dependencies, include=FALSE}
# Force DT and Plotly JS dependencies to be included in the self-contained HTML
# Note: filter="top" forces noUiSlider dependency (needed for numeric range filters)
DT::datatable(data.frame(x = 1), filter = "top")
if (requireNamespace("plotly", quietly = TRUE)) {
  plotly::plot_ly(data.frame(x = 1, y = 1), x = ~x, y = ~y,
                  type = "scatter", mode = "markers")
}
```

```{r unpack-results, include=FALSE}
pre            <- params$pre
qc_res         <- params$qc_res
de_res         <- params$de_res
rf_res         <- params$rf_res
plsda_res      <- params$plsda_res
class_res      <- params$class_res
clustering_res <- params$clustering_res
biomarker_res  <- params$biomarker_res
pathway_res    <- params$pathway_res
qc_enhanced    <- params$qc_enhanced

n_features <- if (!is.null(pre)) nrow(pre$expr_work) else NA
n_samples  <- if (!is.null(pre)) ncol(pre$expr_work) else NA

# Resolve group / condition column
cond_col <- lipid_cfg$de$condition_column %||%
            lipid_cfg$effects$color %||% "sample_type"
n_groups <- NA
if (!is.null(pre$meta) && cond_col %in% colnames(pre$meta)) {
  n_groups <- length(unique(pre$meta[[cond_col]]))
}

# Normalization summary
norm_method <- "N/A"
if (!is.null(pre$info$normalization)) {
  ni    <- pre$info$normalization
  parts <- character(0)
  if (!is.null(ni$sample_norm) && tolower(ni$sample_norm) != "none")
    parts <- c(parts, toupper(ni$sample_norm))
  if (!is.null(ni$transform) && tolower(ni$transform) != "none")
    parts <- c(parts, ni$transform)
  if (!is.null(ni$scaling) && tolower(ni$scaling) != "none")
    parts <- c(parts, ni$scaling)
  if (length(parts) > 0) norm_method <- paste(parts, collapse = " + ")
}

de_method <- if (!is.null(de_res$method)) de_res$method else "N/A"

# DE contrasts
de_contrasts <- NULL
if (!is.null(de_res) && !is.null(de_res$summary_df)) {
  de_cfg       <- lipid_cfg$de %||% list()
  de_contrasts <- de_cfg$contrasts
  if (is.list(de_contrasts)) de_contrasts <- unlist(de_contrasts)
}

# Lipid classes count
n_classes <- NA
if (!is.null(class_res) && !is.null(class_res$class_summary)) {
  n_classes <- nrow(class_res$class_summary)
}
```

# Executive Summary

```{r executive-summary, results='asis'}
# ---- Gradient panel (matching proteomics style) --------------------------
cat("<div style='background: linear-gradient(135deg, #f8f9fa 0%, #e9ecef 100%); border: 1px solid #dee2e6; border-radius: 8px; padding: 20px 24px; margin: 10px 0 20px 0;'>\n")
cat("<div style='display: flex; flex-wrap: wrap; gap: 16px; margin-bottom: 16px;'>\n")

# Metric card helper
metric_card <- function(value, label, color = "#2c3e50") {
  cat(sprintf(
    "<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: %s;'>%s</div><div style='font-size: 0.85em; color: #666;'>%s</div></div>\n",
    color, as.character(value), label
  ))
}

# Samples
if (!is.na(n_samples)) metric_card(n_samples, "Samples")

# Groups
if (!is.na(n_groups)) metric_card(n_groups, "Groups")

# Lipid Features
if (!is.na(n_features)) metric_card(format(n_features, big.mark = ","), "Lipid Features")

# Lipid Classes
if (!is.na(n_classes)) metric_card(n_classes, "Lipid Classes", color = "#8e44ad")

# Contrasts
n_contrasts <- length(de_contrasts)
if (n_contrasts > 0) metric_card(n_contrasts, "Contrasts")

# Total DE lipids
n_de_total <- NA
if (!is.null(de_res$summary_df) && "pass_any_contrast" %in% colnames(de_res$summary_df)) {
  n_de_total <- sum(de_res$summary_df$pass_any_contrast == 1, na.rm = TRUE)
  metric_card(format(n_de_total, big.mark = ","), "DE Lipids", color = "#c0392b")
}

# Significant Pathways
n_sig_pathways <- if (!is.null(pathway_res$pathway_scores)) sum(pathway_res$pathway_scores$p_value < 0.05, na.rm = TRUE) else 0
if (n_sig_pathways > 0) metric_card(n_sig_pathways, "Pathways", color = "#16a085")

cat("</div>\n")

# ---- DE breakdown per contrast -------------------------------------------
if (!is.null(de_contrasts) && length(de_contrasts) > 0 && !is.null(de_res$summary_df)) {
  padj_cut <- lipid_cfg$de$p_cutoff %||% 0.05
  if (!is.null(lipid_cfg$de$logfc_cutoff)) {
    log2_lfc <- lipid_cfg$de$logfc_cutoff
  } else {
    lfc_cut  <- lipid_cfg$de$linear_fc_cutoff %||% 1.5
    log2_lfc <- log2(lfc_cut)
  }

  cat("<div style='margin: 12px 0;'>\n")
  cat("<strong>Differential Expression per Contrast:</strong>\n<ul style='margin: 6px 0; padding-left: 20px;'>\n")

  for (ctr in de_contrasts) {
    ctr_label <- make_contrast_label(ctr)
    ctr_tbl   <- tryCatch(extract_contrast_table(de_res$summary_df, ctr_label),
                           error = function(e) NULL)
    if (!is.null(ctr_tbl)) {
      sig   <- !is.na(ctr_tbl$adj.P.Val) & ctr_tbl$adj.P.Val <= padj_cut &
               !is.na(ctr_tbl$logFC) & abs(ctr_tbl$logFC) >= log2_lfc
      n_up   <- sum(sig & ctr_tbl$logFC > 0, na.rm = TRUE)
      n_down <- sum(sig & ctr_tbl$logFC < 0, na.rm = TRUE)
      if (exists("cutoff_summary_span")) {
        cat(sprintf("<li><strong>%s</strong>: %s</li>\n",
                    ctr, cutoff_summary_span(ctr, n_up, n_down, "lipids")))
      } else {
        cat(sprintf("<li><strong>%s</strong>: %d DE lipids (%d up, %d down)</li>\n",
                    ctr, n_up + n_down, n_up, n_down))
      }
    }
  }
  cat("</ul>\n</div>\n")
}

cat("</div>\n")
```

---

# Analysis Parameters {.tabset}

## Project

```{r params-project}
proj <- data.frame(
  Parameter = c("Project Name", "Analyst", "Project Directory",
                "Raw Data Path", "Output Path"),
  Value = c(
    cfg$project$name %||% "N/A",
    gsub("_", " ", cfg$project$analyst %||% "N/A"),
    cfg$project$dir %||% "N/A",
    cfg$paths$raw %||% "N/A",
    cfg$paths$out %||% "N/A"
  ),
  stringsAsFactors = FALSE
)
make_dt(proj, "Project Parameters")
```

## Experimental Design

```{r params-design}
if (!is.null(pre$meta) && cond_col %in% colnames(pre$meta)) {
  design_tbl <- as.data.frame(table(pre$meta[[cond_col]]))
  colnames(design_tbl) <- c("Group", "n (replicates)")
  make_dt(design_tbl, "Experimental Design")
}
```

## Pipeline Settings

```{r params-pipeline}
settings <- data.frame(
  Parameter = c(
    "Normalization",
    "DE Method",
    "Adjusted p-value Cutoff",
    "Linear FC Cutoff",
    "Random Forest",
    "PLS-DA",
    "Lipid Class Analysis"
  ),
  Value = c(
    norm_method,
    de_method,
    as.character(lipid_cfg$de$p_cutoff %||% 0.05),
    as.character(lipid_cfg$de$linear_fc_cutoff %||% 1.5),
    if (!is.null(rf_res)) "Enabled" else "Disabled",
    if (!is.null(plsda_res)) "Enabled" else "Disabled",
    if (!is.null(class_res)) "Enabled" else "Disabled"
  ),
  stringsAsFactors = FALSE
)
make_dt(settings, "Pipeline Settings")
```

## Contrasts

```{r params-contrasts}
if (!is.null(de_contrasts) && length(de_contrasts) > 0) {
  ctr_df <- data.frame(
    `#` = seq_along(de_contrasts),
    Contrast = de_contrasts,
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
  make_dt(ctr_df, "Contrasts")
} else {
  cat("No contrasts defined.\n")
}
```

---

# Module Status

```{r module-status}
module_status <- data.frame(
  Module = c("Preprocessing", "QC Diagnostics", "Enhanced QC",
             "Differential Expression",
             "Random Forest", "PLS-DA", "Lipid Class Analysis",
             "Biomarker Discovery", "Lipid Pathway Analysis"),
  Status = c(
    if (!is.null(pre)) "Completed" else "Skipped",
    if (!is.null(qc_res)) "Completed" else "Skipped",
    if (!is.null(qc_enhanced)) "Completed" else "Skipped",
    if (!is.null(de_res) && !is.null(de_res$summary_df)) "Completed" else "Skipped",
    if (!is.null(rf_res)) "Completed" else "Skipped",
    if (!is.null(plsda_res)) "Completed" else "Skipped",
    if (!is.null(class_res)) "Completed" else "Skipped",
    if (!is.null(biomarker_res)) "Completed" else "Skipped",
    if (!is.null(pathway_res)) "Completed" else "Skipped"
  ),
  stringsAsFactors = FALSE
)
make_dt(module_status, "Module Status", pageLength = 10)
```

---

```{r technical-summary-heading-lip, eval=show_technical_summary_lip, results='asis'}
cat("# Technical Summary\n\n")
```

```{r technical-summary-lip, eval=show_technical_summary_lip, results='asis'}
tech <- lipid_cfg$technical_report

n_samples_ts <- if (!is.null(pre)) ncol(pre$expr_work) else NA
n_groups_ts  <- if (!is.null(pre$meta) && cond_col %in% colnames(pre$meta)) length(unique(pre$meta[[cond_col]])) else NA

cat("## Sample & Acquisition Overview\n\n")
overview <- character(0)
if (!is.na(n_samples_ts) && !is.na(n_groups_ts)) {
    overview <- c(overview, sprintf("This analysis comprises **%d samples** across **%d experimental groups** (grouping variable: *%s*).",
                                     n_samples_ts, n_groups_ts, cond_col))
}
if (!is.null(tech$facility) && nzchar(tech$facility)) {
    overview <- c(overview, sprintf("Data were acquired at the **%s**.", tech$facility))
}
if (length(overview) > 0) cat(paste(overview, collapse = " "), "\n\n")

if (!is.null(tech)) {
    if (!is.null(tech$sample_prep) && nzchar(tech$sample_prep)) {
        cat("## Lipid Extraction & Sample Preparation\n\n", tech$sample_prep, "\n\n")
    }
    if (!is.null(tech$ms_acquisition) && nzchar(tech$ms_acquisition)) {
        cat("## MS Acquisition\n\n", tech$ms_acquisition, "\n\n")
    }
    if (!is.null(tech$search_engine) && nzchar(tech$search_engine)) {
        cat("## Software / Identification\n\n", tech$search_engine, "\n\n")
    }
    if (!is.null(tech$search_parameters) && nzchar(tech$search_parameters)) {
        cat("## Processing Parameters\n\n", tech$search_parameters, "\n\n")
    }
}

cat("## Downstream Analysis (multiomics-core)\n\n")
norm_cfg <- lipid_cfg$normalization %||% list()
cat(sprintf("- **Normalization:** %s + %s + %s\n",
            norm_cfg$sample_norm %||% "none",
            norm_cfg$transform %||% "none",
            norm_cfg$scaling %||% "none"))
if (identical(norm_cfg$sample_norm, "bio_factor") && nzchar(norm_cfg$biological_factor_col %||% "")) {
    cat(sprintf("  - Biological factor column: `%s`\n", norm_cfg$biological_factor_col))
}
cat(sprintf("- **DE method:** %s (p <= %s, |FC| >= %s)\n",
            lipid_cfg$de$method %||% "limma",
            lipid_cfg$de$p_cutoff %||% 0.05,
            lipid_cfg$de$linear_fc_cutoff %||% 1.5))
cat(sprintf("- **Pathway analysis:** %s\n",
            if (isTRUE(lipid_cfg$pathway$enabled)) "enabled" else "disabled"))
```

---

`r if (is.null(qc_res) || !show_qc_lip) "<!--"`

# QC Diagnostics {.tabset}

## PCA {.tabset}

```{r qc-pca-helpers-lip, include=FALSE}
.pick_pca_key_lip <- function(plots, axes_prefix, subset = c("with_qc", "no_qc")) {
  subset <- match.arg(subset)
  keys <- grep(paste0("^", axes_prefix), names(plots), value = TRUE)
  if (subset == "no_qc") keys <- keys[grepl("_noQC$", keys)]
  else                   keys <- keys[!grepl("_noQC$", keys)]
  if (length(keys) == 0) NA_character_ else keys[1]
}

.pca_has_noqc_lip <- any(grepl("_noQC$", names(qc_res$plots %||% list())))
.pca_subsets_lip <- if (.pca_has_noqc_lip) {
  list(list(tab = "With QC samples",    sub = "with_qc"),
       list(tab = "Without QC samples", sub = "no_qc"))
} else {
  list(list(tab = "All samples", sub = "with_qc"))
}
```

### Static {.tabset}

```{r qc-pca, fig.width=9, fig.height=7, results='asis'}
if (!is.null(qc_res$plots)) {
  for (sub in .pca_subsets_lip) {
    cat(sprintf("\n#### %s\n\n", sub$tab))
    pca_main <- .pick_pca_key_lip(qc_res$plots, "pca_1_2", sub$sub)
    if (!is.na(pca_main)) {
      safe_print_plot(qc_res$plots[[pca_main]])
      figure_legend(sprintf(
        "PCA of normalized lipidomics data (PC1 vs PC2) — %s. Each point represents one sample, colored by experimental group.",
        tolower(sub$tab)))
      render_commentary(paste0("pca_plot", if (sub$sub == "no_qc") "_noQC" else ""))
    }
    pca13_main <- .pick_pca_key_lip(qc_res$plots, "pca_1_3", sub$sub)
    if (!is.na(pca13_main)) {
      safe_print_plot(qc_res$plots[[pca13_main]])
      figure_legend(sprintf(
        "PCA projection onto PC1 vs PC3 — %s. Captures additional variance beyond the first two components.",
        tolower(sub$tab)))
      render_commentary(paste0("pca_plot_13", if (sub$sub == "no_qc") "_noQC" else ""))
    }
  }
}
```

### With Sample Names {.tabset}

```{r qc-pca-labeled, fig.width=9, fig.height=7, results='asis'}
if (!is.null(qc_res$plots)) {
  for (sub in .pca_subsets_lip) {
    cat(sprintf("\n#### %s\n\n", sub$tab))
    pca_main <- .pick_pca_key_lip(qc_res$plots, "pca_1_2", sub$sub)
    if (!is.na(pca_main)) {
      p <- qc_res$plots[[pca_main]]
      if ("sample" %in% colnames(p$data)) {
        p_lab <- p + ggplot2::geom_text(ggplot2::aes(label = sample), size = 2.5, vjust = -0.8, show.legend = FALSE)
        safe_print_plot(p_lab)
      }
    }
    pca13_main <- .pick_pca_key_lip(qc_res$plots, "pca_1_3", sub$sub)
    if (!is.na(pca13_main)) {
      p13 <- qc_res$plots[[pca13_main]]
      if ("sample" %in% colnames(p13$data)) {
        p13_lab <- p13 + ggplot2::geom_text(ggplot2::aes(label = sample), size = 2.5, vjust = -0.8, show.legend = FALSE)
        safe_print_plot(p13_lab)
      }
    }
  }
}
```

### Interactive

```{r qc-pca-interactive, results='asis'}
if (!is.null(qc_res$plots) && requireNamespace("plotly", quietly = TRUE)) {
  pca_keys <- grep("^pca_1_2", names(qc_res$plots), value = TRUE)
  pca_main <- pca_keys[!grepl("_noQC$", pca_keys)][1]
  if (!is.na(pca_main)) {
    p <- qc_res$plots[[pca_main]]
    sc <- p$data
    if ("sample" %in% colnames(sc)) {
      # Use condition column for coloring (prefer over sample_id)
      non_pc <- setdiff(colnames(sc), c("sample", "sample_id", grep("^PC", colnames(sc), value = TRUE)))
      color_var <- if (length(non_pc) > 0) non_pc[1] else NA
      sc$hover_text <- paste0("Sample: ", sc$sample,
                              if (!is.na(color_var)) paste0("<br>", color_var, ": ", sc[[color_var]]) else "")
      pl <- plotly::plot_ly(sc, x = ~PC1, y = ~PC2,
                            color = if (!is.na(color_var)) as.formula(paste0("~`", color_var, "`")) else NULL,
                            text = ~hover_text, hoverinfo = "text",
                            type = "scatter", mode = "markers",
                            marker = list(size = 10)) |>
          plotly::layout(title = "PCA: PC1 vs PC2 — Hover for sample names")
      print(htmltools::tagList(pl))
    }
  }
} else {
  cat("*Interactive PCA not available (plotly package missing).*\n")
}
```

## Density

```{r qc-density, fig.width=9, fig.height=6, results='asis'}
if (!is.null(qc_res$plots)) {
  if (!is.null(qc_res$plots[["density_raw"]])) {
    safe_print_plot(qc_res$plots[["density_raw"]])
    figure_legend("Per-sample density of log2(raw intensities) before any normalization. Well-overlapping curves indicate consistent sample loading and acquisition.")
    render_commentary("density_raw")
  }
  ni_dens <- pre$info$normalization
  has_scaling_dens <- !is.null(ni_dens$scaling) && tolower(ni_dens$scaling) != "none"
  if (has_scaling_dens && !is.null(qc_res$plots[["density_transformed"]])) {
    safe_print_plot(qc_res$plots[["density_transformed"]])
    figure_legend("Per-sample density after the transform step. This is the matrix used for differential expression testing. Well-overlapping curves confirm the transform brought samples onto a comparable scale.")
    render_commentary("density_normalized")
  }
  if (!is.null(qc_res$plots[["density_norm"]])) {
    safe_print_plot(qc_res$plots[["density_norm"]])
    if (has_scaling_dens) {
      figure_legend(sprintf(
        "Per-sample density after full normalization (%s + %s). %s scaling standardizes each lipid independently (mean=0, SD=1), so divergent per-sample curves reflect biological differences between groups. This matrix is used for PCA, heatmaps, and multivariate analyses (RF, PLS-DA).",
        ni_dens$transform %||% "", ni_dens$scaling %||% "",
        tools::toTitleCase(ni_dens$scaling %||% "")
      ))
    } else {
      figure_legend("Per-sample density after normalization. Well-aligned curves confirm effective normalization across samples.")
    }
  }
}
```

## Boxplots

```{r qc-boxplots, fig.width=10, fig.height=6, results='asis'}
if (!is.null(qc_res$plots)) {
  if (!is.null(qc_res$plots[["boxplot_raw"]])) {
    safe_print_plot(qc_res$plots[["boxplot_raw"]])
    figure_legend("Per-sample boxplots of log2-transformed raw lipid intensities before normalization. Log2 transformation is applied for visualization to make intensity distributions comparable across the dynamic range.")
    render_commentary("boxplot_raw")
  }
  # Show transform-only boxplot when scaling is applied; otherwise show
  # the fully normalized boxplot (they are identical when no scaling).
  ni_box <- pre$info$normalization
  has_scaling_box <- !is.null(ni_box$scaling) && tolower(ni_box$scaling) != "none"
  if (has_scaling_box && !is.null(qc_res$plots[["boxplot_transformed"]])) {
    safe_print_plot(qc_res$plots[["boxplot_transformed"]])
    figure_legend("Per-sample boxplots after the transform step. This is the matrix used for differential expression testing. Aligned medians confirm the transform effectively standardized sample distributions.")
  }
  if (!is.null(qc_res$plots[["boxplot"]])) {
    safe_print_plot(qc_res$plots[["boxplot"]])
    if (has_scaling_box) {
      figure_legend(sprintf(
        "Per-sample boxplots after full normalization (%s + %s). %s scaling standardizes each lipid independently (mean=0, SD=1), so shifted medians between groups reflect biological differences. This matrix is used for PCA, heatmaps, and multivariate analyses (RF, PLS-DA).",
        ni_box$transform %||% "", ni_box$scaling %||% "",
        tools::toTitleCase(ni_box$scaling %||% "")
      ))
    } else {
      figure_legend("Per-sample boxplots after normalization. Comparable medians and spreads indicate successful normalization.")
    }
  }
}
```

## Heatmaps {.tabset}

```{r qc-heatmaps, out.width="100%", results='asis'}
# Distance heatmap only (correlation heatmap intentionally omitted from the
# report after Apr 2026 reviewer feedback — pipeline still writes it to disk).
# Tabs split by sample subset when QC samples exist.
if (!is.null(qc_res$files)) {
  fnames <- basename(qc_res$files)
  dist_with <- qc_res$files[file.exists(qc_res$files) &
                            fnames == "sample_distance_heatmap.png"]
  dist_no   <- qc_res$files[file.exists(qc_res$files) &
                            fnames == "sample_distance_heatmap_noQC.png"]

  hm_subsets <- list()
  if (length(dist_with) > 0) {
    hm_subsets[[length(hm_subsets) + 1]] <- list(
      tab  = if (length(dist_no) > 0) "With QC samples" else "All samples",
      path = dist_with[1],
      cid  = "sample_distance"
    )
  }
  if (length(dist_no) > 0) {
    hm_subsets[[length(hm_subsets) + 1]] <- list(
      tab  = "Without QC samples",
      path = dist_no[1],
      cid  = "sample_distance_noQC"
    )
  }

  if (length(hm_subsets) == 0) {
    cat("*No sample distance heatmap available.*\n\n")
  } else {
    for (sub in hm_subsets) {
      cat(sprintf("\n### %s\n\n", sub$tab))
      cat(sprintf("![Sample Distance Heatmap -- %s](%s){width=100%%}\n\n",
                  sub$tab, sub$path))
      figure_legend(sprintf(
        "Sample distance heatmap based on Euclidean distances between normalized lipid profiles (%s). Darker colors indicate larger distances (less similar samples).",
        tolower(sub$tab)))
      render_commentary(sub$cid)
    }
  }
}
```

`r if (is.null(qc_res) || !show_qc_lip) "-->"`

`r if (is.null(qc_enhanced)) "<!--"`

## PCA Scree

```{r qc-enh-scree, fig.width=9, fig.height=6, results='asis'}
if (!is.null(qc_enhanced$plots$pca_scree)) {
  safe_print_plot(qc_enhanced$plots$pca_scree)
  figure_legend("PCA scree plot showing the proportion of total variance explained by each principal component. The elbow point indicates the number of components that capture the majority of biological variation in the lipidome.")
  render_commentary("pca_scree")
}
```

## PCA Loading

```{r qc-enh-loading, fig.width=9, fig.height=7, results='asis'}
if (!is.null(qc_enhanced$plots$pca_loading)) {
  safe_print_plot(qc_enhanced$plots$pca_loading)
  figure_legend("PCA loading plot showing the contribution of individual lipid features to the first two principal components. Lipids with large loadings drive sample separation along the corresponding PC axis.")
  render_commentary("pca_loading")
}
```

## Category Donut {.tabset}

```{r qc-enh-donut, fig.width=9, fig.height=9, results='asis'}
if (!is.null(qc_enhanced$plots$category_donuts) && length(qc_enhanced$plots$category_donuts) > 0) {
  for (grp in names(qc_enhanced$plots$category_donuts)) {
    cat(paste0("\n### ", grp, "\n\n"))
    safe_print_plot(qc_enhanced$plots$category_donuts[[grp]])
    cat("\n\n")
  }
  figure_legend("Donut charts showing the distribution of detected lipid features across major lipid categories (e.g., glycerophospholipids, sphingolipids, glycerolipids) per condition. This provides an overview of lipidome coverage and allows comparison of category composition between groups.")
  render_commentary("category_donut")
} else if (!is.null(qc_enhanced$plots$category_donut)) {
  safe_print_plot(qc_enhanced$plots$category_donut)
  figure_legend("Donut chart showing the distribution of detected lipid features across major lipid categories (e.g., glycerophospholipids, sphingolipids, glycerolipids). This provides an overview of lipidome coverage.")
  render_commentary("category_donut")
}
```

## Class Correlation

```{r qc-enh-classcor, fig.width=10, fig.height=8, results='asis'}
if (!is.null(qc_enhanced$plots$class_correlation)) {
  safe_print_plot(qc_enhanced$plots$class_correlation)
  figure_legend("Correlation matrix of lipid class-level mean intensities across samples. Strong positive correlations between classes may indicate co-regulation or shared metabolic pathways.")
  render_commentary("class_correlation")
}
```

## PERMANOVA

```{r qc-enh-permanova, results='asis'}
if (!is.null(qc_enhanced)) {
  if (!is.null(qc_enhanced$permanova)) {
    cat("<div style='background: linear-gradient(135deg, #f8f9fa 0%, #e9ecef 100%); border: 1px solid #dee2e6; border-radius: 8px; padding: 20px 24px; margin: 10px 0 20px 0;'>\n")
    cat("<h4>PERMANOVA Results</h4>\n")
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 12px;'><em>Permutational multivariate analysis of variance testing whether lipid profiles differ significantly between experimental groups.</em></p>\n")

    perm_df <- tryCatch({
      perm <- qc_enhanced$permanova
      if (is.data.frame(perm)) {
        perm
      } else if (!is.null(perm$aov.tab)) {
        as.data.frame(perm$aov.tab)
      } else if (!is.null(perm$table)) {
        as.data.frame(perm$table)
      } else if (is.list(perm) && !is.null(perm$F_statistic)) {
        # Handle list(F_statistic, R2, p_value) from run_permanova()
        data.frame(
          Statistic = c("F statistic", "R-squared", "p-value"),
          Value = c(
            round(perm$F_statistic, 2),
            round(perm$R2, 2),
            sprintf("%.2e", perm$p_value)
          ),
          stringsAsFactors = FALSE
        )
      } else {
        NULL
      }
    }, error = function(e) NULL)

    if (!is.null(perm_df)) {
      print(make_dt(perm_df, caption = "PERMANOVA Results"))
    }
    cat("</div>\n")

    figure_legend("PERMANOVA (Permutational Multivariate Analysis of Variance) tests whether the centroids and dispersion of lipid profiles differ between groups. A significant p-value (< 0.05) indicates that group membership explains a significant proportion of the overall variation in the lipidome.")
  } else {
    cat("<div style='background: #fff3cd; border: 1px solid #ffc107; border-radius: 8px; padding: 16px 20px; margin: 10px 0 20px 0;'>\n")
    cat("<p><strong>PERMANOVA not available</strong> — the <code>vegan</code> package is not installed. Install it with <code>install.packages('vegan')</code> to enable permutational multivariate analysis of variance.</p>\n")
    cat("</div>\n")
  }
}
```

`r if (is.null(qc_enhanced)) "-->"`

---

`r if (is.null(de_res) || is.null(de_res$summary_df) || !show_de_lip) "<!--"`

# Differential Expression

```{r cutoff-panel, results='asis'}
if (exists("cutoff_panel_html")) {
    .lfc_default <- if (!is.null(lipid_cfg$de$logfc_cutoff)) {
        lipid_cfg$de$logfc_cutoff
    } else {
        log2(lipid_cfg$de$linear_fc_cutoff %||% 1.5)
    }
    .p_default <- lipid_cfg$de$p_cutoff %||% 0.05
    cat(cutoff_panel_html(.lfc_default, .p_default, default_ptype = "pval"))
}
```

```{r de-prep, include=FALSE}
# Prepare shared data used across DE sub-sections
padj_cut_de <- lipid_cfg$de$p_cutoff %||% 0.05
# logfc_cutoff is applied directly to |logFC| (in the transform's scale).
# linear_fc_cutoff is a backwards-compat alias converted via log2().
if (!is.null(lipid_cfg$de$logfc_cutoff)) {
  log2_lfc_de <- lipid_cfg$de$logfc_cutoff
} else {
  lfc_cut_de  <- lipid_cfg$de$linear_fc_cutoff %||% 1.5
  log2_lfc_de <- log2(lfc_cut_de)
}

# Build feature_id -> display_name lookup (full lipid names)
name_map_v <- NULL
if (!is.null(pre$row_data)) {
  # Prefer display_name (e.g. "Phosphatidylcholine (16:0/18:1)") over Name
  name_col <- if ("display_name" %in% colnames(pre$row_data)) "display_name" else "Name"
  if (name_col %in% colnames(pre$row_data)) {
    name_map_v <- stats::setNames(
      as.character(pre$row_data[[name_col]]),
      as.character(pre$row_data$feature_id)
    )
  }
}

resolve_name <- function(fid) {
  if (!is.null(name_map_v) && fid %in% names(name_map_v) && !is.na(name_map_v[fid])) {
    name_map_v[fid]
  } else {
    fid
  }
}

# Build norm intensity y-label
norm_y_label_report <- "Normalized intensity"
ni_r <- pre$info$normalization
if (!is.null(ni_r)) {
  np <- character(0)
  if (!is.null(ni_r$transform) && tolower(ni_r$transform) != "none") np <- c(np, ni_r$transform)
  if (!is.null(ni_r$sample_norm) && tolower(ni_r$sample_norm) != "none") np <- c(np, toupper(ni_r$sample_norm))
  if (!is.null(ni_r$scaling) && tolower(ni_r$scaling) != "none") np <- c(np, ni_r$scaling)
  if (length(np) > 0) norm_y_label_report <- paste0("Normalized intensity (", paste(np, collapse = " + "), ")")
}
```

```{r de-note, results='asis'}
if (!is.null(de_res) && !is.null(de_res$summary_df)) {
  n_sig_any <- sum(de_res$summary_df$pass_any_contrast == 1, na.rm = TRUE)
  if (n_sig_any == 0) {
    n_samp <- ncol(pre$expr_work)
    cat("\n> **Note:** No features reached significance (p-value < ",
        padj_cut_de, " and |logFC| >= ", round(log2_lfc_de, 2),
        "). With only **", n_samp, " samples**, the test has limited power. ",
        "Consider increasing sample size, relaxing thresholds, or using a ",
        "rank-based method (e.g. GSEA).\n\n", sep = "")
  }
}
```

```{r de-sections, results='asis', fig.width=10, fig.height=7}
if (!is.null(de_contrasts)) {
  for (ctr in de_contrasts) {
    ctr_label <- make_contrast_label(ctr)

    ctr_tbl <- tryCatch(
      extract_contrast_table(de_res$summary_df, ctr_label),
      error = function(e) NULL
    )
    if (is.null(ctr_tbl)) next

    # Add Name column
    ctr_tbl$Name <- vapply(ctr_tbl$feature_id, resolve_name, character(1), USE.NAMES = FALSE)

    # Direction column for coloring (uses adj.P.Val for consistency with cutoff panel)
    ctr_tbl$Direction <- ifelse(
      !is.na(ctr_tbl$adj.P.Val) & ctr_tbl$adj.P.Val <= padj_cut_de &
        !is.na(ctr_tbl$logFC) & abs(ctr_tbl$logFC) >= log2_lfc_de,
      ifelse(ctr_tbl$logFC > 0, "Up", "Down"), "NS"
    )
    ctr_tbl$Direction <- factor(ctr_tbl$Direction, levels = c("NS", "Down", "Up"))
    ctr_tbl$neglog10_pval <- pmin(-log10(pmax(
      ifelse(is.na(ctr_tbl$P.Value), 1, ctr_tbl$P.Value), 1e-300
    )), 50)

    # ================================================================
    # VOLCANO
    # ================================================================
    cat("\n## Volcano: ", ctr, " {.tabset}\n\n", sep = "")

    # -- Static Volcano (sub-tabbed by p-value type) --
    # Default sub-tab matches the cutoff-panel default (`default_ptype = "pval"`)
    default_ptype_lip <- "pval"
    vol_padj_png <- de_res$files[grepl(paste0("volcano_", ctr_label, "_padj\\.png$"),
                                       basename(de_res$files))]
    vol_pval_png <- de_res$files[grepl(paste0("volcano_", ctr_label, "_pval\\.png$"),
                                       basename(de_res$files))]
    vol_padj_png <- vol_padj_png[file.exists(vol_padj_png)]
    vol_pval_png <- vol_pval_png[file.exists(vol_pval_png)]

    if (length(vol_padj_png) > 0 || length(vol_pval_png) > 0) {
      cat("### Static {.tabset}\n\n")
      order_ptypes <- if (default_ptype_lip == "padj") c("padj", "pval") else c("pval", "padj")
      for (pt in order_ptypes) {
        png_path <- if (pt == "padj") vol_padj_png else vol_pval_png
        tab_label <- if (pt == "padj") "Adjusted P-value" else "Raw P-value"
        if (length(png_path) > 0) {
          cat(sprintf("#### %s\n\n", tab_label))
          cat(sprintf("\n![Volcano: %s -- %s](%s){width=100%%}\n\n",
                      ctr, tab_label, png_path[1]))
          figure_legend(paste0(
            "Volcano plot for contrast ", ctr,
            " (y-axis: -log10 ",
            if (pt == "padj") "adjusted P-value" else "raw P-value", "). ",
            "Coloured points exceed both fold-change and significance thresholds."
          ))
        }
      }
      render_commentary(paste0("volcano_", ctr_label))
      cat("\n\n")
    } else {
      vol_png <- de_res$files[grepl(paste0("volcano_", ctr_label, "\\.png$"),
                                    basename(de_res$files))]
      vol_png <- vol_png[file.exists(vol_png)]
      if (length(vol_png) > 0) {
        cat("### Static\n\n")
        cat(sprintf("\n![Volcano: %s](%s){width=100%%}\n\n", ctr, vol_png[1]))
        figure_legend(paste0("Volcano plot for contrast ", ctr, ". Each point is a lipid; x-axis shows log2(fold change) and y-axis shows -log10(p-value). Colored points exceed both fold-change and significance thresholds."))
        render_commentary(paste0("volcano_", ctr_label))
        cat("\n\n")
      }
    }

    # -- Interactive Volcano (matching proteomics pattern) --
    if (requireNamespace("plotly", quietly = TRUE)) {
      cat("### Interactive\n\n")

      vol_cid <- gsub("[^a-zA-Z0-9]", "_", ctr_label)

      vol_p <- ggplot2::ggplot(
        ctr_tbl,
        ggplot2::aes(x = logFC, y = neglog10_pval, color = Direction,
                     text = paste0("Lipid: ", Name,
                                   "\nlog2(FC): ", round(logFC, 2),
                                   "\nP.Value: ", sprintf("%.2e", P.Value),
                                   "\nadj.P.Val: ", sprintf("%.2e", adj.P.Val)))
      ) +
        ggplot2::geom_point(alpha = 0.6, size = 1.5) +
        ggplot2::scale_color_manual(values = c("NS" = "grey70", "Down" = "blue", "Up" = "red")) +
        ggplot2::geom_vline(xintercept = c(-log2_lfc_de, log2_lfc_de), linetype = "dashed", alpha = 0.5) +
        ggplot2::geom_hline(yintercept = -log10(padj_cut_de), linetype = "dashed", alpha = 0.5) +
        ggplot2::labs(x = "log2(FC)", y = "-log10(P.Value)") +
        ggplot2::theme_minimal()

      vol_plotly <- plotly::ggplotly(vol_p, tooltip = "text")

      cat(sprintf('
<div style="margin-bottom: 15px; position: relative;">
<label for="volcano_genes_%s"><strong>Highlight lipids:</strong></label>
<div style="display: inline-block; position: relative; margin-left: 10px;">
<input type="text" id="volcano_genes_%s" placeholder="Start typing a lipid name..." autocomplete="off"
  style="padding: 8px; border-radius: 4px; border: 1px solid #ccc; width: 350px;"
  oninput="volcanoAutocomplete_%s(this)" onkeydown="volcanoAutoKey_%s(event)">
<div id="volcano_ac_%s" style="position: absolute; top: 100%%; left: 0; width: 350px; max-height: 200px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 4px 4px; z-index: 1000; display: none; box-shadow: 0 4px 6px rgba(0,0,0,0.1);"></div>
</div>
<button onclick="highlightVolcanoGenes_%s()" style="padding: 8px 15px; margin-left: 5px; background: #007bff; color: white; border: none; border-radius: 4px; cursor: pointer;">Highlight</button>
<button onclick="clearVolcanoHighlight_%s()" style="padding: 8px 15px; margin-left: 5px; background: #6c757d; color: white; border: none; border-radius: 4px; cursor: pointer;">Clear</button>
<div id="volcano_tags_%s" style="margin-top: 5px; display: flex; flex-wrap: wrap; gap: 4px;"></div>
</div>
', vol_cid, vol_cid, vol_cid, vol_cid, vol_cid, vol_cid, vol_cid, vol_cid))

      vol_plotly$elementId <- paste0("volcano_plot_", vol_cid)
      print(htmltools::tagList(vol_plotly))

      # Register volcano plot for cutoff panel
      if (exists("cutoff_register_plot")) {
        vol_reg_df <- data.frame(
          name = as.character(ctr_tbl$Name),
          logFC = ctr_tbl$logFC,
          pval = ctr_tbl$P.Value,
          adjPval = ctr_tbl$adj.P.Val,
          stringsAsFactors = FALSE
        )
        pass_col_l <- paste0("pass.", ctr_label)
        if (!is.null(de_res$summary_df) && pass_col_l %in% names(de_res$summary_df)) {
          pf <- as.numeric(de_res$summary_df[[pass_col_l]])
          pf_aligned <- pf[match(ctr_tbl$feature_id, de_res$summary_df$feature_id)]
          vol_reg_df$passFlag <- ifelse(is.na(pf_aligned), 0, pf_aligned)
        }
        cat(cutoff_register_plot(
          paste0("volcano_plot_", vol_cid),
          vol_reg_df, "volcano", "Lipid", ctr
        ))
      }

      # JS for autocomplete + highlight (proteomics pattern)
      cat(sprintf('
<script>
(function() {
    var cid = "%s";
    var selectedGenes = [];
    var acIndex = -1;
    function getLipidNames() {
        var plotDiv = document.getElementById("volcano_plot_" + cid);
        if (!plotDiv || !plotDiv.data) return [];
        var names = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var data = plotDiv.data[t];
            if (!data.text) continue;
            for (var i = 0; i < data.text.length; i++) {
                var parts = data.text[i].split(/<br\\s*\\/?>/i);
                if (parts.length > 0) {
                    var g = parts[0].replace(/Lipid:\\s*/i, "").trim();
                    if (g && names.indexOf(g) === -1) names.push(g);
                }
            }
        }
        return names.sort();
    }
    window.volcanoAutocomplete_%s = function(el) {
        var acDiv = document.getElementById("volcano_ac_" + cid);
        var val = el.value.trim().toLowerCase();
        acDiv.innerHTML = ""; acIndex = -1;
        if (val.length < 1) { acDiv.style.display = "none"; return; }
        var allNames = getLipidNames();
        var matches = allNames.filter(function(g) { return g.toLowerCase().indexOf(val) !== -1 && selectedGenes.indexOf(g) === -1; }).slice(0, 15);
        if (matches.length === 0) { acDiv.style.display = "none"; return; }
        matches.forEach(function(g) {
            var item = document.createElement("div");
            item.textContent = g;
            item.style.cssText = "padding: 6px 10px; cursor: pointer; font-size: 13px;";
            item.onmouseenter = function() { this.style.background = "#e9ecef"; };
            item.onmouseleave = function() { this.style.background = "white"; };
            item.onclick = function() { selectGene(g); };
            acDiv.appendChild(item);
        });
        acDiv.style.display = "block";
    };
    window.volcanoAutoKey_%s = function(e) {
        var acDiv = document.getElementById("volcano_ac_" + cid);
        var items = acDiv.children;
        if (items.length === 0) return;
        if (e.key === "ArrowDown") { e.preventDefault(); acIndex = Math.min(acIndex + 1, items.length - 1); updateAcHighlight(items); }
        else if (e.key === "ArrowUp") { e.preventDefault(); acIndex = Math.max(acIndex - 1, 0); updateAcHighlight(items); }
        else if (e.key === "Enter" && acIndex >= 0) { e.preventDefault(); items[acIndex].click(); }
    };
    function updateAcHighlight(items) { for (var i = 0; i < items.length; i++) items[i].style.background = i === acIndex ? "#e9ecef" : "white"; }
    function selectGene(gene) {
        if (selectedGenes.indexOf(gene) === -1) selectedGenes.push(gene);
        document.getElementById("volcano_genes_" + cid).value = "";
        document.getElementById("volcano_ac_" + cid).style.display = "none";
        renderTags(); applyHighlight();
    }
    function renderTags() {
        var container = document.getElementById("volcano_tags_" + cid);
        container.innerHTML = "";
        selectedGenes.forEach(function(g) {
            var tag = document.createElement("span");
            tag.style.cssText = "display: inline-flex; align-items: center; background: #007bff; color: white; padding: 2px 8px; border-radius: 12px; font-size: 12px; gap: 4px;";
            tag.innerHTML = g + " <span style=\\"cursor:pointer; font-weight:bold; margin-left:3px;\\" onclick=\\"removeVolcanoGene_" + cid + "(\'" + g + "\')\\">x</span>";
            container.appendChild(tag);
        });
    }
    window["removeVolcanoGene_" + cid] = function(gene) { selectedGenes = selectedGenes.filter(function(g) { return g !== gene; }); renderTags(); applyHighlight(); };
    function applyHighlight() {
        var plotDiv = document.getElementById("volcano_plot_" + cid);
        if (!plotDiv || !plotDiv.data) return;
        if (selectedGenes.length === 0) { Plotly.relayout(plotDiv, { annotations: [] }); return; }
        var genesLower = selectedGenes.map(function(g) { return g.toLowerCase(); });
        var annotations = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var data = plotDiv.data[t];
            if (!data.text) continue;
            for (var i = 0; i < data.text.length; i++) {
                var parts = data.text[i].split(/<br\\s*\\/?>/i);
                if (parts.length > 0) {
                    var lipidName = parts[0].replace(/Lipid:\\s*/i, "").trim();
                    if (genesLower.indexOf(lipidName.toLowerCase()) !== -1) {
                        annotations.push({ x: data.x[i], y: data.y[i], text: lipidName, showarrow: true, arrowhead: 2, ax: 20, ay: -30, font: { size: 12, color: "black" }, bgcolor: "rgba(255,255,255,0.9)", bordercolor: "black", borderwidth: 1 });
                    }
                }
            }
        }
        Plotly.relayout(plotDiv, { annotations: annotations });
    }
    window.highlightVolcanoGenes_%s = function() {
        var input = document.getElementById("volcano_genes_" + cid).value;
        var manual = input.split(",").map(function(s) { return s.trim(); }).filter(function(s) { return s.length > 0; });
        var allNames = getLipidNames();
        manual.forEach(function(m) { var mLower = m.toLowerCase(); var match = allNames.find(function(g) { return g.toLowerCase() === mLower; }); if (match && selectedGenes.indexOf(match) === -1) selectedGenes.push(match); });
        document.getElementById("volcano_genes_" + cid).value = "";
        document.getElementById("volcano_ac_" + cid).style.display = "none";
        renderTags(); applyHighlight();
    };
    window.clearVolcanoHighlight_%s = function() {
        selectedGenes = [];
        document.getElementById("volcano_genes_" + cid).value = "";
        renderTags();
        var plotDiv = document.getElementById("volcano_plot_" + cid);
        if (plotDiv) Plotly.relayout(plotDiv, { annotations: [] });
    };
    document.addEventListener("click", function(e) {
        var acDiv = document.getElementById("volcano_ac_" + cid);
        var input = document.getElementById("volcano_genes_" + cid);
        if (acDiv && e.target !== input && !acDiv.contains(e.target)) acDiv.style.display = "none";
    });
})();
</script>
', vol_cid, vol_cid, vol_cid, vol_cid, vol_cid))

      figure_legend(paste0("Interactive volcano plot for ", ctr, ". Type lipid names in the search box to annotate them on the plot. Hover over points for details."))
      cat("\n\n")
    }

    # ================================================================
    # MA PLOT  (default-off after Apr 2026 reviewer feedback;
    #           toggle via report_sections$de$ma: true to restore)
    # ================================================================
    ma_png <- de_res$files[grepl(paste0("MA_", ctr_label, "\\.png$"), basename(de_res$files))]
    ma_png <- ma_png[file.exists(ma_png)]
    if (show_de_ma_lip && length(ma_png) > 0) {
      cat("\n## MA Plot: ", ctr, " {.tabset}\n\n", sep = "")

      cat("### Static\n\n")
      cat(sprintf("\n![MA Plot: %s](%s){width=100%%}\n\n", ctr, ma_png[1]))
      figure_legend(paste0("MA plot for contrast ", ctr, ". The x-axis shows average expression and the y-axis shows log2 fold change. Significant features are highlighted."))
      render_commentary(paste0("ma_", ctr_label))
      cat("\n\n")

      # Interactive MA (matching proteomics pattern)
      if (requireNamespace("plotly", quietly = TRUE)) {
        cat("### Interactive\n\n")

        ma_cid <- paste0("ma_", gsub("[^a-zA-Z0-9]", "_", ctr_label))

        ma_p <- ggplot2::ggplot(
          ctr_tbl,
          ggplot2::aes(x = AveExpr, y = logFC, color = Direction,
                       text = paste0("Lipid: ", Name,
                                     "\nlog2(FC): ", round(logFC, 2),
                                     "\nAveExpr: ", round(AveExpr, 2),
                                     "\nP.Value: ", sprintf("%.2e", P.Value),
                                     "\nadj.P.Val: ", sprintf("%.2e", adj.P.Val)))
        ) +
          ggplot2::geom_point(alpha = 0.6, size = 1.5) +
          ggplot2::scale_color_manual(values = c("NS" = "grey70", "Down" = "blue", "Up" = "red")) +
          ggplot2::geom_hline(yintercept = c(-log2_lfc_de, 0, log2_lfc_de),
                              linetype = c("dashed", "solid", "dashed"),
                              alpha = c(0.5, 0.3, 0.5)) +
          ggplot2::labs(x = "Average Expression", y = "log2(FC)") +
          ggplot2::theme_minimal()

        ma_plotly <- plotly::ggplotly(ma_p, tooltip = "text")

        cat(sprintf('
<div style="margin-bottom: 15px; position: relative;">
<label for="ma_genes_%s"><strong>Highlight lipids:</strong></label>
<div style="display: inline-block; position: relative; margin-left: 10px;">
<input type="text" id="ma_genes_%s" placeholder="Start typing a lipid name..." autocomplete="off"
  style="padding: 8px; border-radius: 4px; border: 1px solid #ccc; width: 350px;"
  oninput="maAutocomplete_%s(this)" onkeydown="maAutoKey_%s(event)">
<div id="ma_ac_%s" style="position: absolute; top: 100%%; left: 0; width: 350px; max-height: 200px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 4px 4px; z-index: 1000; display: none; box-shadow: 0 4px 6px rgba(0,0,0,0.1);"></div>
</div>
<button onclick="highlightMAGenes_%s()" style="padding: 8px 15px; margin-left: 5px; background: #007bff; color: white; border: none; border-radius: 4px; cursor: pointer;">Highlight</button>
<button onclick="clearMAHighlight_%s()" style="padding: 8px 15px; margin-left: 5px; background: #6c757d; color: white; border: none; border-radius: 4px; cursor: pointer;">Clear</button>
<div id="ma_tags_%s" style="margin-top: 5px; display: flex; flex-wrap: wrap; gap: 4px;"></div>
</div>
', ma_cid, ma_cid, ma_cid, ma_cid, ma_cid, ma_cid, ma_cid, ma_cid))

        ma_plotly$elementId <- paste0("ma_plot_", ma_cid)
        print(htmltools::tagList(ma_plotly))

        # Register MA plot for cutoff panel
        if (exists("cutoff_register_plot")) {
          ma_reg_df <- data.frame(
            name = as.character(ctr_tbl$Name),
            logFC = ctr_tbl$logFC,
            pval = ctr_tbl$P.Value,
            avgExpr = ctr_tbl$AveExpr,
            adjPval = ctr_tbl$adj.P.Val,
            stringsAsFactors = FALSE
          )
          pass_col_l <- paste0("pass.", ctr_label)
          if (!is.null(de_res$summary_df) && pass_col_l %in% names(de_res$summary_df)) {
            pf <- as.numeric(de_res$summary_df[[pass_col_l]])
            pf_aligned <- pf[match(ctr_tbl$feature_id, de_res$summary_df$feature_id)]
            ma_reg_df$passFlag <- ifelse(is.na(pf_aligned), 0, pf_aligned)
          }
          cat(cutoff_register_plot(
            paste0("ma_plot_", ma_cid),
            ma_reg_df, "ma", "Lipid", ctr
          ))
        }

        cat(sprintf('
<script>
(function() {
    var cid = "%s"; var selectedGenes = []; var acIndex = -1;
    function getLipidNames() {
        var plotDiv = document.getElementById("ma_plot_" + cid);
        if (!plotDiv || !plotDiv.data) return [];
        var names = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var data = plotDiv.data[t];
            if (!data.text) continue;
            for (var i = 0; i < data.text.length; i++) {
                var parts = data.text[i].split(/<br\\s*\\/?>/i);
                if (parts.length > 0) {
                    var g = parts[0].replace(/Lipid:\\s*/i, "").trim();
                    if (g && names.indexOf(g) === -1) names.push(g);
                }
            }
        }
        return names.sort();
    }
    window.maAutocomplete_%s = function(el) {
        var acDiv = document.getElementById("ma_ac_" + cid);
        var val = el.value.trim().toLowerCase();
        acDiv.innerHTML = ""; acIndex = -1;
        if (val.length < 1) { acDiv.style.display = "none"; return; }
        var allNames = getLipidNames();
        var matches = allNames.filter(function(g) { return g.toLowerCase().indexOf(val) !== -1 && selectedGenes.indexOf(g) === -1; }).slice(0, 15);
        if (matches.length === 0) { acDiv.style.display = "none"; return; }
        matches.forEach(function(g) {
            var item = document.createElement("div");
            item.textContent = g;
            item.style.cssText = "padding: 6px 10px; cursor: pointer; font-size: 13px;";
            item.onmouseenter = function() { this.style.background = "#e9ecef"; };
            item.onmouseleave = function() { this.style.background = "white"; };
            item.onclick = function() { selectGene(g); };
            acDiv.appendChild(item);
        });
        acDiv.style.display = "block";
    };
    window.maAutoKey_%s = function(e) {
        var acDiv = document.getElementById("ma_ac_" + cid);
        var items = acDiv.children;
        if (items.length === 0) return;
        if (e.key === "ArrowDown") { e.preventDefault(); acIndex = Math.min(acIndex + 1, items.length - 1); updateAcHighlight(items); }
        else if (e.key === "ArrowUp") { e.preventDefault(); acIndex = Math.max(acIndex - 1, 0); updateAcHighlight(items); }
        else if (e.key === "Enter" && acIndex >= 0) { e.preventDefault(); items[acIndex].click(); }
    };
    function updateAcHighlight(items) { for (var i = 0; i < items.length; i++) items[i].style.background = i === acIndex ? "#e9ecef" : "white"; }
    function selectGene(gene) {
        if (selectedGenes.indexOf(gene) === -1) selectedGenes.push(gene);
        document.getElementById("ma_genes_" + cid).value = "";
        document.getElementById("ma_ac_" + cid).style.display = "none";
        renderTags(); applyHighlight();
    }
    function renderTags() {
        var container = document.getElementById("ma_tags_" + cid);
        container.innerHTML = "";
        selectedGenes.forEach(function(g) {
            var tag = document.createElement("span");
            tag.style.cssText = "display: inline-flex; align-items: center; background: #007bff; color: white; padding: 2px 8px; border-radius: 12px; font-size: 12px; gap: 4px;";
            tag.innerHTML = g + " <span style=\\"cursor:pointer; font-weight:bold; margin-left:3px;\\" onclick=\\"removeMAGene_" + cid + "(\'" + g + "\')\\">x</span>";
            container.appendChild(tag);
        });
    }
    window["removeMAGene_" + cid] = function(gene) { selectedGenes = selectedGenes.filter(function(g) { return g !== gene; }); renderTags(); applyHighlight(); };
    function applyHighlight() {
        var plotDiv = document.getElementById("ma_plot_" + cid);
        if (!plotDiv || !plotDiv.data) return;
        if (selectedGenes.length === 0) { Plotly.relayout(plotDiv, { annotations: [] }); return; }
        var genesLower = selectedGenes.map(function(g) { return g.toLowerCase(); });
        var annotations = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var data = plotDiv.data[t];
            if (!data.text) continue;
            for (var i = 0; i < data.text.length; i++) {
                var parts = data.text[i].split(/<br\\s*\\/?>/i);
                if (parts.length > 0) {
                    var lipidName = parts[0].replace(/Lipid:\\s*/i, "").trim();
                    if (genesLower.indexOf(lipidName.toLowerCase()) !== -1) {
                        annotations.push({ x: data.x[i], y: data.y[i], text: lipidName, showarrow: true, arrowhead: 2, ax: 20, ay: -30, font: { size: 12, color: "black" }, bgcolor: "rgba(255,255,255,0.9)", bordercolor: "black", borderwidth: 1 });
                    }
                }
            }
        }
        Plotly.relayout(plotDiv, { annotations: annotations });
    }
    window.highlightMAGenes_%s = function() {
        var input = document.getElementById("ma_genes_" + cid).value;
        var manual = input.split(",").map(function(s) { return s.trim(); }).filter(function(s) { return s.length > 0; });
        var allNames = getLipidNames();
        manual.forEach(function(m) { var mLower = m.toLowerCase(); var match = allNames.find(function(g) { return g.toLowerCase() === mLower; }); if (match && selectedGenes.indexOf(match) === -1) selectedGenes.push(match); });
        document.getElementById("ma_genes_" + cid).value = "";
        document.getElementById("ma_ac_" + cid).style.display = "none";
        renderTags(); applyHighlight();
    };
    window.clearMAHighlight_%s = function() {
        selectedGenes = [];
        document.getElementById("ma_genes_" + cid).value = "";
        renderTags();
        var plotDiv = document.getElementById("ma_plot_" + cid);
        if (plotDiv) Plotly.relayout(plotDiv, { annotations: [] });
    };
    document.addEventListener("click", function(e) {
        var acDiv = document.getElementById("ma_ac_" + cid);
        var input = document.getElementById("ma_genes_" + cid);
        if (acDiv && e.target !== input && !acDiv.contains(e.target)) acDiv.style.display = "none";
    });
})();
</script>
', ma_cid, ma_cid, ma_cid, ma_cid, ma_cid))
        figure_legend(paste0("Interactive MA plot for ", ctr, ". Search and highlight specific lipids."))
        cat("\n\n")
      }
    }

    # ================================================================
    # P-VALUE DISTRIBUTION  (default-off after Apr 2026 reviewer feedback;
    #                        toggle via report_sections$de$pval_hist: true)
    # ================================================================
    phist_png <- de_res$files[grepl(paste0("pvalue_hist_", ctr_label, "\\.png$"), basename(de_res$files))]
    phist_png <- phist_png[file.exists(phist_png)]
    if (show_de_pvalhist_lip && length(phist_png) > 0) {
      cat("\n## P-value Distribution: ", ctr, "\n\n", sep = "")
      cat(sprintf("\n![P-value Distribution: %s](%s){width=85%%}\n\n", ctr, phist_png[1]))
      figure_legend(paste0("Distribution of raw p-values for contrast ", ctr, ". A uniform distribution suggests no differential signal; enrichment near zero indicates true positives."))
      render_commentary(paste0("pvalue_hist_", ctr_label))
      cat("\n\n")
    }

    # ================================================================
    # DE RESULTS TABLE (statistics + normalized intensity)
    # ================================================================
    cat("\n## DE Results: ", ctr, " {.tabset}\n\n", sep = "")

    # Top 50 table
    top_tbl <- ctr_tbl[order(ctr_tbl$P.Value), ]
    top_tbl <- head(top_tbl, 50)
    top_display <- top_tbl[, c("feature_id", "Name", "logFC", "AveExpr", "P.Value", "adj.P.Val", "Direction")]
    cat("### Top 50 DE Lipids\n\n")
    print(make_dt(top_display, caption = paste0("Top 50 DE lipids: ", ctr)))
    cat("\n\n")

    # Full table with normalized intensities
    cat("### Full Results + Expression\n\n")
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>",
        "Full DE results merged with normalized expression values. ",
        "Use the search box to find specific lipids and the buttons to export.</em></p>\n")

    full_tbl <- ctr_tbl[order(ctr_tbl$P.Value), ,drop=FALSE]
    full_display <- full_tbl[, c("feature_id", "Name", "logFC", "AveExpr", "P.Value", "adj.P.Val", "Direction")]

    # Merge transformed expression (glog10, before scaling) for meaningful values
    de_tbl_expr <- pre$expr_log %||% pre$expr_work
    if (!is.null(de_tbl_expr)) {
      expr_df2 <- as.data.frame(de_tbl_expr, check.names = FALSE)
      expr_df2$feature_id <- rownames(de_tbl_expr)
      merged <- merge(full_display, expr_df2, by = "feature_id", all.x = TRUE)
      if (nrow(merged) > 0) full_display <- merged
    }

    # Format p-value columns as scientific notation strings before DT
    pval_pat <- "^(p[._]?val|padj|adj[._]?p|fdr|q[._]?val|BH|bonferroni|p\\.value|adj\\.P\\.Val|PValue|FDR)"
    pval_cols_de <- grep(pval_pat, names(full_display), ignore.case = TRUE, value = TRUE)
    for (pc in pval_cols_de) {
      if (is.numeric(full_display[[pc]])) {
        full_display[[pc]] <- ifelse(is.na(full_display[[pc]]), NA_character_,
                                     sprintf("%.2e", full_display[[pc]]))
      }
    }
    num_cols_de <- names(full_display)[vapply(full_display, is.numeric, logical(1))]
    tbl_de <- DT::datatable(
      full_display,
      caption = paste0("DE Results + Normalized Expression: ", ctr),
      extensions = "Buttons",
      options = list(
        scrollX = TRUE, pageLength = 10, dom = "Bfrtip",
        buttons = list(
          list(extend = "csv", filename = paste0("DE_expression_", ctr_label)),
          list(extend = "excel", filename = paste0("DE_expression_", ctr_label))
        ),
        search = list(regex = TRUE, caseInsensitive = TRUE)
      ),
      rownames = FALSE, filter = "top",
      class = "compact stripe hover"
    )
    if (length(num_cols_de) > 0) tbl_de <- DT::formatRound(tbl_de, columns = num_cols_de, digits = 2)
    print(htmltools::tagList(tbl_de))
    cat("\n\n")

  } # end contrast loop
}
```

```{r lipid-explorer-prep, include=FALSE}
# Prepare data for the lipid expression explorer
explorer_ready <- FALSE
explorer_json_str <- "{}"

# Use expr_log (transform-only, e.g. glog10) for the explorer so that means
# reflect actual intensity levels rather than z-scored zeros.
# Fall back to expr_work if expr_log is not available.
explorer_expr <- pre$expr_log %||% pre$expr_work

if (!is.null(explorer_expr) && !is.null(pre$meta)) {
  # Build lipid display names
  lipid_ids <- rownames(explorer_expr)
  lipid_display <- vapply(lipid_ids, resolve_name, character(1), USE.NAMES = FALSE)

  # Build sample -> group mapping
  sample_col_exp <- lipid_cfg$effects$samples %||% "sample_id"
  group_col_exp  <- cond_col
  meta_exp       <- pre$meta

  if (sample_col_exp %in% names(meta_exp) && group_col_exp %in% names(meta_exp)) {
    group_map_exp <- stats::setNames(as.character(meta_exp[[group_col_exp]]),
                                      meta_exp[[sample_col_exp]])
    sample_groups_exp <- as.character(group_map_exp[colnames(explorer_expr)])
    sample_groups_exp[is.na(sample_groups_exp)] <- "Unknown"
  } else {
    sample_groups_exp <- rep("All", ncol(explorer_expr))
  }

  # Build expression matrix as list of numeric vectors
  expr_matrix_list <- lapply(seq_len(nrow(explorer_expr)), function(i) {
    round(as.numeric(explorer_expr[i, ]), 2)
  })

  explorer_data <- list(
    genes   = lipid_display,
    samples = colnames(explorer_expr),
    groups  = sample_groups_exp,
    expr    = expr_matrix_list
  )

  explorer_json_str <- jsonlite::toJSON(explorer_data, auto_unbox = FALSE)
  explorer_ready <- TRUE
}
```

`r if (explorer_ready) "" else "<!--"`

## Lipid Explorer

Search for lipids and visualize their transformed expression (glog10, before scaling) across groups.

```{r lipid-explorer, results='asis'}
if (explorer_ready) {
  cat(sprintf('
<div style="margin: 15px 0 20px 0;">
<div style="display: flex; align-items: center; gap: 10px; flex-wrap: wrap;">
<label for="boxplot_gene_input" style="font-weight: bold; white-space: nowrap;">Search lipid:</label>
<div style="position: relative; flex: 1; max-width: 400px;">
<input type="text" id="boxplot_gene_input" placeholder="Start typing a lipid name..."
  autocomplete="off"
  style="padding: 10px 12px; border-radius: 6px; border: 1px solid #ccc; width: 100%%; font-size: 14px; box-sizing: border-box;"
  oninput="boxplotAutocomplete(this)" onkeydown="boxplotAutoKey(event)">
<div id="boxplot_ac" style="position: absolute; top: 100%%; left: 0; width: 100%%; max-height: 220px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 6px 6px; z-index: 1000; display: none; box-shadow: 0 4px 8px rgba(0,0,0,0.12);"></div>
</div>
<button onclick="addGeneToPlot()" style="padding: 10px 20px; background: #007bff; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Add</button>
<button onclick="clearBoxplot()" style="padding: 10px 20px; background: #6c757d; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Clear All</button>
</div>
<div id="boxplot_gene_tags" style="margin-top: 8px; display: flex; flex-wrap: wrap; gap: 6px;"></div>
<div id="boxplot_gene_info" style="margin-top: 8px; font-size: 0.9em; color: #555;"></div>
</div>
<div id="boxplot_container" style="width: 100%%; min-height: 450px; margin-bottom: 10px;"></div>
<div id="boxplot_export_bar" style="display: none; margin-bottom: 20px; gap: 8px; align-items: center;">
  <span style="font-size: 0.85em; color: #555; margin-right: 4px;">Export:</span>
  <button onclick="exportBoxplot(\'svg\')" style="padding: 5px 14px; background: #28a745; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">SVG</button>
  <button onclick="exportBoxplot(\'png\')" style="padding: 5px 14px; background: #17a2b8; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">PNG</button>
</div>
'))

  # Inject JS data and rendering logic
  cat(sprintf('<script>
var geneExprData = %s;
var bpGroupLabel = "%s";
var bpYLabel = "%s";
var boxplotAcIndex = -1;
var selectedGenes = [];

function boxplotAutocomplete(el) {
  var ac = document.getElementById("boxplot_ac");
  var val = el.value.trim().toLowerCase(); ac.innerHTML = ""; boxplotAcIndex = -1;
  if (val.length < 1) { ac.style.display = "none"; return; }
  var matches = [];
  for (var i = 0; i < geneExprData.genes.length; i++) {
    if (geneExprData.genes[i].toLowerCase().indexOf(val) !== -1) {
      if (selectedGenes.some(function(g){return g.idx===i;})) continue;
      matches.push({name: geneExprData.genes[i], idx: i});
      if (matches.length >= 20) break;
    }
  }
  if (matches.length === 0) { ac.style.display = "none"; return; }
  matches.forEach(function(m) {
    var item = document.createElement("div"); item.textContent = m.name;
    item.style.cssText = "padding: 8px 12px; cursor: pointer; font-size: 13px; border-bottom: 1px solid #f0f0f0;";
    item.onmouseenter = function() { this.style.background = "#e9ecef"; };
    item.onmouseleave = function() { this.style.background = "white"; };
    item.onclick = function() { document.getElementById("boxplot_gene_input").value = ""; ac.style.display = "none"; addGeneByIdx(m.idx, m.name); };
    ac.appendChild(item);
  });
  ac.style.display = "block";
}
function boxplotAutoKey(e) {
  var ac = document.getElementById("boxplot_ac"); var items = ac.children;
  if (e.key==="ArrowDown"){boxplotAcIndex=Math.min(boxplotAcIndex+1,items.length-1);for(var i=0;i<items.length;i++)items[i].style.background=i===boxplotAcIndex?"#e9ecef":"white";}
  else if(e.key==="ArrowUp"){boxplotAcIndex=Math.max(boxplotAcIndex-1,0);for(var i=0;i<items.length;i++)items[i].style.background=i===boxplotAcIndex?"#e9ecef":"white";}
  else if(e.key==="Enter"){if(boxplotAcIndex>=0&&items[boxplotAcIndex])items[boxplotAcIndex].click(); else addGeneToPlot(); e.preventDefault();}
}
function addGeneToPlot() {
  var el = document.getElementById("boxplot_gene_input");
  var val = el.value.trim(); if (!val) return;
  var vl = val.toLowerCase();
  for (var i = 0; i < geneExprData.genes.length; i++) {
    if (geneExprData.genes[i].toLowerCase() === vl) { addGeneByIdx(i, geneExprData.genes[i]); el.value = ""; return; }
  }
  // Partial match
  for (var i = 0; i < geneExprData.genes.length; i++) {
    if (geneExprData.genes[i].toLowerCase().indexOf(vl) !== -1) { addGeneByIdx(i, geneExprData.genes[i]); el.value = ""; return; }
  }
}
function addGeneByIdx(idx, name) {
  if (selectedGenes.some(function(g){return g.idx===idx;})) return;
  selectedGenes.push({name:name, idx:idx}); renderGeneTags(); renderMultiBoxplot();
}
function removeGene(idx) {
  selectedGenes = selectedGenes.filter(function(g){return g.idx!==idx;});
  renderGeneTags(); if (selectedGenes.length>0) renderMultiBoxplot();
  else { document.getElementById("boxplot_container").innerHTML=""; document.getElementById("boxplot_gene_info").innerHTML=""; document.getElementById("boxplot_export_bar").style.display="none"; }
}
function clearBoxplot() {
  selectedGenes = []; renderGeneTags();
  document.getElementById("boxplot_container").innerHTML = "";
  document.getElementById("boxplot_gene_info").innerHTML = "";
  document.getElementById("boxplot_export_bar").style.display = "none";
}
function renderGeneTags() {
  var c = document.getElementById("boxplot_gene_tags"); c.innerHTML = "";
  selectedGenes.forEach(function(g) {
    var tag = document.createElement("span");
    tag.style.cssText = "display:inline-flex;align-items:center;background:#007bff;color:white;padding:3px 10px;border-radius:14px;font-size:12px;gap:4px;";
    var txt = document.createTextNode(g.name + " ");
    tag.appendChild(txt);
    var btn = document.createElement("span");
    btn.textContent = String.fromCharCode(215);
    btn.style.cssText = "cursor:pointer;font-weight:bold;margin-left:3px;";
    btn.addEventListener("click", (function(idx){return function(){removeGene(idx);};})(g.idx));
    tag.appendChild(btn);
    c.appendChild(tag);
  });
}
function renderMultiBoxplot() {
  if (selectedGenes.length === 0) return;
  var groups = geneExprData.groups; var samples = geneExprData.samples;
  var uGroups = []; groups.forEach(function(g){if(uGroups.indexOf(g)===-1)uGroups.push(g);});
  var colors = ["#e41a1c","#377eb8","#4daf4a","#984ea3","#ff7f00","#a65628","#f781bf","#999999"];
  var nGenes = selectedGenes.length; var traces = [];
  if (nGenes === 1) {
    var gIdx = selectedGenes[0].idx; var values = geneExprData.expr[gIdx];
    uGroups.forEach(function(g, gi) {
      var gv = [], gs = [];
      for (var i=0;i<values.length;i++) { if(groups[i]===g){gv.push(values[i]);gs.push(samples[i]);} }
      traces.push({y:gv,type:"box",name:g,boxpoints:"all",jitter:0.4,pointpos:0,text:gs,
        hovertemplate:"Sample: %%{text}<br>Expression: %%{y:.3f}<extra>"+g+"</extra>",
        marker:{color:colors[gi%%colors.length],size:6,opacity:0.7},
        line:{color:colors[gi%%colors.length]},fillcolor:colors[gi%%colors.length]+"33"});
    });
    var layout = {title:{text:selectedGenes[0].name+" Expression",font:{size:16}},yaxis:{title:bpYLabel},xaxis:{title:bpGroupLabel},showlegend:false,margin:{t:50,b:80,l:70,r:30},boxgap:0.3};
  } else {
    uGroups.forEach(function(g, gi) {
      var xv=[],yv=[],st=[];
      selectedGenes.forEach(function(sg){var vals=geneExprData.expr[sg.idx];for(var i=0;i<vals.length;i++){if(groups[i]===g){xv.push(sg.name);yv.push(vals[i]);st.push(samples[i]);}}});
      traces.push({x:xv,y:yv,type:"box",name:g,boxpoints:"all",jitter:0.4,pointpos:0,text:st,
        hovertemplate:"%%{x}<br>Sample: %%{text}<br>Expression: %%{y:.3f}<extra>"+g+"</extra>",
        marker:{color:colors[gi%%colors.length],size:5,opacity:0.7},
        line:{color:colors[gi%%colors.length]},fillcolor:colors[gi%%colors.length]+"33"});
    });
    var gNames = selectedGenes.map(function(sg){return sg.name;});
    var layout = {title:{text:"Lipid Expression Comparison",font:{size:16}},yaxis:{title:bpYLabel},xaxis:{title:"Lipid",categoryorder:"array",categoryarray:gNames},showlegend:true,legend:{title:{text:bpGroupLabel}},boxmode:"group",margin:{t:50,b:100,l:70,r:30},boxgap:0.15,boxgroupgap:0.1};
  }
  var ch = Math.max(450, 350 + nGenes * 30);
  document.getElementById("boxplot_container").style.height = ch + "px";
  Plotly.newPlot("boxplot_container", traces, layout, {responsive:true});
  var info = selectedGenes.map(function(sg){var vals=geneExprData.expr[sg.idx].filter(function(v){return !isNaN(v);});var mean=vals.reduce(function(a,b){return a+b;},0)/vals.length;return "<strong>"+sg.name+"</strong>: mean "+mean.toFixed(3)+" ("+vals.length+" samples)";});
  document.getElementById("boxplot_gene_info").innerHTML = info.join(" &nbsp;|&nbsp; ");
  document.getElementById("boxplot_export_bar").style.display = "flex";
}
function exportBoxplot(fmt) {
  var c = document.getElementById("boxplot_container");
  if (!c||!c.data){alert("No plot to export.");return;}
  var gNames = selectedGenes.map(function(g){return g.name;}).join("_");
  var w = fmt==="svg"?1200:2400, h = fmt==="svg"?800:1600, s = fmt==="svg"?1:3;
  Plotly.downloadImage(c,{format:fmt,width:w,height:h,scale:s,filename:"lipid_expression_"+gNames});
}
</script>\n', explorer_json_str, group_col_exp, norm_y_label_report))

  figure_legend("Lipid Expression Explorer. Search for lipids by name, add multiple for side-by-side comparison. Each box shows the distribution of normalized expression values per group. Individual sample points are overlaid. Use the export buttons to save as SVG or PNG.")
}
```

`r if (explorer_ready) "" else "-->"`

`r if (is.null(de_res) || is.null(de_res$summary_df) || !show_de_lip) "-->"`

---

`r if (is.null(de_res) || is.null(de_res$summary_df) || !show_de_heatmaps_lip) "<!--"`

# Top DE Heatmaps {.tabset}

Heatmaps of differentially expressed lipids (adjusted p-value < `r padj_cut_de`, z-score scaled) with hierarchical clustering.

```{r de-heatmaps, results='asis', fig.width=10, fig.height=8}
if (!is.null(de_res$summary_df) && requireNamespace("pheatmap", quietly = TRUE)) {
  # Find the Diagnostic_plots directory for heatmap images
  hm_out_dir <- NULL
  if (length(de_res$files) > 0) {
    png_files <- de_res$files[grepl("\\.png$", de_res$files)]
    if (length(png_files) > 0) {
      diag_dirs <- unique(dirname(png_files))
      diag_dirs <- diag_dirs[dir.exists(diag_dirs)]
      if (length(diag_dirs) > 0) hm_out_dir <- diag_dirs[1]
    }
  }
  if (is.null(hm_out_dir)) hm_out_dir <- file.path(dirname(knitr::current_input(dir = TRUE)), "lipidomics", "Diagnostic_plots")
  dir.create(hm_out_dir, recursive = TRUE, showWarnings = FALSE)

  # Identify DE features for heatmaps using raw P.Value (ignoring FC cutoff).
  # Heatmaps benefit from showing all significant lipids; the clustering
  # pattern reveals biology even for features with modest fold changes.
  sum_df <- de_res$summary_df
  pval_cols <- grep("^P\\.Value", colnames(sum_df), value = TRUE)
  if (length(pval_cols) > 0) {
    pval_mat <- sum_df[, pval_cols, drop = FALSE]
    min_pval_per_feat <- apply(pval_mat, 1, function(x) min(x, na.rm = TRUE))
    de_feats <- sum_df$feature_id[min_pval_per_feat <= padj_cut_de]
  } else {
    de_feats <- character(0)
  }

  # Sort by minimum raw p-value
  if (length(de_feats) > 0) {
    feat_pval <- min_pval_per_feat[sum_df$feature_id %in% de_feats]
    de_feats <- de_feats[order(feat_pval)]
  }

  # Use expr_log (transform-only) for heatmaps -- z-scoring removes scale anyway,
  # but expr_log preserves natural variance structure better than auto-scaled data.
  hm_expr <- pre$expr_log %||% pre$expr_work

  if (length(de_feats) >= 2 && !is.null(hm_expr)) {
    # Build annotation
    anno_df <- NULL
    if (!is.null(pre$meta) && cond_col %in% colnames(pre$meta)) {
      sample_col_hm <- lipid_cfg$effects$samples %||% "sample_id"
      anno_df <- data.frame(
        row.names = pre$meta[[sample_col_hm]],
        Group = pre$meta[[cond_col]],
        stringsAsFactors = FALSE
      )
    }

    # Build display labels (use Name if available)
    hm_label_map <- if (!is.null(name_map_v)) name_map_v else stats::setNames(de_feats, de_feats)

    for (n_top in c(10, 25, 50, 100)) {
      if (n_top > length(de_feats)) next
      top_ids <- utils::head(de_feats, n_top)

      mat_hm <- hm_expr[top_ids, , drop = FALSE]
      # Z-score scale rows
      mat_z <- t(scale(t(mat_hm)))
      mat_z[is.na(mat_z)] <- 0

      # Use display names for rows
      row_labels <- vapply(rownames(mat_z), function(fid) {
        nm <- hm_label_map[fid]
        if (!is.na(nm) && nzchar(nm)) nm else fid
      }, character(1), USE.NAMES = FALSE)
      rownames(mat_z) <- row_labels

      cat(sprintf("\n## Top %d\n\n", n_top))

      # Adjust figure height based on number of features
      fig_h <- max(6, min(20, 4 + n_top * 0.15))
      show_rownames <- n_top <= 60

      hm_file <- file.path(hm_out_dir, sprintf("heatmap_top%d_DE.png", n_top))
      hm_ok <- tryCatch({
        p_hm <- pheatmap::pheatmap(
          mat_z,
          annotation_col = anno_df,
          cluster_rows = TRUE,
          cluster_cols = TRUE,
          show_rownames = show_rownames,
          show_colnames = TRUE,
          fontsize_row = max(5, min(9, 120 / n_top)),
          fontsize_col = 8,
          main = sprintf("Top %d DE Lipids (z-score)", n_top),
          color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
          border_color = NA,
          silent = TRUE
        )
        ggplot2::ggsave(hm_file, p_hm, width = 10, height = fig_h, dpi = 150)
        TRUE
      }, error = function(e) { message("Heatmap top ", n_top, " failed: ", e$message); FALSE })
      if (hm_ok && file.exists(hm_file)) {
        cat(sprintf("\n![Top %d DE Heatmap](%s){width=100%%}\n\n", n_top, hm_file))
        figure_legend(sprintf(
          "Heatmap of the top %d differentially expressed lipids (ranked by adjusted p-value). Values are z-score normalized per lipid. Rows and columns are hierarchically clustered.",
          n_top
        ))
        cat("\n\n")
      }
    }

    # All DE heatmap (if reasonable size and different from last top-N)
    if (length(de_feats) >= 2 && length(de_feats) <= 500) {
      n_all <- length(de_feats)
      mat_all <- hm_expr[de_feats, , drop = FALSE]
      mat_z_all <- t(scale(t(mat_all)))
      mat_z_all[is.na(mat_z_all)] <- 0

      row_labels_all <- vapply(rownames(mat_z_all), function(fid) {
        nm <- hm_label_map[fid]
        if (!is.na(nm) && nzchar(nm)) nm else fid
      }, character(1), USE.NAMES = FALSE)
      rownames(mat_z_all) <- row_labels_all

      cat(sprintf("\n## All DE (%d)\n\n", n_all))

      fig_h_all <- max(8, min(30, 4 + n_all * 0.06))
      hm_all_file <- file.path(hm_out_dir, sprintf("heatmap_all%d_DE.png", n_all))
      hm_all_ok <- tryCatch({
        p_hm_all <- pheatmap::pheatmap(
          mat_z_all,
          annotation_col = anno_df,
          cluster_rows = TRUE,
          cluster_cols = TRUE,
          show_rownames = n_all <= 60,
          show_colnames = TRUE,
          fontsize_row = max(4, min(8, 100 / n_all)),
          fontsize_col = 8,
          main = sprintf("All %d DE Lipids (z-score)", n_all),
          color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
          border_color = NA,
          silent = TRUE
        )
        ggplot2::ggsave(hm_all_file, p_hm_all, width = 10, height = fig_h_all, dpi = 150)
        TRUE
      }, error = function(e) { message("Heatmap all DE failed: ", e$message); FALSE })
      if (hm_all_ok && file.exists(hm_all_file)) {
        cat(sprintf("\n![All DE Heatmap](%s){width=100%%}\n\n", hm_all_file))
      }
      figure_legend(sprintf(
        "Heatmap of all %d differentially expressed lipids. Values are z-score normalized per lipid.",
        n_all
      ))
      cat("\n\n")
    }
  } else if (length(de_feats) < 2) {
    cat("\n*Fewer than 2 DE lipids -- heatmap not generated.*\n\n")
  }
}
```

`r if (is.null(de_res) || is.null(de_res$summary_df) || !show_de_heatmaps_lip) "-->"`

---

```{r clustering-discover, include=FALSE}
# ---- Discover clustering outputs on disk for lipidomics ----
clustering_root_lip <- file.path(dirname(knitr::current_input(dir = TRUE)),
                                 "lipidomics", "Clustering")
hier_dir_l   <- file.path(clustering_root_lip, "Hierarchical")
part_root_l  <- file.path(clustering_root_lip, "Partition_clustering")
binary_dir_l <- file.path(clustering_root_lip, "Binary_patterns")

hier_hm_l <- file.path(hier_dir_l, "Hierarchical_DE_heatmap.png")

part_subdirs_l <- if (dir.exists(part_root_l)) {
  list.dirs(part_root_l, recursive = FALSE)
} else character(0)
part_subdir_l <- if (length(part_subdirs_l) > 0) part_subdirs_l[1] else NA_character_
part_hm_l     <- if (!is.na(part_subdir_l)) file.path(part_subdir_l, "Partition_clustering_heatmap.png") else NA_character_

part_profile_pngs_l <- if (!is.na(part_subdir_l) && dir.exists(part_subdir_l)) {
  list.files(part_subdir_l, pattern = "^cluster_profiles_cluster[0-9]+\\.png$",
             full.names = TRUE)
} else character(0)
if (length(part_profile_pngs_l) > 0) {
  ord <- order(as.integer(sub(".*cluster([0-9]+)\\.png$", "\\1",
                              basename(part_profile_pngs_l))))
  part_profile_pngs_l <- part_profile_pngs_l[ord]
}

binary_pngs_l <- if (dir.exists(binary_dir_l)) {
  list.files(binary_dir_l, pattern = "^Heatmap_[01]+\\.png$", full.names = TRUE)
} else character(0)

clustering_enabled_l <- isTRUE(lipid_cfg$clustering$enabled)
has_hier_l   <- file.exists(hier_hm_l)
has_part_l   <- !is.na(part_hm_l) && file.exists(part_hm_l)
has_binary_l <- length(binary_pngs_l) > 0
clustering_show_l <- clustering_enabled_l && (has_hier_l || has_part_l || has_binary_l)
```

```{r clustering-header-l, eval=clustering_show_l, results='asis'}
cat("# Clustering {.tabset}\n\n")
cat("Clustering was performed on the differentially expressed lipids. Each method below answers a different question:\n\n")
cat("- **Hierarchical** — bottom-up tree built from feature-by-feature distances; reveals nested groupings.\n")
cat("- **Partition** — splits features into a chosen number of clusters with similar trajectories across groups.\n")
cat("- **Binary patterns** — assigns each feature to an on/off (1/0) pattern across groups based on correlation.\n\n")
```

```{r clustering-hier-l, eval=clustering_show_l && has_hier_l, results='asis'}
cat("## Hierarchical\n\n")
cat(sprintf("![Hierarchical clustering heatmap](%s){width=100%%}\n\n", hier_hm_l))
figure_legend("Hierarchical clustering heatmap of differentially expressed lipids (rows z-scored). Row dendrogram reflects feature similarity; samples are annotated by experimental group.")
render_commentary("clustering_hierarchical")
```

```{r clustering-partition-l, eval=clustering_show_l && has_part_l, results='asis'}
cat("## Partition {.tabset}\n\n")
cat("### Heatmap\n\n")
cat(sprintf("![Partition clustering heatmap](%s){width=100%%}\n\n", part_hm_l))
figure_legend("Partition clustering heatmap. Rows are grouped by cluster (gaps separate clusters); within-cluster ordering follows the feature similarity.")

if (length(part_profile_pngs_l) > 0) {
  cat("\n### Cluster Profiles {.tabset}\n\n")
  for (pf in part_profile_pngs_l) {
    k <- sub(".*cluster([0-9]+)\\.png$", "\\1", basename(pf))
    cat(sprintf("\n#### Cluster %s\n\n", k))
    cat(sprintf("![Cluster %s expression profile](%s){width=80%%}\n\n", k, pf))
    figure_legend(sprintf(
      "Mean (±SE) expression profile across experimental groups for lipids in cluster %s.", k))
  }
}
render_commentary("clustering_partition")
```

```{r clustering-binary-l, eval=clustering_show_l && has_binary_l, results='asis'}
cat("## Binary Patterns {.tabset}\n\n")
for (bf in binary_pngs_l) {
  pat <- sub("^Heatmap_([01]+)\\.png$", "\\1", basename(bf))
  cat(sprintf("\n### Pattern %s\n\n", pat))
  cat(sprintf("![Binary pattern %s](%s){width=100%%}\n\n", pat, bf))
  figure_legend(sprintf(
    "Heatmap of lipids assigned to binary on/off pattern '%s' across experimental groups.", pat))
}
render_commentary("clustering_binary")
```

```{r clustering-divider-l, eval=clustering_show_l, results='asis'}
cat("\n---\n\n")
```

`r if (is.null(class_res) || !show_class_lip) "<!--"`

# Lipid Class Analysis {.tabset}

## Class Composition {.tabset}

### Normalized

```{r class-composition-norm, fig.width=10, fig.height=7, results='asis'}
if (!is.null(class_res$plots$class_normalized)) {
  safe_print_plot(class_res$plots$class_normalized)
  figure_legend("Normalized lipid class composition showing the relative proportion of each lipid class across samples or groups. Bars are scaled to 100% to facilitate comparison of class distributions.")
  render_commentary("class_normalized")
}
```

### Absolute

```{r class-composition-abs, fig.width=10, fig.height=7, results='asis'}
if (!is.null(class_res$plots$class_absolute)) {
  safe_print_plot(class_res$plots$class_absolute)
  figure_legend("Absolute lipid class composition showing the total intensity or count of each lipid class across samples or groups.")
  render_commentary("class_absolute")
}
```

## Chain Saturation

Lipids are grouped by the number of carbon-carbon double bonds (DB) in their fatty acyl chains:

- **SFA — Saturated Fatty Acids (0 DB):** no double bonds; chains pack tightly, increasing membrane rigidity and contributing to lipid storage.
- **MUFA — Monounsaturated Fatty Acids (1 DB):** a single double bond introduces a kink that increases membrane fluidity; common in dietary oils (e.g. oleic acid).
- **PUFA — Polyunsaturated Fatty Acids (2+ DB):** multiple double bonds further increase fluidity and serve as precursors for signaling lipids (eicosanoids, resolvins). Includes essential omega-3 and omega-6 fatty acids.

**Biological interpretation of shifts between groups:**

| Category | ↑ Increase typically reflects | ↓ Decrease typically reflects |
|----------|-------------------------------|-------------------------------|
| **SFA**  | Greater membrane rigidity; active de novo lipogenesis; pro-inflammatory signaling (e.g. palmitate → TLR4); commonly seen with high-fat diet, lipotoxicity, insulin resistance. | Increased membrane fluidity; reduced lipogenic flux; generally a metabolically healthier profile. |
| **MUFA** | Elevated SCD1 (Δ9-desaturase) activity converting SFA → MUFA; adaptive / cytoprotective response to lipid stress; oleic-acid-rich (Mediterranean-style) diets; anti-inflammatory tendency. | Reduced desaturation capacity; possible accumulation of toxic SFA species. |
| **PUFA** | Larger pool of essential fatty acids and signaling-lipid precursors (eicosanoids, resolvins); higher membrane fluidity; ω-3-driven shifts often anti-inflammatory; greater susceptibility to lipid peroxidation. | Reduced essential FA availability or dietary deficiency; smaller signaling-lipid pool; reduced membrane fluidity; possible oxidative losses. |

```{r class-chain-saturation, fig.width=10, fig.height=7, results='asis'}
if (!is.null(class_res$plots$chain_saturation)) {
  safe_print_plot(class_res$plots$chain_saturation)
  figure_legend("Distribution of fatty acyl chain saturation across lipid classes. This analysis reveals the degree of unsaturation (number of double bonds) in the detected lipidome, which is relevant to membrane fluidity and lipid signaling.")
  render_commentary("chain_saturation")
}
```

## Chain Length

```{r class-chain-length, fig.width=10, fig.height=7, results='asis'}
if (!is.null(class_res$plots$chain_length)) {
  safe_print_plot(class_res$plots$chain_length)
  figure_legend("Distribution of fatty acyl chain lengths across lipid classes. Chain length influences membrane structure and lipid function.")
  render_commentary("chain_length")
}
```

## Per-Class Chain Profiles {.tabset}

```{r chain-profiles-static, fig.width=10, fig.height=6, results='asis'}
if (!is.null(class_res$chain_profiles) && length(class_res$chain_profiles) > 0) {
  for (cls in names(class_res$chain_profiles)) {
    cat(paste0("\n### ", cls, "\n\n"))
    p <- class_res$plots[[paste0("chain_profile_", cls)]]
    if (!is.null(p)) {
      safe_print_plot(p)
    }
    cat("\n\n")
  }
  figure_legend("Per-class chain profiles showing individual fatty acid chain abundances within each lipid class. Select a lipid class tab to view its chain profile. Error bars show standard error.")
}
```

## Class Enrichment {.tabset}

### ORA Plot

```{r class-ora-plot, fig.width=10, fig.height=7, results='asis'}
if (!is.null(class_res$plots$class_ora)) {
  safe_print_plot(class_res$plots$class_ora)
  figure_legend("Over-representation analysis (ORA) of lipid classes among differentially expressed lipids. This test evaluates whether specific lipid classes are enriched among DE lipids compared to the full lipidome background.")
  render_commentary("class_ora")
}
```

### ORA Table

```{r class-ora-table, results='asis'}
if (!is.null(class_res$class_ora)) {
  print(htmltools::tagList(make_dt(class_res$class_ora, caption = "Lipid Class Over-Representation Analysis")))
}
```

`r if (is.null(class_res) || !show_class_lip) "-->"`

---

`r if ((is.null(rf_res) && is.null(plsda_res)) || !show_fs_lip) "<!--"`

# Feature Selection {.tabset}

`r if (is.null(rf_res)) "<!--"`

## Random Forest {.tabset}

### Importance Plot

```{r rf-plot, fig.width=9, fig.height=7, results='asis'}
if (!is.null(rf_res$plots$rf_importance)) {
  safe_print_plot(rf_res$plots$rf_importance)
  figure_legend("Random Forest variable importance plot showing the top features ranked by their contribution to classification accuracy. Higher Mean Decrease Accuracy indicates greater importance.")
  render_commentary("rf_importance")
}
```

### Importance Table

```{r rf-table, results='asis'}
if (!is.null(rf_res$importance_df)) {
  imp_df <- rf_res$importance_df
  sort_col <- if ("MeanDecreaseAccuracy" %in% names(imp_df)) "MeanDecreaseAccuracy" else "importance"
  imp_df <- imp_df[order(-imp_df[[sort_col]]), ]
  print(htmltools::tagList(make_dt(head(imp_df, 50), caption = "RF Feature Importance (top 50)")))
}
```

`r if (is.null(rf_res)) "-->"`

`r if (is.null(plsda_res)) "<!--"`

## PLS-DA {.tabset}

### Scores Plot

```{r plsda-scores, fig.width=9, fig.height=7, results='asis'}
if (!is.null(plsda_res$plots$plsda_scores)) {
  safe_print_plot(plsda_res$plots$plsda_scores)
  figure_legend("PLS-DA scores plot showing sample separation along the first two latent variables. Clustering by group indicates good discriminatory power of the lipid profiles.")
  render_commentary("plsda_scores")
}
```

### VIP Plot

```{r plsda-vip-plot, fig.width=9, fig.height=7, results='asis'}
if (!is.null(plsda_res$plots$plsda_vip)) {
  safe_print_plot(plsda_res$plots$plsda_vip)
  figure_legend("PLS-DA Variable Importance in Projection (VIP) scores. Features with VIP > 1 are generally considered important for group discrimination.")
  render_commentary("plsda_vip")
}
```

### VIP Table

```{r plsda-vip-table, results='asis'}
if (!is.null(plsda_res$vip_df)) {
  vip_df <- plsda_res$vip_df
  vip_col <- if ("VIP" %in% names(vip_df)) "VIP" else grep("^VIP", names(vip_df), value = TRUE)[1]
  vip_df <- vip_df[order(-vip_df[[vip_col]]), ]

  # Summary statistics to help interpret VIP scores
  vip_vals <- vip_df[[vip_col]]
  vip_mean <- round(mean(vip_vals, na.rm = TRUE), 2)
  vip_median <- round(median(vip_vals, na.rm = TRUE), 2)
  n_above_1 <- sum(vip_vals > 1, na.rm = TRUE)
  cat(sprintf(
    "<div style='background: #f0f7ff; border: 1px solid #b8daff; border-radius: 6px; padding: 12px 16px; margin-bottom: 12px; font-size: 0.92em;'><strong>VIP Summary:</strong> Mean = %s, Median = %s, Features with VIP &gt; 1.0: %d/%d. <em>VIP &gt; 1.0 is generally considered important for group discrimination.</em></div>\n",
    vip_mean, vip_median, n_above_1, length(vip_vals)
  ))

  print(htmltools::tagList(make_dt(head(vip_df, 50), caption = "PLS-DA VIP Scores (top 50)")))
}
```

`r if (is.null(plsda_res)) "-->"`

`r if ((is.null(rf_res) && is.null(plsda_res)) || !show_fs_lip) "-->"`

---

`r if (is.null(biomarker_res) || !show_biomarker_lip) "<!--"`

# Biomarker Discovery {.tabset}

## Biomarker Table

```{r biomarker-table-plot, fig.width=10, fig.height=8, results='asis'}
if (!is.null(biomarker_res$plots$biomarker_table)) {
  safe_print_plot(biomarker_res$plots$biomarker_table)
  figure_legend("Biomarker candidate summary showing key performance metrics for each candidate lipid. AUC (Area Under the ROC Curve) measures classification accuracy, Cohen's d quantifies effect size, and statistical power estimates the probability of detecting the observed effect at the given sample size.")
  render_commentary("biomarker_table")
}
```

## ROC Curves

```{r biomarker-roc, fig.width=10, fig.height=7, results='asis'}
if (!is.null(biomarker_res$plots$roc_curves)) {
  safe_print_plot(biomarker_res$plots$roc_curves)
  figure_legend("Receiver Operating Characteristic (ROC) curves for top biomarker candidates. Each curve shows the trade-off between sensitivity (true positive rate) and 1-specificity (false positive rate). AUC values closer to 1.0 indicate better discriminatory performance; the diagonal dashed line represents random classification (AUC = 0.5).")
  render_commentary("roc_curves")
}
```

## Multivariate ROC

```{r biomarker-roc-multi, fig.width=10, fig.height=7, results='asis'}
if (!is.null(biomarker_res$plots$roc_multivariate)) {
  safe_print_plot(biomarker_res$plots$roc_multivariate)
  figure_legend("Multivariate ROC curve combining multiple biomarker candidates into a single classifier. This evaluates whether a panel of lipids collectively achieves higher classification accuracy than individual markers alone.")
  render_commentary("roc_multivariate")
}
```

## Top Candidates

```{r biomarker-candidates, results='asis'}
if (!is.null(biomarker_res$biomarker_df)) {
  bm_df <- biomarker_res$biomarker_df
  if ("is_candidate" %in% colnames(bm_df)) {
    bm_show <- bm_df[bm_df$is_candidate == TRUE, , drop = FALSE]
    if (nrow(bm_show) == 0) bm_show <- bm_df
  } else {
    bm_show <- bm_df
  }
  bm_show <- head(bm_show, 20)
  cat("<div style='background: linear-gradient(135deg, #f8f9fa 0%, #e9ecef 100%); border: 1px solid #dee2e6; border-radius: 8px; padding: 20px 24px; margin: 10px 0 20px 0;'>\n")
  cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 12px;'><em>Top biomarker candidates ranked by classification performance. Candidates are selected based on AUC, effect size, and statistical significance.</em></p>\n")
  print(make_dt(bm_show, caption = "Top Biomarker Candidates"))
  cat("</div>\n")
}
```

`r if (is.null(biomarker_res) || !show_biomarker_lip) "-->"`

---

`r if (is.null(pathway_res) || !show_pathway_lip) "<!--"`

# Lipid Pathway Analysis {.tabset}

## Pathway Network

```{r pathway-network, fig.width=11, fig.height=9, results='asis'}
if (!is.null(pathway_res$plots$pathway_network)) {
  safe_print_plot(pathway_res$plots$pathway_network)
  figure_legend("Lipid pathway network showing metabolic relationships between lipid classes. Nodes represent lipid classes (sized by number of detected species), edges represent known enzymatic reactions. Node colors indicate the direction and magnitude of change: red shades denote upregulation, blue shades denote downregulation, and grey indicates no significant change.")
  render_commentary("pathway_network")
}
```

## Pathway Scores {.tabset}

### Barplot

```{r pathway-barplot, fig.width=10, fig.height=7, results='asis'}
if (!is.null(pathway_res$plots$pathway_barplot)) {
  safe_print_plot(pathway_res$plots$pathway_barplot)
  figure_legend("Pathway activity scores for each lipid metabolic pathway. Bars represent the aggregate dysregulation score combining fold-change direction and significance of member lipids. Pathways are ranked by statistical significance.")
  render_commentary("pathway_barplot")
}
```

### Scores Table

```{r pathway-scores-table, results='asis'}
if (!is.null(pathway_res$pathway_scores)) {
  cat("<div style='background: linear-gradient(135deg, #f8f9fa 0%, #e9ecef 100%); border: 1px solid #dee2e6; border-radius: 8px; padding: 20px 24px; margin: 10px 0 20px 0;'>\n")
  cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 12px;'><em>Pathway-level scores summarizing the direction and magnitude of lipid changes within each metabolic pathway. Lower p-values indicate coordinated dysregulation.</em></p>\n")
  print(make_dt(pathway_res$pathway_scores, caption = "Lipid Pathway Scores"))
  cat("</div>\n")
}
```

## Predicted Enzymes

```{r pathway-enzymes, results='asis'}
if (!is.null(pathway_res$enzyme_predictions)) {
  cat("<div style='background: linear-gradient(135deg, #f8f9fa 0%, #e9ecef 100%); border: 1px solid #dee2e6; border-radius: 8px; padding: 20px 24px; margin: 10px 0 20px 0;'>\n")
  cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 12px;'><em>Predicted enzymes involved in the conversion between differentially abundant lipid species. These predictions are inferred from known lipid metabolic reactions and the observed changes in substrate/product pairs.</em></p>\n")
  print(make_dt(pathway_res$enzyme_predictions, caption = "Predicted Enzymes in Dysregulated Pathways"))
  cat("</div>\n")

  figure_legend("Table of predicted enzymes catalysing reactions between differentially expressed lipid pairs. Enzyme annotations are derived from the LIPID MAPS and Reactome pathway databases. Reaction scores reflect the confidence and magnitude of the inferred enzymatic activity change.")
}
```

## Enzyme Network

```{r pathway-enzyme-net, fig.width=11, fig.height=9, results='asis'}
if (!is.null(pathway_res$plots$enzyme_network)) {
  safe_print_plot(pathway_res$plots$enzyme_network)
  figure_legend("Enzyme interaction network for predicted dysregulated enzymes. Nodes represent enzymes, and edges represent known protein-protein interactions from the STRING database. Clusters of interacting enzymes may indicate coordinated metabolic pathway perturbations.")
  render_commentary("enzyme_network")
} else if (!is.null(pathway_res)) {
  if (!requireNamespace("STRINGdb", quietly = TRUE)) {
    cat("<div style='background: #fff3cd; border: 1px solid #ffc107; border-radius: 8px; padding: 16px 20px; margin: 10px 0 20px 0;'>\n")
    cat("<p><strong>Enzyme interaction network not available</strong> — the <code>STRINGdb</code> package is not installed. Install it with <code>BiocManager::install('STRINGdb')</code> to enable enzyme interaction network visualization.</p>\n")
    cat("</div>\n")
  } else {
    cat("<div style='background: #f0f7ff; border: 1px solid #b8daff; border-radius: 8px; padding: 16px 20px; margin: 10px 0 20px 0;'>\n")
    cat("<p><strong>Enzyme interaction network not available</strong> — fewer than 2 enzymes mapped to STRING, or no interactions found above the score threshold.</p>\n")
    cat("</div>\n")
  }
}
```

`r if (is.null(pathway_res) || !show_pathway_lip) "-->"`

---

`r if (!show_output_files_lip) "<!--"`

# Output Files

```{r output-files}
files_tbl <- data.frame(
  File = c(
    "lipidomics_report.html",
    "de_summary_all.csv",
    "expression_matrix_normalized.csv"
  ),
  Description = c(
    "This HTML report with all QC, DE, and lipid class analysis results",
    "Full differential expression summary table across all contrasts",
    "Normalized lipid expression matrix used for downstream analysis"
  ),
  stringsAsFactors = FALSE
)

# Add lipid class analysis output rows if class analysis was run
if (!is.null(class_res)) {
  class_rows <- data.frame(
    File = "lipid_class_analysis.csv",
    Description = "Lipid class composition and enrichment results",
    stringsAsFactors = FALSE
  )
  files_tbl <- rbind(files_tbl, class_rows)
}

# Add feature selection outputs
if (!is.null(rf_res))
  files_tbl <- rbind(files_tbl, data.frame(
    File = "rf_importance.csv",
    Description = "Random Forest feature importance scores",
    stringsAsFactors = FALSE))
if (!is.null(plsda_res))
  files_tbl <- rbind(files_tbl, data.frame(
    File = "plsda_vip_scores.csv",
    Description = "PLS-DA VIP scores for all features",
    stringsAsFactors = FALSE))

# Add biomarker discovery outputs
if (!is.null(biomarker_res))
  files_tbl <- rbind(files_tbl, data.frame(
    File = "biomarker_candidates.csv",
    Description = "Biomarker candidate lipids with AUC, Cohen's d, and power metrics",
    stringsAsFactors = FALSE))

# Add pathway analysis outputs
if (!is.null(pathway_res)) {
  files_tbl <- rbind(files_tbl, data.frame(
    File = c("pathway_scores.csv", "enzyme_predictions.csv"),
    Description = c("Lipid pathway activity scores and significance",
                    "Predicted enzymes in dysregulated lipid pathways"),
    stringsAsFactors = FALSE))
}

make_dt(files_tbl, "Key Output Files")
```

`r if (!show_output_files_lip) "-->"`

---

`r if (!show_methods_lip) "<!--"`

# Methods

```{r methods-text, results='asis'}
padj_cut_m <- lipid_cfg$de$p_cutoff %||% 0.05
lfc_cut_m  <- lipid_cfg$de$linear_fc_cutoff %||% 1.5

n_lipids <- if (!is.null(pre)) nrow(pre$expr_work) else "N"
n_de_m  <- "N"
if (!is.null(de_res$summary_df) && "pass_any_contrast" %in% colnames(de_res$summary_df)) {
  n_de_m <- sum(de_res$summary_df$pass_any_contrast == 1, na.rm = TRUE)
}

cat("**Lipidomics Analysis Methods**\n\n")

cat("Lipidomics data were processed using the multiomics-core pipeline. ")

if (norm_method != "N/A") {
  cat(sprintf("Data were normalized using %s. ", norm_method))
}

cat(sprintf(
  "Differential expression analysis was performed using %s. ", de_method
))

cat(sprintf(
  "Lipids were considered differentially expressed at adjusted p-value <= %.2f and |fold change| >= %.2f. ",
  padj_cut_m, lfc_cut_m
))

cat(sprintf(
  "A total of %s lipid features were quantified, with %s identified as differentially expressed in at least one contrast.",
  format(as.numeric(n_lipids), big.mark = ","),
  format(as.numeric(n_de_m), big.mark = ",")
))

if (!is.null(class_res)) {
  cat(" Lipid class analysis was performed to characterize class composition, chain saturation, chain length distributions, and class-level over-representation among DE lipids.")
}

if (!is.null(rf_res) || !is.null(plsda_res)) {
  fs_methods <- character(0)
  if (!is.null(rf_res))    fs_methods <- c(fs_methods, "Random Forest")
  if (!is.null(plsda_res)) fs_methods <- c(fs_methods, "PLS-DA")
  cat(sprintf(" Feature selection was performed using %s.",
              paste(fs_methods, collapse = " and ")))
}

if (!is.null(biomarker_res)) {
  cat(" Biomarker discovery was performed by evaluating each lipid's classification performance using ROC analysis (AUC), effect size (Cohen's d), and statistical power.")
}

if (!is.null(pathway_res)) {
  cat(" Lipid pathway analysis was conducted to identify dysregulated metabolic pathways and predict involved enzymes based on substrate-product relationships in the LIPID MAPS and Reactome databases.")
}

cat("\n\n")

cat("**Copy-ready version:**\n\n")
cat("<div style='background-color: #f5f5f5; padding: 15px; border-radius: 5px; font-size: 0.9em;'>\n")

cat(sprintf(
  "Lipidomics analysis was performed using the multiomics-core pipeline (v1.0). %sDifferential expression was assessed with %s (adjusted p <= %.2f, |FC| >= %.2f). A total of %s lipid features were quantified, with %s differentially expressed.",
  if (norm_method != "N/A") sprintf("Data were normalized using %s. ", norm_method) else "",
  de_method, padj_cut_m, lfc_cut_m,
  format(as.numeric(n_lipids), big.mark = ","),
  format(as.numeric(n_de_m), big.mark = ",")
))

cat("\n</div>\n")
```

## Statistical Analysis

```{r methods-stats, results='asis'}
de_method_s <- de_method

test_desc <- switch(tolower(de_method_s),
  "t_test"       = "Welch's t-tests (unequal variance) were performed for each lipid.",
  "t_test_equal" = "Student's t-tests (equal variance assumed, MetaboAnalyst convention) were performed for each lipid.",
  "wilcoxon"     = "Wilcoxon rank-sum tests were used as a non-parametric alternative for differential expression testing.",
  "limma"        = "Moderated t-tests were performed using the *limma* R/Bioconductor package, which applies empirical Bayes shrinkage of variance estimates to improve statistical power.",
  sprintf("Differential expression was assessed using the *%s* method.", de_method_s)
)

cat(test_desc, "\n\n")

cat(sprintf(
  "Multiple testing correction was applied using the Benjamini-Hochberg procedure to control the false discovery rate (FDR). Lipids were considered differentially expressed at adjusted p-value $\\leq$ %s and |linear fold change| $\\geq$ %s.\n\n",
  padj_cut_m, lfc_cut_m
))

cat(sprintf(
  "All analyses were performed in R (version %s) using the multiomics-core pipeline.\n\n",
  paste(R.version$major, R.version$minor, sep = ".")
))
```

`r if (!show_methods_lip) "-->"`

---

`r if (!show_session_lip) "<!--"`

# Session Info

<details>
<summary>Click to expand</summary>

```{r session-info}
sessionInfo()
```

</details>

`r if (!show_session_lip) "-->"`

```{r dt-tab-fix, results='asis', echo=FALSE}
# Fix DT tables inside hidden tabs: redraw when tab becomes visible
cat('
<script>
$(document).on("shown.bs.tab", function(e) {
  $($(e.target).attr("href")).find(".datatables.html-widget").each(function() {
    var widget = HTMLWidgets.find("#" + this.id);
    if (widget && widget.table) {
      widget.table.columns.adjust().draw(false);
    }
  });
});
// Also trigger for any tables that may be in the initially active tab
$(document).ready(function() {
  setTimeout(function() {
    $(".datatables.html-widget").each(function() {
      var widget = HTMLWidgets.find("#" + this.id);
      if (widget && widget.table) {
        widget.table.columns.adjust().draw(false);
      }
    });
  }, 500);
});
</script>
')
```
