---
title: "`r local({ .c <- yaml::read_yaml(file.path(normalizePath(dirname(knitr::current_input())), 'execution_info', 'config_used.yaml')); pname <- .c$project$name %||% ''; analyst <- gsub('_', ' ', .c$project$analyst %||% ''); parts <- c('Proteomics Analysis Report'); if (nzchar(pname)) parts <- c(parts, pname); if (nzchar(analyst)) parts <- c(parts, analyst); paste(parts, collapse = ' — ') })`"
subtitle: "multiomics-core Pipeline"
author: "Computational analysis: `r yaml::read_yaml(file.path(normalizePath(dirname(knitr::current_input())), 'execution_info', 'config_used.yaml'))$project$analyst`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    fig_width: 10
    fig_height: 7
    self_contained: true
---

```{r setup, include=FALSE}
# error = TRUE: if any individual chunk errors (missing upstream data, empty
# results, absent OrgDb/network), print the error in place and keep rendering
# — the report ALWAYS completes, even if some sections are degraded.
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, error = TRUE,
  fig.align = "center", out.width = "100%"
)

library(yaml)
library(knitr)
library(DT)
library(plotly)
library(ggplot2)
library(htmltools)
library(jsonlite)

# Source interactive cutoff panel helpers
tryCatch({
  .cp_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
  .cp_found <- FALSE
  for (.cp_i in 1:6) {
    .cp_try <- file.path(.cp_dir, "R", "core", "17_cutoff_panel.R")
    if (file.exists(.cp_try)) { source(.cp_try); .cp_found <- TRUE; break }
    .cp_dir <- dirname(.cp_dir)
  }
  if (!.cp_found) message("cutoff_panel helpers not found (walked 6 levels up)")
}, error = function(e) message("cutoff_panel helpers not found: ", e$message))

# Load ggVennDiagram if available
has_ggVennDiagram <- FALSE
tryCatch({
    library(ggVennDiagram)
    has_ggVennDiagram <- TRUE
}, error = function(e) {
    message("ggVennDiagram not available: ", e$message)
})

# Helper function for downloadable DT tables
make_dt <- function(df, caption = NULL, pageLength = 15) {
    # Format p-value/padj columns to scientific notation
    pval_cols <- grep("^(pval|padj|p\\.val|adj\\.p\\.val|p_value|p\\.adjust|fdr|pvalue)$",
                      colnames(df), ignore.case = TRUE)
    for (col_idx in pval_cols) {
        vals <- df[[col_idx]]
        if (is.numeric(vals)) {
            df[[col_idx]] <- ifelse(is.na(vals), NA_character_,
                                     ifelse(vals < 0.001, sprintf("%.2e", vals),
                                            sprintf("%.4f", vals)))
        }
    }

    # Round other numeric columns to 3 decimal places
    num_cols <- setdiff(which(vapply(df, is.numeric, logical(1))), pval_cols)
    for (col_idx in num_cols) {
        df[[col_idx]] <- round(df[[col_idx]], 3)
    }

    DT::datatable(
        df,
        caption = caption,
        extensions = 'Buttons',
        options = list(
            scrollX = TRUE,
            pageLength = pageLength,
            dom = 'Bfrtip',
            buttons = list(
                list(extend = 'csv', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data")),
                list(extend = 'excel', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data"))
            )
        ),
        rownames = FALSE
    )
}

# Helper: render figure legend (paper-style caption below figure)
figure_legend <- function(text) {
    cat(sprintf("<p class='figure-legend' style='text-align: center; font-size: 0.9em; color: #555; margin: 4px auto 12px auto; max-width: 88%%; line-height: 1.45; font-style: italic;'>%s</p>\n", text))
}

# Paths
run_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
# Detect if we are already inside the proteomics output dir
if (basename(run_dir) == "proteomics") {
    prot_dir <- run_dir
    config_path <- file.path(dirname(run_dir), "execution_info", "config_used.yaml")
} else {
    prot_dir <- file.path(run_dir, "proteomics")
    config_path <- file.path(run_dir, "execution_info", "config_used.yaml")
}
diag_dir <- file.path(prot_dir, "Diagnostic_plots")
qc_post_dir <- file.path(diag_dir, "QC_post")
# Backwards-compat: older runs wrote DE PNGs (volcano / MA / heatmaps) to
# Diagnostic_plots/DE_plots/ before the QC_post layout was introduced.
# Use that directory as a fallback when QC_post is absent or empty.
de_plots_dir_legacy <- file.path(diag_dir, "DE_plots")
de_png_dir <- if (dir.exists(qc_post_dir) &&
                  length(list.files(qc_post_dir, pattern = "\\.png$")) > 0) {
    qc_post_dir
} else if (dir.exists(de_plots_dir_legacy) &&
           length(list.files(de_plots_dir_legacy, pattern = "\\.png$")) > 0) {
    de_plots_dir_legacy
} else {
    qc_post_dir
}
datasets_dir <- file.path(prot_dir, "Datasets")
clustering_dir <- file.path(prot_dir, "Clustering")
enrich_dir <- file.path(prot_dir, "Enrichment")
enrich_plot_dir <- file.path(enrich_dir, "plots")
ppi_dir <- file.path(prot_dir, "ppi_networks")
ppi_plot_dir <- file.path(ppi_dir, "plots")
adv_stats_dir <- file.path(prot_dir, "advanced_stats")
adv_stats_plot_dir <- file.path(adv_stats_dir, "plots")
commentary_dir <- file.path(prot_dir, "commentary")

# Load commentary if available
commentary <- NULL
has_commentary <- FALSE
commentary_json <- file.path(commentary_dir, "commentary_all.json")
if (file.exists(commentary_json)) {
    commentary <- tryCatch(
        jsonlite::read_json(commentary_json),
        error = function(e) NULL
    )
    has_commentary <- !is.null(commentary) && length(commentary) > 0
}

# Helper: render commentary block for a figure_id (collapsible)
render_commentary <- function(figure_id) {
    if (!has_commentary) return(invisible(NULL))
    c <- commentary[[figure_id]]
    if (is.null(c)) return(invisible(NULL))

    cat("<details class='commentary-details' style='margin: 4px 0 18px 0;'>\n")
    cat("<summary style='cursor: pointer; color: #4a90d9; font-weight: 500; font-size: 0.93em; padding: 3px 0;'>Show AI Interpretation</summary>\n")

    html <- "<div class='commentary-block' style='background: #f8f9fa; border-left: 4px solid #4a90d9; padding: 12px 16px; margin: 6px 0 4px 0; border-radius: 0 4px 4px 0; font-size: 0.92em;'>\n"
    html <- paste0(html, "<p style='margin: 0 0 8px 0;'><strong>What this shows:</strong> ",
                   htmltools::htmlEscape(c$what_is_this %||% ""), "</p>\n")

    bio_interp <- c$biological_interpretation %||% ""
    if (nzchar(bio_interp)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Interpretation:</strong> ",
                       htmltools::htmlEscape(bio_interp), "</p>\n")
    }

    obs <- c$observations
    if (length(obs) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Key observations:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (o in obs) html <- paste0(html, "<li>", htmltools::htmlEscape(o), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    stat_assess <- c$statistical_assessment %||% ""
    if (nzchar(stat_assess)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Statistical note:</strong> ",
                       htmltools::htmlEscape(stat_assess), "</p>\n")
    }

    issues <- c$issues_checks
    if (length(issues) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Checks:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (issue in issues) html <- paste0(html, "<li>", htmltools::htmlEscape(issue), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    steps <- c$next_steps
    if (length(steps) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Next steps:</strong></p>\n<ul style='margin: 2px 0 4px 0; padding-left: 20px;'>\n")
        for (s in steps) html <- paste0(html, "<li>", htmltools::htmlEscape(s), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    backend <- c$backend %||% "unknown"
    confidence <- c$confidence %||% ""
    backend_label <- if (grepl("claude", backend, ignore.case = TRUE)) {
        "Claude (Anthropic) + K-Dense AI scientific-skills"
    } else if (grepl("openai", backend, ignore.case = TRUE)) {
        "OpenAI + K-Dense AI scientific-skills"
    } else if (backend == "data_driven") {
        "data-driven + K-Dense AI scientific-skills"
    } else {
        backend
    }
    if (nzchar(confidence) && confidence != "none") {
        backend_label <- paste0(backend_label, " | confidence: ", confidence)
    }
    html <- paste0(html, "<p style='margin: 4px 0 0 0; color: #888; font-size: 0.85em;'><em>Generated by: ",
                   htmltools::htmlEscape(backend_label), "</em></p>\n</div>\n")

    cat(html)
    cat("</details>\n")
}

# Load config
cfg <- yaml::read_yaml(config_path)
prot_cfg <- cfg$modes$proteomics
report_cfg <- prot_cfg$report %||% list()  # new report-level parameters

# Resolve project.dir to absolute path (handles "." which is relative to repo root, not Rmd location)
proj_dir_resolved <- cfg$project$dir
if (is.null(proj_dir_resolved) || proj_dir_resolved == "." || proj_dir_resolved == "") {
    candidate <- normalizePath(run_dir, mustWork = FALSE)
    for (.i in 1:6) {
        if (dir.exists(file.path(candidate, "R")) && dir.exists(file.path(candidate, "config"))) break
        candidate <- dirname(candidate)
    }
    proj_dir_resolved <- candidate
}

# QC plot visibility flags (all default TRUE except distance heatmap)
show_pca       <- report_cfg$show_pca %||% TRUE
show_umap      <- report_cfg$show_umap %||% FALSE   # generated but hidden by default
show_pca_3d    <- report_cfg$show_pca_3d %||% TRUE
show_pca_subsets <- report_cfg$show_pca_subsets %||% TRUE
show_pca_topvar  <- report_cfg$show_pca_topvar %||% FALSE  # generated but hidden by default
show_density  <- report_cfg$show_density %||% TRUE
show_distance <- report_cfg$show_distance %||% TRUE
# Correlation heatmap dropped from report after Apr 2026 reviewer feedback
# (still produced in the pipeline). Override with `report.show_correlation: true`.
show_correlation <- report_cfg$show_correlation %||% FALSE
show_expression_heatmap <- report_cfg$show_expression_heatmap %||% TRUE
show_imputation_qc <- report_cfg$show_imputation_qc %||% TRUE
# Default-off after April 2026 reviewer feedback (Michal/Vered): keep volcano only.
show_ma_plots <- report_cfg$show_ma_plots %||% FALSE
show_top_de_heatmaps <- report_cfg$show_top_de_heatmaps %||% TRUE
show_protein_explorer <- report_cfg$show_protein_explorer %||% TRUE
show_peptide_explorer <- report_cfg$show_peptide_explorer %||% TRUE
show_technical_summary <- report_cfg$show_technical_summary %||% TRUE
show_executive_summary <- report_cfg$show_executive_summary %||% TRUE
show_methods <- report_cfg$show_methods %||% TRUE
show_session_info <- report_cfg$show_session_info %||% TRUE
force_pca <- report_cfg$force_pca %||% FALSE

# Load DE summary data (used in multiple sections). Use the same QC_post /
# DE_plots fallback resolver as the PNG chunks so older runs (which wrote
# de_summary_all.tsv under DE_plots/) still load the table.
de_summary_file <- file.path(de_png_dir, "de_summary_all.tsv")
de_data <- NULL
if (file.exists(de_summary_file)) {
    de_data <- read.delim(de_summary_file, stringsAsFactors = FALSE, check.names = FALSE)
}

# Get contrast names from column patterns
contrast_names <- character(0)
if (!is.null(de_data)) {
    padj_cols <- grep("^padj\\.imputs\\.", names(de_data), value = TRUE)
    contrast_names <- gsub("^padj\\.imputs\\.", "", padj_cols)
}
```

```{r force-dependencies, include=FALSE}
# Force DT and Plotly dependencies to be included in the HTML
DT::datatable(data.frame(x = 1))
plotly::plot_ly(data.frame(x = 1, y = 1), x = ~x, y = ~y, type = "scatter", mode = "markers")
```

```{r introduction, results='asis'}
# Introduction — optional and flexible.  Rendered only when at least one
# text field is provided (user_notes, introduction_text, introduction_background,
# introduction_aim).  Each sub-part is omitted silently when empty.
intro_text  <- report_cfg$introduction_text %||% ""
intro_bg    <- report_cfg$introduction_background %||% ""
intro_aim   <- report_cfg$introduction_aim %||% ""
user_notes  <- cfg$project$user_notes %||% ""

# Project summary HTML (generated from technical report / user notes)
summary_file <- file.path(prot_dir, "project_summary.html")
has_summary_file <- file.exists(summary_file)

has_intro <- any(nzchar(trimws(c(intro_text, intro_bg, intro_aim, user_notes)))) || has_summary_file

if (has_intro) {
    cat("# Introduction\n\n")

    # Project summary (from technical report / AI blend) takes priority
    if (has_summary_file) {
        cat(paste(readLines(summary_file, warn = FALSE), collapse = "\n"))
        cat("\n\n")
    } else {
        # Fallback: render user_notes or intro text as styled block
        body_text <- if (nzchar(trimws(intro_text))) intro_text else user_notes
        if (nzchar(trimws(body_text))) {
            notes_html <- gsub("\n", "<br>\n", htmltools::htmlEscape(trimws(body_text)))
            cat(sprintf("<div style='background:#f8f9fa; border-left:4px solid #4a90d9; padding:16px 20px; border-radius:6px; margin-bottom:20px; font-size:14px; line-height:1.7; color:#333;'>%s</div>\n", notes_html))
            cat("\n")
        }
    }

    if (nzchar(trimws(intro_bg))) {
        cat("## Background\n\n", trimws(intro_bg), "\n\n")
    }

    if (nzchar(trimws(intro_aim))) {
        cat("## Aim\n\n", trimws(intro_aim), "\n\n")
    }
}
```

```{r experimental-design-section, results='asis'}
# Standalone Experimental Design block intentionally suppressed; the sample-key
# table is rendered inside the Analysis Parameters > Experimental Design tab.
```

# Analysis Parameters {.tabset}

## Project

```{r params-project}
proj <- data.frame(
  Parameter = c("Project Name", "Analysis Round", "Project Directory",
                 "Raw Data Path", "Output Path"),
  Value = c(cfg$project$name, cfg$project$analysis_round, cfg$project$dir,
            cfg$paths$raw, cfg$paths$out),
  stringsAsFactors = FALSE
)
make_dt(proj, "Analysis Parameters")
```

## Experimental Design

```{r params-design, results='asis'}
meta_file <- file.path(proj_dir_resolved, cfg$paths$raw, prot_cfg$files$metadata)
if (file.exists(meta_file)) {
    meta <- read.csv(meta_file, stringsAsFactors = FALSE)
    group_col <- prot_cfg$effects$color %||% "Condition"

    if (group_col %in% names(meta)) {
        design_tbl <- as.data.frame(table(meta[[group_col]]))
        colnames(design_tbl) <- c("Group", "n (replicates)")
        cat("**Replicates per group:**\n\n")
        print(htmltools::tagList(make_dt(design_tbl, "Replicates per group")))
        cat("\n\n")
    }

    # Optional design file/link (replaces the previous standalone block)
    design_file <- report_cfg$experimental_design_file %||% ""
    if (nzchar(trimws(design_file))) {
        if (grepl("^https?://", design_file)) {
            cat(sprintf("Full experimental design: [%s](%s)\n\n", design_file, design_file))
        } else if (file.exists(design_file)) {
            cat(sprintf("Design file: `%s`\n\n", design_file))
        } else {
            cat(sprintf("Design file reference: `%s`\n\n", design_file))
        }
    }

    # Sample-key datatable (full metadata)
    cat("**Sample key:**\n\n")
    print(htmltools::tagList(make_dt(meta, "Sample Metadata")))
    cat("\n\n")
}
```

## Pipeline Settings

```{r params-pipeline}
settings <- data.frame(
  Parameter = c(
    "Search Engine",
    "Imputation Method",
    "DE Method",
    "Adjusted p-value Cutoff",
    "Linear FC Cutoff",
    "Clustering Enabled",
    "Pathway Analysis",
    "PPI Networks",
    "Advanced Statistics"
  ),
  Value = c(
    prot_cfg$engine %||% "unknown",
    prot_cfg$imputation$method %||% "none",
    prot_cfg$de$method %||% "limma",
    prot_cfg$de$p_cutoff %||% 0.05,
    prot_cfg$de$linear_fc_cutoff %||% 1.5,
    ifelse(isTRUE(prot_cfg$clustering$enabled), "Yes", "No"),
    ifelse(isTRUE(prot_cfg$pathway$enabled), "Yes", "No"),
    ifelse(isTRUE(prot_cfg$ppi$enabled), "Yes", "No"),
    ifelse(isTRUE(prot_cfg$advanced_stats$enabled), "Yes", "No")
  ),
  stringsAsFactors = FALSE
)
make_dt(settings, "Pipeline Settings")
```

## Contrasts

```{r params-contrasts}
contrasts_file <- file.path(proj_dir_resolved, cfg$paths$raw, prot_cfg$files$contrasts)
if (file.exists(contrasts_file)) {
    contrasts <- read.csv(contrasts_file, stringsAsFactors = FALSE)
    make_dt(contrasts, "Contrasts")
} else {
    cat("No contrasts file specified.\n")
}
```

---

```{r technical-summary-heading, eval=show_technical_summary, results='asis'}
cat("# Technical Summary\n\n")
```

```{r technical-summary, eval=show_technical_summary, results='asis'}
tech <- prot_cfg$technical_report

meta_ts <- tryCatch(
    read.csv(file.path(proj_dir_resolved, cfg$paths$raw, prot_cfg$files$metadata),
             stringsAsFactors = FALSE),
    error = function(e) NULL
)
group_col_ts <- prot_cfg$effects$color

n_samples_ts <- if (!is.null(meta_ts)) nrow(meta_ts) else NA
n_groups_ts <- if (!is.null(meta_ts) && group_col_ts %in% names(meta_ts)) length(unique(meta_ts[[group_col_ts]])) else NA

cat("## Sample & Acquisition Overview\n\n")
overview <- character(0)
if (!is.na(n_samples_ts) && !is.na(n_groups_ts)) {
    overview <- c(overview, sprintf("This analysis comprises **%d samples** across **%d experimental groups** (grouping variable: *%s*).",
                                     n_samples_ts, n_groups_ts, group_col_ts))
}
if (!is.null(tech$facility) && nzchar(tech$facility)) {
    overview <- c(overview, sprintf("Data were acquired at the **%s**.", tech$facility))
}
if (length(overview) > 0) cat(paste(overview, collapse = " "), "\n\n")

if (!is.null(tech)) {
    if (!is.null(tech$sample_prep) && nzchar(tech$sample_prep)) {
        cat("## Sample Preparation\n\n", tech$sample_prep, "\n\n")
    }
    if (!is.null(tech$ms_acquisition) && nzchar(tech$ms_acquisition)) {
        cat("## MS Acquisition\n\n", tech$ms_acquisition, "\n\n")
    }
    if (!is.null(tech$search_engine) && nzchar(tech$search_engine)) {
        cat("## Database Search\n\n", tech$search_engine, "\n\n")
    }
    if (!is.null(tech$search_parameters) && nzchar(tech$search_parameters)) {
        cat("## Search Parameters\n\n", tech$search_parameters, "\n\n")
    }
}

cat("## Downstream Analysis (multiomics-core)\n\n")
cat(sprintf("- **Imputation:** %s", prot_cfg$imputation$method %||% "none"))
if (!is.null(prot_cfg$imputation$width)) cat(sprintf(" (width=%.1f, downshift=%.1f)", prot_cfg$imputation$width, prot_cfg$imputation$downshift))
cat("\n")
cat(sprintf("- **DE method:** %s (adj.p <= %s, |FC| >= %s)\n",
            prot_cfg$de$method %||% "limma", prot_cfg$de$p_cutoff %||% 0.05, prot_cfg$de$linear_fc_cutoff %||% 1.5))
cat(sprintf("- **Pathway:** %s\n", if (isTRUE(prot_cfg$pathway$enabled)) paste(prot_cfg$pathway$databases, collapse = ", ") else "disabled"))
cat(sprintf("- **PPI:** %s\n", if (isTRUE(prot_cfg$ppi$enabled)) "STRING" else "disabled"))
```

---

```{r executive-summary-heading, eval=show_executive_summary, results='asis'}
cat("# Executive Summary\n\n")
```

```{r executive-summary, eval=show_executive_summary, results='asis'}
exec_summary_path <- file.path(prot_dir, "executive_summary.md")
exec_available <- file.exists(exec_summary_path)

# Build key metrics panel
cat("<div style='background: linear-gradient(135deg, #f8f9fa 0%, #e9ecef 100%); border: 1px solid #dee2e6; border-radius: 8px; padding: 20px 24px; margin: 10px 0 20px 0;'>\n")
cat("<div style='display: flex; flex-wrap: wrap; gap: 16px; margin-bottom: 16px;'>\n")

# Samples
if (exists("meta") && is.data.frame(meta)) {
    n_samp <- nrow(meta)
    cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #2c3e50;'>%d</div><div style='font-size: 0.85em; color: #666;'>Samples</div></div>\n", n_samp))
}

# Groups
if (exists("meta") && is.data.frame(meta) && exists("group_col") && group_col %in% names(meta)) {
    n_groups <- length(unique(meta[[group_col]]))
    cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #2c3e50;'>%d</div><div style='font-size: 0.85em; color: #666;'>Groups</div></div>\n", n_groups))
}

# Proteins quantified
if (!is.null(de_data)) {
    n_proteins <- nrow(de_data)
    cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #2c3e50;'>%s</div><div style='font-size: 0.85em; color: #666;'>Proteins Quantified</div></div>\n", format(n_proteins, big.mark = ",")))
}

# Contrasts
n_contrasts <- length(contrast_names)
if (n_contrasts > 0) {
    cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #2c3e50;'>%d</div><div style='font-size: 0.85em; color: #666;'>Contrasts</div></div>\n", n_contrasts))
}

# Total DE proteins
if (!is.null(de_data)) {
    pass_any_col <- grep("pass_any", names(de_data), value = TRUE)[1]
    if (!is.na(pass_any_col)) {
        n_de_total <- sum(de_data[[pass_any_col]] == TRUE, na.rm = TRUE)
        cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #c0392b;'>%s</div><div style='font-size: 0.85em; color: #666;'>DE Proteins</div></div>\n", format(n_de_total, big.mark = ",")))
    }
}

cat("</div>\n")

# DE breakdown per contrast
if (!is.null(de_data) && length(contrast_names) > 0) {
    padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- prot_cfg$de$linear_fc_cutoff %||% 1.5

    cat("<div style='margin: 12px 0;'>\n")
    cat("<strong>Differential Expression per Contrast:</strong>\n<ul style='margin: 6px 0; padding-left: 20px;'>\n")
    for (cn in contrast_names) {
        pass_col <- paste0("pass.imputs.", cn)
        lfc_col <- paste0("linearFC.imputs.", cn)
        padj_col <- paste0("padj.imputs.", cn)
        n_up <- n_down <- NULL
        if (pass_col %in% names(de_data) && lfc_col %in% names(de_data)) {
            is_sig <- !is.na(de_data[[pass_col]]) & de_data[[pass_col]] == 1
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            n_up <- sum(is_sig & lfc_vals > 0, na.rm = TRUE)
            n_down <- sum(is_sig & lfc_vals < 0, na.rm = TRUE)
        } else if (padj_col %in% names(de_data) && lfc_col %in% names(de_data)) {
            padj_vals <- as.numeric(de_data[[padj_col]])
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            log2fc <- signed_fc_to_log2(lfc_vals)
            is_sig <- !is.na(padj_vals) & padj_vals <= padj_cut & abs(log2fc) >= log2(lfc_cut)
            n_up <- sum(is_sig & log2fc > 0, na.rm = TRUE)
            n_down <- sum(is_sig & log2fc < 0, na.rm = TRUE)
        }
        if (!is.null(n_up)) {
            cat(sprintf("<li><strong>%s</strong>: %d DE proteins (%d up, %d down)</li>\n",
                        cn, n_up + n_down, n_up, n_down))
        }
    }
    cat("</ul>\n</div>\n")
}

# Pipeline-generated executive summary
if (exec_available) {
    exec_md_raw <- paste(readLines(exec_summary_path, warn = FALSE), collapse = "\n")
    exec_md_raw <- gsub("^##?\\s*Executive\\s*Summary\\s*\n+", "", exec_md_raw)

    # Drop verbose / redundant lines that already appear in dedicated downstream
    # sections (avoids duplication in the Executive Summary panel):
    #   - Top DE proteins list  -> shown in DE Results / volcano sections
    #   - PPI network counts    -> shown in PPI Network Analysis
    #   - Top hub proteins list -> shown in Hub Proteins table
    .lines <- strsplit(exec_md_raw, "\n", fixed = TRUE)[[1]]
    .drop_pat <- "(?i)(top de proteins|top hub proteins|ppi network)"
    # Also drop any indented continuation line that follows a dropped header
    keep <- rep(TRUE, length(.lines))
    for (i in seq_along(.lines)) {
        if (grepl(.drop_pat, .lines[i], perl = TRUE)) keep[i] <- FALSE
    }
    exec_md_raw <- paste(.lines[keep], collapse = "\n")

    if (nzchar(trimws(exec_md_raw)) && trimws(exec_md_raw) != "No summary available.") {
        cat("<div style='margin-top: 12px; padding-top: 12px; border-top: 1px solid #dee2e6;'>\n")
        cat(exec_md_raw, "\n")
        cat("</div>\n")
    }
}

cat("</div>\n")
```

```{r technical-metadata-table, results='asis'}
# Technical metadata table — what ran, and feature counts before/after filtering.
.tm_n_after <- if (!is.null(de_data)) nrow(de_data) else NA_integer_

# Try to count raw proteins from the source file (best-effort; may be NA for
# preprocessed inputs).
.tm_n_before <- NA_integer_
.tm_raw_path <- tryCatch({
    raw_rel <- prot_cfg$files$protein %||% prot_cfg$files$preprocessed_protein %||% ""
    if (nzchar(raw_rel)) file.path(proj_dir_resolved, cfg$paths$raw, raw_rel) else ""
}, error = function(e) "")
if (nzchar(.tm_raw_path) && file.exists(.tm_raw_path)) {
    .tm_n_before <- tryCatch({
        sep <- if (grepl("\\.tsv$", .tm_raw_path, ignore.case = TRUE)) "\t" else ","
        nrow(read.delim(.tm_raw_path, sep = sep, stringsAsFactors = FALSE,
                         check.names = FALSE))
    }, error = function(e) NA_integer_)
}

.tm_steps <- list(
    list(label = "Filtering",
         ran   = !is.na(.tm_n_before) && !is.na(.tm_n_after) && .tm_n_before != .tm_n_after),
    list(label = "Imputation",
         ran   = (tolower(prot_cfg$imputation$method %||% "none") != "none")),
    list(label = "Normalization",
         ran   = (tolower(prot_cfg$normalization$method %||% "none") != "none")),
    list(label = "Differential expression",
         ran   = !is.null(de_data) && nrow(de_data) > 0),
    list(label = "Clustering",
         ran   = isTRUE(prot_cfg$clustering$enabled)),
    list(label = "Pathway enrichment",
         ran   = isTRUE(prot_cfg$pathway$enabled)),
    list(label = "PPI network",
         ran   = isTRUE(prot_cfg$ppi$enabled)),
    list(label = "Advanced statistics",
         ran   = isTRUE(prot_cfg$advanced_stats$enabled))
)
.tm_steps_str <- paste(vapply(.tm_steps, function(s) {
    sprintf("%s: %s", s$label, if (isTRUE(s$ran)) "yes" else "skipped")
}, character(1)), collapse = "<br>")

.tm_fmt_n <- function(n) if (is.na(n)) "n/a" else format(n, big.mark = ",")

cat('<table style="margin: 16px 0 24px 0; border-collapse: collapse; width: 100%; border: 1px solid #d0d7de; border-radius: 6px; overflow: hidden; font-size: 14px;">\n')
cat('<thead><tr style="background: #f6f8fa; color: #24292f;">\n')
cat('<th style="text-align:left; padding: 10px 14px; border-bottom: 1px solid #d0d7de;">Technical metadata</th>\n')
cat('<th style="text-align:left; padding: 10px 14px; border-bottom: 1px solid #d0d7de;">Value</th>\n')
cat('</tr></thead><tbody>\n')

.tm_row <- function(k, v) cat(sprintf(
    '<tr><td style="padding: 8px 14px; border-bottom: 1px solid #eaecef; font-weight: 600; color: #24292f;">%s</td><td style="padding: 8px 14px; border-bottom: 1px solid #eaecef; color: #57606a;">%s</td></tr>\n',
    k, v))

.tm_row("Analysis steps performed", .tm_steps_str)
.tm_row("Proteins before filtering", .tm_fmt_n(.tm_n_before))
.tm_row("Proteins after filtering",  .tm_fmt_n(.tm_n_after))
if (!is.na(.tm_n_before) && !is.na(.tm_n_after) && .tm_n_before > 0) {
    .tm_row("Retained", sprintf("%.1f%%", 100 * .tm_n_after / .tm_n_before))
}
.tm_row("DE method",
        sprintf("%s (padj &le; %s, |linear FC| &ge; %s)",
                toupper(prot_cfg$de$method %||% "limma"),
                signif(prot_cfg$de$p_cutoff %||% 0.05, 3),
                prot_cfg$de$linear_fc_cutoff %||% 1.5))
cat('</tbody></table>\n')
```

---

# Results

```{r data-file-links, results='asis'}
raw_loc <- report_cfg$raw_files_location %||% ""
analysis_loc <- report_cfg$analysis_files_location %||% ""
if (nzchar(trimws(raw_loc)) || nzchar(trimws(analysis_loc))) {
    cat("<div style='background:#f0f4f8; border:1px solid #d0d7de; border-radius:6px; padding:12px 18px; margin-bottom:20px;'>\n")
    cat("<strong>Data Locations:</strong><br>\n")
    if (nzchar(trimws(raw_loc))) {
        if (grepl("^https?://", raw_loc)) {
            cat(sprintf("Raw instrument files: <a href='%s'>%s</a><br>\n", raw_loc, raw_loc))
        } else {
            cat(sprintf("Raw instrument files: <code>%s</code><br>\n", raw_loc))
        }
    }
    if (nzchar(trimws(analysis_loc))) {
        if (grepl("^https?://", analysis_loc)) {
            cat(sprintf("Analysis output files: <a href='%s'>%s</a><br>\n", analysis_loc, analysis_loc))
        } else {
            cat(sprintf("Analysis output files: <code>%s</code><br>\n", analysis_loc))
        }
    }
    cat("</div>\n\n")
}
```

## Quality Control — Pre-DE

### Filtering Statistics

```{r filtering-stats, results='asis'}
# Compact summary table only — full strategy/threshold rationale lives
# in the Technical Metadata Table at the top of the report.
filt_stats_file <- file.path(prot_dir, "filtering_stats.csv")
rows <- list()
if (file.exists(filt_stats_file)) {
    filt_stats <- read.csv(filt_stats_file, stringsAsFactors = FALSE)
    for (i in seq_len(nrow(filt_stats))) {
        rows[[length(rows) + 1L]] <- c(filt_stats$metric[i], filt_stats$value[i])
    }
} else if (!is.null(de_data)) {
    rows[[length(rows) + 1L]] <- c("Proteins in final dataset",
                                    format(nrow(de_data), big.mark = ","))
    imp_method <- prot_cfg$imputation$method %||% "none"
if (imp_method == "perseus") imp_method <- "perseus_like"
    if (tolower(imp_method) != "none") {
        rows[[length(rows) + 1L]] <- c("Imputation method", imp_method)
    }
}
if (length(rows) > 0) {
    fs_df <- do.call(rbind, lapply(rows, function(r) data.frame(Metric = r[1], Value = r[2], stringsAsFactors = FALSE)))
    print(htmltools::tagList(make_dt(fs_df, "Filtering Statistics")))
}
```

```{r expected-proteins-bar, results='asis', fig.height=5, fig.width=8}
# Optional bar chart of expected / key proteins
show_expected_bar <- isTRUE(report_cfg$show_expected_proteins_bar)
expected_file <- report_cfg$expected_proteins_file %||% ""

if (show_expected_bar && nzchar(trimws(expected_file)) && file.exists(expected_file)) {
    cat("### Expected Proteins\n\n")
    exp_df <- tryCatch(read.csv(expected_file, stringsAsFactors = FALSE), error = function(e) NULL)
    if (!is.null(exp_df) && nrow(exp_df) > 0) {
        # Expect columns: protein (or protein_id), value (optional)
        pcol <- intersect(c("protein", "protein_id", "gene", "gene_name"), tolower(names(exp_df)))[1]
        vcol <- intersect(c("value", "intensity", "abundance", "log2fc"), tolower(names(exp_df)))[1]
        if (!is.na(pcol)) {
            names(exp_df)[tolower(names(exp_df)) == pcol] <- "protein"
            if (!is.na(vcol)) {
                names(exp_df)[tolower(names(exp_df)) == vcol] <- "value"
            } else {
                exp_df$value <- seq_len(nrow(exp_df))
            }
            exp_df <- exp_df[order(exp_df$value, decreasing = TRUE), ]
            p <- ggplot2::ggplot(exp_df,
                     ggplot2::aes(x = reorder(protein, value), y = value)) +
                ggplot2::geom_bar(stat = "identity", fill = "steelblue") +
                ggplot2::coord_flip() +
                ggplot2::labs(title = "Expected Proteins", x = NULL, y = "Value") +
                ggplot2::theme_minimal()
            print(p)
            figure_legend("Bar chart of expected/key proteins relevant to this experiment.")
        }
    }
} else if (show_expected_bar) {
    cat("### Expected Proteins\n\n")
    cat("*Expected proteins bar chart enabled but no data file provided (`report.expected_proteins_file`).*\n\n")
}
```

```{r pca-replicate-check, include=FALSE}
# PCA is only meaningful when there are replicates (>=2 samples per condition)
has_replicates <- FALSE
if (exists("meta") && is.data.frame(meta) && exists("group_col") && group_col %in% names(meta)) {
    group_counts <- table(meta[[group_col]])
    has_replicates <- any(group_counts >= 2)
}
show_pca_section <- (isTRUE(show_pca) && has_replicates) || isTRUE(force_pca)
```

```{r pca-skip-msg, eval=!show_pca_section, results='asis'}
if (isTRUE(show_pca) && !has_replicates) {
    cat("## PCA\n\n*PCA plots are omitted because no condition has replicate samples. Set `report.force_pca: true` to override.*\n\n")
}
```

```{r pca-heading, eval=show_pca_section, results='asis'}
cat("## PCA {.tabset}\n\n")
cat("Principal Component Analysis (PCA) reduces high-dimensional expression data to a low-dimensional representation that captures the largest sources of variation. Samples that cluster together share similar expression profiles; separation along PC1 indicates the dominant source of biological or technical variation.\n\n")
cat("### PC1 vs PC2 {.tabset}\n")
cat("#### Static\n")
```

```{r pca-12, eval=show_pca_section, fig.cap="PCA -- PC1 vs PC2"}
p <- file.path(diag_dir, "PCA_PC1.vs.PC2.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-pca-12, eval=show_pca_section, results='asis'}
figure_legend("Principal Component Analysis (PCA) of imputed protein expression data, showing sample separation along PC1 and PC2. Each point represents one sample, colored by experimental group.")
```

```{r commentary-pca, eval=show_pca_section, results='asis'}
render_commentary("pca_plot")
```

```{r pca-12-labeled-heading, eval=show_pca_section, results='asis'}
cat("#### With Sample Names\n")
```

```{r pca-12-labeled, eval=show_pca_section, fig.cap="PCA -- PC1 vs PC2 (labeled)"}
p_lab <- file.path(diag_dir, "PCA_PC1.vs.PC2_labeled.png")
if (file.exists(p_lab)) knitr::include_graphics(p_lab) else cat("*Labeled PCA plot not available.*\n")
```

```{r pca-12-interactive-heading, eval=show_pca_section, results='asis'}
cat("#### Interactive\n")
```

```{r pca-12-interactive, eval=show_pca_section, fig.width=8, fig.height=6}
scores_csv <- file.path(diag_dir, "PCA_PC1.vs.PC2_scores.csv")
if (file.exists(scores_csv) && requireNamespace("plotly", quietly = TRUE)) {
    sc <- read.csv(scores_csv, stringsAsFactors = FALSE)
    color_var <- prot_cfg$effects$color %||% "Condition"
    if (!color_var %in% colnames(sc)) color_var <- setdiff(colnames(sc), c("sample", grep("^PC", colnames(sc), value = TRUE)))[1]
    sc$hover_text <- paste0("Sample: ", sc$sample, "<br>", color_var, ": ", sc[[color_var]])
    pl <- plotly::plot_ly(sc, x = ~PC1, y = ~PC2, color = as.formula(paste0("~`", color_var, "`")),
                          text = ~hover_text, hoverinfo = "text",
                          type = "scatter", mode = "markers",
                          marker = list(size = 10)) |>
        plotly::layout(xaxis = list(title = colnames(sc)[1]),
                       yaxis = list(title = colnames(sc)[2]),
                       title = "PCA: PC1 vs PC2 — Hover for sample names")
    pl
} else {
    cat("*Interactive PCA not available (scores CSV or plotly package missing).*\n")
}
```

```{r pca-topvar-block, eval=show_pca_section && show_pca_topvar, echo=FALSE, results='asis'}
# By Top Variable Proteins (PCA): generated by the pipeline, hidden from
# report by default. Set report_cfg$show_pca_topvar = TRUE to include.
pca_files <- list.files(diag_dir, pattern = "^PCA_top[0-9]+\\.png$", full.names = TRUE)

pca_data_map <- list()
available_n_top <- c()

if (length(pca_files) > 0) {
    for (f in pca_files) {
        n_val <- as.numeric(sub(".*PCA_top([0-9]+)\\.png$", "\\1", basename(f)))
        if (file.exists(f)) {
            pca_data_map[[as.character(n_val)]] <- knitr::image_uri(f)
            available_n_top <- c(available_n_top, n_val)
        }
    }
}

if (length(available_n_top) > 0) {
    available_n_top <- sort(unique(available_n_top))
    default_n_top <- available_n_top[1]
    default_img_uri <- pca_data_map[[as.character(default_n_top)]]
} else {
    default_pca_path <- file.path(diag_dir, "PCA_PC1.vs.PC2.png")
    if (file.exists(default_pca_path)) {
        default_n_top <- 0
        default_img_uri <- knitr::image_uri(default_pca_path)
        pca_data_map[["0"]] <- default_img_uri
        available_n_top <- 0
    } else {
        default_n_top <- NULL
        default_img_uri <- NULL
    }
}

cat("### By Top Variable Proteins\n\nSelect the number of top variable proteins used for PCA calculation:\n\n")
cat('<div style="margin-bottom: 20px;">\n')
cat('<label for="pca_prot_selector"><strong>Number of top variable proteins:</strong></label>\n')
cat('<select id="pca_prot_selector" onchange="updateProtPCAPlot(this.value)" style="padding: 8px; border-radius: 4px; border: 1px solid #ccc; font-size: 14px;">\n')
if (length(available_n_top) > 0 && available_n_top[1] != 0) {
    for (n_val in available_n_top) {
        selected <- if (n_val == default_n_top) " selected" else ""
        cat(sprintf('<option value="%d"%s>%s proteins</option>\n', n_val, selected, format(n_val, big.mark = ",")))
    }
} else if (!is.null(default_img_uri)) {
    cat('<option value="0" selected>All proteins (default)</option>\n')
} else {
    cat('<option disabled>No PCA plots found</option>\n')
}
cat('</select>\n</div>\n')

cat('<div id="pca_prot_container" style="text-align: center;">\n')
if (!is.null(default_img_uri)) {
    cat(sprintf('<img id="pca_prot_image" src="%s" style="width:90%%; max-width: 900px; height: auto;" alt="PCA Plot" />\n', default_img_uri))
} else {
    cat('<p><em>PCA plot not available.</em></p>\n')
}
cat('</div>\n')

cat('<script>\n')
cat(sprintf('var embeddedProtPcaImages = %s;\n',
            if (length(pca_data_map) > 0) jsonlite::toJSON(pca_data_map, auto_unbox = TRUE) else "{}"))
cat('function updateProtPCAPlot(n_top) {\n')
cat('    var img = document.getElementById("pca_prot_image");\n')
cat('    if (embeddedProtPcaImages && embeddedProtPcaImages[n_top]) {\n')
cat('        img.src = embeddedProtPcaImages[n_top];\n')
cat('    }\n')
cat('}\n')
cat('</script>\n')
```

```{r pca-13-heading, eval=show_pca_section, results='asis'}
cat("### PC1 vs PC3 {.tabset}\n")
cat("#### Static\n")
```

```{r pca-13, eval=show_pca_section, fig.cap="PCA -- PC1 vs PC3"}
p <- file.path(diag_dir, "PCA_PC1.vs.PC3.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-pca-13, eval=show_pca_section, results='asis'}
figure_legend("PCA plot showing sample separation along PC1 and PC3. PC3 may capture additional biological variation or batch effects not visible in the PC1-PC2 projection.")
```

```{r pca-13-labeled-heading, eval=show_pca_section, results='asis'}
cat("#### With Sample Names\n")
```

```{r pca-13-labeled, eval=show_pca_section, fig.cap="PCA -- PC1 vs PC3 (labeled)"}
p_lab <- file.path(diag_dir, "PCA_PC1.vs.PC3_labeled.png")
if (file.exists(p_lab)) knitr::include_graphics(p_lab) else cat("*Labeled PCA plot not available.*\n")
```

```{r pca-13-interactive-heading, eval=show_pca_section, results='asis'}
cat("#### Interactive\n")
```

```{r pca-13-interactive, eval=show_pca_section, fig.width=8, fig.height=6}
scores_csv <- file.path(diag_dir, "PCA_PC1.vs.PC3_scores.csv")
if (file.exists(scores_csv) && requireNamespace("plotly", quietly = TRUE)) {
    sc <- read.csv(scores_csv, stringsAsFactors = FALSE)
    color_var <- prot_cfg$effects$color %||% "Condition"
    if (!color_var %in% colnames(sc)) color_var <- setdiff(colnames(sc), c("sample", grep("^PC", colnames(sc), value = TRUE)))[1]
    sc$hover_text <- paste0("Sample: ", sc$sample, "<br>", color_var, ": ", sc[[color_var]])
    pl <- plotly::plot_ly(sc, x = ~PC1, y = ~PC3, color = as.formula(paste0("~`", color_var, "`")),
                          text = ~hover_text, hoverinfo = "text",
                          type = "scatter", mode = "markers",
                          marker = list(size = 10)) |>
        plotly::layout(xaxis = list(title = "PC1"),
                       yaxis = list(title = "PC3"),
                       title = "PCA: PC1 vs PC3 — Hover for sample names")
    pl
} else {
    cat("*Interactive PCA not available (scores CSV or plotly package missing).*\n")
}
```

```{r pca-3d-heading, eval=show_pca_section && show_pca_3d, results='asis'}
cat("### 3D Interactive PCA\n")
```

```{r pca-3d-link, eval=show_pca_section && show_pca_3d, results='asis'}
pca3d_file <- file.path(diag_dir, "PCA_3D.html")
if (file.exists(pca3d_file)) {
    cat('<div style="margin: 20px 0; padding: 20px 24px; border-left: 4px solid #0969da; background: #ddf4ff; border-radius: 6px;">\n')
    cat('<p style="color: #0550ae; font-weight: 600; font-size: 1.05em; margin-top: 0; margin-bottom: 8px;">Explore sample separation in three dimensions.</p>\n')
    cat('<p style="color: #0a3069; font-size: 0.95em; margin-bottom: 14px;">The interactive 3D PCA plot shows sample separation along PC1, PC2, and PC3. Rotate, zoom, and hover over points to explore clustering structure beyond what a 2D PCA can reveal.</p>\n')
    cat('<a href="proteomics/Diagnostic_plots/PCA_3D.html" target="_blank" style="display: inline-block; padding: 10px 20px; background: #0969da; color: #fff; font-weight: 600; text-decoration: none; border-radius: 6px; font-size: 1em;">Open Interactive 3D PCA &rarr;</a>\n')
    cat('</div>\n')
} else {
    cat("*3D PCA plot not available.*")
}
```

```{r umap-heading, eval=show_pca_section && show_umap, results='asis'}
cat("### UMAP\n")
```

```{r umap, eval=show_pca_section && show_umap, fig.cap="UMAP embedding of protein expression"}
umap_file <- file.path(diag_dir, "UMAP.png")
if (file.exists(umap_file)) {
    knitr::include_graphics(umap_file)
} else {
    cat("*UMAP plot not available. Enable with `qc.run_umap: true` in config.*\n")
}
```

```{r legend-umap, eval=show_pca_section && show_umap, results='asis'}
if (file.exists(file.path(diag_dir, "UMAP.png"))) {
    figure_legend("Uniform Manifold Approximation and Projection (UMAP) of imputed protein expression data. UMAP preserves local structure and may reveal non-linear groupings not apparent in PCA.")
}
```

```{r commentary-umap, eval=show_pca_section && show_umap, results='asis'}
render_commentary("umap_plot")
```

```{r pca-subsets, eval=show_pca_section && show_pca_subsets, results='asis'}
subset_pngs <- list.files(diag_dir, pattern = "^PCA_.*\\.png$", full.names = TRUE)
standard_pat <- "^PCA_(PC1\\.vs\\.PC[23](_labeled)?|top[0-9]+(_labeled)?)\\.png$"
subset_pngs <- subset_pngs[!grepl(standard_pat, basename(subset_pngs))]
if (length(subset_pngs) > 0) {
    cat("### Subsets\n\n")
    cat("PCA computed on user-defined subsets of samples (configured via the wizard's *PCA Subsets* section). Each plot below uses PC1 vs PC2 restricted to the listed conditions.\n\n")
    for (f in subset_pngs) {
        subset_name <- sub("^PCA_(.*)\\.png$", "\\1", basename(f))
        img_uri <- knitr::image_uri(f)
        cat(sprintf("#### %s\n\n", subset_name))
        cat(sprintf('<img src="%s" style="width:90%%; max-width: 900px; height: auto;" alt="PCA subset: %s" />\n\n', img_uri, subset_name))
    }
}
```

```{r imputation-check, include=FALSE}
imp_method <- prot_cfg$imputation$method %||% "none"
if (imp_method == "perseus") imp_method <- "perseus_like"
has_imputation <- tolower(imp_method) != "none" && isTRUE(show_imputation_qc)
```

```{r distribution-heading, eval=show_density, results='asis'}
cat("## Protein Distribution & Imputation QC {.tabset}\n\n")
cat("Density and boxplot views of per-sample protein expression after the same preprocessing (normalization + imputation if enabled). When imputation is enabled, additional tabs show observed-vs-imputed distributions.\n\n")
```

```{r density-tab-heading, eval=show_density, results='asis'}
cat("### Density\n")
```

```{r density, eval=show_density, fig.cap="Protein expression density"}
hp <- file.path(diag_dir, "protein_histograms_summary.png")
if (file.exists(hp)) knitr::include_graphics(hp)
```

```{r legend-density, eval=show_density, results='asis'}
figure_legend("Kernel density estimates of per-sample protein expression distributions. Overlapping curves indicate consistent expression profiles across samples.")
```

```{r commentary-density, eval=show_density, results='asis'}
render_commentary("density_plot")
```

```{r boxplot-tab-heading, eval=show_density && !has_imputation, results='asis'}
cat("### Boxplot\n")
```

```{r distribution-boxplot, eval=show_density && !has_imputation, fig.cap="Protein expression boxplot"}
bp <- file.path(diag_dir, "imputed_boxplot.png")
if (file.exists(bp)) {
    knitr::include_graphics(bp)
} else {
    cat("*Boxplot not available.*")
}
```

```{r legend-distribution-boxplot, eval=show_density && !has_imputation, results='asis'}
figure_legend("Per-sample boxplots of protein expression after normalization. Boxes show the interquartile range. Aligned medians and IQR widths across samples indicate successful normalization.")
```

```{r imp-hist-tab-heading, eval=has_imputation, results='asis'}
cat("### Imputation Histograms\n")
cat(sprintf("Imputation method: **%s**.\n\n", imp_method))
if (tolower(imp_method) == "perseus_like") {
    imp_w <- prot_cfg$imputation$width %||% 0.3
    imp_s <- prot_cfg$imputation$downshift %||% 1.8
    cat(sprintf("Imputed values drawn from a downshifted normal distribution (downshift = %s SD, width = %s SD).\n\n",
                imp_s, imp_w))
}
```

```{r imp-hist, eval=has_imputation, fig.cap="Imputed vs original intensity distributions"}
imp_w <- prot_cfg$imputation$width %||% "NA"
imp_s <- prot_cfg$imputation$downshift %||% "NA"
f_imp <- file.path(diag_dir, sprintf("imputed_histograms_samples_summary_w%s_s%s.png", imp_w, imp_s))
if (file.exists(f_imp)) knitr::include_graphics(f_imp)
```

```{r legend-imp-hist, eval=has_imputation, results='asis'}
figure_legend("Histograms comparing the distribution of observed (blue) and imputed (red) protein intensities per sample. Imputed values should appear as a low-abundance shoulder distinct from the observed distribution.")
```

```{r commentary-imp-hist, eval=has_imputation, results='asis'}
render_commentary("imputation_histogram")
```

```{r imp-box-tab-heading, eval=has_imputation, results='asis'}
cat("### Boxplot After Imputation\n")
```

```{r imp-boxplot, eval=has_imputation, fig.cap="Expression boxplot after imputation"}
f_box <- file.path(diag_dir, "imputed_boxplot.png")
if (file.exists(f_box)) knitr::include_graphics(f_box)
```

```{r legend-imp-boxplot, eval=has_imputation, results='asis'}
figure_legend("Boxplot of protein expression values after imputation. Boxes show the interquartile range (IQR) for each sample. Consistent medians and IQR widths across samples indicate successful normalization.")
```

```{r commentary-boxplot-post, results='asis'}
render_commentary("imputation_boxplot")
```

```{r similarity-heading, eval=show_distance || show_correlation, results='asis'}
cat("## Sample Similarity {.tabset}\n\n")
cat("Pairwise sample correlation and distance metrics assess overall reproducibility within and between experimental groups. High within-group correlation and clear between-group separation indicate good data quality and a detectable biological signal.\n\n")
```

```{r dist-heading, eval=show_distance, results='asis'}
cat("### Distance Heatmap (Imputed)\n")
```

```{r dist-heatmap, eval=show_distance, fig.cap="Sample-to-sample distance heatmap (imputed data)"}
p <- file.path(diag_dir, "sample_distance_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-dist-heatmap, eval=show_distance, results='asis'}
figure_legend("Euclidean distance matrix of imputed protein expression values. Darker colors indicate larger distances (less similar samples). Samples from the same experimental group should cluster together with smaller (lighter) inter-sample distances.")
```

```{r commentary-distance, eval=show_distance, results='asis'}
render_commentary("sample_distance")
```

```{r dist-na-heading, eval=show_distance, results='asis'}
cat("### Distance Heatmap (Raw, with NAs)\n")
```

```{r dist-heatmap-na, eval=show_distance, fig.cap="Sample-to-sample distance heatmap (raw data, NAs present)"}
p <- file.path(diag_dir, "sample_distance_heatmap_NA.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r corr-heading, eval=show_correlation, results='asis'}
cat("### Correlation Heatmap\n")
```

```{r corr-heatmap, eval=show_correlation, fig.cap="Sample-to-sample correlation heatmap"}
p <- file.path(diag_dir, "sample_correlation_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-corr-heatmap, eval=show_correlation, results='asis'}
figure_legend("Pairwise Pearson correlation heatmap of protein expression profiles. Within-group correlations should be higher than between-group correlations.")
```

```{r commentary-correlation, eval=show_correlation, results='asis'}
render_commentary("sample_correlation")
```

```{r expr-heatmap-heading, eval=show_expression_heatmap, results='asis'}
cat("## Expression Heatmap {.tabset}\n\n")
cat("Both views below cluster proteins (rows) by expression similarity. They differ only in whether the samples (columns) are also clustered — comparing them helps reveal whether sample clustering meaningfully reorganises the experimental groups.\n\n")
```

```{r expr-heatmap-uncl-heading, eval=show_expression_heatmap, results='asis'}
cat("### Without Samples Clustering\n")
```

```{r expr-heatmap-uncl, eval=show_expression_heatmap, fig.cap="Sample expression heatmap (rows clustered, sample order preserved)"}
p_wo_col <- file.path(diag_dir, "samples_protein_heatmap_wo_col.png")
if (file.exists(p_wo_col)) {
    knitr::include_graphics(p_wo_col)
} else {
    cat("*Expression heatmap (no sample clustering) not available.*")
}
```

```{r legend-expr-heatmap-uncl, eval=show_expression_heatmap, results='asis'}
figure_legend("Expression heatmap with proteins (rows) clustered hierarchically; samples (columns) kept in their input order. Useful for inspecting sample-by-sample variation without imposing a sample dendrogram.")
```

```{r expr-heatmap-cl-heading, eval=show_expression_heatmap, results='asis'}
cat("### With Samples Clustering\n")
```

```{r expr-heatmap-cl, eval=show_expression_heatmap, fig.cap="Sample expression heatmap (rows + samples clustered)"}
p <- file.path(diag_dir, "samples_protein_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-expr-heatmap-cl, eval=show_expression_heatmap, results='asis'}
figure_legend("Unsupervised hierarchical clustering heatmap based on the top most variable proteins. Both rows (proteins) and columns (samples) are clustered by expression similarity.")
```

```{r commentary-expr-heatmap, eval=show_expression_heatmap, results='asis'}
render_commentary("expr_heatmap")
```

---

# Differential Expression Analysis

```{r cutoff-panel, results='asis'}
if (exists("cutoff_panel_html")) {
    .lfc_default <- log2(prot_cfg$de$linear_fc_cutoff %||% 1.5)
    .p_default   <- prot_cfg$de$p_cutoff %||% 0.05
    cat(cutoff_panel_html(.lfc_default, .p_default, default_ptype = "padj"))
}
```

```{r de-stats-description, results='asis'}
de_method_name <- prot_cfg$de$method %||% "limma"
padj_cut_de <- prot_cfg$de$p_cutoff %||% 0.05
lfc_cut_de <- prot_cfg$de$linear_fc_cutoff %||% 1.5

# Custom stats text takes priority; otherwise auto-generate from config
custom_stats_text <- report_cfg$stats_methods_text %||% ""
if (nzchar(trimws(custom_stats_text))) {
    cat(trimws(custom_stats_text), "\n\n")
} else {
    # Auto-generate description
    test_desc <- switch(tolower(de_method_name),
        limma = "moderated t-tests via the *limma* R/Bioconductor package (empirical Bayes shrinkage of variance estimates)",
        "limma-voom" = "limma-voom (log-CPM transformed, precision-weighted linear models)",
        deseq2 = "Wald tests via *DESeq2* (negative binomial generalized linear models)",
        paste0("*", de_method_name, "* method")
    )
    cat(sprintf(
        "Differential expression was assessed using %s. Multiple testing correction was applied using the **Benjamini-Hochberg** procedure (FDR). Proteins were considered differentially expressed at adjusted p-value $\\leq$ %s and |linear fold change| $\\geq$ %s.\n\n",
        test_desc, padj_cut_de, lfc_cut_de
    ))
}

# Link to full results table
full_results_link <- report_cfg$full_results_table_link %||% ""
if (nzchar(trimws(full_results_link))) {
    if (grepl("^https?://", full_results_link)) {
        cat(sprintf("Full differential results table: [download](%s)\n\n", full_results_link))
    } else {
        cat(sprintf("Full differential results table: `%s`\n\n", full_results_link))
    }
}
```

## DE Summary

Differential expression analysis identifies features with statistically significant abundance changes between experimental conditions. Results are corrected for multiple testing to control the false discovery rate.

```{r de-summary, results='asis'}
if (!is.null(de_data) && length(contrast_names) > 0) {
    padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- prot_cfg$de$linear_fc_cutoff %||% 1.5

    total_proteins <- nrow(de_data)

    de_counts_list <- list()
    for (cn in contrast_names) {
        pass_col <- paste0("pass.imputs.", cn)
        lfc_col <- paste0("linearFC.imputs.", cn)
        padj_col <- paste0("padj.imputs.", cn)

        if (pass_col %in% names(de_data) && lfc_col %in% names(de_data)) {
            # Imputation pipeline: use consensus pass column for consistency
            # with executive summary (pass_any_contrast) and Excel exports
            is_sig <- !is.na(de_data[[pass_col]]) & de_data[[pass_col]] == 1
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            n_up <- sum(is_sig & lfc_vals > 0, na.rm = TRUE)
            n_down <- sum(is_sig & lfc_vals < 0, na.rm = TRUE)
        } else if (padj_col %in% names(de_data) && lfc_col %in% names(de_data)) {
            # Non-imputation fallback: use padj + linearFC thresholds directly
            padj_vals <- as.numeric(de_data[[padj_col]])
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            log2fc <- signed_fc_to_log2(lfc_vals)
            is_sig <- !is.na(padj_vals) & padj_vals <= padj_cut & abs(log2fc) >= log2(lfc_cut)
            n_up <- sum(is_sig & log2fc > 0, na.rm = TRUE)
            n_down <- sum(is_sig & log2fc < 0, na.rm = TRUE)
        } else {
            next
        }
        de_counts_list[[cn]] <- data.frame(
            Contrast = cn, Up = n_up, Down = n_down, Total = sum(is_sig),
            stringsAsFactors = FALSE
        )
    }

    if (length(de_counts_list) > 0) {
        de_counts_df <- do.call(rbind, de_counts_list)
        cat(sprintf("**Total proteins quantified:** %s\n\n", format(total_proteins, big.mark = ",")))
        if (exists("cutoff_summary_span")) {
            cat("<div class='cutoff-de-summary' style='margin-bottom:12px;'>\n")
            for (row_i in seq_len(nrow(de_counts_df))) {
                cat(sprintf("<div style='margin:3px 0;'><strong>%s:</strong> %s</div>\n",
                    de_counts_df$Contrast[row_i],
                    cutoff_summary_span(de_counts_df$Contrast[row_i],
                                        de_counts_df$Up[row_i],
                                        de_counts_df$Down[row_i], "proteins")))
            }
            cat("</div>\n")
        }
        make_dt(de_counts_df, "DE Protein Counts per Contrast")
    }
} else {
    cat("DE summary data not available.\n")
}
```

```{r de-overlap-cond, include=FALSE}
.show_overlap <- !is.null(de_data) && length(contrast_names) >= 2
```

`r if (!.show_overlap) "<!--"`

## DE Protein Overlap {.tabset}

### All DE Proteins

```{r de-overlap-venn-all, fig.height=6, fig.width=10, results='asis'}
if (!is.null(de_data) && length(contrast_names) >= 2) {
    padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- log2(prot_cfg$de$linear_fc_cutoff %||% 1.5)
    id_col <- prot_cfg$de_table$id_col %||% "FeatureID"

    de_protein_lists <- list()
    for (cn in contrast_names) {
        padj_col <- paste0("padj.imputs.", cn)
        lfc_col <- paste0("linearFC.imputs.", cn)
        if (padj_col %in% names(de_data) && lfc_col %in% names(de_data)) {
            padj_vals <- as.numeric(de_data[[padj_col]])
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            log2fc <- signed_fc_to_log2(lfc_vals)
            is_de <- !is.na(padj_vals) & padj_vals <= padj_cut & abs(log2fc) >= lfc_cut
            protein_col <- if (id_col %in% names(de_data)) id_col else names(de_data)[1]
            de_protein_lists[[gsub("_", " ", cn)]] <- de_data[[protein_col]][is_de]
        }
    }

    if (length(de_protein_lists) >= 2 && length(de_protein_lists) <= 7) {
        if (has_ggVennDiagram) {
            total_unique <- length(unique(unlist(de_protein_lists)))
            short_names <- gsub(" vs ", "\nvs ", names(de_protein_lists))
            names(de_protein_lists) <- short_names
            venn_plot <- ggVennDiagram(de_protein_lists, label_alpha = 0,
                                       label = "both", label_percent_digit = 1,
                                       set_size = 2.8) +
                scale_fill_gradient(low = "white", high = "steelblue") +
                scale_x_continuous(expand = expansion(mult = 0.4)) +
                labs(title = "DE Protein Overlap (All)",
                     subtitle = paste0("Total unique DE proteins: ", total_unique)) +
                theme(legend.position = "none",
                      plot.title = element_text(hjust = 0),
                      plot.subtitle = element_text(hjust = 0),
                      plot.margin = margin(10, 60, 10, 60))
            print(venn_plot)
        } else {
            cat("**DE protein counts per contrast:**\n\n")
            for (nm in names(de_protein_lists)) {
                cat(sprintf("- **%s**: %d proteins\n", nm, length(de_protein_lists[[nm]])))
            }
        }
    }
} else if (length(contrast_names) < 2) {
    cat("Venn diagram requires at least 2 contrasts.\n")
} else {
    cat("DE data not available for Venn diagram.\n")
}
```

### Up-regulated

```{r de-overlap-venn-up, fig.height=6, fig.width=10, results='asis'}
if (!is.null(de_data) && length(contrast_names) >= 2) {
    padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- log2(prot_cfg$de$linear_fc_cutoff %||% 1.5)
    id_col <- prot_cfg$de_table$id_col %||% "FeatureID"

    up_protein_lists <- list()
    for (cn in contrast_names) {
        padj_col <- paste0("padj.imputs.", cn)
        lfc_col <- paste0("linearFC.imputs.", cn)
        if (padj_col %in% names(de_data) && lfc_col %in% names(de_data)) {
            padj_vals <- as.numeric(de_data[[padj_col]])
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            log2fc <- signed_fc_to_log2(lfc_vals)
            is_up <- !is.na(padj_vals) & padj_vals <= padj_cut & log2fc >= lfc_cut
            protein_col <- if (id_col %in% names(de_data)) id_col else names(de_data)[1]
            up_protein_lists[[gsub("_", " ", cn)]] <- de_data[[protein_col]][is_up]
        }
    }

    if (length(up_protein_lists) >= 2 && length(up_protein_lists) <= 7 && has_ggVennDiagram) {
        total_up <- length(unique(unlist(up_protein_lists)))
        short_names_up <- gsub(" vs ", "\nvs ", names(up_protein_lists))
        names(up_protein_lists) <- short_names_up
        venn_plot <- ggVennDiagram(up_protein_lists, label_alpha = 0,
                                   label = "both", label_percent_digit = 1,
                                   set_size = 2.8) +
            scale_fill_gradient(low = "white", high = "#E41A1C") +
            scale_x_continuous(expand = expansion(mult = 0.4)) +
            labs(title = "Up-regulated Protein Overlap",
                 subtitle = paste0("Total unique up-regulated: ", total_up)) +
            theme(legend.position = "none",
                  plot.title = element_text(hjust = 0),
                  plot.subtitle = element_text(hjust = 0),
                  plot.margin = margin(10, 60, 10, 60))
        print(venn_plot)
    }
}
```

### Down-regulated

```{r de-overlap-venn-down, fig.height=6, fig.width=10, results='asis'}
if (!is.null(de_data) && length(contrast_names) >= 2) {
    padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- log2(prot_cfg$de$linear_fc_cutoff %||% 1.5)
    id_col <- prot_cfg$de_table$id_col %||% "FeatureID"

    down_protein_lists <- list()
    for (cn in contrast_names) {
        padj_col <- paste0("padj.imputs.", cn)
        lfc_col <- paste0("linearFC.imputs.", cn)
        if (padj_col %in% names(de_data) && lfc_col %in% names(de_data)) {
            padj_vals <- as.numeric(de_data[[padj_col]])
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            log2fc <- signed_fc_to_log2(lfc_vals)
            is_down <- !is.na(padj_vals) & padj_vals <= padj_cut & log2fc <= -lfc_cut
            protein_col <- if (id_col %in% names(de_data)) id_col else names(de_data)[1]
            down_protein_lists[[gsub("_", " ", cn)]] <- de_data[[protein_col]][is_down]
        }
    }

    if (length(down_protein_lists) >= 2 && length(down_protein_lists) <= 7 && has_ggVennDiagram) {
        total_down <- length(unique(unlist(down_protein_lists)))
        short_names_dn <- gsub(" vs ", "\nvs ", names(down_protein_lists))
        names(down_protein_lists) <- short_names_dn
        venn_plot <- ggVennDiagram(down_protein_lists, label_alpha = 0,
                                   label = "both", label_percent_digit = 1,
                                   set_size = 2.8) +
            scale_fill_gradient(low = "white", high = "#377EB8") +
            scale_x_continuous(expand = expansion(mult = 0.4)) +
            labs(title = "Down-regulated Protein Overlap",
                 subtitle = paste0("Total unique down-regulated: ", total_down)) +
            theme(legend.position = "none",
                  plot.title = element_text(hjust = 0),
                  plot.subtitle = element_text(hjust = 0),
                  plot.margin = margin(10, 60, 10, 60))
        print(venn_plot)
    }
}
```

`r if (!.show_overlap) "-->"`

```{r identified-proteins-venn, results='asis', fig.height=6, fig.width=10}
# Venn diagram of all identified proteins per sample/condition (pre-imputation)
show_id_venn <- isTRUE(report_cfg$show_identified_proteins_venn)

if (show_id_venn && has_ggVennDiagram) {
    cat("## Identified Proteins (Pre-imputation) Venn\n\n")
    cat("This Venn diagram shows proteins *identified* (non-missing) in each condition,\n")
    cat("**before** imputation of missing values.\n\n")

    # Load raw (pre-imputation) expression data — prefer unimputed file, exclude imputed
    raw_expr_file <- list.files(datasets_dir, pattern = "unimputed.*\\.(tsv|csv)$",
                                 full.names = TRUE)[1]
    if (is.na(raw_expr_file)) {
        raw_expr_file <- list.files(datasets_dir, pattern = "(raw_|pre_imputation).*\\.(tsv|csv)$",
                                     full.names = TRUE)[1]
    }
    if (!is.na(raw_expr_file) && file.exists(raw_expr_file)) {
        raw_df <- tryCatch({
            if (grepl("\\.csv$", raw_expr_file)) read.csv(raw_expr_file, stringsAsFactors = FALSE, check.names = FALSE)
            else read.delim(raw_expr_file, stringsAsFactors = FALSE, check.names = FALSE)
        }, error = function(e) NULL)

        if (!is.null(raw_df) && exists("meta") && is.data.frame(meta)) {
            id_col_v <- prot_cfg$de_table$id_col %||% "FeatureID"
            # Fallback: raw/unimputed file may use original protein_id column, not FeatureID
            if (!(id_col_v %in% names(raw_df))) {
                id_col_v <- prot_cfg$id_columns$protein_id %||% "Protein.Group"
            }
            sample_col_v <- prot_cfg$effects$samples %||% "SampleID"
            group_col_v <- prot_cfg$effects$color %||% "Condition"

            if (id_col_v %in% names(raw_df) && group_col_v %in% names(meta) && sample_col_v %in% names(meta)) {
                sample_ids <- meta[[sample_col_v]]
                groups <- meta[[group_col_v]]
                expr_cols <- intersect(sample_ids, names(raw_df))

                if (length(expr_cols) >= 2) {
                    group_map <- setNames(groups, sample_ids)
                    unique_groups <- unique(groups)

                    # For each group, list proteins detected in at least one sample
                    id_lists <- lapply(setNames(unique_groups, unique_groups), function(g) {
                        g_samples <- names(group_map)[group_map == g]
                        g_samples <- intersect(g_samples, names(raw_df))
                        if (length(g_samples) == 0) return(character(0))
                        detected <- apply(raw_df[, g_samples, drop = FALSE], 1, function(row) any(!is.na(row) & row != 0))
                        raw_df[[id_col_v]][detected]
                    })

                    if (length(id_lists) >= 2 && length(id_lists) <= 7) {
                        names(id_lists) <- gsub("_", " ", names(id_lists))
                        total <- length(unique(unlist(id_lists)))
                        venn_plot <- ggVennDiagram(id_lists, label_alpha = 0,
                                                   label = "both", label_percent_digit = 1,
                                                   set_size = 2.8) +
                            scale_fill_gradient(low = "white", high = "#2ca25f") +
                            scale_x_continuous(expand = expansion(mult = 0.4)) +
                            labs(title = "All Identified Proteins per Condition (Pre-imputation)",
                                 subtitle = paste0("Total unique identified: ", total)) +
                            theme(legend.position = "none",
                                  plot.title = element_text(hjust = 0),
                                  plot.subtitle = element_text(hjust = 0),
                                  plot.margin = margin(10, 60, 10, 60))
                        print(venn_plot)
                        figure_legend("Venn diagram of all identified (non-missing) proteins per experimental condition, before imputation.")
                    }
                }
            }
        }
    } else {
        cat("*Pre-imputation data not available for identified proteins Venn diagram.*\n\n")
    }
}
```

## DE Results Table

```{r de-table, results='asis'}
if (!is.null(de_data)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Full differential expression results merged with imputed expression values. Columns include per-contrast adjusted p-values, fold changes, significance flags, and per-sample expression. Use the search box to filter and the buttons to export.</em></p>\n")
    show_df <- de_data

    # Merge imputed expression into DE results
    imp_files_de <- list.files(datasets_dir, pattern = "imputed.*\\.tsv$", full.names = TRUE)
    if (length(imp_files_de) > 0) {
        imp_expr <- tryCatch(
            read.delim(imp_files_de[1], stringsAsFactors = FALSE, check.names = FALSE),
            error = function(e) NULL
        )
        if (!is.null(imp_expr)) {
            imp_id_col <- names(imp_expr)[1]
            de_id_col <- prot_cfg$de_table$id_col %||% "FeatureID"
            if (de_id_col %in% names(show_df)) {
                merged_de <- merge(show_df, imp_expr, by.x = de_id_col, by.y = imp_id_col, all.x = TRUE)
                if (nrow(merged_de) > 0) show_df <- merged_de
            }
        }
    }

    make_dt(show_df, caption = "DE Results + Expression", pageLength = 15)
}
```

## Volcano Plots {.tabset}

### Static {.tabset}

```{r volcano-plots-static, results='asis'}
# Per-contrast static volcano: show padj + raw-pval variants as sub-tabs.
# Default tab matches the cutoff-panel default (`default_ptype` in cutoff_panel_html).
volcano_files <- list.files(de_png_dir, pattern = "^volcano_.*_padj\\.png$",
                            full.names = TRUE)
padj_cut_vol <- prot_cfg$de$p_cutoff %||% 0.05
lfc_cut_vol  <- prot_cfg$de$linear_fc_cutoff %||% 1.5
default_ptype_vol <- "padj"  # matches cutoff_panel_html(default_ptype = "padj") above
if (length(volcano_files) > 0) {
    for (vf_padj in volcano_files) {
        suffix <- gsub("^volcano_|_padj\\.png$", "", basename(vf_padj))
        # suffix may include "<contrast>_<de_source>" for proteomics (e.g. "A_vs_B_summary")
        contrast_clean <- gsub("_", " ", suffix)
        vf_pval <- sub("_padj\\.png$", "_pval.png", vf_padj)
        cat(sprintf("\n#### %s {.tabset}\n\n", contrast_clean))

        # Order tabs so the configured default p-value type renders first
        order_ptypes <- if (default_ptype_vol == "padj") c("padj", "pval") else c("pval", "padj")
        for (pt in order_ptypes) {
            tab_label <- if (pt == "padj") "Adjusted P-value" else "Raw P-value"
            vf <- if (pt == "padj") vf_padj else vf_pval
            if (file.exists(vf)) {
                cat(sprintf("\n##### %s\n\n", tab_label))
                cat(sprintf("![Volcano -- %s -- %s](%s)\n\n",
                            contrast_clean, tab_label, vf))
                figure_legend(sprintf(
                    "Volcano plot for the %s contrast (y-axis: -log10 %s). Red: significantly up-regulated; blue: significantly down-regulated (cutoff p &le; %s, |FC| &ge; %s).",
                    contrast_clean,
                    if (pt == "padj") "adjusted P-value" else "raw P-value",
                    padj_cut_vol, lfc_cut_vol))
            }
        }
        render_commentary(paste0("volcano_", suffix))
    }
} else {
    # Backwards-compat: fall back to legacy single volcano if pipeline did not
    # yet produce padj/pval variants
    legacy_files <- list.files(de_png_dir, pattern = "^volcano_.*\\.png$",
                               full.names = TRUE)
    legacy_files <- legacy_files[!grepl("_(padj|pval)\\.png$", legacy_files)]
    if (length(legacy_files) > 0) {
        for (vf in legacy_files) {
            cn <- gsub("^volcano_|\\.png$", "", basename(vf))
            contrast_clean <- gsub("_", " ", cn)
            cat(sprintf("\n#### %s\n\n", contrast_clean))
            cat(sprintf("![Volcano -- %s](%s)\n\n", contrast_clean, vf))
            figure_legend(sprintf(
                "Volcano plot for the %s contrast. Red: significantly up-regulated; blue: significantly down-regulated (adj. p &le; %s, |FC| &ge; %s).",
                contrast_clean, padj_cut_vol, lfc_cut_vol))
            render_commentary(paste0("volcano_", cn))
        }
    } else {
        cat("No volcano plots found.\n")
    }
}
```

### Interactive {.tabset}

```{r volcano-interactive-setup, include=FALSE}
# Create interactive volcano plot
make_volcano_plotly_prot <- function(df, contrast, padj_cut = 0.05, lfc_cut = log2(1.5), title = NULL) {
    padj_col <- paste0("padj.imputs.", contrast)
    lfc_col <- paste0("linearFC.imputs.", contrast)
    if (!(padj_col %in% names(df)) || !(lfc_col %in% names(df))) return(NULL)

    id_col <- prot_cfg$de_table$id_col %||% "FeatureID"
    gene_col <- if ("Genes" %in% names(df)) "Genes" else if (id_col %in% names(df)) id_col else names(df)[1]
    # Pull supporting identifier columns if present so the tooltip can
    # carry gene symbol + accession + protein description together.
    accession_vals <- if (id_col %in% names(df)) df[[id_col]] else rep(NA_character_, nrow(df))
    desc_vals      <- if ("First.Protein.Description" %in% names(df)) {
        df[["First.Protein.Description"]]
    } else rep(NA_character_, nrow(df))

    plot_df <- data.frame(
        protein = df[[gene_col]],
        accession = accession_vals,
        description = desc_vals,
        logFC = signed_fc_to_log2(as.numeric(df[[lfc_col]])),
        padj = as.numeric(df[[padj_col]]),
        stringsAsFactors = FALSE
    )
    plot_df$neglog10padj <- -log10(pmax(plot_df$padj, 1e-300))
    plot_df$Direction <- ifelse(
        !is.na(plot_df$padj) & plot_df$padj <= padj_cut & abs(plot_df$logFC) >= lfc_cut,
        ifelse(plot_df$logFC > 0, "Up", "Down"), "NS"
    )
    plot_df$Direction <- factor(plot_df$Direction, levels = c("NS", "Down", "Up"))

    # Tooltip: gene symbol + UniProt accession + protein description
    # (skip lines that are NA / blank / equal to the gene symbol)
    tooltip_lines <- vapply(seq_len(nrow(plot_df)), function(i) {
        lines <- c(paste0("Gene: ", plot_df$protein[i]))
        if (!is.na(plot_df$accession[i]) && nzchar(plot_df$accession[i]) &&
            !identical(plot_df$accession[i], plot_df$protein[i])) {
            lines <- c(lines, paste0("Accession: ", plot_df$accession[i]))
        }
        if (!is.na(plot_df$description[i]) && nzchar(plot_df$description[i])) {
            lines <- c(lines, paste0("Description: ", plot_df$description[i]))
        }
        lines <- c(lines, paste0("log2FC: ", round(plot_df$logFC[i], 3)),
                          paste0("padj: ", signif(plot_df$padj[i], 3)))
        paste(lines, collapse = "\n")
    }, character(1))
    plot_df$tooltip <- tooltip_lines

    p <- ggplot(plot_df, aes(x = logFC, y = neglog10padj, color = Direction,
                              text = tooltip)) +
        geom_point(alpha = 0.6, size = 1.5) +
        scale_color_manual(values = c("NS" = "grey70", "Down" = "blue", "Up" = "red")) +
        geom_vline(xintercept = c(-lfc_cut, lfc_cut), linetype = "dashed") +
        geom_hline(yintercept = -log10(padj_cut), linetype = "dashed") +
        labs(title = title, x = "log2 Fold Change", y = "-log10(adj.P.Val)") +
        theme_minimal()
    widget <- ggplotly(p, tooltip = "text")
    # Return widget + registration data for cutoff panel
    # Include both raw p-value and adjusted p-value if available
    pval_col <- paste0("pvalue.imputs.", contrast)
    raw_pval <- if (pval_col %in% names(df)) as.numeric(df[[pval_col]]) else plot_df$padj
    # Imputation-consensus pass flag — ensures interactive counts match pipeline
    pass_col <- paste0("pass.imputs.", contrast)
    pass_flag <- if (pass_col %in% names(df)) as.numeric(df[[pass_col]]) else NULL
    reg_df <- data.frame(name = as.character(plot_df$protein),
                         logFC = plot_df$logFC,
                         pval = raw_pval,
                         adjPval = plot_df$padj,
                         stringsAsFactors = FALSE)
    if (!is.null(pass_flag)) reg_df$passFlag <- pass_flag
    list(widget = widget, plot_df = reg_df)
}
```

Volcano plots display the relationship between statistical significance (-log10 adjusted p-value) and biological effect size (log2 fold-change) for each feature. Points in the upper corners represent features that are both statistically significant and have large fold changes.

Hover over points to see protein names and statistics. Enter protein names to highlight them.

```{r volcano-interactive, results='asis', fig.height=6}
if (!is.null(de_data) && length(contrast_names) > 0) {
    for (i in seq_along(contrast_names)) {
        cn <- contrast_names[i]
        contrast_clean <- gsub("_", " ", cn)
        clean_id <- gsub("[^a-zA-Z0-9]", "_", cn)

        cat(sprintf("\n#### %s\n\n", contrast_clean))

        cat(sprintf('
<div style="margin-bottom: 15px; position: relative;">
<label for="volcano_genes_%s"><strong>Highlight proteins:</strong></label>
<div style="display: inline-block; position: relative; margin-left: 10px;">
<input type="text" id="volcano_genes_%s" placeholder="Start typing a protein name..." autocomplete="off"
  style="padding: 8px; border-radius: 4px; border: 1px solid #ccc; width: 350px;"
  oninput="volcanoAutocomplete_%s(this)" onkeydown="volcanoAutoKey_%s(event)">
<div id="volcano_ac_%s" style="position: absolute; top: 100%%; left: 0; width: 350px; max-height: 200px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 4px 4px; z-index: 1000; display: none; box-shadow: 0 4px 6px rgba(0,0,0,0.1);"></div>
</div>
<button onclick="highlightVolcanoGenes_%s()" style="padding: 8px 15px; margin-left: 5px; background: #007bff; color: white; border: none; border-radius: 4px; cursor: pointer;">Highlight</button>
<button onclick="clearVolcanoHighlight_%s()" style="padding: 8px 15px; margin-left: 5px; background: #6c757d; color: white; border: none; border-radius: 4px; cursor: pointer;">Clear</button>
<div id="volcano_tags_%s" style="margin-top: 5px; display: flex; flex-wrap: wrap; gap: 4px;"></div>
</div>
', clean_id, clean_id, clean_id, clean_id, clean_id, clean_id, clean_id, clean_id))

        vol_res <- make_volcano_plotly_prot(de_data, cn, prot_cfg$de$p_cutoff %||% 0.05,
                                      log2(prot_cfg$de$linear_fc_cutoff %||% 1.5), contrast_clean)
        if (!is.null(vol_res)) {
            p <- vol_res$widget
            p$elementId <- paste0("volcano_plot_", clean_id)
            print(htmltools::tagList(p))
            if (exists("cutoff_register_plot")) {
                cat(cutoff_register_plot(
                    paste0("volcano_plot_", clean_id),
                    vol_res$plot_df, "volcano", "Protein", cn
                ))
            }

            cat(sprintf('
<script>
(function() {
    var cid = "%s";
    var selectedGenes = [];
    var acIndex = -1;
    function getGeneNames() {
        var plotDiv = document.getElementById("volcano_plot_" + cid);
        if (!plotDiv || !plotDiv.data) return [];
        var names = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var data = plotDiv.data[t];
            if (!data.text) continue;
            for (var i = 0; i < data.text.length; i++) {
                var parts = data.text[i].split(/<br\\s*\\/?>/i);
                if (parts.length > 0) {
                    var g = parts[0].replace(/Protein:\\s*/i, "").trim();
                    if (g && names.indexOf(g) === -1) names.push(g);
                }
            }
        }
        return names.sort();
    }
    window.volcanoAutocomplete_%s = function(el) {
        var acDiv = document.getElementById("volcano_ac_" + cid);
        var val = el.value.trim().toLowerCase();
        acDiv.innerHTML = ""; acIndex = -1;
        if (val.length < 1) { acDiv.style.display = "none"; return; }
        var parts = val.split(","); var current = parts[parts.length - 1].trim();
        if (current.length < 1) { acDiv.style.display = "none"; return; }
        var allGenes = getGeneNames();
        var matches = allGenes.filter(function(g) { return g.toLowerCase().indexOf(current) !== -1 && selectedGenes.indexOf(g) === -1; }).slice(0, 15);
        if (matches.length === 0) { acDiv.style.display = "none"; return; }
        matches.forEach(function(g) {
            var item = document.createElement("div");
            item.textContent = g;
            item.style.cssText = "padding: 6px 10px; cursor: pointer; font-size: 13px;";
            item.onmouseenter = function() { this.style.background = "#e9ecef"; };
            item.onmouseleave = function() { this.style.background = "white"; };
            item.onclick = function() { selectGene(g); };
            acDiv.appendChild(item);
        });
        acDiv.style.display = "block";
    };
    window.volcanoAutoKey_%s = function(e) {
        var acDiv = document.getElementById("volcano_ac_" + cid);
        var items = acDiv.children;
        if (items.length === 0) return;
        if (e.key === "ArrowDown") { e.preventDefault(); acIndex = Math.min(acIndex + 1, items.length - 1); updateAcHighlight(items); }
        else if (e.key === "ArrowUp") { e.preventDefault(); acIndex = Math.max(acIndex - 1, 0); updateAcHighlight(items); }
        else if (e.key === "Enter" && acIndex >= 0) { e.preventDefault(); items[acIndex].click(); }
    };
    function updateAcHighlight(items) { for (var i = 0; i < items.length; i++) items[i].style.background = i === acIndex ? "#e9ecef" : "white"; }
    function selectGene(gene) {
        if (selectedGenes.indexOf(gene) === -1) selectedGenes.push(gene);
        document.getElementById("volcano_genes_" + cid).value = "";
        document.getElementById("volcano_ac_" + cid).style.display = "none";
        renderTags(); applyHighlight();
    }
    function renderTags() {
        var container = document.getElementById("volcano_tags_" + cid);
        container.innerHTML = "";
        selectedGenes.forEach(function(g) {
            var tag = document.createElement("span");
            tag.style.cssText = "display: inline-flex; align-items: center; background: #007bff; color: white; padding: 2px 8px; border-radius: 12px; font-size: 12px; gap: 4px;";
            tag.innerHTML = g + " <span style=\\"cursor:pointer; font-weight:bold; margin-left:3px;\\" onclick=\\"removeVolcanoGene_" + cid + "(\'" + g + "\')\\">x</span>";
            container.appendChild(tag);
        });
    }
    window["removeVolcanoGene_" + cid] = function(gene) { selectedGenes = selectedGenes.filter(function(g) { return g !== gene; }); renderTags(); applyHighlight(); };
    function applyHighlight() {
        var plotDiv = document.getElementById("volcano_plot_" + cid);
        if (!plotDiv || !plotDiv.data) return;
        if (selectedGenes.length === 0) { Plotly.relayout(plotDiv, { annotations: [] }); return; }
        var genesLower = selectedGenes.map(function(g) { return g.toLowerCase(); });
        var annotations = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var data = plotDiv.data[t];
            if (!data.text) continue;
            for (var i = 0; i < data.text.length; i++) {
                var parts = data.text[i].split(/<br\\s*\\/?>/i);
                if (parts.length > 0) {
                    var geneName = parts[0].replace(/Protein:\\s*/i, "").trim();
                    if (genesLower.indexOf(geneName.toLowerCase()) !== -1) {
                        annotations.push({ x: data.x[i], y: data.y[i], text: geneName, showarrow: true, arrowhead: 2, ax: 20, ay: -30, font: { size: 12, color: "black" }, bgcolor: "rgba(255,255,255,0.9)", bordercolor: "black", borderwidth: 1 });
                    }
                }
            }
        }
        Plotly.relayout(plotDiv, { annotations: annotations });
    }
    window.highlightVolcanoGenes_%s = function() {
        var input = document.getElementById("volcano_genes_" + cid).value;
        var manual = input.split(",").map(function(s) { return s.trim(); }).filter(function(s) { return s.length > 0; });
        var allGenes = getGeneNames();
        manual.forEach(function(m) { var mLower = m.toLowerCase(); var match = allGenes.find(function(g) { return g.toLowerCase() === mLower; }); if (match && selectedGenes.indexOf(match) === -1) selectedGenes.push(match); });
        document.getElementById("volcano_genes_" + cid).value = "";
        document.getElementById("volcano_ac_" + cid).style.display = "none";
        renderTags(); applyHighlight();
    };
    window.clearVolcanoHighlight_%s = function() {
        selectedGenes = [];
        document.getElementById("volcano_genes_" + cid).value = "";
        renderTags();
        var plotDiv = document.getElementById("volcano_plot_" + cid);
        if (plotDiv) Plotly.relayout(plotDiv, { annotations: [] });
    };
    document.addEventListener("click", function(e) {
        var acDiv = document.getElementById("volcano_ac_" + cid);
        var input = document.getElementById("volcano_genes_" + cid);
        if (acDiv && e.target !== input && !acDiv.contains(e.target)) acDiv.style.display = "none";
    });
})();
</script>
', clean_id, clean_id, clean_id, clean_id, clean_id))
        }
        cat("\n\n")
    }
} else {
    cat("Interactive plots require DE data.\n")
}
```

```{r ma-plots-heading, eval=show_ma_plots, results='asis'}
cat("## MA Plots {.tabset}\n\n### Static {.tabset}\n\n")
```

```{r ma-plots-static, eval=show_ma_plots, results='asis'}
ma_files <- list.files(de_png_dir, pattern = "^ma_.*\\.png$", full.names = TRUE)
if (length(ma_files) > 0) {
    for (mf in ma_files) {
        cn <- gsub("^ma_|\\.png$", "", basename(mf))
        contrast_clean <- gsub("_", " ", cn)
        cat(sprintf("\n#### %s\n\n", contrast_clean))
        cat(sprintf("![MA -- %s](%s)\n\n", contrast_clean, mf))
        figure_legend(sprintf("MA plot for the %s contrast. Red: significantly up-regulated; blue: significantly down-regulated.", contrast_clean))
        render_commentary(paste0("ma_", cn))
    }
} else {
    cat("No MA plots found.\n")
}
```

```{r ma-interactive-heading, eval=show_ma_plots, results='asis'}
cat("### Interactive {.tabset}\n\n")
cat("MA plots show the relationship between average expression (A, x-axis) and log-fold-change (M, y-axis). They help assess whether fold-change estimates are independent of expression level and whether the data is properly normalized.\n\n")
```

```{r ma-interactive-setup, eval=show_ma_plots, include=FALSE}
# Load expression data from final_results.tsv for MA plot (has per-sample columns)
ma_expr_df <- NULL
ma_final_file <- file.path(datasets_dir, "final_results.tsv")
if (file.exists(ma_final_file)) {
    ma_expr_df <- tryCatch(
        read.delim(ma_final_file, stringsAsFactors = FALSE, check.names = FALSE),
        error = function(e) NULL
    )
}

make_ma_plotly_prot <- function(df, expr_df, contrast, padj_cut = 0.05, lfc_cut = log2(1.5), title = NULL) {
    padj_col <- paste0("padj.imputs.", contrast)
    lfc_col <- paste0("linearFC.imputs.", contrast)
    if (!(padj_col %in% names(df)) || !(lfc_col %in% names(df))) return(NULL)

    id_col <- prot_cfg$de_table$id_col %||% "FeatureID"
    gene_col <- if ("Genes" %in% names(df)) "Genes" else if (id_col %in% names(df)) id_col else names(df)[1]

    # Get per-sample expression columns for baseMean calculation
    # Use expr_df (final_results.tsv) if available, otherwise try df itself
    src <- if (!is.null(expr_df) && id_col %in% names(expr_df)) expr_df else df

    # Primary: match columns to metadata sample IDs (most reliable)
    expr_cols <- character(0)
    if (exists("meta_df_exp") && !is.null(meta_df_exp) && exists("sample_col_exp")) {
        meta_ids <- as.character(meta_df_exp[[sample_col_exp]])
        expr_cols <- intersect(meta_ids, names(src))
    }
    # Fallback: exclude known stat columns, keep numeric
    if (length(expr_cols) == 0) {
        stat_patterns <- "^(padj|pvalue|linearFC|linearRatio|upDown|pass|sum\\.pass|n_pass|manual_cutoffs|FeatureID|Protein\\.Names|Genes|First\\.Protein)"
        expr_cols <- names(src)[!grepl(stat_patterns, names(src))]
        expr_cols <- expr_cols[vapply(src[expr_cols], is.numeric, logical(1))]
    }

    if (length(expr_cols) == 0) return(NULL)

    # If using expr_df, align rows to df by id_col
    if (!identical(src, df) && id_col %in% names(src)) {
        m <- match(df[[id_col]], src[[id_col]])
        baseMean <- rowMeans(as.matrix(src[m, expr_cols, drop = FALSE]), na.rm = TRUE)
    } else {
        baseMean <- rowMeans(as.matrix(src[, expr_cols, drop = FALSE]), na.rm = TRUE)
    }

    # Expression values are already log2-transformed, so baseMean is the
    # average log2 intensity — use directly as the A axis (no extra log)
    plot_df <- data.frame(
        protein = df[[gene_col]],
        logFC = signed_fc_to_log2(as.numeric(df[[lfc_col]])),
        padj = as.numeric(df[[padj_col]]),
        baseMean = baseMean,
        stringsAsFactors = FALSE
    )
    plot_df <- plot_df[!is.na(plot_df$baseMean) & plot_df$baseMean > 0, ]
    plot_df$Direction <- ifelse(
        !is.na(plot_df$padj) & plot_df$padj <= padj_cut & abs(plot_df$logFC) >= lfc_cut,
        ifelse(plot_df$logFC > 0, "Up", "Down"), "NS"
    )
    plot_df$Direction <- factor(plot_df$Direction, levels = c("NS", "Down", "Up"))

    p <- ggplot(plot_df, aes(x = baseMean, y = logFC, color = Direction,
                              text = paste0("Protein: ", protein, "\nlog2FC: ", round(logFC, 3),
                                           "\nMean log2 intensity: ", round(baseMean, 2),
                                           "\npadj: ", signif(padj, 3)))) +
        geom_point(alpha = 0.6, size = 1.5) +
        scale_color_manual(values = c("NS" = "grey70", "Down" = "blue", "Up" = "red")) +
        geom_hline(yintercept = c(-lfc_cut, lfc_cut), linetype = "dashed", alpha = 0.5) +
        geom_hline(yintercept = 0, color = "black", linewidth = 0.3) +
        labs(title = title, x = "Mean log2 Intensity", y = "log2 Fold Change") +
        theme_minimal()
    widget <- ggplotly(p, tooltip = "text")
    pval_col <- paste0("pvalue.imputs.", contrast)
    raw_pval <- if (pval_col %in% names(df)) as.numeric(df[[pval_col]]) else plot_df$padj
    # Match raw_pval and pass flag to filtered plot_df rows
    keep_mask <- !is.na(baseMean) & baseMean > 0
    raw_pval_filt <- raw_pval[keep_mask]
    pass_col <- paste0("pass.imputs.", contrast)
    pass_flag <- if (pass_col %in% names(df)) as.numeric(df[[pass_col]])[keep_mask] else NULL
    reg_df <- data.frame(name = as.character(plot_df$protein),
                         logFC = plot_df$logFC,
                         pval = raw_pval_filt,
                         adjPval = plot_df$padj,
                         avgExpr = plot_df$baseMean,
                         stringsAsFactors = FALSE)
    if (!is.null(pass_flag)) reg_df$passFlag <- pass_flag
    list(widget = widget, plot_df = reg_df)
}
```

```{r ma-interactive, eval=show_ma_plots, results='asis', fig.height=6}
if (!is.null(de_data) && length(contrast_names) > 0) {
    for (i in seq_along(contrast_names)) {
        cn <- contrast_names[i]
        contrast_clean <- gsub("_", " ", cn)
        clean_id <- gsub("[^a-zA-Z0-9]", "_", cn)

        cat(sprintf("\n#### %s\n\n", contrast_clean))

        ma_res <- make_ma_plotly_prot(de_data, ma_expr_df, cn, prot_cfg$de$p_cutoff %||% 0.05,
                                  log2(prot_cfg$de$linear_fc_cutoff %||% 1.5), contrast_clean)
        if (!is.null(ma_res)) {
            p <- ma_res$widget
            p$elementId <- paste0("ma_plot_", clean_id)
            print(htmltools::tagList(p))
            if (exists("cutoff_register_plot")) {
                cat(cutoff_register_plot(
                    paste0("ma_plot_", clean_id),
                    ma_res$plot_df, "ma", "Protein", cn
                ))
            }
        }
        cat("\n\n")
    }
} else {
    cat("Interactive MA plots require DE data.\n")
}
```

```{r top-de-heatmaps-heading, eval=show_top_de_heatmaps, results='asis'}
cat("## Top DE Heatmaps {.tabset}\n\n")
```

```{r top-de-heatmaps, eval=show_top_de_heatmaps, results='asis'}
heatmap_files <- list.files(de_png_dir, pattern = "^heatmap_top.*\\.png$", full.names = TRUE)
if (length(heatmap_files) > 0) {
    # Parse contrast names and sizes
    for (hf in heatmap_files) {
        label <- gsub("^heatmap_|\\.png$", "", basename(hf))
        label_clean <- gsub("_", " ", label)
        cat(sprintf("\n### %s\n\n", label_clean))
        cat(sprintf("![%s](%s)\n\n", label_clean, hf))
        figure_legend(sprintf("Heatmap of %s across samples. Rows are z-score normalized. Samples are annotated by experimental group.", label_clean))
        render_commentary(paste0("heatmap_", label))
    }
} else {
    cat("No top DE heatmaps found. Enable with `qc.top_de_heatmap_sizes` in config.\n")
}
```

```{r protein-explorer-heading, eval=show_protein_explorer, results='asis'}
cat("## Protein Expression Explorer\n\n")
cat("The expression explorer provides an interactive interface for examining individual protein expression patterns across experimental groups. Box plots show the distribution of expression values for each condition, enabling visual assessment of effect sizes and within-group variability.\n\n")
```

```{r protein-explorer-data, eval=show_protein_explorer, include=FALSE}
explorer_ready <- FALSE
explorer_json_str <- "{}"

# Load expression data: use final_results.tsv which has per-sample columns + annotations
meta_file_exp <- file.path(proj_dir_resolved, cfg$paths$raw, prot_cfg$files$metadata)
meta_df_exp <- if (file.exists(meta_file_exp)) read.csv(meta_file_exp, stringsAsFactors = FALSE) else NULL
group_col_exp <- prot_cfg$effects$color %||% "Condition"
sample_col_exp <- prot_cfg$effects$samples %||% "SampleID"

norm_expr_exp <- NULL

if (!is.null(meta_df_exp) && sample_col_exp %in% names(meta_df_exp)) {
    meta_sample_ids <- as.character(meta_df_exp[[sample_col_exp]])

    # Primary: final_results.tsv (has expression + annotation columns)
    final_file <- file.path(datasets_dir, "final_results.tsv")
    if (file.exists(final_file)) {
        final_df <- tryCatch(
            read.delim(final_file, stringsAsFactors = FALSE, check.names = FALSE),
            error = function(e) NULL
        )
        if (!is.null(final_df)) {
            id_col_exp <- prot_cfg$de_table$id_col %||% "FeatureID"
            # Extract only sample columns that match metadata
            expr_cols_exp <- intersect(meta_sample_ids, names(final_df))
            if (length(expr_cols_exp) > 0 && id_col_exp %in% names(final_df)) {
                norm_expr_exp <- final_df[, expr_cols_exp, drop = FALSE]
                rownames(norm_expr_exp) <- final_df[[id_col_exp]]
            }
        }
    }

    # Fallback: imputed expression file
    if (is.null(norm_expr_exp)) {
        imp_files <- list.files(datasets_dir, pattern = "imputed.*\\.tsv$", full.names = TRUE)
        if (length(imp_files) > 0) {
            norm_expr_exp <- tryCatch(
                read.delim(imp_files[1], stringsAsFactors = FALSE, row.names = 1, check.names = FALSE),
                error = function(e) NULL
            )
            if (!is.null(norm_expr_exp)) {
                # Keep only columns matching metadata sample IDs
                keep <- intersect(meta_sample_ids, colnames(norm_expr_exp))
                if (length(keep) > 0) norm_expr_exp <- norm_expr_exp[, keep, drop = FALSE]
                else norm_expr_exp <- NULL
            }
        }
    }
}

if (!is.null(norm_expr_exp) && !is.null(meta_df_exp) && ncol(norm_expr_exp) >= 2) {
    protein_ids <- rownames(norm_expr_exp)
    protein_display <- protein_ids

    # Use Genes column from de_data if available
    if (!is.null(de_data) && "Genes" %in% names(de_data)) {
        id_col_map <- prot_cfg$de_table$id_col %||% "FeatureID"
        if (id_col_map %in% names(de_data)) {
            gene_map <- setNames(de_data$Genes, de_data[[id_col_map]])
            mapped <- gene_map[protein_ids]
            has_map <- !is.na(mapped) & nzchar(mapped)
            protein_display[has_map] <- mapped[has_map]
        }
    }

    # Build sample -> group mapping
    if (sample_col_exp %in% names(meta_df_exp) && group_col_exp %in% names(meta_df_exp)) {
        group_map_exp <- setNames(as.character(meta_df_exp[[group_col_exp]]), meta_df_exp[[sample_col_exp]])
        sample_groups_exp <- as.character(group_map_exp[colnames(norm_expr_exp)])
        sample_groups_exp[is.na(sample_groups_exp)] <- "Unknown"
    } else {
        sample_groups_exp <- rep("All", ncol(norm_expr_exp))
    }

    expr_matrix_list <- lapply(seq_len(nrow(norm_expr_exp)), function(i) {
        round(as.numeric(norm_expr_exp[i, ]), 3)
    })

    explorer_data <- list(
        genes = protein_display,
        samples = colnames(norm_expr_exp),
        groups = sample_groups_exp,
        expr = expr_matrix_list
    )
    explorer_json_str <- jsonlite::toJSON(explorer_data, auto_unbox = TRUE)
    explorer_ready <- TRUE
}
```

```{r protein-explorer-ui, eval=show_protein_explorer && explorer_ready, results='asis'}
cat('
<div style="margin: 15px 0 20px 0;">
<div style="display: flex; align-items: center; gap: 10px; flex-wrap: wrap;">
<label for="boxplot_gene_input" style="font-weight: bold; white-space: nowrap;">Search protein:</label>
<div style="position: relative; flex: 1; max-width: 400px;">
<input type="text" id="boxplot_gene_input" placeholder="Start typing a protein name..."
  autocomplete="off"
  style="padding: 10px 12px; border-radius: 6px; border: 1px solid #ccc; width: 100%; font-size: 14px; box-sizing: border-box;"
  oninput="boxplotAutocomplete(this)" onkeydown="boxplotAutoKey(event)">
<div id="boxplot_ac" style="position: absolute; top: 100%; left: 0; width: 100%; max-height: 220px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 6px 6px; z-index: 1000; display: none; box-shadow: 0 4px 8px rgba(0,0,0,0.12);"></div>
</div>
<button onclick="addGeneToPlot()" style="padding: 10px 20px; background: #007bff; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Add Protein</button>
<button onclick="clearBoxplot()" style="padding: 10px 20px; background: #6c757d; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Clear All</button>
</div>
<div id="boxplot_gene_tags" style="margin-top: 8px; display: flex; flex-wrap: wrap; gap: 6px;"></div>
<div id="boxplot_gene_info" style="margin-top: 8px; font-size: 0.9em; color: #555;"></div>
</div>
<div id="boxplot_container" style="width: 100%; min-height: 450px; margin-bottom: 10px;"></div>
<div id="boxplot_export_bar" style="display: none; margin-bottom: 20px; gap: 8px; align-items: center;">
  <span style="font-size: 0.85em; color: #555; margin-right: 4px;">Export:</span>
  <button onclick="exportBoxplot(\'svg\')" style="padding: 5px 14px; background: #28a745; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">SVG</button>
  <button onclick="exportBoxplot(\'png\')" style="padding: 5px 14px; background: #17a2b8; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">PNG</button>
</div>
')
```

```{r protein-explorer-js, eval=show_protein_explorer && explorer_ready, results='asis'}
cat('<script>\nvar geneExprData = ')
cat(explorer_json_str)
cat(';\n')
cat(sprintf('var bpGroupLabel = "%s";\n', group_col_exp))
cat('
var boxplotAcIndex = -1;
var selectedGenes = [];

function boxplotAutocomplete(el) {
    var acDiv = document.getElementById("boxplot_ac");
    var val = el.value.trim().toLowerCase();
    acDiv.innerHTML = ""; boxplotAcIndex = -1;
    if (val.length < 1) { acDiv.style.display = "none"; return; }
    var matches = [];
    for (var i = 0; i < geneExprData.genes.length; i++) {
        if (geneExprData.genes[i].toLowerCase().indexOf(val) !== -1) {
            matches.push({name: geneExprData.genes[i], idx: i});
            if (matches.length >= 20) break;
        }
    }
    if (matches.length === 0) { acDiv.style.display = "none"; return; }
    matches.forEach(function(m) {
        var item = document.createElement("div");
        item.textContent = m.name;
        item.style.cssText = "padding: 8px 12px; cursor: pointer; font-size: 13px; border-bottom: 1px solid #f0f0f0;";
        item.onmouseenter = function() { this.style.background = "#e9ecef"; };
        item.onmouseleave = function() { this.style.background = "white"; };
        item.onclick = function() { document.getElementById("boxplot_gene_input").value = m.name; acDiv.style.display = "none"; addGeneToPlot(); };
        acDiv.appendChild(item);
    });
    acDiv.style.display = "block";
}

function boxplotAutoKey(e) {
    var acDiv = document.getElementById("boxplot_ac");
    var items = acDiv.children;
    if (items.length === 0) { if (e.key === "Enter") addGeneToPlot(); return; }
    if (e.key === "ArrowDown") { e.preventDefault(); boxplotAcIndex = Math.min(boxplotAcIndex + 1, items.length - 1); updateBpAcHighlight(items); }
    else if (e.key === "ArrowUp") { e.preventDefault(); boxplotAcIndex = Math.max(boxplotAcIndex - 1, 0); updateBpAcHighlight(items); }
    else if (e.key === "Enter") { e.preventDefault(); if (boxplotAcIndex >= 0) items[boxplotAcIndex].click(); else addGeneToPlot(); }
}

function updateBpAcHighlight(items) { for (var i = 0; i < items.length; i++) items[i].style.background = i === boxplotAcIndex ? "#e9ecef" : "white"; }

function addGeneToPlot() {
    var input = document.getElementById("boxplot_gene_input");
    var val = input.value.trim();
    if (!val) return;
    var valLower = val.toLowerCase();
    var found = -1;
    for (var i = 0; i < geneExprData.genes.length; i++) {
        if (geneExprData.genes[i].toLowerCase() === valLower) { found = i; break; }
    }
    if (found === -1) {
        for (var i = 0; i < geneExprData.genes.length; i++) {
            if (geneExprData.genes[i].toLowerCase().indexOf(valLower) !== -1) { found = i; break; }
        }
    }
    if (found === -1) { document.getElementById("boxplot_gene_info").textContent = "Protein not found: " + val; return; }
    if (selectedGenes.some(function(g) { return g.idx === found; })) { document.getElementById("boxplot_gene_info").textContent = "Already added."; input.value = ""; return; }
    selectedGenes.push({name: geneExprData.genes[found], idx: found});
    input.value = "";
    document.getElementById("boxplot_ac").style.display = "none";
    document.getElementById("boxplot_gene_info").textContent = "";
    renderBoxplotTags();
    renderMultiBoxplot();
}

function renderBoxplotTags() {
    var container = document.getElementById("boxplot_gene_tags");
    container.innerHTML = "";
    selectedGenes.forEach(function(g, i) {
        var tag = document.createElement("span");
        tag.style.cssText = "display: inline-flex; align-items: center; background: #007bff; color: white; padding: 3px 10px; border-radius: 14px; font-size: 13px; gap: 5px;";
        tag.innerHTML = g.name + " <span style=\\"cursor:pointer; font-weight:bold;\\" onclick=\\"removeBoxplotGene(" + i + ")\\">&times;</span>";
        container.appendChild(tag);
    });
}

function removeBoxplotGene(idx) {
    selectedGenes.splice(idx, 1);
    renderBoxplotTags();
    if (selectedGenes.length > 0) { renderMultiBoxplot(); }
    else { Plotly.purge("boxplot_container"); document.getElementById("boxplot_export_bar").style.display = "none"; document.getElementById("boxplot_gene_info").textContent = ""; }
}
function clearBoxplot() {
    selectedGenes = [];
    renderBoxplotTags();
    Plotly.purge("boxplot_container");
    document.getElementById("boxplot_gene_info").textContent = "";
    document.getElementById("boxplot_export_bar").style.display = "none";
}

function renderMultiBoxplot() {
    if (selectedGenes.length === 0) return;

    var groups = geneExprData.groups;
    var samples = geneExprData.samples;
    var uniqueGroups = [];
    groups.forEach(function(g) { if (uniqueGroups.indexOf(g) === -1) uniqueGroups.push(g); });

    var colors = ["#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00", "#a65628", "#f781bf", "#999999"];
    var nGenes = selectedGenes.length;
    var traces = [];

    if (nGenes === 1) {
        // Single protein: one box per group, distinct colors
        var geneIdx = selectedGenes[0].idx;
        var values = geneExprData.expr[geneIdx];
        uniqueGroups.forEach(function(g, gi) {
            var gValues = [], gSamples = [];
            for (var i = 0; i < values.length; i++) {
                if (groups[i] === g) { gValues.push(values[i]); gSamples.push(samples[i]); }
            }
            traces.push({
                y: gValues, type: "box", name: g,
                boxpoints: "all", jitter: 0.4, pointpos: 0,
                text: gSamples,
                hovertemplate: "Sample: %{text}<br>Expression: %{y:.3f}<extra>" + g + "</extra>",
                marker: { color: colors[gi % colors.length], size: 6, opacity: 0.7 },
                line: { color: colors[gi % colors.length] },
                fillcolor: colors[gi % colors.length] + "33"
            });
        });
        var layout = {
            title: { text: selectedGenes[0].name + " Expression", font: { size: 16 } },
            yaxis: { title: "Expression (log2 intensity)" },
            xaxis: { title: bpGroupLabel },
            showlegend: false,
            margin: { t: 50, b: 80, l: 70, r: 30 },
            boxgap: 0.3
        };
        var containerH = 450;
    } else {
        // Multiple proteins: genes on x-axis, grouped by condition
        var containerH = Math.max(450, 350 + nGenes * 30);
        uniqueGroups.forEach(function(g, gi) {
            var xVals = [], yVals = [], sampleTexts = [];
            selectedGenes.forEach(function(sg) {
                var values = geneExprData.expr[sg.idx];
                for (var i = 0; i < values.length; i++) {
                    if (groups[i] === g) {
                        xVals.push(sg.name);
                        yVals.push(values[i]);
                        sampleTexts.push(samples[i]);
                    }
                }
            });
            traces.push({
                x: xVals, y: yVals, type: "box", name: g,
                boxpoints: "all", jitter: 0.4, pointpos: 0,
                text: sampleTexts,
                hovertemplate: "%{x}<br>Sample: %{text}<br>Expression: %{y:.3f}<extra>" + g + "</extra>",
                marker: { color: colors[gi % colors.length], size: 5, opacity: 0.7 },
                line: { color: colors[gi % colors.length] },
                fillcolor: colors[gi % colors.length] + "33"
            });
        });

        var protNames = selectedGenes.map(function(sg) { return sg.name; });
        var layout = {
            title: { text: "Protein Expression Comparison", font: { size: 16 } },
            yaxis: { title: "Expression (log2 intensity)" },
            xaxis: { title: "Protein", categoryorder: "array", categoryarray: protNames },
            showlegend: true,
            legend: { title: { text: bpGroupLabel } },
            boxmode: "group",
            margin: { t: 50, b: 100, l: 70, r: 30 },
            boxgap: 0.15,
            boxgroupgap: 0.1
        };
    }

    document.getElementById("boxplot_container").style.height = containerH + "px";
    Plotly.newPlot("boxplot_container", traces, layout, { responsive: true });

    // Show info
    var infoLines = selectedGenes.map(function(sg) {
        var vals = geneExprData.expr[sg.idx].filter(function(v) { return !isNaN(v); });
        var mean = vals.reduce(function(a, b) { return a + b; }, 0) / vals.length;
        return "<strong>" + sg.name + "</strong>: mean " + mean.toFixed(3) + " (" + vals.length + " samples)";
    });
    document.getElementById("boxplot_gene_info").innerHTML = infoLines.join(" &nbsp;|&nbsp; ");
    document.getElementById("boxplot_export_bar").style.display = "flex";
}

function exportBoxplot(format) {
    var container = document.getElementById("boxplot_container");
    if (!container || !container.data) return;
    var names = selectedGenes.map(function(g) { return g.name; }).join("_");
    if (format === "svg") Plotly.downloadImage(container, {format: "svg", width: 1200, height: 800, filename: "protein_expression_" + names});
    else Plotly.downloadImage(container, {format: "png", width: 2400, height: 1600, scale: 3, filename: "protein_expression_" + names});
}

document.addEventListener("click", function(e) {
    var acDiv = document.getElementById("boxplot_ac");
    var input = document.getElementById("boxplot_gene_input");
    if (acDiv && e.target !== input && !acDiv.contains(e.target)) acDiv.style.display = "none";
});
</script>
')
```

```{r protein-explorer-fallback, eval=show_protein_explorer && !explorer_ready, results='asis'}
cat("*Protein expression explorer not available. Normalized expression matrix not found.*\n")
```

```{r peptide-explorer-heading, eval=show_peptide_explorer, results='asis'}
cat("## Peptide Expression Explorer\n\n")
cat("The peptide explorer allows you to search for a protein and view the expression of its individual peptides across experimental groups. This provides insight into peptide-level variability, potential isoform-specific effects, and data quality at the peptide level.\n\n")
```

```{r peptide-explorer-data, eval=show_peptide_explorer, include=FALSE}
peptide_ready <- FALSE
peptide_json_str <- "{}"

# Load peptide data if available in config
peptide_file_path <- NULL
if (!is.null(prot_cfg$files$peptides) && nzchar(prot_cfg$files$peptides)) {
    peptide_file_path <- file.path(proj_dir_resolved, cfg$paths$raw, prot_cfg$files$peptides)
    if (!file.exists(peptide_file_path)) peptide_file_path <- NULL
}

# Also check for peptide file in datasets directory
if (is.null(peptide_file_path)) {
    pep_candidates <- list.files(datasets_dir, pattern = "peptide", full.names = TRUE, ignore.case = TRUE)
    if (length(pep_candidates) > 0) peptide_file_path <- pep_candidates[1]
}

peptide_df <- NULL
if (!is.null(peptide_file_path) && file.exists(peptide_file_path)) {
    peptide_df <- tryCatch({
        if (grepl("\\.tsv$|\\.tab$|\\.txt$", peptide_file_path, ignore.case = TRUE)) {
            read.delim(peptide_file_path, stringsAsFactors = FALSE, check.names = FALSE)
        } else {
            read.csv(peptide_file_path, stringsAsFactors = FALSE, check.names = FALSE)
        }
    }, error = function(e) NULL)
}

if (!is.null(peptide_df) && !is.null(meta_df_exp) && nrow(peptide_df) > 0) {
    # Auto-detect protein ID column in peptide data
    pep_protein_col <- NULL
    for (cand in c("Protein.Group", "Protein.Ids", "Protein.ID", "Protein",
                    "Leading.razor.protein", "Master.Protein.Accessions",
                    "ProteinID", "Protein.Accession")) {
        if (cand %in% colnames(peptide_df)) { pep_protein_col <- cand; break }
    }

    # Auto-detect peptide sequence column
    pep_seq_col <- NULL
    for (cand in c("Stripped.Sequence", "Peptide.Sequence", "Sequence",
                    "Modified.Sequence", "Annotated.Sequence", "Peptide")) {
        if (cand %in% colnames(peptide_df)) { pep_seq_col <- cand; break }
    }

    if (!is.null(pep_protein_col) && !is.null(pep_seq_col)) {
        meta_sample_ids_pep <- as.character(meta_df_exp[[sample_col_exp]])
        pep_expr_cols <- intersect(meta_sample_ids_pep, colnames(peptide_df))

        # DIA-NN raw path column mapping: extract numeric prefix from raw file paths
        # e.g., "D:\...\Seq93724_DIA_480Ex2.raw" -> match "93724" to "93724_prot-OXY-1A"
        if (length(pep_expr_cols) == 0) {
            raw_path_cols <- grep("\\.(raw|mzML|d)$", colnames(peptide_df),
                                   value = TRUE, ignore.case = TRUE)
            if (length(raw_path_cols) > 0) {
                # Extract numeric ID from raw file name (handles both original
                # backslash paths and R-mangled dot paths like D..DIANN.projects...)
                # Strategy: find the last Seq/seq prefix followed by digits, or
                # extract the first 4+ digit number after the last separator
                raw_nums <- vapply(raw_path_cols, function(col) {
                    # Try Seq<digits> pattern first
                    m <- regmatches(col, regexpr("[Ss]eq(\\d{4,})", col))
                    if (length(m) > 0) return(sub("[Ss]eq", "", m[1]))
                    # Fallback: last 4+ digit group in the string
                    digits <- regmatches(col, gregexpr("\\d{4,}", col))[[1]]
                    if (length(digits) > 0) return(digits[length(digits)])
                    return(col)  # no match
                }, character(1))
                meta_nums <- sub("^(\\d+).*$", "\\1", meta_sample_ids_pep)
                # Build mapping: raw_path_col -> metadata SampleID
                pep_col_map <- setNames(character(length(raw_path_cols)), raw_path_cols)
                for (i in seq_along(raw_nums)) {
                    match_idx <- which(meta_nums == raw_nums[i])
                    if (length(match_idx) == 1) {
                        pep_col_map[raw_path_cols[i]] <- meta_sample_ids_pep[match_idx]
                    }
                }
                mapped_cols <- pep_col_map[nzchar(pep_col_map)]
                if (length(mapped_cols) >= 2) {
                    # Rename columns in peptide_df
                    col_idx <- match(names(mapped_cols), colnames(peptide_df))
                    colnames(peptide_df)[col_idx] <- unname(mapped_cols)
                    pep_expr_cols <- unname(mapped_cols)
                    message("Peptide Explorer: mapped ", length(mapped_cols),
                            " DIA-NN raw columns to sample IDs")
                }
            }
        }

        if (length(pep_expr_cols) >= 2) {
            # Build sample -> group mapping
            if (sample_col_exp %in% names(meta_df_exp) && group_col_exp %in% names(meta_df_exp)) {
                pep_group_map <- setNames(as.character(meta_df_exp[[group_col_exp]]),
                                           meta_df_exp[[sample_col_exp]])
                pep_sample_groups <- as.character(pep_group_map[pep_expr_cols])
                pep_sample_groups[is.na(pep_sample_groups)] <- "Unknown"
            } else {
                pep_sample_groups <- rep("All", length(pep_expr_cols))
            }

            # Build protein -> peptide mapping
            proteins_with_peptides <- unique(peptide_df[[pep_protein_col]])
            # Take first protein in group (handle semicolon-separated)
            peptide_df$.prot_first <- vapply(peptide_df[[pep_protein_col]], function(x) {
                trimws(strsplit(as.character(x), ";")[[1]][1])
            }, character(1))

            # Map protein IDs to display names (gene symbols)
            pep_protein_display <- unique(peptide_df$.prot_first)
            if (!is.null(de_data) && "Genes" %in% names(de_data)) {
                id_col_pep <- prot_cfg$de_table$id_col %||% "FeatureID"
                if (id_col_pep %in% names(de_data)) {
                    gene_map_pep <- setNames(de_data$Genes, de_data[[id_col_pep]])
                    mapped_pep <- gene_map_pep[pep_protein_display]
                    has_map_pep <- !is.na(mapped_pep) & nzchar(mapped_pep)
                    pep_display_map <- setNames(pep_protein_display, pep_protein_display)
                    pep_display_map[has_map_pep] <- paste0(mapped_pep[has_map_pep],
                                                            " (", pep_protein_display[has_map_pep], ")")
                } else {
                    pep_display_map <- setNames(pep_protein_display, pep_protein_display)
                }
            } else {
                pep_display_map <- setNames(pep_protein_display, pep_protein_display)
            }

            # Aggregate precursors to peptide level (mean intensity per protein+sequence)
            peptide_df$.display <- unname(pep_display_map[peptide_df$.prot_first])
            na_display <- is.na(peptide_df$.display)
            peptide_df$.display[na_display] <- peptide_df$.prot_first[na_display]

            agg_key <- paste(peptide_df$.display, peptide_df[[pep_seq_col]], sep = "|||")
            agg_expr <- aggregate(
                peptide_df[, pep_expr_cols, drop = FALSE],
                by = list(.key = agg_key),
                FUN = function(x) mean(x, na.rm = TRUE)
            )
            agg_parts <- strsplit(agg_expr$.key, "\\|\\|\\|")
            agg_protein <- vapply(agg_parts, `[`, character(1), 1)
            agg_sequence <- vapply(agg_parts, `[`, character(1), 2)

            # Log2-transform raw intensities for display (peptide data from
            # DIA-NN pr_matrix is raw unnormalized intensity, not log2)
            for (ec in pep_expr_cols) {
                v <- agg_expr[[ec]]
                agg_expr[[ec]] <- ifelse(is.na(v) | v <= 0, NA_real_, log2(v))
            }

            # Build JSON
            pep_entries <- lapply(seq_len(nrow(agg_expr)), function(i) {
                list(
                    protein = agg_protein[i],
                    sequence = agg_sequence[i],
                    expr = round(as.numeric(agg_expr[i, pep_expr_cols]), 3)
                )
            })

            peptide_explorer_data <- list(
                proteins = sort(unique(agg_protein)),
                peptides = pep_entries,
                samples = pep_expr_cols,
                groups = pep_sample_groups
            )

            peptide_json_str <- jsonlite::toJSON(peptide_explorer_data, auto_unbox = TRUE)
            peptide_ready <- TRUE
        }
    }
}
```

```{r peptide-explorer-ui, eval=show_peptide_explorer && peptide_ready, results='asis'}
cat('
<div style="margin: 15px 0 20px 0;">
<div style="display: flex; align-items: center; gap: 10px; flex-wrap: wrap;">
<label for="pep_protein_input" style="font-weight: bold; white-space: nowrap;">Search protein:</label>
<div style="position: relative; flex: 1; max-width: 400px;">
<input type="text" id="pep_protein_input" placeholder="Start typing a protein name..."
  autocomplete="off"
  style="padding: 10px 12px; border-radius: 6px; border: 1px solid #ccc; width: 100%; font-size: 14px; box-sizing: border-box;"
  oninput="pepAutocomplete(this)" onkeydown="pepAutoKey(event)">
<div id="pep_ac" style="position: absolute; top: 100%; left: 0; width: 100%; max-height: 220px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 6px 6px; z-index: 1000; display: none; box-shadow: 0 4px 8px rgba(0,0,0,0.12);"></div>
</div>
<button onclick="showPeptides()" style="padding: 10px 20px; background: #28a745; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Show Peptides</button>
<button onclick="clearPeptidePlot()" style="padding: 10px 20px; background: #6c757d; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Clear</button>
</div>
<div id="pep_info" style="margin-top: 8px; font-size: 0.9em; color: #555;"></div>
</div>
<div id="pep_container" style="width: 100%; min-height: 500px; margin-bottom: 10px;"></div>
<div id="pep_export_bar" style="display: none; margin-bottom: 20px; gap: 8px; align-items: center;">
  <span style="font-size: 0.85em; color: #555; margin-right: 4px;">Export:</span>
  <button onclick="exportPepPlot(\'svg\')" style="padding: 5px 14px; background: #28a745; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">SVG</button>
  <button onclick="exportPepPlot(\'png\')" style="padding: 5px 14px; background: #17a2b8; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">PNG</button>
</div>
')
```

```{r peptide-explorer-js, eval=show_peptide_explorer && peptide_ready, results='asis'}
cat('<script>\nvar pepData = ')
cat(peptide_json_str)
cat(';\n')
cat(sprintf('var pepGroupLabel = "%s";\n', group_col_exp))
cat('
var pepAcIndex = -1;

function pepAutocomplete(el) {
    var acDiv = document.getElementById("pep_ac");
    var val = el.value.trim().toLowerCase();
    acDiv.innerHTML = ""; pepAcIndex = -1;
    if (val.length < 1) { acDiv.style.display = "none"; return; }
    var matches = [];
    for (var i = 0; i < pepData.proteins.length; i++) {
        if (pepData.proteins[i].toLowerCase().indexOf(val) !== -1) {
            matches.push(pepData.proteins[i]);
            if (matches.length >= 20) break;
        }
    }
    if (matches.length === 0) { acDiv.style.display = "none"; return; }
    matches.forEach(function(m) {
        var item = document.createElement("div");
        item.textContent = m;
        item.style.cssText = "padding: 8px 12px; cursor: pointer; font-size: 13px; border-bottom: 1px solid #f0f0f0;";
        item.onmouseenter = function() { this.style.background = "#e9ecef"; };
        item.onmouseleave = function() { this.style.background = "white"; };
        item.onclick = function() { document.getElementById("pep_protein_input").value = m; acDiv.style.display = "none"; showPeptides(); };
        acDiv.appendChild(item);
    });
    acDiv.style.display = "block";
}

function pepAutoKey(e) {
    var acDiv = document.getElementById("pep_ac");
    var items = acDiv.children;
    if (items.length === 0) { if (e.key === "Enter") showPeptides(); return; }
    if (e.key === "ArrowDown") { e.preventDefault(); pepAcIndex = Math.min(pepAcIndex + 1, items.length - 1); for (var i = 0; i < items.length; i++) items[i].style.background = i === pepAcIndex ? "#e9ecef" : "white"; }
    else if (e.key === "ArrowUp") { e.preventDefault(); pepAcIndex = Math.max(pepAcIndex - 1, 0); for (var i = 0; i < items.length; i++) items[i].style.background = i === pepAcIndex ? "#e9ecef" : "white"; }
    else if (e.key === "Enter") { e.preventDefault(); if (pepAcIndex >= 0) items[pepAcIndex].click(); else showPeptides(); }
}

function showPeptides() {
    var input = document.getElementById("pep_protein_input");
    var val = input.value.trim();
    if (!val) return;
    var valLower = val.toLowerCase();

    // Find matching protein
    var protName = null;
    for (var i = 0; i < pepData.proteins.length; i++) {
        if (pepData.proteins[i].toLowerCase() === valLower) { protName = pepData.proteins[i]; break; }
    }
    if (!protName) {
        for (var i = 0; i < pepData.proteins.length; i++) {
            if (pepData.proteins[i].toLowerCase().indexOf(valLower) !== -1) { protName = pepData.proteins[i]; break; }
        }
    }
    if (!protName) { document.getElementById("pep_info").textContent = "Protein not found: " + val; return; }

    document.getElementById("pep_ac").style.display = "none";

    // Collect peptides for this protein
    var peptides = [];
    for (var i = 0; i < pepData.peptides.length; i++) {
        if (pepData.peptides[i].protein === protName) {
            peptides.push(pepData.peptides[i]);
        }
    }

    if (peptides.length === 0) {
        document.getElementById("pep_info").textContent = "No peptides found for: " + protName;
        return;
    }

    var groups = pepData.groups;
    var samples = pepData.samples;
    var uniqueGroups = [];
    groups.forEach(function(g) { if (uniqueGroups.indexOf(g) === -1) uniqueGroups.push(g); });
    var colors = ["#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00", "#a65628", "#f781bf", "#999999"];

    var traces = [];
    uniqueGroups.forEach(function(g, gi) {
        var xVals = [], yVals = [], sampleTexts = [];
        peptides.forEach(function(pep) {
            var seq = pep.sequence.length > 20 ? pep.sequence.substring(0, 17) + "..." : pep.sequence;
            for (var i = 0; i < pep.expr.length; i++) {
                if (groups[i] === g && pep.expr[i] !== null && !isNaN(pep.expr[i])) {
                    xVals.push(seq);
                    yVals.push(pep.expr[i]);
                    sampleTexts.push(samples[i] + "\\n" + pep.sequence);
                }
            }
        });
        traces.push({
            x: xVals, y: yVals, type: "box", name: g,
            boxpoints: "all", jitter: 0.4, pointpos: 0,
            text: sampleTexts,
            hovertemplate: "%{x}<br>%{text}<br>Intensity: %{y:.3f}<extra>" + g + "</extra>",
            marker: { color: colors[gi % colors.length], size: 5, opacity: 0.7 },
            line: { color: colors[gi % colors.length] },
            fillcolor: colors[gi % colors.length] + "33"
        });
    });

    var pepNames = peptides.map(function(p) {
        return p.sequence.length > 20 ? p.sequence.substring(0, 17) + "..." : p.sequence;
    });

    var containerH = Math.max(500, 350 + peptides.length * 25);
    var layout = {
        title: { text: protName + " — Peptide Expression (" + peptides.length + " peptides)", font: { size: 15 } },
        yaxis: { title: "Raw Intensity (log2)" },
        xaxis: { title: "Peptide Sequence", categoryorder: "array", categoryarray: pepNames,
                 tickangle: -45, tickfont: { size: 10, family: "monospace" } },
        showlegend: true,
        legend: { title: { text: pepGroupLabel } },
        boxmode: "group",
        margin: { t: 50, b: 160, l: 70, r: 30 },
        boxgap: 0.15, boxgroupgap: 0.1
    };

    document.getElementById("pep_container").style.height = containerH + "px";
    Plotly.newPlot("pep_container", traces, layout, { responsive: true });

    document.getElementById("pep_info").innerHTML =
        "<strong>" + protName + "</strong>: " + peptides.length + " peptides detected";
    document.getElementById("pep_export_bar").style.display = "flex";
}

function clearPeptidePlot() {
    Plotly.purge("pep_container");
    document.getElementById("pep_info").textContent = "";
    document.getElementById("pep_export_bar").style.display = "none";
}

function exportPepPlot(format) {
    var container = document.getElementById("pep_container");
    if (!container || !container.data) return;
    var name = document.getElementById("pep_protein_input").value || "peptides";
    if (format === "svg") Plotly.downloadImage(container, {format: "svg", width: 1400, height: 900, filename: "peptide_expr_" + name});
    else Plotly.downloadImage(container, {format: "png", width: 2800, height: 1800, scale: 3, filename: "peptide_expr_" + name});
}

document.addEventListener("click", function(e) {
    var acDiv = document.getElementById("pep_ac");
    var input = document.getElementById("pep_protein_input");
    if (acDiv && e.target !== input && !acDiv.contains(e.target)) acDiv.style.display = "none";
});
</script>
')
```

```{r peptide-explorer-fallback, eval=show_peptide_explorer && !peptide_ready, results='asis'}
cat("*Peptide expression explorer not available. No peptide-level data file was provided in the configuration (`files.peptides`). To enable this feature, provide a peptide intensity matrix (e.g., from DIA-NN or MaxQuant) in the config file.*\n")
```

---

```{r clustering-check, include=FALSE}
clustering_enabled <- isTRUE(prot_cfg$clustering$enabled)

# ---- Discover clustering outputs on disk (independent of in-memory results) ----
hier_dir   <- file.path(clustering_dir, "Hierarchical")
part_root  <- file.path(clustering_dir, "Partition_clustering")
binary_dir <- file.path(clustering_dir, "Binary_patterns")

hier_hm <- file.path(hier_dir, "Hierarchical_DE_heatmap.png")

# Partition: <Partition_clustering>/<Partition_clustering_<k>_clusters>/
part_subdirs <- if (dir.exists(part_root)) {
  list.dirs(part_root, recursive = FALSE)
} else character(0)
part_subdir <- if (length(part_subdirs) > 0) part_subdirs[1] else NA_character_
part_hm     <- if (!is.na(part_subdir)) file.path(part_subdir, "Partition_clustering_heatmap.png") else NA_character_

# Per-cluster profile PNGs (cluster_profiles_clusterN.png)
part_profile_pngs <- if (!is.na(part_subdir) && dir.exists(part_subdir)) {
  list.files(part_subdir, pattern = "^cluster_profiles_cluster[0-9]+\\.png$",
             full.names = TRUE)
} else character(0)

# Sort by numeric cluster number
if (length(part_profile_pngs) > 0) {
  ord <- order(as.integer(sub(".*cluster([0-9]+)\\.png$", "\\1",
                              basename(part_profile_pngs))))
  part_profile_pngs <- part_profile_pngs[ord]
}

# Binary patterns: Heatmap_<pattern>.png per pattern
binary_pngs <- if (dir.exists(binary_dir)) {
  list.files(binary_dir, pattern = "^Heatmap_[01]+\\.png$", full.names = TRUE)
} else character(0)

has_hier   <- file.exists(hier_hm)
has_part   <- !is.na(part_hm) && file.exists(part_hm)
has_binary <- length(binary_pngs) > 0
clustering_show <- clustering_enabled && (has_hier || has_part || has_binary)
```

```{r clustering-header, eval=clustering_show, results='asis'}
cat("# Clustering {.tabset}\n\n")
cat("Clustering was performed on the differentially expressed proteins. Each method below answers a different question:\n\n")
cat("- **Hierarchical** — bottom-up tree built from feature-by-feature distances; reveals nested groupings.\n")
cat("- **Partition** — splits features into a chosen number of clusters with similar expression trajectories across groups.\n")
cat("- **Binary patterns** — assigns each feature to an on/off (1/0) pattern across groups based on correlation.\n\n")
```

```{r clustering-hier, eval=clustering_show && has_hier, results='asis'}
cat("## Hierarchical\n\n")
cat(sprintf("![Hierarchical clustering heatmap](%s){width=100%%}\n\n", hier_hm))
figure_legend("Hierarchical clustering heatmap of differentially expressed proteins (rows z-scored). Row dendrogram reflects feature similarity; samples are annotated by experimental group.")
render_commentary("clustering_hierarchical")
```

```{r clustering-partition, eval=clustering_show && has_part, results='asis'}
cat("## Partition {.tabset}\n\n")

# 1. Heatmap first
cat("### Heatmap\n\n")
cat(sprintf("![Partition clustering heatmap](%s){width=100%%}\n\n", part_hm))
figure_legend("Partition clustering heatmap. Rows are grouped by cluster (gaps separate clusters); within-cluster ordering follows the feature similarity.")

# 2. Per-cluster expression profiles (sub-tabs, one per cluster)
if (length(part_profile_pngs) > 0) {
  cat("\n### Cluster Profiles {.tabset}\n\n")
  for (pf in part_profile_pngs) {
    k <- sub(".*cluster([0-9]+)\\.png$", "\\1", basename(pf))
    cat(sprintf("\n#### Cluster %s\n\n", k))
    cat(sprintf("![Cluster %s expression profile](%s){width=80%%}\n\n", k, pf))
    figure_legend(sprintf(
      "Mean (±SE) expression profile across experimental groups for proteins in cluster %s.", k))
  }
}
render_commentary("clustering_partition")
```

```{r clustering-binary, eval=clustering_show && has_binary, results='asis'}
cat("## Binary Patterns {.tabset}\n\n")
for (bf in binary_pngs) {
  pat <- sub("^Heatmap_([01]+)\\.png$", "\\1", basename(bf))
  cat(sprintf("\n### Pattern %s\n\n", pat))
  cat(sprintf("![Binary pattern %s](%s){width=100%%}\n\n", pat, bf))
  figure_legend(sprintf(
    "Heatmap of features assigned to binary on/off pattern '%s' across experimental groups (correlation-based assignment).", pat))
}
render_commentary("clustering_binary")
```

---

# Enrichment / Pathway Analysis

Gene set enrichment analysis tests whether predefined biological pathways show coordinated expression changes beyond what is expected by chance. fGSEA uses a ranked list of all genes (rank-based), while ORA tests only the significant gene list against a background (threshold-based).

```{r pathway-setup, include=FALSE}
has_enrichment <- dir.exists(enrich_dir) && length(list.files(enrich_dir, pattern = "\\.csv$")) > 0
has_pathway_plots <- dir.exists(enrich_plot_dir) && length(list.files(enrich_plot_dir, pattern = "\\.png$")) > 0
pathway_enabled <- isTRUE(prot_cfg$pathway$enabled)
```

```{r pathway-skip, eval=!has_enrichment, results='asis'}
if (!pathway_enabled) {
    cat("Pathway analysis was disabled. To enable it, set `pathway.enabled: true` in the configuration file.\n")
} else {
    cat("Pathway analysis was enabled but produced no results.\n")
}
```

```{r pathway-method-note, eval=has_enrichment, results='asis'}
cat("**Method:** Pathway enrichment analysis was performed using gene set enrichment approaches. Proteins were mapped to gene symbols for compatibility with standard pathway databases.\n\n")
cat("**Key metrics:**\n\n")
cat("- **NES (Normalized Enrichment Score)**: Positive NES = enrichment in up-regulated proteins; negative NES = enrichment in down-regulated proteins.\n")
cat("- **padj**: Benjamini-Hochberg adjusted p-value. Pathways with padj < 0.05 are significant.\n")
cat("- **size**: Number of mapped proteins in the pathway.\n\n")
```

## fGSEA Results {.tabset}

```{r fgsea-caption, eval=has_enrichment, results='asis'}
cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Fast Gene Set Enrichment Analysis (fGSEA) results. NES &gt; 0 indicates enrichment among up-regulated proteins; NES &lt; 0 indicates enrichment among down-regulated proteins. Pathways with padj &lt; 0.05 are shown.</em></p>\n")
```

```{r fgsea-tables, eval=has_enrichment, results='asis'}
fgsea_files <- list.files(enrich_dir, pattern = "pathway_.*_fgsea\\.csv$", full.names = TRUE)
if (length(fgsea_files) > 0) {
    for (ff in fgsea_files) {
        label <- gsub("^pathway_|_fgsea\\.csv$", "", basename(ff))
        label_clean <- gsub("_", " ", label)

        fgsea_df <- read.csv(ff, stringsAsFactors = FALSE)
        sig <- fgsea_df[!is.na(fgsea_df$padj) & fgsea_df$padj < 0.05, ]

        cat(sprintf("\n### %s {.tabset}\n\n", label_clean))

        if (nrow(sig) > 0) {
            sig <- sig[order(sig$padj), ]
            if ("pathway_name" %in% names(sig)) {
                show_cols <- intersect(c("pathway_name", "pval", "padj", "NES", "size"), names(sig))
            } else {
                show_cols <- intersect(c("pathway", "pval", "padj", "NES", "size"), names(sig))
            }
            sig_show <- sig[, show_cols]
            sig_show$Direction <- ifelse(sig_show$NES > 0, "Up-regulated", "Down-regulated")
            n_up <- sum(sig_show$Direction == "Up-regulated")
            n_down <- sum(sig_show$Direction == "Down-regulated")

            cat(sprintf("\n#### All (%d pathways)\n\n", nrow(sig_show)))
            cat(sprintf("**Up-regulated: %d** | **Down-regulated: %d**\n\n", n_up, n_down))
            print(htmltools::tagList(make_dt(sig_show, paste("All significant pathways -", label_clean))))
            cat("\n\n")

            if (n_up > 0) {
                cat(sprintf("\n#### Up-regulated (%d)\n\n", n_up))
                sig_up <- sig_show[sig_show$Direction == "Up-regulated", ]
                print(htmltools::tagList(make_dt(sig_up[order(sig_up$padj), ], paste("Up-regulated -", label_clean))))
                cat("\n\n")
            }

            if (n_down > 0) {
                cat(sprintf("\n#### Down-regulated (%d)\n\n", n_down))
                sig_down <- sig_show[sig_show$Direction == "Down-regulated", ]
                print(htmltools::tagList(make_dt(sig_down[order(sig_down$padj), ], paste("Down-regulated -", label_clean))))
                cat("\n\n")
            }
        } else {
            cat("No significant pathways (padj < 0.05) for this contrast/database.\n\n")
        }
    }
} else if (has_enrichment) {
    cat("No fGSEA result files found.\n")
}
```

## ORA Results {.tabset}

Over-Representation Analysis (ORA) tests whether differentially expressed proteins are enriched in specific pathways compared to the background proteome. Unlike fGSEA, ORA uses a binary classification (significant vs. not significant) and applies a hypergeometric test.

```{r ora-caption, eval=has_enrichment, results='asis'}
cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Over-Representation Analysis results. Pathways with padj &lt; 0.05 are shown, ranked by significance.</em></p>\n")
```

```{r ora-tables, eval=has_enrichment, results='asis'}
ora_files <- list.files(enrich_dir, pattern = "pathway_.*_ora.*\\.csv$", full.names = TRUE)
if (length(ora_files) > 0) {
    for (of in ora_files) {
        label <- gsub("^pathway_|_ora(_up|_down)?\\.csv$", "", basename(of))
        label_clean <- gsub("_", " ", label)

        ora_df <- read.csv(of, stringsAsFactors = FALSE)
        # Support both column names: padj (our pipeline) and p.adjust (clusterProfiler)
        padj_col <- if ("padj" %in% names(ora_df)) "padj" else "p.adjust"
        sig <- ora_df[!is.na(ora_df[[padj_col]]) & ora_df[[padj_col]] < 0.05, ]

        cat(sprintf("\n### %s\n\n", label_clean))

        if (nrow(sig) > 0) {
            sig <- sig[order(sig[[padj_col]]), ]
            # Support both clusterProfiler format (Description) and our pipeline format (pathway_name)
            if ("pathway_name" %in% names(sig)) {
                show_cols <- intersect(c("pathway_name", "size", "overlap", "pvalue", "padj", "odds_ratio"), names(sig))
            } else {
                show_cols <- intersect(c("Description", "GeneRatio", "BgRatio", "pvalue", "p.adjust", "Count"), names(sig))
            }
            if (length(show_cols) > 0) {
                print(htmltools::tagList(make_dt(head(sig[, show_cols], 30), paste("ORA -", label_clean))))
            }
        } else {
            cat("No significant pathways (padj < 0.05).\n\n")
        }
    }
}
```

```{r clustered-setup, include=FALSE}
clustered_dir <- file.path(enrich_dir, "clustered")
has_clustered <- dir.exists(clustered_dir) && length(list.files(clustered_dir, pattern = "\\.csv$")) > 0
```

`r if (!has_clustered) "<!--"`

## Clustered Enrichment (Redundancy-Reduced) {.tabset}

Enrichment results often contain many similar terms (e.g., "ribosome biogenesis", "large ribosomal subunit assembly", "small ribosomal subunit assembly"). Semantic clustering groups related terms into clusters and selects a single representative per cluster, revealing a broader functional landscape. For GO terms, clustering uses semantic similarity ([rrvgo](https://bioconductor.org/packages/rrvgo/)); for KEGG/custom databases, it uses Jaccard similarity on shared genes.

```{r clustered-method-note, eval=has_clustered, results='asis'}
cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Enrichment terms clustered by semantic similarity (GO) or gene overlap (KEGG/custom). Each row represents a cluster of related terms. The representative term is the most significant member. \"Members\" lists other terms merged into the cluster.</em></p>\n")
```

```{r clustered-tables, eval=has_clustered, results='asis'}
clustered_files <- list.files(clustered_dir, pattern = "\\.csv$", full.names = TRUE)
for (cf in clustered_files) {
    label <- gsub("^clustered_|\\.csv$", "", basename(cf))
    label_clean <- gsub("_", " ", label)

    cl_df <- read.csv(cf, stringsAsFactors = FALSE)
    if (nrow(cl_df) == 0) next

    cat(sprintf("\n### %s\n\n", label_clean))
    cat(sprintf("**%d clusters** from %d enriched terms\n\n",
                nrow(cl_df), sum(cl_df$n_members)))

    # Select display columns based on what's available (fgsea vs ORA)
    has_nes <- "NES" %in% names(cl_df) && !all(is.na(cl_df$NES))
    if (has_nes) {
        show_cols <- intersect(
            c("representative_term", "n_members", "padj", "NES", "size", "member_terms"),
            names(cl_df))
    } else {
        show_cols <- intersect(
            c("representative_term", "n_members", "padj", "pvalue", "size", "member_terms"),
            names(cl_df))
    }

    display_df <- cl_df[, show_cols, drop = FALSE]
    # Rename for clarity
    names(display_df)[names(display_df) == "representative_term"] <- "Representative Term"
    names(display_df)[names(display_df) == "n_members"] <- "# Merged"
    names(display_df)[names(display_df) == "member_terms"] <- "Clustered Terms"

    print(htmltools::tagList(make_dt(display_df, paste("Clustered -", label_clean))))
    cat("\n\n")
}
```

### Clustered Dotplots {.tabset}

```{r clustered-dotplots-setup, include=FALSE}
clustered_plot_dir <- file.path(enrich_dir, "clustered", "plots")
has_clustered_plots <- dir.exists(clustered_plot_dir) && length(list.files(clustered_plot_dir, pattern = "\\.png$")) > 0
```

```{r clustered-dotplots, eval=has_clustered_plots, results='asis'}
cl_pngs <- list.files(clustered_plot_dir, pattern = "\\.png$", full.names = TRUE)
for (pp in cl_pngs) {
    label_raw <- gsub("\\.png$", "", basename(pp))
    label <- gsub("_", " ", label_raw)
    cat(sprintf("\n#### %s\n\n", label))
    cat(sprintf("![%s](%s)\n\n", label, pp))
    figure_legend(sprintf("Clustered dotplot for %s. Each dot represents a cluster of semantically similar terms. Dot size reflects the number of merged terms; color intensity reflects significance. Numbers in parentheses indicate cluster size.", label))
}
```

```{r clustered-dotplots-skip, eval=has_clustered && !has_clustered_plots, results='asis'}
cat("Clustered dotplots could not be generated.\n")
```

`r if (!has_clustered) "-->"`

## Pathway Dotplots {.tabset}

```{r pathway-dotplots, eval=has_pathway_plots, results='asis'}
plot_pngs <- list.files(enrich_plot_dir, pattern = "\\.png$", full.names = TRUE)
if (length(plot_pngs) > 0) {
    for (pp in plot_pngs) {
        label_raw <- gsub("^pathway_|\\.png$", "", basename(pp))
        label <- gsub("_", " ", label_raw)
        cat(sprintf("\n### %s\n\n", label))
        cat(sprintf("![%s](%s)\n\n", label, pp))
        figure_legend(sprintf("Dotplot of top enriched pathways for %s. Dot size represents the number of pathway members; color intensity reflects statistical significance (adjusted p-value).", label))
        render_commentary(paste0("pathway_", label_raw))
    }
}
```

## Leading Edge Genes {.tabset}

Leading edge genes are the core subset of pathway members that drive the enrichment signal. They represent the genes most responsible for a pathway's statistical significance and are candidates for validation experiments.

```{r leading-edge-caption, eval=has_enrichment, results='asis'}
cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Leading edge proteins for the top 10 most significant pathways per contrast. These proteins drive the enrichment signal and represent the strongest candidates for follow-up.</em></p>\n")
```

```{r leading-edge, eval=has_enrichment, results='asis'}
fgsea_files <- list.files(enrich_dir, pattern = "pathway_.*_fgsea\\.csv$", full.names = TRUE)
if (length(fgsea_files) > 0) {
    for (ff in fgsea_files) {
        label <- gsub("^pathway_|_fgsea\\.csv$", "", basename(ff))
        label_clean <- gsub("_", " ", label)
        cat(sprintf("\n### %s\n\n", label_clean))

        fgsea_df <- read.csv(ff, stringsAsFactors = FALSE)
        sig <- fgsea_df[!is.na(fgsea_df$padj) & fgsea_df$padj < 0.05, ]

        if (nrow(sig) > 0 && "leadingEdge" %in% names(sig)) {
            sig <- head(sig[order(sig$padj), ], 10)
            le_list <- list()
            for (i in seq_len(nrow(sig))) {
                pathway_name <- if ("pathway_name" %in% names(sig)) sig$pathway_name[i] else sig$pathway[i]
                le_genes <- strsplit(sig$leadingEdge[i], ",")[[1]]
                le_list[[i]] <- data.frame(
                    Pathway = pathway_name,
                    NES = round(sig$NES[i], 2),
                    padj = signif(sig$padj[i], 3),
                    `Leading Edge Proteins` = paste(head(le_genes, 15), collapse = ", "),
                    check.names = FALSE,
                    stringsAsFactors = FALSE
                )
            }
            le_df <- do.call(rbind, le_list)
            print(htmltools::tagList(make_dt(le_df, paste("Leading Edge Proteins -", label_clean))))
            cat("\n\n")
        } else {
            cat("No significant pathways with leading edge information.\n\n")
        }
    }
}
```

## Pathway Volcano {.tabset}

```{r pathway-volcano-setup, include=FALSE}
pw_volcano_files <- list.files(enrich_dir, pattern = "^pathway_volcano_data_.*\\.csv$", full.names = TRUE)
has_pathway_volcano <- length(pw_volcano_files) > 0
if (has_pathway_volcano) {
    # Verify at least one file has a populated pathways column
    has_pathway_volcano <- any(vapply(pw_volcano_files, function(f) {
        df <- read.csv(f, stringsAsFactors = FALSE, nrows = 5)
        "pathways" %in% colnames(df) && any(nzchar(df$pathways), na.rm = TRUE)
    }, logical(1)))
}
```

```{r pathway-volcano-skip, eval=!has_pathway_volcano, results='asis'}
cat("No pathway volcano data available. Enable with `pathway_volcano: yes` in the pathway configuration and ensure pathway analysis produces significant results.\n")
```

```{r pathway-volcano-legend, eval=has_pathway_volcano, results='asis'}
figure_legend("Pathway volcano plots overlay enriched pathway membership onto the standard DE volcano plot. Select a pathway from the dropdown to highlight its member proteins in orange. Non-member proteins are dimmed for contrast.")
```

```{r pathway-volcano-plots, eval=has_pathway_volcano, results='asis'}
for (vf in pw_volcano_files) {
    contrast_name <- gsub("^pathway_volcano_data_|\\.csv$", "", basename(vf))
    contrast_clean <- gsub("_", " ", contrast_name)
    clean_id <- gsub("[^a-zA-Z0-9]", "_", contrast_name)

    volcano_df <- read.csv(vf, stringsAsFactors = FALSE)
    if (!"pathways" %in% colnames(volcano_df)) next

    # Cap -log10(p) at 99.5th percentile to keep plot readable
    yvals <- volcano_df$neg_log10_pvalue[is.finite(volcano_df$neg_log10_pvalue)]
    ycap <- max(quantile(yvals, 0.995, na.rm = TRUE), 10)
    volcano_df$neg_log10_pvalue <- pmin(volcano_df$neg_log10_pvalue, ycap)

    cat(sprintf("\n### %s\n\n", contrast_clean))

    # Collect fGSEA info for NES / padj display
    fgsea_files_for_contrast <- list.files(enrich_dir,
        pattern = paste0("pathway_", contrast_name, "_.*_fgsea\\.csv$"),
        full.names = TRUE)
    pathway_info <- list()
    for (ff in fgsea_files_for_contrast) {
        fdf <- read.csv(ff, stringsAsFactors = FALSE)
        sig <- fdf[!is.na(fdf$padj) & fdf$padj < 0.05, , drop = FALSE]
        for (r in seq_len(nrow(sig))) {
            # Use pathway_name (human-readable) to match the volcano data keys
            pname <- if ("pathway_name" %in% names(sig) && nzchar(sig$pathway_name[r] %||% "")) {
                sig$pathway_name[r]
            } else {
                sig$pathway[r]
            }
            pathway_info[[pname]] <- list(
                NES = round(sig$NES[r], 3),
                padj = signif(sig$padj[r], 3),
                size = sig$size[r]
            )
        }
    }

    # Assign direction colors
    padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- log2(prot_cfg$de$linear_fc_cutoff %||% 1.5)
    volcano_df$Direction <- ifelse(
        is.na(volcano_df$padj) | volcano_df$padj >= padj_cut, "NS",
        ifelse(volcano_df$log2FC > lfc_cut, "Up",
               ifelse(volcano_df$log2FC < -lfc_cut, "Down", "NS")))

    dir_colors <- c("Up" = "#e74c3c", "Down" = "#3498db", "NS" = "#bdc3c7")

    p <- plotly::plot_ly(
        data = volcano_df,
        x = ~log2FC,
        y = ~neg_log10_pvalue,
        type = "scatter",
        mode = "markers",
        color = ~Direction,
        colors = dir_colors,
        marker = list(size = 5, opacity = 0.7),
        text = ~paste0("Protein: ", FeatureID,
                        "<br>log2FC: ", round(log2FC, 3),
                        "<br>padj: ", signif(padj, 3),
                        "<br>Pathways: ", ifelse(nzchar(pathways), pathways, "none")),
        hoverinfo = "text",
        customdata = ~FeatureID,
        source = paste0("pw_volcano_", clean_id)
    ) |>
        plotly::layout(
            title = list(text = paste("Pathway Volcano -", contrast_clean), font = list(size = 14)),
            xaxis = list(title = "log2 Fold Change"),
            yaxis = list(title = "-log10(p-value)", rangemode = "tozero"),
            shapes = list(
                list(type = "line", x0 = -lfc_cut, x1 = -lfc_cut, y0 = 0, y1 = 1,
                     yref = "paper", line = list(dash = "dash", color = "grey50")),
                list(type = "line", x0 = lfc_cut, x1 = lfc_cut, y0 = 0, y1 = 1,
                     yref = "paper", line = list(dash = "dash", color = "grey50")),
                list(type = "line", x0 = 0, x1 = 1, xref = "paper",
                     y0 = -log10(padj_cut), y1 = -log10(padj_cut),
                     line = list(dash = "dash", color = "grey50"))
            )
        ) |>
        plotly::config(
            toImageButtonOptions = list(format = "svg", filename = paste0("pathway_volcano_", clean_id),
                                         width = 1200, height = 900, scale = 3),
            modeBarButtonsToAdd = list(
                list(name = "Download PNG (300 dpi)", icon = list(
                    path = "M12 2C6.48 2 2 6.48 2 12s4.48 10 10 10 10-4.48 10-10S17.52 2 12 2zm-1 14H8l4-4 4 4h-3v3h-2v-3z",
                    width = 24, height = 24
                ), click = htmlwidgets::JS(sprintf(
                    "function(gd) { Plotly.downloadImage(gd, {format: 'png', width: 2400, height: 1800, scale: 3, filename: 'pathway_volcano_%s'}); }",
                    clean_id
                )))
            )
        )

    p$elementId <- paste0("pw_volcano_plot_", clean_id)

    # Build pathway→genes mapping from the pathways column
    all_pathways <- unique(unlist(strsplit(volcano_df$pathways[nzchar(volcano_df$pathways)], "; ")))
    all_pathways <- sort(all_pathways[nzchar(all_pathways)])

    pw_genes_map <- list()
    for (pw in all_pathways) {
        members <- as.character(volcano_df$FeatureID[grepl(pw, volcano_df$pathways, fixed = TRUE)])
        pw_genes_map[[pw]] <- I(members)
    }

    pw_genes_json <- jsonlite::toJSON(pw_genes_map, auto_unbox = TRUE)
    pw_info_json <- jsonlite::toJSON(pathway_info, auto_unbox = TRUE)

    # Build dropdown options
    options_html <- '<option value="">None (show all)</option>'
    for (pw in all_pathways) {
        pw_escaped <- gsub("\"", "&quot;", pw)
        info <- pathway_info[[pw]]
        label <- if (!is.null(info)) {
            sprintf("%s (NES=%.2f, padj=%.3g)", pw, info$NES, info$padj)
        } else { pw }
        label_escaped <- gsub("\"", "&quot;", label)
        options_html <- paste0(options_html,
            sprintf('<option value="%s">%s</option>', pw_escaped, label_escaped))
    }

    cat(sprintf('
<div style="margin-bottom: 10px;">
<label for="pw_select_%s"><strong>Highlight pathway:</strong></label>
<select id="pw_select_%s" style="padding: 6px; border-radius: 4px; border: 1px solid #ccc; margin-left: 8px; max-width: 600px;">
%s
</select>
<div id="pw_info_%s" style="margin-top: 6px; font-size: 13px; color: #555;"></div>
<div id="pw_legend_%s" style="margin-top: 4px; font-size: 13px;"></div>
</div>
', clean_id, clean_id, options_html, clean_id, clean_id))

    print(htmltools::tagList(p))

    cat(sprintf('
<script>
(function() {
    var cid = "%s";
    var plotId = "pw_volcano_plot_" + cid;
    var pwGenes = %s;
    var pwInfo = %s;
    var origColors = null;
    var origOpacities = null;

    function captureOriginal() {
        var plotDiv = document.getElementById(plotId);
        if (!plotDiv || !plotDiv.data) return;
        origColors = [];
        origOpacities = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var mc = plotDiv.data[t].marker;
            origColors.push(mc ? (mc.color || null) : null);
            origOpacities.push(mc ? (mc.opacity || 0.7) : 0.7);
        }
    }

    function onSelect() {
        var plotDiv = document.getElementById(plotId);
        if (!plotDiv || !plotDiv.data) return;
        if (!origColors) captureOriginal();

        var pw = document.getElementById("pw_select_" + cid).value;
        var infoDiv = document.getElementById("pw_info_" + cid);
        var legendDiv = document.getElementById("pw_legend_" + cid);

        // Freeze axes so restyle cannot trigger auto-range
        var xRange = plotDiv.layout.xaxis.range ? plotDiv.layout.xaxis.range.slice() : null;
        var yRange = plotDiv.layout.yaxis.range ? plotDiv.layout.yaxis.range.slice() : null;
        if (xRange && yRange) {
            Plotly.relayout(plotDiv, {
                "xaxis.autorange": false, "xaxis.range": xRange,
                "yaxis.autorange": false, "yaxis.range": yRange
            });
        }

        if (!pw) {
            // Restore original
            for (var t = 0; t < plotDiv.data.length; t++) {
                Plotly.restyle(plotDiv, {
                    "marker.color": [origColors[t]],
                    "marker.opacity": [origOpacities[t]],
                    "marker.size": [5]
                }, [t]);
            }
            infoDiv.innerHTML = "";
            if (legendDiv) legendDiv.innerHTML = "";
            return;
        }

        var genes = pwGenes[pw] || [];
        if (typeof genes === "string") genes = [genes];
        var genesSet = {};
        for (var g = 0; g < genes.length; g++) genesSet[genes[g]] = true;

        // Show pathway info
        var info = pwInfo[pw];
        if (info) {
            infoDiv.innerHTML = "<strong>" + pw + "</strong> &mdash; NES: " +
                info.NES + " | padj: " + info.padj + " | Members in data: " + genes.length;
        } else {
            infoDiv.innerHTML = "<strong>" + pw + "</strong> &mdash; Members in data: " + genes.length;
        }
        if (legendDiv) {
            legendDiv.innerHTML = "<font color=#e67e22>&#9679;</font> Pathway member &nbsp;" +
                "<font color=#d5d8dc>&#9679;</font> Non-member";
        }

        // Color traces
        for (var t = 0; t < plotDiv.data.length; t++) {
            var cd = plotDiv.data[t].customdata;
            if (!cd) { continue; }
            var colors = [];
            var opacities = [];
            var sizes = [];
            for (var i = 0; i < cd.length; i++) {
                var fid = Array.isArray(cd[i]) ? cd[i][0] : cd[i];
                if (genesSet[fid]) {
                    colors.push("#e67e22");
                    opacities.push(0.9);
                    sizes.push(8);
                } else {
                    colors.push("#d5d8dc");
                    opacities.push(0.15);
                    sizes.push(4);
                }
            }
            Plotly.restyle(plotDiv, {
                "marker.color": [colors],
                "marker.opacity": [opacities],
                "marker.size": [sizes]
            }, [t]);
        }
    }

    // Attach event after plot renders
    var attempts = 0;
    var timer = setInterval(function() {
        var plotDiv = document.getElementById(plotId);
        if ((plotDiv && plotDiv.data) || attempts > 50) {
            clearInterval(timer);
            if (plotDiv && plotDiv.data) {
                captureOriginal();
                document.getElementById("pw_select_" + cid).addEventListener("change", onSelect);
            }
        }
        attempts++;
    }, 200);
})();
</script>
', clean_id, pw_genes_json, pw_info_json))

    cat("\n\n")
}
```

---

# PPI Network Analysis

```{r ppi-setup, include=FALSE}
has_ppi <- dir.exists(ppi_dir) && (
    length(list.files(ppi_dir, pattern = "\\.csv$")) > 0 ||
    length(list.files(ppi_plot_dir, pattern = "\\.png$")) > 0
)
ppi_enabled <- isTRUE(prot_cfg$ppi$enabled)
```

```{r ppi-skip, eval=!has_ppi, results='asis'}
if (!ppi_enabled) {
    cat("PPI network analysis was disabled. Enable with `ppi.enabled: true`.\n")
} else {
    cat("PPI analysis was enabled but produced no results.\n")
}
```

```{r ppi-explanation, eval=has_ppi, results='asis'}
cat("Protein-protein interaction networks map physical and functional associations between differentially expressed proteins. Network topology reveals hub proteins (highly connected), bottleneck proteins (bridging modules), and functional communities that may correspond to coordinated biological processes.\n\n")
```

```{r ppi-summary, eval=has_ppi, results='asis'}
cat("Protein-protein interaction networks were constructed using the STRING database.\n\n")

# Load summary if available
ppi_summary_file <- file.path(ppi_dir, "ppi_summary.csv")
if (file.exists(ppi_summary_file)) {
    ppi_summary <- read.csv(ppi_summary_file, stringsAsFactors = FALSE)
    make_dt(ppi_summary, caption = "PPI Network Summary")
}
```

## Network Plots {.tabset}

```{r ppi-plots, eval=has_ppi, results='asis'}
# Load STRING info for building web links
string_info_file <- file.path(ppi_dir, "string_info.csv")
string_proteins <- NULL
string_species <- 9606
if (file.exists(string_info_file)) {
    si <- read.csv(string_info_file, stringsAsFactors = FALSE)
    string_proteins <- si$protein
    string_species <- si$species_id[1]
}

if (dir.exists(ppi_plot_dir)) {
    ppi_pngs <- list.files(ppi_plot_dir, pattern = "\\.png$", full.names = TRUE)
    if (length(ppi_pngs) > 0) {
        for (pp in ppi_pngs) {
            label_raw <- gsub("\\.png$", "", basename(pp))
            label <- gsub("_", " ", label_raw)
            cat(sprintf("\n### %s\n\n", label))
            cat(sprintf("![%s](%s)\n\n", label, pp))
            figure_legend(sprintf("PPI network visualization: %s. Nodes represent proteins; edges represent known or predicted interactions from the STRING database. Node size and color may reflect connectivity or expression change.", label))
            cat('<div style="margin: 8px 0; font-size: 13px;">
  <strong>Node color:</strong>
  <span style="color: #3498db;">&#9679;</span> Down-regulated &nbsp;
  <span style="color: #ccc;">&#9679;</span> No change &nbsp;
  <span style="color: #e74c3c;">&#9679;</span> Up-regulated &nbsp;
  <em>(log2FC gradient, capped at &plusmn;2)</em>
</div>\n')
            # Add "View on STRING" link
            if (!is.null(string_proteins) && length(string_proteins) > 0) {
                string_url <- sprintf(
                    "https://string-db.org/cgi/network?identifiers=%s&species=%s",
                    paste(string_proteins, collapse = "%%0d"), string_species
                )
                cat(sprintf('<a href="%s" target="_blank" style="display: inline-block; margin: 6px 0 12px 0; padding: 6px 14px; background: #2c7fb8; color: white; border-radius: 4px; text-decoration: none; font-size: 13px;">View on STRING &rarr;</a>\n', string_url))
            }
            render_commentary(paste0("ppi_", label_raw))
        }
    } else {
        cat("No PPI plots found.\n")
    }
}
```

## Network Statistics

```{r ppi-topology, eval=has_ppi, results='asis'}
topo_file <- file.path(ppi_dir, "network_topology.csv")
if (file.exists(topo_file)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Global network topology metrics including node count, edge count, density, clustering coefficient, and characteristic path length.</em></p>\n")
    topo <- read.csv(topo_file, stringsAsFactors = FALSE)
    make_dt(topo, caption = "Network Topology Statistics")
}
```

## Hub Proteins

```{r ppi-hubs, eval=has_ppi, results='asis'}
hubs_file <- file.path(ppi_dir, "hub_proteins.csv")
if (file.exists(hubs_file)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Hub proteins ranked by network connectivity (degree). Hub proteins are highly connected nodes that may play central roles in the biological process under study.</em></p>\n")
    hubs <- read.csv(hubs_file, stringsAsFactors = FALSE)
    make_dt(hubs, caption = "Hub Proteins (top connectivity)")
}
```

## Communities

```{r ppi-communities, eval=has_ppi, results='asis'}
comm_file <- file.path(ppi_dir, "communities.csv")
if (file.exists(comm_file)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Network communities detected by graph clustering algorithms. Communities represent densely connected subnetworks that may correspond to functional protein modules or complexes.</em></p>\n")
    comm <- read.csv(comm_file, stringsAsFactors = FALSE)
    make_dt(comm, caption = "Network Communities")
}
```

```{r string-enrich-setup, include=FALSE}
string_enrich_dir <- file.path(ppi_dir, "string_enrichment")
has_string_enrich <- dir.exists(string_enrich_dir) &&
    length(list.files(string_enrich_dir, pattern = "\\.csv$")) > 0
```

`r if (!has_string_enrich) "<!--"`

## STRING Functional Enrichment {.tabset}

STRING provides functional enrichment analysis across multiple annotation categories — including disease associations, tissue expression, subcellular compartments, and literature references — directly from the protein interaction network.

```{r string-enrich-tables, eval=has_string_enrich, results='asis'}
string_files <- list.files(string_enrich_dir, pattern = "^string_.*\\.csv$", full.names = TRUE)
# Exclude the full combined file
string_files <- string_files[!grepl("_full\\.csv$", string_files)]

if (length(string_files) > 0) {
    # Priority order for display
    priority_cats <- c("DISEASES", "TISSUES", "COMPARTMENTS", "Reactome",
                       "KEGG", "Process", "Component", "Function",
                       "PMID", "InterPro", "Pfam", "Keyword")

    # Sort files by priority
    file_labels <- gsub("^string_|\\.csv$", "", basename(string_files))
    file_order <- match(file_labels, priority_cats)
    file_order[is.na(file_order)] <- 100
    string_files <- string_files[order(file_order)]

    for (sf in string_files) {
        label <- gsub("^string_|\\.csv$", "", basename(sf))
        label_clean <- gsub("_", " ", label)

        df <- tryCatch(read.csv(sf, stringsAsFactors = FALSE), error = function(e) NULL)
        if (is.null(df) || nrow(df) == 0) next

        # Filter to FDR < 0.05
        fdr_col <- intersect(c("fdr", "FDR", "p_value", "pvalue"), colnames(df))
        if (length(fdr_col) > 0) {
            sig <- df[!is.na(df[[fdr_col[1]]]) & df[[fdr_col[1]]] < 0.05, ]
        } else {
            sig <- df
        }
        if (nrow(sig) == 0) next

        cat(sprintf("\n### %s (%d)\n\n", label_clean, nrow(sig)))

        # Select display columns
        show_cols <- intersect(
            c("term", "description", "number_of_genes", "fdr",
              "inputGenes", "preferredNames", "category"),
            colnames(sig)
        )
        if (length(show_cols) == 0) show_cols <- colnames(sig)[1:min(6, ncol(sig))]

        print(htmltools::tagList(make_dt(
            head(sig[, show_cols, drop = FALSE], 30),
            paste("STRING", label_clean)
        )))
        cat("\n\n")
    }
} else {
    cat("No STRING functional enrichment results available.\n")
}
```

`r if (!has_string_enrich) "-->"`

---

```{r domain-setup, include=FALSE}
domain_dir <- file.path(prot_dir, "domain_analysis")
has_domains <- dir.exists(domain_dir) && (
    length(list.files(domain_dir, pattern = "\\.csv$")) > 0 ||
    length(list.files(file.path(domain_dir, "plots"), pattern = "\\.png$")) > 0
)
```

`r if (!has_domains) "<!--"`

# Protein Domain & Functional Feature Analysis

```{r domain-intro, eval=has_domains, results='asis'}
cat("Protein domain and functional feature analysis identifies structural and functional motifs enriched among differentially expressed proteins. This reveals whether specific protein families, binding domains, or post-translational modification sites are over-represented in the DE set — complementing pathway-level enrichment with protein-level structural insights.\n\n")
```

## Domain Enrichment {.tabset}

```{r domain-enrichment, eval=has_domains, results='asis'}
domain_enrich_file <- file.path(domain_dir, "domain_enrichment.csv")
if (file.exists(domain_enrich_file)) {
    enr <- read.csv(domain_enrich_file, stringsAsFactors = FALSE)
    sig_enr <- enr[!is.na(enr$padj) & enr$padj < 0.05, ]

    if (nrow(sig_enr) > 0) {
        # Split by database (InterPro vs Pfam)
        for (db in unique(sig_enr$database)) {
            db_df <- sig_enr[sig_enr$database == db, ]
            db_df <- db_df[order(db_df$padj), ]

            cat(sprintf("\n### %s (%d significant)\n\n", db, nrow(db_df)))
            show_cols <- intersect(
                c("domain_id", "domain_description", "domain_size", "overlap",
                  "pvalue", "padj", "odds_ratio", "genes"),
                colnames(db_df)
            )
            print(htmltools::tagList(make_dt(
                head(db_df[, show_cols, drop = FALSE], 30),
                paste("Enriched", db, "Domains")
            )))
            cat("\n\n")
        }
    } else {
        cat("\n### Results\n\nNo significantly enriched domains (padj < 0.05) detected.\n\n")
    }
} else {
    cat("\n### Results\n\nDomain enrichment results not available.\n\n")
}
```

## Domain Enrichment Plot

```{r domain-dotplot, eval=has_domains, results='asis'}
dotplot_file <- file.path(domain_dir, "plots", "domain_enrichment_dotplot.png")
if (file.exists(dotplot_file)) {
    cat(sprintf("![Domain Enrichment](%s)\n\n", dotplot_file))
    figure_legend("Dotplot of significantly enriched protein domains among DE proteins. Dot size reflects the number of DE proteins carrying the domain; color distinguishes InterPro from Pfam annotations.")
}
```

## UniProt Functional Features {.tabset}

UniProt functional annotations include binding sites (ligand/metal/DNA binding), active sites (catalytic residues), post-translational modifications (phosphorylation, glycosylation), and functional domains. These annotations highlight specific molecular functions affected by the observed expression changes.

```{r uniprot-features, eval=has_domains, results='asis'}
features_file <- file.path(domain_dir, "uniprot_features.csv")
if (file.exists(features_file)) {
    feat <- read.csv(features_file, stringsAsFactors = FALSE)
    if (nrow(feat) > 0 && "feature_type" %in% colnames(feat)) {
        for (ft in unique(feat$feature_type)) {
            ft_df <- feat[feat$feature_type == ft, ]
            cat(sprintf("\n### %s (%d)\n\n", ft, nrow(ft_df)))
            print(htmltools::tagList(make_dt(
                head(ft_df, 50),
                paste("UniProt", ft, "Annotations")
            )))
            cat("\n\n")
        }
    } else {
        cat("\n### Results\n\nNo UniProt functional features retrieved.\n\n")
    }
} else {
    cat("\n### Results\n\nUniProt feature data not available.\n\n")
}
```

## Feature Type Summary

```{r feature-barchart, eval=has_domains, results='asis'}
barchart_file <- file.path(domain_dir, "plots", "feature_type_barchart.png")
if (file.exists(barchart_file)) {
    cat(sprintf("![Feature Types](%s)\n\n", barchart_file))
    figure_legend("Distribution of UniProt functional feature types among DE proteins. Shows the relative prevalence of binding sites, active sites, PTMs, domains, and other structural/functional annotations.")
}
```

`r if (!has_domains) "-->"`

```{r adv-stats-setup, include=FALSE}
has_adv_stats <- dir.exists(adv_stats_dir) && (
    length(list.files(adv_stats_dir, pattern = "\\.csv$")) > 0 ||
    length(list.files(adv_stats_plot_dir, pattern = "\\.png$")) > 0
)
adv_stats_enabled <- isTRUE(prot_cfg$advanced_stats$enabled)
```

`r if (!has_adv_stats) "<!--"`

---

# Advanced Statistics

```{r adv-stats-intro, eval=has_adv_stats, results='asis'}
cat("Advanced statistical analyses provide supplementary metrics including effect size confidence intervals, robust regression estimates, and statistical power calculations. These complement the primary differential expression results by quantifying uncertainty and detection sensitivity.\n\n")
```

## Effect Sizes

```{r adv-stats-effect-sizes, eval=has_adv_stats, results='asis'}
es_file <- file.path(adv_stats_dir, "effect_sizes.csv")
if (file.exists(es_file)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Effect size estimates with bootstrap confidence intervals for the top 100 most significant proteins. Confidence intervals quantify uncertainty in fold-change estimates.</em></p>\n")
    es <- read.csv(es_file, stringsAsFactors = FALSE)
    make_dt(head(es, 100), caption = "Effect Sizes (top 100 by significance)")
}
```

## Plots {.tabset}

```{r adv-stats-plots, eval=has_adv_stats, results='asis'}
if (dir.exists(adv_stats_plot_dir)) {
    adv_pngs <- list.files(adv_stats_plot_dir, pattern = "\\.png$", full.names = TRUE)
    if (length(adv_pngs) > 0) {
        for (ap in adv_pngs) {
            label_raw <- gsub("\\.png$", "", basename(ap))
            label <- gsub("_", " ", label_raw)
            cat(sprintf("\n### %s\n\n", label))
            cat(sprintf("![%s](%s)\n\n", label, ap))
            figure_legend(sprintf("%s from the advanced statistics module.", label))
            render_commentary(paste0("adv_stats_", label_raw))
        }
    } else {
        cat("No advanced statistics plots found.\n")
    }
}
```

## Power Analysis

```{r adv-stats-power, eval=has_adv_stats, results='asis'}
power_file <- file.path(adv_stats_dir, "power_analysis.csv")
if (file.exists(power_file)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Statistical power analysis estimating the probability of detecting true differential expression at various effect sizes given the current sample size. Power &ge; 0.8 is considered adequate.</em></p>\n")
    power <- read.csv(power_file, stringsAsFactors = FALSE)
    make_dt(power, caption = "Statistical Power Analysis")
}
```

## Multiple Testing Comparison

```{r adv-stats-mtc, eval=has_adv_stats, results='asis'}
mtc_file <- file.path(adv_stats_dir, "mtc_comparison.csv")
if (file.exists(mtc_file)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Comparison of different multiple testing correction methods (BH, Bonferroni, BY, q-value) showing the number of significant proteins at each threshold. BH (Benjamini-Hochberg) is the default method.</em></p>\n")
    mtc <- read.csv(mtc_file, stringsAsFactors = FALSE)
    make_dt(mtc, caption = "Multiple Testing Correction Comparison")
}
```

`r if (!has_adv_stats) "-->"`

---

# Output Files

```{r output-files}
files_tbl <- data.frame(
  File = c(
    "Final_results_ALL_P_0.05.xlsx",
    "Final_results_DE_P_0.05.xlsx",
    "Datasets/",
    "Diagnostic_plots/",
    "Diagnostic_plots/QC_post/"
  ),
  Description = c(
    "Excel workbook with all proteins and DE statistics (p < 0.05)",
    "Excel workbook with DE proteins only (p < 0.05)",
    "Intermediate datasets (expression matrices, DE tables)",
    "QC and diagnostic plots (PCA, density, heatmaps, UMAP)",
    "Post-DE QC plots (volcano, MA, top DE heatmaps)"
  ),
  stringsAsFactors = FALSE
)

# Add enrichment files if they exist
if (has_enrichment) {
    enrich_csvs <- list.files(enrich_dir, pattern = "\\.csv$")
    if (length(enrich_csvs) > 0) {
        enrich_tbl <- data.frame(
            File = paste0("Enrichment/", enrich_csvs),
            Description = paste("Pathway analysis results -",
                                gsub("^pathway_|\\.csv$", "", enrich_csvs)),
            stringsAsFactors = FALSE
        )
        files_tbl <- rbind(files_tbl, enrich_tbl)
    }
}

if (has_ppi) {
    files_tbl <- rbind(files_tbl, data.frame(
        File = "ppi_networks/", Description = "PPI network analysis results and plots",
        stringsAsFactors = FALSE
    ))
}

if (has_adv_stats) {
    files_tbl <- rbind(files_tbl, data.frame(
        File = "advanced_stats/", Description = "Advanced statistics results and plots",
        stringsAsFactors = FALSE
    ))
}

make_dt(files_tbl, "Key Output Files")
```

---

```{r methods-heading, eval=show_methods, results='asis'}
cat("# Methods\n\n")
```

```{r methods-text, eval=show_methods, results='asis'}
imp_method <- prot_cfg$imputation$method %||% "none"
if (imp_method == "perseus") imp_method <- "perseus_like"
de_method <- prot_cfg$de$method %||% "limma"
padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
lfc_cut <- prot_cfg$de$linear_fc_cutoff %||% 1.5
engine <- prot_cfg$engine %||% "unknown"

n_proteins <- if (!is.null(de_data)) nrow(de_data) else "N"
n_de <- "N"
if (!is.null(de_data)) {
    pass_col <- grep("pass_any", names(de_data), value = TRUE)[1]
    if (!is.na(pass_col)) n_de <- sum(de_data[[pass_col]] == TRUE, na.rm = TRUE)
}

cat("**Proteomics Analysis Methods**\n\n")

cat(sprintf(
"Proteomics data were acquired using %s and processed using the multiomics-core pipeline. ", engine
))

if (tolower(imp_method) != "none") {
    cat(sprintf(
    "Missing values were imputed using the %s method. ", imp_method
    ))
}

cat(sprintf(
"Differential expression analysis was performed using %s. ", de_method
))

cat(sprintf(
"Proteins were considered differentially expressed at adjusted p-value <= %.2f and |fold change| >= %.1f. ",
padj_cut, lfc_cut
))

cat(sprintf(
"A total of %s proteins were quantified, with %s identified as differentially expressed in at least one contrast.",
format(as.numeric(n_proteins), big.mark = ","),
format(as.numeric(n_de), big.mark = ",")
))

if (pathway_enabled) {
    pathway_method <- prot_cfg$pathway$method %||% "both"
    pathway_dbs <- paste(prot_cfg$pathway$databases %||% c("GO", "KEGG"), collapse = " and ")
    cat(sprintf(" Pathway enrichment analysis was performed using %s against %s databases.",
                pathway_method, pathway_dbs))
}

if (ppi_enabled) {
    cat(" Protein-protein interaction networks were constructed using the STRING database.")
}

cat("\n\n")

cat("**Copy-ready version:**\n\n")
cat("<div style='background-color: #f5f5f5; padding: 15px; border-radius: 5px; font-size: 0.9em;'>\n")

cat(sprintf(
"Proteomics analysis was performed using the multiomics-core pipeline (v1.0). Data were acquired using %s. %sDifferential expression was assessed with %s (adjusted p <= %.2f, |FC| >= %.1f).",
engine,
if (tolower(imp_method) != "none") sprintf("Missing values were imputed using the %s method. ", imp_method) else "",
de_method, padj_cut, lfc_cut
))

cat("\n</div>\n")
```

```{r methods-statistical-heading, eval=show_methods, results='asis'}
cat("## Statistical Analysis\n\n")
```

```{r methods-statistical-analysis, eval=show_methods, results='asis'}
# Parameterized statistical analysis subsection for the Methods section.
# If report_cfg$stats_section_text is non-empty, print it verbatim;
# otherwise auto-generate from pipeline config values.
stats_section_text <- report_cfg$stats_section_text %||% ""
if (nzchar(trimws(stats_section_text))) {
    cat(trimws(stats_section_text), "\n\n")
} else {
    # Auto-generate from config
    de_method_m <- prot_cfg$de$method %||% "limma"
    imp_method_m <- prot_cfg$imputation$method %||% "none"
    padj_m <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_m <- prot_cfg$de$linear_fc_cutoff %||% 1.5

    test_desc_m <- switch(tolower(de_method_m),
        limma = "Moderated t-tests were performed using the *limma* R/Bioconductor package, which applies empirical Bayes shrinkage of variance estimates to improve statistical power in small-sample proteomics studies.",
        "limma-voom" = "Differential expression was assessed using *limma-voom*, which transforms count data to log-CPM and applies precision-weighted linear modelling.",
        deseq2 = "Differential expression was assessed using Wald tests implemented in *DESeq2*, which fits negative binomial generalized linear models.",
        sprintf("Differential expression was assessed using the *%s* method.", de_method_m)
    )

    cat(test_desc_m, "\n\n")

    if (tolower(imp_method_m) != "none") {
        imp_desc <- switch(tolower(imp_method_m),
            perseus_like = "Missing values were imputed using the Perseus-like method (down-shifted normal distribution for MNAR, KNN for MCAR).",
            minprob = "Missing values were imputed using the MinProb method (random draws from a low-abundance distribution).",
            knn = "Missing values were imputed using K-nearest neighbours (KNN).",
            sprintf("Missing values were imputed using the %s method.", imp_method_m)
        )
        cat(imp_desc, "\n\n")
    }

    cat(sprintf(
        "Multiple testing correction was applied using the Benjamini-Hochberg procedure to control the false discovery rate (FDR). Proteins were considered differentially expressed at adjusted p-value $\\leq$ %s and |linear fold change| $\\geq$ %s.\n\n",
        padj_m, lfc_m
    ))

    cat(sprintf(
        "All analyses were performed in R (version %s) using the multiomics-core pipeline.\n\n",
        paste(R.version$major, R.version$minor, sep = ".")
    ))
}
```

---

```{r session-info-heading, eval=show_session_info, results='asis'}
cat("# Session Info\n\n")
```

```{r session-info, eval=show_session_info, comment=""}
si_file <- file.path(run_dir, "execution_info", "sessionInfo.txt")
if (file.exists(si_file)) {
    si <- readLines(si_file)
    cat(paste(si, collapse = "\n"))
} else {
    cat(capture.output(sessionInfo()), sep = "\n")
}
```

```{r disclaimer-section, results='asis'}
disclaimer_text <- report_cfg$disclaimer_text %||% ""
if (nzchar(trimws(disclaimer_text))) {
    cat("---\n\n# Disclaimer\n\n")
    cat(trimws(disclaimer_text), "\n\n")
}
```

---

<p style="text-align:center; color:grey; font-size:0.9em;">
Report generated by multiomics-core pipeline | `r format(Sys.time(), '%Y-%m-%d %H:%M')`
</p>
