---
title: "`r local({ .c <- yaml::read_yaml(file.path(normalizePath(dirname(knitr::current_input())), 'execution_info', 'config_used.yaml')); pname <- .c$project$name %||% ''; analyst <- gsub('_', ' ', .c$project$analyst %||% ''); parts <- c('Proteomics Analysis Report'); if (nzchar(pname)) parts <- c(parts, pname); if (nzchar(analyst)) parts <- c(parts, analyst); paste(parts, collapse = ' — ') })`"
subtitle: "multiomics-core Pipeline"
author: "Computational analysis: `r yaml::read_yaml(file.path(normalizePath(dirname(knitr::current_input())), 'execution_info', 'config_used.yaml'))$project$analyst`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    fig_width: 10
    fig_height: 7
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%"
)

library(yaml)
library(knitr)
library(DT)
library(plotly)
library(ggplot2)
library(htmltools)
library(jsonlite)

# Load ggVennDiagram if available
has_ggVennDiagram <- FALSE
tryCatch({
    library(ggVennDiagram)
    has_ggVennDiagram <- TRUE
}, error = function(e) {
    message("ggVennDiagram not available: ", e$message)
})

# Helper function for downloadable DT tables
make_dt <- function(df, caption = NULL, pageLength = 15) {
    DT::datatable(
        df,
        caption = caption,
        extensions = 'Buttons',
        options = list(
            scrollX = TRUE,
            pageLength = pageLength,
            dom = 'Bfrtip',
            buttons = list(
                list(extend = 'csv', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data")),
                list(extend = 'excel', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data"))
            )
        ),
        rownames = FALSE
    )
}

# Helper: render figure legend (paper-style caption below figure)
figure_legend <- function(text) {
    cat(sprintf("<p class='figure-legend' style='text-align: center; font-size: 0.9em; color: #555; margin: 4px auto 12px auto; max-width: 88%%; line-height: 1.45; font-style: italic;'>%s</p>\n", text))
}

# Paths
run_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
# Detect if we are already inside the proteomics output dir
if (basename(run_dir) == "proteomics") {
    prot_dir <- run_dir
    config_path <- file.path(dirname(run_dir), "execution_info", "config_used.yaml")
} else {
    prot_dir <- file.path(run_dir, "proteomics")
    config_path <- file.path(run_dir, "execution_info", "config_used.yaml")
}
diag_dir <- file.path(prot_dir, "Diagnostic_plots")
qc_post_dir <- file.path(diag_dir, "QC_post")
datasets_dir <- file.path(prot_dir, "Datasets")
clustering_dir <- file.path(prot_dir, "Clustering")
enrich_dir <- file.path(prot_dir, "Enrichment")
enrich_plot_dir <- file.path(enrich_dir, "plots")
ppi_dir <- file.path(prot_dir, "ppi_networks")
ppi_plot_dir <- file.path(ppi_dir, "plots")
adv_stats_dir <- file.path(prot_dir, "advanced_stats")
adv_stats_plot_dir <- file.path(adv_stats_dir, "plots")
commentary_dir <- file.path(prot_dir, "commentary")

# Load commentary if available
commentary <- NULL
has_commentary <- FALSE
commentary_json <- file.path(commentary_dir, "commentary_all.json")
if (file.exists(commentary_json)) {
    commentary <- tryCatch(
        jsonlite::read_json(commentary_json),
        error = function(e) NULL
    )
    has_commentary <- !is.null(commentary) && length(commentary) > 0
}

# Helper: render commentary block for a figure_id (collapsible)
render_commentary <- function(figure_id) {
    if (!has_commentary) return(invisible(NULL))
    c <- commentary[[figure_id]]
    if (is.null(c)) return(invisible(NULL))

    cat("<details class='commentary-details' style='margin: 4px 0 18px 0;'>\n")
    cat("<summary style='cursor: pointer; color: #4a90d9; font-weight: 500; font-size: 0.93em; padding: 3px 0;'>Show AI Interpretation</summary>\n")

    html <- "<div class='commentary-block' style='background: #f8f9fa; border-left: 4px solid #4a90d9; padding: 12px 16px; margin: 6px 0 4px 0; border-radius: 0 4px 4px 0; font-size: 0.92em;'>\n"
    html <- paste0(html, "<p style='margin: 0 0 8px 0;'><strong>What this shows:</strong> ",
                   htmltools::htmlEscape(c$what_is_this %||% ""), "</p>\n")

    bio_interp <- c$biological_interpretation %||% ""
    if (nzchar(bio_interp)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Interpretation:</strong> ",
                       htmltools::htmlEscape(bio_interp), "</p>\n")
    }

    obs <- c$observations
    if (length(obs) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Key observations:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (o in obs) html <- paste0(html, "<li>", htmltools::htmlEscape(o), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    stat_assess <- c$statistical_assessment %||% ""
    if (nzchar(stat_assess)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Statistical note:</strong> ",
                       htmltools::htmlEscape(stat_assess), "</p>\n")
    }

    issues <- c$issues_checks
    if (length(issues) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Checks:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (issue in issues) html <- paste0(html, "<li>", htmltools::htmlEscape(issue), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    steps <- c$next_steps
    if (length(steps) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Next steps:</strong></p>\n<ul style='margin: 2px 0 4px 0; padding-left: 20px;'>\n")
        for (s in steps) html <- paste0(html, "<li>", htmltools::htmlEscape(s), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    backend <- c$backend %||% "unknown"
    confidence <- c$confidence %||% ""
    backend_label <- if (grepl("claude", backend, ignore.case = TRUE)) {
        "Claude (Anthropic) + K-Dense AI scientific-skills"
    } else if (grepl("openai", backend, ignore.case = TRUE)) {
        "OpenAI + K-Dense AI scientific-skills"
    } else if (backend == "data_driven") {
        "data-driven + K-Dense AI scientific-skills"
    } else {
        backend
    }
    if (nzchar(confidence) && confidence != "none") {
        backend_label <- paste0(backend_label, " | confidence: ", confidence)
    }
    html <- paste0(html, "<p style='margin: 4px 0 0 0; color: #888; font-size: 0.85em;'><em>Generated by: ",
                   htmltools::htmlEscape(backend_label), "</em></p>\n</div>\n")

    cat(html)
    cat("</details>\n")
}

# Load config
cfg <- yaml::read_yaml(config_path)
prot_cfg <- cfg$modes$proteomics
report_cfg <- prot_cfg$report %||% list()  # new report-level parameters

# QC plot visibility flags (all default TRUE except distance heatmap)
show_pca      <- report_cfg$show_pca %||% TRUE
show_density  <- report_cfg$show_density %||% TRUE
show_distance <- report_cfg$show_distance %||% FALSE   # disabled per facility feedback
show_correlation <- report_cfg$show_correlation %||% TRUE
show_expression_heatmap <- report_cfg$show_expression_heatmap %||% TRUE
show_imputation_qc <- report_cfg$show_imputation_qc %||% TRUE
force_pca <- report_cfg$force_pca %||% FALSE

# Load DE summary data (used in multiple sections)
de_summary_file <- file.path(qc_post_dir, "de_summary_all.tsv")
de_data <- NULL
if (file.exists(de_summary_file)) {
    de_data <- read.delim(de_summary_file, stringsAsFactors = FALSE, check.names = FALSE)
}

# Get contrast names from column patterns
contrast_names <- character(0)
if (!is.null(de_data)) {
    padj_cols <- grep("^padj\\.imputs\\.", names(de_data), value = TRUE)
    contrast_names <- gsub("^padj\\.imputs\\.", "", padj_cols)
}
```

```{r force-dependencies, include=FALSE}
# Force DT and Plotly dependencies to be included in the HTML
DT::datatable(data.frame(x = 1))
plotly::plot_ly(data.frame(x = 1, y = 1), x = ~x, y = ~y, type = "scatter", mode = "markers")
```

```{r introduction, results='asis'}
# Introduction — optional and flexible.  Rendered only when at least one
# text field is provided (user_notes, introduction_text, introduction_background,
# introduction_aim).  Each sub-part is omitted silently when empty.
intro_text  <- report_cfg$introduction_text %||% ""
intro_bg    <- report_cfg$introduction_background %||% ""
intro_aim   <- report_cfg$introduction_aim %||% ""
user_notes  <- cfg$project$user_notes %||% ""

has_intro <- any(nzchar(trimws(c(intro_text, intro_bg, intro_aim, user_notes))))

if (has_intro) {
    cat("# Introduction\n\n")

    # Main intro text (or legacy user_notes)
    body_text <- if (nzchar(trimws(intro_text))) intro_text else user_notes
    if (nzchar(trimws(body_text))) {
        summary_file <- file.path(run_dir, "proteomics", "project_summary.html")
        if (file.exists(summary_file)) {
            cat(paste(readLines(summary_file, warn = FALSE), collapse = "\n"))
        } else {
            notes_html <- gsub("\n", "<br>\n", htmltools::htmlEscape(trimws(body_text)))
            cat(sprintf("<div style='background:#f8f9fa; border-left:4px solid #4a90d9; padding:16px 20px; border-radius:6px; margin-bottom:20px; font-size:14px; line-height:1.7; color:#333;'>%s</div>\n", notes_html))
        }
        cat("\n")
    }

    if (nzchar(trimws(intro_bg))) {
        cat("## Background\n\n", trimws(intro_bg), "\n\n")
    }

    if (nzchar(trimws(intro_aim))) {
        cat("## Aim\n\n", trimws(intro_aim), "\n\n")
    }
}
```

```{r experimental-design-section, results='asis'}
# Experimental design subsection — shown when a design file/link is provided
# or when metadata is available.  Provides a sample-name key.
design_file <- report_cfg$experimental_design_file %||% ""
has_design_file <- nzchar(trimws(design_file))

meta_file_ed <- file.path(cfg$project$dir, cfg$paths$raw, prot_cfg$files$metadata)
has_meta_ed <- file.exists(meta_file_ed)

if (has_design_file || has_meta_ed) {
    cat("## Experimental Design\n\n")

    # Link / file reference
    if (has_design_file) {
        if (grepl("^https?://", design_file)) {
            cat(sprintf("Full experimental design: [%s](%s)\n\n", design_file, design_file))
        } else if (file.exists(design_file)) {
            cat(sprintf("Design file: `%s`\n\n", design_file))
        } else {
            cat(sprintf("Design file reference: `%s`\n\n", design_file))
        }
    }

    # Always show a sample-name key table from metadata
    if (has_meta_ed) {
        meta_ed <- read.csv(meta_file_ed, stringsAsFactors = FALSE)
        cat("**Sample key:**\n\n")
        print(htmltools::tagList(make_dt(meta_ed, "Sample Metadata")))
        cat("\n\n")
    }
} else if (has_intro) {
    cat("## Experimental Design\n\n")
    cat("*No experimental design file was provided. Add `report.experimental_design_file` to the config to include one.*\n\n")
}
```

# Analysis Parameters {.tabset}

## Project

```{r params-project}
proj <- data.frame(
  Parameter = c("Project Name", "Analysis Round", "Project Directory",
                 "Raw Data Path", "Output Path"),
  Value = c(cfg$project$name, cfg$project$analysis_round, cfg$project$dir,
            cfg$paths$raw, cfg$paths$out),
  stringsAsFactors = FALSE
)
make_dt(proj, "Analysis Parameters")
```

## Experimental Design

```{r params-design}
meta_file <- file.path(cfg$project$dir, cfg$paths$raw, prot_cfg$files$metadata)
if (file.exists(meta_file)) {
    meta <- read.csv(meta_file, stringsAsFactors = FALSE)
    group_col <- prot_cfg$effects$color %||% "Condition"

    if (group_col %in% names(meta)) {
        design_tbl <- as.data.frame(table(meta[[group_col]]))
        colnames(design_tbl) <- c("Group", "n (replicates)")
        make_dt(design_tbl, "Experimental Design")
    }
}
```

## Pipeline Settings

```{r params-pipeline}
settings <- data.frame(
  Parameter = c(
    "Search Engine",
    "Imputation Method",
    "DE Method",
    "Adjusted p-value Cutoff",
    "Linear FC Cutoff",
    "Clustering Enabled",
    "Pathway Analysis",
    "PPI Networks",
    "Advanced Statistics"
  ),
  Value = c(
    prot_cfg$engine %||% "unknown",
    prot_cfg$imputation$method %||% "none",
    prot_cfg$de$method %||% "limma",
    prot_cfg$de$p_cutoff %||% 0.05,
    prot_cfg$de$linear_fc_cutoff %||% 1.5,
    ifelse(isTRUE(prot_cfg$clustering$enabled), "Yes", "No"),
    ifelse(isTRUE(prot_cfg$pathway$enabled), "Yes", "No"),
    ifelse(isTRUE(prot_cfg$ppi$enabled), "Yes", "No"),
    ifelse(isTRUE(prot_cfg$advanced_stats$enabled), "Yes", "No")
  ),
  stringsAsFactors = FALSE
)
make_dt(settings, "Pipeline Settings")
```

## Contrasts

```{r params-contrasts}
contrasts_file <- file.path(cfg$project$dir, cfg$paths$raw, prot_cfg$files$contrasts)
if (file.exists(contrasts_file)) {
    contrasts <- read.csv(contrasts_file, stringsAsFactors = FALSE)
    make_dt(contrasts, "Contrasts")
} else {
    cat("No contrasts file specified.\n")
}
```

---

# Technical Summary

```{r technical-summary, results='asis'}
tech <- prot_cfg$technical_report

meta_ts <- tryCatch(
    read.csv(file.path(cfg$project$dir, cfg$paths$raw, prot_cfg$files$metadata),
             stringsAsFactors = FALSE),
    error = function(e) NULL
)
group_col_ts <- prot_cfg$effects$color

n_samples_ts <- if (!is.null(meta_ts)) nrow(meta_ts) else NA
n_groups_ts <- if (!is.null(meta_ts) && group_col_ts %in% names(meta_ts)) length(unique(meta_ts[[group_col_ts]])) else NA

cat("## Sample & Acquisition Overview\n\n")
overview <- character(0)
if (!is.na(n_samples_ts) && !is.na(n_groups_ts)) {
    overview <- c(overview, sprintf("This analysis comprises **%d samples** across **%d experimental groups** (grouping variable: *%s*).",
                                     n_samples_ts, n_groups_ts, group_col_ts))
}
if (!is.null(tech$facility) && nzchar(tech$facility)) {
    overview <- c(overview, sprintf("Data were acquired at the **%s**.", tech$facility))
}
if (length(overview) > 0) cat(paste(overview, collapse = " "), "\n\n")

if (!is.null(tech)) {
    if (!is.null(tech$sample_prep) && nzchar(tech$sample_prep)) {
        cat("## Sample Preparation\n\n", tech$sample_prep, "\n\n")
    }
    if (!is.null(tech$ms_acquisition) && nzchar(tech$ms_acquisition)) {
        cat("## MS Acquisition\n\n", tech$ms_acquisition, "\n\n")
    }
    if (!is.null(tech$search_engine) && nzchar(tech$search_engine)) {
        cat("## Database Search\n\n", tech$search_engine, "\n\n")
    }
    if (!is.null(tech$search_parameters) && nzchar(tech$search_parameters)) {
        cat("## Search Parameters\n\n", tech$search_parameters, "\n\n")
    }
}

cat("## Downstream Analysis (multiomics-core)\n\n")
cat(sprintf("- **Imputation:** %s", prot_cfg$imputation$method %||% "none"))
if (!is.null(prot_cfg$imputation$width)) cat(sprintf(" (width=%.1f, downshift=%.1f)", prot_cfg$imputation$width, prot_cfg$imputation$downshift))
cat("\n")
cat(sprintf("- **DE method:** %s (adj.p <= %s, |FC| >= %s)\n",
            prot_cfg$de$method %||% "limma", prot_cfg$de$p_cutoff %||% 0.05, prot_cfg$de$linear_fc_cutoff %||% 1.5))
cat(sprintf("- **Pathway:** %s\n", if (isTRUE(prot_cfg$pathway$enabled)) paste(prot_cfg$pathway$databases, collapse = ", ") else "disabled"))
cat(sprintf("- **PPI:** %s\n", if (isTRUE(prot_cfg$ppi$enabled)) "STRING" else "disabled"))
```

---

# Executive Summary

```{r executive-summary, results='asis'}
exec_summary_path <- file.path(prot_dir, "executive_summary.md")
exec_available <- file.exists(exec_summary_path)

# Build key metrics panel
cat("<div style='background: linear-gradient(135deg, #f8f9fa 0%, #e9ecef 100%); border: 1px solid #dee2e6; border-radius: 8px; padding: 20px 24px; margin: 10px 0 20px 0;'>\n")
cat("<div style='display: flex; flex-wrap: wrap; gap: 16px; margin-bottom: 16px;'>\n")

# Samples
if (exists("meta") && is.data.frame(meta)) {
    n_samp <- nrow(meta)
    cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #2c3e50;'>%d</div><div style='font-size: 0.85em; color: #666;'>Samples</div></div>\n", n_samp))
}

# Groups
if (exists("meta") && is.data.frame(meta) && exists("group_col") && group_col %in% names(meta)) {
    n_groups <- length(unique(meta[[group_col]]))
    cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #2c3e50;'>%d</div><div style='font-size: 0.85em; color: #666;'>Groups</div></div>\n", n_groups))
}

# Proteins quantified
if (!is.null(de_data)) {
    n_proteins <- nrow(de_data)
    cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #2c3e50;'>%s</div><div style='font-size: 0.85em; color: #666;'>Proteins Quantified</div></div>\n", format(n_proteins, big.mark = ",")))
}

# Contrasts
n_contrasts <- length(contrast_names)
if (n_contrasts > 0) {
    cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #2c3e50;'>%d</div><div style='font-size: 0.85em; color: #666;'>Contrasts</div></div>\n", n_contrasts))
}

# Total DE proteins
if (!is.null(de_data)) {
    pass_any_col <- grep("pass_any", names(de_data), value = TRUE)[1]
    if (!is.na(pass_any_col)) {
        n_de_total <- sum(de_data[[pass_any_col]] == TRUE, na.rm = TRUE)
        cat(sprintf("<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: #c0392b;'>%s</div><div style='font-size: 0.85em; color: #666;'>DE Proteins</div></div>\n", format(n_de_total, big.mark = ",")))
    }
}

cat("</div>\n")

# DE breakdown per contrast
if (!is.null(de_data) && length(contrast_names) > 0) {
    padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- prot_cfg$de$linear_fc_cutoff %||% 1.5

    cat("<div style='margin: 12px 0;'>\n")
    cat("<strong>Differential Expression per Contrast:</strong>\n<ul style='margin: 6px 0; padding-left: 20px;'>\n")
    for (cn in contrast_names) {
        padj_col <- paste0("padj.imputs.", cn)
        lfc_col <- paste0("linearFC.imputs.", cn)
        if (padj_col %in% names(de_data) && lfc_col %in% names(de_data)) {
            padj_vals <- as.numeric(de_data[[padj_col]])
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            log2fc <- log2(abs(lfc_vals)) * sign(lfc_vals)
            is_sig <- !is.na(padj_vals) & padj_vals <= padj_cut & abs(log2fc) >= log2(lfc_cut)
            n_up <- sum(is_sig & log2fc > 0, na.rm = TRUE)
            n_down <- sum(is_sig & log2fc < 0, na.rm = TRUE)
            cat(sprintf("<li><strong>%s</strong>: %d DE proteins (%d up, %d down)</li>\n",
                        cn, n_up + n_down, n_up, n_down))
        }
    }
    cat("</ul>\n</div>\n")
}

# Pipeline-generated executive summary
if (exec_available) {
    exec_md_raw <- paste(readLines(exec_summary_path, warn = FALSE), collapse = "\n")
    exec_md_raw <- gsub("^##?\\s*Executive\\s*Summary\\s*\n+", "", exec_md_raw)
    if (nzchar(trimws(exec_md_raw)) && trimws(exec_md_raw) != "No summary available.") {
        cat("<div style='margin-top: 12px; padding-top: 12px; border-top: 1px solid #dee2e6;'>\n")
        cat(exec_md_raw, "\n")
        cat("</div>\n")
    }
}

cat("</div>\n")
```

---

# Results

```{r data-file-links, results='asis'}
raw_loc <- report_cfg$raw_files_location %||% ""
analysis_loc <- report_cfg$analysis_files_location %||% ""
if (nzchar(trimws(raw_loc)) || nzchar(trimws(analysis_loc))) {
    cat("<div style='background:#f0f4f8; border:1px solid #d0d7de; border-radius:6px; padding:12px 18px; margin-bottom:20px;'>\n")
    cat("<strong>Data Locations:</strong><br>\n")
    if (nzchar(trimws(raw_loc))) {
        if (grepl("^https?://", raw_loc)) {
            cat(sprintf("Raw instrument files: <a href='%s'>%s</a><br>\n", raw_loc, raw_loc))
        } else {
            cat(sprintf("Raw instrument files: <code>%s</code><br>\n", raw_loc))
        }
    }
    if (nzchar(trimws(analysis_loc))) {
        if (grepl("^https?://", analysis_loc)) {
            cat(sprintf("Analysis output files: <a href='%s'>%s</a><br>\n", analysis_loc, analysis_loc))
        } else {
            cat(sprintf("Analysis output files: <code>%s</code><br>\n", analysis_loc))
        }
    }
    cat("</div>\n\n")
}
```

## Quality Control — Pre-DE

### Filtering Statistics

Low-abundance features with insufficient signal are removed to reduce noise and improve statistical power. The filtering threshold balances retaining biologically relevant features against eliminating measurement noise.

```{r filtering-stats, results='asis'}
# Try to load filtering stats from pre-computed file or compute inline
filt_stats_file <- file.path(prot_dir, "filtering_stats.csv")
if (file.exists(filt_stats_file)) {
    filt_stats <- read.csv(filt_stats_file, stringsAsFactors = FALSE)
    for (i in seq_len(nrow(filt_stats))) {
        cat(sprintf("**%s:** %s\n\n", filt_stats$metric[i], filt_stats$value[i]))
    }
} else if (!is.null(de_data)) {
    n_quantified <- nrow(de_data)
    cat(sprintf("**Proteins in final dataset:** %s\n\n", format(n_quantified, big.mark = ",")))

    imp_method <- prot_cfg$imputation$method %||% "none"
    if (tolower(imp_method) != "none") {
        cat(sprintf("**Imputation method:** %s\n\n", imp_method))
    }
}
```

```{r expected-proteins-bar, results='asis', fig.height=5, fig.width=8}
# Optional bar chart of expected / key proteins
show_expected_bar <- isTRUE(report_cfg$show_expected_proteins_bar)
expected_file <- report_cfg$expected_proteins_file %||% ""

if (show_expected_bar && nzchar(trimws(expected_file)) && file.exists(expected_file)) {
    cat("### Expected Proteins\n\n")
    exp_df <- tryCatch(read.csv(expected_file, stringsAsFactors = FALSE), error = function(e) NULL)
    if (!is.null(exp_df) && nrow(exp_df) > 0) {
        # Expect columns: protein (or protein_id), value (optional)
        pcol <- intersect(c("protein", "protein_id", "gene", "gene_name"), tolower(names(exp_df)))[1]
        vcol <- intersect(c("value", "intensity", "abundance", "log2fc"), tolower(names(exp_df)))[1]
        if (!is.na(pcol)) {
            names(exp_df)[tolower(names(exp_df)) == pcol] <- "protein"
            if (!is.na(vcol)) {
                names(exp_df)[tolower(names(exp_df)) == vcol] <- "value"
            } else {
                exp_df$value <- seq_len(nrow(exp_df))
            }
            exp_df <- exp_df[order(exp_df$value, decreasing = TRUE), ]
            p <- ggplot2::ggplot(exp_df,
                     ggplot2::aes(x = reorder(protein, value), y = value)) +
                ggplot2::geom_bar(stat = "identity", fill = "steelblue") +
                ggplot2::coord_flip() +
                ggplot2::labs(title = "Expected Proteins", x = NULL, y = "Value") +
                ggplot2::theme_minimal()
            print(p)
            figure_legend("Bar chart of expected/key proteins relevant to this experiment.")
        }
    }
} else if (show_expected_bar) {
    cat("### Expected Proteins\n\n")
    cat("*Expected proteins bar chart enabled but no data file provided (`report.expected_proteins_file`).*\n\n")
}
```

```{r pca-replicate-check, include=FALSE}
# PCA is only meaningful when there are replicates (>=2 samples per condition)
has_replicates <- FALSE
if (exists("meta") && is.data.frame(meta) && exists("group_col") && group_col %in% names(meta)) {
    group_counts <- table(meta[[group_col]])
    has_replicates <- any(group_counts >= 2)
}
show_pca_section <- (isTRUE(show_pca) && has_replicates) || isTRUE(force_pca)
```

```{r pca-skip-msg, eval=!show_pca_section, results='asis'}
if (isTRUE(show_pca) && !has_replicates) {
    cat("## PCA\n\n*PCA plots are omitted because no condition has replicate samples. Set `report.force_pca: true` to override.*\n\n")
}
```

```{r pca-heading, eval=show_pca_section, results='asis'}
cat("## PCA {.tabset}\n\n")
cat("Principal Component Analysis (PCA) reduces high-dimensional expression data to a low-dimensional representation that captures the largest sources of variation. Samples that cluster together share similar expression profiles; separation along PC1 indicates the dominant source of biological or technical variation.\n\n")
cat("### PC1 vs PC2\n")
```

```{r pca-12, eval=show_pca_section, fig.cap="PCA -- PC1 vs PC2"}
p <- file.path(diag_dir, "PCA_PC1.vs.PC2.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-pca-12, eval=show_pca_section, results='asis'}
figure_legend("Principal Component Analysis (PCA) of imputed protein expression data, showing sample separation along PC1 and PC2. Each point represents one sample, colored by experimental group.")
```

```{r commentary-pca, eval=show_pca_section, results='asis'}
render_commentary("pca_plot")
```

```{r pca-topvar-heading, eval=show_pca_section, results='asis'}
cat("### By Top Variable Proteins\n\nSelect the number of top variable proteins used for PCA calculation:\n")
```

```{r pca-var-prot-setup, eval=show_pca_section, echo=FALSE}
pca_files <- list.files(diag_dir, pattern = "^PCA_top[0-9]+\\.png$", full.names = TRUE)

pca_data_map <- list()
available_n_top <- c()

if (length(pca_files) > 0) {
    for (f in pca_files) {
        n_val <- as.numeric(sub(".*PCA_top([0-9]+)\\.png$", "\\1", basename(f)))
        if (file.exists(f)) {
            pca_data_map[[as.character(n_val)]] <- knitr::image_uri(f)
            available_n_top <- c(available_n_top, n_val)
        }
    }
}

if (length(available_n_top) > 0) {
    available_n_top <- sort(unique(available_n_top))
    default_n_top <- available_n_top[1]
    default_img_uri <- pca_data_map[[as.character(default_n_top)]]
} else {
    default_pca_path <- file.path(diag_dir, "PCA_PC1.vs.PC2.png")
    if (file.exists(default_pca_path)) {
        default_n_top <- 0
        default_img_uri <- knitr::image_uri(default_pca_path)
        pca_data_map[["0"]] <- default_img_uri
        available_n_top <- 0
    } else {
        default_n_top <- NULL
        default_img_uri <- NULL
    }
}
```

<div style="margin-bottom: 20px;">
<label for="pca_prot_selector"><strong>Number of top variable proteins:</strong></label>
<select id="pca_prot_selector" onchange="updateProtPCAPlot(this.value)" style="padding: 8px; border-radius: 4px; border: 1px solid #ccc; font-size: 14px;">
```{r pca-prot-dropdown, eval=show_pca_section, echo=FALSE, results='asis'}
if (length(available_n_top) > 0 && available_n_top[1] != 0) {
    for (n_val in available_n_top) {
        selected <- if (n_val == default_n_top) " selected" else ""
        cat(sprintf('<option value="%d"%s>%s proteins</option>\n', n_val, selected, format(n_val, big.mark = ",")))
    }
} else if (!is.null(default_img_uri)) {
    cat('<option value="0" selected>All proteins (default)</option>\n')
} else {
    cat('<option disabled>No PCA plots found</option>\n')
}
```
</select>
</div>

<div id="pca_prot_container" style="text-align: center;">
```{r pca-prot-render, eval=show_pca_section, echo=FALSE, results='asis'}
if (!is.null(default_img_uri)) {
    cat(sprintf('<img id="pca_prot_image" src="%s" style="width:90%%; max-width: 900px; height: auto;" alt="PCA Plot" />', default_img_uri))
} else {
    cat('<p><em>PCA plot not available.</em></p>')
}
```
</div>

<script>
var embeddedProtPcaImages = `r if (length(pca_data_map) > 0) jsonlite::toJSON(pca_data_map, auto_unbox = TRUE) else "{}"`;

function updateProtPCAPlot(n_top) {
    var img = document.getElementById("pca_prot_image");
    if (embeddedProtPcaImages && embeddedProtPcaImages[n_top]) {
        img.src = embeddedProtPcaImages[n_top];
    }
}
</script>

```{r pca-13-heading, eval=show_pca_section, results='asis'}
cat("### PC1 vs PC3\n")
```

```{r pca-13, eval=show_pca_section, fig.cap="PCA -- PC1 vs PC3"}
p <- file.path(diag_dir, "PCA_PC1.vs.PC3.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-pca-13, eval=show_pca_section, results='asis'}
figure_legend("PCA plot showing sample separation along PC1 and PC3. PC3 may capture additional biological variation or batch effects not visible in the PC1-PC2 projection.")
```

```{r pca-3d-heading, eval=show_pca_section, results='asis'}
cat("### 3D Interactive\n")
```

```{r pca-3d-link, eval=show_pca_section, results='asis'}
pca3d_file <- file.path(diag_dir, "PCA_3D.html")
if (file.exists(pca3d_file)) {
    cat(sprintf('<p><a href="proteomics/Diagnostic_plots/PCA_3D.html" target="_blank" style="font-size: 1.1em;">Open 3D PCA (new tab)</a></p>'))
} else {
    cat("*3D PCA plot not available.*")
}
```

```{r umap-heading, eval=show_pca_section, results='asis'}
cat("### UMAP\n")
```

```{r umap, eval=show_pca_section, fig.cap="UMAP embedding of protein expression"}
umap_file <- file.path(diag_dir, "UMAP.png")
if (file.exists(umap_file)) {
    knitr::include_graphics(umap_file)
} else {
    cat("*UMAP plot not available. Enable with `qc.run_umap: true` in config.*\n")
}
```

```{r legend-umap, eval=show_pca_section, results='asis'}
if (file.exists(file.path(diag_dir, "UMAP.png"))) {
    figure_legend("Uniform Manifold Approximation and Projection (UMAP) of imputed protein expression data. UMAP preserves local structure and may reveal non-linear groupings not apparent in PCA.")
}
```

```{r commentary-umap, eval=show_pca_section, results='asis'}
render_commentary("umap_plot")
```

```{r density-heading, eval=show_density, results='asis'}
cat("## Protein Distribution\n")
```

```{r density, eval=show_density, fig.cap="Protein expression density"}
hp <- file.path(diag_dir, "protein_histograms_summary.png")
if (file.exists(hp)) knitr::include_graphics(hp)
```

```{r legend-density, eval=show_density, results='asis'}
figure_legend("Kernel density estimates of per-sample protein expression distributions. Overlapping curves indicate consistent expression profiles across samples.")
```

```{r commentary-density, eval=show_density, results='asis'}
render_commentary("density_plot")
```

```{r imputation-check, include=FALSE}
imp_method <- prot_cfg$imputation$method %||% "none"
has_imputation <- tolower(imp_method) != "none" && isTRUE(show_imputation_qc)
```

```{r imputation-heading, eval=has_imputation, results='asis'}
cat("## Imputation QC {.tabset}\n")
```

```{r imputation-section, eval=has_imputation, results='asis'}
cat("Imputation was performed using the **", imp_method, "** method.\n\n")

if (tolower(imp_method) == "perseus") {
    imp_w <- prot_cfg$imputation$width %||% 0.3
    imp_s <- prot_cfg$imputation$downshift %||% 1.8
    imp_n <- prot_cfg$imputation$n_imputations %||% 1
    imp_min <- prot_cfg$imputation$min_passed %||% 1

    cat("#### How Perseus-style imputation works\n\n")
    cat("Missing values in label-free proteomics are often *missing not at random* (MNAR): ",
        "they arise because the protein abundance is below the instrument's detection limit. ",
        "Perseus-style imputation models these missing values by drawing random intensities ",
        "from a **downshifted normal distribution** relative to each sample's observed values.\n\n")
    cat("For each sample, the imputed values are drawn from:\n\n")
    cat(sprintf("- **Mean** = sample mean - %.1f x sample SD (downshift)\n", imp_s))
    cat(sprintf("- **SD** = %.1f x sample SD (width)\n\n", imp_w))
    cat("This places imputed values in the low-abundance tail of the distribution, ",
        "consistent with the MNAR assumption. ",
        "The narrow width (relative to the observed distribution) prevents imputed values ",
        "from inflating variance estimates.\n\n")
    if (imp_n > 1) {
        cat(sprintf("To account for imputation uncertainty, the procedure is repeated **%d times** ", imp_n))
        cat("and differential expression is computed on each imputed dataset independently. ",
            "Final p-values and fold changes are aggregated across imputations.\n\n")
    }
    cat(sprintf("Only proteins detected in at least **%d sample(s)** per group were retained for analysis.\n\n", imp_min))
}
```

### Imputation Histograms

```{r imp-hist, eval=has_imputation, fig.cap="Imputed vs original intensity distributions"}
imp_w <- prot_cfg$imputation$width %||% "NA"
imp_s <- prot_cfg$imputation$downshift %||% "NA"
f_imp <- file.path(diag_dir, sprintf("imputed_histograms_samples_summary_w%s_s%s.png", imp_w, imp_s))
if (file.exists(f_imp)) knitr::include_graphics(f_imp)
```

```{r legend-imp-hist, eval=has_imputation, results='asis'}
figure_legend("Histograms comparing the distribution of observed (blue) and imputed (red) protein intensities per sample. Imputed values should appear as a low-abundance shoulder distinct from the observed distribution.")
```

```{r commentary-imp-hist, results='asis'}
render_commentary("imputation_histogram")
```

### Boxplot After Imputation

```{r imp-boxplot, eval=has_imputation, fig.cap="Expression boxplot after imputation"}
f_box <- file.path(diag_dir, "imputed_boxplot.png")
if (file.exists(f_box)) knitr::include_graphics(f_box)
```

```{r legend-imp-boxplot, eval=has_imputation, results='asis'}
figure_legend("Boxplot of protein expression values after imputation. Boxes show the interquartile range (IQR) for each sample. Consistent medians and IQR widths across samples indicate successful normalization.")
```

```{r commentary-boxplot-post, results='asis'}
render_commentary("imputation_boxplot")
```

```{r similarity-heading, eval=show_distance || show_correlation, results='asis'}
cat("## Sample Similarity {.tabset}\n\n")
cat("Pairwise sample correlation and distance metrics assess overall reproducibility within and between experimental groups. High within-group correlation and clear between-group separation indicate good data quality and a detectable biological signal.\n\n")
```

```{r dist-heading, eval=show_distance, results='asis'}
cat("### Distance Heatmap (Imputed)\n")
```

```{r dist-heatmap, eval=show_distance, fig.cap="Sample-to-sample distance heatmap (imputed data)"}
p <- file.path(diag_dir, "sample_distance_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-dist-heatmap, eval=show_distance, results='asis'}
figure_legend("Euclidean distance matrix of imputed protein expression values. Darker colors indicate smaller distances (higher similarity). Samples from the same experimental group should cluster together.")
```

```{r commentary-distance, eval=show_distance, results='asis'}
render_commentary("sample_distance")
```

```{r dist-na-heading, eval=show_distance, results='asis'}
cat("### Distance Heatmap (Raw, with NAs)\n")
```

```{r dist-heatmap-na, eval=show_distance, fig.cap="Sample-to-sample distance heatmap (raw data, NAs present)"}
p <- file.path(diag_dir, "sample_distance_heatmap_NA.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r corr-heading, eval=show_correlation, results='asis'}
cat("### Correlation Heatmap\n")
```

```{r corr-heatmap, eval=show_correlation, fig.cap="Sample-to-sample correlation heatmap"}
p <- file.path(diag_dir, "sample_correlation_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-corr-heatmap, eval=show_correlation, results='asis'}
figure_legend("Pairwise Pearson correlation heatmap of protein expression profiles. Within-group correlations should be higher than between-group correlations.")
```

```{r commentary-correlation, eval=show_correlation, results='asis'}
render_commentary("sample_correlation")
```

```{r expr-heatmap-heading, eval=show_expression_heatmap, results='asis'}
cat("## Expression Heatmap\n\n")
cat("Expression heatmaps display the relative abundance of features across samples, with hierarchical clustering grouping features and/or samples by expression similarity. They reveal global expression patterns, condition-specific signatures, and potential outlier samples.\n\n")
```

```{r expr-heatmap, eval=show_expression_heatmap, fig.cap="Sample expression heatmap (top variable proteins)"}
p <- file.path(diag_dir, "samples_protein_heatmap.png")
if (file.exists(p)) knitr::include_graphics(p)
```

```{r legend-expr-heatmap, eval=show_expression_heatmap, results='asis'}
figure_legend("Unsupervised hierarchical clustering heatmap based on the top most variable proteins. Both rows (proteins) and columns (samples) are clustered by expression similarity.")
```

```{r commentary-expr-heatmap, eval=show_expression_heatmap, results='asis'}
render_commentary("expr_heatmap")
```

## Outlier Detection

Outlier detection identifies samples with expression profiles that deviate substantially from their experimental group. Samples flagged as outliers may indicate technical failures, mislabeling, or genuine biological differences that warrant further investigation.

```{r outlier-detection, results='asis'}
outlier_file <- file.path(diag_dir, "outlier_report.csv")
if (file.exists(outlier_file)) {
    outlier_df <- read.csv(outlier_file, stringsAsFactors = FALSE)
    if (nrow(outlier_df) > 0) {
        cat(sprintf("**%d potential outlier sample(s) detected:**\n\n", nrow(outlier_df)))
        make_dt(outlier_df, caption = "Outlier Samples", pageLength = 10)
    } else {
        cat("No outlier samples detected.\n")
    }
} else {
    cat("Outlier detection results not available.\n")
}
```

---

# Differential Expression Analysis

```{r de-stats-description, results='asis'}
de_method_name <- prot_cfg$de$method %||% "limma"
padj_cut_de <- prot_cfg$de$p_cutoff %||% 0.05
lfc_cut_de <- prot_cfg$de$linear_fc_cutoff %||% 1.5

# Custom stats text takes priority; otherwise auto-generate from config
custom_stats_text <- report_cfg$stats_methods_text %||% ""
if (nzchar(trimws(custom_stats_text))) {
    cat(trimws(custom_stats_text), "\n\n")
} else {
    # Auto-generate description
    test_desc <- switch(tolower(de_method_name),
        limma = "moderated t-tests via the *limma* R/Bioconductor package (empirical Bayes shrinkage of variance estimates)",
        "limma-voom" = "limma-voom (log-CPM transformed, precision-weighted linear models)",
        deseq2 = "Wald tests via *DESeq2* (negative binomial generalized linear models)",
        paste0("*", de_method_name, "* method")
    )
    cat(sprintf(
        "Differential expression was assessed using %s. Multiple testing correction was applied using the **Benjamini-Hochberg** procedure (FDR). Proteins were considered differentially expressed at adjusted p-value $\\leq$ %s and |linear fold change| $\\geq$ %s.\n\n",
        test_desc, padj_cut_de, lfc_cut_de
    ))
}

# Link to full results table
full_results_link <- report_cfg$full_results_table_link %||% ""
if (nzchar(trimws(full_results_link))) {
    if (grepl("^https?://", full_results_link)) {
        cat(sprintf("Full differential results table: [download](%s)\n\n", full_results_link))
    } else {
        cat(sprintf("Full differential results table: `%s`\n\n", full_results_link))
    }
}
```

## DE Summary

Differential expression analysis identifies features with statistically significant abundance changes between experimental conditions. Results are corrected for multiple testing to control the false discovery rate.

```{r de-summary}
if (!is.null(de_data) && length(contrast_names) > 0) {
    padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- prot_cfg$de$linear_fc_cutoff %||% 1.5

    total_proteins <- nrow(de_data)

    de_counts_list <- list()
    for (cn in contrast_names) {
        padj_col <- paste0("padj.imputs.", cn)
        lfc_col <- paste0("linearFC.imputs.", cn)
        if (padj_col %in% names(de_data) && lfc_col %in% names(de_data)) {
            padj_vals <- as.numeric(de_data[[padj_col]])
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            log2fc <- log2(abs(lfc_vals)) * sign(lfc_vals)
            is_sig <- !is.na(padj_vals) & padj_vals <= padj_cut & abs(log2fc) >= log2(lfc_cut)
            n_up <- sum(is_sig & log2fc > 0, na.rm = TRUE)
            n_down <- sum(is_sig & log2fc < 0, na.rm = TRUE)
            de_counts_list[[cn]] <- data.frame(
                Contrast = cn, Up = n_up, Down = n_down, Total = n_up + n_down,
                stringsAsFactors = FALSE
            )
        }
    }

    if (length(de_counts_list) > 0) {
        de_counts_df <- do.call(rbind, de_counts_list)
        cat(sprintf("**Total proteins quantified:** %s\n\n", format(total_proteins, big.mark = ",")))
        make_dt(de_counts_df, "DE Protein Counts per Contrast")
    }
} else {
    cat("DE summary data not available.\n")
}
```

## DE Protein Overlap {.tabset}

### All DE Proteins

```{r de-overlap-venn-all, fig.height=6, fig.width=10, results='asis'}
if (!is.null(de_data) && length(contrast_names) >= 2) {
    padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- log2(prot_cfg$de$linear_fc_cutoff %||% 1.5)
    id_col <- prot_cfg$de_table$id_col %||% "FeatureID"

    de_protein_lists <- list()
    for (cn in contrast_names) {
        padj_col <- paste0("padj.imputs.", cn)
        lfc_col <- paste0("linearFC.imputs.", cn)
        if (padj_col %in% names(de_data) && lfc_col %in% names(de_data)) {
            padj_vals <- as.numeric(de_data[[padj_col]])
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            log2fc <- log2(abs(lfc_vals)) * sign(lfc_vals)
            is_de <- !is.na(padj_vals) & padj_vals <= padj_cut & abs(log2fc) >= lfc_cut
            protein_col <- if (id_col %in% names(de_data)) id_col else names(de_data)[1]
            de_protein_lists[[gsub("_", " ", cn)]] <- de_data[[protein_col]][is_de]
        }
    }

    if (length(de_protein_lists) >= 2 && length(de_protein_lists) <= 7) {
        if (has_ggVennDiagram) {
            total_unique <- length(unique(unlist(de_protein_lists)))
            short_names <- gsub(" vs ", "\nvs ", names(de_protein_lists))
            names(de_protein_lists) <- short_names
            venn_plot <- ggVennDiagram(de_protein_lists, label_alpha = 0,
                                       label = "both", label_percent_digit = 1,
                                       set_size = 2.8) +
                scale_fill_gradient(low = "white", high = "steelblue") +
                scale_x_continuous(expand = expansion(mult = 0.4)) +
                labs(title = "DE Protein Overlap (All)",
                     subtitle = paste0("Total unique DE proteins: ", total_unique)) +
                theme(legend.position = "none",
                      plot.title = element_text(hjust = 0),
                      plot.subtitle = element_text(hjust = 0),
                      plot.margin = margin(10, 60, 10, 60))
            print(venn_plot)
        } else {
            cat("**DE protein counts per contrast:**\n\n")
            for (nm in names(de_protein_lists)) {
                cat(sprintf("- **%s**: %d proteins\n", nm, length(de_protein_lists[[nm]])))
            }
        }
    }
} else if (length(contrast_names) < 2) {
    cat("Venn diagram requires at least 2 contrasts.\n")
} else {
    cat("DE data not available for Venn diagram.\n")
}
```

### Up-regulated

```{r de-overlap-venn-up, fig.height=6, fig.width=10, results='asis'}
if (!is.null(de_data) && length(contrast_names) >= 2) {
    padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- log2(prot_cfg$de$linear_fc_cutoff %||% 1.5)
    id_col <- prot_cfg$de_table$id_col %||% "FeatureID"

    up_protein_lists <- list()
    for (cn in contrast_names) {
        padj_col <- paste0("padj.imputs.", cn)
        lfc_col <- paste0("linearFC.imputs.", cn)
        if (padj_col %in% names(de_data) && lfc_col %in% names(de_data)) {
            padj_vals <- as.numeric(de_data[[padj_col]])
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            log2fc <- log2(abs(lfc_vals)) * sign(lfc_vals)
            is_up <- !is.na(padj_vals) & padj_vals <= padj_cut & log2fc >= lfc_cut
            protein_col <- if (id_col %in% names(de_data)) id_col else names(de_data)[1]
            up_protein_lists[[gsub("_", " ", cn)]] <- de_data[[protein_col]][is_up]
        }
    }

    if (length(up_protein_lists) >= 2 && length(up_protein_lists) <= 7 && has_ggVennDiagram) {
        total_up <- length(unique(unlist(up_protein_lists)))
        short_names_up <- gsub(" vs ", "\nvs ", names(up_protein_lists))
        names(up_protein_lists) <- short_names_up
        venn_plot <- ggVennDiagram(up_protein_lists, label_alpha = 0,
                                   label = "both", label_percent_digit = 1,
                                   set_size = 2.8) +
            scale_fill_gradient(low = "white", high = "#E41A1C") +
            scale_x_continuous(expand = expansion(mult = 0.4)) +
            labs(title = "Up-regulated Protein Overlap",
                 subtitle = paste0("Total unique up-regulated: ", total_up)) +
            theme(legend.position = "none",
                  plot.title = element_text(hjust = 0),
                  plot.subtitle = element_text(hjust = 0),
                  plot.margin = margin(10, 60, 10, 60))
        print(venn_plot)
    }
}
```

### Down-regulated

```{r de-overlap-venn-down, fig.height=6, fig.width=10, results='asis'}
if (!is.null(de_data) && length(contrast_names) >= 2) {
    padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- log2(prot_cfg$de$linear_fc_cutoff %||% 1.5)
    id_col <- prot_cfg$de_table$id_col %||% "FeatureID"

    down_protein_lists <- list()
    for (cn in contrast_names) {
        padj_col <- paste0("padj.imputs.", cn)
        lfc_col <- paste0("linearFC.imputs.", cn)
        if (padj_col %in% names(de_data) && lfc_col %in% names(de_data)) {
            padj_vals <- as.numeric(de_data[[padj_col]])
            lfc_vals <- as.numeric(de_data[[lfc_col]])
            log2fc <- log2(abs(lfc_vals)) * sign(lfc_vals)
            is_down <- !is.na(padj_vals) & padj_vals <= padj_cut & log2fc <= -lfc_cut
            protein_col <- if (id_col %in% names(de_data)) id_col else names(de_data)[1]
            down_protein_lists[[gsub("_", " ", cn)]] <- de_data[[protein_col]][is_down]
        }
    }

    if (length(down_protein_lists) >= 2 && length(down_protein_lists) <= 7 && has_ggVennDiagram) {
        total_down <- length(unique(unlist(down_protein_lists)))
        short_names_dn <- gsub(" vs ", "\nvs ", names(down_protein_lists))
        names(down_protein_lists) <- short_names_dn
        venn_plot <- ggVennDiagram(down_protein_lists, label_alpha = 0,
                                   label = "both", label_percent_digit = 1,
                                   set_size = 2.8) +
            scale_fill_gradient(low = "white", high = "#377EB8") +
            scale_x_continuous(expand = expansion(mult = 0.4)) +
            labs(title = "Down-regulated Protein Overlap",
                 subtitle = paste0("Total unique down-regulated: ", total_down)) +
            theme(legend.position = "none",
                  plot.title = element_text(hjust = 0),
                  plot.subtitle = element_text(hjust = 0),
                  plot.margin = margin(10, 60, 10, 60))
        print(venn_plot)
    }
}
```

```{r identified-proteins-venn, results='asis', fig.height=6, fig.width=10}
# Venn diagram of all identified proteins per sample/condition (pre-imputation)
show_id_venn <- isTRUE(report_cfg$show_identified_proteins_venn)

if (show_id_venn && has_ggVennDiagram) {
    cat("## Identified Proteins (Pre-imputation) Venn\n\n")
    cat("This Venn diagram shows proteins *identified* (non-missing) in each condition,\n")
    cat("**before** imputation of missing values.\n\n")

    # Load raw (pre-imputation) expression data — prefer unimputed file, exclude imputed
    raw_expr_file <- list.files(datasets_dir, pattern = "unimputed.*\\.(tsv|csv)$",
                                 full.names = TRUE)[1]
    if (is.na(raw_expr_file)) {
        raw_expr_file <- list.files(datasets_dir, pattern = "(raw_|pre_imputation).*\\.(tsv|csv)$",
                                     full.names = TRUE)[1]
    }
    if (!is.na(raw_expr_file) && file.exists(raw_expr_file)) {
        raw_df <- tryCatch({
            if (grepl("\\.csv$", raw_expr_file)) read.csv(raw_expr_file, stringsAsFactors = FALSE, check.names = FALSE)
            else read.delim(raw_expr_file, stringsAsFactors = FALSE, check.names = FALSE)
        }, error = function(e) NULL)

        if (!is.null(raw_df) && exists("meta") && is.data.frame(meta)) {
            id_col_v <- prot_cfg$de_table$id_col %||% "FeatureID"
            # Fallback: raw/unimputed file may use original protein_id column, not FeatureID
            if (!(id_col_v %in% names(raw_df))) {
                id_col_v <- prot_cfg$id_columns$protein_id %||% "Protein.Group"
            }
            sample_col_v <- prot_cfg$effects$samples %||% "SampleID"
            group_col_v <- prot_cfg$effects$color %||% "Condition"

            if (id_col_v %in% names(raw_df) && group_col_v %in% names(meta) && sample_col_v %in% names(meta)) {
                sample_ids <- meta[[sample_col_v]]
                groups <- meta[[group_col_v]]
                expr_cols <- intersect(sample_ids, names(raw_df))

                if (length(expr_cols) >= 2) {
                    group_map <- setNames(groups, sample_ids)
                    unique_groups <- unique(groups)

                    # For each group, list proteins detected in at least one sample
                    id_lists <- lapply(setNames(unique_groups, unique_groups), function(g) {
                        g_samples <- names(group_map)[group_map == g]
                        g_samples <- intersect(g_samples, names(raw_df))
                        if (length(g_samples) == 0) return(character(0))
                        detected <- apply(raw_df[, g_samples, drop = FALSE], 1, function(row) any(!is.na(row) & row != 0))
                        raw_df[[id_col_v]][detected]
                    })

                    if (length(id_lists) >= 2 && length(id_lists) <= 7) {
                        names(id_lists) <- gsub("_", " ", names(id_lists))
                        total <- length(unique(unlist(id_lists)))
                        venn_plot <- ggVennDiagram(id_lists, label_alpha = 0,
                                                   label = "both", label_percent_digit = 1,
                                                   set_size = 2.8) +
                            scale_fill_gradient(low = "white", high = "#2ca25f") +
                            scale_x_continuous(expand = expansion(mult = 0.4)) +
                            labs(title = "All Identified Proteins per Condition (Pre-imputation)",
                                 subtitle = paste0("Total unique identified: ", total)) +
                            theme(legend.position = "none",
                                  plot.title = element_text(hjust = 0),
                                  plot.subtitle = element_text(hjust = 0),
                                  plot.margin = margin(10, 60, 10, 60))
                        print(venn_plot)
                        figure_legend("Venn diagram of all identified (non-missing) proteins per experimental condition, before imputation.")
                    }
                }
            }
        }
    } else {
        cat("*Pre-imputation data not available for identified proteins Venn diagram.*\n\n")
    }
}
```

## DE Results Table

```{r de-table, results='asis'}
if (!is.null(de_data)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Full differential expression results merged with imputed expression values. Columns include per-contrast adjusted p-values, fold changes, significance flags, and per-sample expression. Use the search box to filter and the buttons to export.</em></p>\n")
    show_df <- de_data

    # Merge imputed expression into DE results
    imp_files_de <- list.files(datasets_dir, pattern = "imputed.*\\.tsv$", full.names = TRUE)
    if (length(imp_files_de) > 0) {
        imp_expr <- tryCatch(
            read.delim(imp_files_de[1], stringsAsFactors = FALSE, check.names = FALSE),
            error = function(e) NULL
        )
        if (!is.null(imp_expr)) {
            imp_id_col <- names(imp_expr)[1]
            de_id_col <- prot_cfg$de_table$id_col %||% "FeatureID"
            if (de_id_col %in% names(show_df)) {
                merged_de <- merge(show_df, imp_expr, by.x = de_id_col, by.y = imp_id_col, all.x = TRUE)
                if (nrow(merged_de) > 0) show_df <- merged_de
            }
        }
    }

    make_dt(show_df, caption = "DE Results + Expression", pageLength = 15)
}
```

## Volcano Plots {.tabset}

### Static {.tabset}

```{r volcano-plots-static, results='asis'}
volcano_files <- list.files(qc_post_dir, pattern = "^volcano_.*\\.png$", full.names = TRUE)
padj_cut_vol <- prot_cfg$de$p_cutoff %||% 0.05
lfc_cut_vol <- prot_cfg$de$linear_fc_cutoff %||% 1.5
if (length(volcano_files) > 0) {
    for (vf in volcano_files) {
        cn <- gsub("^volcano_|\\.png$", "", basename(vf))
        contrast_clean <- gsub("_", " ", cn)
        cat(sprintf("\n#### %s\n\n", contrast_clean))
        cat(sprintf("![Volcano -- %s](%s)\n\n", contrast_clean, vf))
        figure_legend(sprintf("Volcano plot for the %s contrast. Red: significantly up-regulated; blue: significantly down-regulated (adj. p &le; %s, |FC| &ge; %s).", contrast_clean, padj_cut_vol, lfc_cut_vol))
        render_commentary(paste0("volcano_", cn))
    }
} else {
    cat("No volcano plots found.\n")
}
```

### Interactive {.tabset}

```{r volcano-interactive-setup, include=FALSE}
# Create interactive volcano plot
make_volcano_plotly_prot <- function(df, contrast, padj_cut = 0.05, lfc_cut = log2(1.5), title = NULL) {
    padj_col <- paste0("padj.imputs.", contrast)
    lfc_col <- paste0("linearFC.imputs.", contrast)
    if (!(padj_col %in% names(df)) || !(lfc_col %in% names(df))) return(NULL)

    id_col <- prot_cfg$de_table$id_col %||% "FeatureID"
    gene_col <- if ("Genes" %in% names(df)) "Genes" else if (id_col %in% names(df)) id_col else names(df)[1]

    plot_df <- data.frame(
        protein = df[[gene_col]],
        logFC = log2(abs(as.numeric(df[[lfc_col]]))) * sign(as.numeric(df[[lfc_col]])),
        padj = as.numeric(df[[padj_col]]),
        stringsAsFactors = FALSE
    )
    plot_df$neglog10padj <- -log10(pmax(plot_df$padj, 1e-300))
    plot_df$Direction <- ifelse(
        !is.na(plot_df$padj) & plot_df$padj <= padj_cut & abs(plot_df$logFC) >= lfc_cut,
        ifelse(plot_df$logFC > 0, "Up", "Down"), "NS"
    )
    plot_df$Direction <- factor(plot_df$Direction, levels = c("NS", "Down", "Up"))

    p <- ggplot(plot_df, aes(x = logFC, y = neglog10padj, color = Direction,
                              text = paste0("Protein: ", protein, "\nlog2FC: ", round(logFC, 3),
                                           "\npadj: ", signif(padj, 3)))) +
        geom_point(alpha = 0.6, size = 1.5) +
        scale_color_manual(values = c("NS" = "grey70", "Down" = "blue", "Up" = "red")) +
        geom_vline(xintercept = c(-lfc_cut, lfc_cut), linetype = "dashed") +
        geom_hline(yintercept = -log10(padj_cut), linetype = "dashed") +
        labs(title = title, x = "log2 Fold Change", y = "-log10(adj.P.Val)") +
        theme_minimal()
    ggplotly(p, tooltip = "text")
}
```

Volcano plots display the relationship between statistical significance (-log10 adjusted p-value) and biological effect size (log2 fold-change) for each feature. Points in the upper corners represent features that are both statistically significant and have large fold changes.

Hover over points to see protein names and statistics. Enter protein names to highlight them.

```{r volcano-interactive, results='asis', fig.height=6}
if (!is.null(de_data) && length(contrast_names) > 0) {
    for (i in seq_along(contrast_names)) {
        cn <- contrast_names[i]
        contrast_clean <- gsub("_", " ", cn)
        clean_id <- gsub("[^a-zA-Z0-9]", "_", cn)

        cat(sprintf("\n#### %s\n\n", contrast_clean))

        cat(sprintf('
<div style="margin-bottom: 15px; position: relative;">
<label for="volcano_genes_%s"><strong>Highlight proteins:</strong></label>
<div style="display: inline-block; position: relative; margin-left: 10px;">
<input type="text" id="volcano_genes_%s" placeholder="Start typing a protein name..." autocomplete="off"
  style="padding: 8px; border-radius: 4px; border: 1px solid #ccc; width: 350px;"
  oninput="volcanoAutocomplete_%s(this)" onkeydown="volcanoAutoKey_%s(event)">
<div id="volcano_ac_%s" style="position: absolute; top: 100%%; left: 0; width: 350px; max-height: 200px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 4px 4px; z-index: 1000; display: none; box-shadow: 0 4px 6px rgba(0,0,0,0.1);"></div>
</div>
<button onclick="highlightVolcanoGenes_%s()" style="padding: 8px 15px; margin-left: 5px; background: #007bff; color: white; border: none; border-radius: 4px; cursor: pointer;">Highlight</button>
<button onclick="clearVolcanoHighlight_%s()" style="padding: 8px 15px; margin-left: 5px; background: #6c757d; color: white; border: none; border-radius: 4px; cursor: pointer;">Clear</button>
<div id="volcano_tags_%s" style="margin-top: 5px; display: flex; flex-wrap: wrap; gap: 4px;"></div>
</div>
', clean_id, clean_id, clean_id, clean_id, clean_id, clean_id, clean_id, clean_id))

        p <- make_volcano_plotly_prot(de_data, cn, prot_cfg$de$p_cutoff %||% 0.05,
                                      log2(prot_cfg$de$linear_fc_cutoff %||% 1.5), contrast_clean)
        if (!is.null(p)) {
            p$elementId <- paste0("volcano_plot_", clean_id)
            print(htmltools::tagList(p))

            cat(sprintf('
<script>
(function() {
    var cid = "%s";
    var selectedGenes = [];
    var acIndex = -1;
    function getGeneNames() {
        var plotDiv = document.getElementById("volcano_plot_" + cid);
        if (!plotDiv || !plotDiv.data) return [];
        var names = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var data = plotDiv.data[t];
            if (!data.text) continue;
            for (var i = 0; i < data.text.length; i++) {
                var parts = data.text[i].split(/<br\\s*\\/?>/i);
                if (parts.length > 0) {
                    var g = parts[0].replace(/Protein:\\s*/i, "").trim();
                    if (g && names.indexOf(g) === -1) names.push(g);
                }
            }
        }
        return names.sort();
    }
    window.volcanoAutocomplete_%s = function(el) {
        var acDiv = document.getElementById("volcano_ac_" + cid);
        var val = el.value.trim().toLowerCase();
        acDiv.innerHTML = ""; acIndex = -1;
        if (val.length < 1) { acDiv.style.display = "none"; return; }
        var parts = val.split(","); var current = parts[parts.length - 1].trim();
        if (current.length < 1) { acDiv.style.display = "none"; return; }
        var allGenes = getGeneNames();
        var matches = allGenes.filter(function(g) { return g.toLowerCase().indexOf(current) !== -1 && selectedGenes.indexOf(g) === -1; }).slice(0, 15);
        if (matches.length === 0) { acDiv.style.display = "none"; return; }
        matches.forEach(function(g) {
            var item = document.createElement("div");
            item.textContent = g;
            item.style.cssText = "padding: 6px 10px; cursor: pointer; font-size: 13px;";
            item.onmouseenter = function() { this.style.background = "#e9ecef"; };
            item.onmouseleave = function() { this.style.background = "white"; };
            item.onclick = function() { selectGene(g); };
            acDiv.appendChild(item);
        });
        acDiv.style.display = "block";
    };
    window.volcanoAutoKey_%s = function(e) {
        var acDiv = document.getElementById("volcano_ac_" + cid);
        var items = acDiv.children;
        if (items.length === 0) return;
        if (e.key === "ArrowDown") { e.preventDefault(); acIndex = Math.min(acIndex + 1, items.length - 1); updateAcHighlight(items); }
        else if (e.key === "ArrowUp") { e.preventDefault(); acIndex = Math.max(acIndex - 1, 0); updateAcHighlight(items); }
        else if (e.key === "Enter" && acIndex >= 0) { e.preventDefault(); items[acIndex].click(); }
    };
    function updateAcHighlight(items) { for (var i = 0; i < items.length; i++) items[i].style.background = i === acIndex ? "#e9ecef" : "white"; }
    function selectGene(gene) {
        if (selectedGenes.indexOf(gene) === -1) selectedGenes.push(gene);
        document.getElementById("volcano_genes_" + cid).value = "";
        document.getElementById("volcano_ac_" + cid).style.display = "none";
        renderTags(); applyHighlight();
    }
    function renderTags() {
        var container = document.getElementById("volcano_tags_" + cid);
        container.innerHTML = "";
        selectedGenes.forEach(function(g) {
            var tag = document.createElement("span");
            tag.style.cssText = "display: inline-flex; align-items: center; background: #007bff; color: white; padding: 2px 8px; border-radius: 12px; font-size: 12px; gap: 4px;";
            tag.innerHTML = g + " <span style=\\"cursor:pointer; font-weight:bold; margin-left:3px;\\" onclick=\\"removeVolcanoGene_" + cid + "(\'" + g + "\')\\">x</span>";
            container.appendChild(tag);
        });
    }
    window["removeVolcanoGene_" + cid] = function(gene) { selectedGenes = selectedGenes.filter(function(g) { return g !== gene; }); renderTags(); applyHighlight(); };
    function applyHighlight() {
        var plotDiv = document.getElementById("volcano_plot_" + cid);
        if (!plotDiv || !plotDiv.data) return;
        if (selectedGenes.length === 0) { Plotly.relayout(plotDiv, { annotations: [] }); return; }
        var genesLower = selectedGenes.map(function(g) { return g.toLowerCase(); });
        var annotations = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var data = plotDiv.data[t];
            if (!data.text) continue;
            for (var i = 0; i < data.text.length; i++) {
                var parts = data.text[i].split(/<br\\s*\\/?>/i);
                if (parts.length > 0) {
                    var geneName = parts[0].replace(/Protein:\\s*/i, "").trim();
                    if (genesLower.indexOf(geneName.toLowerCase()) !== -1) {
                        annotations.push({ x: data.x[i], y: data.y[i], text: geneName, showarrow: true, arrowhead: 2, ax: 20, ay: -30, font: { size: 12, color: "black" }, bgcolor: "rgba(255,255,255,0.9)", bordercolor: "black", borderwidth: 1 });
                    }
                }
            }
        }
        Plotly.relayout(plotDiv, { annotations: annotations });
    }
    window.highlightVolcanoGenes_%s = function() {
        var input = document.getElementById("volcano_genes_" + cid).value;
        var manual = input.split(",").map(function(s) { return s.trim(); }).filter(function(s) { return s.length > 0; });
        var allGenes = getGeneNames();
        manual.forEach(function(m) { var mLower = m.toLowerCase(); var match = allGenes.find(function(g) { return g.toLowerCase() === mLower; }); if (match && selectedGenes.indexOf(match) === -1) selectedGenes.push(match); });
        document.getElementById("volcano_genes_" + cid).value = "";
        document.getElementById("volcano_ac_" + cid).style.display = "none";
        renderTags(); applyHighlight();
    };
    window.clearVolcanoHighlight_%s = function() {
        selectedGenes = [];
        document.getElementById("volcano_genes_" + cid).value = "";
        renderTags();
        var plotDiv = document.getElementById("volcano_plot_" + cid);
        if (plotDiv) Plotly.relayout(plotDiv, { annotations: [] });
    };
    document.addEventListener("click", function(e) {
        var acDiv = document.getElementById("volcano_ac_" + cid);
        var input = document.getElementById("volcano_genes_" + cid);
        if (acDiv && e.target !== input && !acDiv.contains(e.target)) acDiv.style.display = "none";
    });
})();
</script>
', clean_id, clean_id, clean_id, clean_id, clean_id))
        }
        cat("\n\n")
    }
} else {
    cat("Interactive plots require DE data.\n")
}
```

## MA Plots {.tabset}

### Static {.tabset}

```{r ma-plots-static, results='asis'}
ma_files <- list.files(qc_post_dir, pattern = "^ma_.*\\.png$", full.names = TRUE)
if (length(ma_files) > 0) {
    for (mf in ma_files) {
        cn <- gsub("^ma_|\\.png$", "", basename(mf))
        contrast_clean <- gsub("_", " ", cn)
        cat(sprintf("\n#### %s\n\n", contrast_clean))
        cat(sprintf("![MA -- %s](%s)\n\n", contrast_clean, mf))
        figure_legend(sprintf("MA plot for the %s contrast. Red: significantly up-regulated; blue: significantly down-regulated.", contrast_clean))
        render_commentary(paste0("ma_", cn))
    }
} else {
    cat("No MA plots found.\n")
}
```

### Interactive {.tabset}

MA plots show the relationship between average expression (A, x-axis) and log-fold-change (M, y-axis). They help assess whether fold-change estimates are independent of expression level and whether the data is properly normalized.

```{r ma-interactive-setup, include=FALSE}
make_ma_plotly_prot <- function(df, contrast, padj_cut = 0.05, lfc_cut = log2(1.5), title = NULL) {
    padj_col <- paste0("padj.imputs.", contrast)
    lfc_col <- paste0("linearFC.imputs.", contrast)
    if (!(padj_col %in% names(df)) || !(lfc_col %in% names(df))) return(NULL)

    id_col <- prot_cfg$de_table$id_col %||% "FeatureID"
    gene_col <- if ("Genes" %in% names(df)) "Genes" else if (id_col %in% names(df)) id_col else names(df)[1]

    # Calculate baseMean from numeric non-stat columns
    stat_patterns <- "^(padj|pvalue|linearFC|upDown|pass|FeatureID|Protein\\.Names|Genes|First\\.Protein)"
    expr_cols <- names(df)[!grepl(stat_patterns, names(df))]
    expr_cols <- expr_cols[sapply(df[expr_cols], is.numeric)]

    if (length(expr_cols) == 0) return(NULL)
    baseMean <- rowMeans(as.matrix(df[, expr_cols, drop = FALSE]), na.rm = TRUE)

    plot_df <- data.frame(
        protein = df[[gene_col]],
        logFC = log2(abs(as.numeric(df[[lfc_col]]))) * sign(as.numeric(df[[lfc_col]])),
        padj = as.numeric(df[[padj_col]]),
        baseMean = baseMean,
        stringsAsFactors = FALSE
    )
    plot_df <- plot_df[!is.na(plot_df$baseMean) & plot_df$baseMean > 0, ]
    plot_df$log10baseMean <- log10(plot_df$baseMean)
    plot_df$Direction <- ifelse(
        !is.na(plot_df$padj) & plot_df$padj <= padj_cut & abs(plot_df$logFC) >= lfc_cut,
        ifelse(plot_df$logFC > 0, "Up", "Down"), "NS"
    )
    plot_df$Direction <- factor(plot_df$Direction, levels = c("NS", "Down", "Up"))

    p <- ggplot(plot_df, aes(x = log10baseMean, y = logFC, color = Direction,
                              text = paste0("Protein: ", protein, "\nlog2FC: ", round(logFC, 3),
                                           "\nbaseMean: ", round(baseMean, 1),
                                           "\npadj: ", signif(padj, 3)))) +
        geom_point(alpha = 0.6, size = 1.5) +
        scale_color_manual(values = c("NS" = "grey70", "Down" = "blue", "Up" = "red")) +
        geom_hline(yintercept = c(-lfc_cut, lfc_cut), linetype = "dashed", alpha = 0.5) +
        geom_hline(yintercept = 0, color = "black", linewidth = 0.3) +
        labs(title = title, x = "log10(Mean Intensity)", y = "log2 Fold Change") +
        theme_minimal()
    ggplotly(p, tooltip = "text")
}
```

```{r ma-interactive, results='asis', fig.height=6}
if (!is.null(de_data) && length(contrast_names) > 0) {
    for (i in seq_along(contrast_names)) {
        cn <- contrast_names[i]
        contrast_clean <- gsub("_", " ", cn)
        clean_id <- gsub("[^a-zA-Z0-9]", "_", cn)

        cat(sprintf("\n#### %s\n\n", contrast_clean))

        p <- make_ma_plotly_prot(de_data, cn, prot_cfg$de$p_cutoff %||% 0.05,
                                  log2(prot_cfg$de$linear_fc_cutoff %||% 1.5), contrast_clean)
        if (!is.null(p)) {
            p$elementId <- paste0("ma_plot_", clean_id)
            print(htmltools::tagList(p))
        }
        cat("\n\n")
    }
} else {
    cat("Interactive MA plots require DE data.\n")
}
```

## Top DE Heatmaps {.tabset}

```{r top-de-heatmaps, results='asis'}
heatmap_files <- list.files(qc_post_dir, pattern = "^heatmap_top.*\\.png$", full.names = TRUE)
if (length(heatmap_files) > 0) {
    # Parse contrast names and sizes
    for (hf in heatmap_files) {
        label <- gsub("^heatmap_|\\.png$", "", basename(hf))
        label_clean <- gsub("_", " ", label)
        cat(sprintf("\n### %s\n\n", label_clean))
        cat(sprintf("![%s](%s)\n\n", label_clean, hf))
        figure_legend(sprintf("Heatmap of %s across samples. Rows are z-score normalized. Samples are annotated by experimental group.", label_clean))
        render_commentary(paste0("heatmap_", label))
    }
} else {
    cat("No top DE heatmaps found. Enable with `qc.top_de_heatmap_sizes` in config.\n")
}
```

## Protein Expression Explorer

The expression explorer provides an interactive interface for examining individual protein expression patterns across experimental groups. Box plots show the distribution of expression values for each condition, enabling visual assessment of effect sizes and within-group variability.

```{r protein-explorer-data, include=FALSE}
explorer_ready <- FALSE
explorer_json_str <- "{}"

# Load expression data: use final_results.tsv which has per-sample columns + annotations
meta_file_exp <- file.path(cfg$project$dir, cfg$paths$raw, prot_cfg$files$metadata)
meta_df_exp <- if (file.exists(meta_file_exp)) read.csv(meta_file_exp, stringsAsFactors = FALSE) else NULL
group_col_exp <- prot_cfg$effects$color %||% "Condition"
sample_col_exp <- prot_cfg$effects$samples %||% "SampleID"

norm_expr_exp <- NULL

if (!is.null(meta_df_exp) && sample_col_exp %in% names(meta_df_exp)) {
    meta_sample_ids <- as.character(meta_df_exp[[sample_col_exp]])

    # Primary: final_results.tsv (has expression + annotation columns)
    final_file <- file.path(datasets_dir, "final_results.tsv")
    if (file.exists(final_file)) {
        final_df <- tryCatch(
            read.delim(final_file, stringsAsFactors = FALSE, check.names = FALSE),
            error = function(e) NULL
        )
        if (!is.null(final_df)) {
            id_col_exp <- prot_cfg$de_table$id_col %||% "FeatureID"
            # Extract only sample columns that match metadata
            expr_cols_exp <- intersect(meta_sample_ids, names(final_df))
            if (length(expr_cols_exp) > 0 && id_col_exp %in% names(final_df)) {
                norm_expr_exp <- final_df[, expr_cols_exp, drop = FALSE]
                rownames(norm_expr_exp) <- final_df[[id_col_exp]]
            }
        }
    }

    # Fallback: imputed expression file
    if (is.null(norm_expr_exp)) {
        imp_files <- list.files(datasets_dir, pattern = "imputed.*\\.tsv$", full.names = TRUE)
        if (length(imp_files) > 0) {
            norm_expr_exp <- tryCatch(
                read.delim(imp_files[1], stringsAsFactors = FALSE, row.names = 1, check.names = FALSE),
                error = function(e) NULL
            )
            if (!is.null(norm_expr_exp)) {
                # Keep only columns matching metadata sample IDs
                keep <- intersect(meta_sample_ids, colnames(norm_expr_exp))
                if (length(keep) > 0) norm_expr_exp <- norm_expr_exp[, keep, drop = FALSE]
                else norm_expr_exp <- NULL
            }
        }
    }
}

if (!is.null(norm_expr_exp) && !is.null(meta_df_exp) && ncol(norm_expr_exp) >= 2) {
    protein_ids <- rownames(norm_expr_exp)
    protein_display <- protein_ids

    # Use Genes column from de_data if available
    if (!is.null(de_data) && "Genes" %in% names(de_data)) {
        id_col_map <- prot_cfg$de_table$id_col %||% "FeatureID"
        if (id_col_map %in% names(de_data)) {
            gene_map <- setNames(de_data$Genes, de_data[[id_col_map]])
            mapped <- gene_map[protein_ids]
            has_map <- !is.na(mapped) & nzchar(mapped)
            protein_display[has_map] <- mapped[has_map]
        }
    }

    # Build sample -> group mapping
    if (sample_col_exp %in% names(meta_df_exp) && group_col_exp %in% names(meta_df_exp)) {
        group_map_exp <- setNames(as.character(meta_df_exp[[group_col_exp]]), meta_df_exp[[sample_col_exp]])
        sample_groups_exp <- as.character(group_map_exp[colnames(norm_expr_exp)])
        sample_groups_exp[is.na(sample_groups_exp)] <- "Unknown"
    } else {
        sample_groups_exp <- rep("All", ncol(norm_expr_exp))
    }

    expr_matrix_list <- lapply(seq_len(nrow(norm_expr_exp)), function(i) {
        round(as.numeric(norm_expr_exp[i, ]), 3)
    })

    explorer_data <- list(
        genes = protein_display,
        samples = colnames(norm_expr_exp),
        groups = sample_groups_exp,
        expr = expr_matrix_list
    )
    explorer_json_str <- jsonlite::toJSON(explorer_data, auto_unbox = FALSE)
    explorer_ready <- TRUE
}
```

```{r protein-explorer-ui, eval=explorer_ready, results='asis'}
cat('
<div style="margin: 15px 0 20px 0;">
<div style="display: flex; align-items: center; gap: 10px; flex-wrap: wrap;">
<label for="boxplot_gene_input" style="font-weight: bold; white-space: nowrap;">Search protein:</label>
<div style="position: relative; flex: 1; max-width: 400px;">
<input type="text" id="boxplot_gene_input" placeholder="Start typing a protein name..."
  autocomplete="off"
  style="padding: 10px 12px; border-radius: 6px; border: 1px solid #ccc; width: 100%; font-size: 14px; box-sizing: border-box;"
  oninput="boxplotAutocomplete(this)" onkeydown="boxplotAutoKey(event)">
<div id="boxplot_ac" style="position: absolute; top: 100%; left: 0; width: 100%; max-height: 220px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 6px 6px; z-index: 1000; display: none; box-shadow: 0 4px 8px rgba(0,0,0,0.12);"></div>
</div>
<button onclick="addGeneToPlot()" style="padding: 10px 20px; background: #007bff; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Add Protein</button>
<button onclick="clearBoxplot()" style="padding: 10px 20px; background: #6c757d; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Clear All</button>
</div>
<div id="boxplot_gene_tags" style="margin-top: 8px; display: flex; flex-wrap: wrap; gap: 6px;"></div>
<div id="boxplot_gene_info" style="margin-top: 8px; font-size: 0.9em; color: #555;"></div>
</div>
<div id="boxplot_container" style="width: 100%; min-height: 450px; margin-bottom: 10px;"></div>
<div id="boxplot_export_bar" style="display: none; margin-bottom: 20px; gap: 8px; align-items: center;">
  <span style="font-size: 0.85em; color: #555; margin-right: 4px;">Export:</span>
  <button onclick="exportBoxplot(\'svg\')" style="padding: 5px 14px; background: #28a745; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">SVG</button>
  <button onclick="exportBoxplot(\'png\')" style="padding: 5px 14px; background: #17a2b8; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">PNG</button>
</div>
')
```

```{r protein-explorer-js, eval=explorer_ready, results='asis'}
cat('<script>\nvar geneExprData = ')
cat(explorer_json_str)
cat(';\n')
cat(sprintf('var bpGroupLabel = "%s";\n', group_col_exp))
cat('
var boxplotAcIndex = -1;
var selectedGenes = [];

function boxplotAutocomplete(el) {
    var acDiv = document.getElementById("boxplot_ac");
    var val = el.value.trim().toLowerCase();
    acDiv.innerHTML = ""; boxplotAcIndex = -1;
    if (val.length < 1) { acDiv.style.display = "none"; return; }
    var matches = [];
    for (var i = 0; i < geneExprData.genes.length; i++) {
        if (geneExprData.genes[i].toLowerCase().indexOf(val) !== -1) {
            matches.push({name: geneExprData.genes[i], idx: i});
            if (matches.length >= 20) break;
        }
    }
    if (matches.length === 0) { acDiv.style.display = "none"; return; }
    matches.forEach(function(m) {
        var item = document.createElement("div");
        item.textContent = m.name;
        item.style.cssText = "padding: 8px 12px; cursor: pointer; font-size: 13px; border-bottom: 1px solid #f0f0f0;";
        item.onmouseenter = function() { this.style.background = "#e9ecef"; };
        item.onmouseleave = function() { this.style.background = "white"; };
        item.onclick = function() { document.getElementById("boxplot_gene_input").value = m.name; acDiv.style.display = "none"; addGeneToPlot(); };
        acDiv.appendChild(item);
    });
    acDiv.style.display = "block";
}

function boxplotAutoKey(e) {
    var acDiv = document.getElementById("boxplot_ac");
    var items = acDiv.children;
    if (items.length === 0) { if (e.key === "Enter") addGeneToPlot(); return; }
    if (e.key === "ArrowDown") { e.preventDefault(); boxplotAcIndex = Math.min(boxplotAcIndex + 1, items.length - 1); updateBpAcHighlight(items); }
    else if (e.key === "ArrowUp") { e.preventDefault(); boxplotAcIndex = Math.max(boxplotAcIndex - 1, 0); updateBpAcHighlight(items); }
    else if (e.key === "Enter") { e.preventDefault(); if (boxplotAcIndex >= 0) items[boxplotAcIndex].click(); else addGeneToPlot(); }
}

function updateBpAcHighlight(items) { for (var i = 0; i < items.length; i++) items[i].style.background = i === boxplotAcIndex ? "#e9ecef" : "white"; }

function addGeneToPlot() {
    var input = document.getElementById("boxplot_gene_input");
    var val = input.value.trim();
    if (!val) return;
    var valLower = val.toLowerCase();
    var found = -1;
    for (var i = 0; i < geneExprData.genes.length; i++) {
        if (geneExprData.genes[i].toLowerCase() === valLower) { found = i; break; }
    }
    if (found === -1) {
        for (var i = 0; i < geneExprData.genes.length; i++) {
            if (geneExprData.genes[i].toLowerCase().indexOf(valLower) !== -1) { found = i; break; }
        }
    }
    if (found === -1) { document.getElementById("boxplot_gene_info").textContent = "Protein not found: " + val; return; }
    if (selectedGenes.some(function(g) { return g.idx === found; })) { document.getElementById("boxplot_gene_info").textContent = "Already added."; input.value = ""; return; }
    selectedGenes.push({name: geneExprData.genes[found], idx: found});
    input.value = "";
    document.getElementById("boxplot_ac").style.display = "none";
    document.getElementById("boxplot_gene_info").textContent = "";
    renderBoxplotTags();
    renderMultiBoxplot();
}

function renderBoxplotTags() {
    var container = document.getElementById("boxplot_gene_tags");
    container.innerHTML = "";
    selectedGenes.forEach(function(g, i) {
        var tag = document.createElement("span");
        tag.style.cssText = "display: inline-flex; align-items: center; background: #007bff; color: white; padding: 3px 10px; border-radius: 14px; font-size: 13px; gap: 5px;";
        tag.innerHTML = g.name + " <span style=\\"cursor:pointer; font-weight:bold;\\" onclick=\\"removeBoxplotGene(" + i + ")\\">&times;</span>";
        container.appendChild(tag);
    });
}

function removeBoxplotGene(idx) {
    selectedGenes.splice(idx, 1);
    renderBoxplotTags();
    if (selectedGenes.length > 0) { renderMultiBoxplot(); }
    else { Plotly.purge("boxplot_container"); document.getElementById("boxplot_export_bar").style.display = "none"; document.getElementById("boxplot_gene_info").textContent = ""; }
}
function clearBoxplot() {
    selectedGenes = [];
    renderBoxplotTags();
    Plotly.purge("boxplot_container");
    document.getElementById("boxplot_gene_info").textContent = "";
    document.getElementById("boxplot_export_bar").style.display = "none";
}

function renderMultiBoxplot() {
    if (selectedGenes.length === 0) return;

    var groups = geneExprData.groups;
    var samples = geneExprData.samples;
    var uniqueGroups = [];
    groups.forEach(function(g) { if (uniqueGroups.indexOf(g) === -1) uniqueGroups.push(g); });

    var colors = ["#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00", "#a65628", "#f781bf", "#999999"];
    var nGenes = selectedGenes.length;
    var traces = [];

    if (nGenes === 1) {
        // Single protein: one box per group, distinct colors
        var geneIdx = selectedGenes[0].idx;
        var values = geneExprData.expr[geneIdx];
        uniqueGroups.forEach(function(g, gi) {
            var gValues = [], gSamples = [];
            for (var i = 0; i < values.length; i++) {
                if (groups[i] === g) { gValues.push(values[i]); gSamples.push(samples[i]); }
            }
            traces.push({
                y: gValues, type: "box", name: g,
                boxpoints: "all", jitter: 0.4, pointpos: 0,
                text: gSamples,
                hovertemplate: "Sample: %{text}<br>Expression: %{y:.3f}<extra>" + g + "</extra>",
                marker: { color: colors[gi % colors.length], size: 6, opacity: 0.7 },
                line: { color: colors[gi % colors.length] },
                fillcolor: colors[gi % colors.length] + "33"
            });
        });
        var layout = {
            title: { text: selectedGenes[0].name + " Expression", font: { size: 16 } },
            yaxis: { title: "Expression (log2 intensity)" },
            xaxis: { title: bpGroupLabel },
            showlegend: false,
            margin: { t: 50, b: 80, l: 70, r: 30 },
            boxgap: 0.3
        };
        var containerH = 450;
    } else {
        // Multiple proteins: genes on x-axis, grouped by condition
        var containerH = Math.max(450, 350 + nGenes * 30);
        uniqueGroups.forEach(function(g, gi) {
            var xVals = [], yVals = [], sampleTexts = [];
            selectedGenes.forEach(function(sg) {
                var values = geneExprData.expr[sg.idx];
                for (var i = 0; i < values.length; i++) {
                    if (groups[i] === g) {
                        xVals.push(sg.name);
                        yVals.push(values[i]);
                        sampleTexts.push(samples[i]);
                    }
                }
            });
            traces.push({
                x: xVals, y: yVals, type: "box", name: g,
                boxpoints: "all", jitter: 0.4, pointpos: 0,
                text: sampleTexts,
                hovertemplate: "%{x}<br>Sample: %{text}<br>Expression: %{y:.3f}<extra>" + g + "</extra>",
                marker: { color: colors[gi % colors.length], size: 5, opacity: 0.7 },
                line: { color: colors[gi % colors.length] },
                fillcolor: colors[gi % colors.length] + "33"
            });
        });

        var protNames = selectedGenes.map(function(sg) { return sg.name; });
        var layout = {
            title: { text: "Protein Expression Comparison", font: { size: 16 } },
            yaxis: { title: "Expression (log2 intensity)" },
            xaxis: { title: "Protein", categoryorder: "array", categoryarray: protNames },
            showlegend: true,
            legend: { title: { text: bpGroupLabel } },
            boxmode: "group",
            margin: { t: 50, b: 100, l: 70, r: 30 },
            boxgap: 0.15,
            boxgroupgap: 0.1
        };
    }

    document.getElementById("boxplot_container").style.height = containerH + "px";
    Plotly.newPlot("boxplot_container", traces, layout, { responsive: true });

    // Show info
    var infoLines = selectedGenes.map(function(sg) {
        var vals = geneExprData.expr[sg.idx].filter(function(v) { return !isNaN(v); });
        var mean = vals.reduce(function(a, b) { return a + b; }, 0) / vals.length;
        return "<strong>" + sg.name + "</strong>: mean " + mean.toFixed(3) + " (" + vals.length + " samples)";
    });
    document.getElementById("boxplot_gene_info").innerHTML = infoLines.join(" &nbsp;|&nbsp; ");
    document.getElementById("boxplot_export_bar").style.display = "flex";
}

function exportBoxplot(format) {
    var container = document.getElementById("boxplot_container");
    if (!container || !container.data) return;
    var names = selectedGenes.map(function(g) { return g.name; }).join("_");
    if (format === "svg") Plotly.downloadImage(container, {format: "svg", width: 1200, height: 800, filename: "protein_expression_" + names});
    else Plotly.downloadImage(container, {format: "png", width: 2400, height: 1600, scale: 3, filename: "protein_expression_" + names});
}

document.addEventListener("click", function(e) {
    var acDiv = document.getElementById("boxplot_ac");
    var input = document.getElementById("boxplot_gene_input");
    if (acDiv && e.target !== input && !acDiv.contains(e.target)) acDiv.style.display = "none";
});
</script>
')
```

```{r protein-explorer-fallback, eval=!explorer_ready, results='asis'}
cat("*Protein expression explorer not available. Normalized expression matrix not found.*\n")
```

---

```{r clustering-check, include=FALSE}
clustering_enabled <- isTRUE(prot_cfg$clustering$enabled)
```

```{r clustering-header, eval=clustering_enabled, results='asis'}
cat("# Clustering\n\n")
cat("Clustering was performed on differentially expressed proteins.\n\n")
```

```{r clustering-hierarchical, eval=clustering_enabled, results='asis', fig.cap="Hierarchical clustering heatmap"}
hm_file <- file.path(clustering_dir, "hierarchical_heatmap.png")
if (file.exists(hm_file)) {
    cat("## Hierarchical Clustering\n\n")
    cat(sprintf("![Hierarchical clustering](%s)\n\n", hm_file))
}
```

```{r clustering-partition, eval=clustering_enabled, results='asis'}
part_hm <- file.path(clustering_dir, "partition_heatmap.png")
part_profiles <- file.path(clustering_dir, "partition_profiles.png")
if (file.exists(part_hm) || file.exists(part_profiles)) {
    cat("## Partition Clustering {.tabset}\n\n")
    if (file.exists(part_hm)) {
        cat("### Heatmap\n\n")
        cat(sprintf("![Partition heatmap](%s)\n\n", part_hm))
    }
    if (file.exists(part_profiles)) {
        cat("### Cluster Profiles\n\n")
        cat(sprintf("![Partition profiles](%s)\n\n", part_profiles))
    }
}
```

---

# Enrichment / Pathway Analysis

Gene set enrichment analysis tests whether predefined biological pathways show coordinated expression changes beyond what is expected by chance. fGSEA uses a ranked list of all genes (rank-based), while ORA tests only the significant gene list against a background (threshold-based).

```{r pathway-setup, include=FALSE}
has_enrichment <- dir.exists(enrich_dir) && length(list.files(enrich_dir, pattern = "\\.csv$")) > 0
has_pathway_plots <- dir.exists(enrich_plot_dir) && length(list.files(enrich_plot_dir, pattern = "\\.png$")) > 0
pathway_enabled <- isTRUE(prot_cfg$pathway$enabled)
```

```{r pathway-skip, eval=!has_enrichment, results='asis'}
if (!pathway_enabled) {
    cat("Pathway analysis was disabled. To enable it, set `pathway.enabled: true` in the configuration file.\n")
} else {
    cat("Pathway analysis was enabled but produced no results.\n")
}
```

```{r pathway-method-note, eval=has_enrichment, results='asis'}
cat("**Method:** Pathway enrichment analysis was performed using gene set enrichment approaches. Proteins were mapped to gene symbols for compatibility with standard pathway databases.\n\n")
cat("**Key metrics:**\n\n")
cat("- **NES (Normalized Enrichment Score)**: Positive NES = enrichment in up-regulated proteins; negative NES = enrichment in down-regulated proteins.\n")
cat("- **padj**: Benjamini-Hochberg adjusted p-value. Pathways with padj < 0.05 are significant.\n")
cat("- **size**: Number of mapped proteins in the pathway.\n\n")
```

## fGSEA Results {.tabset}

```{r fgsea-caption, eval=has_enrichment, results='asis'}
cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Fast Gene Set Enrichment Analysis (fGSEA) results. NES &gt; 0 indicates enrichment among up-regulated proteins; NES &lt; 0 indicates enrichment among down-regulated proteins. Pathways with padj &lt; 0.05 are shown.</em></p>\n")
```

```{r fgsea-tables, eval=has_enrichment, results='asis'}
fgsea_files <- list.files(enrich_dir, pattern = "pathway_.*_fgsea\\.csv$", full.names = TRUE)
if (length(fgsea_files) > 0) {
    for (ff in fgsea_files) {
        label <- gsub("^pathway_|_fgsea\\.csv$", "", basename(ff))
        label_clean <- gsub("_", " ", label)

        fgsea_df <- read.csv(ff, stringsAsFactors = FALSE)
        sig <- fgsea_df[!is.na(fgsea_df$padj) & fgsea_df$padj < 0.05, ]

        cat(sprintf("\n### %s {.tabset}\n\n", label_clean))

        if (nrow(sig) > 0) {
            sig <- sig[order(sig$padj), ]
            if ("pathway_name" %in% names(sig)) {
                show_cols <- intersect(c("pathway_name", "pval", "padj", "NES", "size"), names(sig))
            } else {
                show_cols <- intersect(c("pathway", "pval", "padj", "NES", "size"), names(sig))
            }
            sig_show <- sig[, show_cols]
            sig_show$Direction <- ifelse(sig_show$NES > 0, "Up-regulated", "Down-regulated")
            n_up <- sum(sig_show$Direction == "Up-regulated")
            n_down <- sum(sig_show$Direction == "Down-regulated")

            cat(sprintf("\n#### All (%d pathways)\n\n", nrow(sig_show)))
            cat(sprintf("**Up-regulated: %d** | **Down-regulated: %d**\n\n", n_up, n_down))
            print(htmltools::tagList(make_dt(sig_show, paste("All significant pathways -", label_clean))))
            cat("\n\n")

            if (n_up > 0) {
                cat(sprintf("\n#### Up-regulated (%d)\n\n", n_up))
                sig_up <- sig_show[sig_show$Direction == "Up-regulated", ]
                print(htmltools::tagList(make_dt(sig_up[order(sig_up$padj), ], paste("Up-regulated -", label_clean))))
                cat("\n\n")
            }

            if (n_down > 0) {
                cat(sprintf("\n#### Down-regulated (%d)\n\n", n_down))
                sig_down <- sig_show[sig_show$Direction == "Down-regulated", ]
                print(htmltools::tagList(make_dt(sig_down[order(sig_down$padj), ], paste("Down-regulated -", label_clean))))
                cat("\n\n")
            }
        } else {
            cat("No significant pathways (padj < 0.05) for this contrast/database.\n\n")
        }
    }
} else if (has_enrichment) {
    cat("No fGSEA result files found.\n")
}
```

## ORA Results {.tabset}

Over-Representation Analysis (ORA) tests whether differentially expressed proteins are enriched in specific pathways compared to the background proteome. Unlike fGSEA, ORA uses a binary classification (significant vs. not significant) and applies a hypergeometric test.

```{r ora-caption, eval=has_enrichment, results='asis'}
cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Over-Representation Analysis results. Pathways with padj &lt; 0.05 are shown, ranked by significance.</em></p>\n")
```

```{r ora-tables, eval=has_enrichment, results='asis'}
ora_files <- list.files(enrich_dir, pattern = "pathway_.*_ora.*\\.csv$", full.names = TRUE)
if (length(ora_files) > 0) {
    for (of in ora_files) {
        label <- gsub("^pathway_|_ora(_up|_down)?\\.csv$", "", basename(of))
        label_clean <- gsub("_", " ", label)

        ora_df <- read.csv(of, stringsAsFactors = FALSE)
        # Support both column names: padj (our pipeline) and p.adjust (clusterProfiler)
        padj_col <- if ("padj" %in% names(ora_df)) "padj" else "p.adjust"
        sig <- ora_df[!is.na(ora_df[[padj_col]]) & ora_df[[padj_col]] < 0.05, ]

        cat(sprintf("\n### %s\n\n", label_clean))

        if (nrow(sig) > 0) {
            sig <- sig[order(sig[[padj_col]]), ]
            # Support both clusterProfiler format (Description) and our pipeline format (pathway_name)
            if ("pathway_name" %in% names(sig)) {
                show_cols <- intersect(c("pathway_name", "size", "overlap", "pvalue", "padj", "odds_ratio"), names(sig))
            } else {
                show_cols <- intersect(c("Description", "GeneRatio", "BgRatio", "pvalue", "p.adjust", "Count"), names(sig))
            }
            if (length(show_cols) > 0) {
                print(htmltools::tagList(make_dt(head(sig[, show_cols], 30), paste("ORA -", label_clean))))
            }
        } else {
            cat("No significant pathways (padj < 0.05).\n\n")
        }
    }
}
```

## Pathway Dotplots {.tabset}

```{r pathway-dotplots, eval=has_pathway_plots, results='asis'}
plot_pngs <- list.files(enrich_plot_dir, pattern = "\\.png$", full.names = TRUE)
if (length(plot_pngs) > 0) {
    for (pp in plot_pngs) {
        label_raw <- gsub("^pathway_|\\.png$", "", basename(pp))
        label <- gsub("_", " ", label_raw)
        cat(sprintf("\n### %s\n\n", label))
        cat(sprintf("![%s](%s)\n\n", label, pp))
        figure_legend(sprintf("Dotplot of top enriched pathways for %s. Dot size represents the number of pathway members; color intensity reflects statistical significance (adjusted p-value).", label))
        render_commentary(paste0("pathway_", label_raw))
    }
}
```

## Leading Edge Genes {.tabset}

Leading edge genes are the core subset of pathway members that drive the enrichment signal. They represent the genes most responsible for a pathway's statistical significance and are candidates for validation experiments.

```{r leading-edge-caption, eval=has_enrichment, results='asis'}
cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Leading edge proteins for the top 10 most significant pathways per contrast. These proteins drive the enrichment signal and represent the strongest candidates for follow-up.</em></p>\n")
```

```{r leading-edge, eval=has_enrichment, results='asis'}
fgsea_files <- list.files(enrich_dir, pattern = "pathway_.*_fgsea\\.csv$", full.names = TRUE)
if (length(fgsea_files) > 0) {
    for (ff in fgsea_files) {
        label <- gsub("^pathway_|_fgsea\\.csv$", "", basename(ff))
        label_clean <- gsub("_", " ", label)
        cat(sprintf("\n### %s\n\n", label_clean))

        fgsea_df <- read.csv(ff, stringsAsFactors = FALSE)
        sig <- fgsea_df[!is.na(fgsea_df$padj) & fgsea_df$padj < 0.05, ]

        if (nrow(sig) > 0 && "leadingEdge" %in% names(sig)) {
            sig <- head(sig[order(sig$padj), ], 10)
            le_list <- list()
            for (i in seq_len(nrow(sig))) {
                pathway_name <- if ("pathway_name" %in% names(sig)) sig$pathway_name[i] else sig$pathway[i]
                le_genes <- strsplit(sig$leadingEdge[i], ",")[[1]]
                le_list[[i]] <- data.frame(
                    Pathway = pathway_name,
                    NES = round(sig$NES[i], 2),
                    padj = signif(sig$padj[i], 3),
                    `Leading Edge Proteins` = paste(head(le_genes, 15), collapse = ", "),
                    check.names = FALSE,
                    stringsAsFactors = FALSE
                )
            }
            le_df <- do.call(rbind, le_list)
            print(htmltools::tagList(make_dt(le_df, paste("Leading Edge Proteins -", label_clean))))
            cat("\n\n")
        } else {
            cat("No significant pathways with leading edge information.\n\n")
        }
    }
}
```

## Pathway Volcano {.tabset}

```{r pathway-volcano-setup, include=FALSE}
pw_volcano_files <- list.files(enrich_dir, pattern = "^pathway_volcano_data_.*\\.csv$", full.names = TRUE)
has_pathway_volcano <- length(pw_volcano_files) > 0
if (has_pathway_volcano) {
    # Verify at least one file has a populated pathways column
    has_pathway_volcano <- any(vapply(pw_volcano_files, function(f) {
        df <- read.csv(f, stringsAsFactors = FALSE, nrows = 5)
        "pathways" %in% colnames(df) && any(nzchar(df$pathways), na.rm = TRUE)
    }, logical(1)))
}
```

```{r pathway-volcano-skip, eval=!has_pathway_volcano, results='asis'}
cat("No pathway volcano data available. Enable with `pathway_volcano: yes` in the pathway configuration and ensure pathway analysis produces significant results.\n")
```

```{r pathway-volcano-legend, eval=has_pathway_volcano, results='asis'}
figure_legend("Pathway volcano plots overlay enriched pathway membership onto the standard DE volcano plot. Select a pathway from the dropdown to highlight its member proteins in orange. Non-member proteins are dimmed for contrast.")
```

```{r pathway-volcano-plots, eval=has_pathway_volcano, results='asis'}
for (vf in pw_volcano_files) {
    contrast_name <- gsub("^pathway_volcano_data_|\\.csv$", "", basename(vf))
    contrast_clean <- gsub("_", " ", contrast_name)
    clean_id <- gsub("[^a-zA-Z0-9]", "_", contrast_name)

    volcano_df <- read.csv(vf, stringsAsFactors = FALSE)
    if (!"pathways" %in% colnames(volcano_df)) next

    # Cap -log10(p) at 99.5th percentile to keep plot readable
    yvals <- volcano_df$neg_log10_pvalue[is.finite(volcano_df$neg_log10_pvalue)]
    ycap <- max(quantile(yvals, 0.995, na.rm = TRUE), 10)
    volcano_df$neg_log10_pvalue <- pmin(volcano_df$neg_log10_pvalue, ycap)

    cat(sprintf("\n### %s\n\n", contrast_clean))

    # Collect fGSEA info for NES / padj display
    fgsea_files_for_contrast <- list.files(enrich_dir,
        pattern = paste0("pathway_", contrast_name, "_.*_fgsea\\.csv$"),
        full.names = TRUE)
    pathway_info <- list()
    for (ff in fgsea_files_for_contrast) {
        fdf <- read.csv(ff, stringsAsFactors = FALSE)
        sig <- fdf[!is.na(fdf$padj) & fdf$padj < 0.05, , drop = FALSE]
        for (r in seq_len(nrow(sig))) {
            # Use pathway_name (human-readable) to match the volcano data keys
            pname <- if ("pathway_name" %in% names(sig) && nzchar(sig$pathway_name[r] %||% "")) {
                sig$pathway_name[r]
            } else {
                sig$pathway[r]
            }
            pathway_info[[pname]] <- list(
                NES = round(sig$NES[r], 3),
                padj = signif(sig$padj[r], 3),
                size = sig$size[r]
            )
        }
    }

    # Assign direction colors
    padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_cut <- log2(prot_cfg$de$linear_fc_cutoff %||% 1.5)
    volcano_df$Direction <- ifelse(
        is.na(volcano_df$padj) | volcano_df$padj >= padj_cut, "NS",
        ifelse(volcano_df$log2FC > lfc_cut, "Up",
               ifelse(volcano_df$log2FC < -lfc_cut, "Down", "NS")))

    dir_colors <- c("Up" = "#e74c3c", "Down" = "#3498db", "NS" = "#bdc3c7")

    p <- plotly::plot_ly(
        data = volcano_df,
        x = ~log2FC,
        y = ~neg_log10_pvalue,
        type = "scatter",
        mode = "markers",
        color = ~Direction,
        colors = dir_colors,
        marker = list(size = 5, opacity = 0.7),
        text = ~paste0("Protein: ", FeatureID,
                        "<br>log2FC: ", round(log2FC, 3),
                        "<br>padj: ", signif(padj, 3),
                        "<br>Pathways: ", ifelse(nzchar(pathways), pathways, "none")),
        hoverinfo = "text",
        customdata = ~FeatureID,
        source = paste0("pw_volcano_", clean_id)
    ) |>
        plotly::layout(
            title = list(text = paste("Pathway Volcano -", contrast_clean), font = list(size = 14)),
            xaxis = list(title = "log2 Fold Change"),
            yaxis = list(title = "-log10(p-value)", rangemode = "tozero"),
            shapes = list(
                list(type = "line", x0 = -lfc_cut, x1 = -lfc_cut, y0 = 0, y1 = 1,
                     yref = "paper", line = list(dash = "dash", color = "grey50")),
                list(type = "line", x0 = lfc_cut, x1 = lfc_cut, y0 = 0, y1 = 1,
                     yref = "paper", line = list(dash = "dash", color = "grey50")),
                list(type = "line", x0 = 0, x1 = 1, xref = "paper",
                     y0 = -log10(padj_cut), y1 = -log10(padj_cut),
                     line = list(dash = "dash", color = "grey50"))
            )
        ) |>
        plotly::config(
            toImageButtonOptions = list(format = "svg", filename = paste0("pathway_volcano_", clean_id),
                                         width = 1200, height = 900, scale = 3),
            modeBarButtonsToAdd = list(
                list(name = "Download PNG (300 dpi)", icon = list(
                    path = "M12 2C6.48 2 2 6.48 2 12s4.48 10 10 10 10-4.48 10-10S17.52 2 12 2zm-1 14H8l4-4 4 4h-3v3h-2v-3z",
                    width = 24, height = 24
                ), click = htmlwidgets::JS(sprintf(
                    "function(gd) { Plotly.downloadImage(gd, {format: 'png', width: 2400, height: 1800, scale: 3, filename: 'pathway_volcano_%s'}); }",
                    clean_id
                )))
            )
        )

    p$elementId <- paste0("pw_volcano_plot_", clean_id)

    # Build pathway→genes mapping from the pathways column
    all_pathways <- unique(unlist(strsplit(volcano_df$pathways[nzchar(volcano_df$pathways)], "; ")))
    all_pathways <- sort(all_pathways[nzchar(all_pathways)])

    pw_genes_map <- list()
    for (pw in all_pathways) {
        members <- as.character(volcano_df$FeatureID[grepl(pw, volcano_df$pathways, fixed = TRUE)])
        pw_genes_map[[pw]] <- I(members)
    }

    pw_genes_json <- jsonlite::toJSON(pw_genes_map, auto_unbox = TRUE)
    pw_info_json <- jsonlite::toJSON(pathway_info, auto_unbox = TRUE)

    # Build dropdown options
    options_html <- '<option value="">None (show all)</option>'
    for (pw in all_pathways) {
        pw_escaped <- gsub("\"", "&quot;", pw)
        info <- pathway_info[[pw]]
        label <- if (!is.null(info)) {
            sprintf("%s (NES=%.2f, padj=%.3g)", pw, info$NES, info$padj)
        } else { pw }
        label_escaped <- gsub("\"", "&quot;", label)
        options_html <- paste0(options_html,
            sprintf('<option value="%s">%s</option>', pw_escaped, label_escaped))
    }

    cat(sprintf('
<div style="margin-bottom: 10px;">
<label for="pw_select_%s"><strong>Highlight pathway:</strong></label>
<select id="pw_select_%s" style="padding: 6px; border-radius: 4px; border: 1px solid #ccc; margin-left: 8px; max-width: 600px;">
%s
</select>
<div id="pw_info_%s" style="margin-top: 6px; font-size: 13px; color: #555;"></div>
<div id="pw_legend_%s" style="margin-top: 4px; font-size: 13px;"></div>
</div>
', clean_id, clean_id, options_html, clean_id, clean_id))

    print(htmltools::tagList(p))

    cat(sprintf('
<script>
(function() {
    var cid = "%s";
    var plotId = "pw_volcano_plot_" + cid;
    var pwGenes = %s;
    var pwInfo = %s;
    var origColors = null;
    var origOpacities = null;

    function captureOriginal() {
        var plotDiv = document.getElementById(plotId);
        if (!plotDiv || !plotDiv.data) return;
        origColors = [];
        origOpacities = [];
        for (var t = 0; t < plotDiv.data.length; t++) {
            var mc = plotDiv.data[t].marker;
            origColors.push(mc ? (mc.color || null) : null);
            origOpacities.push(mc ? (mc.opacity || 0.7) : 0.7);
        }
    }

    function onSelect() {
        var plotDiv = document.getElementById(plotId);
        if (!plotDiv || !plotDiv.data) return;
        if (!origColors) captureOriginal();

        var pw = document.getElementById("pw_select_" + cid).value;
        var infoDiv = document.getElementById("pw_info_" + cid);
        var legendDiv = document.getElementById("pw_legend_" + cid);

        // Freeze axes so restyle cannot trigger auto-range
        var xRange = plotDiv.layout.xaxis.range ? plotDiv.layout.xaxis.range.slice() : null;
        var yRange = plotDiv.layout.yaxis.range ? plotDiv.layout.yaxis.range.slice() : null;
        if (xRange && yRange) {
            Plotly.relayout(plotDiv, {
                "xaxis.autorange": false, "xaxis.range": xRange,
                "yaxis.autorange": false, "yaxis.range": yRange
            });
        }

        if (!pw) {
            // Restore original
            for (var t = 0; t < plotDiv.data.length; t++) {
                Plotly.restyle(plotDiv, {
                    "marker.color": [origColors[t]],
                    "marker.opacity": [origOpacities[t]],
                    "marker.size": [5]
                }, [t]);
            }
            infoDiv.innerHTML = "";
            if (legendDiv) legendDiv.innerHTML = "";
            return;
        }

        var genes = pwGenes[pw] || [];
        if (typeof genes === "string") genes = [genes];
        var genesSet = {};
        for (var g = 0; g < genes.length; g++) genesSet[genes[g]] = true;

        // Show pathway info
        var info = pwInfo[pw];
        if (info) {
            infoDiv.innerHTML = "<strong>" + pw + "</strong> &mdash; NES: " +
                info.NES + " | padj: " + info.padj + " | Members in data: " + genes.length;
        } else {
            infoDiv.innerHTML = "<strong>" + pw + "</strong> &mdash; Members in data: " + genes.length;
        }
        if (legendDiv) {
            legendDiv.innerHTML = "<font color=#e67e22>&#9679;</font> Pathway member &nbsp;" +
                "<font color=#d5d8dc>&#9679;</font> Non-member";
        }

        // Color traces
        for (var t = 0; t < plotDiv.data.length; t++) {
            var cd = plotDiv.data[t].customdata;
            if (!cd) { continue; }
            var colors = [];
            var opacities = [];
            var sizes = [];
            for (var i = 0; i < cd.length; i++) {
                var fid = Array.isArray(cd[i]) ? cd[i][0] : cd[i];
                if (genesSet[fid]) {
                    colors.push("#e67e22");
                    opacities.push(0.9);
                    sizes.push(8);
                } else {
                    colors.push("#d5d8dc");
                    opacities.push(0.15);
                    sizes.push(4);
                }
            }
            Plotly.restyle(plotDiv, {
                "marker.color": [colors],
                "marker.opacity": [opacities],
                "marker.size": [sizes]
            }, [t]);
        }
    }

    // Attach event after plot renders
    var attempts = 0;
    var timer = setInterval(function() {
        var plotDiv = document.getElementById(plotId);
        if ((plotDiv && plotDiv.data) || attempts > 50) {
            clearInterval(timer);
            if (plotDiv && plotDiv.data) {
                captureOriginal();
                document.getElementById("pw_select_" + cid).addEventListener("change", onSelect);
            }
        }
        attempts++;
    }, 200);
})();
</script>
', clean_id, pw_genes_json, pw_info_json))

    cat("\n\n")
}
```

---

# PPI Network Analysis

```{r ppi-setup, include=FALSE}
has_ppi <- dir.exists(ppi_dir) && (
    length(list.files(ppi_dir, pattern = "\\.csv$")) > 0 ||
    length(list.files(ppi_plot_dir, pattern = "\\.png$")) > 0
)
ppi_enabled <- isTRUE(prot_cfg$ppi$enabled)
```

```{r ppi-skip, eval=!has_ppi, results='asis'}
if (!ppi_enabled) {
    cat("PPI network analysis was disabled. Enable with `ppi.enabled: true`.\n")
} else {
    cat("PPI analysis was enabled but produced no results.\n")
}
```

```{r ppi-explanation, eval=has_ppi, results='asis'}
cat("Protein-protein interaction networks map physical and functional associations between differentially expressed proteins. Network topology reveals hub proteins (highly connected), bottleneck proteins (bridging modules), and functional communities that may correspond to coordinated biological processes.\n\n")
```

```{r ppi-summary, eval=has_ppi, results='asis'}
cat("Protein-protein interaction networks were constructed using the STRING database.\n\n")

# Load summary if available
ppi_summary_file <- file.path(ppi_dir, "ppi_summary.csv")
if (file.exists(ppi_summary_file)) {
    ppi_summary <- read.csv(ppi_summary_file, stringsAsFactors = FALSE)
    make_dt(ppi_summary, caption = "PPI Network Summary")
}
```

## Network Plots {.tabset}

```{r ppi-plots, eval=has_ppi, results='asis'}
# Load STRING info for building web links
string_info_file <- file.path(ppi_dir, "string_info.csv")
string_proteins <- NULL
string_species <- 9606
if (file.exists(string_info_file)) {
    si <- read.csv(string_info_file, stringsAsFactors = FALSE)
    string_proteins <- si$protein
    string_species <- si$species_id[1]
}

if (dir.exists(ppi_plot_dir)) {
    ppi_pngs <- list.files(ppi_plot_dir, pattern = "\\.png$", full.names = TRUE)
    if (length(ppi_pngs) > 0) {
        for (pp in ppi_pngs) {
            label_raw <- gsub("\\.png$", "", basename(pp))
            label <- gsub("_", " ", label_raw)
            cat(sprintf("\n### %s\n\n", label))
            cat(sprintf("![%s](%s)\n\n", label, pp))
            figure_legend(sprintf("PPI network visualization: %s. Nodes represent proteins; edges represent known or predicted interactions from the STRING database. Node size and color may reflect connectivity or expression change.", label))
            cat('<div style="margin: 8px 0; font-size: 13px;">
  <strong>Node color:</strong>
  <span style="color: #3498db;">&#9679;</span> Down-regulated &nbsp;
  <span style="color: #ccc;">&#9679;</span> No change &nbsp;
  <span style="color: #e74c3c;">&#9679;</span> Up-regulated &nbsp;
  <em>(log2FC gradient, capped at &plusmn;2)</em>
</div>\n')
            # Add "View on STRING" link
            if (!is.null(string_proteins) && length(string_proteins) > 0) {
                string_url <- sprintf(
                    "https://string-db.org/cgi/network?identifiers=%s&species=%s",
                    paste(string_proteins, collapse = "%%0d"), string_species
                )
                cat(sprintf('<a href="%s" target="_blank" style="display: inline-block; margin: 6px 0 12px 0; padding: 6px 14px; background: #2c7fb8; color: white; border-radius: 4px; text-decoration: none; font-size: 13px;">View on STRING &rarr;</a>\n', string_url))
            }
            render_commentary(paste0("ppi_", label_raw))
        }
    } else {
        cat("No PPI plots found.\n")
    }
}
```

## Network Statistics

```{r ppi-topology, eval=has_ppi, results='asis'}
topo_file <- file.path(ppi_dir, "network_topology.csv")
if (file.exists(topo_file)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Global network topology metrics including node count, edge count, density, clustering coefficient, and characteristic path length.</em></p>\n")
    topo <- read.csv(topo_file, stringsAsFactors = FALSE)
    make_dt(topo, caption = "Network Topology Statistics")
}
```

## Hub Proteins

```{r ppi-hubs, eval=has_ppi, results='asis'}
hubs_file <- file.path(ppi_dir, "hub_proteins.csv")
if (file.exists(hubs_file)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Hub proteins ranked by network connectivity (degree). Hub proteins are highly connected nodes that may play central roles in the biological process under study.</em></p>\n")
    hubs <- read.csv(hubs_file, stringsAsFactors = FALSE)
    make_dt(hubs, caption = "Hub Proteins (top connectivity)")
}
```

## Communities

```{r ppi-communities, eval=has_ppi, results='asis'}
comm_file <- file.path(ppi_dir, "communities.csv")
if (file.exists(comm_file)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Network communities detected by graph clustering algorithms. Communities represent densely connected subnetworks that may correspond to functional protein modules or complexes.</em></p>\n")
    comm <- read.csv(comm_file, stringsAsFactors = FALSE)
    make_dt(comm, caption = "Network Communities")
}
```

---

# Advanced Statistics

```{r adv-stats-setup, include=FALSE}
has_adv_stats <- dir.exists(adv_stats_dir) && (
    length(list.files(adv_stats_dir, pattern = "\\.csv$")) > 0 ||
    length(list.files(adv_stats_plot_dir, pattern = "\\.png$")) > 0
)
adv_stats_enabled <- isTRUE(prot_cfg$advanced_stats$enabled)
```

```{r adv-stats-skip, eval=!has_adv_stats, results='asis'}
if (!adv_stats_enabled) {
    cat("Advanced statistics was disabled. Enable with `advanced_stats.enabled: true`.\n")
} else {
    cat("Advanced statistics was enabled but produced no results.\n")
}
```

```{r adv-stats-intro, eval=has_adv_stats, results='asis'}
cat("Advanced statistical analyses provide supplementary metrics including effect size confidence intervals, robust regression estimates, and statistical power calculations. These complement the primary differential expression results by quantifying uncertainty and detection sensitivity.\n\n")
```

## Effect Sizes

```{r adv-stats-effect-sizes, eval=has_adv_stats, results='asis'}
es_file <- file.path(adv_stats_dir, "effect_sizes.csv")
if (file.exists(es_file)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Effect size estimates with bootstrap confidence intervals for the top 100 most significant proteins. Confidence intervals quantify uncertainty in fold-change estimates.</em></p>\n")
    es <- read.csv(es_file, stringsAsFactors = FALSE)
    make_dt(head(es, 100), caption = "Effect Sizes (top 100 by significance)")
}
```

## Plots {.tabset}

```{r adv-stats-plots, eval=has_adv_stats, results='asis'}
if (dir.exists(adv_stats_plot_dir)) {
    adv_pngs <- list.files(adv_stats_plot_dir, pattern = "\\.png$", full.names = TRUE)
    if (length(adv_pngs) > 0) {
        for (ap in adv_pngs) {
            label_raw <- gsub("\\.png$", "", basename(ap))
            label <- gsub("_", " ", label_raw)
            cat(sprintf("\n### %s\n\n", label))
            cat(sprintf("![%s](%s)\n\n", label, ap))
            figure_legend(sprintf("%s from the advanced statistics module.", label))
            render_commentary(paste0("adv_stats_", label_raw))
        }
    } else {
        cat("No advanced statistics plots found.\n")
    }
}
```

## Power Analysis

```{r adv-stats-power, eval=has_adv_stats, results='asis'}
power_file <- file.path(adv_stats_dir, "power_analysis.csv")
if (file.exists(power_file)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Statistical power analysis estimating the probability of detecting true differential expression at various effect sizes given the current sample size. Power &ge; 0.8 is considered adequate.</em></p>\n")
    power <- read.csv(power_file, stringsAsFactors = FALSE)
    make_dt(power, caption = "Statistical Power Analysis")
}
```

## Multiple Testing Comparison

```{r adv-stats-mtc, eval=has_adv_stats, results='asis'}
mtc_file <- file.path(adv_stats_dir, "mtc_comparison.csv")
if (file.exists(mtc_file)) {
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>Table: Comparison of different multiple testing correction methods (BH, Bonferroni, BY, q-value) showing the number of significant proteins at each threshold. BH (Benjamini-Hochberg) is the default method.</em></p>\n")
    mtc <- read.csv(mtc_file, stringsAsFactors = FALSE)
    make_dt(mtc, caption = "Multiple Testing Correction Comparison")
}
```

---

# Output Files

```{r output-files}
files_tbl <- data.frame(
  File = c(
    "Final_results_ALL_P_0.05.xlsx",
    "Final_results_DE_P_0.05.xlsx",
    "Datasets/",
    "Diagnostic_plots/",
    "Diagnostic_plots/QC_post/"
  ),
  Description = c(
    "Excel workbook with all proteins and DE statistics (p < 0.05)",
    "Excel workbook with DE proteins only (p < 0.05)",
    "Intermediate datasets (expression matrices, DE tables)",
    "QC and diagnostic plots (PCA, density, heatmaps, UMAP)",
    "Post-DE QC plots (volcano, MA, top DE heatmaps)"
  ),
  stringsAsFactors = FALSE
)

# Add enrichment files if they exist
if (has_enrichment) {
    enrich_csvs <- list.files(enrich_dir, pattern = "\\.csv$")
    if (length(enrich_csvs) > 0) {
        enrich_tbl <- data.frame(
            File = paste0("Enrichment/", enrich_csvs),
            Description = paste("Pathway analysis results -",
                                gsub("^pathway_|\\.csv$", "", enrich_csvs)),
            stringsAsFactors = FALSE
        )
        files_tbl <- rbind(files_tbl, enrich_tbl)
    }
}

if (has_ppi) {
    files_tbl <- rbind(files_tbl, data.frame(
        File = "ppi_networks/", Description = "PPI network analysis results and plots",
        stringsAsFactors = FALSE
    ))
}

if (has_adv_stats) {
    files_tbl <- rbind(files_tbl, data.frame(
        File = "advanced_stats/", Description = "Advanced statistics results and plots",
        stringsAsFactors = FALSE
    ))
}

make_dt(files_tbl, "Key Output Files")
```

---

# Methods

```{r methods-text, results='asis'}
imp_method <- prot_cfg$imputation$method %||% "none"
de_method <- prot_cfg$de$method %||% "limma"
padj_cut <- prot_cfg$de$p_cutoff %||% 0.05
lfc_cut <- prot_cfg$de$linear_fc_cutoff %||% 1.5
engine <- prot_cfg$engine %||% "unknown"

n_proteins <- if (!is.null(de_data)) nrow(de_data) else "N"
n_de <- "N"
if (!is.null(de_data)) {
    pass_col <- grep("pass_any", names(de_data), value = TRUE)[1]
    if (!is.na(pass_col)) n_de <- sum(de_data[[pass_col]] == TRUE, na.rm = TRUE)
}

cat("**Proteomics Analysis Methods**\n\n")

cat(sprintf(
"Proteomics data were acquired using %s and processed using the multiomics-core pipeline. ", engine
))

if (tolower(imp_method) != "none") {
    cat(sprintf(
    "Missing values were imputed using the %s method. ", imp_method
    ))
}

cat(sprintf(
"Differential expression analysis was performed using %s. ", de_method
))

cat(sprintf(
"Proteins were considered differentially expressed at adjusted p-value <= %.2f and |fold change| >= %.1f. ",
padj_cut, lfc_cut
))

cat(sprintf(
"A total of %s proteins were quantified, with %s identified as differentially expressed in at least one contrast.",
format(as.numeric(n_proteins), big.mark = ","),
format(as.numeric(n_de), big.mark = ",")
))

if (pathway_enabled) {
    pathway_method <- prot_cfg$pathway$method %||% "both"
    pathway_dbs <- paste(prot_cfg$pathway$databases %||% c("GO", "KEGG"), collapse = " and ")
    cat(sprintf(" Pathway enrichment analysis was performed using %s against %s databases.",
                pathway_method, pathway_dbs))
}

if (ppi_enabled) {
    cat(" Protein-protein interaction networks were constructed using the STRING database.")
}

cat("\n\n")

cat("**Copy-ready version:**\n\n")
cat("<div style='background-color: #f5f5f5; padding: 15px; border-radius: 5px; font-size: 0.9em;'>\n")

cat(sprintf(
"Proteomics analysis was performed using the multiomics-core pipeline (v1.0). Data were acquired using %s. %sDifferential expression was assessed with %s (adjusted p <= %.2f, |FC| >= %.1f).",
engine,
if (tolower(imp_method) != "none") sprintf("Missing values were imputed using the %s method. ", imp_method) else "",
de_method, padj_cut, lfc_cut
))

cat("\n</div>\n")
```

## Statistical Analysis

```{r methods-statistical-analysis, results='asis'}
# Parameterized statistical analysis subsection for the Methods section.
# If report_cfg$stats_section_text is non-empty, print it verbatim;
# otherwise auto-generate from pipeline config values.
stats_section_text <- report_cfg$stats_section_text %||% ""
if (nzchar(trimws(stats_section_text))) {
    cat(trimws(stats_section_text), "\n\n")
} else {
    # Auto-generate from config
    de_method_m <- prot_cfg$de$method %||% "limma"
    imp_method_m <- prot_cfg$imputation$method %||% "none"
    padj_m <- prot_cfg$de$p_cutoff %||% 0.05
    lfc_m <- prot_cfg$de$linear_fc_cutoff %||% 1.5

    test_desc_m <- switch(tolower(de_method_m),
        limma = "Moderated t-tests were performed using the *limma* R/Bioconductor package, which applies empirical Bayes shrinkage of variance estimates to improve statistical power in small-sample proteomics studies.",
        "limma-voom" = "Differential expression was assessed using *limma-voom*, which transforms count data to log-CPM and applies precision-weighted linear modelling.",
        deseq2 = "Differential expression was assessed using Wald tests implemented in *DESeq2*, which fits negative binomial generalized linear models.",
        sprintf("Differential expression was assessed using the *%s* method.", de_method_m)
    )

    cat(test_desc_m, "\n\n")

    if (tolower(imp_method_m) != "none") {
        imp_desc <- switch(tolower(imp_method_m),
            perseus = "Missing values were imputed using the Perseus-style method (down-shifted normal distribution for MNAR, KNN for MCAR).",
            minprob = "Missing values were imputed using the MinProb method (random draws from a low-abundance distribution).",
            knn = "Missing values were imputed using K-nearest neighbours (KNN).",
            sprintf("Missing values were imputed using the %s method.", imp_method_m)
        )
        cat(imp_desc, "\n\n")
    }

    cat(sprintf(
        "Multiple testing correction was applied using the Benjamini-Hochberg procedure to control the false discovery rate (FDR). Proteins were considered differentially expressed at adjusted p-value $\\leq$ %s and |linear fold change| $\\geq$ %s.\n\n",
        padj_m, lfc_m
    ))

    cat(sprintf(
        "All analyses were performed in R (version %s) using the multiomics-core pipeline.\n\n",
        paste(R.version$major, R.version$minor, sep = ".")
    ))
}
```

---

# Session Info

```{r session-info, comment=""}
si_file <- file.path(run_dir, "execution_info", "sessionInfo.txt")
if (file.exists(si_file)) {
    si <- readLines(si_file)
    cat(paste(si, collapse = "\n"))
} else {
    cat(capture.output(sessionInfo()), sep = "\n")
}
```

```{r disclaimer-section, results='asis'}
disclaimer_text <- report_cfg$disclaimer_text %||% ""
if (nzchar(trimws(disclaimer_text))) {
    cat("---\n\n# Disclaimer\n\n")
    cat(trimws(disclaimer_text), "\n\n")
}
```

---

<p style="text-align:center; color:grey; font-size:0.9em;">
Report generated by multiomics-core pipeline | `r format(Sys.time(), '%Y-%m-%d %H:%M')`
</p>
