---
title: "Multi-Omics Integration Report"
subtitle: "multiomics-core Pipeline"
author: "Bioinformatician: `r yaml::read_yaml(file.path(normalizePath(dirname(knitr::current_input())), 'execution_info', 'config_used.yaml'))$project$analyst`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    fig_width: 10
    fig_height: 7
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%"
)

library(yaml)
library(knitr)
library(DT)
library(plotly)
library(ggplot2)
library(htmltools)
library(jsonlite)

# Helper function for downloadable DT tables
make_dt <- function(df, caption = NULL, pageLength = 5) {
    # Format real-number columns: < 0.001 → scientific, else 3 decimals
    for (col in colnames(df)) {
        if (is.numeric(df[[col]])) {
            vals <- df[[col]]
            formatted <- ifelse(
                is.na(vals), NA_character_,
                ifelse(vals == as.integer(vals) & !grepl("\\.", as.character(vals)),
                       as.character(vals),
                       ifelse(abs(vals) < 0.001 & vals != 0,
                              sprintf("%.3e", vals),
                              sprintf("%.3f", vals))))
            df[[col]] <- formatted
        }
    }
    DT::datatable(
        df,
        caption = caption,
        extensions = 'Buttons',
        options = list(
            scrollX = TRUE,
            pageLength = pageLength,
            lengthMenu = c(5, 10, 25, 50, 100),
            dom = 'Blfrtip',
            buttons = list(
                list(extend = 'csv', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data")),
                list(extend = 'excel', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data"))
            ),
            columnDefs = list(list(className = 'dt-left', targets = '_all'))
        ),
        rownames = FALSE
    )
}

# Helper: render figure legend (paper-style caption below figure)
figure_legend <- function(text) {
    cat(sprintf("<p class='figure-legend' style='text-align: center; font-size: 0.9em; color: #555; margin: 4px auto 12px auto; max-width: 88%%; line-height: 1.45; font-style: italic;'>%s</p>\n", text))
}

# Helper: render commentary block for a figure_id (collapsible)
render_commentary <- function(figure_id) {
    if (!has_commentary) return(invisible(NULL))
    c <- commentary[[figure_id]]
    if (is.null(c)) return(invisible(NULL))

    cat("<details class='commentary-details' style='margin: 4px 0 18px 0;'>\n")
    cat("<summary style='cursor: pointer; color: #4a90d9; font-weight: 500; font-size: 0.93em; padding: 3px 0;'>Show AI Interpretation</summary>\n")

    html <- "<div class='commentary-block' style='background: #f8f9fa; border-left: 4px solid #4a90d9; padding: 12px 16px; margin: 6px 0 4px 0; border-radius: 0 4px 4px 0; font-size: 0.92em;'>\n"
    html <- paste0(html, "<p style='margin: 0 0 8px 0;'><strong>What this shows:</strong> ",
                   htmltools::htmlEscape(c$what_is_this %||% ""), "</p>\n")

    bio_interp <- c$biological_interpretation %||% ""
    if (nzchar(bio_interp)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Interpretation:</strong> ",
                       htmltools::htmlEscape(bio_interp), "</p>\n")
    }

    obs <- c$observations
    if (length(obs) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Key observations:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (o in obs) html <- paste0(html, "<li>", htmltools::htmlEscape(o), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    stat_assess <- c$statistical_assessment %||% ""
    if (nzchar(stat_assess)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Statistical note:</strong> ",
                       htmltools::htmlEscape(stat_assess), "</p>\n")
    }

    issues <- c$issues_checks
    if (length(issues) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Checks:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (issue in issues) html <- paste0(html, "<li>", htmltools::htmlEscape(issue), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    steps <- c$next_steps
    if (length(steps) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Next steps:</strong></p>\n<ul style='margin: 2px 0 4px 0; padding-left: 20px;'>\n")
        for (s in steps) html <- paste0(html, "<li>", htmltools::htmlEscape(s), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    backend <- c$backend %||% "unknown"
    confidence <- c$confidence %||% ""
    backend_label <- if (grepl("claude", backend, ignore.case = TRUE)) {
        "Claude (Anthropic) + K-Dense AI scientific-skills"
    } else if (grepl("openai", backend, ignore.case = TRUE)) {
        "OpenAI + K-Dense AI scientific-skills"
    } else if (backend == "data_driven") {
        "data-driven + K-Dense AI scientific-skills"
    } else {
        backend
    }
    if (nzchar(confidence) && confidence != "none") {
        backend_label <- paste0(backend_label, " | confidence: ", confidence)
    }
    html <- paste0(html, "<p style='margin: 4px 0 0 0; color: #888; font-size: 0.85em;'><em>Generated by: ",
                   htmltools::htmlEscape(backend_label), "</em></p>\n</div>\n")

    cat(html)
    cat("</details>\n")
}

# Paths
run_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
config_path <- file.path(run_dir, "execution_info", "config_used.yaml")

# Multiomics output subdirectories
integration_dir  <- file.path(run_dir, "integration")
diablo_dir       <- file.path(integration_dir, "diablo")
snf_dir          <- file.path(integration_dir, "snf")
mofa_dir         <- file.path(integration_dir, "mofa")
enrichment_dir   <- file.path(run_dir, "cross_enrichment")
loadings_enrich_dir <- file.path(run_dir, "loadings_enrichment")
rna_prot_dir     <- file.path(run_dir, "rna_protein_correlation")
multigsea_dir    <- file.path(run_dir, "multigsea")
commentary_dir   <- file.path(run_dir, "commentary")

# Load commentary if available
commentary <- NULL
has_commentary <- FALSE
commentary_json <- file.path(commentary_dir, "all_commentary.json")
if (file.exists(commentary_json)) {
    commentary <- tryCatch(
        jsonlite::read_json(commentary_json),
        error = function(e) NULL
    )
    has_commentary <- !is.null(commentary) && length(commentary) > 0
}

# Load config
cfg <- yaml::read_yaml(config_path)
multi_cfg <- cfg$modes$multiomics

# Optional per-section switches, so a project can drop a section it cannot
# interpret without editing this shared template. Every section is ON unless a
# project sets it explicitly to FALSE -- an absent `report` block, an absent
# `sections` block and an unlisted name all mean "render it", so configs written
# before this existed behave exactly as they did.
section_cfg <- (multi_cfg$report %||% list())$sections %||% list()
show_section <- function(name) !identical(section_cfg[[name]], FALSE)

show_translation_eff    <- show_section("translation_efficiency")
show_group_distance     <- show_section("group_distance")
# Whole-section switches: FALSE drops the section, heading included, so a
# project that turned one off does not read an empty heading in its report.
show_harmonization      <- show_section("harmonization")
show_sample_concordance <- show_section("sample_concordance")
show_mechanistic        <- show_section("mechanistic")

# Detect which sections have data
has_diablo      <- dir.exists(diablo_dir) && length(list.files(diablo_dir, "\\.png$")) > 0
has_snf         <- dir.exists(snf_dir) && length(list.files(snf_dir, "\\.png$")) > 0
has_mofa        <- dir.exists(mofa_dir) && length(list.files(mofa_dir, "\\.png$")) > 0
has_integration <- has_diablo || has_snf || has_mofa
has_enrichment  <- dir.exists(enrichment_dir) && length(list.files(enrichment_dir, "\\.png$|\\.csv$")) > 0
has_transcriptomics <- "transcriptomics" %in% (cfg$global$omics_present %||% character(0)) ||
    !is.null(cfg$modes$rna)
has_rna_prot    <- has_transcriptomics &&
    dir.exists(file.path(rna_prot_dir, "plots")) &&
    length(list.files(file.path(rna_prot_dir, "plots"), "\\.png$")) > 0
has_multigsea   <- dir.exists(multigsea_dir) && length(list.files(multigsea_dir, "\\.png$")) > 0
# Tables count too: an entry whose rows cannot be drawn still writes its CSV.
has_loadings_enrich <- dir.exists(loadings_enrich_dir) &&
    length(list.files(loadings_enrich_dir, "\\.(png|csv)$", recursive = TRUE)) > 0

# How many contrasts the run compared, from the config and from the outputs
# the run actually wrote; see report_contrast_layout().
contrast_layout <- report_contrast_layout(cfg$design$contrasts, enrichment_dir,
                                          multigsea_dir)
# With one contrast every per-contrast view repeats the combined one, so each
# section shows a single tab named after that contrast instead of two copies.
single_contrast <- contrast_layout$single_contrast
single_contrast_label <- contrast_layout$label
# GO terms annotate genes and gene products; metabolites carry none, so a GO
# figure can never hold a metabolomics column. Said beside the figure, because an
# absent column otherwise reads as "tested and found nothing".
has_metabolomics_layer <- !is.null(cfg$modes$metabolomics) ||
    "metabolomics" %in% (cfg$global$omics_present %||% character(0))
collection_note <- function(coll) {
    if (has_metabolomics_layer && grepl("^GO", coll, ignore.case = TRUE)) {
        figure_legend(paste("GO annotates genes and proteins only; metabolites have",
                            "no GO terms, so metabolomics cannot appear in GO figures."))
    }
}

combined_tab <- function(level, multi_label) {
    label <- if (single_contrast) single_contrast_label else multi_label
    paste0(strrep("#", level), " ", label, " {.unnumbered}")
}

# Load gene-protein mapping if available
mapping_file <- file.path(run_dir, "gene_protein_mapping.csv")
gene_protein_mapping <- NULL
if (file.exists(mapping_file)) {
    gene_protein_mapping <- tryCatch(
        read.csv(mapping_file, stringsAsFactors = FALSE),
        error = function(e) NULL
    )
}
```

```{r force-dependencies, include=FALSE}
# Force DT and Plotly dependencies to be included in the HTML
DT::datatable(data.frame(x = 1))
plotly::plot_ly(data.frame(x = 1, y = 1), x = ~x, y = ~y, type = "scatter", mode = "markers")
```

# Analysis Parameters {.tabset}

## Project

```{r params-project}
proj <- data.frame(
  Parameter = c("Project Name", "Analysis Round", "Project Directory",
                 "Raw Data Path", "Output Path"),
  Value = c(cfg$project$name %||% "N/A",
            cfg$project$analysis_round %||% "N/A",
            cfg$project$dir %||% "N/A",
            cfg$paths$raw %||% "N/A",
            cfg$paths$out %||% "N/A"),
  stringsAsFactors = FALSE
)
kable(proj, col.names = c("Parameter", "Value"))
```

## Integration Methods

```{r params-integration}
methods <- multi_cfg$integration$methods %||% c("DIABLO", "SNF")
feat_sel <- multi_cfg$feature_selection$method %||% "variance"
# feature_selection$per_omics overrides top_n per layer; reporting top_n alone
# stated a number no layer actually used.
feat_n <- multi_cfg$feature_selection$top_n %||% 500
per_omics_n <- multi_cfg$feature_selection$per_omics
if (!is.null(per_omics_n) && length(per_omics_n) > 0) {
    feat_n <- paste0(
        feat_n, " (per omic: ",
        paste(sprintf("%s=%s", names(per_omics_n), unlist(per_omics_n)),
              collapse = ", "),
        ")"
    )
}
cond_col <- multi_cfg$condition_column %||% cfg$design$condition_column %||% "condition"

settings <- data.frame(
  Parameter = c(
    "Integration Methods",
    "Feature Selection Method",
    "Top N Features",
    "Condition Column",
    "Require 1:1 Gene-Protein Mapping"
  ),
  Value = c(
    paste(methods, collapse = ", "),
    feat_sel,
    feat_n,
    cond_col,
    ifelse(isTRUE(multi_cfg$require_one_to_one_mapping), "Yes", "No")
  ),
  stringsAsFactors = FALSE
)
kable(settings, col.names = c("Parameter", "Value"))
```

## Omics Layers

```{r params-omics}
omics_modes <- setdiff(names(cfg$modes), "multiomics")
omics_tbl <- data.frame(
    Omics = omics_modes,
    Configured = "Yes",
    stringsAsFactors = FALSE
)
kable(omics_tbl, col.names = c("Omics Layer", "Configured"))
```

---

`r if (has_rna_prot) "# RNA-Protein Correlation {.tabset}" else "# RNA-Protein Correlation"`

```{r rna-prot-note, eval=!has_rna_prot, results='asis'}
if (!has_transcriptomics) {
    cat("*Not run: this study has no RNA-seq layer, and the RNA-protein analyses",
        "(concordance, DE scatter, per-contrast DE details, top proteins by RNA",
        "agreement, translation efficiency) need one.*\n")
} else {
    cat("*No RNA-protein correlation results found. This analysis requires both transcriptomics and proteomics data.*\n")
}
```

`r if (has_rna_prot) "## Concordance Distribution"`

```{r rna-prot-concordance, eval=has_transcriptomics && has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "rna_protein_concordance.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r rna-prot-concordance-legend, eval=has_transcriptomics && has_rna_prot, results='asis'}
figure_legend("Distribution of per-gene RNA-protein Pearson correlations across samples. The histogram shows how well transcript levels predict protein abundance for individual genes.")
render_commentary("rna_protein_concordance")
```

`r if (has_rna_prot) "## DE Scatter (All Contrasts)"`

```{r rna-prot-de-scatter, eval=has_transcriptomics && has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "rna_protein_de_scatter.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
} else {
    cat("*Not produced for this run.*\n\n")
}
```

```{r rna-prot-de-legend, eval=has_transcriptomics && has_rna_prot && file.exists(file.path(rna_prot_dir, "plots", "rna_protein_de_scatter.png")), results='asis'}
figure_legend("RNA vs protein differential expression scatter plot per contrast. Each point is a gene/protein pair. Diagonal concordance indicates matching fold-change direction; off-diagonal points suggest post-transcriptional regulation.")
render_commentary("rna_protein_de_scatter")
```

```{r rna-prot-per-contrast-setup, include=FALSE}
# A contrast that joined nothing writes no files, and per_contrast/ exists only
# for successful DE joins. One predicate decides both which contrasts get a tab
# and whether the tabset's heading appears at all, so it never stands empty.
rna_prot_contrast_files <- function(cdir) {
    c(file.path(cdir, "plots", "rna_protein_de_scatter.png"),
      file.path(cdir, "plots", "translation_efficiency_hist.png"),
      file.path(cdir, "plots", "translation_efficiency_scatter.png"),
      file.path(cdir, "tables", "rna_protein_de_concordance.csv"))
}
rna_prot_pc_dir <- file.path(rna_prot_dir, "per_contrast")
rna_prot_contrast_dirs <- if (has_rna_prot && dir.exists(rna_prot_pc_dir)) {
    Filter(function(cdir) any(file.exists(rna_prot_contrast_files(cdir))),
           list.dirs(rna_prot_pc_dir, full.names = TRUE, recursive = FALSE))
} else character(0)
has_rna_prot_contrasts <- length(rna_prot_contrast_dirs) > 0
```

`r if (has_rna_prot_contrasts) "## Per-Contrast DE Details {.tabset}"`

```{r rna-prot-per-contrast, eval=has_rna_prot_contrasts, results='asis'}
for (cdir in rna_prot_contrast_dirs) {
    cname <- basename(cdir)
    de_csv <- file.path(cdir, "tables", "rna_protein_de_concordance.csv")

    cat(sprintf("\n### %s\n\n", gsub("_", " ", cname)))

    de_scatter <- file.path(cdir, "plots", "rna_protein_de_scatter.png")
    if (file.exists(de_scatter)) {
        cat(knitr::knit_child(text = sprintf(
            '```{r rna-prot-scatter-%s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            gsub("[^a-zA-Z0-9]", "", cname), de_scatter
        ), quiet = TRUE))
    }

    te_hist <- file.path(cdir, "plots", "translation_efficiency_hist.png")
    if (show_translation_eff && file.exists(te_hist)) {
        cat(knitr::knit_child(text = sprintf(
            '```{r rna-prot-te-%s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            gsub("[^a-zA-Z0-9]", "", cname), te_hist
        ), quiet = TRUE))
    }

    te_scatter <- file.path(cdir, "plots", "translation_efficiency_scatter.png")
    if (show_translation_eff && file.exists(te_scatter)) {
        cat(knitr::knit_child(text = sprintf(
            '```{r rna-prot-tes-%s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            gsub("[^a-zA-Z0-9]", "", cname), te_scatter
        ), quiet = TRUE))
    }

    # Summary line plus the per-pair table it is computed from.
    if (file.exists(de_csv)) {
        de_df <- read.csv(de_csv, stringsAsFactors = FALSE)
        n_total <- nrow(de_df)
        n_conc <- sum(de_df$concordant, na.rm = TRUE)
        r_val <- cor(de_df$rna_log2FC, de_df$protein_log2FC, use = "complete.obs")
        # Spearman alongside Pearson: the two layers' fold changes come from
        # different models, so a rank agreement and a linear one can differ
        # and a reader should see both rather than one chosen for them.
        rho_val <- suppressWarnings(
            cor(de_df$rna_log2FC, de_df$protein_log2FC,
                method = "spearman", use = "complete.obs"))
        tf_msg <- ""
        if ("is_transcription_factor" %in% colnames(de_df)) {
            n_tf <- sum(de_df$is_transcription_factor, na.rm = TRUE)
            if (n_tf > 0) tf_msg <- sprintf(", %d known TFs", n_tf)
        }
        cat(sprintf(paste0("\n**%s:** %d gene-protein pairs, Pearson r = %.3f, ",
                           "Spearman rho = %.3f, %d/%d (%.0f%%) concordant%s\n\n"),
                    gsub("_", " ", cname), n_total, r_val, rho_val,
                    n_conc, n_total, 100 * n_conc / n_total, tf_msg))

        # The table the line above is computed from. Rendered through
        # knit_child because print() inside a results='asis' chunk emits
        # nothing for an htmlwidget -- the idiom the rest of this file uses.
        # Columns lead with both identifiers: the layers use different ID
        # spaces, and a reader looking a gene up in the proteomics tables
        # needs the protein id to join on.
        show_cols <- intersect(
            c("gene_id", "protein_id", "rna_log2FC", "rna_padj",
              "protein_log2FC", "protein_padj", "concordant", "category"),
            names(de_df))
        de_csv_abs <- normalizePath(de_csv, winslash = "/", mustWork = FALSE)
        cat(knitr::knit_child(text = sprintf(
            '```{r rna-prot-tbl-%s, echo=FALSE}\nd <- read.csv("%s", stringsAsFactors = FALSE)\nd <- d[, intersect(c(%s), names(d)), drop = FALSE]\nmake_dt(d, caption = "RNA vs protein log2FC per gene-protein pair - %s", pageLength = 15)\n```\n',
            gsub("[^a-zA-Z0-9]", "", cname), de_csv_abs,
            paste(sprintf('"%s"', show_cols), collapse = ", "),
            gsub("_", " ", cname)
        ), quiet = TRUE))
    }
}
```

`r if (has_rna_prot) "## Top Proteins by RNA Agreement"`

```{r rna-prot-topn-bars, eval=has_transcriptomics && has_rna_prot, fig.width=10, fig.height=7}
# The protein side is the limma moderated log2FC, the RNA side the DESeq2
# log2FC. They are NOT the same statistic, so this is read as agreement of
# direction, not of magnitude.
TOPN <- 50
fc_thr <- log2(1.5)

pairs_csv <- file.path(rna_prot_dir, "tables", "rna_protein_log2FC_pairs.csv")
tb_ok <- file.exists(pairs_csv)
if (tb_ok) {
    pr <- read.csv(pairs_csv, stringsAsFactors = FALSE)
    # Only the file and its columns are checked here. Whether there are enough
    # pairs to draw is decided per contrast below, since that is the unit each
    # panel is built from: a pooled total says nothing about whether any single
    # contrast can fill a panel, in either direction.
    tb_ok <- all(c("contrast", "rna_log2FC", "protein_log2FC") %in% names(pr))
}
if (tb_ok) {
    pr <- pr[is.finite(pr$rna_log2FC) & is.finite(pr$protein_log2FC), , drop = FALSE]
    tb_ok <- nrow(pr) > 0
}

if (!tb_ok) {
    cat("*Not enough mapped gene-protein pairs for the top-protein bar plot.*\n")
} else {
    # Hypergeometric: of the n most extreme proteins, how many have an RNA fold
    # change past the threshold in the SAME direction, against the rate over
    # every mapped pair. lower.tail = FALSE with q - 1 gives P(X >= k).
    enrich_p <- function(k, K, n, N) {
        if (K == 0 || n == 0) return(NA_real_)
        stats::phyper(k - 1, K, N - K, n, lower.tail = FALSE)
    }

    panels <- list()
    for (cn in unique(pr$contrast)) {
        d <- pr[pr$contrast == cn, , drop = FALSE]
        N <- nrow(d)

        for (dir in c("up", "down")) {
            # Restricted to proteins that actually moved that way BEFORE the
            # tail is taken. Sorting alone fills a short tail with proteins
            # moving the other way -- in an all-positive contrast the "down"
            # panel would hold positive protein fold changes, and `agree` would
            # then count a falling transcript as agreeing with a rising
            # protein and feed that into the p-value.
            d_dir <- if (dir == "up") {
                d[d$protein_log2FC > 0, , drop = FALSE]
            } else {
                d[d$protein_log2FC < 0, , drop = FALSE]
            }
            n <- min(TOPN, nrow(d_dir))
            if (n < 10) next

            sel <- if (dir == "up") {
                d_dir[order(-d_dir$protein_log2FC), ][seq_len(n), ]
            } else {
                d_dir[order(d_dir$protein_log2FC), ][seq_len(n), ]
            }
            agree <- if (dir == "up") sel$rna_log2FC >= fc_thr else sel$rna_log2FC <= -fc_thr
            # The background stays every mapped pair in the contrast: the
            # question is whether the strongest movers agree more often than
            # the layer does at large.
            K <- if (dir == "up") sum(d$rna_log2FC >= fc_thr) else sum(d$rna_log2FC <= -fc_thr)
            k <- sum(agree)
            p <- enrich_p(k, K, n, N)

            sel$rank <- seq_len(n)
            sel$agree <- agree
            # n, not a literal: a contrast with few pairs takes fewer than TOPN,
            # and a label naming a count the panel does not show is worse than
            # no label at all.
            sel$direction <- sprintf("top %d %s in protein", n,
                                     if (dir == "up") "up" else "down")
            sel$contrast_lab <- gsub("_", " ", cn)
            sel$note <- sprintf("%d/%d agree (expected %.1f), hypergeometric p = %s",
                                k, n, n * K / N,
                                if (is.na(p)) "NA" else format.pval(p, digits = 2))
            panels[[paste(cn, dir)]] <- sel
        }
    }

    if (length(panels) == 0) {
        cat("*Not enough pairs per contrast for the top-protein bar plot.*\n")
    } else {
        tb <- do.call(rbind, panels)
        long <- rbind(
            transform(tb[, c("rank", "direction", "contrast_lab", "agree", "note")],
                      value = tb$protein_log2FC, layer = "protein log2FC"),
            transform(tb[, c("rank", "direction", "contrast_lab", "agree", "note")],
                      value = tb$rna_log2FC, layer = "RNA log2FC")
        )
        long$layer <- factor(long$layer, levels = c("protein log2FC", "RNA log2FC"))

        lab <- unique(tb[, c("direction", "contrast_lab", "note")])
        lab$layer <- factor("protein log2FC", levels = levels(long$layer))

        print(
            ggplot2::ggplot(long, ggplot2::aes(rank, value, fill = agree)) +
                ggplot2::geom_col(width = 1) +
                ggplot2::geom_hline(yintercept = 0, linewidth = 0.2, colour = "grey40") +
                ggplot2::geom_hline(
                    data = data.frame(y = c(fc_thr, -fc_thr)),
                    ggplot2::aes(yintercept = y), inherit.aes = FALSE,
                    linetype = "dashed", linewidth = 0.25, colour = "grey55") +
                ggplot2::geom_text(data = lab,
                    ggplot2::aes(x = Inf, y = Inf, label = note), inherit.aes = FALSE,
                    hjust = 1.02, vjust = 1.4, size = 2.6) +
                ggplot2::facet_grid(layer ~ direction + contrast_lab, scales = "free_y") +
                ggplot2::scale_fill_manual(
                    values = c("TRUE" = "#2166ac", "FALSE" = "grey78"),
                    labels = c("TRUE" = "RNA agrees (past 1.5-fold, same direction)",
                               "FALSE" = "RNA does not"),
                    name = NULL) +
                ggplot2::theme_bw(base_size = 9) +
                ggplot2::theme(legend.position = "top") +
                ggplot2::labs(
                    x = "protein rank (1 = most extreme)", y = "log2 fold change",
                    title = "Top proteins by log2FC, with the RNA fold change of the same features")
        )
    }
}
```

```{r rna-prot-topn-bars-legend, eval=has_transcriptomics && has_rna_prot, results='asis'}
figure_legend(paste(
    "Each column is one gene-protein pair, ranked by protein log2FC, most extreme",
    "first; the upper row is the protein fold change, the lower row the RNA fold",
    "change of the same pair. Blue marks pairs whose RNA moved past 1.5-fold in the",
    "same direction (dashed lines). The panel label gives the observed count, the",
    "count expected if the top proteins were a random draw of mapped pairs, and a",
    "hypergeometric p. Protein is the limma log2FC and RNA the DESeq2",
    "log2FC, so this shows agreement of direction, not size. Unmapped proteins are",
    "absent. Source: rna_protein_log2FC_pairs.csv. Hypothesis-generating."))
```

`r if (has_rna_prot && show_translation_eff) "## Translation Efficiency (All Contrasts)"`

```{r rna-prot-te-hist, eval=has_transcriptomics && has_rna_prot && show_translation_eff, results='asis'}
p <- file.path(rna_prot_dir, "plots", "translation_efficiency_hist.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
} else {
    cat("*Not produced for this run.*\n\n")
}
```

```{r rna-prot-te-hist-legend, eval=has_transcriptomics && has_rna_prot && show_translation_eff && file.exists(file.path(rna_prot_dir, "plots", "translation_efficiency_hist.png")), results='asis'}
figure_legend("Distribution of translation efficiency changes per contrast (protein log2FC - RNA log2FC). Values near 0 indicate transcriptional control; large positive/negative values suggest post-transcriptional regulation.")
render_commentary("translation_efficiency_hist")
```

```{r rna-prot-te-scatter, eval=has_transcriptomics && has_rna_prot && show_translation_eff, results='asis'}
p <- file.path(rna_prot_dir, "plots", "translation_efficiency_scatter.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
} else {
    cat("*Not produced for this run.*\n\n")
}
```

```{r rna-prot-te-scatter-legend, eval=has_transcriptomics && has_rna_prot && show_translation_eff && file.exists(file.path(rna_prot_dir, "plots", "translation_efficiency_scatter.png")), results='asis'}
figure_legend("Translation efficiency scatter plot per contrast comparing RNA and protein fold-changes.")
render_commentary("translation_efficiency_scatter")
```

---

# MultiGSEA, Multi-ORA and Compound Enrichment {.tabset}

```{r multigsea-pair-helpers, include=FALSE}
# The pairwise MultiGSEA figures appear per contrast, and for a single-contrast
# run under that contrast's name; both places draw and caption them alike.
show_multigsea_pair <- function(png, chunk_prefix) {
    pair_name <- gsub("multigsea_|\\.png$", "", basename(png))
    chunk_id <- paste0(chunk_prefix, "-", gsub("[^a-zA-Z0-9]", "", pair_name))
    cat(knitr::knit_child(text = sprintf(
        '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        chunk_id, png
    ), quiet = TRUE))
    cat("\n\n")
}
multigsea_pair_legend <- function() {
    figure_legend(paste(
        "Two layers compared pathway by pathway. Each point is a pathway either",
        "layer scored; each axis is -log10 of that layer's adjusted p-value (raw p",
        "where a result has none), and a pathway one layer did not score sits at zero",
        "on its axis. The dashed line marks equal significance. A size or colour key,",
        "where shown, is the two layers' average of what it names, an unscored layer",
        "counting as zero. The subtitle correlation runs over the union of pathways,",
        "zeros included; pathways share genes, so its p-value is nominal, and it is",
        "not a measure of biological agreement."))
}
```

`r if (has_multigsea && single_contrast) combined_tab(2, "")`

```{r multigsea-single-contrast, eval=has_multigsea && single_contrast, results='asis'}
# With one contrast the pipeline writes no per_contrast/ directory: the pairwise
# figures at the top of multigsea/ are that contrast's own view.
pair_pngs <- sort(list.files(multigsea_dir, pattern = "^multigsea_.*_vs_.*\\.png$",
                             full.names = TRUE))
if (length(pair_pngs) == 0) {
    cat("*No pairwise MultiGSEA plots for this contrast.*\n\n")
} else {
    for (pp in pair_pngs) show_multigsea_pair(pp, "mg-sc")
    multigsea_pair_legend()
}
```

`r if (!single_contrast) "## Per-Contrast MultiGSEA {.tabset .unnumbered}"`

```{r multigsea-per-contrast, eval=has_multigsea && !single_contrast, results='asis'}
multigsea_per_contrast_dir <- file.path(multigsea_dir, "per_contrast")
if (dir.exists(multigsea_per_contrast_dir)) {
    contrast_dirs <- list.dirs(multigsea_per_contrast_dir, full.names = TRUE, recursive = FALSE)
    for (cdir in contrast_dirs) {
        cname <- basename(cdir)
        cat(sprintf("\n### %s {.unnumbered}\n\n", gsub("_", " ", cname)))

        # Per-contrast dirs hold only the pairwise multigsea_<a>_vs_<b>.png
        # plots — the combined figure is written at the top level only, so
        # filtering "_vs_" out here (as the top-level block does) left nothing.
        cpngs <- list.files(cdir, pattern = "^multigsea_.*\\.png$", full.names = TRUE)
        if (length(cpngs) == 0) {
            cat("*No MultiGSEA plots for this contrast.*\n\n")
            next
        }
        for (cp in cpngs) {
            show_multigsea_pair(cp, paste0("mg-pc-", gsub("[^a-zA-Z0-9]", "", cname)))
        }
        multigsea_pair_legend()
    }
} else {
    # Single-contrast runs never reach this chunk (see single_contrast), so a
    # missing directory here means the per-contrast step produced nothing.
    cat("*Per-contrast MultiGSEA results not available.*\n")
}
```

```{r multi-ora-setup, include=FALSE}
multi_ora_dir <- file.path(multigsea_dir, "multi_ora")
has_multi_ora <- dir.exists(multi_ora_dir) && length(list.files(multi_ora_dir, "\\.png$")) > 0
```

## Multi-ORA (Combined Cross-Omics ORA) {.tabset .unnumbered}

```{r multi-ora-note, eval=!has_multi_ora, results='asis'}
ora_issues_file <- file.path(multi_ora_dir, "multi_ora_issues.txt")
if (file.exists(ora_issues_file)) {
    issues <- readLines(ora_issues_file)
    cat("**Multi-ORA did not complete.** The following issues were encountered:\n\n")
    for (iss in issues) {
        cat(sprintf("- %s\n", iss))
    }
    cat("\n*This is often caused by KEGG REST API timeouts or network issues. Re-running the pipeline may resolve it.*\n\n")
} else {
    cat("*No Multi-ORA results found.*\n")
}
```

`r combined_tab(3, "All Contrasts (Pooled)")`

```{r multi-ora-plots, eval=has_multi_ora, results='asis'}
# Pooled barplot
pooled_bp <- file.path(multi_ora_dir, "multi_ora_pooled_barplot.png")
if (file.exists(pooled_bp)) {
    cat("\n#### Pooled Multi-ORA {.unnumbered}\n\n")
    cat(sprintf("![Pooled Multi-ORA](%s)\n\n", pooled_bp))
    figure_legend(paste(
        "Pooled multi-ORA: significant genes from all gene-based layers, pooled into",
        "one set and tested with Fisher's exact test against KEGG pathways, or the",
        "supplied GMT gene sets where KEGG is unavailable. Bar length and the red",
        "dashed line show the adjusted p-value where usable, otherwise the raw",
        "p-value (see the axis label). With several collections, terms are taken in",
        "turn from each, and bars are tagged and coloured by collection, strongest",
        "first."))
}

# Dot plot
dotplot <- file.path(multi_ora_dir, "multi_ora_dotplot.png")
if (file.exists(dotplot)) {
    cat("\n#### Pooled vs Per-Omics Comparison {.unnumbered}\n\n")
    cat(sprintf("![Multi-ORA Dot Plot](%s)\n\n", dotplot))
    figure_legend("Multi-ORA dot plot comparing per-omics and pooled enrichment results. Each dot represents a pathway; x-axis is -log10(padj), dot size reflects gene count, and color indicates the source (pooled or individual omics). Red dashed line marks padj = 0.05.")
}

# Omics support barplot
support_bp <- file.path(multi_ora_dir, "multi_ora_support_barplot.png")
if (file.exists(support_bp)) {
    cat("\n#### Omics Support per Pathway {.unnumbered}\n\n")
    cat(sprintf("![Multi-ORA Support](%s)\n\n", support_bp))
    figure_legend("Multi-ORA omics support barplot. Bar length shows pooled enrichment significance (-log10 p-value). Color indicates the number of individual omics layers that independently support each pathway (padj < 0.05).")
}

# Summary table
multi_ora_csv <- file.path(multi_ora_dir, "multi_ora_results.csv")
if (file.exists(multi_ora_csv)) {
    cat("\n#### Top Multi-ORA Pathways {.unnumbered}\n\n")
    mora_df <- read.csv(multi_ora_csv, stringsAsFactors = FALSE)
    mora_show <- mora_df[seq_len(min(25, nrow(mora_df))), ]
    display_cols <- intersect(c("pathway", "pooled_padj", "pooled_count", "metabolomics_padj", "n_omics_support"), colnames(mora_show))
    mora_show <- mora_show[, display_cols, drop = FALSE]
    col_labels <- c(pathway = "Pathway", pooled_padj = "Pooled padj", pooled_count = "Gene Count",
                    metabolomics_padj = "Metab padj", n_omics_support = "# Omics Supporting")
    colnames(mora_show) <- col_labels[display_cols]
    # Format numeric columns (signif handles NA gracefully)
    for (nc in intersect(c("Pooled padj", "Metab padj"), colnames(mora_show))) {
        mora_show[[nc]] <- signif(mora_show[[nc]], 3)
    }
    knitr::kable(mora_show, row.names = FALSE, caption = "Top multi-ORA enriched KEGG pathways (gene-based + metabolite compound ORA)")
}

render_commentary("multi_ora")
```

`r if (!single_contrast) "### Per-Contrast Multi-ORA {.tabset .unnumbered}"`

```{r multi-ora-per-contrast, eval=has_multi_ora && !single_contrast, results='asis'}
mora_per_contrast_dir <- file.path(multi_ora_dir, "per_contrast")
if (dir.exists(mora_per_contrast_dir)) {
    mora_contrast_dirs <- list.dirs(mora_per_contrast_dir, full.names = TRUE, recursive = FALSE)
    for (cdir in mora_contrast_dirs) {
        cname <- basename(cdir)
        cat(sprintf("\n#### %s {.unnumbered}\n\n", gsub("_", " ", cname)))

        # Barplot
        c_bp <- file.path(cdir, "multi_ora_pooled_barplot.png")
        if (file.exists(c_bp)) {
            chunk_id <- paste0("mora-pc-bp-", gsub("[^a-zA-Z0-9]", "", cname))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, c_bp
            ), quiet = TRUE))
            cat("\n\n")
        }

        # Results table
        c_csv <- file.path(cdir, "multi_ora_results.csv")
        if (file.exists(c_csv)) {
            c_df <- read.csv(c_csv, stringsAsFactors = FALSE)
            c_show <- c_df[seq_len(min(20, nrow(c_df))), ]
            display_cols <- intersect(c("pathway", "pooled_padj", "pooled_count", "metabolomics_padj", "n_omics_support"), colnames(c_show))
            if (length(display_cols) > 0) {
                c_show <- c_show[, display_cols, drop = FALSE]
                cat(sprintf("\n**%s:** %d enriched pathways\n\n",
                            gsub("_", " ", cname), nrow(c_df)))
            }
        } else {
            cat("*No enriched pathways for this contrast.*\n\n")
        }
    }
} else {
    cat("*Per-contrast Multi-ORA results not available.*\n")
}
```

```{r metab-ora-setup, include=FALSE}
# Compound ORA comes out of run_compound_ora() via two callers, and neither
# depends on the gene-based multi-ORA having succeeded — but they handle
# contrasts differently, so the two files do not read the same way:
#
#   compound_ora_results.csv          pools every contrast into a single test,
#                                     one row per pathway
#   metabolomics_kegg_enrichment.csv  the only one produced for an organism with
#                                     no OrgDb/KEGG code; tests each contrast
#                                     separately and row-binds the results under
#                                     a `contrast` column, so one pathway can
#                                     appear once per contrast
#
# That column decides how the section below reports: reading a per-contrast file
# as if it were pooled counts the same pathway several times.
metab_ora_csv <- file.path(multi_ora_dir, "compound_ora_results.csv")
if (!file.exists(metab_ora_csv)) {
    metab_ora_csv <- file.path(enrichment_dir, "metabolomics",
                               "metabolomics_kegg_enrichment.csv")
}
has_metab_ora <- file.exists(metab_ora_csv)
metab_df <- if (has_metab_ora) read.csv(metab_ora_csv, stringsAsFactors = FALSE) else NULL
metab_ora_by_contrast <- !is.null(metab_df) && "contrast" %in% colnames(metab_df)
# Each ORA call sorts its own results, so a row-bound file arrives in contrast
# order rather than p-value order. Restore a global ordering before anything
# below takes a top-N slice of it.
if (metab_ora_by_contrast && "pvalue" %in% colnames(metab_df)) {
    metab_df <- metab_df[order(metab_df$pvalue), , drop = FALSE]
}
```

## Metabolomics Compound Enrichment {.unnumbered}

```{r metab-ora-note, eval=!has_metab_ora, results='asis'}
cat("*No metabolomics compound ORA results found. Compound enrichment requires metabolomics DE results with KEGG compound ID mappings.*\n")
```

```{r metab-ora-results, eval=has_metab_ora, results='asis'}
if (nrow(metab_df) > 0) {
    if (metab_ora_by_contrast) {
        cat(sprintf(paste0("\nCompound-based over-representation analysis identified **%d** ",
                           "enriched KEGG pathways from significantly altered metabolites, ",
                           "tested separately within each of **%d** contrasts.\n\n"),
                    length(unique(metab_df$pathway)),
                    length(unique(metab_df$contrast))))
    } else {
        cat(sprintf(paste0("\nCompound-based over-representation analysis identified **%d** ",
                           "enriched KEGG pathways from significantly altered metabolites.\n\n"),
                    nrow(metab_df)))
    }
}
```

```{r metab-ora-barplot, eval=has_metab_ora, results='asis'}
metab_bp <- file.path(multi_ora_dir, "compound_ora_barplot.png")
if (file.exists(metab_bp)) {
    cat(sprintf("![Compound ORA](%s)\n\n", metab_bp))
    figure_legend("Metabolomics compound ORA barplot. Significant metabolites (KEGG compound IDs) are tested for over-representation in KEGG metabolic pathways using a hypergeometric test. Bar length and the red dashed threshold line both show the adjusted p-value where the result carries usable ones and the raw p-value otherwise; the axis label states which.")
}
```

```{r metab-ora-table, eval=has_metab_ora}
# Compound ORA writes the readable name in `pathway` and the accession in `ID`;
# a table that has been through add_pathway_names() carries `pathway_name` too,
# but it can be filled for some rows and not others. pathway_display_label()
# resolves that per row, so a row without a name falls back to its own `pathway`
# instead of rendering blank -- which is what choosing one column for the whole
# table would do. The label is built into the presentation copy only; the raw
# compound-ORA result is not modified.
metab_show <- metab_df[seq_len(min(30, nrow(metab_df))), , drop = FALSE]
metab_show$pathway <- pathway_display_label(metab_show)

display_cols <- intersect(c("contrast", "pathway", "ID", "pvalue", "padj", "setSize", "compounds"),
                          colnames(metab_show))
metab_show <- metab_show[, display_cols, drop = FALSE]
col_labels <- c(contrast = "Contrast", pathway = "Pathway", ID = "KEGG ID", pvalue = "P-value",
                padj = "Adj. P-value", setSize = "Compound Count", compounds = "Compounds")
colnames(metab_show) <- col_labels[display_cols]
knitr::kable(metab_show, row.names = FALSE,
             caption = if (metab_ora_by_contrast)
                 "Metabolomics compound ORA - enriched KEGG pathways, one row per pathway and contrast"
             else "Metabolomics compound ORA - enriched KEGG pathways")
```

```{r metab-ora-commentary, eval=has_metab_ora, results='asis'}
render_commentary("compound_ora")
```

```{r metab-gsea-setup, include=FALSE}
# Compound GSEA is written by its own orchestrator, not by the compound ORA
# path, and it is kept out of the per-layer cross-omics tables and the ORA
# figure. Its p-values do reach the meta-analysis, as the metabolomics layer's
# rank-based evidence (rank_tables in mod_multiomics_enrichment()). One row per
# pathway per contrast, and every
# pathway fgsea scored -- the producer applies no significance cutoff, so the
# subset shown below is a reading convenience and the CSV is the whole result.
metab_gsea_csv <- file.path(enrichment_dir, "metabolomics_compound_gsea.csv")
has_metab_gsea <- file.exists(metab_gsea_csv)
metab_gsea_df <- if (has_metab_gsea) {
    read.csv(metab_gsea_csv, stringsAsFactors = FALSE)
} else {
    NULL
}
has_metab_gsea <- !is.null(metab_gsea_df) && nrow(metab_gsea_df) > 0
```

## Metabolomics Compound GSEA {.unnumbered}

```{r metab-gsea-note, eval=!has_metab_gsea, results='asis'}
cat("*No metabolomics compound GSEA results found. Compound GSEA needs metabolomics DE results, KEGG compound ID mappings, and the compound-pathway table the compound ORA step caches.*\n")
```

```{r metab-gsea-table, eval=has_metab_gsea}
# Ordered on the adjusted p-value for display only; the exported file keeps
# every scored pathway. pathway_display_label() resolves the readable name per
# row, so a row whose name did not resolve falls back to its own accession
# rather than rendering blank.
gsea_show <- metab_gsea_df[order(metab_gsea_df$padj, metab_gsea_df$ID), , drop = FALSE]
gsea_show <- gsea_show[seq_len(min(30, nrow(gsea_show))), , drop = FALSE]
gsea_show$pathway <- pathway_display_label(gsea_show)

gsea_cols <- intersect(c("contrast", "pathway", "ID", "NES", "pvalue", "padj", "setSize"),
                       colnames(gsea_show))
gsea_show <- gsea_show[, gsea_cols, drop = FALSE]
gsea_labels <- c(contrast = "Contrast", pathway = "Pathway", ID = "KEGG ID",
                 NES = "NES", pvalue = "P-value", padj = "Adj. P-value",
                 setSize = "Compound Count")
colnames(gsea_show) <- gsea_labels[gsea_cols]
knitr::kable(gsea_show, row.names = FALSE,
             caption = "Metabolomics compound GSEA - top pathways by adjusted p-value")
```

```{r metab-gsea-legend, eval=has_metab_gsea, results='asis'}
figure_legend(paste(
    "Compound GSEA over KEGG metabolic pathways, beside the compound ORA above.",
    "ORA asks whether a pathway holds many significant metabolites; GSEA asks",
    "whether its metabolites sit high or low in the ranked list, and NES gives",
    "that direction (positive: concentrated at the top for the contrast).",
    "Ranking: the moderated statistic where available, else sign(log2FC) x",
    "-log10(p). Adjusted p is BH over the pathways scored here, so it is not",
    "comparable with the ORA table's. The strongest 30 rows are shown; all are in",
    "metabolomics_compound_gsea.csv."))
render_commentary("compound_gsea")
```


---

# Cross-Omics Enrichment {.tabset}

```{r enrichment-note, eval=!has_enrichment, results='asis'}
cat("*No cross-omics enrichment results found. Enrichment analysis requires pathway results from at least 1 omics layer.*\n")
```

`r combined_tab(2, "All Contrasts (Combined)")`

```{r enrichment-per-omics, eval=has_enrichment, results='asis'}
# Show per-omics enrichment barplots
omics_barplots <- list.files(enrichment_dir, pattern = "_top_pathways\\.png$", full.names = TRUE)
if (length(omics_barplots) > 0) {
    cat("\n### Per-Omics Pathway Enrichment {.unnumbered}\n\n")
    for (i in seq_along(omics_barplots)) {
        bp <- omics_barplots[i]
        omics_name <- gsub("_top_pathways\\.png$", "", basename(bp))
        cat(sprintf("\n#### %s {.unnumbered}\n\n", tools::toTitleCase(omics_name)))
        chunk_id <- paste0("enrich-bp-", i)
        cat(knitr::knit_child(text = sprintf(
            '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            chunk_id, bp
        ), quiet = TRUE))
        cat("\n\n")
    }
}
```

```{r enrichment-per-omics-legend, eval=has_enrichment, results='asis'}
figure_legend("Per-omics pathway enrichment across every gene-set collection configured for that layer (all contrasts combined), not KEGG alone. Bar length represents -log10(p-value). Red dashed line indicates p = 0.05 significance threshold.")
```

```{r enrichment-per-omics-table, eval=has_enrichment}
# Show enrichment tables for each omics
csv_files <- list.files(enrichment_dir, pattern = "_enriched_pathways\\.csv$", full.names = TRUE)
for (cf in csv_files) {
    omics_name <- gsub("_enriched_pathways\\.csv$", "", basename(cf))
    tbl <- read.csv(cf, stringsAsFactors = FALSE)
    if (nrow(tbl) > 0) {
        print(make_dt(
            head(tbl, 200),
            caption = paste0(tools::toTitleCase(omics_name), " Enriched KEGG Pathways"),
            pageLength = 15
        ))
    }
}
```

```{r enrichment-heatmap, eval=has_enrichment, results='asis'}
# One figure per gene-set collection. The collections cover neither the same
# layers nor the same identifier space, and differ by an order of magnitude in
# size, so a single combined figure gave its rows to whichever was largest.
# Discovered by glob rather than by a fixed name: which collections a run
# produces depends on the gene sets it was given.
nes_pngs <- sort(list.files(
    enrichment_dir, pattern = "^cross_omics_pathway_heatmap_.*\\.png$",
    full.names = TRUE))
if (length(nes_pngs) > 0) {
    cat("\n### Cross-Omics Pathway Comparison {.unnumbered}\n\n")
    for (p in nes_pngs) {
        coll <- sub("^cross_omics_pathway_heatmap_", "",
                    sub("\\.png$", "", basename(p)))
        cat(sprintf("\n#### %s gene sets {.unnumbered}\n\n", gsub("_", " ", coll)))
        cat(knitr::knit_child(text = sprintf(
            '```{r enrich-heatmap-img-%s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            gsub("[^a-zA-Z0-9]", "", coll), p
        ), quiet = TRUE))
        collection_note(coll)
    }
}
```

```{r enrichment-heatmap-legend, eval=has_enrichment, results='asis'}
if (length(list.files(enrichment_dir, pattern = "^cross_omics_pathway_heatmap_.*\\.png$")) > 0) {
    # Within 100 words: how to read the colour, the grey and the row order, and
    # the limits a reader must not read past. The reasoning behind them lives in
    # the code comments of analyze_cross_omics_enrichment() and
    # select_multi_omics_pathways().
    figure_legend(paste(
        "One heatmap per gene-set collection. Colour: -log10 p of each layer's own",
        "test (GSEA, else ORA; the table names which). Grey: no p-value there, not p",
        "near 1. A missing layer was not necessarily tested and found negative. Rows:",
        "top 30 pathways by layers at p < 0.05, then Stouffer p; that count orders the",
        "display and is not a significance threshold. A contrast or collection a layer",
        "scored only with its other test is left out for that layer; the run log names",
        "it.",
        if (!single_contrast) "Each cell is the layer's smallest p across contrasts."))
    render_commentary("cross_omics_pathway_heatmap")
}
```

```{r enrichment-ora-heatmap, eval=has_enrichment, results='asis'}
ora_pngs <- sort(list.files(
    enrichment_dir, pattern = "^cross_omics_ora_heatmap_.*\\.png$",
    full.names = TRUE))
if (length(ora_pngs) > 0) {
    cat("\n### Cross-Omics ORA Evidence {.unnumbered}\n\n")
    for (p in ora_pngs) {
        coll <- sub("^cross_omics_ora_heatmap_", "",
                    sub("\\.png$", "", basename(p)))
        cat(sprintf("\n#### %s gene sets {.unnumbered}\n\n", gsub("_", " ", coll)))
        cat(knitr::knit_child(text = sprintf(
            '```{r enrich-ora-heatmap-img-%s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            gsub("[^a-zA-Z0-9]", "", coll), p
        ), quiet = TRUE))
        collection_note(coll)
    }
}
```

```{r enrichment-ora-heatmap-legend, eval=has_enrichment, results='asis'}
if (length(list.files(enrichment_dir, pattern = "^cross_omics_ora_heatmap_.*\\.png$")) > 0) {
    figure_legend(paste(
        "ORA evidence, one heatmap per gene-set collection. Colour: -log10 of each",
        "layer's own adjusted ORA p-value, capped at 1e-10; nothing is re-adjusted",
        "across contrasts, and GSEA results do not enter. Grey: no ORA result in that",
        "layer, not p near 1; a missing layer was not necessarily tested. Rows: top 30",
        "pathways by number of layers with a result, then smallest adjusted p.",
        "Compare colours within a column.",
        if (!single_contrast) paste("Each cell is the layer's smallest adjusted p across",
                                    "contrasts: a display choice, not an error rate",
                                    "controlled across contrasts.")))
    render_commentary("cross_omics_ora_heatmap")
}
```

```{r enrichment-cross-lookup, eval=has_enrichment, results='asis'}
# Each layer's top pathways read in every other layer, both directions.
# Discovered by glob, like the heatmaps: which directions exist depends on the
# layers a run enriched, and the lookup can be switched off in the config.
lookup_tsvs <- sort(list.files(enrichment_dir, pattern = "^cross_lookup_.*\\.tsv$",
                               full.names = TRUE))
if (length(lookup_tsvs) > 0) {
    cat("\n### Top Pathways of One Layer, Looked Up in Another {.unnumbered}\n\n")
    for (tsv in lookup_tsvs) {
        stem <- sub("\\.tsv$", "", tsv)
        pair <- sub("^cross_lookup_", "", basename(stem))
        from <- sub("_to_.*$", "", pair)
        to <- sub("^.*_to_", "", pair)
        cat(sprintf("\n#### %s to %s {.unnumbered}\n\n",
                    tools::toTitleCase(from), tools::toTitleCase(to)))
        for (suffix in c("", "_tested")) {
            png <- paste0(stem, suffix, ".png")
            if (!file.exists(png)) next
            if (nzchar(suffix)) {
                cat(sprintf("\nRestricted to pathways with a %s result:\n\n", to))
            }
            chunk_id <- paste0("enr-lookup-", gsub("[^a-zA-Z0-9]", "", pair),
                               gsub("_", "-", suffix))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, png
            ), quiet = TRUE))
            cat("\n\n")
        }
    }
}
```

```{r enrichment-cross-lookup-legend, eval=has_enrichment, results='asis'}
if (length(list.files(enrichment_dir, pattern = "^cross_lookup_.*\\.png$")) > 0) {
    figure_legend(paste(
        "Each layer's top pathways by its own raw p-value (GSEA, else ORA), among",
        "those its enrichment reported, beside the same pathways in the other layer",
        "and contrast (see the contrast columns), both directions. A KEGG-from-DE",
        "layer reports only pathways past its own cutoff, so it is not ranked over",
        "all it scored. Colour/number: NES, grey at zero; * raw p < 0.05. Unfilled:",
        "\"ORA\" where that layer has only ORA there, a dash where it has none. The",
        "second heatmap keeps only pathways the other layer tested. P-values are",
        "nominal; a shared NES sign is not by itself biological agreement."))
}
```

```{r enrichment-cross-lookup-table, eval=has_enrichment}
lookup_tsvs <- sort(list.files(enrichment_dir, pattern = "^cross_lookup_.*\\.tsv$",
                               full.names = TRUE))
lookup_cols <- c("ranking", "rank", "pathway", "source_method",
                 "source_contrast", "source_NES", "source_p", "source_padj",
                 "target_status", "target_method", "target_contrast",
                 "target_NES", "target_p", "target_padj", "target_ora_p",
                 "target_ora_padj")
for (tsv in lookup_tsvs) {
    tbl <- read.delim(tsv, stringsAsFactors = FALSE, check.names = FALSE)
    if (nrow(tbl) == 0) next
    pair <- gsub("_", " ", sub("^cross_lookup_", "", sub("\\.tsv$", "", basename(tsv))))
    print(make_dt(
        tbl[, intersect(lookup_cols, names(tbl)), drop = FALSE],
        caption = paste0("Top-pathway lookup, ", pair, " (nominal p-values)"),
        pageLength = 15
    ))
}
```

```{r enrichment-table, eval=has_enrichment}
meta_file <- file.path(enrichment_dir, "cross_omics_pathways_meta_analysis.csv")
if (file.exists(meta_file)) {
    meta_tbl <- read.csv(meta_file, stringsAsFactors = FALSE)
    make_dt(
        head(meta_tbl, 200),
        caption = "Cross-Omics Pathway Meta-Analysis (Stouffer combined p-values)",
        pageLength = 15
    )
}
```

```{r enrichment-table-legend, eval=has_enrichment, results='asis'}
meta_file <- file.path(enrichment_dir, "cross_omics_pathways_meta_analysis.csv")
if (file.exists(meta_file)) {
    figure_legend(paste(
        "Top pathways from cross-omics meta-analysis: per-layer p-values combined",
        "with Stouffer's method. Each layer contributes one kind of test -- its",
        "rank-based (GSEA) p-values where it has them, otherwise its ORA p-values --",
        "so no layer mixes methods; the method_<layer> columns record which. A",
        "contrast or collection a layer scored only with its other test is therefore",
        "left out of that layer's contribution, and the run log names any such case.",
        "Layers can still differ from each other in method; read the combined p-value",
        "with that in mind. Pathways significant in multiple omics are most informative."))
    render_commentary("cross_omics_meta_analysis")
}
```

```{r enzyme-metabolite-setup, include=FALSE}
enzyme_metab_file <- file.path(enrichment_dir, "enzyme_metabolite_pairs.tsv")
has_enzyme_metab <- file.exists(enzyme_metab_file)
```

`r if (has_enzyme_metab) "## Enzymes and the Metabolites They Act On {.unnumbered}"`

```{r enzyme-metabolite-table, eval=has_enzyme_metab}
# The writer stores missing values as blanks, so blanks are read back as NA;
# otherwise an unpaired enzyme's empty compound would count as a pair.
em <- read.delim(enzyme_metab_file, stringsAsFactors = FALSE, check.names = FALSE,
                 na.strings = "")
# The hit rules are shown: in a contrast where nothing cleared FDR, "changed"
# rests on raw p, and the reader has to be able to see that.
em_cols <- c("contrast", "gene_symbol", "protein", "ec", "enzyme_log2fc",
             "enzyme_p", "enzyme_padj", "enzyme_hit_rule", "compound_name",
             "compound", "metabolite", "metabolite_log2fc", "metabolite_p",
             "metabolite_padj", "metabolite_hit_rule", "role", "pathway_name",
             "same_direction", "note")
em_show <- head(em[, intersect(em_cols, names(em)), drop = FALSE], 200)
dt <- make_dt(em_show,
              caption = paste0("Enzymes and the measured metabolites they act on (",
                               sum(!is.na(em$compound)), " pairs, ",
                               sum(is.na(em$compound)), " enzymes with nothing ",
                               "measured to pair; the full table is ",
                               "enzyme_metabolite_pairs.tsv)"),
              pageLength = 15)
# The rows worth reading first are the ones where the metabolite itself cleared
# FDR; bold says so without hiding the rest.
if ("metabolite_padj" %in% names(em_show)) {
    dt <- DT::formatStyle(dt, "metabolite_padj", target = "row",
                          fontWeight = DT::styleInterval(0.05, c("bold", "normal")))
}
dt
```

```{r enzyme-metabolite-legend, eval=has_enzyme_metab, results='asis'}
# Built from the table and the settings it was made with, not fixed text.
figure_legend(describe_enzyme_metabolite_table(em, enzyme_metabolite_config(cfg)))
```

`r if (!single_contrast) "## Per-Contrast Enrichment {.tabset .unnumbered}"`

```{r enrichment-per-contrast, eval=has_enrichment && !single_contrast, results='asis'}
enrich_per_contrast_dir <- file.path(enrichment_dir, "per_contrast")
if (dir.exists(enrich_per_contrast_dir)) {
    enrich_contrast_dirs <- list.dirs(enrich_per_contrast_dir, full.names = TRUE, recursive = FALSE)
    for (cdir in enrich_contrast_dirs) {
        cname <- basename(cdir)
        cat(sprintf("\n### %s {.unnumbered}\n\n", gsub("_", " ", cname)))

        # Per-omics barplots for this contrast
        c_barplots <- list.files(cdir, pattern = "_top_pathways\\.png$", full.names = TRUE)
        for (cbp in c_barplots) {
            omics_name <- gsub("_top_pathways\\.png$", "", basename(cbp))
            cat(sprintf("\n#### %s {.unnumbered}\n\n", tools::toTitleCase(omics_name)))
            chunk_id <- paste0("enr-pc-", gsub("[^a-zA-Z0-9]", "", cname), "-",
                               gsub("[^a-zA-Z0-9]", "", omics_name))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, cbp
            ), quiet = TRUE))
            cat("\n\n")
        }

        # Cross-omics heatmaps for this contrast, one per gene-set collection.
        # Same contract as the run-level figures above: the analysis function
        # is called again per contrast with out_dir pointing here, so these
        # directories carry the same per-collection filenames.
        show_collection_pngs <- function(prefix, heading, id_stem) {
            pngs <- sort(list.files(
                cdir, pattern = paste0("^", prefix, "_.*\\.png$"),
                full.names = TRUE))
            if (length(pngs) == 0) return(FALSE)
            cat(sprintf("\n#### %s {.unnumbered}\n\n", heading))
            for (pg in pngs) {
                coll <- sub(paste0("^", prefix, "_"), "",
                            sub("\\.png$", "", basename(pg)))
                cat(sprintf("\n##### %s gene sets {.unnumbered}\n\n",
                            gsub("_", " ", coll)))
                chunk_id <- paste0(id_stem, gsub("[^a-zA-Z0-9]", "", cname), "-",
                                   gsub("[^a-zA-Z0-9]", "", coll))
                cat(knitr::knit_child(text = sprintf(
                    '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                    chunk_id, pg
                ), quiet = TRUE))
                collection_note(coll)
                cat("\n\n")
            }
            TRUE
        }
        drew_heatmap <- show_collection_pngs("cross_omics_pathway_heatmap",
                                             "Cross-Omics Heatmap", "enr-pc-hm-")
        drew_ora <- show_collection_pngs("cross_omics_ora_heatmap",
                                         "Cross-Omics ORA Evidence", "enr-pc-ora-")

        # Top-pathway lookups for this contrast; the "_tested" variants stay in
        # the directory beside them.
        c_lookups <- sort(list.files(cdir, pattern = "^cross_lookup_.*\\.png$",
                                     full.names = TRUE))
        c_lookups <- c_lookups[!grepl("_tested\\.png$", c_lookups)]
        if (length(c_lookups) > 0) {
            cat("\n#### Top Pathways of One Layer, Looked Up in Another {.unnumbered}\n\n")
            for (lp in c_lookups) {
                pair <- sub("^cross_lookup_", "", sub("\\.png$", "", basename(lp)))
                cat(sprintf("\n##### %s {.unnumbered}\n\n", gsub("_", " ", pair)))
                chunk_id <- paste0("enr-pc-lookup-", gsub("[^a-zA-Z0-9]", "", cname),
                                   "-", gsub("[^a-zA-Z0-9]", "", pair))
                cat(knitr::knit_child(text = sprintf(
                    '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                    chunk_id, lp
                ), quiet = TRUE))
                cat("\n\n")
            }
        }

        # Meta-analysis table for this contrast
        c_meta <- file.path(cdir, "cross_omics_pathways_meta_analysis.csv")
        if (file.exists(c_meta)) {
            c_meta_df <- read.csv(c_meta, stringsAsFactors = FALSE)
            cat(sprintf("\n**%s:** %d pathways in cross-omics meta-analysis\n\n",
                        gsub("_", " ", cname), nrow(c_meta_df)))
        }

        if (length(c_barplots) == 0 && !drew_heatmap && !drew_ora &&
            length(c_lookups) == 0) {
            cat("*No enrichment results for this contrast.*\n\n")
        }
    }
} else {
    cat("*Per-contrast enrichment results not available.*\n")
}
```

---

# Pathway Maps (Pathview) {.tabset}

```{r pathview-setup, include=FALSE}
# Two renderers can fill this section, on mutually exclusive branches of
# run_multi_ora(), and they do not make the same claim: "supported" means the
# pathway was enriched in two or more omics layers, while the no-OrgDb fallback
# unions the hits of either gene layer, so a map there can rest on one layer
# alone. The section says which of the two it is showing.
pathview_supported_pdf <- file.path(multi_ora_dir, "multi_ora_pathview_supported.pdf")
pathview_union_pdf <- file.path(multi_ora_dir, "multi_ora_pathview_union.pdf")
pathview_is_union <- !file.exists(pathview_supported_pdf) && file.exists(pathview_union_pdf)
pathview_pdf <- if (pathview_is_union) pathview_union_pdf else pathview_supported_pdf
pathview_dir <- file.path(multi_ora_dir, "pathview")
has_pathview <- file.exists(pathview_pdf)

# Whether compound nodes actually carry metabolomics values is recorded by the
# renderer beside its PDF, because the section must not promise a colour that is
# not on the map. Absent sidecar: assume not, and say less rather than more.
pathview_union_meta <- if (pathview_is_union) {
    tryCatch(yaml::read_yaml(file.path(multi_ora_dir,
                                       "multi_ora_pathview_union.yaml")),
             error = function(e) list())
} else {
    list()
}
# Both renderers record this now. The supported one used to be assumed to have
# compounds whenever metabolomics was in the run, which stopped being true once
# a metabolite had to clear the node thresholds to reach the map: every one of
# them can fail while the gene layers still produce a figure.
pathview_supported_meta <- if (pathview_is_union) {
    list()
} else {
    tryCatch(yaml::read_yaml(file.path(multi_ora_dir,
                                       "multi_ora_pathview_supported.yaml")),
             error = function(e) list())
}
pathview_has_compounds <- if (pathview_is_union) {
    isTRUE(pathview_union_meta$compound_nodes)
} else {
    isTRUE(pathview_supported_meta$compound_nodes)
}
# Map filenames carry a contrast key that cannot be read back into the
# contrast's own spelling; each renderer records the spelling beside its PDF.
# Read from whichever renderer drew these maps. A run from before the supported
# renderer recorded them falls back to reading the key.
pathview_active_meta <- if (pathview_is_union) pathview_union_meta else pathview_supported_meta
pathview_contrast_labels <- pathview_active_meta$contrast_labels %||% list()
# Set name and opening sentence from what the renderer recorded, so neither
# claims more layers than the maps rest on.
pathview_support <- pathview_multi_ora_support(pathview_is_union,
                                               pathview_supported_meta$support_layers)
```

```{r per-omics-pathview-setup, include=FALSE}
pv_metab_pdf <- file.path(multi_ora_dir, "pathview_top_metabolomics_pathways.pdf")
pv_prot_pdf <- file.path(multi_ora_dir, "pathview_top_proteomics_pathways.pdf")
has_pv_metab <- file.exists(pv_metab_pdf)
has_pv_prot <- file.exists(pv_prot_pdf)
has_per_omics_pv <- has_pv_metab || has_pv_prot
has_any_pv <- has_pathview || has_per_omics_pv

# One index feeds the overview table and every tab, so a map is named and
# assigned a contrast the same way in both; see pathview_map_index().
pv_index <- pathview_map_index(pathview_dir, multi_ora = has_pathview,
                               metab_top = has_pv_metab, prot_top = has_pv_prot,
                               multi_ora_label = pathview_support$label,
                               contrast_labels = pathview_contrast_labels)
```

```{r pathview-all-pdf, eval=has_any_pv, include=FALSE}
# Collect ALL pathview PNGs from all sections into one combined PDF; the
# overview tab links to it.
all_pv_pngs <- character(0)
has_combined_pv_pdf <- FALSE
combined_pdf <- file.path(multi_ora_dir, "all_pathview_combined.pdf")
if (dir.exists(pathview_dir)) {
    # Only the overlaid maps. pathview also leaves the blank KEGG reference PNG
    # it downloaded (ko00010.png beside ko00010.multi_ora.multi.png), which
    # filled half the PDF with pages carrying no data.
    all_pv_pngs <- sort(list.files(
        pathview_dir,
        pattern = "\\.(multi_ora|metab_top|prot_top)[^/]*\\.png$",
        full.names = TRUE))
}

if (length(all_pv_pngs) > 0) {
    compile_ok <- tryCatch({
        grDevices::pdf(combined_pdf, width = 12, height = 8)
        for (png_file in all_pv_pngs) {
            img <- png::readPNG(png_file)
            grid::grid.newpage()
            # Add pathway ID as title at top
            pw_label <- sub("\\..*$", "", basename(png_file))
            grid::grid.text(pw_label, x = 0.5, y = 0.98,
                            gp = grid::gpar(fontsize = 10, col = "grey40"))
            grid::grid.raster(img, width = 0.95, height = 0.92, y = 0.48)
        }
        grDevices::dev.off()
        TRUE
    }, error = function(e) {
        tryCatch(grDevices::dev.off(), error = function(e2) NULL)
        FALSE
    })

    has_combined_pv_pdf <- compile_ok && file.exists(combined_pdf)
}
```

## Overview {.unnumbered}

```{r pathview-overview, results='asis'}
if (!has_any_pv) {
    cat(pathview_missing_note(multi_cfg), "\n")
} else {
    cat("KEGG pathway maps with each layer's log2 fold changes drawn on the nodes.",
        "Each tab holds one set of maps; this table lists all of them.\n\n")
    if (isTRUE(has_combined_pv_pdf)) {
        cat(sprintf(
            '\n<div style="text-align: center; margin: 16px 0;">\n<a href="%s" download style="display: inline-block; padding: 10px 24px; background-color: #4a90d9; color: white; text-decoration: none; border-radius: 6px; font-weight: 600; font-size: 1em;">Download all pathway maps (PDF, %d figures)</a>\n</div>\n\n',
            combined_pdf, length(all_pv_pngs)))
    }
}
```

```{r pathview-index-table, eval=has_any_pv}
if (nrow(pv_index) > 0) {
    pv_table <- pv_index[, c("map_set", "contrast", "kegg_id", "pathway")]
    names(pv_table) <- c("Map set", "Contrast", "KEGG id", "Pathway")
    make_dt(pv_table, caption = "Pathway maps in this report", pageLength = 10)
}
```

`r if (has_pathview) "## Multi-Omics Maps {.unnumbered}"`

```{r pathview-plots, eval=has_pathview, results='asis'}
cat(pathview_support$intro,
    " Gene nodes are colored by logFC (transcriptomics and/or proteomics)",
    if (pathview_has_compounds) {
        "; compound nodes are colored by metabolomics logFC.\n\n"
    } else {
        paste(". No metabolomics compound values were available for these maps,",
              "so compound nodes are uncolored.\n\n")
    })
cat(sprintf("[Download these maps (PDF)](%s)\n\n", pathview_pdf))

pv_multi <- pv_index[pv_index$set == "multi_ora", ]
for (i in seq_len(nrow(pv_multi))) {
    cat(sprintf("\n### %s {.unnumbered}\n\n", pv_multi$heading[i]))
    cat(sprintf("![%s](%s)\n\n", pv_multi$heading[i], pv_multi$png[i]))
}
figure_legend(paste0(
    "KEGG maps drawn with pathview, per contrast. Gene nodes: transcriptomics",
    " (left) and proteomics (right) logFC, where available. ",
    if (pathview_has_compounds) "Compound nodes: metabolomics logFC. " else "",
    "Red up, green/blue down. ",
    # Taken from the same constants the selection and the node filter
    # read, so the caption cannot describe a rule the figure did not apply.
    pathview_significance_caption()))
render_commentary("multi_ora_pathview")
```

```{r per-omics-pathview-metab, eval=has_pv_metab, results='asis'}
cat("\n## Top Metabolomics Pathways (with Proteomics Overlay) {.unnumbered}\n\n")
cat("Top enriched metabolomics KEGG pathways with compound logFC (metabolomics) and ",
    "enzyme/gene logFC (proteomics) overlaid on the same pathway map.\n\n")
cat(sprintf("[Download PDF](%s)\n\n", pv_metab_pdf))
pv_metab <- pv_index[pv_index$set == "metab_top", ]
for (i in seq_len(nrow(pv_metab))) {
    cat(sprintf("\n### %s {.unnumbered}\n\n", pv_metab$heading[i]))
    cat(sprintf("![%s](%s)\n\n", pv_metab$heading[i], pv_metab$png[i]))
}
figure_legend("Top metabolomics-enriched KEGG pathways. Compound nodes show metabolomics logFC; gene/enzyme nodes show proteomics logFC. Red = upregulated, blue/green = downregulated.")
```

```{r per-omics-pathview-prot, eval=has_pv_prot, results='asis'}
cat("\n## Top Proteomics Pathways (with Metabolomics Overlay) {.unnumbered}\n\n")
cat("Top enriched proteomics KEGG pathways with enzyme/gene logFC (proteomics) and ",
    "compound logFC (metabolomics) overlaid on the same pathway map.\n\n")
cat(sprintf("[Download PDF](%s)\n\n", pv_prot_pdf))
pv_prot <- pv_index[pv_index$set == "prot_top", ]
for (i in seq_len(nrow(pv_prot))) {
    cat(sprintf("\n### %s {.unnumbered}\n\n", pv_prot$heading[i]))
    cat(sprintf("![%s](%s)\n\n", pv_prot$heading[i], pv_prot$png[i]))
}
figure_legend("Top proteomics-enriched KEGG pathways. Gene/enzyme nodes show proteomics logFC; compound nodes show metabolomics logFC. Red = upregulated, blue/green = downregulated.")
```

---

`r if (show_harmonization) "# Data Harmonization"`

```{r harmonization-summary, eval=show_harmonization, results='asis'}
# Look for harmonization summary files
harmony_summary_file <- file.path(run_dir, "harmonization_summary.csv")
if (file.exists(harmony_summary_file)) {
    harmony_summary <- read.csv(harmony_summary_file, stringsAsFactors = FALSE)
    cat("## Harmonization Overview\n\n")
    kable(harmony_summary)
} else {
    cat("Multi-omics data were harmonized into a unified **MultiAssayExperiment** object. ",
        "Each omics layer was aligned to a common set of samples.\n\n")

    # These are the integration/ subdirectories — one per method that produced
    # output (diablo, mofa, snf), not the omics layers.
    method_dirs <- list.dirs(integration_dir, full.names = FALSE, recursive = FALSE)
    if (length(method_dirs) > 0) {
        cat("**Integration methods with output:** ", paste(method_dirs, collapse = ", "), "\n\n")
    }
}
```

```{r gene-protein-mapping, eval=show_harmonization && !is.null(gene_protein_mapping), results='asis'}
cat("## Gene-Protein Mapping\n\n")
cat(sprintf("A total of **%d gene-protein pairs** were mapped across transcriptomics and proteomics layers.\n\n",
            nrow(gene_protein_mapping)))
```

```{r gene-protein-table, eval=show_harmonization && !is.null(gene_protein_mapping) && nrow(gene_protein_mapping) > 0}
make_dt(
    head(gene_protein_mapping, 500),
    caption = "Gene-Protein Mapping (first 500 rows)",
    pageLength = 10
)
```

```{r harmonization-legend, eval=show_harmonization && !is.null(gene_protein_mapping), results='asis'}
figure_legend("Gene-protein mapping table showing matched features between transcriptomics and proteomics layers.")
render_commentary("gene_protein_mapping")
```

---

```{r mantel-setup, include=FALSE}
foundational_dir <- file.path(run_dir, "foundational", "foundational")
mantel_heatmap <- file.path(foundational_dir, "plots", "mantel_test_heatmap.png")
mantel_csv <- file.path(foundational_dir, "tables", "mantel_test_results.csv")
clustering_consistency_plot <- file.path(foundational_dir, "plots", "sample_clustering_consistency.png")
condition_separation_plot <- file.path(foundational_dir, "plots", "sample_condition_separation.png")
has_mantel <- file.exists(mantel_heatmap)
has_clustering <- file.exists(clustering_consistency_plot)
has_separation <- file.exists(condition_separation_plot)

# Find per-omics correlation heatmaps
mantel_corr_heatmaps <- list.files(
    file.path(foundational_dir, "plots"),
    pattern = "^mantel_corr_heatmap_.*\\.png$",
    full.names = TRUE
)
has_corr_heatmaps <- length(mantel_corr_heatmaps) > 0
```

```{r group-dist-setup, include=FALSE}
group_dist_scatter <- file.path(foundational_dir, "plots", "sample_within_vs_cross_group_distance.png")
group_dist_boxplot <- file.path(foundational_dir, "plots", "sample_group_distance_boxplot.png")
# The switch is folded into has_group_dist so the heading, both figures and
# the table go together. Gating the figures alone -- which is what happened
# before -- left the heading standing over nothing.
has_group_dist <- show_sample_concordance && show_group_distance &&
    (file.exists(group_dist_scatter) || file.exists(group_dist_boxplot))
```

`r if (has_group_dist) "# Sample Concordance Across Omics"`

```{r clustering-consistency, eval=FALSE, results='asis'}
# Removed: Clustering Consistency and Condition Separation plots
# were not informative for this analysis
```

```{r group-dist-section, eval=has_group_dist, results='asis'}
cat("## Per-Sample Within vs Cross-Group Distance\n\n")
cat("For each sample, the mean Euclidean distance to other samples in the same condition group (within-group) and to samples in different groups (cross-group) is computed per omics layer. The x axis is the within-group distance and the y axis the cross-group distance, so samples ABOVE the diagonal are closer to their own group; samples below it are ambiguous or misplaced.\n\n")
```

```{r group-dist-scatter, eval=has_group_dist && file.exists(group_dist_scatter), results='asis'}
knitr::include_graphics(group_dist_scatter)
figure_legend("Per-sample within-group vs cross-group mean Euclidean distance, faceted by omics layer. Points above the diagonal line indicate samples that are closer to their own condition group (x = within-group, y = cross-group distance). Labeled points (if any) are samples closer to the other group than their own, which may indicate outliers, mislabeling, or biological heterogeneity.")
render_commentary("sample_group_distances")
```

```{r group-dist-boxplot, eval=has_group_dist && file.exists(group_dist_boxplot), results='asis'}
knitr::include_graphics(group_dist_boxplot)
figure_legend("Distribution of within-group and cross-group mean distances across all samples per omics layer. Clear separation between the two distributions indicates strong condition effect in that omics layer.")
```

```{r group-dist-table, eval=has_group_dist}
group_dist_csv <- file.path(foundational_dir, "tables", "sample_within_vs_cross_group_distances.csv")
if (file.exists(group_dist_csv)) {
    df <- read.csv(group_dist_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        df$within_group_dist <- round(df$within_group_dist, 2)
        df$cross_group_dist <- round(df$cross_group_dist, 2)
        df$dist_ratio <- round(df$dist_ratio, 3)
        make_dt(df, caption = "Per-sample within-group vs cross-group distances")
    }
}
```

---

# Integration Results {.tabset}

```{r integration-note, eval=!has_integration, results='asis'}
cat("*No integration results found. Integration methods may not have been configured or may have failed.*\n")
```

## DIABLO

```{r diablo-sample-plot, eval=has_diablo, results='asis'}
p <- file.path(diablo_dir, "diablo_sample_plot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r diablo-sample-legend, eval=has_diablo, results='asis'}
figure_legend("DIABLO sample scores plot. Samples are projected onto the first two components of the multi-omics discriminant analysis. Colors represent experimental groups.")
render_commentary("diablo_sample_plot")
```

```{r diablo-variable-plot, eval=has_diablo, results='asis'}
p <- file.path(diablo_dir, "diablo_variable_plot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r diablo-variable-legend, eval=has_diablo, results='asis'}
figure_legend("DIABLO variable loadings plot. Top contributing features from each omics layer are shown on the first two components.")
render_commentary("diablo_variable_plot")
```

```{r diablo-circos, eval=has_diablo, results='asis'}
circos_path <- file.path(diablo_dir, "diablo_circos.png")
if (file.exists(circos_path)) {
    cat("\n### Cross-Omics Circos Plot {.unnumbered}\n\n")
    knitr::include_graphics(circos_path)
}
```

```{r diablo-circos-legend, eval=has_diablo, results='asis'}
if (file.exists(file.path(diablo_dir, "diablo_circos.png"))) {
    figure_legend("DIABLO circos plot showing cross-omics feature correlations. Sectors represent omics layers; bar heights show DIABLO loading magnitude (component 1). Chords connect feature pairs with absolute Pearson correlation above the cutoff. Red chords indicate positive correlation; blue chords indicate negative correlation. Chord intensity and width scale with correlation strength.")
    render_commentary("diablo_circos")
}
```

```{r diablo-loadings-interactive, eval=has_diablo, results='asis'}
int_html <- file.path(diablo_dir, "diablo_loadings_interactive.html")
int_rds <- file.path(diablo_dir, "diablo_loadings_interactive.rds")
if (file.exists(int_rds)) {
    cat("\n### Interactive Loadings Plot {.unnumbered}\n\n")
    widget <- readRDS(int_rds)
    cat(knitr::knit_child(text = c(
        '```{r diablo-interactive-widget, echo=FALSE, fig.height=7}',
        'widget',
        '```'
    ), envir = environment(), quiet = TRUE))
    cat("\n\n")
    figure_legend("Interactive DIABLO loadings plot. Features are shaped by omics type and coloured by KEGG pathway membership where one is known; for organisms with no KEGG mapping every feature shows as Unmapped. Hover for feature details.")
} else if (file.exists(int_html)) {
    cat("\n### Interactive Loadings Plot {.unnumbered}\n\n")
    cat(sprintf('<iframe src="%s" width="100%%" height="650" frameborder="0" style="border:1px solid #ddd; border-radius:4px;"></iframe>\n\n', int_html))
    figure_legend("Interactive DIABLO loadings plot. Features are shaped by omics type and coloured by KEGG pathway membership where one is known; for organisms with no KEGG mapping every feature shows as Unmapped. Hover for feature details.")
}
```

```{r diablo-feature-heatmap, eval=has_diablo, results='asis'}
hm <- file.path(diablo_dir, "diablo_feature_heatmap.png")
hm_latent <- file.path(diablo_dir, "diablo_feature_heatmap_latent.png")
has_latent_hm <- file.exists(hm_latent)

if (file.exists(hm) && has_latent_hm) {
    cat("\n### Multi-Omics Feature Heatmap {.tabset .unnumbered}\n\n")
    cat("#### Expression Distance {.unnumbered}\n\n")
    cat(sprintf("![DIABLO Feature Heatmap](%s)\n\n", hm))
    figure_legend("Heatmap of top DIABLO features across omics layers. Values are z-scored per feature. Rows are clustered by Euclidean distance on expression values. The color bar on the left indicates the omics source layer.")

    cat("\n#### Latent Space Distance {.unnumbered}\n\n")
    cat(sprintf("![DIABLO Latent Space Heatmap](%s)\n\n", hm_latent))
    figure_legend("Same features as above, but rows are clustered by distance in the DIABLO latent component space (loading values). Features from different omics that contribute similarly to the discriminant components cluster together, highlighting cross-omics integration patterns.")
    render_commentary("diablo_feature_heatmap")
} else if (file.exists(hm)) {
    cat("\n### Multi-Omics Feature Heatmap {.unnumbered}\n\n")
    cat(sprintf("![DIABLO Feature Heatmap](%s)\n\n", hm))
    figure_legend("Heatmap of top DIABLO features across omics layers. Values are z-scored per feature. Rows are clustered by Euclidean distance. The color bar on the left indicates the omics source layer.")
    render_commentary("diablo_feature_heatmap")
}
```

```{r diablo-top-features, eval=has_diablo, results='asis'}
# Load top feature tables. results='asis' + knit_child is required here: without
# it the cat()ed headers were emitted as literal "## ### Top DIABLO Features"
# text and the DT widgets never rendered.
diablo_feat_files <- list.files(diablo_dir, pattern = "diablo_top_features_.*\\.csv$", full.names = TRUE)
if (length(diablo_feat_files) > 0) {
    for (ff in diablo_feat_files) {
        omics_name <- gsub("diablo_top_features_|\\.csv$", "", basename(ff))
        df <- tryCatch(read.csv(ff, stringsAsFactors = FALSE), error = function(e) NULL)
        if (!is.null(df) && nrow(df) > 0) {
            cat(sprintf("\n### Top DIABLO Features: %s\n\n", omics_name))
            chunk_id <- paste0("diablo-feat-", gsub("[^a-zA-Z0-9]", "", omics_name))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nmake_dt(read.csv("%s", stringsAsFactors = FALSE), caption = "DIABLO top features - %s")\n```\n',
                chunk_id, ff, omics_name
            ), quiet = TRUE))
        }
    }
}
```

## SNF

```{r snf-heatmap, eval=has_snf, results='asis'}
p <- file.path(snf_dir, "snf_network_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r snf-heatmap-legend, eval=has_snf, results='asis'}
figure_legend("SNF fused similarity network heatmap. Rows and columns are samples; color intensity represents similarity across all omics layers after network fusion.")
render_commentary("snf_network_heatmap")
```

```{r snf-silhouette, eval=has_snf, results='asis'}
p <- file.path(snf_dir, "snf_silhouette.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r snf-silhouette-legend, eval=has_snf, results='asis'}
figure_legend(paste(
    "SNF cluster silhouette plot. Silhouette width compares how well each sample",
    "fits its own cluster with the nearest alternative: near +1 a good fit, near 0",
    "on a boundary, negative a better fit elsewhere. The average across samples",
    "summarises structure: above 0.7 strong, 0.5-0.7 reasonable, 0.25-0.5 weak,",
    "below 0.25 none. It can guide the choice of K and flag outlier samples."))
render_commentary("snf_silhouette")
```

```{r snf-clusters-table, eval=has_snf}
clusters_file <- file.path(snf_dir, "snf_clusters.csv")
if (file.exists(clusters_file)) {
    clusters <- read.csv(clusters_file, stringsAsFactors = FALSE)
    make_dt(clusters, caption = "SNF Cluster Assignments")
}
```

## MOFA2

```{r mofa-variance, eval=has_mofa, results='asis'}
p <- file.path(mofa_dir, "mofa_variance_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mofa-variance-legend, eval=has_mofa, results='asis'}
figure_legend("MOFA2 variance explained heatmap. Each cell shows the proportion of variance in a given omics view (rows) explained by a latent factor (columns). Factors capturing shared variation across multiple views are most informative.")
render_commentary("mofa_variance_heatmap")
```

```{r mofa-factors-tabs, eval=has_mofa, results='asis'}
p12 <- file.path(mofa_dir, "mofa_factors_1_2.png")
p12_labeled <- file.path(mofa_dir, "mofa_factors_1_2_labeled.png")
p13 <- file.path(mofa_dir, "mofa_factors_1_3.png")

has_f12 <- file.exists(p12)
has_f12_lab <- file.exists(p12_labeled)
has_f13 <- file.exists(p13)

if (has_f12 || has_f12_lab) {
    cat("\n### Factor 1 vs Factor 2 {.tabset .unnumbered}\n\n")

    if (has_f12) {
        cat("#### By Condition {.unnumbered}\n\n")
        cat(knitr::knit_child(text = sprintf(
            '```{r mofa-f12-cond, echo=FALSE}\nknitr::include_graphics("%s")\n```\n', p12
        ), quiet = TRUE))
        figure_legend("MOFA2 Factor 1 vs Factor 2 scatter plot. Each point is a sample, colored by experimental condition.")
    }

    if (has_f12_lab) {
        cat("\n#### With Sample Labels {.unnumbered}\n\n")
        cat(knitr::knit_child(text = sprintf(
            '```{r mofa-f12-lab, echo=FALSE}\nknitr::include_graphics("%s")\n```\n', p12_labeled
        ), quiet = TRUE))
        figure_legend("MOFA2 Factor 1 vs Factor 2 with sample labels. Separation along factors indicates biologically relevant latent dimensions.")
    }

    render_commentary("mofa_factors_1_2")
}

if (has_f13) {
    cat("\n### Factor 1 vs Factor 3 {.unnumbered}\n\n")
    cat(knitr::knit_child(text = sprintf(
        '```{r mofa-f13, echo=FALSE}\nknitr::include_graphics("%s")\n```\n', p13
    ), quiet = TRUE))
    figure_legend("MOFA2 Factor 1 vs Factor 3 scatter plot.")
    render_commentary("mofa_factors_1_3")
}
```

```{r mofa-weights, eval=has_mofa, results='asis'}
weight_files <- list.files(mofa_dir, pattern = "^mofa_top_weights_.*\\.png$", full.names = TRUE)
# Separate the gene-symbol proteomics plot from the rest
genesym_file <- grep("proteomics_genesymbol", weight_files, value = TRUE)
main_files <- setdiff(weight_files, genesym_file)

if (length(main_files) > 0) {
    for (wf in main_files) {
        view_name <- gsub("mofa_top_weights_|\\.png$", "", basename(wf))
        chunk_id <- paste0("mofa-wt-", gsub("[^a-zA-Z0-9]", "", view_name))

        # For proteomics: create a tabset with UniProt and Gene Symbol tabs
        if (view_name == "proteomics" && length(genesym_file) == 1) {
            cat(sprintf("\n### Top Weights: %s {.tabset .unnumbered}\n\n", view_name))
            cat("#### UniProt ID {.unnumbered}\n\n")
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, wf
            ), quiet = TRUE))
            figure_legend(sprintf("Top 20 features by absolute weight for MOFA Factor in the %s view (UniProt IDs).", view_name))

            cat("\n#### Gene Symbol {.unnumbered}\n\n")
            gs_id <- paste0(chunk_id, "gs")
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                gs_id, genesym_file
            ), quiet = TRUE))
            figure_legend(sprintf("Top 20 features by absolute weight for MOFA Factor in the %s view (Gene Symbols).", view_name))
        } else {
            cat(sprintf("\n### Top Weights: %s\n\n", view_name))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, wf
            ), quiet = TRUE))
            figure_legend(sprintf("Top 20 features by absolute weight for MOFA Factor in the %s view.", view_name))
        }

        fig_id <- paste0("mofa_top_weights_", view_name)
        render_commentary(fig_id)
    }
}
```

## Loadings-Based Pathway Enrichment {.tabset}

```{r loadings-enrich-note, eval=!has_loadings_enrich, results='asis'}
cat("*No loadings-based enrichment results available. DIABLO/MOFA2 may not have been configured or enrichment may have failed.*\n")
```

```{r loadings-enrich-setup, include=FALSE}
# Both tests write into the same two directories, so which one this run used
# comes from the record run_loadings_enrichment() writes, not from which files
# happen to be there. Runs from before the record existed were ORA.
loadings_method_file <- file.path(loadings_enrich_dir, "loadings_enrichment_method.txt")
loadings_method <- if (file.exists(loadings_method_file)) {
    trimws(readLines(loadings_method_file, n = 1, warn = FALSE))
} else {
    "ora"
}

show_loadings_plots <- function(dir, pattern, strip, chunk_prefix) {
    plots <- sort(list.files(dir, pattern = pattern, full.names = TRUE))
    for (i in seq_along(plots)) {
        cat(sprintf("\n#### %s {.unnumbered}\n\n", gsub(strip, "", basename(plots[i]))))
        cat(knitr::knit_child(text = sprintf(
            '```{r %s-%d, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            chunk_prefix, i, plots[i]
        ), quiet = TRUE))
        cat("\n\n")
    }
    length(plots)
}

# print() inside a results='asis' chunk emits nothing for an htmlwidget, so the
# table goes through knit_child like the figures.
show_loadings_table <- function(csv, chunk_id, heading, caption) {
    if (!file.exists(csv)) return(invisible(FALSE))
    if (nrow(read.csv(csv, stringsAsFactors = FALSE)) == 0) return(invisible(FALSE))
    cat(sprintf("\n#### %s {.unnumbered}\n\n", heading))
    cat(knitr::knit_child(text = sprintf(
        '```{r %s, echo=FALSE}\nmake_dt(read.csv("%s", stringsAsFactors = FALSE), caption = "%s")\n```\n',
        chunk_id, csv, caption
    ), quiet = TRUE))
    invisible(TRUE)
}

loadings_gsea_legend <- paste(
    "GSEA over every mapped feature of the layer, ranked by its signed loading (DIABLO) or weight (MOFA2);",
    "genes against KEGG or the layer's GMTs, metabolites against KEGG compound pathways.",
    "Bars show NES, FDR beside each; the dashed red lines at NES = -2 and +2 are a visual guide, not a significance test.",
    "Axis signs are arbitrary: the subtitle names the condition scoring higher on the axis, and positive NES points towards it.")

# DIABLO is supervised and MOFA2 is not, so only the DIABLO legends carry this.
diablo_loadings_caveat <- paste(
    "DIABLO loadings were fitted to separate the conditions, so enrichment on them is descriptive:",
    "the FDR is not an independent, confirmatory error rate for a condition effect.")
```

```{r loadings-enrich-diablo, eval=has_loadings_enrich, results='asis'}
diablo_le_dir <- file.path(loadings_enrich_dir, "diablo_loadings")

if (identical(loadings_method, "gsea")) {
    if (dir.exists(diablo_le_dir)) {
        cat("\n### DIABLO Loadings Enrichment\n\n")
        n_plots <- show_loadings_plots(diablo_le_dir, "_gsea_nes\\.png$",
                                       "^DIABLO_|_gsea_nes\\.png$", "diablo-gsea")
        if (n_plots > 0) {
            figure_legend(paste(loadings_gsea_legend, diablo_loadings_caveat))
        } else if (!file.exists(file.path(diablo_le_dir, "diablo_loadings_gsea_all.csv"))) {
            cat("*No pathways were scored for the DIABLO loadings. This can occur when features do not map to the gene sets or compound pathways.*\n\n")
        }
        show_loadings_table(file.path(diablo_le_dir, "diablo_loadings_gsea_all.csv"),
                            "diablo-gsea-table", "DIABLO Loadings GSEA Table",
                            "DIABLO loadings GSEA")
    }
} else {
# ORA display, as before; only its own `_enrichment.png` files, so GSEA
# figures left in the directory by an earlier run are not shown as ORA.
diablo_le_plots <- list.files(diablo_le_dir, pattern = "_enrichment\\.png$", full.names = TRUE)

if (length(diablo_le_plots) > 0) {
    cat("\n### DIABLO Loadings Enrichment\n\n")
    for (i in seq_along(diablo_le_plots)) {
        pf <- diablo_le_plots[i]
        label <- gsub("_enrichment\\.png$", "", basename(pf))
        label <- gsub("^DIABLO_", "", label)
        cat(sprintf("\n#### %s {.unnumbered}\n\n", label))
        chunk_id <- paste0("diablo-le-", i)
        cat(knitr::knit_child(text = sprintf(
            '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            chunk_id, pf
        ), quiet = TRUE))
        cat("\n\n")
    }
    figure_legend(paste("KEGG pathway enrichment of top DIABLO loading features per component and omics layer. For gene-based omics (transcriptomics, proteomics), standard KEGG gene ORA is used. For metabolomics, compound-based ORA maps metabolites to KEGG metabolic pathways. Red dashed line indicates p = 0.05.", diablo_loadings_caveat))
} else if (dir.exists(diablo_le_dir)) {
    cat("\n### DIABLO Loadings Enrichment\n\n")
    cat("*No enriched pathways found for DIABLO loadings. This can occur when top features do not map to KEGG pathways or when enrichment p-values do not meet the significance threshold.*\n\n")
}

# Show combined table
diablo_le_csv <- file.path(diablo_le_dir, "diablo_loadings_enrichment_all.csv")
if (file.exists(diablo_le_csv)) {
    df <- read.csv(diablo_le_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        cat("\n#### DIABLO Loadings Enrichment Table {.unnumbered}\n\n")
        print(make_dt(df, caption = "DIABLO loadings KEGG enrichment"))
    }
}
}
```

```{r loadings-enrich-mofa, eval=has_loadings_enrich, results='asis'}
mofa_le_dir <- file.path(loadings_enrich_dir, "mofa_loadings")

if (identical(loadings_method, "gsea")) {
    if (dir.exists(mofa_le_dir)) {
        cat("\n### MOFA2 Weights Enrichment\n\n")
        n_plots <- show_loadings_plots(mofa_le_dir, "_gsea_nes\\.png$",
                                       "^MOFA_|_gsea_nes\\.png$", "mofa-gsea")
        if (n_plots > 0) {
            figure_legend(loadings_gsea_legend)
        } else if (!file.exists(file.path(mofa_le_dir, "mofa_weights_gsea_all.csv"))) {
            cat("*No pathways were scored for the MOFA2 weights. This can occur when features do not map to the gene sets or compound pathways.*\n\n")
        }
        show_loadings_table(file.path(mofa_le_dir, "mofa_weights_gsea_all.csv"),
                            "mofa-gsea-table", "MOFA2 Weights GSEA Table",
                            "MOFA2 weights GSEA")
    }
} else {
# ORA display, as before; only its own `_enrichment.png` files, so GSEA
# figures left in the directory by an earlier run are not shown as ORA.
mofa_le_plots <- list.files(mofa_le_dir, pattern = "_enrichment\\.png$", full.names = TRUE)

if (length(mofa_le_plots) > 0) {
    cat("\n### MOFA2 Weights Enrichment\n\n")
    for (i in seq_along(mofa_le_plots)) {
        pf <- mofa_le_plots[i]
        label <- gsub("_enrichment\\.png$", "", basename(pf))
        label <- gsub("^MOFA_", "", label)
        cat(sprintf("\n#### %s {.unnumbered}\n\n", label))
        chunk_id <- paste0("mofa-le-", i)
        cat(knitr::knit_child(text = sprintf(
            '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            chunk_id, pf
        ), quiet = TRUE))
        cat("\n\n")
    }
    figure_legend("KEGG pathway enrichment of top MOFA2 weight features per factor and omics view. For gene-based omics, standard KEGG gene ORA is used. For metabolomics, compound-based ORA maps metabolites to KEGG metabolic pathways. Red dashed line indicates p = 0.05.")
} else if (dir.exists(mofa_le_dir)) {
    cat("\n### MOFA2 Weights Enrichment\n\n")
    cat("*No enriched pathways found for MOFA2 weights. This can occur when top features do not map to KEGG pathways or when enrichment p-values do not meet the significance threshold.*\n\n")
}

# Show combined table
mofa_le_csv <- file.path(mofa_le_dir, "mofa_weights_enrichment_all.csv")
if (file.exists(mofa_le_csv)) {
    df <- read.csv(mofa_le_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        cat("\n#### MOFA2 Weights Enrichment Table {.unnumbered}\n\n")
        # print() inside a results='asis' chunk emits nothing for an htmlwidget,
        # so the heading appeared with no table under it.
        cat(knitr::knit_child(text = sprintf(
            '```{r mofa-le-table, echo=FALSE}\nmake_dt(read.csv("%s", stringsAsFactors = FALSE), caption = "MOFA2 weights KEGG enrichment")\n```\n',
            mofa_le_csv
        ), quiet = TRUE))
    }
}
}
```

---

# Consensus Analysis {.tabset}

```{r consensus-setup, include=FALSE}
consensus_dir <- file.path(run_dir, "consensus", "consensus")
has_consensus <- dir.exists(consensus_dir) &&
    length(list.files(consensus_dir, "\\.png$|\\.csv$", recursive = TRUE)) > 0
```

```{r consensus-note, eval=!has_consensus, results='asis'}
cat("*No consensus analysis results found. Requires >= 2 integration methods (e.g. DIABLO + MOFA2 + SNF).*\n")
```

## Meta-Integration PCA {.tabset .unnumbered}

```{r consensus-meta-pca, eval=has_consensus, results='asis'}
meta_pca_plot <- file.path(consensus_dir, "plots", "meta_integration_pca.png")
meta_pca_labeled <- file.path(consensus_dir, "plots", "meta_integration_pca_labeled.png")
has_meta_pca <- file.exists(meta_pca_plot)
has_meta_pca_labeled <- file.exists(meta_pca_labeled)

if (has_meta_pca || has_meta_pca_labeled) {
    if (has_meta_pca) {
        cat("\n### By Cluster {.unnumbered}\n\n")
        cat(knitr::knit_child(text = sprintf(
            '```{r meta-pca-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            meta_pca_plot
        ), quiet = TRUE))
        figure_legend("Meta-integration PCA: latent factors from all integration methods (DIABLO, MOFA2, SNF) are concatenated and projected into a combined principal component space. Points represent samples, coloured by the meta-clustering of the concatenated latent space (meta_cluster), which need not agree with the per-method consensus column. A 95% confidence ellipse is drawn only for clusters with enough samples to fit one.")
    }

    if (has_meta_pca_labeled) {
        cat("\n### With Sample Labels {.unnumbered}\n\n")
        cat(knitr::knit_child(text = sprintf(
            '```{r meta-pca-labeled-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            meta_pca_labeled
        ), quiet = TRUE))
        figure_legend("Meta-integration PCA with sample labels. Samples are projected into a combined latent space from all integration methods. Labels identify individual samples for cross-referencing with other analyses.")
    }

    render_commentary("meta_integration_pca")
} else {
    cat("*Meta-integration PCA plot not available.*\n")
}
```

```{r consensus-method-agreement, eval=FALSE}
# Removed: Method Agreement section was not informative
```

## Feature Importance Correlation {.unnumbered}

```{r consensus-feature-corr, eval=has_consensus, results='asis'}
feat_corr_plot <- file.path(consensus_dir, "plots", "feature_rank_correlation.png")
if (file.exists(feat_corr_plot)) {
    cat(knitr::knit_child(text = sprintf(
        '```{r feat-corr-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        feat_corr_plot
    ), quiet = TRUE))
    figure_legend("Spearman rank correlation of feature importance scores between integration methods. High correlations indicate that different methods identify similar key features.")
    render_commentary("feature_rank_correlation")
}
```

## Top Robust Features Across Methods {.unnumbered}

```{r consensus-robust-features, eval=has_consensus, results='asis'}
robust_plot <- file.path(consensus_dir, "plots", "robust_features.png")
if (file.exists(robust_plot)) {
    cat(knitr::knit_child(text = sprintf(
        '```{r robust-feat-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        robust_plot
    ), quiet = TRUE))
    figure_legend("Top robust features identified consistently across multiple integration methods. Tiles show each feature's importance rank percentile within a method. Features are kept when at least two methods rank them and their mean rank percentile is <= 0.2; the heatmap shows the top 30 of those, while the table below lists more.")
    render_commentary("robust_features")
} else {
    cat("*No robust features found across methods. This can occur when methods identify largely different feature sets.*\n")
}
```

```{r consensus-robust-table, eval=has_consensus}
robust_csv <- file.path(consensus_dir, "robust_features.csv")
if (file.exists(robust_csv)) {
    robust_df <- read.csv(robust_csv, stringsAsFactors = FALSE)
    if (nrow(robust_df) > 0) {
        make_dt(robust_df, caption = "Robust Features Across Integration Methods", pageLength = 15)
    }
}
```

---

`r if (show_mechanistic) "# Mechanistic & Causal Inference {.tabset}"`

```{r mech-setup, include=FALSE}
mechanistic_dir <- file.path(run_dir, "mechanistic", "mechanistic")
has_mechanistic <- show_mechanistic && dir.exists(mechanistic_dir) &&
    (length(list.files(file.path(mechanistic_dir, "plots"), "\\.png$")) > 0 ||
     length(list.files(file.path(mechanistic_dir, "tables"), "\\.csv$")) > 0)
```

```{r mech-note, eval=show_mechanistic && !has_mechanistic, results='asis'}
cat("*Mechanistic inference was not run or produced no results. Enable in config: `mechanistic.run_mechanistic: true`*\n")
```

`r if (has_mechanistic) "## Protein-Metabolite Correlations"`

```{r mech-prot-metab-summary, eval=has_mechanistic, results='asis'}
pm_csv <- file.path(mechanistic_dir, "tables", "protein_metabolite_significant_pairs.csv")
if (file.exists(pm_csv)) {
    pm_df <- read.csv(pm_csv, stringsAsFactors = FALSE)
    n_prot <- length(unique(pm_df$protein))
    n_metab <- length(unique(pm_df$metabolite))
    cat(sprintf("\n**Features shown:** %d proteins x %d metabolites. ", n_prot, n_metab))
    cat("Feature selection: differentially expressed features (pass\\_any filter from DE analysis) are used when available; otherwise the top variable features by variance are selected. ")
    cat(sprintf("Up to 100 proteins and 50 metabolites are included.\n\n"))
    render_commentary("protein_metabolite_correlations")
} else {
    cat("*No significant protein-metabolite pairs available. Requires both proteomics and metabolomics data with sufficient overlapping samples.*\n\n")
}
```

```{r mech-prot-metab-table, eval=has_mechanistic}
pm_pairs <- file.path(mechanistic_dir, "tables", "protein_metabolite_significant_pairs.csv")
if (file.exists(pm_pairs)) {
    df <- read.csv(pm_pairs, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        display_cols <- c("protein", "gene_symbol", "metabolite", "correlation",
                          "padj", "ec_number", "enzyme_metabolite_link",
                          "is_transcription_factor")
        display_cols <- intersect(display_cols, colnames(df))
        df_display <- df[, display_cols, drop = FALSE]
        make_dt(df_display,
                caption = "Significant protein-metabolite pairs (padj < 0.2, |r| > 0.7). EC = enzyme commission number; enzyme_metabolite_link = known KEGG enzyme-substrate/product relationship; is_transcription_factor = dorothea TF database.")
    }
} else {
    cat("*No significant protein-metabolite pairs found at the current thresholds.*\n\n")
}
```

`r if (has_mechanistic) "## Mediation: RNA -> Protein -> Phenotype"`

```{r mech-mediation-prop, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_mediation_proportion_barplot.png")
if (file.exists(p)) {
    cat(knitr::knit_child(text = sprintf(
        '```{r mech-med-prop-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        p
    ), quiet = TRUE))
    figure_legend("Proportion of the RNA effect on phenotype that is mediated through protein levels. Red dashed line marks 10% mediation threshold. Results are hypothesis-generating — the sequential ignorability assumption (no unmeasured confounders) is untestable.")
    render_commentary("mediation_proportion")
} else {
    cat("*RNA -> Protein -> Phenotype mediation not available. Requires both transcriptomics and proteomics data with >= 10 common samples and the `mediation` R package.*\n\n")
}
```

```{r mech-mediation-decomp, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_mediation_effect_decomposition.png")
if (file.exists(p)) {
    cat(knitr::knit_child(text = sprintf(
        '```{r mech-med-decomp-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        p
    ), quiet = TRUE))
    figure_legend("Decomposition of total effect into ACME (Average Causal Mediation Effect, the indirect/mediated path) and ADE (Average Direct Effect, not through protein).")
}
```

```{r mech-mediation-table, eval=has_mechanistic}
med_csv <- file.path(mechanistic_dir, "tables", "mech_mediation_analysis_results.csv")
if (file.exists(med_csv)) {
    df <- read.csv(med_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        if ("rho_at_acme_zero" %in% colnames(df)) {
            df$sensitivity <- ifelse(
                is.na(df$rho_at_acme_zero), "N/A",
                ifelse(is.infinite(df$rho_at_acme_zero), "Robust (ACME never zero)",
                ifelse(abs(df$rho_at_acme_zero) > 0.3, "Robust",
                ifelse(abs(df$rho_at_acme_zero) > 0.1, "Moderate", "Fragile")))
            )
        }
        make_dt(df, caption = "Mediation analysis: RNA -> Protein -> Phenotype")
    }
}
```

```{r mech-mediation-sens-legend, eval=has_mechanistic, results='asis'}
med_csv <- file.path(mechanistic_dir, "tables", "mech_mediation_analysis_results.csv")
if (file.exists(med_csv)) {
    df <- read.csv(med_csv, stringsAsFactors = FALSE)
    if ("rho_at_acme_zero" %in% colnames(df)) {
        cat("\n**Sensitivity interpretation (rho\\_at\\_acme\\_zero):** The rho value at which the mediation effect (ACME) would become zero under unmeasured confounding. |rho| > 0.3 = robust; |rho| < 0.1 = fragile (easily explained by confounders).\n\n")
    }
}
```

`r if (has_mechanistic) "## Protein -> Metabolite -> Phenotype Mediation"`

```{r mech-prot-metab-med-plot, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_prot_metab_mediation_barplot.png")
if (file.exists(p)) {
    cat(knitr::knit_child(text = sprintf(
        '```{r mech-pm-med-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        p
    ), quiet = TRUE))
    figure_legend("Protein -> Metabolite -> Phenotype mediation. Tests whether each protein's effect on the phenotype is mediated through its best-correlated metabolite. Does not require RNA-seq — uses proteomics + metabolomics directly. Red dashed line marks 10% mediation threshold. Results are hypothesis-generating.")
    render_commentary("prot_metab_mediation")
} else {
    cat("*No significant Protein -> Metabolite -> Phenotype mediation results. This can occur when protein-metabolite correlations do not significantly mediate the phenotype association.*\n\n")
}
```

```{r mech-prot-metab-med-table, eval=has_mechanistic}
pm_csv <- file.path(mechanistic_dir, "tables", "mech_prot_metab_mediation_results.csv")
if (file.exists(pm_csv)) {
    df <- read.csv(pm_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        if ("rho_at_acme_zero" %in% colnames(df)) {
            df$sensitivity <- ifelse(
                is.na(df$rho_at_acme_zero), "N/A",
                ifelse(is.infinite(df$rho_at_acme_zero), "Robust (ACME never zero)",
                ifelse(abs(df$rho_at_acme_zero) > 0.3, "Robust",
                ifelse(abs(df$rho_at_acme_zero) > 0.1, "Moderate", "Fragile")))
            )
        }
        make_dt(df, caption = "Protein -> Metabolite -> Phenotype mediation")
    }
}
```

`r if (has_mechanistic) "## Multi-Layer Mediation: Protein -> Metabolite -> Phenotype"`

```{r mech-multilayer-plot, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_multilayer_mediation_barplot.png")
if (file.exists(p)) {
    cat(knitr::knit_child(text = sprintf(
        '```{r mech-ml-med-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        p
    ), quiet = TRUE))
    figure_legend("Multi-layer mediation testing whether protein effects on phenotype are mediated through metabolite levels.")
    render_commentary("multilayer_mediation")
} else {
    cat("*Multi-layer mediation not available. Requires transcriptomics, proteomics, and metabolomics data with overlapping features.*\n\n")
}
```

```{r mech-multilayer-table, eval=has_mechanistic}
ml_csv <- file.path(mechanistic_dir, "tables", "mech_multilayer_mediation_results.csv")
if (file.exists(ml_csv)) {
    df <- read.csv(ml_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        make_dt(df, caption = "Multi-layer mediation: Protein -> Metabolite -> Phenotype")
    }
}
```

`r if (has_mechanistic) "## COSMOS Causal Network"`

```{r cosmos-network, eval=has_mechanistic, results='asis'}
cosmos_plot <- file.path(mechanistic_dir, "plots", "cosmos_causal_network.png")
if (file.exists(cosmos_plot)) {
    cat(knitr::knit_child(text = sprintf(
        '```{r cosmos-net-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        cosmos_plot
    ), quiet = TRUE))
    figure_legend("COSMOS causal network connecting deregulated transcription factors to metabolites through known signaling and metabolic interactions.")
    render_commentary("cosmos_network")
} else {
    organism <- cfg$global$organism %||% "unknown"
    if (tolower(organism) %in% c("human", "homo_sapiens", "mouse", "mus_musculus")) {
        cat("*COSMOS causal network was not generated. Requires TF activity estimation plus cosmosR, CARNIVAL, and decoupleR packages.*\n\n")
    } else {
        cat(sprintf("*COSMOS causal network is not available for %s. COSMOS requires human or mouse prior knowledge networks.*\n\n", organism))
    }
}
```

```{r cosmos-tables, eval=has_mechanistic}
cosmos_edges <- file.path(mechanistic_dir, "tables", "cosmos_network_edges.csv")
cosmos_nodes <- file.path(mechanistic_dir, "tables", "cosmos_network_nodes.csv")

if (file.exists(cosmos_nodes)) {
    df <- read.csv(cosmos_nodes, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        make_dt(df, caption = "COSMOS network nodes")
    }
}

if (file.exists(cosmos_edges)) {
    df <- read.csv(cosmos_edges, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        make_dt(df, caption = "COSMOS network edges")
    }
}
```

`r if (has_mechanistic) "## TF Activity"`

```{r mech-tf-heatmap, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_tf_activity_heatmap.png")
if (file.exists(p)) {
    cat(knitr::knit_child(text = sprintf(
        '```{r mech-tf-hm-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        p
    ), quiet = TRUE))
    figure_legend("Transcription factor activity inferred from target gene expression using dorothea/decoupleR regulons.")
    render_commentary("tf_activity")
} else {
    organism <- cfg$global$organism %||% "unknown"
    if (tolower(organism) %in% c("human", "homo_sapiens", "mouse", "mus_musculus")) {
        cat("*TF activity inference not available. Requires transcriptomics expression data and the decoupleR/dorothea package.*\n\n")
    } else {
        cat(sprintf("*TF activity inference is not available for %s. Dorothea TF regulons only cover human and mouse.*\n\n", organism))
    }
}
```

```{r mech-tf-barplot, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_tf_activity_barplot.png")
if (file.exists(p)) {
    cat(knitr::knit_child(text = sprintf(
        '```{r mech-tf-bp-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        p
    ), quiet = TRUE))
    figure_legend("Top differentially active transcription factors between conditions.")
}
```

`r if (has_mechanistic) "## Pathway Activity (PROGENy)"`

```{r mech-pathway-heatmap, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_pathway_activity_heatmap.png")
if (file.exists(p)) {
    cat(knitr::knit_child(text = sprintf(
        '```{r mech-pa-hm-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        p
    ), quiet = TRUE))
    figure_legend("PROGENy pathway activity inferred from proteomics data using decoupleR weighted mean.")
    render_commentary("pathway_activity")
} else {
    organism <- cfg$global$organism %||% "unknown"
    if (tolower(organism) %in% c("human", "homo_sapiens", "mouse", "mus_musculus")) {
        cat("*Pathway activity estimation via PROGENy was not available. The decoupleR package may not be installed.*\n\n")
    } else {
        cat(sprintf("*PROGENy pathway activity is not available for %s. PROGENy only supports human and mouse.*\n\n", organism))
    }
}
```

```{r mech-pathway-barplot, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_pathway_activity_barplot.png")
if (file.exists(p)) {
    cat(knitr::knit_child(text = sprintf(
        '```{r mech-pa-bp-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        p
    ), quiet = TRUE))
    figure_legend("Pathway activity differences between conditions.")
}
```

```{r mech-pathway-table, eval=has_mechanistic}
pa_csv <- file.path(mechanistic_dir, "tables", "mech_pathway_activity_associations.csv")
if (file.exists(pa_csv)) {
    df <- read.csv(pa_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        make_dt(df, caption = "Pathway activity associations with condition")
    }
}
```

`r if (has_mechanistic) "## Regulatory / Co-expression Network {.tabset}"`

```{r mech-network, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_regulatory_network.png")
network_rds <- file.path(mechanistic_dir, "plots", "mech_regulatory_network_interactive.rds")

p_gs <- file.path(mechanistic_dir, "plots", "mech_regulatory_network_genesymbol.png")

if (file.exists(p)) {
    cat("\n### Static (Feature ID) {.unnumbered}\n\n")
    cat(knitr::knit_child(text = sprintf(
        '```{r mech-network-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        p
    ), quiet = TRUE))
    figure_legend("Co-expression network inferred using GENIE3/ARACNe/correlation. Red nodes = hub regulators (high degree); blue nodes = non-hubs. Node size is proportional to degree (number of connections). Edges represent predicted co-regulatory relationships (association-based, not causal).")
    render_commentary("regulatory_network")
}

if (file.exists(p_gs)) {
    cat("\n### Static (Gene Symbol) {.unnumbered}\n\n")
    cat(knitr::knit_child(text = sprintf(
        '```{r mech-network-gs-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        p_gs
    ), quiet = TRUE))
    figure_legend("Same network as above but with gene symbol labels where available. Hub nodes (red) and non-hub nodes (blue) are labeled with gene symbols for easier biological interpretation.")
}

if (file.exists(network_rds)) {
    cat("\n### Interactive {.unnumbered}\n\n")
    net_widget <- readRDS(network_rds)
    cat(knitr::knit_child(text = c(
        '```{r mech-network-interactive, echo=FALSE, fig.height=8}',
        'net_widget',
        '```'
    ), envir = environment(), quiet = TRUE))
    cat("\n\n")
    figure_legend("Interactive network visualization. Hover over nodes to see feature name, gene symbol (if available), and degree. Drag nodes to rearrange. Scroll to zoom. Red = hub nodes, blue = non-hubs.")
}

if (!file.exists(p) && !file.exists(network_rds)) {
    cat("*No regulatory network plot available.*\n")
}
```

---

# Output Files

```{r output-files, results='asis'}
cat(sprintf("All multi-omics integration results have been saved to:\n\n`%s`\n\n", run_dir))

cat("**Subdirectories:**\n\n")

subdirs <- list.dirs(run_dir, full.names = FALSE, recursive = FALSE)
subdirs <- subdirs[subdirs != "execution_info"]
if (length(subdirs) > 0) {
    for (sd in sort(subdirs)) {
        n_files <- length(list.files(file.path(run_dir, sd), recursive = TRUE))
        cat(sprintf("- **%s/** (%d files)\n", sd, n_files))
    }
}
```

---

# Methods

```{r methods, results='asis'}
cat("## Multi-Omics Integration Pipeline\n\n")

cat("This report was generated by the **multiomics-core** pipeline. ",
    "The following analytical steps were performed:\n\n")

cat("1. **Data Harmonization**: Individual omics layers (preprocessed by their respective sub-pipelines) ",
    "were aligned to a common set of samples and assembled into a `MultiAssayExperiment` object.\n\n")

methods <- multi_cfg$integration$methods %||% c("DIABLO", "SNF")
step_num <- 2
if ("DIABLO" %in% methods) {
    cat(sprintf("%d. **DIABLO (Data Integration Analysis for Biomarker discovery using Latent cOmponents)**: ", step_num),
        "A supervised multi-omics integration method from the mixOmics package that identifies ",
        "discriminant features across omics layers using sparse PLS-DA.\n\n")
    step_num <- step_num + 1
}
if ("SNF" %in% methods) {
    cat(sprintf("%d. **SNF (Similarity Network Fusion)**: ", step_num),
        "An unsupervised method that constructs sample similarity networks for each omics layer ",
        "and fuses them into a single network for clustering.\n\n")
    step_num <- step_num + 1
}
if ("MOFA2" %in% methods) {
    cat(sprintf("%d. **MOFA2 (Multi-Omics Factor Analysis v2)**: ", step_num),
        "An unsupervised Bayesian factor model that identifies latent factors capturing ",
        "shared and omics-specific sources of variation.\n\n")
    step_num <- step_num + 1
}

cat(sprintf("%d. **Cross-Omics Enrichment**: Pathway enrichment results were aggregated across omics ", step_num),
    "using Stouffer's method for combined p-values.\n\n")

cat("## Software\n\n")
cat("```\n")
cat(capture.output(sessionInfo()), sep = "\n")
cat("\n```\n")
```
