---
title: "Multi-Omics Integration Report"
subtitle: "multiomics-core Pipeline"
author: "Bioinformatician: `r yaml::read_yaml(file.path(normalizePath(dirname(knitr::current_input())), 'execution_info', 'config_used.yaml'))$project$analyst`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    fig_width: 10
    fig_height: 7
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%"
)

library(yaml)
library(knitr)
library(DT)
library(plotly)
library(ggplot2)
library(htmltools)
library(jsonlite)

# Helper function for downloadable DT tables
make_dt <- function(df, caption = NULL, pageLength = 5) {
    # Format real-number columns: < 0.001 → scientific, else 3 decimals
    for (col in colnames(df)) {
        if (is.numeric(df[[col]])) {
            vals <- df[[col]]
            formatted <- ifelse(
                is.na(vals), NA_character_,
                ifelse(vals == as.integer(vals) & !grepl("\\.", as.character(vals)),
                       as.character(vals),
                       ifelse(abs(vals) < 0.001 & vals != 0,
                              sprintf("%.3e", vals),
                              sprintf("%.3f", vals))))
            df[[col]] <- formatted
        }
    }
    DT::datatable(
        df,
        caption = caption,
        extensions = 'Buttons',
        options = list(
            scrollX = TRUE,
            pageLength = pageLength,
            lengthMenu = c(5, 10, 25, 50, 100),
            dom = 'Blfrtip',
            buttons = list(
                list(extend = 'csv', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data")),
                list(extend = 'excel', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data"))
            ),
            columnDefs = list(list(className = 'dt-left', targets = '_all'))
        ),
        rownames = FALSE
    )
}

# Helper: render figure legend (paper-style caption below figure)
figure_legend <- function(text) {
    cat(sprintf("<p class='figure-legend' style='text-align: center; font-size: 0.9em; color: #555; margin: 4px auto 12px auto; max-width: 88%%; line-height: 1.45; font-style: italic;'>%s</p>\n", text))
}

# Helper: render commentary block for a figure_id (collapsible)
render_commentary <- function(figure_id) {
    if (!has_commentary) return(invisible(NULL))
    c <- commentary[[figure_id]]
    if (is.null(c)) return(invisible(NULL))

    cat("<details class='commentary-details' style='margin: 4px 0 18px 0;'>\n")
    cat("<summary style='cursor: pointer; color: #4a90d9; font-weight: 500; font-size: 0.93em; padding: 3px 0;'>Show AI Interpretation</summary>\n")

    html <- "<div class='commentary-block' style='background: #f8f9fa; border-left: 4px solid #4a90d9; padding: 12px 16px; margin: 6px 0 4px 0; border-radius: 0 4px 4px 0; font-size: 0.92em;'>\n"
    html <- paste0(html, "<p style='margin: 0 0 8px 0;'><strong>What this shows:</strong> ",
                   htmltools::htmlEscape(c$what_is_this %||% ""), "</p>\n")

    bio_interp <- c$biological_interpretation %||% ""
    if (nzchar(bio_interp)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Interpretation:</strong> ",
                       htmltools::htmlEscape(bio_interp), "</p>\n")
    }

    obs <- c$observations
    if (length(obs) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Key observations:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (o in obs) html <- paste0(html, "<li>", htmltools::htmlEscape(o), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    stat_assess <- c$statistical_assessment %||% ""
    if (nzchar(stat_assess)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Statistical note:</strong> ",
                       htmltools::htmlEscape(stat_assess), "</p>\n")
    }

    issues <- c$issues_checks
    if (length(issues) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Checks:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (issue in issues) html <- paste0(html, "<li>", htmltools::htmlEscape(issue), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    steps <- c$next_steps
    if (length(steps) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Next steps:</strong></p>\n<ul style='margin: 2px 0 4px 0; padding-left: 20px;'>\n")
        for (s in steps) html <- paste0(html, "<li>", htmltools::htmlEscape(s), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    backend <- c$backend %||% "unknown"
    confidence <- c$confidence %||% ""
    backend_label <- if (grepl("claude", backend, ignore.case = TRUE)) {
        "Claude (Anthropic) + K-Dense AI scientific-skills"
    } else if (grepl("openai", backend, ignore.case = TRUE)) {
        "OpenAI + K-Dense AI scientific-skills"
    } else if (backend == "data_driven") {
        "data-driven + K-Dense AI scientific-skills"
    } else {
        backend
    }
    if (nzchar(confidence) && confidence != "none") {
        backend_label <- paste0(backend_label, " | confidence: ", confidence)
    }
    html <- paste0(html, "<p style='margin: 4px 0 0 0; color: #888; font-size: 0.85em;'><em>Generated by: ",
                   htmltools::htmlEscape(backend_label), "</em></p>\n</div>\n")

    cat(html)
    cat("</details>\n")
}

# Paths
run_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
config_path <- file.path(run_dir, "execution_info", "config_used.yaml")

# Multiomics output subdirectories
integration_dir  <- file.path(run_dir, "integration")
diablo_dir       <- file.path(integration_dir, "diablo")
snf_dir          <- file.path(integration_dir, "snf")
mofa_dir         <- file.path(integration_dir, "mofa")
enrichment_dir   <- file.path(run_dir, "cross_enrichment")
loadings_enrich_dir <- file.path(run_dir, "loadings_enrichment")
rna_prot_dir     <- file.path(run_dir, "rna_protein_correlation")
multigsea_dir    <- file.path(run_dir, "multigsea")
commentary_dir   <- file.path(run_dir, "commentary")

# Load commentary if available
commentary <- NULL
has_commentary <- FALSE
commentary_json <- file.path(commentary_dir, "all_commentary.json")
if (file.exists(commentary_json)) {
    commentary <- tryCatch(
        jsonlite::read_json(commentary_json),
        error = function(e) NULL
    )
    has_commentary <- !is.null(commentary) && length(commentary) > 0
}

# Load config
cfg <- yaml::read_yaml(config_path)
multi_cfg <- cfg$modes$multiomics

# Detect which sections have data
has_diablo      <- dir.exists(diablo_dir) && length(list.files(diablo_dir, "\\.png$")) > 0
has_snf         <- dir.exists(snf_dir) && length(list.files(snf_dir, "\\.png$")) > 0
has_mofa        <- dir.exists(mofa_dir) && length(list.files(mofa_dir, "\\.png$")) > 0
has_integration <- has_diablo || has_snf || has_mofa
has_enrichment  <- dir.exists(enrichment_dir) && length(list.files(enrichment_dir, "\\.png$|\\.csv$")) > 0
has_transcriptomics <- "transcriptomics" %in% (cfg$global$omics_present %||% character(0)) ||
    !is.null(cfg$modes$rna)
has_rna_prot    <- has_transcriptomics &&
    dir.exists(file.path(rna_prot_dir, "plots")) &&
    length(list.files(file.path(rna_prot_dir, "plots"), "\\.png$")) > 0
has_multigsea   <- dir.exists(multigsea_dir) && length(list.files(multigsea_dir, "\\.png$")) > 0
has_loadings_enrich <- dir.exists(loadings_enrich_dir) &&
    length(list.files(loadings_enrich_dir, "\\.png$", recursive = TRUE)) > 0

# Load gene-protein mapping if available
mapping_file <- file.path(run_dir, "gene_protein_mapping.csv")
gene_protein_mapping <- NULL
if (file.exists(mapping_file)) {
    gene_protein_mapping <- tryCatch(
        read.csv(mapping_file, stringsAsFactors = FALSE),
        error = function(e) NULL
    )
}

# Join descriptions onto the mapping table (order-preserving, appends columns)
if (!is.null(gene_protein_mapping)) {
    rna_annot_file <- file.path(cfg$project$dir, cfg$paths$raw,
        "Serge_Ankri_June2026/RNAseq/ehi_gene_annotation.csv")
    if (file.exists(rna_annot_file)) {
        ra <- tryCatch(read.csv(rna_annot_file, stringsAsFactors = FALSE),
                       error = function(e) NULL)
        if (!is.null(ra) && all(c("gene_id", "gene_description") %in% colnames(ra))) {
            idx <- match(gene_protein_mapping$gene_id, ra$gene_id)
            gene_protein_mapping$gene_name        <- ra$gene_name[idx]
            gene_protein_mapping$gene_description <- ra$gene_description[idx]
        }
    }
    prot_annot_file <- file.path(cfg$project$dir, cfg$paths$raw,
                                 cfg$modes$proteomics$files$protein)
    if (file.exists(prot_annot_file)) {
        pa <- tryCatch(read.delim(prot_annot_file, stringsAsFactors = FALSE,
                                  check.names = FALSE), error = function(e) NULL)
        if (!is.null(pa) && "Protein.Group" %in% colnames(pa) && "Genes" %in% colnames(pa)) {
            idx <- match(gene_protein_mapping$protein_id, pa$Protein.Group)
            gene_protein_mapping$protein_symbol <- pa$Genes[idx]
        }
    }

    # Join per-contrast log2FC (RNA and proteomics) for the run's shared contrast
    # from the volcano tables, keyed by gene_id (EHI_) and protein_id (XP_).
    rp_parent  <- dirname(run_dir)
    rna_enr_m  <- file.path(rp_parent, "rna", "Enrichment")
    prot_enr_m <- file.path(rp_parent, "proteomics", "Enrichment")
    shared_bn <- intersect(list.files(rna_enr_m,  "^pathway_volcano_data_.*\\.csv$"),
                           list.files(prot_enr_m, "^pathway_volcano_data_.*\\.csv$"))
    if (length(shared_bn) >= 1) {
        rv <- tryCatch(read.csv(file.path(rna_enr_m,  shared_bn[1]), stringsAsFactors = FALSE),
                       error = function(e) NULL)
        pv <- tryCatch(read.csv(file.path(prot_enr_m, shared_bn[1]), stringsAsFactors = FALSE),
                       error = function(e) NULL)
        if (!is.null(rv) && all(c("FeatureID", "log2FC") %in% colnames(rv))) {
            gene_protein_mapping$rna_log2FC <-
                round(rv$log2FC[match(gene_protein_mapping$gene_id, rv$FeatureID)], 3)
        }
        if (!is.null(pv) && all(c("FeatureID", "log2FC") %in% colnames(pv))) {
            gene_protein_mapping$protein_log2FC <-
                round(pv$log2FC[match(gene_protein_mapping$protein_id, pv$FeatureID)], 3)
        }
    }
}
```

```{r force-dependencies, include=FALSE}
# Force DT and Plotly dependencies to be included in the HTML
DT::datatable(data.frame(x = 1))
plotly::plot_ly(data.frame(x = 1, y = 1), x = ~x, y = ~y, type = "scatter", mode = "markers")
```

# Analysis Parameters {.tabset}

## Project

```{r params-project}
proj <- data.frame(
  Parameter = c("Project Name", "Analysis Round", "Project Directory",
                 "Raw Data Path", "Output Path"),
  Value = c(cfg$project$name %||% "N/A",
            cfg$project$analysis_round %||% "N/A",
            cfg$project$dir %||% "N/A",
            cfg$paths$raw %||% "N/A",
            cfg$paths$out %||% "N/A"),
  stringsAsFactors = FALSE
)
kable(proj, col.names = c("Parameter", "Value"))
```

## Integration Methods

```{r params-integration}
methods <- multi_cfg$integration$methods %||% c("DIABLO", "SNF")
feat_sel <- multi_cfg$feature_selection$method %||% "variance"
feat_n <- multi_cfg$feature_selection$top_n %||% 500
cond_col <- multi_cfg$condition_column %||% cfg$design$condition_column %||% "condition"

settings <- data.frame(
  Parameter = c(
    "Integration Methods",
    "Feature Selection Method",
    "Top N Features",
    "Condition Column",
    "Require 1:1 Gene-Protein Mapping"
  ),
  Value = c(
    paste(methods, collapse = ", "),
    feat_sel,
    feat_n,
    cond_col,
    ifelse(isTRUE(multi_cfg$require_one_to_one_mapping), "Yes", "No")
  ),
  stringsAsFactors = FALSE
)
kable(settings, col.names = c("Parameter", "Value"))
```

## Omics Layers

```{r params-omics}
omics_modes <- setdiff(names(cfg$modes), "multiomics")
omics_tbl <- data.frame(
    Omics = omics_modes,
    Configured = "Yes",
    stringsAsFactors = FALSE
)
kable(omics_tbl, col.names = c("Omics Layer", "Configured"))
```

---

`r if (has_transcriptomics) "# RNA-Protein Correlation {.tabset}"`

```{r rna-prot-note, eval=has_transcriptomics && !has_rna_prot, results='asis'}
cat("*No RNA-protein correlation results found. This analysis requires both transcriptomics and proteomics data.*\n")
```

## RNA vs Protein log2FC

```{r rna-prot-log2fc, eval=has_transcriptomics}
# RNA-vs-protein differential concordance from the per-contrast DE tables that
# feed the volcano plots. RNA FeatureID = EHI_ gene_id; protein FeatureID = XP_
# Protein.Group; the EHI_<->XP_ bridge is gene_protein_mapping.csv. Contrasts are
# matched by identical file name so RNA and protein compare the same condition.
results_parent <- dirname(run_dir)
rna_enr  <- file.path(results_parent, "rna", "Enrichment")
prot_enr <- file.path(results_parent, "proteomics", "Enrichment")

# A protein is "not imputed" in a contrast when it has a measured (non-missing,
# positive) intensity in EVERY sample of the contrast's two groups in the raw
# pg_matrix; the pipeline min-val-imputes the missing ones. Raw sample columns
# are named "<Group>_<rep>", so the group is the column name minus the replicate.
prot_fully_observed <- function(cfg, groups) {
    pg_path <- file.path(cfg$project$dir %||% ".", cfg$paths$raw %||% "data",
                         cfg$modes$proteomics$files$protein %||% "")
    if (!file.exists(pg_path)) return(NULL)
    pg <- tryCatch(read.delim(pg_path, check.names = FALSE, stringsAsFactors = FALSE),
                   error = function(e) NULL)
    if (is.null(pg)) return(NULL)
    id_col <- cfg$modes$proteomics$id_columns$protein_id %||% "Protein.Group"
    if (!id_col %in% colnames(pg)) return(NULL)
    grp_of <- sub("_[0-9]+$", "", colnames(pg))
    scols  <- colnames(pg)[grp_of %in% groups & colnames(pg) != id_col]
    if (length(scols) == 0) return(NULL)
    obs <- vapply(pg[, scols, drop = FALSE], function(col) {
        x <- suppressWarnings(as.numeric(as.character(col)))
        is.finite(x) & x > 0
    }, logical(nrow(pg)))
    if (is.null(dim(obs))) obs <- matrix(obs, nrow = nrow(pg))
    data.frame(protein_id  = pg[[id_col]],
               not_imputed = rowSums(!obs) == 0,
               stringsAsFactors = FALSE)
}

# Reusable RNA-vs-protein log2FC scatter with per-panel Pearson/Spearman.
rp_scatter <- function(d, title) {
    labs <- sapply(split(d, d$contrast), function(x) {
        sprintf("Pearson r = %.3f | Spearman rho = %.3f | n = %d",
                cor(x$rna_log2FC, x$protein_log2FC, method = "pearson"),
                cor(x$rna_log2FC, x$protein_log2FC, method = "spearman"), nrow(x))
    })
    d$facet <- paste0(d$contrast, "\n(", labs[d$contrast], ")")
    ggplot2::ggplot(d, ggplot2::aes(x = rna_log2FC, y = protein_log2FC)) +
        ggplot2::geom_hline(yintercept = 0, color = "gray90") +
        ggplot2::geom_vline(xintercept = 0, color = "gray90") +
        ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
        ggplot2::geom_point(alpha = 0.5, size = 1.4, color = "steelblue") +
        ggplot2::geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.5) +
        ggplot2::facet_wrap(~ facet, scales = "free") +
        ggplot2::theme_minimal() +
        ggplot2::labs(title = title, x = "RNA log2 fold change", y = "Protein log2 fold change")
}

rp_pairs <- NULL
if (!is.null(gene_protein_mapping) &&
    all(c("gene_id", "protein_id") %in% colnames(gene_protein_mapping)) &&
    dir.exists(rna_enr) && dir.exists(prot_enr)) {

    rna_v  <- list.files(rna_enr,  "^pathway_volcano_data_.*\\.csv$")
    prot_v <- list.files(prot_enr, "^pathway_volcano_data_.*\\.csv$")
    shared <- intersect(rna_v, prot_v)
    map_sub <- unique(gene_protein_mapping[, c("gene_id", "protein_id")])

    rp_pairs <- do.call(rbind, lapply(shared, function(bn) {
        rna_df  <- tryCatch(read.csv(file.path(rna_enr, bn),  stringsAsFactors = FALSE), error = function(e) NULL)
        prot_df <- tryCatch(read.csv(file.path(prot_enr, bn), stringsAsFactors = FALSE), error = function(e) NULL)
        if (is.null(rna_df) || is.null(prot_df)) return(NULL)
        if (!all(c("FeatureID", "log2FC") %in% colnames(rna_df)))  return(NULL)
        if (!all(c("FeatureID", "log2FC") %in% colnames(prot_df))) return(NULL)

        rna_cols <- c("FeatureID", "log2FC",
                      if ("simple_log2FC" %in% colnames(rna_df)) "simple_log2FC")
        m <- merge(map_sub, rna_df[, rna_cols, drop = FALSE],
                   by.x = "gene_id", by.y = "FeatureID")
        names(m)[names(m) == "log2FC"] <- "rna_log2FC"
        names(m)[names(m) == "simple_log2FC"] <- "rna_simple_log2FC"
        m <- merge(m, prot_df[, c("FeatureID", "log2FC")],
                   by.x = "protein_id", by.y = "FeatureID")
        names(m)[names(m) == "log2FC"] <- "protein_log2FC"
        if (nrow(m) == 0) return(NULL)

        m$te_log2FC    <- m$protein_log2FC - m$rna_log2FC
        m$contrast_raw <- sub("\\.csv$", "", sub("^pathway_volcano_data_", "", bn))
        m$contrast     <- gsub("_", " ", m$contrast_raw)
        m[is.finite(m$rna_log2FC) & is.finite(m$protein_log2FC), ]
    }))
}

# Flag each pair's protein as not-imputed (fully observed) for its contrast.
if (!is.null(rp_pairs) && nrow(rp_pairs) > 0) {
    rp_pairs$prot_not_imputed <- NA
    for (cr in unique(rp_pairs$contrast_raw)) {
        groups <- strsplit(cr, "_vs_", fixed = TRUE)[[1]]
        look <- tryCatch(prot_fully_observed(cfg, groups), error = function(e) NULL)
        if (is.null(look)) next
        idx <- rp_pairs$contrast_raw == cr
        rp_pairs$prot_not_imputed[idx] <-
            look$not_imputed[match(rp_pairs$protein_id[idx], look$protein_id)]
    }
}

# RNA simple log2FC (naive group-mean log-ratio); falls back to the DESeq2
# log2FC on older runs whose volcano CSVs predate the simple_log2FC column.
if (!is.null(rp_pairs) && nrow(rp_pairs) > 0) {
    if (!"rna_simple_log2FC" %in% names(rp_pairs) || all(is.na(rp_pairs$rna_simple_log2FC))) {
        rp_pairs$rna_simple_log2FC <- rp_pairs$rna_log2FC
    }
}

# The correlation test uses ONLY fully-measured proteins: every replicate in
# both conditions has a real (non-missing) intensity, so the fold change carries
# no min-val-imputed values. rp_fm is that subset; rp_pairs is kept for the
# direction-agreement comparison below. Fall back to all pairs only if imputation
# status could not be derived (raw pg_matrix unavailable).
fm_ok  <- !is.null(rp_pairs) && "prot_not_imputed" %in% colnames(rp_pairs) &&
    any(rp_pairs$prot_not_imputed %in% TRUE)
rp_fm  <- if (fm_ok) rp_pairs[which(rp_pairs$prot_not_imputed), , drop = FALSE] else rp_pairs
fm_ttl <- if (fm_ok) {
    "RNA vs Protein log2FC — fully-measured proteins (quantified in every replicate)"
} else {
    "RNA vs Protein log2FC — all mapped pairs (imputation status unavailable)"
}

if (is.null(rp_fm) || nrow(rp_fm) < 3) {
    cat("*Fewer than 3 fully-measured gene-protein pairs (protein quantified in every replicate) were found for this run.*\n")
} else {
    print(rp_scatter(rp_fm, fm_ttl))

    tbl <- rp_fm[, c("contrast", "gene_id", "protein_id",
                     "rna_log2FC", "protein_log2FC", "te_log2FC")]
    tbl <- tbl[order(-(abs(tbl$rna_log2FC) + abs(tbl$protein_log2FC))), ]
    rp_dt <- make_dt(tbl, caption = "RNA vs protein log2FC per fully-measured gene-protein pair (protein quantified in every replicate)")
}
# Widgets only auto-render as a top-level expression, not from inside if/for.
if (exists("rp_dt")) rp_dt
```

```{r rna-prot-log2fc-legend, eval=has_transcriptomics, results='asis'}
figure_legend("RNA-seq log2 fold change (x) versus proteomics log2 fold change (y), one panel per contrast, restricted to fully-measured proteins: a real (non-missing) intensity in every replicate of both conditions in the raw pg_matrix (proteomics files.protein), so the protein fold change carries no min-val-imputed values. Solid line is an OLS fit; the dashed diagonal marks equal RNA and protein change. Sources: RNA DESeq2 and proteomics limma per-contrast fold changes (pathway_volcano_data_<contrast>.csv in rna/Enrichment and proteomics/Enrichment); pairing via gene_protein_mapping.csv (gene_id EHI_ to protein_id XP_). Pearson/Spearman and n (per panel) are computed on this fully-measured subset. Single contrast, hypothesis-generating.")
```

### Fold-change direction agreement

```{r rna-prot-concordance-pie, eval=has_transcriptomics, fig.width=8, fig.height=3.8}
# How often RNA and protein move the same way (sign agreement), for all mapped
# pairs and for the fully-measured subset (protein quantified in every replicate).
# Pairs with a zero fold change on either side are excluded from the count.
if (exists("rp_pairs") && !is.null(rp_pairs) && nrow(rp_pairs) >= 3) {
    dir_counts <- function(d, panel) {
        d <- d[d$rna_log2FC != 0 & d$protein_log2FC != 0, , drop = FALSE]
        same <- sign(d$rna_log2FC) == sign(d$protein_log2FC)
        data.frame(panel = sprintf("%s (n = %d)", panel, nrow(d)),
                   direction = c("Same direction", "Opposite direction"),
                   n = c(sum(same), sum(!same)),
                   stringsAsFactors = FALSE)
    }
    pieces <- list(dir_counts(rp_pairs, "All mapped pairs"))
    if ("prot_not_imputed" %in% colnames(rp_pairs)) {
        ni <- rp_pairs[which(rp_pairs$prot_not_imputed), , drop = FALSE]
        if (nrow(ni) > 0) pieces[[2]] <- dir_counts(ni, "Fully-measured proteins")
    }
    pie_df <- do.call(rbind, pieces)
    pie_df$panel     <- factor(pie_df$panel, levels = unique(pie_df$panel))
    pie_df$direction <- factor(pie_df$direction,
                               levels = c("Same direction", "Opposite direction"))
    pie_df$pct <- ave(pie_df$n, pie_df$panel, FUN = function(x) 100 * x / sum(x))
    pie_df$lab <- ifelse(pie_df$n > 0, sprintf("%d (%.0f%%)", pie_df$n, pie_df$pct), "")

    p <- ggplot2::ggplot(pie_df, ggplot2::aes(x = "", y = n, fill = direction)) +
        ggplot2::geom_col(width = 1, color = "white") +
        ggplot2::coord_polar(theta = "y") +
        ggplot2::facet_wrap(~ panel) +
        ggplot2::geom_text(ggplot2::aes(label = lab),
                           position = ggplot2::position_stack(vjust = 0.5), size = 3) +
        ggplot2::scale_fill_manual(
            values = c("Same direction" = "#4C72B0", "Opposite direction" = "#DD8452"),
            name = NULL) +
        ggplot2::theme_void() +
        ggplot2::labs(title = "RNA / protein fold-change direction agreement")
    print(p)
} else {
    cat("*Not enough matched pairs to summarise fold-change direction agreement.*\n")
}
```

```{r rna-prot-concordance-pie-legend, eval=has_transcriptomics, results='asis'}
figure_legend("Share of mapped gene-protein pairs whose RNA and protein log2 fold changes have the same sign (same direction) versus opposite signs. Left: all mapped pairs; right: fully-measured proteins only (a real intensity in every replicate of both conditions in the raw pg_matrix). Pairs with a zero fold change on either side are excluded. Sources: per-contrast log2FC (pathway_volcano_data_<contrast>.csv), mapping gene_protein_mapping.csv, observedness from proteomics files.protein. Descriptive, hypothesis-generating.")
```

## Rank Concordance

```{r rna-prot-rank-concordance, eval=has_transcriptomics, fig.width=11, fig.height=7.2}
# Directional concordance of RNA vs the protein log2FC ranking. Per contrast and
# direction, proteins are ranked by (simple) log2FC and we track the running % of
# included proteins whose paired RNA (simple) log2FC agrees beyond +/-0.5. Each
# point is tested against a random-set null (permutation, seeded) and the
# hypergeometric distribution at rank 250; the dashed horizontal line is the
# marginal (null) rate. Proteins are ranked by the proteomics limma moderated
# log2FC; the RNA side uses the group-mean simple log2FC. Proteomics has no
# simple log2FC column, so the two sides are not the same statistic.
rc_ok <- exists("rp_pairs") && !is.null(rp_pairs) && nrow(rp_pairs) >= 40 &&
    all(c("rna_simple_log2FC", "protein_log2FC", "contrast") %in% colnames(rp_pairs))
if (rc_ok) {
    up_thr <- 0.5; dn_thr <- -0.5; ranks <- c(100, 250); nperm <- 1000
    seed0  <- cfg$params$seed %||% 42
    lab_up <- "protein high->low: %RNA > 0.5"
    lab_dn <- "protein low->high: %RNA < -0.5"
    d0 <- rp_pairs[is.finite(rp_pairs$rna_simple_log2FC) & is.finite(rp_pairs$protein_log2FC), , drop = FALSE]
    set.seed(seed0)
    curves <- list(); pts <- list(); refs <- list()
    for (cn in unique(d0$contrast)) {
        dc <- d0[d0$contrast == cn, , drop = FALSE]; N <- nrow(dc)
        if (N < 40) next
        for (dir in c("up", "down")) {
            if (dir == "up") { ord <- order(-dc$protein_log2FC); hit <- dc$rna_simple_log2FC > up_thr; lab <- lab_up }
            else             { ord <- order(dc$protein_log2FC);  hit <- dc$rna_simple_log2FC < dn_thr; lab <- lab_dn }
            K <- sum(hit)
            curves[[paste(cn, dir)]] <- data.frame(contrast = cn, direction = lab,
                rank = seq_len(N), cum_pct = 100 * cumsum(hit[ord]) / seq_len(N), stringsAsFactors = FALSE)
            refs[[paste(cn, dir)]] <- data.frame(contrast = cn, direction = lab,
                base_pct = 100 * K / N, stringsAsFactors = FALSE)
            for (R in ranks[ranks <= N]) {
                obs <- sum(hit[ord[seq_len(R)]])
                nd  <- vapply(seq_len(nperm), function(i) sum(sample(hit, R)), integer(1))
                p_perm <- (1 + sum(nd >= obs)) / (1 + nperm)
                p_hyp  <- stats::phyper(obs - 1, K, N - K, R, lower.tail = FALSE)
                lbl <- if (R == max(ranks)) sprintf("r%d perm=%.2g/hyp=%.2g", R, p_perm, p_hyp)
                       else sprintf("r%d perm=%.2g", R, p_perm)
                pts[[paste(cn, dir, R)]] <- data.frame(contrast = cn, direction = lab, rank = R,
                    obs_pct = 100 * obs / R, label = lbl,
                    vj = if (R == min(ranks)) -0.8 else 1.6, stringsAsFactors = FALSE)
            }
        }
    }
    if (length(curves) > 0) {
        cur <- do.call(rbind, curves); pt <- do.call(rbind, pts); rf <- do.call(rbind, refs)
        cols <- setNames(c("#c0392b", "#2c6fbb"), c(lab_up, lab_dn))
        print(
            ggplot2::ggplot(cur, ggplot2::aes(rank, cum_pct, colour = direction)) +
                ggplot2::geom_vline(xintercept = ranks, colour = "grey60", linetype = "dashed", linewidth = 0.3) +
                ggplot2::geom_hline(data = rf, ggplot2::aes(yintercept = base_pct, colour = direction),
                                    linetype = "dashed", linewidth = 0.35) +
                ggplot2::geom_line(linewidth = 0.55) +
                ggplot2::geom_point(data = pt, ggplot2::aes(rank, obs_pct, colour = direction),
                                    inherit.aes = FALSE, size = 1.5) +
                ggplot2::geom_text(data = pt, ggplot2::aes(rank, obs_pct, colour = direction, label = label, vjust = vj),
                                   inherit.aes = FALSE, hjust = -0.06, size = 2.3, show.legend = FALSE) +
                ggplot2::facet_wrap(~ contrast, scales = "free_x") +
                ggplot2::scale_colour_manual(values = cols, name = NULL) +
                ggplot2::theme_bw(base_size = 10) + ggplot2::theme(legend.position = "top") +
                ggplot2::labs(x = "proteins included, ranked from the extreme (rank)",
                              y = "cumulative % RNA concordant beyond +/-0.5",
                              title = "RNA-protein rank concordance (permutation p at 100 & 250; hypergeometric at 250)")
        )
    } else cat("*Not enough mapped pairs per contrast for the rank-concordance analysis.*\n")
} else {
    cat("*Not enough mapped gene-protein pairs for the rank-concordance analysis.*\n")
}
```

```{r rna-prot-rank-concordance-legend, eval=has_transcriptomics, results='asis'}
figure_legend("Cumulative directional agreement as proteins are added from the extreme of the protein log2FC ranking. Red: proteins ranked most-up first, y = running % with RNA simple log2FC > 0.5; blue: most-down first, y = running % with RNA simple log2FC < -0.5. Vertical dashed lines mark ranks 100 and 250; the horizontal dashed line is the marginal (null) rate for that direction. Labels give the permutation p (1000 random equal-size sets, seeded) at ranks 100 and 250 and the hypergeometric p at 250. The two axes are not the same statistic: proteins are ranked by the proteomics limma moderated log2FC, while the RNA threshold uses the RNA group-mean simple log2FC (the naive log-ratio of group means), because the proteomics table carries no simple log2FC column. On runs whose RNA table predates that column the RNA side falls back to the DESeq2 shrunken log2FC. Both are read from pathway_volcano_data_<contrast>.csv and paired via gene_protein_mapping.csv. Hypothesis-generating.")
```

## Top Proteins by RNA Agreement

```{r rna-prot-top100-bars, eval=has_transcriptomics, fig.width=10.5, fig.height=11}
# Top-100 up- and down-regulated proteins per contrast, ranked by the proteomics
# limma moderated log2FC, coloured by whether the paired RNA group-mean simple
# log2FC agrees beyond the +/-0.5 threshold in that direction (up panel:
# RNA > 0.5; down panel: RNA < -0.5).
tb_ok <- exists("rp_pairs") && !is.null(rp_pairs) && nrow(rp_pairs) >= 40 &&
    all(c("rna_simple_log2FC", "protein_log2FC", "contrast") %in% colnames(rp_pairs))
if (tb_ok) {
    up_thr <- 0.5; dn_thr <- -0.5; TOPN <- 100
    pan_dn <- "top down-regulated (RNA < -0.5)"; pan_up <- "top up-regulated (RNA > 0.5)"
    d0 <- rp_pairs[is.finite(rp_pairs$rna_simple_log2FC) & is.finite(rp_pairs$protein_log2FC), , drop = FALSE]
    rows <- list()
    for (cn in unique(d0$contrast)) {
        dc <- d0[d0$contrast == cn, , drop = FALSE]; n <- min(TOPN, nrow(dc))
        if (n < 10) next
        dn <- dc[order(dc$protein_log2FC), ][seq_len(n), ]
        dn$panel <- pan_dn; dn$rank <- seq_len(n); dn$concordant <- dn$rna_simple_log2FC < dn_thr
        up <- dc[order(-dc$protein_log2FC), ][seq_len(n), ]
        up$panel <- pan_up; up$rank <- seq_len(n); up$concordant <- up$rna_simple_log2FC > up_thr
        rows[[cn]] <- rbind(dn, up)[, c("contrast", "panel", "rank", "protein_log2FC", "rna_simple_log2FC", "concordant")]
    }
    if (length(rows) > 0) {
        tb  <- do.call(rbind, rows)
        agg <- do.call(rbind, lapply(split(tb, list(tb$contrast, tb$panel), drop = TRUE), function(s)
            data.frame(contrast = s$contrast[1], panel = s$panel[1],
                       lab = sprintf("concordant %d/%d", sum(s$concordant), nrow(s)), stringsAsFactors = FALSE)))
        print(
            ggplot2::ggplot(tb, ggplot2::aes(rank, protein_log2FC, fill = concordant)) +
                ggplot2::geom_col(width = 1) +
                ggplot2::geom_hline(yintercept = 0, linewidth = 0.2, colour = "grey40") +
                ggplot2::geom_text(data = agg, ggplot2::aes(x = Inf, y = Inf, label = lab),
                                   inherit.aes = FALSE, hjust = 1.05, vjust = 1.3, size = 2.5) +
                ggplot2::facet_grid(contrast ~ panel, scales = "free_y") +
                ggplot2::scale_fill_manual(values = c("TRUE" = "#2166ac", "FALSE" = "grey75"),
                    labels = c("TRUE" = "RNA concordant (beyond +/-0.5)", "FALSE" = "not concordant"), name = NULL) +
                ggplot2::theme_bw(base_size = 9) + ggplot2::theme(legend.position = "top") +
                ggplot2::labs(x = "protein rank (1 = most extreme)", y = "protein log2FC (simple)",
                    title = "Top up/down proteins per contrast, coloured by RNA directional agreement")
        )
    } else cat("*Not enough pairs per contrast for the top-protein bar plot.*\n")
} else {
    cat("*Not enough mapped gene-protein pairs for the top-protein bar plot.*\n")
}
```

```{r rna-prot-top100-bars-legend, eval=has_transcriptomics, results='asis'}
figure_legend("Each bar is one of the top-100 most up- or down-regulated proteins per contrast, ranked by the proteomics limma moderated log2FC; bar height is that same protein log2FC. Blue marks proteins whose paired RNA agrees beyond the threshold in that direction (up panel: RNA simple log2FC > 0.5; down panel: RNA simple log2FC < -0.5), grey otherwise; the per-panel label gives the concordant count. The two sides are not the same statistic: the protein ranking uses the limma moderated log2FC, while the RNA threshold uses the RNA group-mean simple log2FC (the naive log-ratio of group means), because the proteomics table carries no simple log2FC column. Sources: pathway_volcano_data_<contrast>.csv for both fold changes, gene_protein_mapping.csv for the pairing. Descriptive, hypothesis-generating.")
```

---

# MultiGSEA Plots {.tabset}

```{r multigsea-note, eval=!has_multigsea, results='asis'}
cat("*No MultiGSEA plots found.*\n")
```

## All Contrasts (Combined) {.unnumbered}

```{r multigsea-plots, eval=has_multigsea, results='asis'}
# Exclude pairwise "omics_vs_omics" scatter plots; keep combined/summary plots only.
gsea_pngs <- list.files(multigsea_dir, pattern = "^multigsea_.*\\.png$", full.names = TRUE)
gsea_pngs <- gsea_pngs[!grepl("_vs_", basename(gsea_pngs))]
for (gp in gsea_pngs) {
    pair_name <- gsub("multigsea_|\\.png$", "", basename(gp))
    display_name <- gsub("_", " ", pair_name)
    cat(sprintf("\n### %s {.unnumbered}\n\n", display_name))
    cat(sprintf("![%s](%s)\n\n", display_name, gp))
    figure_legend(sprintf("MultiGSEA enrichment correlation plot for %s. Each point is a pathway; axis values represent enrichment scores. Color indicates joint significance status.", display_name))
    render_commentary(paste0("multigsea_", pair_name))
}
```

```{r multigsea-table, eval=has_multigsea}
# GSEA results as a table: GO term name + GO ID + per-omic -log10(padj) scores.
# Source = pairwise scatter CSVs written by 07b_multigsea_plots.R (cols term, label, x, y).
gsea_csvs <- list.files(multigsea_dir, pattern = "^multigsea_.*_vs_.*\\.csv$", full.names = TRUE)
gsea_tables <- list()
for (gc in gsea_csvs) {
    pair_name <- gsub("^multigsea_|\\.csv$", "", basename(gc))
    parts <- strsplit(pair_name, "_vs_")[[1]]
    om1 <- tools::toTitleCase(parts[1]); om2 <- tools::toTitleCase(parts[2])
    gsea_df <- read.csv(gc, stringsAsFactors = FALSE)
    if (nrow(gsea_df) == 0) next
    if (all(c("x", "y") %in% names(gsea_df)))
        gsea_df <- gsea_df[order(-(gsea_df$x + gsea_df$y)), ]
    display_cols <- intersect(c("label", "term", "x", "y"), colnames(gsea_df))
    tbl <- gsea_df[, display_cols, drop = FALSE]
    col_labels <- c(label = "GO Term", term = "GO ID",
                    x = paste0(om1, " -log10(padj)"), y = paste0(om2, " -log10(padj)"))
    colnames(tbl) <- col_labels[display_cols]
    gsea_tables[[length(gsea_tables) + 1]] <- make_dt(head(tbl, 200),
        caption = paste0("MultiGSEA enriched GO terms - ", gsub("_", " ", pair_name)),
        pageLength = 15)
}
# tagList renders every collected widget (a bare for-loop print() emits nothing).
htmltools::tagList(gsea_tables)
```

## Per-Contrast MultiGSEA {.tabset .unnumbered}

```{r multigsea-per-contrast, eval=has_multigsea, results='asis'}
multigsea_per_contrast_dir <- file.path(multigsea_dir, "per_contrast")
if (dir.exists(multigsea_per_contrast_dir)) {
    contrast_dirs <- list.dirs(multigsea_per_contrast_dir, full.names = TRUE, recursive = FALSE)
    for (cdir in contrast_dirs) {
        cname <- basename(cdir)
        cat(sprintf("\n### %s {.unnumbered}\n\n", gsub("_", " ", cname)))

        cpngs <- list.files(cdir, pattern = "^multigsea_.*\\.png$", full.names = TRUE)
        cpngs <- cpngs[!grepl("_vs_", basename(cpngs))]
        if (length(cpngs) == 0) {
            cat("*No MultiGSEA plots for this contrast.*\n\n")
            next
        }
        for (cp in cpngs) {
            pair_name <- gsub("multigsea_|\\.png$", "", basename(cp))
            display_name <- gsub("_", " ", pair_name)
            chunk_id <- paste0("mg-pc-", gsub("[^a-zA-Z0-9]", "", cname), "-",
                               gsub("[^a-zA-Z0-9]", "", pair_name))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, cp
            ), quiet = TRUE))
            cat("\n\n")
        }
    }
} else {
    cat("*Per-contrast MultiGSEA results not available.*\n")
}
```

```{r multi-ora-setup, include=FALSE}
multi_ora_dir <- file.path(multigsea_dir, "multi_ora")
has_multi_ora <- dir.exists(multi_ora_dir) && length(list.files(multi_ora_dir, "\\.png$")) > 0
```

## Multi-ORA (Combined Cross-Omics ORA) {.tabset .unnumbered}

```{r multi-ora-note, eval=!has_multi_ora, results='asis'}
ora_issues_file <- file.path(multi_ora_dir, "multi_ora_issues.txt")
if (file.exists(ora_issues_file)) {
    issues <- readLines(ora_issues_file)
    cat("**Multi-ORA did not complete.** The following issues were encountered:\n\n")
    for (iss in issues) {
        cat(sprintf("- %s\n", iss))
    }
    cat("\n*This is often caused by KEGG REST API timeouts or network issues. Re-running the pipeline may resolve it.*\n\n")
} else {
    cat("*No jointly-significant pathways from the combined cross-omics ORA.* This test pools the significant RNA and protein features into a single over-representation test against the combined gene sets; for this run no pathway reached significance in the pooled test. This does not mean either omic lacks enrichment — the per-omic ORAs can still be enriched (see the per-omic enrichment tables, and the KEGG pathview maps where available). An empty combined result means the RNA and protein hits did not concentrate in the *same* pathways strongly enough to pass the pooled test.\n")
}
```

### Per-Contrast Multi-ORA {.tabset .unnumbered}

```{r multi-ora-per-contrast, eval=has_multi_ora, results='asis'}
mora_per_contrast_dir <- file.path(multi_ora_dir, "per_contrast")
if (dir.exists(mora_per_contrast_dir)) {
    mora_contrast_dirs <- list.dirs(mora_per_contrast_dir, full.names = TRUE, recursive = FALSE)
    for (cdir in mora_contrast_dirs) {
        cname <- basename(cdir)
        cat(sprintf("\n#### %s {.unnumbered}\n\n", gsub("_", " ", cname)))

        # Barplot
        c_bp <- file.path(cdir, "multi_ora_pooled_barplot.png")
        if (file.exists(c_bp)) {
            chunk_id <- paste0("mora-pc-bp-", gsub("[^a-zA-Z0-9]", "", cname))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, c_bp
            ), quiet = TRUE))
            cat("\n\n")
        }

        # Results table
        c_csv <- file.path(cdir, "multi_ora_results.csv")
        if (file.exists(c_csv)) {
            c_df <- read.csv(c_csv, stringsAsFactors = FALSE)
            c_show <- c_df[seq_len(min(20, nrow(c_df))), ]
            display_cols <- intersect(c("pathway", "pooled_padj", "pooled_count", "metabolomics_padj", "n_omics_support"), colnames(c_show))
            if (length(display_cols) > 0) {
                c_show <- c_show[, display_cols, drop = FALSE]
                cat(sprintf("\n**%s:** %d enriched pathways\n\n",
                            gsub("_", " ", cname), nrow(c_df)))
            }
        } else {
            cat("*No enriched pathways for this contrast.*\n\n")
        }
    }
} else {
    cat("*Per-contrast Multi-ORA results not available.*\n")
}
```

```{r metab-ora-setup, include=FALSE}
metab_ora_csv <- file.path(multi_ora_dir, "compound_ora_results.csv")
has_metab_ora <- has_multi_ora && file.exists(metab_ora_csv)
```

## Metabolomics Compound Enrichment {.unnumbered}

```{r metab-ora-note, eval=!has_metab_ora, results='asis'}
cat("*No metabolomics compound ORA results found. Compound enrichment requires metabolomics DE results with KEGG compound ID mappings.*\n")
```

```{r metab-ora-results, eval=has_metab_ora, results='asis'}
metab_df <- read.csv(metab_ora_csv, stringsAsFactors = FALSE)
if (nrow(metab_df) > 0) {
    cat("\nCompound-based over-representation analysis identified **", nrow(metab_df),
        "** enriched KEGG pathways from significantly altered metabolites.\n\n")
}
```

```{r metab-ora-barplot, eval=has_metab_ora, results='asis'}
metab_bp <- file.path(multi_ora_dir, "compound_ora_barplot.png")
if (file.exists(metab_bp)) {
    cat(sprintf("![Compound ORA](%s)\n\n", metab_bp))
    figure_legend("Metabolomics compound ORA barplot. Significant metabolites (KEGG compound IDs) are tested for over-representation in KEGG metabolic pathways using a hypergeometric test. Red dashed line marks p = 0.05.")
}
```

```{r metab-ora-table, eval=has_metab_ora}
metab_df <- read.csv(metab_ora_csv, stringsAsFactors = FALSE)
display_cols <- intersect(c("pathway", "ID", "pvalue", "padj", "setSize", "compounds"), colnames(metab_df))
metab_show <- metab_df[seq_len(min(30, nrow(metab_df))), display_cols, drop = FALSE]
col_labels <- c(pathway = "Pathway", ID = "KEGG ID", pvalue = "P-value",
                padj = "Adj. P-value", setSize = "Compound Count", compounds = "Compounds")
colnames(metab_show) <- col_labels[display_cols]
knitr::kable(metab_show, row.names = FALSE, caption = "Metabolomics compound ORA - enriched KEGG pathways")
```

```{r metab-ora-commentary, eval=has_metab_ora, results='asis'}
render_commentary("compound_ora")
```

```{r pathview-setup, include=FALSE}
pathview_pdf <- file.path(multi_ora_dir, "multi_ora_pathview_supported.pdf")
pathview_dir <- file.path(multi_ora_dir, "pathview")
has_pathview <- file.exists(pathview_pdf)
```

## Multi-Omics Pathway Maps (Pathview) {.unnumbered}

```{r pathview-note, eval=!has_pathview, results='asis'}
cat("*No pathview maps generated. Requires pathways enriched in >= 2 omics layers.*\n")
```

```{r pathview-plots, eval=has_pathview, results='asis'}
pv_pngs <- sort(list.files(pathview_dir, pattern = "\\.multi_ora.*\\.(multi\\.)?png$", full.names = TRUE))
cat("KEGG pathway maps for pathways enriched in >= 2 omics layers, shown **per contrast**. ",
    "Gene nodes are colored by logFC (transcriptomics and/or proteomics); ",
    "compound nodes are colored by metabolomics logFC.\n\n")
cat(sprintf("[Download compiled PDF](%s)\n\n", pathview_pdf))

# Group PNGs by contrast (extracted from filename suffix)
for (pv_png in pv_pngs) {
    bn <- basename(pv_png)
    # Extract pathway ID and contrast from filename
    # e.g. "cel00053.multi_ora_X1.56ppm.vs..0ppm.multi.png"
    pw_id <- sub("^[a-z]+", "", sub("\\.multi_ora.*$", "", bn))
    contrast_part <- sub("^.*\\.multi_ora_?", "", bn)
    contrast_part <- sub("\\.(multi\\.)?png$", "", contrast_part)
    contrast_label <- gsub("\\.", " ", contrast_part)
    if (nzchar(contrast_label)) {
        label <- paste0("Pathway ", pw_id, " (", contrast_label, ")")
    } else {
        label <- paste0("Pathway ", pw_id)
    }
    cat(sprintf("\n### %s {.unnumbered}\n\n", label))
    cat(sprintf("![%s](%s)\n\n", label, pv_png))
}
figure_legend("KEGG pathway maps generated with pathview, shown per contrast. Gene nodes show logFC from transcriptomics (left) and proteomics (right) where available. Compound nodes show metabolomics logFC. Red = upregulated, green/blue = downregulated.")
render_commentary("multi_ora_pathview")
```

```{r per-omics-pathview-setup, include=FALSE}
pv_metab_pdf <- file.path(multi_ora_dir, "pathview_top_metabolomics_pathways.pdf")
pv_prot_pdf <- file.path(multi_ora_dir, "pathview_top_proteomics_pathways.pdf")
has_pv_metab <- file.exists(pv_metab_pdf)
has_pv_prot <- file.exists(pv_prot_pdf)
has_per_omics_pv <- has_pv_metab || has_pv_prot
```

```{r per-omics-pathview-metab, eval=has_pv_metab, results='asis'}
cat("\n## Top Metabolomics Pathways (with Proteomics Overlay) {.unnumbered}\n\n")
cat("Top enriched metabolomics KEGG pathways with compound logFC (metabolomics) and ",
    "enzyme/gene logFC (proteomics) overlaid on the same pathway map.\n\n")
cat(sprintf("[Download PDF](%s)\n\n", pv_metab_pdf))
metab_pngs <- sort(list.files(pathview_dir, pattern = "\\.metab_top\\.png$", full.names = TRUE))
for (pv_png in metab_pngs) {
    pw_id <- sub("^[a-z]+", "", sub("\\.metab_top\\.png$", "", basename(pv_png)))
    cat(sprintf("\n### Pathway %s {.unnumbered}\n\n", pw_id))
    cat(sprintf("![Pathway %s](%s)\n\n", pw_id, pv_png))
}
figure_legend("Top metabolomics-enriched KEGG pathways. Compound nodes show metabolomics logFC; gene/enzyme nodes show proteomics logFC. Red = upregulated, blue/green = downregulated.")
```

```{r per-omics-pathview-prot, eval=has_pv_prot, results='asis'}
cat("\n## Top Proteomics Pathways (with Metabolomics Overlay) {.unnumbered}\n\n")
cat("Top enriched proteomics KEGG pathways with enzyme/gene logFC (proteomics) and ",
    "compound logFC (metabolomics) overlaid on the same pathway map.\n\n")
cat(sprintf("[Download PDF](%s)\n\n", pv_prot_pdf))
prot_pngs <- sort(list.files(pathview_dir, pattern = "\\.prot_top\\.png$", full.names = TRUE))
for (pv_png in prot_pngs) {
    pw_id <- sub("^[a-z]+", "", sub("\\.prot_top\\.png$", "", basename(pv_png)))
    cat(sprintf("\n### Pathway %s {.unnumbered}\n\n", pw_id))
    cat(sprintf("![Pathway %s](%s)\n\n", pw_id, pv_png))
}
figure_legend("Top proteomics-enriched KEGG pathways. Gene/enzyme nodes show proteomics logFC; compound nodes show metabolomics logFC. Red = upregulated, blue/green = downregulated.")
```

```{r pathview-all-pdf, eval=has_pathview || has_pv_metab || has_pv_prot, results='asis'}
# Collect ALL pathview PNGs from all sections into one combined PDF
all_pv_pngs <- character(0)
if (dir.exists(pathview_dir)) {
    all_pv_pngs <- sort(list.files(pathview_dir, pattern = "\\.png$", full.names = TRUE))
}

if (length(all_pv_pngs) > 0) {
    combined_pdf <- file.path(multi_ora_dir, "all_pathview_combined.pdf")
    compile_ok <- tryCatch({
        grDevices::pdf(combined_pdf, width = 12, height = 8)
        for (png_file in all_pv_pngs) {
            img <- png::readPNG(png_file)
            grid::grid.newpage()
            # Add pathway ID as title at top
            pw_label <- sub("\\..*$", "", basename(png_file))
            grid::grid.text(pw_label, x = 0.5, y = 0.98,
                            gp = grid::gpar(fontsize = 10, col = "grey40"))
            grid::grid.raster(img, width = 0.95, height = 0.92, y = 0.48)
        }
        grDevices::dev.off()
        TRUE
    }, error = function(e) {
        tryCatch(grDevices::dev.off(), error = function(e2) NULL)
        FALSE
    })

    if (compile_ok && file.exists(combined_pdf)) {
        cat(sprintf(
            '\n<div style="text-align: center; margin: 16px 0;">\n<a href="%s" download style="display: inline-block; padding: 10px 24px; background-color: #4a90d9; color: white; text-decoration: none; border-radius: 6px; font-weight: 600; font-size: 1em;">Download All Pathview Plots (PDF, %d figures)</a>\n</div>\n\n',
            combined_pdf, length(all_pv_pngs)
        ))
    }
}
```

---

# Cross-Omics Enrichment {.tabset}

```{r enrichment-note, eval=!has_enrichment, results='asis'}
cat("*No cross-omics enrichment results found. Enrichment analysis requires pathway results from at least 1 omics layer.*\n")
```

## All Contrasts (Combined) {.unnumbered}

```{r enrichment-per-omics, eval=has_enrichment, results='asis'}
# Show per-omics enrichment barplots
omics_barplots <- list.files(enrichment_dir, pattern = "_top_pathways\\.png$", full.names = TRUE)
if (length(omics_barplots) > 0) {
    cat("\n### Per-Omics Pathway Enrichment {.unnumbered}\n\n")
    for (i in seq_along(omics_barplots)) {
        bp <- omics_barplots[i]
        omics_name <- gsub("_top_pathways\\.png$", "", basename(bp))
        cat(sprintf("\n#### %s {.unnumbered}\n\n", tools::toTitleCase(omics_name)))
        chunk_id <- paste0("enrich-bp-", i)
        cat(knitr::knit_child(text = sprintf(
            '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            chunk_id, bp
        ), quiet = TRUE))
        cat("\n\n")
    }
}
```

```{r enrichment-per-omics-legend, eval=has_enrichment, results='asis'}
figure_legend("Per-omics KEGG pathway enrichment (all contrasts combined). Bar length represents -log10(p-value). Red dashed line indicates p = 0.05 significance threshold.")
```

```{r enrichment-per-omics-table, eval=has_enrichment}
# Show enrichment tables for each omics
csv_files <- list.files(enrichment_dir, pattern = "_enriched_pathways\\.csv$", full.names = TRUE)
for (cf in csv_files) {
    omics_name <- gsub("_enriched_pathways\\.csv$", "", basename(cf))
    tbl <- read.csv(cf, stringsAsFactors = FALSE)
    if (nrow(tbl) > 0) {
        print(make_dt(
            head(tbl, 200),
            caption = paste0(tools::toTitleCase(omics_name), " Enriched KEGG Pathways"),
            pageLength = 15
        ))
    }
}
```

```{r enrichment-heatmap, eval=has_enrichment, results='asis'}
p <- file.path(enrichment_dir, "cross_omics_pathway_heatmap.png")
if (file.exists(p)) {
    cat("\n### Cross-Omics Pathway Comparison {.unnumbered}\n\n")
    cat(knitr::knit_child(text = sprintf(
        '```{r enrich-heatmap-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        p
    ), quiet = TRUE))
}
```

```{r enrichment-heatmap-legend, eval=has_enrichment, results='asis'}
p <- file.path(enrichment_dir, "cross_omics_pathway_heatmap.png")
if (file.exists(p)) {
    figure_legend("Cross-omics pathway enrichment heatmap (all contrasts combined). Showing the top 30 pathways ordered by Stouffer combined p-value; the complete list is in the table below. Each cell is the fgsea normalized enrichment score (NES) for that pathway in that omics layer, with the sign giving the direction of enrichment: red = up (positive NES), blue = down (negative NES), white ~ 0; the number in each cell is the NES. Note: for each pathway the per-omics value is taken from that layer's enriched-pathways table, so pathways with consistent but individually sub-significant signals across multiple contrasts may appear here but not in the per-contrast views.")
    render_commentary("cross_omics_pathway_heatmap")
}
```

```{r enrichment-table, eval=has_enrichment}
meta_file <- file.path(enrichment_dir, "cross_omics_pathways_meta_analysis.csv")
if (file.exists(meta_file)) {
    meta_tbl <- read.csv(meta_file, stringsAsFactors = FALSE)
    make_dt(
        head(meta_tbl, 200),
        caption = "Cross-Omics Pathway Meta-Analysis (Fisher combined p-values)",
        pageLength = 15
    )
}
```

```{r enrichment-table-legend, eval=has_enrichment, results='asis'}
meta_file <- file.path(enrichment_dir, "cross_omics_pathways_meta_analysis.csv")
if (file.exists(meta_file)) {
    figure_legend("Top pathways from cross-omics meta-analysis. P-values from individual omics layers were combined using Fisher's method. Fisher's method combines evidence regardless of the direction of change, so a pathway can reach joint significance while moving in opposite directions in the two layers. The section below separates these cases.")
    render_commentary("cross_omics_meta_analysis")
}
```

## Jointly Significant Pathways and Their Direction {.unnumbered}

```{r enrichment-concordant, eval=has_enrichment, results='asis'}
meta_file <- file.path(enrichment_dir, "cross_omics_pathways_meta_analysis.csv")
alpha_joint <- 0.05

if (file.exists(meta_file)) {
    meta_tbl <- read.csv(meta_file, stringsAsFactors = FALSE)
    pcols <- grep("^pval_", names(meta_tbl), value = TRUE)
    layer_names <- sub("^pval_", "", pcols)

    if (length(pcols) >= 2 && "combined_padj" %in% names(meta_tbl)) {
        # Jointly significant = nominally significant in every layer AND
        # surviving FDR on the Fisher-combined p.
        keep <- rep(TRUE, nrow(meta_tbl))
        for (pc in pcols) keep <- keep & !is.na(meta_tbl[[pc]]) & meta_tbl[[pc]] < alpha_joint
        keep <- keep & !is.na(meta_tbl$combined_padj) & meta_tbl$combined_padj < alpha_joint
        joint <- meta_tbl[keep, , drop = FALSE]
        joint <- joint[order(joint$combined_padj), , drop = FALSE]

        # Attach the per-layer NES so direction can be compared across layers.
        # Guard on nrow: when no pathway is jointly significant, joint has 0 rows
        # and assigning a scalar column errors ("replacement has 1 row, data has 0").
        nes_cols <- paste0("NES_", layer_names)
        if (nrow(joint) > 0) {
            for (ln in layer_names) {
                nes_file <- file.path(enrichment_dir, sprintf("%s_enriched_pathways.csv", ln))
                joint[[paste0("NES_", ln)]] <- NA_real_
                if (file.exists(nes_file)) {
                    nes_tbl <- tryCatch(read.csv(nes_file, stringsAsFactors = FALSE),
                                        error = function(e) NULL)
                    if (!is.null(nes_tbl) && all(c("pathway_name", "NES") %in% names(nes_tbl))) {
                        nes_tbl <- nes_tbl[!duplicated(nes_tbl$pathway_name), , drop = FALSE]
                        idx <- match(joint$pathway_name, nes_tbl$pathway_name)
                        joint[[paste0("NES_", ln)]] <- suppressWarnings(as.numeric(nes_tbl$NES[idx]))
                    }
                }
            }

            nes_mat <- as.matrix(joint[, nes_cols, drop = FALSE])
            joint$direction <- apply(nes_mat, 1, function(v) {
                if (any(is.na(v))) return("undetermined")
                if (all(v > 0)) "same (up in all layers)"
                else if (all(v < 0)) "same (down in all layers)"
                else "opposite"
            })
        }

        n_joint <- nrow(joint)
        n_same  <- sum(grepl("^same", joint$direction))
        n_opp   <- sum(joint$direction == "opposite")

        if (n_joint > 0) {
            cat(sprintf(
                paste0("Of the %d pathways in the cross-omics meta-analysis, %d are nominally ",
                       "significant (p < %.2f) in every omics layer and also pass FDR on the ",
                       "Fisher-combined p. Of those %d, %d change in the same direction in all ",
                       "layers and %d change in opposite directions.\n\n"),
                nrow(meta_tbl), n_joint, alpha_joint, n_joint, n_same, n_opp))

            if (n_opp > 0) {
                cat(paste0("The opposite-direction pathways are reported here rather than dropped. ",
                           "They are jointly significant but the transcript and protein evidence ",
                           "disagrees on sign, so they do not support a single coherent statement ",
                           "about the pathway.\n\n"))
            }

            show_cols <- intersect(
                c("pathway_name", pcols, "combined_pval", "combined_padj", nes_cols, "direction"),
                names(joint))
            make_dt(joint[, show_cols, drop = FALSE],
                    caption = "Pathways jointly significant across omics layers, annotated by direction",
                    pageLength = 15)

            # Persist so the numbers quoted above are checkable off-report.
            out_tsv <- file.path(enrichment_dir, "cross_omics_jointly_significant_pathways.tsv")
            tryCatch(
                write.table(joint[, show_cols, drop = FALSE], out_tsv,
                            sep = "\t", row.names = FALSE, quote = FALSE),
                error = function(e) invisible(NULL))
        } else {
            cat(sprintf(paste0("No pathway reached nominal significance (p < %.2f) in every ",
                               "omics layer while also passing FDR on the Fisher-combined p.\n\n"),
                        alpha_joint))
        }
    }
}
```

```{r enrichment-concordant-legend, eval=has_enrichment, results='asis'}
if (file.exists(file.path(enrichment_dir, "cross_omics_pathways_meta_analysis.csv"))) {
    figure_legend(paste0(
        "Pathways significant in every omics layer at nominal p < 0.05 and passing FDR on the ",
        "Fisher-combined p. Direction is taken from the sign of the per-layer NES in ",
        "<omics>_enriched_pathways.csv; 'undetermined' means the pathway had no NES in at least ",
        "one layer (for example an ORA-only result). Sources: ",
        "cross_omics_pathways_meta_analysis.csv for the p-values, ",
        "<omics>_enriched_pathways.csv for NES, both in the cross_enrichment directory. ",
        "The full annotated list is written to cross_omics_jointly_significant_pathways.tsv. ",
        "Direction agreement is descriptive and is not itself tested; with few replicates and a ",
        "single contrast these results are hypothesis-generating."))
}
```

## Per-Contrast Enrichment {.tabset .unnumbered}

```{r enrichment-per-contrast, eval=has_enrichment, results='asis'}
enrich_per_contrast_dir <- file.path(enrichment_dir, "per_contrast")
if (dir.exists(enrich_per_contrast_dir)) {
    enrich_contrast_dirs <- list.dirs(enrich_per_contrast_dir, full.names = TRUE, recursive = FALSE)
    for (cdir in enrich_contrast_dirs) {
        cname <- basename(cdir)
        cat(sprintf("\n### %s {.unnumbered}\n\n", gsub("_", " ", cname)))

        # Per-omics barplots for this contrast
        c_barplots <- list.files(cdir, pattern = "_top_pathways\\.png$", full.names = TRUE)
        for (cbp in c_barplots) {
            omics_name <- gsub("_top_pathways\\.png$", "", basename(cbp))
            cat(sprintf("\n#### %s {.unnumbered}\n\n", tools::toTitleCase(omics_name)))
            chunk_id <- paste0("enr-pc-", gsub("[^a-zA-Z0-9]", "", cname), "-",
                               gsub("[^a-zA-Z0-9]", "", omics_name))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, cbp
            ), quiet = TRUE))
            cat("\n\n")
        }

        # Cross-omics heatmap for this contrast
        c_heatmap <- file.path(cdir, "cross_omics_pathway_heatmap.png")
        if (file.exists(c_heatmap)) {
            cat("\n#### Cross-Omics Heatmap {.unnumbered}\n\n")
            chunk_id <- paste0("enr-pc-hm-", gsub("[^a-zA-Z0-9]", "", cname))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, c_heatmap
            ), quiet = TRUE))
            cat("\n\n")
        }

        # Meta-analysis table for this contrast
        c_meta <- file.path(cdir, "cross_omics_pathways_meta_analysis.csv")
        if (file.exists(c_meta)) {
            c_meta_df <- read.csv(c_meta, stringsAsFactors = FALSE)
            cat(sprintf("\n**%s:** %d pathways in cross-omics meta-analysis\n\n",
                        gsub("_", " ", cname), nrow(c_meta_df)))
        }

        if (length(c_barplots) == 0 && !file.exists(c_heatmap)) {
            cat("*No enrichment results for this contrast.*\n\n")
        }
    }
} else {
    cat("*Per-contrast enrichment results not available.*\n")
}
```

---

# Data Harmonization

```{r harmonization-summary, results='asis'}
# Look for harmonization summary files
harmony_summary_file <- file.path(run_dir, "harmonization_summary.csv")
if (file.exists(harmony_summary_file)) {
    harmony_summary <- read.csv(harmony_summary_file, stringsAsFactors = FALSE)
    cat("## Harmonization Overview\n\n")
    kable(harmony_summary)
} else {
    cat("Multi-omics data were harmonized into a unified **MultiAssayExperiment** object. ",
        "Each omics layer was aligned to a common set of samples.\n\n")

    # Show omics layers
    omics_dirs <- list.dirs(integration_dir, full.names = FALSE, recursive = FALSE)
    if (length(omics_dirs) > 0) {
        cat("**Omics layers in integration:** ", paste(omics_dirs, collapse = ", "), "\n\n")
    }
}
```

```{r gene-protein-mapping, eval=!is.null(gene_protein_mapping), results='asis'}
cat("## Gene-Protein Mapping\n\n")
cat(sprintf("A total of **%d gene-protein pairs** were mapped across transcriptomics and proteomics layers.\n\n",
            nrow(gene_protein_mapping)))
```

```{r gene-protein-table, eval=!is.null(gene_protein_mapping) && nrow(gene_protein_mapping) > 0}
make_dt(
    head(gene_protein_mapping, 500),
    caption = "Gene-Protein Mapping (first 500 rows)",
    pageLength = 10
)
```

```{r harmonization-legend, eval=!is.null(gene_protein_mapping), results='asis'}
figure_legend("Gene-protein mapping table showing matched features between transcriptomics and proteomics layers.")
render_commentary("gene_protein_mapping")
```

---

# Integration Results {.tabset}

```{r integration-note, eval=!has_integration, results='asis'}
cat("*No integration results found. Integration methods may not have been configured or may have failed.*\n")
```

## DIABLO

```{r diablo-sample-plot, eval=has_diablo, results='asis'}
p <- file.path(diablo_dir, "diablo_sample_plot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r diablo-sample-legend, eval=has_diablo, results='asis'}
figure_legend("DIABLO sample scores plot. Samples are projected onto the first two components of the multi-omics discriminant analysis. Colors represent experimental groups.")
render_commentary("diablo_sample_plot")
```

```{r diablo-circos, eval=has_diablo, results='asis'}
circos_path <- file.path(diablo_dir, "diablo_circos.png")
if (file.exists(circos_path)) {
    cat("\n### Cross-Omics Circos Plot {.unnumbered}\n\n")
    knitr::include_graphics(circos_path)
}
```

```{r diablo-circos-legend, eval=has_diablo, results='asis'}
if (file.exists(file.path(diablo_dir, "diablo_circos.png"))) {
    figure_legend("DIABLO circos plot showing cross-omics feature correlations. Sectors represent omics layers; bar heights show DIABLO loading magnitude (component 1). Chords connect feature pairs with absolute Pearson correlation above the cutoff. Red chords indicate positive correlation; blue chords indicate negative correlation. Chord intensity and width scale with correlation strength.")
    render_commentary("diablo_circos")
}
```

```{r diablo-feature-heatmap, eval=has_diablo, results='asis'}
hm <- file.path(diablo_dir, "diablo_feature_heatmap.png")
hm_latent <- file.path(diablo_dir, "diablo_feature_heatmap_latent.png")
has_latent_hm <- file.exists(hm_latent)

if (file.exists(hm) && has_latent_hm) {
    cat("\n### Multi-Omics Feature Heatmap {.tabset .unnumbered}\n\n")
    cat("#### Expression Distance {.unnumbered}\n\n")
    cat(sprintf("![DIABLO Feature Heatmap](%s)\n\n", hm))
    figure_legend("Heatmap of top DIABLO features across omics layers. Values are z-scored per feature. Rows are clustered by Euclidean distance on expression values. The color bar on the left indicates the omics source layer.")

    cat("\n#### Latent Space Distance {.unnumbered}\n\n")
    cat(sprintf("![DIABLO Latent Space Heatmap](%s)\n\n", hm_latent))
    figure_legend("Same features as above, but rows are clustered by distance in the DIABLO latent component space (loading values). Features from different omics that contribute similarly to the discriminant components cluster together, highlighting cross-omics integration patterns.")
    render_commentary("diablo_feature_heatmap")
} else if (file.exists(hm)) {
    cat("\n### Multi-Omics Feature Heatmap {.unnumbered}\n\n")
    cat(sprintf("![DIABLO Feature Heatmap](%s)\n\n", hm))
    figure_legend("Heatmap of top DIABLO features across omics layers. Values are z-scored per feature. Rows are clustered by Euclidean distance. The color bar on the left indicates the omics source layer.")
    render_commentary("diablo_feature_heatmap")
}
```

```{r diablo-top-features, eval=has_diablo}
# Load top feature tables
diablo_feat_files <- list.files(diablo_dir, pattern = "diablo_top_features_.*\\.csv$", full.names = TRUE)
if (length(diablo_feat_files) > 0) {
    for (ff in diablo_feat_files) {
        omics_name <- gsub("diablo_top_features_|\\.csv$", "", basename(ff))
        df <- tryCatch(read.csv(ff, stringsAsFactors = FALSE), error = function(e) NULL)
        if (!is.null(df) && nrow(df) > 0) {
            cat(sprintf("\n### Top DIABLO Features: %s\n\n", omics_name))
            print(make_dt(df, caption = paste("DIABLO top features -", omics_name)))
        }
    }
}
```

## SNF

```{r snf-heatmap, eval=has_snf, results='asis'}
p <- file.path(snf_dir, "snf_network_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r snf-heatmap-legend, eval=has_snf, results='asis'}
figure_legend("SNF fused similarity network heatmap. Rows and columns are samples; color intensity represents similarity across all omics layers after network fusion.")
render_commentary("snf_network_heatmap")
```

```{r snf-silhouette, eval=has_snf, results='asis'}
p <- file.path(snf_dir, "snf_silhouette.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r snf-silhouette-legend, eval=has_snf, results='asis'}
figure_legend("SNF cluster silhouette plot. Silhouette width measures how well each sample matches its assigned cluster compared to the nearest alternative cluster. **Interpretation:** values close to **+1** indicate the sample is well-matched to its cluster and poorly matched to neighboring clusters (strong clustering). Values near **0** mean the sample sits on the boundary between two clusters. **Negative** values indicate the sample may be misclassified and would fit better in the neighboring cluster. The average silhouette width across all samples summarizes overall clustering quality: > 0.7 = strong structure, 0.5--0.7 = reasonable structure, 0.25--0.5 = weak structure, < 0.25 = no substantial structure. This metric can guide the choice of the number of clusters (K) and help identify outlier samples that do not conform to any group.")
render_commentary("snf_silhouette")
```

```{r snf-clusters-table, eval=has_snf}
clusters_file <- file.path(snf_dir, "snf_clusters.csv")
if (file.exists(clusters_file)) {
    clusters <- read.csv(clusters_file, stringsAsFactors = FALSE)
    make_dt(clusters, caption = "SNF Cluster Assignments")
}
```

## MOFA2

```{r mofa-variance, eval=has_mofa, results='asis'}
p <- file.path(mofa_dir, "mofa_variance_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mofa-variance-legend, eval=has_mofa, results='asis'}
figure_legend("MOFA2 variance explained heatmap. Each cell shows the proportion of variance in a given omics view (rows) explained by a latent factor (columns). Factors capturing shared variation across multiple views are most informative.")
render_commentary("mofa_variance_heatmap")
```

```{r mofa-factors-tabs, eval=has_mofa, results='asis'}
p12 <- file.path(mofa_dir, "mofa_factors_1_2.png")
p12_labeled <- file.path(mofa_dir, "mofa_factors_1_2_labeled.png")
p13 <- file.path(mofa_dir, "mofa_factors_1_3.png")

has_f12 <- file.exists(p12)
has_f12_lab <- file.exists(p12_labeled)
has_f13 <- file.exists(p13)

if (has_f12 || has_f12_lab) {
    cat("\n### Factor 1 vs Factor 2 {.tabset .unnumbered}\n\n")

    if (has_f12) {
        cat("#### By Condition {.unnumbered}\n\n")
        cat(knitr::knit_child(text = sprintf(
            '```{r mofa-f12-cond, echo=FALSE}\nknitr::include_graphics("%s")\n```\n', p12
        ), quiet = TRUE))
        figure_legend("MOFA2 Factor 1 vs Factor 2 scatter plot. Each point is a sample, colored by experimental condition.")
    }

    if (has_f12_lab) {
        cat("\n#### With Sample Labels {.unnumbered}\n\n")
        cat(knitr::knit_child(text = sprintf(
            '```{r mofa-f12-lab, echo=FALSE}\nknitr::include_graphics("%s")\n```\n', p12_labeled
        ), quiet = TRUE))
        figure_legend("MOFA2 Factor 1 vs Factor 2 with sample labels. Separation along factors indicates biologically relevant latent dimensions.")
    }

    render_commentary("mofa_factors_1_2")
}

if (has_f13) {
    cat("\n### Factor 1 vs Factor 3 {.unnumbered}\n\n")
    cat(knitr::knit_child(text = sprintf(
        '```{r mofa-f13, echo=FALSE}\nknitr::include_graphics("%s")\n```\n', p13
    ), quiet = TRUE))
    figure_legend("MOFA2 Factor 1 vs Factor 3 scatter plot.")
    render_commentary("mofa_factors_1_3")
}
```

```{r mofa-weights, eval=has_mofa, results='asis'}
weight_files <- list.files(mofa_dir, pattern = "^mofa_top_weights_.*\\.png$", full.names = TRUE)
# Separate the gene-symbol proteomics plot from the rest
genesym_file <- grep("proteomics_genesymbol", weight_files, value = TRUE)
main_files <- setdiff(weight_files, genesym_file)

if (length(main_files) > 0) {
    for (wf in main_files) {
        view_name <- gsub("mofa_top_weights_|\\.png$", "", basename(wf))
        chunk_id <- paste0("mofa-wt-", gsub("[^a-zA-Z0-9]", "", view_name))

        # For proteomics: create a tabset with UniProt and Gene Symbol tabs
        if (view_name == "proteomics" && length(genesym_file) == 1) {
            cat(sprintf("\n### Top Weights: %s {.tabset .unnumbered}\n\n", view_name))
            cat("#### UniProt ID {.unnumbered}\n\n")
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, wf
            ), quiet = TRUE))
            figure_legend(sprintf("Top 20 features by absolute weight for MOFA Factor in the %s view (UniProt IDs).", view_name))

            cat("\n#### Gene Symbol {.unnumbered}\n\n")
            gs_id <- paste0(chunk_id, "gs")
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                gs_id, genesym_file
            ), quiet = TRUE))
            figure_legend(sprintf("Top 20 features by absolute weight for MOFA Factor in the %s view (Gene Symbols).", view_name))
        } else {
            cat(sprintf("\n### Top Weights: %s\n\n", view_name))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, wf
            ), quiet = TRUE))
            figure_legend(sprintf("Top 20 features by absolute weight for MOFA Factor in the %s view.", view_name))
        }

        fig_id <- paste0("mofa_top_weights_", view_name)
        render_commentary(fig_id)
    }
}
```

## Loadings-Based Pathway Enrichment {.tabset}

```{r loadings-enrich-note, eval=!has_loadings_enrich, results='asis'}
cat("*No loadings-based enrichment results available. DIABLO/MOFA2 may not have been configured or enrichment may have failed.*\n")
```

```{r loadings-enrich-diablo, eval=has_loadings_enrich, results='asis'}
diablo_le_dir <- file.path(loadings_enrich_dir, "diablo_loadings")
diablo_le_plots <- list.files(diablo_le_dir, pattern = "\\.png$", full.names = TRUE)

if (length(diablo_le_plots) > 0) {
    cat("\n### DIABLO Loadings Enrichment\n\n")
    for (i in seq_along(diablo_le_plots)) {
        pf <- diablo_le_plots[i]
        label <- gsub("_enrichment\\.png$", "", basename(pf))
        label <- gsub("^DIABLO_", "", label)
        cat(sprintf("\n#### %s {.unnumbered}\n\n", label))
        chunk_id <- paste0("diablo-le-", i)
        cat(knitr::knit_child(text = sprintf(
            '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            chunk_id, pf
        ), quiet = TRUE))
        cat("\n\n")
    }
    figure_legend("GO/KEGG pathway enrichment of top DIABLO loading features per component and omics layer. For gene-based omics (transcriptomics, proteomics), standard KEGG gene ORA is used. For metabolomics, compound-based ORA maps metabolites to KEGG metabolic pathways. Red dashed line indicates p = 0.05.")
} else if (dir.exists(diablo_le_dir)) {
    cat("\n### DIABLO Loadings Enrichment\n\n")
    cat("*No enriched pathways found for DIABLO loadings. This can occur when top features do not map to KEGG pathways or when enrichment p-values do not meet the significance threshold.*\n\n")
}

# Show combined table
diablo_le_csv <- file.path(diablo_le_dir, "diablo_loadings_enrichment_all.csv")
if (file.exists(diablo_le_csv)) {
    df <- read.csv(diablo_le_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        cat("\n#### DIABLO Loadings Enrichment Table {.unnumbered}\n\n")
        print(make_dt(df, caption = "DIABLO loadings KEGG enrichment"))
    }
}
```

```{r loadings-enrich-mofa, eval=has_loadings_enrich, results='asis'}
mofa_le_dir <- file.path(loadings_enrich_dir, "mofa_loadings")
mofa_le_plots <- list.files(mofa_le_dir, pattern = "\\.png$", full.names = TRUE)

if (length(mofa_le_plots) > 0) {
    cat("\n### MOFA2 Weights Enrichment\n\n")
    for (i in seq_along(mofa_le_plots)) {
        pf <- mofa_le_plots[i]
        label <- gsub("_enrichment\\.png$", "", basename(pf))
        label <- gsub("^MOFA_", "", label)
        cat(sprintf("\n#### %s {.unnumbered}\n\n", label))
        chunk_id <- paste0("mofa-le-", i)
        cat(knitr::knit_child(text = sprintf(
            '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            chunk_id, pf
        ), quiet = TRUE))
        cat("\n\n")
    }
    figure_legend("GO/KEGG pathway enrichment of top MOFA2 weight features per factor and omics view. For gene-based omics, standard KEGG gene ORA is used. For metabolomics, compound-based ORA maps metabolites to KEGG metabolic pathways. Red dashed line indicates p = 0.05.")
} else if (dir.exists(mofa_le_dir)) {
    cat("\n### MOFA2 Weights Enrichment\n\n")
    cat("*No enriched pathways found for MOFA2 weights. This can occur when top features do not map to KEGG pathways or when enrichment p-values do not meet the significance threshold.*\n\n")
}

# Show combined table
mofa_le_csv <- file.path(mofa_le_dir, "mofa_weights_enrichment_all.csv")
if (file.exists(mofa_le_csv)) {
    df <- read.csv(mofa_le_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        cat("\n#### MOFA2 Weights Enrichment Table {.unnumbered}\n\n")
        print(make_dt(df, caption = "MOFA2 weights KEGG enrichment"))
    }
}
```

---

# Consensus Analysis {.tabset}

```{r consensus-setup, include=FALSE}
consensus_dir <- file.path(run_dir, "consensus", "consensus")
has_consensus <- dir.exists(consensus_dir) &&
    length(list.files(consensus_dir, "\\.png$|\\.csv$", recursive = TRUE)) > 0
```

```{r consensus-note, eval=!has_consensus, results='asis'}
cat("*No consensus analysis results found. Requires >= 2 integration methods (e.g. DIABLO + MOFA2 + SNF).*\n")
```

## Meta-Integration PCA {.tabset .unnumbered}

```{r consensus-meta-pca, eval=has_consensus, results='asis'}
meta_pca_plot <- file.path(consensus_dir, "plots", "meta_integration_pca.png")
meta_pca_labeled <- file.path(consensus_dir, "plots", "meta_integration_pca_labeled.png")
has_meta_pca <- file.exists(meta_pca_plot)
has_meta_pca_labeled <- file.exists(meta_pca_labeled)

if (has_meta_pca || has_meta_pca_labeled) {
    if (has_meta_pca) {
        cat("\n### By Cluster {.unnumbered}\n\n")
        cat(knitr::knit_child(text = sprintf(
            '```{r meta-pca-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            meta_pca_plot
        ), quiet = TRUE))
        figure_legend("Meta-integration PCA: latent factors from all integration methods (DIABLO, MOFA2, SNF) are concatenated and projected into a combined principal component space. Points represent samples, colored by original group, shaped by consensus cluster assignment.")
    }

    if (has_meta_pca_labeled) {
        cat("\n### With Sample Labels {.unnumbered}\n\n")
        cat(knitr::knit_child(text = sprintf(
            '```{r meta-pca-labeled-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            meta_pca_labeled
        ), quiet = TRUE))
        figure_legend("Meta-integration PCA with sample labels. Samples are projected into a combined latent space from all integration methods. Labels identify individual samples for cross-referencing with other analyses.")
    }

    render_commentary("meta_integration_pca")
} else {
    cat("*Meta-integration PCA plot not available.*\n")
}
```

```{r consensus-method-agreement, eval=FALSE}
# Removed: Method Agreement section was not informative
```

## Feature Importance Correlation {.unnumbered}

```{r consensus-feature-corr, eval=has_consensus, results='asis'}
feat_corr_plot <- file.path(consensus_dir, "plots", "feature_rank_correlation.png")
if (file.exists(feat_corr_plot)) {
    cat(knitr::knit_child(text = sprintf(
        '```{r feat-corr-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        feat_corr_plot
    ), quiet = TRUE))
    figure_legend("Spearman rank correlation of feature importance scores between integration methods. High correlations indicate that different methods identify similar key features.")
    render_commentary("feature_rank_correlation")
}
```

## Top Robust Features Across Methods {.unnumbered}

```{r consensus-robust-features, eval=has_consensus, fig.width=11, fig.height=15}
# Rebuild the robust-features heatmap inline (rather than embedding the pipeline
# PNG) so the row labels can be enlarged and annotated with a short description
# per gene/protein id. Descriptions are only assembled here, on gene_protein_mapping.
# Fill mirrors plot_robust_features(): within-method importance rank as a percentile.
robust_ok  <- FALSE
fic <- file.path(consensus_dir, "feature_importance_comparison.csv")
rfc <- file.path(consensus_dir, "robust_features.csv")
if (file.exists(fic) && file.exists(rfc)) {
    fi <- read.csv(fic, stringsAsFactors = FALSE)
    rf <- read.csv(rfc, stringsAsFactors = FALSE)
    method_cols <- intersect(c("mofa", "diablo", "snf", "spls"), names(fi))
    if (nrow(rf) > 0 && length(method_cols) >= 1 && "feature" %in% names(fi)) {
        imp <- as.matrix(fi[, method_cols, drop = FALSE])
        rownames(imp) <- fi$feature
        rank_pct <- apply(imp, 2, function(x) {
            r <- rep(NA_real_, length(x)); v <- !is.na(x)
            r[v] <- rank(-x[v], ties.method = "average"); r
        })
        n_total <- colSums(!is.na(imp))
        top     <- head(rf$feature, 30)

        # "id — short description" labels, falling back id -> gene desc -> gene name
        # -> protein symbol -> bare id, so non-model features still read clearly.
        label_for <- function(ids) {
            if (is.null(gene_protein_mapping)) return(ids)
            ann <- rep(NA_character_, length(ids))
            mg  <- match(ids, gene_protein_mapping$gene_id)
            if ("gene_description" %in% names(gene_protein_mapping))
                ann <- gene_protein_mapping$gene_description[mg]
            if ("gene_name" %in% names(gene_protein_mapping)) {
                gn  <- gene_protein_mapping$gene_name[mg]
                ann <- ifelse(is.na(ann) | !nzchar(ann), gn, ann)
            }
            if ("protein_symbol" %in% names(gene_protein_mapping)) {
                ps  <- gene_protein_mapping$protein_symbol[match(ids, gene_protein_mapping$protein_id)]
                ann <- ifelse(is.na(ann) | !nzchar(ann), ps, ann)
            }
            ann <- ifelse(is.na(ann), "", ann)
            ann <- ifelse(nchar(ann) > 45, paste0(substr(ann, 1, 42), "..."), ann)
            ifelse(nzchar(ann), paste0(ids, "  —  ", ann), ids)
        }

        plot_df <- do.call(rbind, lapply(method_cols, function(m) {
            data.frame(feature = top, method = m,
                       rank_pct = rank_pct[match(top, rownames(imp)), m] / n_total[m] * 100,
                       stringsAsFactors = FALSE)
        }))
        lvl <- rev(top)
        plot_df$feature <- factor(plot_df$feature, levels = lvl, labels = label_for(lvl))

        p <- ggplot2::ggplot(plot_df,
                 ggplot2::aes(x = method, y = feature, fill = rank_pct)) +
            ggplot2::geom_tile() +
            ggplot2::scale_fill_viridis_c(option = "plasma", na.value = "grey90",
                     direction = -1, labels = function(x) paste0(round(x), "%")) +
            ggplot2::labs(title = "Top Robust Features Across Methods",
                          subtitle = "Features consistently important in multiple methods",
                          x = "Method", y = NULL, fill = "Rank\n(percentile)") +
            ggplot2::theme_minimal() +
            ggplot2::theme(axis.text.y = ggplot2::element_text(size = 12),
                           axis.text.x = ggplot2::element_text(size = 11),
                           plot.margin = ggplot2::margin(6, 6, 6, 14))
        print(p)
        robust_ok <- TRUE
    }
}
```

```{r consensus-robust-features-legend, eval=has_consensus, results='asis'}
if (isTRUE(robust_ok)) {
    figure_legend("Top robust features across integration methods. Fill is each feature's within-method importance rank as a percentile (0% = most important), so scores are comparable across methods. Rows are labelled by gene/protein id with a short description where available (gene_description/gene_name from ehi_gene_annotation.csv; protein symbol from the proteomics table). Sources: feature_importance_comparison.csv and robust_features.csv in the consensus directory.")
    render_commentary("robust_features")
} else {
    cat("*No robust features found across methods. This can occur when methods identify largely different feature sets.*\n")
}
```

```{r consensus-robust-table, eval=has_consensus}
robust_csv <- file.path(consensus_dir, "robust_features.csv")
if (file.exists(robust_csv)) {
    robust_df <- read.csv(robust_csv, stringsAsFactors = FALSE)
    if (nrow(robust_df) > 0) {
        # Resolve the consensus "feature" column to original gene_id + protein_id
        # + description. The feature may be a harmonized GENE_n (GENE_n = row n of
        # gene_protein_mapping, see 02_mae.R) or already a gene_id; handle both,
        # then join protein_id / description by gene_id.
        if (!is.null(gene_protein_mapping) && "feature" %in% names(robust_df)) {
            gid       <- robust_df$feature
            is_gene_n <- grepl("^GENE_\\d+$", gid)
            if (any(is_gene_n)) {
                n  <- suppressWarnings(as.integer(sub("^GENE_", "", gid)))
                ok <- is_gene_n & !is.na(n) & n >= 1 & n <= nrow(gene_protein_mapping)
                gid[ok] <- gene_protein_mapping$gene_id[n[ok]]
            }
            robust_df$gene_id    <- gid
            mi                   <- match(robust_df$gene_id, gene_protein_mapping$gene_id)
            robust_df$protein_id <- gene_protein_mapping$protein_id[mi]
            if ("gene_description" %in% names(gene_protein_mapping)) {
                robust_df$description <- gene_protein_mapping$gene_description[mi]
            }
            front     <- intersect(c("gene_id", "protein_id", "description"), names(robust_df))
            robust_df <- robust_df[, c(front, setdiff(names(robust_df), c(front, "feature")))]
        }
        make_dt(robust_df, caption = "Robust Features Across Integration Methods", pageLength = 15)
    }
}
```

---

# Output Files

```{r output-files, results='asis'}
cat(sprintf("All multi-omics integration results have been saved to:\n\n`%s`\n\n", run_dir))

cat("**Subdirectories:**\n\n")

subdirs <- list.dirs(run_dir, full.names = FALSE, recursive = FALSE)
subdirs <- subdirs[subdirs != "execution_info"]
if (length(subdirs) > 0) {
    for (sd in sort(subdirs)) {
        n_files <- length(list.files(file.path(run_dir, sd), recursive = TRUE))
        cat(sprintf("- **%s/** (%d files)\n", sd, n_files))
    }
}
```

---

# Methods

```{r methods, results='asis'}
cat("## Multi-Omics Integration Pipeline\n\n")

cat("This report was generated by the **multiomics-core** pipeline. ",
    "The following analytical steps were performed:\n\n")

cat("1. **Data Harmonization**: Individual omics layers (preprocessed by their respective sub-pipelines) ",
    "were aligned to a common set of samples and assembled into a `MultiAssayExperiment` object.\n\n")

methods <- multi_cfg$integration$methods %||% c("DIABLO", "SNF")
step_num <- 2
if ("DIABLO" %in% methods) {
    cat(sprintf("%d. **DIABLO (Data Integration Analysis for Biomarker discovery using Latent cOmponents)**: ", step_num),
        "A supervised multi-omics integration method from the mixOmics package that identifies ",
        "discriminant features across omics layers using sparse PLS-DA.\n\n")
    step_num <- step_num + 1
}
if ("SNF" %in% methods) {
    cat(sprintf("%d. **SNF (Similarity Network Fusion)**: ", step_num),
        "An unsupervised method that constructs sample similarity networks for each omics layer ",
        "and fuses them into a single network for clustering.\n\n")
    step_num <- step_num + 1
}
if ("MOFA2" %in% methods) {
    cat(sprintf("%d. **MOFA2 (Multi-Omics Factor Analysis v2)**: ", step_num),
        "An unsupervised Bayesian factor model that identifies latent factors capturing ",
        "shared and omics-specific sources of variation.\n\n")
    step_num <- step_num + 1
}

cat(sprintf("%d. **Cross-Omics Enrichment**: Pathway enrichment results were aggregated across omics ", step_num),
    "using Fisher's method for combined p-values.\n\n")

cat("## Software\n\n")
cat("```\n")
cat(capture.output(sessionInfo()), sep = "\n")
cat("\n```\n")
```
