---
title: "Multi-Omics Integration Report"
subtitle: "multiomics-core Pipeline"
author: "Bioinformatician: `r yaml::read_yaml(file.path(normalizePath(dirname(knitr::current_input())), 'execution_info', 'config_used.yaml'))$project$analyst`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    fig_width: 10
    fig_height: 7
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%"
)

library(yaml)
library(knitr)
library(DT)
library(plotly)
library(ggplot2)
library(htmltools)
library(jsonlite)

# Helper function for downloadable DT tables
make_dt <- function(df, caption = NULL, pageLength = 5) {
    # Format real-number columns: < 0.001 → scientific, else 3 decimals
    for (col in colnames(df)) {
        if (is.numeric(df[[col]])) {
            vals <- df[[col]]
            formatted <- ifelse(
                is.na(vals), NA_character_,
                ifelse(vals == as.integer(vals) & !grepl("\\.", as.character(vals)),
                       as.character(vals),
                       ifelse(abs(vals) < 0.001 & vals != 0,
                              sprintf("%.3e", vals),
                              sprintf("%.3f", vals))))
            df[[col]] <- formatted
        }
    }
    DT::datatable(
        df,
        caption = caption,
        extensions = 'Buttons',
        options = list(
            scrollX = TRUE,
            pageLength = pageLength,
            lengthMenu = c(5, 10, 25, 50, 100),
            dom = 'Blfrtip',
            buttons = list(
                list(extend = 'csv', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data")),
                list(extend = 'excel', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data"))
            ),
            columnDefs = list(list(className = 'dt-left', targets = '_all'))
        ),
        rownames = FALSE
    )
}

# Helper: render figure legend (paper-style caption below figure)
figure_legend <- function(text) {
    cat(sprintf("<p class='figure-legend' style='text-align: center; font-size: 0.9em; color: #555; margin: 4px auto 12px auto; max-width: 88%%; line-height: 1.45; font-style: italic;'>%s</p>\n", text))
}

# Helper: render commentary block for a figure_id (collapsible)
render_commentary <- function(figure_id) {
    if (!has_commentary) return(invisible(NULL))
    c <- commentary[[figure_id]]
    if (is.null(c)) return(invisible(NULL))

    cat("<details class='commentary-details' style='margin: 4px 0 18px 0;'>\n")
    cat("<summary style='cursor: pointer; color: #4a90d9; font-weight: 500; font-size: 0.93em; padding: 3px 0;'>Show AI Interpretation</summary>\n")

    html <- "<div class='commentary-block' style='background: #f8f9fa; border-left: 4px solid #4a90d9; padding: 12px 16px; margin: 6px 0 4px 0; border-radius: 0 4px 4px 0; font-size: 0.92em;'>\n"
    html <- paste0(html, "<p style='margin: 0 0 8px 0;'><strong>What this shows:</strong> ",
                   htmltools::htmlEscape(c$what_is_this %||% ""), "</p>\n")

    bio_interp <- c$biological_interpretation %||% ""
    if (nzchar(bio_interp)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Interpretation:</strong> ",
                       htmltools::htmlEscape(bio_interp), "</p>\n")
    }

    obs <- c$observations
    if (length(obs) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Key observations:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (o in obs) html <- paste0(html, "<li>", htmltools::htmlEscape(o), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    stat_assess <- c$statistical_assessment %||% ""
    if (nzchar(stat_assess)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Statistical note:</strong> ",
                       htmltools::htmlEscape(stat_assess), "</p>\n")
    }

    issues <- c$issues_checks
    if (length(issues) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Checks:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (issue in issues) html <- paste0(html, "<li>", htmltools::htmlEscape(issue), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    steps <- c$next_steps
    if (length(steps) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Next steps:</strong></p>\n<ul style='margin: 2px 0 4px 0; padding-left: 20px;'>\n")
        for (s in steps) html <- paste0(html, "<li>", htmltools::htmlEscape(s), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    backend <- c$backend %||% "unknown"
    confidence <- c$confidence %||% ""
    backend_label <- if (grepl("claude", backend, ignore.case = TRUE)) {
        "Claude (Anthropic) + K-Dense AI scientific-skills"
    } else if (grepl("openai", backend, ignore.case = TRUE)) {
        "OpenAI + K-Dense AI scientific-skills"
    } else if (backend == "data_driven") {
        "data-driven + K-Dense AI scientific-skills"
    } else {
        backend
    }
    if (nzchar(confidence) && confidence != "none") {
        backend_label <- paste0(backend_label, " | confidence: ", confidence)
    }
    html <- paste0(html, "<p style='margin: 4px 0 0 0; color: #888; font-size: 0.85em;'><em>Generated by: ",
                   htmltools::htmlEscape(backend_label), "</em></p>\n</div>\n")

    cat(html)
    cat("</details>\n")
}

# Paths
run_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
config_path <- file.path(run_dir, "execution_info", "config_used.yaml")

# Multiomics output subdirectories
integration_dir  <- file.path(run_dir, "integration")
diablo_dir       <- file.path(integration_dir, "diablo")
snf_dir          <- file.path(integration_dir, "snf")
mofa_dir         <- file.path(integration_dir, "mofa")
enrichment_dir   <- file.path(run_dir, "cross_enrichment")
loadings_enrich_dir <- file.path(run_dir, "loadings_enrichment")
rna_prot_dir     <- file.path(run_dir, "rna_protein_correlation")
multigsea_dir    <- file.path(run_dir, "multigsea")
commentary_dir   <- file.path(run_dir, "commentary")

# Extended analysis directories
feature_network_dir  <- file.path(run_dir, "feature_network")
concordance_dir      <- file.path(run_dir, "concordance")
consensus_base_dir   <- file.path(run_dir, "consensus")
consensus_dir        <- file.path(consensus_base_dir, "consensus")
stability_dir        <- file.path(consensus_base_dir, "stability")
foundational_base_dir <- file.path(run_dir, "foundational")
foundational_dir     <- file.path(foundational_base_dir, "foundational")
mechanistic_base_dir <- file.path(run_dir, "mechanistic")
mechanistic_dir      <- file.path(mechanistic_base_dir, "mechanistic")

# Load commentary if available
commentary <- NULL
has_commentary <- FALSE
commentary_json <- file.path(commentary_dir, "all_commentary.json")
if (file.exists(commentary_json)) {
    commentary <- tryCatch(
        jsonlite::read_json(commentary_json),
        error = function(e) NULL
    )
    has_commentary <- !is.null(commentary) && length(commentary) > 0
}

# Load config
cfg <- yaml::read_yaml(config_path)
multi_cfg <- cfg$modes$multiomics

# Detect which sections have data
has_diablo      <- dir.exists(diablo_dir) && length(list.files(diablo_dir, "\\.png$")) > 0
has_snf         <- dir.exists(snf_dir) && length(list.files(snf_dir, "\\.png$")) > 0
has_mofa        <- dir.exists(mofa_dir) && length(list.files(mofa_dir, "\\.png$")) > 0
has_integration <- has_diablo || has_snf || has_mofa
has_enrichment  <- dir.exists(enrichment_dir) && length(list.files(enrichment_dir, "\\.png$|\\.csv$")) > 0
has_transcriptomics <- "transcriptomics" %in% (cfg$global$omics_present %||% character(0)) ||
    "rna" %in% (cfg$global$omics_present %||% character(0)) ||
    !is.null(cfg$modes$rna)
has_proteomics <- "proteomics" %in% (cfg$global$omics_present %||% character(0)) ||
    !is.null(cfg$modes$proteomics)
has_metabolomics <- "metabolomics" %in% (cfg$global$omics_present %||% character(0)) ||
    !is.null(cfg$modes$metabolomics)
has_rna_prot    <- has_transcriptomics && has_proteomics &&
    dir.exists(file.path(rna_prot_dir, "plots")) &&
    length(list.files(file.path(rna_prot_dir, "plots"), "\\.png$")) > 0
has_multigsea   <- dir.exists(multigsea_dir) && length(list.files(multigsea_dir, "\\.png$")) > 0
has_loadings_enrich <- dir.exists(loadings_enrich_dir) &&
    length(list.files(loadings_enrich_dir, "\\.png$", recursive = TRUE)) > 0
has_feature_network <- dir.exists(file.path(feature_network_dir, "plots")) &&
    length(list.files(file.path(feature_network_dir, "plots"), "\\.png$")) > 0
has_concordance     <- dir.exists(concordance_dir) &&
    length(list.files(concordance_dir, "\\.png$|\\.csv$")) > 0
has_consensus       <- dir.exists(consensus_dir) &&
    (length(list.files(file.path(consensus_dir, "plots"), "\\.png$")) > 0 ||
     length(list.files(consensus_dir, "\\.csv$")) > 0)
has_stability       <- dir.exists(stability_dir) &&
    length(list.files(file.path(stability_dir, "plots"), "\\.png$")) > 0
has_foundational    <- dir.exists(foundational_dir) &&
    (length(list.files(file.path(foundational_dir, "plots"), "\\.png$")) > 0 ||
     length(list.files(foundational_dir, "\\.csv$")) > 0)
has_mechanistic     <- dir.exists(mechanistic_dir) &&
    (length(list.files(file.path(mechanistic_dir, "plots"), "\\.png$")) > 0 ||
     length(list.files(mechanistic_dir, "\\.csv$")) > 0)

# Load gene-protein mapping if available
mapping_file <- file.path(run_dir, "gene_protein_mapping.csv")
gene_protein_mapping <- NULL
if (file.exists(mapping_file)) {
    gene_protein_mapping <- tryCatch(
        read.csv(mapping_file, stringsAsFactors = FALSE),
        error = function(e) NULL
    )
}
```

```{r force-dependencies, include=FALSE}
# Force DT and Plotly dependencies to be included in the HTML
DT::datatable(data.frame(x = 1))
plotly::plot_ly(data.frame(x = 1, y = 1), x = ~x, y = ~y, type = "scatter", mode = "markers")
```

# Analysis Parameters {.tabset}

## Project

```{r params-project}
proj <- data.frame(
  Parameter = c("Project Name", "Analysis Round", "Project Directory",
                 "Raw Data Path", "Output Path"),
  Value = c(cfg$project$name %||% "N/A",
            cfg$project$analysis_round %||% "N/A",
            cfg$project$dir %||% "N/A",
            cfg$paths$raw %||% "N/A",
            cfg$paths$out %||% "N/A"),
  stringsAsFactors = FALSE
)
kable(proj, col.names = c("Parameter", "Value"))
```

## Integration Methods

```{r params-integration}
methods <- multi_cfg$integration$methods %||% c("DIABLO", "SNF")
feat_sel <- multi_cfg$feature_selection$method %||% "variance"
feat_n <- multi_cfg$feature_selection$top_n %||% 500
cond_col <- multi_cfg$condition_column %||% cfg$design$condition_column %||% "condition"

settings <- data.frame(
  Parameter = c(
    "Integration Methods",
    "Feature Selection Method",
    "Top N Features",
    "Condition Column",
    "Require 1:1 Gene-Protein Mapping"
  ),
  Value = c(
    paste(methods, collapse = ", "),
    feat_sel,
    feat_n,
    cond_col,
    ifelse(isTRUE(multi_cfg$require_one_to_one_mapping), "Yes", "No")
  ),
  stringsAsFactors = FALSE
)
kable(settings, col.names = c("Parameter", "Value"))
```

## Omics Layers

```{r params-omics}
omics_modes <- setdiff(names(cfg$modes), "multiomics")
# In payload mode, get omics from global$omics_present or payload_paths
if (length(omics_modes) == 0) {
    omics_modes <- cfg$global$omics_present %||%
                   names(cfg$modes$multiomics$payload_paths) %||%
                   harmonization$omics_available
}
if (length(omics_modes) > 0) {
    omics_tbl <- data.frame(
        Omics = omics_modes,
        Configured = "Yes",
        stringsAsFactors = FALSE
    )
    kable(omics_tbl, col.names = c("Omics Layer", "Configured"))
} else {
    cat("No individual omics modes configured (payload mode).\n")
}
```

## Feature ID Mapping {.unnumbered}

```{r feature-mapping-table, results='asis'}
# Show GENE_xxx to GL/KAE mapping if MAE has harmonized features
mapping_csv <- file.path(run_dir, "gene_protein_mapping.csv")
if (file.exists(mapping_csv)) {
    gpm <- read.csv(mapping_csv, stringsAsFactors = FALSE)
    if (nrow(gpm) > 0) {
        cat("The integration pipeline harmonizes features across omics layers using internal IDs (GENE\\_1, GENE\\_2, etc.). The table below maps these to the original gene and protein identifiers.\n\n")
        display_cols <- intersect(c("gene_id", "protein_id", "gene_symbol", "mapping_source"), colnames(gpm))
        make_dt(gpm[, display_cols, drop = FALSE],
                caption = "Gene-Protein Feature ID Mapping",
                pageLength = 10)
    }
}
```

---

`r if (has_transcriptomics) "# RNA-Protein Correlation {.tabset}"`

```{r rna-prot-note, eval=has_transcriptomics && !has_rna_prot, results='asis'}
cat("*No RNA-protein correlation results found. This analysis requires both transcriptomics and proteomics data.*\n")
```

## Concordance Distribution

```{r rna-prot-concordance, eval=has_transcriptomics && has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "rna_protein_concordance.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r rna-prot-concordance-legend, eval=has_transcriptomics && has_rna_prot, results='asis'}
figure_legend("Distribution of per-gene RNA-protein Pearson correlations across samples. The histogram shows how well transcript levels predict protein abundance for individual genes.")
render_commentary("rna_protein_concordance")
```

## DE Scatter (All Contrasts) {.tabset}

### Static

```{r rna-prot-de-scatter, eval=has_transcriptomics && has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "rna_protein_de_scatter.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r rna-prot-de-legend, eval=has_transcriptomics && has_rna_prot, results='asis'}
figure_legend("RNA vs protein differential expression scatter plot per contrast. Each point is a gene/protein pair. Diagonal concordance indicates matching fold-change direction; off-diagonal points suggest post-transcriptional regulation.")
render_commentary("rna_protein_de_scatter")
```

### Interactive

```{r rna-prot-de-interactive, eval=has_transcriptomics && has_rna_prot, results='asis'}
conc_csv <- file.path(run_dir, "concordance", "concordance_transcriptomics_vs_proteomics.csv")
if (file.exists(conc_csv)) {
    conc <- read.csv(conc_csv, stringsAsFactors = FALSE)
    conc <- conc[is.finite(conc$logFC_rna) & is.finite(conc$logFC_prot), ]

    # Build label from available columns
    conc$label <- if ("gene_symbol" %in% colnames(conc)) conc$gene_symbol
                  else if ("gene_id" %in% colnames(conc)) conc$gene_id
                  else if ("protein_id" %in% colnames(conc)) conc$protein_id
                  else rownames(conc)
    conc$label[is.na(conc$label) | conc$label == ""] <- "unknown"

    cols <- c("concordant" = "#2ca02c", "discordant" = "#d62728",
              "rna_only" = "#1f77b4", "protein_only" = "#ff7f0e", "not_sig" = "#cccccc")

    p <- plotly::plot_ly(
        data = conc,
        x = ~logFC_rna, y = ~logFC_prot,
        text = ~paste0(label, "\nRNA log2FC: ", round(logFC_rna, 2),
                       "\nProtein log2FC: ", round(logFC_prot, 2),
                       "\nConcordance: ", concordance),
        color = ~concordance,
        colors = cols,
        type = "scatter", mode = "markers",
        hoverinfo = "text",
        marker = list(size = 5, opacity = 0.6)
    ) %>%
        plotly::layout(
            title = "RNA vs Protein Fold Changes (Interactive)",
            xaxis = list(title = "RNA log2FC", zeroline = TRUE),
            yaxis = list(title = "Protein log2FC", zeroline = TRUE),
            shapes = list(
                list(type = "line", x0 = -5, x1 = 5, y0 = 0, y1 = 0,
                     line = list(color = "grey", dash = "dash")),
                list(type = "line", x0 = 0, x1 = 0, y0 = -5, y1 = 5,
                     line = list(color = "grey", dash = "dash"))
            )
        )
    print(htmltools::tagList(p))
} else {
    cat("*Concordance data not available for interactive plot.*\n")
}
```

## Per-Contrast DE Details {.tabset}

```{r rna-prot-per-contrast, eval=has_transcriptomics && has_rna_prot, results='asis'}
per_contrast_dir <- file.path(rna_prot_dir, "per_contrast")
if (dir.exists(per_contrast_dir)) {
    contrast_dirs <- list.dirs(per_contrast_dir, full.names = TRUE, recursive = FALSE)
    for (cdir in contrast_dirs) {
        cname <- basename(cdir)
        cat(sprintf("\n### %s\n\n", gsub("_", " ", cname)))

        de_scatter <- file.path(cdir, "plots", "rna_protein_de_scatter.png")
        if (file.exists(de_scatter)) {
            cat(knitr::knit_child(text = sprintf(
                '```{r rna-prot-scatter-%s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                gsub("[^a-zA-Z0-9]", "", cname), de_scatter
            ), quiet = TRUE))
        }

        te_hist <- file.path(cdir, "plots", "translation_efficiency_hist.png")
        if (file.exists(te_hist)) {
            cat(knitr::knit_child(text = sprintf(
                '```{r rna-prot-te-%s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                gsub("[^a-zA-Z0-9]", "", cname), te_hist
            ), quiet = TRUE))
        }

        te_scatter <- file.path(cdir, "plots", "translation_efficiency_scatter.png")
        if (file.exists(te_scatter)) {
            cat(knitr::knit_child(text = sprintf(
                '```{r rna-prot-tes-%s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                gsub("[^a-zA-Z0-9]", "", cname), te_scatter
            ), quiet = TRUE))
        }

        # Summary table
        de_csv <- file.path(cdir, "tables", "rna_protein_de_concordance.csv")
        if (file.exists(de_csv)) {
            de_df <- read.csv(de_csv, stringsAsFactors = FALSE)
            n_total <- nrow(de_df)
            n_conc <- sum(de_df$concordant, na.rm = TRUE)
            r_val <- cor(de_df$rna_log2FC, de_df$protein_log2FC, use = "complete.obs")
            tf_msg <- ""
            if ("is_transcription_factor" %in% colnames(de_df)) {
                n_tf <- sum(de_df$is_transcription_factor, na.rm = TRUE)
                if (n_tf > 0) tf_msg <- sprintf(", %d known TFs", n_tf)
            }
            cat(sprintf("\n**%s:** %d gene-protein pairs, r = %.3f, %d/%d (%.0f%%) concordant%s\n\n",
                        gsub("_", " ", cname), n_total, r_val,
                        n_conc, n_total, 100 * n_conc / n_total, tf_msg))
        }
    }
}
```

## Translation Efficiency (All Contrasts)

```{r rna-prot-te-hist, eval=has_transcriptomics && has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "translation_efficiency_hist.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r rna-prot-te-hist-legend, eval=has_transcriptomics && has_rna_prot, results='asis'}
figure_legend("Distribution of translation efficiency changes per contrast (protein log2FC - RNA log2FC). Values near 0 indicate transcriptional control; large positive/negative values suggest post-transcriptional regulation.")
render_commentary("translation_efficiency_hist")
```

```{r rna-prot-te-scatter, eval=has_transcriptomics && has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "translation_efficiency_scatter.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r rna-prot-te-scatter-legend, eval=has_transcriptomics && has_rna_prot, results='asis'}
figure_legend("Translation efficiency scatter plot per contrast comparing RNA and protein fold-changes.")
render_commentary("translation_efficiency_scatter")
```

## RNA-Protein Concordance Visualization {.tabset}

```{r conc-viz-setup, include=FALSE}
# Load concordance CSV for visualization section
conc_csv_path <- file.path(concordance_dir, "concordance_transcriptomics_vs_proteomics.csv")
has_conc_viz <- file.exists(conc_csv_path)
conc_df <- NULL
if (has_conc_viz) {
    conc_df <- tryCatch(read.csv(conc_csv_path, stringsAsFactors = FALSE), error = function(e) NULL)
    has_conc_viz <- !is.null(conc_df) && nrow(conc_df) > 0
}
if (has_conc_viz) {
    library(patchwork)
    library(forcats)
    # Derive significance flags using padj < 0.05
    conc_df$sig_prot_flag <- !is.na(conc_df$padj_prot) & conc_df$padj_prot < 0.05
    conc_df$sig_rna_flag  <- !is.na(conc_df$padj_rna) & conc_df$padj_rna < 0.05
    # Trend labels for protein
    conc_df$prot_trend <- ifelse(!conc_df$sig_prot_flag, "Not sig",
                          ifelse(conc_df$logFC_prot > 0, "Up", "Down"))
    # RNA change labels
    conc_df$rna_change <- ifelse(!conc_df$sig_rna_flag, "Not sig",
                          ifelse(conc_df$logFC_rna > 0, "Up", "Down"))
    # Gene label for plots — use gene_symbol if available, else gene_id
    conc_df$gene_label <- if ("gene_symbol" %in% colnames(conc_df)) conc_df$gene_symbol
                          else if ("gene_id" %in% colnames(conc_df)) conc_df$gene_id
                          else conc_df$protein_id
}
```

```{r conc-viz-note, eval=!has_conc_viz, results='asis'}
cat("*No RNA-protein concordance data found. This analysis requires both transcriptomics and proteomics differential expression results.*\n")
```

### Lollipop Plots {.tabset}

```{r conc-lollipop-all, eval=has_conc_viz, results='asis', fig.width=12, fig.height=10}
tryCatch({
    # Filter to significant proteins
    sig_prot <- conc_df[conc_df$sig_prot_flag, ]
    if (nrow(sig_prot) > 200) {
        sig_prot <- sig_prot[order(-abs(sig_prot$logFC_prot)), ][1:100, ]
    }
    if (nrow(sig_prot) > 0) {
        sig_prot$gene_label_f <- forcats::fct_reorder(sig_prot$gene_label, sig_prot$logFC_prot)
        prot_colors <- c("Up" = "springgreen4", "Down" = "violetred", "Not sig" = "gray60")
        rna_colors  <- c("Up" = "lightgreen", "Down" = "lightcoral", "Not sig" = "gray60")

        p_top <- ggplot(sig_prot, aes(x = gene_label_f, y = logFC_prot, color = prot_trend)) +
            geom_segment(aes(xend = gene_label_f, y = 0, yend = logFC_prot), linewidth = 0.5) +
            geom_point(size = 2) +
            scale_color_manual(values = prot_colors, name = "Trend") +
            labs(y = "Protein log2FC", x = NULL, title = "All Significant Proteins") +
            theme_minimal(base_size = 11) +
            theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
                  panel.grid.major.x = element_blank())

        p_bot <- ggplot(sig_prot, aes(x = gene_label_f, y = logFC_rna, color = rna_change)) +
            geom_segment(aes(xend = gene_label_f, y = 0, yend = logFC_rna), linewidth = 0.5) +
            geom_point(size = 2) +
            scale_color_manual(values = rna_colors, name = "RNA Change") +
            labs(y = "RNA log2FC", x = "Gene (sorted by protein logFC)") +
            theme_minimal(base_size = 11) +
            theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6),
                  panel.grid.major.x = element_blank())

        print(p_top / p_bot)
    } else {
        cat("*No significant proteins found (padj < 0.05).*\n")
    }
}, error = function(e) cat(sprintf("*Lollipop plot (All) could not be generated: %s*\n", e$message)))
```

```{r conc-lollipop-all-legend, eval=has_conc_viz, results='asis'}
figure_legend("Lollipop plot of all significant proteins (padj < 0.05). Top panel: protein log2 fold-change sorted by magnitude, colored by direction. Bottom panel: corresponding RNA log2 fold-change for the same genes, preserving protein order.")
```

```{r conc-lollipop-up, eval=has_conc_viz, results='asis', fig.width=12, fig.height=10}
tryCatch({
    up_prot <- conc_df[conc_df$sig_prot_flag & conc_df$logFC_prot > 0, ]
    if (nrow(up_prot) > 0) {
        up_prot$gene_label_f <- forcats::fct_reorder(up_prot$gene_label, up_prot$logFC_prot)
        prot_colors <- c("Up" = "springgreen4", "Down" = "violetred", "Not sig" = "gray60")
        rna_colors  <- c("Up" = "lightgreen", "Down" = "lightcoral", "Not sig" = "gray60")

        p_top <- ggplot(up_prot, aes(x = gene_label_f, y = logFC_prot, color = prot_trend)) +
            geom_segment(aes(xend = gene_label_f, y = 0, yend = logFC_prot), linewidth = 0.5) +
            geom_point(size = 2) +
            scale_color_manual(values = prot_colors, name = "Trend") +
            labs(y = "Protein log2FC", x = NULL, title = "Upregulated Proteins") +
            theme_minimal(base_size = 11) +
            theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
                  panel.grid.major.x = element_blank())

        p_bot <- ggplot(up_prot, aes(x = gene_label_f, y = logFC_rna, color = rna_change)) +
            geom_segment(aes(xend = gene_label_f, y = 0, yend = logFC_rna), linewidth = 0.5) +
            geom_point(size = 2) +
            scale_color_manual(values = rna_colors, name = "RNA Change") +
            labs(y = "RNA log2FC", x = "Gene (sorted by protein logFC)") +
            theme_minimal(base_size = 11) +
            theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6),
                  panel.grid.major.x = element_blank())

        print(p_top / p_bot)
    } else {
        cat("*No significantly upregulated proteins found.*\n")
    }
}, error = function(e) cat(sprintf("*Lollipop plot (Up) could not be generated: %s*\n", e$message)))
```

```{r conc-lollipop-up-legend, eval=has_conc_viz, results='asis'}
figure_legend("Lollipop plot of significantly upregulated proteins (logFC > 0, padj < 0.05). Bottom panel shows corresponding RNA fold-changes to reveal concordance or discordance at the transcript level.")
```

```{r conc-lollipop-down, eval=has_conc_viz, results='asis', fig.width=12, fig.height=10}
tryCatch({
    down_prot <- conc_df[conc_df$sig_prot_flag & conc_df$logFC_prot < 0, ]
    if (nrow(down_prot) > 0) {
        down_prot$gene_label_f <- forcats::fct_reorder(down_prot$gene_label, down_prot$logFC_prot)
        prot_colors <- c("Up" = "springgreen4", "Down" = "violetred", "Not sig" = "gray60")
        rna_colors  <- c("Up" = "lightgreen", "Down" = "lightcoral", "Not sig" = "gray60")

        p_top <- ggplot(down_prot, aes(x = gene_label_f, y = logFC_prot, color = prot_trend)) +
            geom_segment(aes(xend = gene_label_f, y = 0, yend = logFC_prot), linewidth = 0.5) +
            geom_point(size = 2) +
            scale_color_manual(values = prot_colors, name = "Trend") +
            labs(y = "Protein log2FC", x = NULL, title = "Downregulated Proteins") +
            theme_minimal(base_size = 11) +
            theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
                  panel.grid.major.x = element_blank())

        p_bot <- ggplot(down_prot, aes(x = gene_label_f, y = logFC_rna, color = rna_change)) +
            geom_segment(aes(xend = gene_label_f, y = 0, yend = logFC_rna), linewidth = 0.5) +
            geom_point(size = 2) +
            scale_color_manual(values = rna_colors, name = "RNA Change") +
            labs(y = "RNA log2FC", x = "Gene (sorted by protein logFC)") +
            theme_minimal(base_size = 11) +
            theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6),
                  panel.grid.major.x = element_blank())

        print(p_top / p_bot)
    } else {
        cat("*No significantly downregulated proteins found.*\n")
    }
}, error = function(e) cat(sprintf("*Lollipop plot (Down) could not be generated: %s*\n", e$message)))
```

```{r conc-lollipop-down-legend, eval=has_conc_viz, results='asis'}
figure_legend("Lollipop plot of significantly downregulated proteins (logFC < 0, padj < 0.05). Bottom panel shows corresponding RNA fold-changes to assess post-transcriptional regulation.")
```

### Pie Charts {.tabset}

```{r conc-pie-up, eval=has_conc_viz, results='asis', fig.width=7, fig.height=7}
tryCatch({
    up_prot <- conc_df[conc_df$sig_prot_flag & conc_df$logFC_prot > 0, ]
    if (nrow(up_prot) > 0) {
        pie_data <- data.frame(table(up_prot$rna_change))
        colnames(pie_data) <- c("Category", "Count")
        pie_data$Pct <- round(100 * pie_data$Count / sum(pie_data$Count), 1)
        pie_data$Label <- paste0(pie_data$Category, "\n", pie_data$Count, " (", pie_data$Pct, "%)")
        pie_colors <- c("Up" = "green3", "Down" = "red3", "Not sig" = "gray70")

        p <- ggplot(pie_data, aes(x = 2, y = Count, fill = Category)) +
            geom_bar(stat = "identity", width = 1, color = "white") +
            coord_polar("y", start = 0) +
            scale_fill_manual(values = pie_colors) +
            xlim(0.5, 2.5) +
            geom_text(aes(label = Label), position = position_stack(vjust = 0.5), size = 3.5) +
            labs(title = "Upregulated Proteins: RNA status") +
            theme_void(base_size = 12) +
            theme(legend.position = "bottom")

        print(p)
    } else {
        cat("*No significantly upregulated proteins found.*\n")
    }
}, error = function(e) cat(sprintf("*Pie chart (Up) could not be generated: %s*\n", e$message)))
```

```{r conc-pie-up-legend, eval=has_conc_viz, results='asis'}
figure_legend("Donut chart showing the RNA change status for significantly upregulated proteins (padj < 0.05). Green = RNA also upregulated, Red = RNA downregulated, Gray = RNA not significantly changed.")
```

```{r conc-pie-down, eval=has_conc_viz, results='asis', fig.width=7, fig.height=7}
tryCatch({
    down_prot <- conc_df[conc_df$sig_prot_flag & conc_df$logFC_prot < 0, ]
    if (nrow(down_prot) > 0) {
        pie_data <- data.frame(table(down_prot$rna_change))
        colnames(pie_data) <- c("Category", "Count")
        pie_data$Pct <- round(100 * pie_data$Count / sum(pie_data$Count), 1)
        pie_data$Label <- paste0(pie_data$Category, "\n", pie_data$Count, " (", pie_data$Pct, "%)")
        pie_colors <- c("Up" = "green3", "Down" = "red3", "Not sig" = "gray70")

        p <- ggplot(pie_data, aes(x = 2, y = Count, fill = Category)) +
            geom_bar(stat = "identity", width = 1, color = "white") +
            coord_polar("y", start = 0) +
            scale_fill_manual(values = pie_colors) +
            xlim(0.5, 2.5) +
            geom_text(aes(label = Label), position = position_stack(vjust = 0.5), size = 3.5) +
            labs(title = "Downregulated Proteins: RNA status") +
            theme_void(base_size = 12) +
            theme(legend.position = "bottom")

        print(p)
    } else {
        cat("*No significantly downregulated proteins found.*\n")
    }
}, error = function(e) cat(sprintf("*Pie chart (Down) could not be generated: %s*\n", e$message)))
```

```{r conc-pie-down-legend, eval=has_conc_viz, results='asis'}
figure_legend("Donut chart showing the RNA change status for significantly downregulated proteins (padj < 0.05). Red = RNA also downregulated, Green = RNA upregulated, Gray = RNA not significantly changed.")
```

### Stacked Bar

```{r conc-stacked-bar, eval=has_conc_viz, results='asis', fig.width=8, fig.height=5}
tryCatch({
    # Classify each gene into concordance categories
    conc_df$conc_category <- ifelse(
        conc_df$sig_prot_flag & conc_df$sig_rna_flag &
            sign(conc_df$logFC_prot) == sign(conc_df$logFC_rna), "Concordant",
    ifelse(
        conc_df$sig_prot_flag & conc_df$sig_rna_flag &
            sign(conc_df$logFC_prot) != sign(conc_df$logFC_rna), "Discordant",
    ifelse(
        conc_df$sig_prot_flag & !conc_df$sig_rna_flag, "Protein only",
    ifelse(
        !conc_df$sig_prot_flag & conc_df$sig_rna_flag, "RNA only",
        "Not sig"))))

    bar_data <- data.frame(table(conc_df$conc_category))
    colnames(bar_data) <- c("Category", "Count")
    bar_data$Category <- factor(bar_data$Category,
        levels = c("Concordant", "Discordant", "Protein only", "RNA only", "Not sig"))
    bar_colors <- c("Concordant" = "green4", "Discordant" = "red3",
                    "Protein only" = "orange", "RNA only" = "steelblue", "Not sig" = "gray70")

    p <- ggplot(bar_data, aes(x = "All Features", y = Count, fill = Category)) +
        geom_bar(stat = "identity", width = 0.6, color = "white") +
        geom_text(aes(label = Count), position = position_stack(vjust = 0.5), size = 3.5) +
        scale_fill_manual(values = bar_colors) +
        labs(y = "Number of features", x = NULL,
             title = "RNA-Protein Concordance Summary") +
        theme_minimal(base_size = 12) +
        theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

    print(p)
}, error = function(e) cat(sprintf("*Stacked bar could not be generated: %s*\n", e$message)))
```

```{r conc-stacked-bar-legend, eval=has_conc_viz, results='asis'}
figure_legend("Stacked bar chart summarizing RNA-protein concordance across all features. Concordant = both significant in the same direction; Discordant = both significant in opposite directions; Protein only = protein significant but RNA not; RNA only = RNA significant but protein not; Not sig = neither significant.")
```

### Volcano Overlay {.tabset}

```{r conc-volcano-prot, eval=has_conc_viz, results='asis', fig.width=9, fig.height=7}
tryCatch({
    vdf <- conc_df[!is.na(conc_df$padj_prot) & conc_df$padj_prot > 0, ]
    if (nrow(vdf) > 0) {
        vdf$neg_log10_padj <- -log10(vdf$padj_prot)
        prot_colors <- c("Up" = "springgreen4", "Down" = "violetred", "Not sig" = "gray70")

        p <- ggplot(vdf, aes(x = logFC_prot, y = neg_log10_padj, color = prot_trend)) +
            geom_point(alpha = 0.6, size = 1.5) +
            scale_color_manual(values = prot_colors, name = "Protein Trend") +
            geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
            labs(x = "Protein log2FC", y = "-log10(padj)",
                 title = "Protein Volcano (colored by protein significance)") +
            theme_minimal(base_size = 12)

        print(p)
    }
}, error = function(e) cat(sprintf("*Volcano (protein) could not be generated: %s*\n", e$message)))
```

```{r conc-volcano-prot-legend, eval=has_conc_viz, results='asis'}
figure_legend("Standard protein volcano plot colored by protein significance status. Dashed line indicates padj = 0.05 threshold.")
```

```{r conc-volcano-rna-overlay, eval=has_conc_viz, results='asis', fig.width=9, fig.height=7}
tryCatch({
    vdf <- conc_df[!is.na(conc_df$padj_prot) & conc_df$padj_prot > 0, ]
    if (nrow(vdf) > 0) {
        vdf$neg_log10_padj <- -log10(vdf$padj_prot)
        rna_colors <- c("Up" = "lightgreen", "Down" = "lightcoral", "Not sig" = "gray70")

        p <- ggplot(vdf, aes(x = logFC_prot, y = neg_log10_padj, color = rna_change)) +
            geom_point(alpha = 0.6, size = 1.5) +
            scale_color_manual(values = rna_colors, name = "RNA Change") +
            geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
            labs(x = "Protein log2FC", y = "-log10(padj)",
                 title = "Protein Volcano (colored by RNA change status)") +
            theme_minimal(base_size = 12)

        print(p)
    }
}, error = function(e) cat(sprintf("*Volcano (RNA overlay) could not be generated: %s*\n", e$message)))
```

```{r conc-volcano-rna-legend, eval=has_conc_viz, results='asis'}
figure_legend("Protein volcano plot with points colored by RNA change status. This overlay reveals whether protein-level changes are supported by corresponding transcript-level changes.")
```

### Venn Diagram

```{r conc-venn, eval=has_conc_viz, results='asis', fig.width=7, fig.height=6}
tryCatch({
    gene_col <- intersect(c("gene_symbol", "gene_id", "protein_id"), colnames(conc_df))[1]
    sig_rna_genes  <- conc_df[[gene_col]][conc_df$sig_rna_flag]
    sig_prot_genes <- conc_df[[gene_col]][conc_df$sig_prot_flag]

    # Try ggVennDiagram first, fall back to basic ggplot
    if (requireNamespace("ggVennDiagram", quietly = TRUE)) {
        gene_list <- list("Significant RNA" = sig_rna_genes,
                          "Significant Protein" = sig_prot_genes)
        p <- ggVennDiagram::ggVennDiagram(gene_list, label_alpha = 0) +
            scale_fill_gradient(low = "white", high = "steelblue") +
            labs(title = "Overlap of Significant Features") +
            theme(legend.position = "none")
        print(p)
    } else {
        # Simple text-based summary as fallback
        n_rna_only  <- length(setdiff(sig_rna_genes, sig_prot_genes))
        n_prot_only <- length(setdiff(sig_prot_genes, sig_rna_genes))
        n_both      <- length(intersect(sig_rna_genes, sig_prot_genes))
        venn_df <- data.frame(
            Region = c("RNA only", "Both", "Protein only"),
            Count  = c(n_rna_only, n_both, n_prot_only)
        )
        p <- ggplot(venn_df, aes(x = Region, y = Count, fill = Region)) +
            geom_col(width = 0.6) +
            geom_text(aes(label = Count), vjust = -0.3, size = 5) +
            scale_fill_manual(values = c("RNA only" = "steelblue", "Both" = "purple3",
                                          "Protein only" = "orange")) +
            labs(title = "Overlap of Significant Features (padj < 0.05)",
                 y = "Number of genes") +
            theme_minimal(base_size = 12) +
            theme(legend.position = "none")
        print(p)
    }
}, error = function(e) cat(sprintf("*Venn diagram could not be generated: %s*\n", e$message)))
```

```{r conc-venn-legend, eval=has_conc_viz, results='asis'}
figure_legend("Venn diagram showing the overlap between genes with significant RNA changes and genes with significant protein changes (padj < 0.05). The intersection highlights genes regulated at both levels.")
```

### Correlation Scatter

```{r conc-scatter, eval=has_conc_viz, results='asis', fig.width=8, fig.height=7}
tryCatch({
    sdf <- conc_df[!is.na(conc_df$logFC_rna) & !is.na(conc_df$logFC_prot), ]
    if (nrow(sdf) > 2) {
        r_val <- cor(sdf$logFC_rna, sdf$logFC_prot, use = "complete.obs")

        # Concordance status for coloring
        sdf$conc_status <- ifelse(
            sdf$sig_prot_flag & sdf$sig_rna_flag &
                sign(sdf$logFC_prot) == sign(sdf$logFC_rna), "Concordant",
        ifelse(
            sdf$sig_prot_flag & sdf$sig_rna_flag &
                sign(sdf$logFC_prot) != sign(sdf$logFC_rna), "Discordant",
        ifelse(
            sdf$sig_prot_flag | sdf$sig_rna_flag, "One sig only",
            "Not sig")))

        scatter_colors <- c("Concordant" = "green4", "Discordant" = "red3",
                            "One sig only" = "orange", "Not sig" = "gray70")

        p <- ggplot(sdf, aes(x = logFC_rna, y = logFC_prot, color = conc_status)) +
            geom_point(alpha = 0.5, size = 1.5) +
            geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.7, linetype = "dashed") +
            scale_color_manual(values = scatter_colors, name = "Status") +
            annotate("text", x = Inf, y = Inf, label = sprintf("Pearson r = %.3f", r_val),
                     hjust = 1.1, vjust = 1.5, size = 5, fontface = "bold") +
            labs(x = "RNA log2FC", y = "Protein log2FC",
                 title = "RNA vs Protein Fold-Change Correlation") +
            theme_minimal(base_size = 12)

        print(p)
    } else {
        cat("*Insufficient data for correlation scatter.*\n")
    }
}, error = function(e) cat(sprintf("*Correlation scatter could not be generated: %s*\n", e$message)))
```

```{r conc-scatter-legend, eval=has_conc_viz, results='asis'}
figure_legend("Scatter plot of RNA log2 fold-change vs protein log2 fold-change with Pearson correlation. Points are colored by concordance status. The dashed line shows the linear regression fit.")
```

### mHG Enrichment Test

```{r conc-mhg, eval=has_conc_viz, results='asis'}
tryCatch({
    if (requireNamespace("mHG", quietly = TRUE)) {
        library(mHG)

        # --- Upregulated proteins: is RNA sig enriched at the top? ---
        up_df <- conc_df[conc_df$sig_prot_flag & conc_df$logFC_prot > 0, ]
        up_df <- up_df[order(-up_df$logFC_prot), ]
        up_binary <- as.integer(up_df$sig_rna_flag)

        mhg_up <- NULL
        if (length(up_binary) > 2 && sum(up_binary) > 0) {
            mhg_up <- tryCatch(mHG::mHG.test(up_binary), error = function(e) NULL)
        }

        # --- Downregulated proteins: is RNA sig enriched at the top? ---
        down_df <- conc_df[conc_df$sig_prot_flag & conc_df$logFC_prot < 0, ]
        down_df <- down_df[order(down_df$logFC_prot), ]
        down_binary <- as.integer(down_df$sig_rna_flag)

        mhg_down <- NULL
        if (length(down_binary) > 2 && sum(down_binary) > 0) {
            mhg_down <- tryCatch(mHG::mHG.test(down_binary), error = function(e) NULL)
        }

        # Build results table
        results <- data.frame(
            Direction = c("Upregulated proteins", "Downregulated proteins"),
            N_proteins = c(nrow(up_df), nrow(down_df)),
            N_RNA_sig = c(sum(up_binary), sum(down_binary)),
            mHG_stat = c(
                if (!is.null(mhg_up)) sprintf("%.4e", mhg_up$mHG) else "N/A",
                if (!is.null(mhg_down)) sprintf("%.4e", mhg_down$mHG) else "N/A"
            ),
            mHG_pvalue = c(
                if (!is.null(mhg_up)) sprintf("%.4e", mhg_up$p.value) else "N/A",
                if (!is.null(mhg_down)) sprintf("%.4e", mhg_down$p.value) else "N/A"
            ),
            stringsAsFactors = FALSE
        )

        cat("\n#### mHG Enrichment Results\n\n")
        cat("Testing whether RNA-significant genes are enriched among the most extreme protein fold-changes.\n\n")
        print(knitr::kable(results, align = "lcccc"))
    } else {
        cat("*mHG package is not available. Install with `BiocManager::install('mHG')` to enable this analysis.*\n")
    }
}, error = function(e) cat(sprintf("*mHG enrichment test could not be generated: %s*\n", e$message)))
```

```{r conc-mhg-legend, eval=has_conc_viz, results='asis'}
figure_legend("Minimum hypergeometric (mHG) enrichment test results. For upregulated proteins (sorted by descending logFC) and downregulated proteins (sorted by ascending logFC), the test assesses whether RNA-significant genes are enriched at the top of the ranked list. A significant p-value indicates that protein fold-change magnitude is predictive of RNA significance.")
```

---

# MultiGSEA Plots {.tabset}

```{r multigsea-note, eval=!has_multigsea, results='asis'}
cat("*No MultiGSEA plots found.*\n")
```

## All Contrasts (Combined) {.unnumbered}

```{r multigsea-plots, eval=has_multigsea, results='asis'}
gsea_pngs <- list.files(multigsea_dir, pattern = "^multigsea_.*\\.png$", full.names = TRUE)
for (gp in gsea_pngs) {
    pair_name <- gsub("multigsea_|\\.png$", "", basename(gp))
    display_name <- gsub("_", " ", pair_name)
    cat(sprintf("\n### %s {.unnumbered}\n\n", display_name))
    cat(sprintf("![%s](%s)\n\n", display_name, gp))
    figure_legend(sprintf("MultiGSEA enrichment correlation plot for %s. Each point is a pathway; axis values represent enrichment scores. Color indicates joint significance status.", display_name))
    render_commentary(paste0("multigsea_", pair_name))
}
```

## Per-Contrast MultiGSEA {.tabset .unnumbered}

```{r multigsea-per-contrast, eval=has_multigsea, results='asis'}
multigsea_per_contrast_dir <- file.path(multigsea_dir, "per_contrast")
if (dir.exists(multigsea_per_contrast_dir)) {
    contrast_dirs <- list.dirs(multigsea_per_contrast_dir, full.names = TRUE, recursive = FALSE)
    for (cdir in contrast_dirs) {
        cname <- basename(cdir)
        cat(sprintf("\n### %s {.unnumbered}\n\n", gsub("_", " ", cname)))

        cpngs <- list.files(cdir, pattern = "^multigsea_.*\\.png$", full.names = TRUE)
        if (length(cpngs) == 0) {
            cat("*No MultiGSEA plots for this contrast.*\n\n")
            next
        }
        for (cp in cpngs) {
            pair_name <- gsub("multigsea_|\\.png$", "", basename(cp))
            display_name <- gsub("_", " ", pair_name)
            chunk_id <- paste0("mg-pc-", gsub("[^a-zA-Z0-9]", "", cname), "-",
                               gsub("[^a-zA-Z0-9]", "", pair_name))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, cp
            ), quiet = TRUE))
            cat("\n\n")
        }
    }
} else {
    cat("*Per-contrast MultiGSEA results not available.*\n")
}
```

```{r multi-ora-setup, include=FALSE}
multi_ora_dir <- file.path(multigsea_dir, "multi_ora")
has_multi_ora <- dir.exists(multi_ora_dir) && length(list.files(multi_ora_dir, "\\.png$")) > 0
```

## Multi-ORA (Combined Cross-Omics ORA) {.tabset .unnumbered}

```{r multi-ora-note, eval=!has_multi_ora, results='asis'}
cat("*No Multi-ORA results found.*\n")
```

### All Contrasts (Pooled) {.unnumbered}

```{r multi-ora-plots, eval=has_multi_ora, results='asis'}
# Pooled barplot
pooled_bp <- file.path(multi_ora_dir, "multi_ora_pooled_barplot.png")
if (file.exists(pooled_bp)) {
    cat("\n#### Pooled Multi-ORA {.unnumbered}\n\n")
    cat(sprintf("![Pooled Multi-ORA](%s)\n\n", pooled_bp))
    figure_legend("Pooled multi-ORA barplot. Significant genes from all gene-based omics (transcriptomics, proteomics) are pooled into a single set and tested for over-representation in KEGG pathways using Fisher's exact test. Red dashed line marks padj = 0.05.")
}

# Dot plot
dotplot <- file.path(multi_ora_dir, "multi_ora_dotplot.png")
if (file.exists(dotplot)) {
    cat("\n#### Pooled vs Per-Omics Comparison {.unnumbered}\n\n")
    cat(sprintf("![Multi-ORA Dot Plot](%s)\n\n", dotplot))
    figure_legend("Multi-ORA dot plot comparing per-omics and pooled enrichment results. Each dot represents a pathway; x-axis is -log10(padj), dot size reflects gene count, and color indicates the source (pooled or individual omics). Red dashed line marks padj = 0.05.")
}

# Omics support barplot
support_bp <- file.path(multi_ora_dir, "multi_ora_support_barplot.png")
if (file.exists(support_bp)) {
    cat("\n#### Omics Support per Pathway {.unnumbered}\n\n")
    cat(sprintf("![Multi-ORA Support](%s)\n\n", support_bp))
    figure_legend("Multi-ORA omics support barplot. Bar length shows pooled enrichment significance (-log10 p-value). Color indicates the number of individual omics layers that independently support each pathway (padj < 0.05).")
}

# Summary table
multi_ora_csv <- file.path(multi_ora_dir, "multi_ora_results.csv")
if (file.exists(multi_ora_csv)) {
    cat("\n#### Top Multi-ORA Pathways {.unnumbered}\n\n")
    mora_df <- read.csv(multi_ora_csv, stringsAsFactors = FALSE)
    mora_show <- mora_df[seq_len(min(25, nrow(mora_df))), ]
    display_cols <- intersect(c("pathway", "pooled_padj", "pooled_count", "metabolomics_padj", "n_omics_support"), colnames(mora_show))
    mora_show <- mora_show[, display_cols, drop = FALSE]
    col_labels <- c(pathway = "Pathway", pooled_padj = "Pooled padj", pooled_count = "Gene Count",
                    metabolomics_padj = "Metab padj", n_omics_support = "# Omics Supporting")
    colnames(mora_show) <- col_labels[display_cols]
    # Format numeric columns (signif handles NA gracefully)
    for (nc in intersect(c("Pooled padj", "Metab padj"), colnames(mora_show))) {
        mora_show[[nc]] <- signif(mora_show[[nc]], 3)
    }
    knitr::kable(mora_show, row.names = FALSE, caption = "Top multi-ORA enriched KEGG pathways (gene-based + metabolite compound ORA)")
}

render_commentary("multi_ora")
```

### Per-Contrast Multi-ORA {.tabset .unnumbered}

```{r multi-ora-per-contrast, eval=has_multi_ora, results='asis'}
mora_per_contrast_dir <- file.path(multi_ora_dir, "per_contrast")
if (dir.exists(mora_per_contrast_dir)) {
    mora_contrast_dirs <- list.dirs(mora_per_contrast_dir, full.names = TRUE, recursive = FALSE)
    for (cdir in mora_contrast_dirs) {
        cname <- basename(cdir)
        cat(sprintf("\n#### %s {.unnumbered}\n\n", gsub("_", " ", cname)))

        # Barplot
        c_bp <- file.path(cdir, "multi_ora_pooled_barplot.png")
        if (file.exists(c_bp)) {
            chunk_id <- paste0("mora-pc-bp-", gsub("[^a-zA-Z0-9]", "", cname))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, c_bp
            ), quiet = TRUE))
            cat("\n\n")
        }

        # Results table
        c_csv <- file.path(cdir, "multi_ora_results.csv")
        if (file.exists(c_csv)) {
            c_df <- read.csv(c_csv, stringsAsFactors = FALSE)
            c_show <- c_df[seq_len(min(20, nrow(c_df))), ]
            display_cols <- intersect(c("pathway", "pooled_padj", "pooled_count", "metabolomics_padj", "n_omics_support"), colnames(c_show))
            if (length(display_cols) > 0) {
                c_show <- c_show[, display_cols, drop = FALSE]
                cat(sprintf("\n**%s:** %d enriched pathways\n\n",
                            gsub("_", " ", cname), nrow(c_df)))
            }
        } else {
            cat("*No enriched pathways for this contrast.*\n\n")
        }
    }
} else {
    cat("*Per-contrast Multi-ORA results not available.*\n")
}
```

```{r metab-ora-setup, include=FALSE}
metab_ora_csv <- file.path(multi_ora_dir, "compound_ora_results.csv")
has_metab_ora <- has_multi_ora && file.exists(metab_ora_csv)
```

## Metabolomics Compound Enrichment {.unnumbered}

```{r metab-ora-note, eval=!has_metab_ora, results='asis'}
cat("*No metabolomics compound ORA results found. Compound enrichment requires metabolomics DE results with KEGG compound ID mappings.*\n")
```

```{r metab-ora-results, eval=has_metab_ora, results='asis'}
metab_df <- read.csv(metab_ora_csv, stringsAsFactors = FALSE)
if (nrow(metab_df) > 0) {
    cat("\nCompound-based over-representation analysis identified **", nrow(metab_df),
        "** enriched KEGG pathways from significantly altered metabolites.\n\n")
}
```

```{r metab-ora-barplot, eval=has_metab_ora, results='asis'}
metab_bp <- file.path(multi_ora_dir, "compound_ora_barplot.png")
if (file.exists(metab_bp)) {
    cat(sprintf("![Compound ORA](%s)\n\n", metab_bp))
    figure_legend("Metabolomics compound ORA barplot. Significant metabolites (KEGG compound IDs) are tested for over-representation in KEGG metabolic pathways using a hypergeometric test. Red dashed line marks p = 0.05.")
}
```

```{r metab-ora-table, eval=has_metab_ora}
metab_df <- read.csv(metab_ora_csv, stringsAsFactors = FALSE)
display_cols <- intersect(c("pathway", "ID", "pvalue", "padj", "setSize", "compounds"), colnames(metab_df))
metab_show <- metab_df[seq_len(min(30, nrow(metab_df))), display_cols, drop = FALSE]
col_labels <- c(pathway = "Pathway", ID = "KEGG ID", pvalue = "P-value",
                padj = "Adj. P-value", setSize = "Compound Count", compounds = "Compounds")
colnames(metab_show) <- col_labels[display_cols]
knitr::kable(metab_show, row.names = FALSE, caption = "Metabolomics compound ORA - enriched KEGG pathways")
```

```{r metab-ora-commentary, eval=has_metab_ora, results='asis'}
render_commentary("compound_ora")
```

```{r pathview-setup, include=FALSE}
pathview_pdf <- file.path(multi_ora_dir, "multi_ora_pathview_supported.pdf")
pathview_dir <- file.path(multi_ora_dir, "pathview")
has_pathview <- file.exists(pathview_pdf)
```

## Multi-Omics Pathway Maps (Pathview) {.unnumbered}

```{r pathview-note, eval=!has_pathview, results='asis'}
cat("*No pathview maps generated. Requires pathways enriched in >= 2 omics layers.*\n")
```

```{r pathview-plots, eval=has_pathview, results='asis'}
pv_pngs <- sort(list.files(pathview_dir, pattern = "\\.multi_ora\\.(multi\\.)?png$", full.names = TRUE))
cat("KEGG pathway maps for **", length(pv_pngs),
    "** pathways enriched in >= 2 omics layers. Gene nodes are colored by logFC ",
    "(transcriptomics and/or proteomics); compound nodes are colored by metabolomics logFC.\n\n")
cat(sprintf("[Download compiled PDF](%s)\n\n", pathview_pdf))
for (pv_png in pv_pngs) {
    pw_id <- sub("^[a-z]+", "", sub("\\.multi_ora\\.png$", "", basename(pv_png)))
    cat(sprintf("\n### Pathway %s {.unnumbered}\n\n", pw_id))
    cat(sprintf("![Pathway %s](%s)\n\n", pw_id, pv_png))
}
figure_legend("KEGG pathway maps generated with pathview. Gene nodes show logFC from transcriptomics (left) and proteomics (right) where available. Compound nodes show metabolomics logFC. Red = upregulated, green/blue = downregulated.")
render_commentary("multi_ora_pathview")
```

```{r per-omics-pathview-setup, include=FALSE}
pv_metab_pdf <- file.path(multi_ora_dir, "pathview_top_metabolomics_pathways.pdf")
pv_prot_pdf <- file.path(multi_ora_dir, "pathview_top_proteomics_pathways.pdf")
has_pv_metab <- file.exists(pv_metab_pdf)
has_pv_prot <- file.exists(pv_prot_pdf)
has_per_omics_pv <- has_pv_metab || has_pv_prot
```

```{r per-omics-pathview-metab, eval=has_pv_metab, results='asis'}
cat("\n## Top Metabolomics Pathways (with Proteomics Overlay) {.unnumbered}\n\n")
cat("Top enriched metabolomics KEGG pathways with compound logFC (metabolomics) and ",
    "enzyme/gene logFC (proteomics) overlaid on the same pathway map.\n\n")
cat(sprintf("[Download PDF](%s)\n\n", pv_metab_pdf))
metab_pngs <- sort(list.files(pathview_dir, pattern = "\\.metab_top\\.png$", full.names = TRUE))
for (pv_png in metab_pngs) {
    pw_id <- sub("^[a-z]+", "", sub("\\.metab_top\\.png$", "", basename(pv_png)))
    cat(sprintf("\n### Pathway %s {.unnumbered}\n\n", pw_id))
    cat(sprintf("![Pathway %s](%s)\n\n", pw_id, pv_png))
}
figure_legend("Top metabolomics-enriched KEGG pathways. Compound nodes show metabolomics logFC; gene/enzyme nodes show proteomics logFC. Red = upregulated, blue/green = downregulated.")
```

```{r per-omics-pathview-prot, eval=has_pv_prot, results='asis'}
cat("\n## Top Proteomics Pathways (with Metabolomics Overlay) {.unnumbered}\n\n")
cat("Top enriched proteomics KEGG pathways with enzyme/gene logFC (proteomics) and ",
    "compound logFC (metabolomics) overlaid on the same pathway map.\n\n")
cat(sprintf("[Download PDF](%s)\n\n", pv_prot_pdf))
prot_pngs <- sort(list.files(pathview_dir, pattern = "\\.prot_top\\.png$", full.names = TRUE))
for (pv_png in prot_pngs) {
    pw_id <- sub("^[a-z]+", "", sub("\\.prot_top\\.png$", "", basename(pv_png)))
    cat(sprintf("\n### Pathway %s {.unnumbered}\n\n", pw_id))
    cat(sprintf("![Pathway %s](%s)\n\n", pw_id, pv_png))
}
figure_legend("Top proteomics-enriched KEGG pathways. Gene/enzyme nodes show proteomics logFC; compound nodes show metabolomics logFC. Red = upregulated, blue/green = downregulated.")
```

---

# Cross-Omics Enrichment {.tabset}

```{r enrichment-note, eval=!has_enrichment, results='asis'}
cat("*No cross-omics enrichment results found. Enrichment analysis requires pathway results from at least 1 omics layer.*\n")
```

## All Contrasts (Combined) {.unnumbered}

```{r enrichment-per-omics, eval=has_enrichment, results='asis'}
# Show per-omics enrichment barplots
omics_barplots <- list.files(enrichment_dir, pattern = "_top_pathways\\.png$", full.names = TRUE)
if (length(omics_barplots) > 0) {
    # Deduplicate: if both "rna" and "transcriptomics" exist, keep "transcriptomics"
    bp_names <- gsub("_top_pathways\\.png$", "", basename(omics_barplots))
    if ("rna" %in% bp_names && "transcriptomics" %in% bp_names) {
        omics_barplots <- omics_barplots[bp_names != "rna"]
    }
    cat("\n### Per-Omics Pathway Enrichment {.unnumbered}\n\n")
    for (i in seq_along(omics_barplots)) {
        bp <- omics_barplots[i]
        omics_name <- gsub("_top_pathways\\.png$", "", basename(bp))
        cat(sprintf("\n#### %s {.unnumbered}\n\n", tools::toTitleCase(omics_name)))
        chunk_id <- paste0("enrich-bp-", i)
        cat(knitr::knit_child(text = sprintf(
            '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            chunk_id, bp
        ), quiet = TRUE))
        cat("\n\n")
    }
}
```

```{r enrichment-per-omics-legend, eval=has_enrichment, results='asis'}
figure_legend("Per-omics KEGG pathway enrichment (all contrasts combined). Bar length represents -log10(p-value). Red dashed line indicates p = 0.05 significance threshold.")
```

```{r enrichment-per-omics-table, eval=has_enrichment}
# Show enrichment tables for each omics
csv_files <- list.files(enrichment_dir, pattern = "_enriched_pathways\\.csv$", full.names = TRUE)
for (cf in csv_files) {
    omics_name <- gsub("_enriched_pathways\\.csv$", "", basename(cf))
    tbl <- read.csv(cf, stringsAsFactors = FALSE)
    if (nrow(tbl) > 0) {
        print(make_dt(
            head(tbl, 200),
            caption = paste0(tools::toTitleCase(omics_name), " Enriched KEGG Pathways"),
            pageLength = 15
        ))
    }
}
```

```{r enrichment-heatmap, eval=has_enrichment, results='asis'}
p <- file.path(enrichment_dir, "cross_omics_pathway_heatmap.png")
if (file.exists(p)) {
    cat("\n### Cross-Omics Pathway Comparison {.unnumbered}\n\n")
    cat(knitr::knit_child(text = sprintf(
        '```{r enrich-heatmap-img, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        p
    ), quiet = TRUE))
}
```

```{r enrichment-heatmap-legend, eval=has_enrichment, results='asis'}
p <- file.path(enrichment_dir, "cross_omics_pathway_heatmap.png")
if (file.exists(p)) {
    figure_legend("Cross-omics pathway enrichment heatmap (all contrasts). Showing the top 30 pathways ordered by Stouffer combined p-value; the complete list is in the table below. Color intensity represents -log10(p-value) for each pathway across omics layers.")
    render_commentary("cross_omics_pathway_heatmap")
}
```

```{r enrichment-top10, eval=has_enrichment, results='asis'}
meta_file <- file.path(enrichment_dir, "cross_omics_pathways_meta_analysis.csv")
if (file.exists(meta_file)) {
    meta_tbl <- read.csv(meta_file, stringsAsFactors = FALSE)
    # Top 10 sorted by n_omics desc then combined_pval asc
    top10 <- meta_tbl[order(-meta_tbl$n_omics, meta_tbl$combined_pval), ]
    top10 <- head(top10, 10)
    cat("\n### Top 10 Cross-Omics Pathways {.unnumbered}\n\n")
    knitr::kable(top10, row.names = FALSE, digits = 3,
                 caption = "Top 10 pathways prioritized by number of omics layers (n_omics) and combined p-value")
}
```

```{r enrichment-table, eval=has_enrichment}
meta_file <- file.path(enrichment_dir, "cross_omics_pathways_meta_analysis.csv")
if (file.exists(meta_file)) {
    meta_tbl <- read.csv(meta_file, stringsAsFactors = FALSE)
    make_dt(
        meta_tbl,
        caption = "Cross-Omics Pathway Meta-Analysis (combined p-values)",
        pageLength = 6
    )
}
```

```{r enrichment-table-legend, eval=has_enrichment, results='asis'}
meta_file <- file.path(enrichment_dir, "cross_omics_pathways_meta_analysis.csv")
if (file.exists(meta_file)) {
    figure_legend("Top pathways from cross-omics meta-analysis. P-values from individual omics layers were combined using Fisher's method. Pathways significant in multiple omics are most informative.")
    render_commentary("cross_omics_meta_analysis")
}
```

## Per-Contrast Enrichment {.tabset .unnumbered}

```{r enrichment-per-contrast, eval=has_enrichment, results='asis'}
enrich_per_contrast_dir <- file.path(enrichment_dir, "per_contrast")
if (dir.exists(enrich_per_contrast_dir)) {
    enrich_contrast_dirs <- list.dirs(enrich_per_contrast_dir, full.names = TRUE, recursive = FALSE)
    for (cdir in enrich_contrast_dirs) {
        cname <- basename(cdir)
        cat(sprintf("\n### %s {.unnumbered}\n\n", gsub("_", " ", cname)))

        # Per-omics barplots for this contrast
        c_barplots <- list.files(cdir, pattern = "_top_pathways\\.png$", full.names = TRUE)
        for (cbp in c_barplots) {
            omics_name <- gsub("_top_pathways\\.png$", "", basename(cbp))
            cat(sprintf("\n#### %s {.unnumbered}\n\n", tools::toTitleCase(omics_name)))
            chunk_id <- paste0("enr-pc-", gsub("[^a-zA-Z0-9]", "", cname), "-",
                               gsub("[^a-zA-Z0-9]", "", omics_name))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, cbp
            ), quiet = TRUE))
            cat("\n\n")
        }

        # Cross-omics heatmap for this contrast
        c_heatmap <- file.path(cdir, "cross_omics_pathway_heatmap.png")
        if (file.exists(c_heatmap)) {
            cat("\n#### Cross-Omics Heatmap {.unnumbered}\n\n")
            chunk_id <- paste0("enr-pc-hm-", gsub("[^a-zA-Z0-9]", "", cname))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, c_heatmap
            ), quiet = TRUE))
            cat("\n\n")
        }

        # Meta-analysis table for this contrast
        c_meta <- file.path(cdir, "cross_omics_pathways_meta_analysis.csv")
        if (file.exists(c_meta)) {
            c_meta_df <- read.csv(c_meta, stringsAsFactors = FALSE)
            cat(sprintf("\n**%s:** %d pathways in cross-omics meta-analysis\n\n",
                        gsub("_", " ", cname), nrow(c_meta_df)))
        }

        if (length(c_barplots) == 0 && !file.exists(c_heatmap)) {
            cat("*No enrichment results for this contrast.*\n\n")
        }
    }
} else {
    cat("*Per-contrast enrichment results not available.*\n")
}
```

---

# Data Harmonization

```{r harmonization-summary, results='asis'}
# Look for harmonization summary files
harmony_summary_file <- file.path(run_dir, "harmonization_summary.csv")
if (file.exists(harmony_summary_file)) {
    harmony_summary <- read.csv(harmony_summary_file, stringsAsFactors = FALSE)
    cat("## Harmonization Overview\n\n")
    kable(harmony_summary)
} else {
    cat("Multi-omics data were harmonized into a unified **MultiAssayExperiment** object. ",
        "Each omics layer was aligned to a common set of samples.\n\n")

    # Show omics layers
    omics_dirs <- list.dirs(integration_dir, full.names = FALSE, recursive = FALSE)
    if (length(omics_dirs) > 0) {
        cat("**Omics layers in integration:** ", paste(omics_dirs, collapse = ", "), "\n\n")
    }
}
```

```{r gene-protein-mapping, eval=!is.null(gene_protein_mapping), results='asis'}
cat("## Gene-Protein Mapping\n\n")
cat(sprintf("A total of **%d gene-protein pairs** were mapped across transcriptomics and proteomics layers.\n\n",
            nrow(gene_protein_mapping)))
```

```{r gene-protein-table, eval=!is.null(gene_protein_mapping) && nrow(gene_protein_mapping) > 0}
make_dt(
    head(gene_protein_mapping, 500),
    caption = "Gene-Protein Mapping (first 500 rows)",
    pageLength = 10
)
```

```{r harmonization-legend, eval=!is.null(gene_protein_mapping), results='asis'}
figure_legend("Gene-protein mapping table showing matched features between transcriptomics and proteomics layers.")
render_commentary("gene_protein_mapping")
```

---

# Sample Concordance Across Omics

```{r mantel-setup, include=FALSE}
foundational_dir <- file.path(run_dir, "foundational", "foundational")
mantel_heatmap <- file.path(foundational_dir, "plots", "mantel_test_heatmap.png")
mantel_csv <- file.path(foundational_dir, "tables", "mantel_test_results.csv")
clustering_consistency_plot <- file.path(foundational_dir, "plots", "sample_clustering_consistency.png")
condition_separation_plot <- file.path(foundational_dir, "plots", "sample_condition_separation.png")
has_mantel <- file.exists(mantel_heatmap)
has_clustering <- file.exists(clustering_consistency_plot)
has_separation <- file.exists(condition_separation_plot)

# Find per-omics correlation heatmaps
mantel_corr_heatmaps <- list.files(
    file.path(foundational_dir, "plots"),
    pattern = "^mantel_corr_heatmap_.*\\.png$",
    full.names = TRUE
)
has_corr_heatmaps <- length(mantel_corr_heatmaps) > 0
```

```{r clustering-consistency, eval=has_clustering, results='asis'}
cat("## Clustering Consistency\n\n")
knitr::include_graphics(clustering_consistency_plot)
figure_legend("Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) comparing unsupervised sample clusters between omics pairs. Values near 1 indicate identical clustering structure.")
```

```{r condition-separation, eval=has_separation, results='asis'}
cat("## Condition Separation\n\n")
knitr::include_graphics(condition_separation_plot)
figure_legend("Within-condition vs between-condition sample similarity per omics layer. Positive separation indicates that samples from the same experimental group are more similar than samples from different groups.")
```

```{r group-dist-setup, include=FALSE}
group_dist_scatter <- file.path(foundational_dir, "plots", "sample_within_vs_cross_group_distance.png")
group_dist_boxplot <- file.path(foundational_dir, "plots", "sample_group_distance_boxplot.png")
has_group_dist <- file.exists(group_dist_scatter) || file.exists(group_dist_boxplot)
```

```{r group-dist-section, eval=has_group_dist, results='asis'}
cat("## Per-Sample Within vs Cross-Group Distance\n\n")
cat("For each sample, the mean Euclidean distance to other samples in the same condition group (within-group) and to samples in different groups (cross-group) is computed per omics layer. Samples below the diagonal are closer to their own group; samples above the diagonal are ambiguous or misplaced.\n\n")
```

```{r group-dist-scatter, eval=file.exists(group_dist_scatter), results='asis'}
knitr::include_graphics(group_dist_scatter)
figure_legend("Per-sample within-group vs cross-group mean Euclidean distance, faceted by omics layer. Points below the diagonal line indicate samples that are closer to their own condition group. Labeled points (if any) are samples closer to the other group than their own, which may indicate outliers, mislabeling, or biological heterogeneity.")
render_commentary("sample_group_distances")
```

```{r group-dist-boxplot, eval=file.exists(group_dist_boxplot), results='asis'}
knitr::include_graphics(group_dist_boxplot)
figure_legend("Distribution of within-group and cross-group mean distances across all samples per omics layer. Clear separation between the two distributions indicates strong condition effect in that omics layer.")
```

```{r group-dist-table, eval=has_group_dist}
group_dist_csv <- file.path(foundational_dir, "tables", "sample_within_vs_cross_group_distances.csv")
if (file.exists(group_dist_csv)) {
    df <- read.csv(group_dist_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        df$within_group_dist <- round(df$within_group_dist, 2)
        df$cross_group_dist <- round(df$cross_group_dist, 2)
        df$dist_ratio <- round(df$dist_ratio, 3)
        make_dt(df, caption = "Per-sample within-group vs cross-group distances")
    }
}
```

---

# Integration Results {.tabset}

```{r integration-note, eval=!has_integration, results='asis'}
cat("*No integration results found. Integration methods may not have been configured or may have failed.*\n")
```

## DIABLO

```{r diablo-sample-plot, eval=has_diablo, results='asis'}
p <- file.path(diablo_dir, "diablo_sample_plot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r diablo-sample-legend, eval=has_diablo, results='asis'}
figure_legend("DIABLO sample scores plot. Samples are projected onto the first two components of the multi-omics discriminant analysis. Colors represent experimental groups.")
render_commentary("diablo_sample_plot")
```

```{r diablo-variable-plot, eval=has_diablo, results='asis'}
p <- file.path(diablo_dir, "diablo_variable_plot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r diablo-variable-legend, eval=has_diablo, results='asis'}
figure_legend("DIABLO variable loadings plot. Top contributing features from each omics layer are shown on the first two components.")
render_commentary("diablo_variable_plot")
```

### Circos Plot {.tabset .unnumbered}

#### Static

```{r diablo-circos-plot, eval=has_diablo}
p <- file.path(diablo_dir, "diablo_circos_plot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

#### Interactive

```{r diablo-circos-interactive, eval=has_diablo, results='asis'}
circos_csv <- file.path(diablo_dir, "diablo_circos_correlations.csv")
if (file.exists(circos_csv)) {
    edges <- read.csv(circos_csv, stringsAsFactors = FALSE)
    edges <- edges[order(abs(edges$weight), decreasing = TRUE), ]
    edges <- head(edges, 80)
    p <- plotly::plot_ly(
        x = edges$from, y = edges$to, z = edges$weight,
        text = paste0(edges$from, " [", edges$from_omics, "] \u2014 ",
                      edges$to, " [", edges$to_omics, "]",
                      "\nWeight: ", round(edges$weight, 3)),
        type = "heatmap", hoverinfo = "text",
        colorscale = list(c(0, "#f7fbff"), c(1, "#08306b"))
    ) %>%
        plotly::layout(
            title = "Cross-Omics Feature Correlations (DIABLO)",
            xaxis = list(title = "", tickangle = 45, tickfont = list(size = 8)),
            yaxis = list(title = "", tickfont = list(size = 8)),
            margin = list(b = 120, l = 120)
        )
    print(htmltools::tagList(p))
} else {
    cat("*Circos correlation CSV not available for interactive plot.*\n")
}
```

### Interactive Loadings Plot {.tabset .unnumbered}

#### Static

```{r diablo-variable-static2, eval=has_diablo}
vp <- file.path(diablo_dir, "diablo_variable_plot.png")
if (file.exists(vp)) {
    knitr::include_graphics(vp)
}
```

#### Interactive

```{r diablo-loadings-interactive, eval=has_diablo, results='asis'}
loadings_csv <- file.path(diablo_dir, "diablo_variable_loadings.csv")
if (file.exists(loadings_csv)) {
    ld <- read.csv(loadings_csv, stringsAsFactors = FALSE)
    omics_col <- intersect(c("view", "omics", "block"), colnames(ld))[1]
    if (all(c("comp1", "comp2", "feature") %in% colnames(ld)) && !is.na(omics_col)) {
        ld$omics_layer <- ld[[omics_col]]
        p <- plotly::plot_ly(
            data = ld,
            x = ~comp1, y = ~comp2,
            text = ~paste0(feature, " [", omics_layer, "]"),
            color = ~omics_layer,
            type = "scatter", mode = "markers",
            hoverinfo = "text",
            marker = list(size = 8, opacity = 0.7)
        ) %>%
            plotly::layout(
                title = "DIABLO Variable Loadings (Interactive)",
                xaxis = list(title = "Component 1"),
                yaxis = list(title = "Component 2")
            )
        print(htmltools::tagList(p))
    } else {
        cat("*Loadings CSV columns not in expected format.*\n")
    }
} else {
    cat("*Loadings CSV not available for interactive plot.*\n")
}
```

```{r diablo-loadings-interactive-legend, eval=has_diablo, results='asis'}
figure_legend("Interactive DIABLO loadings plot. Features are colored by omics type. Hover for feature details.")
```

```{r diablo-feature-heatmap, eval=has_diablo, results='asis'}
hm <- file.path(diablo_dir, "diablo_feature_heatmap.png")
if (file.exists(hm)) {
    cat("\n### Multi-Omics Feature Heatmap {.unnumbered}\n\n")
    cat(sprintf("![DIABLO Feature Heatmap](%s)\n\n", hm))
    figure_legend("Heatmap of top DIABLO features across omics layers. Values are z-scored per feature. Rows are grouped by omics type; columns are samples annotated by condition.")
    render_commentary("diablo_feature_heatmap")
}
```

```{r diablo-loadings-log2fc, eval=has_diablo, results='asis'}
lfc_plots <- list.files(diablo_dir, pattern = "diablo_loadings_log2fc_.*\\.png$", full.names = TRUE)
if (length(lfc_plots) > 0) {
    cat("\n### Loadings Colored by log2FC {.unnumbered}\n\n")
    for (lp in lfc_plots) {
        contrast_label <- gsub("diablo_loadings_log2fc_|\\.png$", "", basename(lp))
        contrast_label <- gsub("\\.", " ", contrast_label)
        cat(sprintf("#### %s {.unnumbered}\n\n", contrast_label))
        cat(sprintf("![DIABLO Loadings log2FC - %s](%s)\n\n", contrast_label, lp))
    }
    figure_legend("DIABLO loadings (Component 1 vs Component 2) colored by log2 fold-change from differential expression. Red indicates upregulation, blue indicates downregulation. Faceted by omics layer; top features are labeled.")
    render_commentary("diablo_loadings_log2fc")
}
```

```{r diablo-top-features, eval=has_diablo}
# Load top feature tables
diablo_feat_files <- list.files(diablo_dir, pattern = "diablo_top_features_.*\\.csv$", full.names = TRUE)
if (length(diablo_feat_files) > 0) {
    for (ff in diablo_feat_files) {
        omics_name <- gsub("diablo_top_features_|\\.csv$", "", basename(ff))
        df <- tryCatch(read.csv(ff, stringsAsFactors = FALSE), error = function(e) NULL)
        if (!is.null(df) && nrow(df) > 0) {
            cat(sprintf("\n### Top DIABLO Features: %s\n\n", omics_name))
            print(make_dt(df, caption = paste("DIABLO top features -", omics_name)))
        }
    }
}
```

## SNF

```{r snf-heatmap, eval=has_snf, results='asis'}
p <- file.path(snf_dir, "snf_network_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r snf-heatmap-legend, eval=has_snf, results='asis'}
figure_legend("SNF fused similarity network heatmap. Rows and columns are samples; color intensity represents similarity across all omics layers after network fusion.")
render_commentary("snf_network_heatmap")
```

```{r snf-silhouette, eval=has_snf, results='asis'}
p <- file.path(snf_dir, "snf_silhouette.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r snf-silhouette-legend, eval=has_snf, results='asis'}
figure_legend("SNF cluster silhouette plot. Silhouette width quantifies how well each sample fits its assigned cluster. Values near 1 indicate good clustering; values near 0 or negative suggest misclassification.")
render_commentary("snf_silhouette")
```

```{r snf-clusters-table, eval=has_snf}
clusters_file <- file.path(snf_dir, "snf_clusters.csv")
if (file.exists(clusters_file)) {
    clusters <- read.csv(clusters_file, stringsAsFactors = FALSE)
    make_dt(clusters, caption = "SNF Cluster Assignments")
}
```

## MOFA2

```{r mofa-variance, eval=has_mofa, results='asis'}
p <- file.path(mofa_dir, "mofa_variance_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mofa-variance-legend, eval=has_mofa, results='asis'}
figure_legend("MOFA2 variance explained heatmap. Each cell shows the proportion of variance in a given omics view (rows) explained by a latent factor (columns). Factors capturing shared variation across multiple views are most informative.")
render_commentary("mofa_variance_heatmap")
```

```{r mofa-factors-tabs, eval=has_mofa, results='asis'}
p12 <- file.path(mofa_dir, "mofa_factors_1_2.png")
p12_labeled <- file.path(mofa_dir, "mofa_factors_1_2_labeled.png")
p13 <- file.path(mofa_dir, "mofa_factors_1_3.png")

has_f12 <- file.exists(p12)
has_f12_lab <- file.exists(p12_labeled)
has_f13 <- file.exists(p13)

if (has_f12 || has_f12_lab) {
    cat("\n### Factor 1 vs Factor 2 {.tabset .unnumbered}\n\n")

    if (has_f12) {
        cat("#### By Condition {.unnumbered}\n\n")
        cat(knitr::knit_child(text = sprintf(
            '```{r mofa-f12-cond, echo=FALSE}\nknitr::include_graphics("%s")\n```\n', p12
        ), quiet = TRUE))
        figure_legend("MOFA2 Factor 1 vs Factor 2 scatter plot. Each point is a sample, colored by experimental condition.")
    }

    if (has_f12_lab) {
        cat("\n#### With Sample Labels {.unnumbered}\n\n")
        cat(knitr::knit_child(text = sprintf(
            '```{r mofa-f12-lab, echo=FALSE}\nknitr::include_graphics("%s")\n```\n', p12_labeled
        ), quiet = TRUE))
        figure_legend("MOFA2 Factor 1 vs Factor 2 with sample labels. Separation along factors indicates biologically relevant latent dimensions.")
    }

    render_commentary("mofa_factors_1_2")
}

if (has_f13) {
    cat("\n### Factor 1 vs Factor 3 {.unnumbered}\n\n")
    cat(knitr::knit_child(text = sprintf(
        '```{r mofa-f13, echo=FALSE}\nknitr::include_graphics("%s")\n```\n', p13
    ), quiet = TRUE))
    figure_legend("MOFA2 Factor 1 vs Factor 3 scatter plot.")
    render_commentary("mofa_factors_1_3")
}
```

```{r mofa-weights, eval=has_mofa, results='asis'}
weight_files <- list.files(mofa_dir, pattern = "^mofa_top_weights_.*\\.png$", full.names = TRUE)
# Separate the gene-symbol proteomics plot from the rest
genesym_file <- grep("proteomics_genesymbol", weight_files, value = TRUE)
main_files <- setdiff(weight_files, genesym_file)

if (length(main_files) > 0) {
    for (wf in main_files) {
        view_name <- gsub("mofa_top_weights_|\\.png$", "", basename(wf))
        chunk_id <- paste0("mofa-wt-", gsub("[^a-zA-Z0-9]", "", view_name))

        # For proteomics: create a tabset with UniProt and Gene Symbol tabs
        if (view_name == "proteomics" && length(genesym_file) == 1) {
            cat(sprintf("\n### Top Weights: %s {.tabset .unnumbered}\n\n", view_name))
            cat("#### UniProt ID {.unnumbered}\n\n")
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, wf
            ), quiet = TRUE))
            figure_legend(sprintf("Top 20 features by absolute weight for MOFA Factor in the %s view (UniProt IDs).", view_name))

            cat("\n#### Gene Symbol {.unnumbered}\n\n")
            gs_id <- paste0(chunk_id, "gs")
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                gs_id, genesym_file
            ), quiet = TRUE))
            figure_legend(sprintf("Top 20 features by absolute weight for MOFA Factor in the %s view (Gene Symbols).", view_name))
        } else {
            cat(sprintf("\n### Top Weights: %s\n\n", view_name))
            cat(knitr::knit_child(text = sprintf(
                '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                chunk_id, wf
            ), quiet = TRUE))
            figure_legend(sprintf("Top 20 features by absolute weight for MOFA Factor in the %s view.", view_name))
        }

        fig_id <- paste0("mofa_top_weights_", view_name)
        render_commentary(fig_id)
    }
}
```

```{r mofa-loadings-log2fc, eval=has_mofa, results='asis'}
mofa_lfc_plots <- list.files(mofa_dir, pattern = "mofa_loadings_log2fc_.*\\.png$", full.names = TRUE)
if (length(mofa_lfc_plots) > 0) {
    cat("\n### Weights Colored by log2FC {.unnumbered}\n\n")
    for (lp in mofa_lfc_plots) {
        contrast_label <- gsub("mofa_loadings_log2fc_|\\.png$", "", basename(lp))
        contrast_label <- gsub("\\.", " ", contrast_label)
        cat(sprintf("#### %s {.unnumbered}\n\n", contrast_label))
        cat(sprintf("![MOFA2 Weights log2FC - %s](%s)\n\n", contrast_label, lp))
    }
    figure_legend("MOFA2 weights (Factor 1 vs Factor 2) colored by log2 fold-change from differential expression. Red indicates upregulation, blue indicates downregulation. Faceted by omics view; top features are labeled.")
    render_commentary("mofa_loadings_log2fc")
}
```

## Loadings-Based Pathway Enrichment {.tabset}

```{r loadings-enrich-note, eval=!has_loadings_enrich, results='asis'}
cat("*Loadings-based enrichment requires a standard organism annotation database (org.db). For non-model organisms like Giardia, loadings enrichment is not available. See the Cross-Omics Pathway Comparison section for GMT-based enrichment results.*\n")
```

```{r loadings-enrich-diablo, eval=has_loadings_enrich, results='asis'}
diablo_le_dir <- file.path(loadings_enrich_dir, "diablo_loadings")
diablo_le_plots <- list.files(diablo_le_dir, pattern = "\\.png$", full.names = TRUE)

if (length(diablo_le_plots) > 0) {
    cat("\n### DIABLO Loadings Enrichment\n\n")
    for (i in seq_along(diablo_le_plots)) {
        pf <- diablo_le_plots[i]
        label <- gsub("_enrichment\\.png$", "", basename(pf))
        label <- gsub("^DIABLO_", "", label)
        cat(sprintf("\n#### %s {.unnumbered}\n\n", label))
        chunk_id <- paste0("diablo-le-", i)
        cat(knitr::knit_child(text = sprintf(
            '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            chunk_id, pf
        ), quiet = TRUE))
        cat("\n\n")
    }
    figure_legend("KEGG pathway enrichment of top DIABLO loading features per component and omics layer. Red dashed line indicates p = 0.05.")
}

# Show combined table
diablo_le_csv <- file.path(diablo_le_dir, "diablo_loadings_enrichment_all.csv")
if (file.exists(diablo_le_csv)) {
    df <- read.csv(diablo_le_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        cat("\n#### DIABLO Loadings Enrichment Table {.unnumbered}\n\n")
        print(make_dt(df, caption = "DIABLO loadings KEGG enrichment"))
    }
}
```

```{r loadings-enrich-mofa, eval=has_loadings_enrich, results='asis'}
mofa_le_dir <- file.path(loadings_enrich_dir, "mofa_loadings")
mofa_le_plots <- list.files(mofa_le_dir, pattern = "\\.png$", full.names = TRUE)

if (length(mofa_le_plots) > 0) {
    cat("\n### MOFA2 Weights Enrichment\n\n")
    for (i in seq_along(mofa_le_plots)) {
        pf <- mofa_le_plots[i]
        label <- gsub("_enrichment\\.png$", "", basename(pf))
        label <- gsub("^MOFA_", "", label)
        cat(sprintf("\n#### %s {.unnumbered}\n\n", label))
        chunk_id <- paste0("mofa-le-", i)
        cat(knitr::knit_child(text = sprintf(
            '```{r %s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
            chunk_id, pf
        ), quiet = TRUE))
        cat("\n\n")
    }
    figure_legend("KEGG pathway enrichment of top MOFA2 weight features per factor and omics view. Red dashed line indicates p = 0.05.")
}

# Show combined table
mofa_le_csv <- file.path(mofa_le_dir, "mofa_weights_enrichment_all.csv")
if (file.exists(mofa_le_csv)) {
    df <- read.csv(mofa_le_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        cat("\n#### MOFA2 Weights Enrichment Table {.unnumbered}\n\n")
        print(make_dt(df, caption = "MOFA2 weights KEGG enrichment"))
    }
}
```

---

# Mechanistic & Causal Inference {.tabset}

```{r mech-setup, include=FALSE}
mechanistic_dir <- file.path(run_dir, "mechanistic", "mechanistic")
has_mechanistic <- dir.exists(mechanistic_dir) &&
    (length(list.files(file.path(mechanistic_dir, "plots"), "\\.png$")) > 0 ||
     length(list.files(file.path(mechanistic_dir, "tables"), "\\.csv$")) > 0)
```

```{r mech-note, eval=!has_mechanistic, results='asis'}
cat("*Mechanistic inference was not run or produced no results. Enable in config: `mechanistic.run_mechanistic: true`*\n")
```

## Mediation: RNA -> Protein -> Phenotype

```{r mech-mediation-prop, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_mediation_proportion_barplot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
    figure_legend("Proportion of the RNA effect on phenotype that is mediated through protein levels. Red dashed line marks 10% mediation threshold. Results are hypothesis-generating — the sequential ignorability assumption (no unmeasured confounders) is untestable.")
    render_commentary("mediation_proportion")
} else {
    cat("*Mediation analysis requires a continuous phenotype/outcome variable in the sample metadata. Not available for this dataset (binary condition only).*\n\n")
}
```

```{r mech-mediation-decomp, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_mediation_effect_decomposition.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
    figure_legend("Decomposition of total effect into ACME (Average Causal Mediation Effect, the indirect/mediated path) and ADE (Average Direct Effect, not through protein).")
}
```

```{r mech-mediation-table, eval=has_mechanistic}
med_csv <- file.path(mechanistic_dir, "tables", "mech_mediation_analysis_results.csv")
if (file.exists(med_csv)) {
    df <- read.csv(med_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        # Flag sensitivity robustness
        if ("rho_at_acme_zero" %in% colnames(df)) {
            df$sensitivity <- ifelse(
                is.na(df$rho_at_acme_zero), "N/A",
                ifelse(is.infinite(df$rho_at_acme_zero), "Robust (ACME never zero)",
                ifelse(abs(df$rho_at_acme_zero) > 0.3, "Robust",
                ifelse(abs(df$rho_at_acme_zero) > 0.1, "Moderate", "Fragile")))
            )
        }
        make_dt(df, caption = "Mediation analysis: RNA -> Protein -> Phenotype")
    }
}
```

```{r mech-mediation-sens-legend, eval=has_mechanistic, results='asis'}
med_csv <- file.path(mechanistic_dir, "tables", "mech_mediation_analysis_results.csv")
if (file.exists(med_csv)) {
    df <- read.csv(med_csv, stringsAsFactors = FALSE)
    if ("rho_at_acme_zero" %in% colnames(df)) {
        cat("\n**Sensitivity interpretation (rho\\_at\\_acme\\_zero):** The rho value at which the mediation effect (ACME) would become zero under unmeasured confounding. |rho| > 0.3 = robust; |rho| < 0.1 = fragile (easily explained by confounders).\n\n")
    }
}
```

## Protein -> Metabolite -> Phenotype Mediation

```{r mech-prot-metab-med-plot, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_prot_metab_mediation_barplot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
    figure_legend("Protein -> Metabolite -> Phenotype mediation. Tests whether each protein's effect on the phenotype is mediated through its best-correlated metabolite. Does not require RNA-seq — uses proteomics + metabolomics directly. Red dashed line marks 10% mediation threshold. Results are hypothesis-generating.")
    render_commentary("prot_metab_mediation")
} else {
    cat("*Protein-metabolite mediation requires a continuous phenotype/outcome variable. Not available for binary conditions.*\n\n")
}
```

```{r mech-prot-metab-med-table, eval=has_mechanistic}
pm_csv <- file.path(mechanistic_dir, "tables", "mech_prot_metab_mediation_results.csv")
if (file.exists(pm_csv)) {
    df <- read.csv(pm_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        if ("rho_at_acme_zero" %in% colnames(df)) {
            df$sensitivity <- ifelse(
                is.na(df$rho_at_acme_zero), "N/A",
                ifelse(is.infinite(df$rho_at_acme_zero), "Robust (ACME never zero)",
                ifelse(abs(df$rho_at_acme_zero) > 0.3, "Robust",
                ifelse(abs(df$rho_at_acme_zero) > 0.1, "Moderate", "Fragile")))
            )
        }
        make_dt(df, caption = "Protein -> Metabolite -> Phenotype mediation")
    }
}
```

## Protein-Metabolite Correlations

```{r mech-prot-metab-heatmap, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "protein_metabolite_correlation_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
    figure_legend("Cross-omics Spearman correlation heatmap between significant/variable proteins and metabolites. Blue = negative correlation, red = positive. Significant pairs (padj < 0.2, |r| > 0.7) are used for downstream network analysis.")
    render_commentary("protein_metabolite_correlations")
}
```

```{r mech-prot-metab-table, eval=has_mechanistic}
pm_pairs <- file.path(mechanistic_dir, "tables", "protein_metabolite_significant_pairs.csv")
if (file.exists(pm_pairs)) {
    df <- read.csv(pm_pairs, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        # Resolve HMDB IDs to metabolite names
        if ("metabolite" %in% colnames(df)) {
            proj_dir <- cfg$project$dir %||% "/home/multiomics-core"
            hmdb_lookup <- file.path(proj_dir, "data", "hmdb_compound_names.tsv")
            if (file.exists(hmdb_lookup)) {
                hmdb_db <- tryCatch(read.delim(hmdb_lookup, stringsAsFactors = FALSE),
                                    error = function(e) NULL)
                if (!is.null(hmdb_db) && all(c("HMDB", "Name") %in% colnames(hmdb_db))) {
                    idx <- match(df$metabolite, hmdb_db$HMDB)
                    resolved <- hmdb_db$Name[idx]
                    has_name <- !is.na(resolved) & resolved != ""
                    df$metabolite_name <- ifelse(has_name, resolved, df$metabolite)
                }
            }
        }
        # Reorder columns: annotation columns first, drop abs_cor (redundant)
        display_cols <- c("protein", "gene_symbol", "metabolite_name", "metabolite",
                          "correlation", "padj", "ec_number", "enzyme_metabolite_link",
                          "is_transcription_factor")
        display_cols <- intersect(display_cols, colnames(df))
        df_display <- df[, display_cols, drop = FALSE]
        make_dt(df_display,
                caption = "Significant protein-metabolite pairs (padj < 0.2, |r| > 0.7). EC = enzyme commission number; enzyme_metabolite_link = known KEGG enzyme-substrate/product relationship; is_transcription_factor = dorothea TF database.")
    }
} else {
    cat("*No significant protein-metabolite pairs found at the current thresholds.*\n\n")
}
```

## Multi-Layer Mediation: Protein -> Metabolite -> Phenotype

```{r mech-multilayer-plot, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_multilayer_mediation_barplot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
    figure_legend("Multi-layer mediation testing whether protein effects on phenotype are mediated through metabolite levels. Each bar shows the proportion of the protein effect carried through the most correlated metabolite. This extends causal inference across all three omics layers but compounds confounding assumptions — interpret as hypothesis-generating.")
    render_commentary("multilayer_mediation")
} else {
    cat("*Multi-layer mediation requires a continuous phenotype/outcome variable. Not available for binary conditions.*\n\n")
}
```

```{r mech-multilayer-table, eval=has_mechanistic}
ml_csv <- file.path(mechanistic_dir, "tables", "mech_multilayer_mediation_results.csv")
if (file.exists(ml_csv)) {
    df <- read.csv(ml_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        make_dt(df, caption = "Multi-layer mediation: Protein -> Metabolite -> Phenotype")
    }
}
```

## COSMOS Causal Network

```{r cosmos-network, eval=has_mechanistic, results='asis'}
cosmos_plot <- file.path(mechanistic_dir, "plots", "cosmos_causal_network.png")
if (file.exists(cosmos_plot)) {
    cat("\n### Signaling-to-Metabolism Network {.unnumbered}\n\n")
    knitr::include_graphics(cosmos_plot)
    figure_legend("COSMOS causal network connecting deregulated transcription factors to metabolites through known signaling and metabolic interactions (OmniPath + STITCH + Recon3D prior knowledge). Red edges = activating interactions; blue = inhibiting. Node size reflects degree centrality. This network represents the smallest coherent subnetwork explaining the observed molecular changes, constrained by prior biological knowledge.")
    render_commentary("cosmos_network")
} else {
    cat("*COSMOS causal network was not generated. Requires: cosmosR, CARNIVAL, and decoupleR packages.*\n\n")
}
```

```{r cosmos-tables, eval=has_mechanistic}
cosmos_edges <- file.path(mechanistic_dir, "tables", "cosmos_network_edges.csv")
cosmos_nodes <- file.path(mechanistic_dir, "tables", "cosmos_network_nodes.csv")

if (file.exists(cosmos_nodes)) {
    df <- read.csv(cosmos_nodes, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        make_dt(df, caption = "COSMOS network nodes")
    }
}

if (file.exists(cosmos_edges)) {
    df <- read.csv(cosmos_edges, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        make_dt(df, caption = "COSMOS network edges")
    }
}
```

## TF Activity

```{r mech-tf-heatmap, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_tf_activity_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
    figure_legend("Transcription factor activity inferred from target gene expression using dorothea/decoupleR regulons. Rows are TFs; columns are samples. Activity is estimated as the weighted mean of target expression, accounting for mode of regulation.")
    render_commentary("tf_activity")
} else {
    cat("*TF activity inference requires dorothea/decoupleR regulons for the organism. Not available for non-model organisms (e.g., Giardia lamblia).*\n\n")
}
```

```{r mech-tf-barplot, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_tf_activity_barplot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
    figure_legend("Top differentially active transcription factors between conditions.")
}
```

## Pathway Activity (PROGENy)

```{r mech-pathway-heatmap, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_pathway_activity_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
    figure_legend("PROGENy pathway activity inferred from proteomics data using decoupleR weighted mean. Rows are pathways; columns are samples. PROGENy uses perturbation-derived pathway-gene weights to estimate signaling pathway activity from protein abundance.")
    render_commentary("pathway_activity")
} else {
    cat("*PROGENy pathway activity requires human/mouse gene annotations. Not available for non-model organisms.*\n\n")
}
```

```{r mech-pathway-barplot, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_pathway_activity_barplot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
    figure_legend("Pathway activity differences between conditions. Bar height represents -log10(p-value) from t-test comparing pathway activity scores between groups.")
}
```

```{r mech-pathway-table, eval=has_mechanistic}
pa_csv <- file.path(mechanistic_dir, "tables", "mech_pathway_activity_associations.csv")
if (file.exists(pa_csv)) {
    df <- read.csv(pa_csv, stringsAsFactors = FALSE)
    if (nrow(df) > 0) {
        make_dt(df, caption = "Pathway activity associations with condition")
    }
}
```

## Regulatory / Co-expression Network

```{r mech-network, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_regulatory_network.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
    figure_legend("Co-expression network inferred using GENIE3/ARACNe/correlation from the available omics data (gene expression or protein abundance). Edges represent predicted co-regulatory relationships (association-based, not causal). Hub nodes with many connections are highlighted in red.")
    render_commentary("regulatory_network")
} else {
    cat("*Regulatory network not generated for this dataset.*\n\n")
}
```

---

# Output Files

```{r output-files, results='asis'}
cat(sprintf("All multi-omics integration results have been saved to:\n\n`%s`\n\n", run_dir))

cat("**Subdirectories:**\n\n")

subdirs <- list.dirs(run_dir, full.names = FALSE, recursive = FALSE)
subdirs <- subdirs[subdirs != "execution_info"]
if (length(subdirs) > 0) {
    for (sd in sort(subdirs)) {
        n_files <- length(list.files(file.path(run_dir, sd), recursive = TRUE))
        cat(sprintf("- **%s/** (%d files)\n", sd, n_files))
    }
}
```

---

# Methods

```{r methods, results='asis'}
cat("## Multi-Omics Integration Pipeline\n\n")

cat("This report was generated by the **multiomics-core** pipeline. ",
    "The following analytical steps were performed:\n\n")

cat("1. **Data Harmonization**: Individual omics layers (preprocessed by their respective sub-pipelines) ",
    "were aligned to a common set of samples and assembled into a `MultiAssayExperiment` object.\n\n")

methods <- multi_cfg$integration$methods %||% c("DIABLO", "SNF")
step_num <- 2
if ("DIABLO" %in% methods) {
    cat(sprintf("%d. **DIABLO (Data Integration Analysis for Biomarker discovery using Latent cOmponents)**: ", step_num),
        "A supervised multi-omics integration method from the mixOmics package that identifies ",
        "discriminant features across omics layers using sparse PLS-DA.\n\n")
    step_num <- step_num + 1
}
if ("SNF" %in% methods) {
    cat(sprintf("%d. **SNF (Similarity Network Fusion)**: ", step_num),
        "An unsupervised method that constructs sample similarity networks for each omics layer ",
        "and fuses them into a single network for clustering.\n\n")
    step_num <- step_num + 1
}
if ("MOFA2" %in% methods) {
    cat(sprintf("%d. **MOFA2 (Multi-Omics Factor Analysis v2)**: ", step_num),
        "An unsupervised Bayesian factor model that identifies latent factors capturing ",
        "shared and omics-specific sources of variation.\n\n")
    step_num <- step_num + 1
}

cat(sprintf("%d. **Cross-Omics Enrichment**: Pathway enrichment results were aggregated across omics ", step_num),
    "using Fisher's method for combined p-values.\n\n")

cat("## Software\n\n")
cat("```\n")
cat(capture.output(sessionInfo()), sep = "\n")
cat("\n```\n")
```
