---
title: "Multi-Omics Integration Report"
subtitle: "multiomics-core Pipeline"
author: "Bioinformatician: `r yaml::read_yaml(file.path(normalizePath(dirname(knitr::current_input())), 'execution_info', 'config_used.yaml'))$project$analyst`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    fig_width: 10
    fig_height: 7
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%"
)

library(yaml)
library(knitr)
library(DT)
library(plotly)
library(ggplot2)
library(htmltools)
library(jsonlite)

# Helper function for downloadable DT tables
make_dt <- function(df, caption = NULL, pageLength = 15) {
    tbl <- DT::datatable(
        df,
        caption = caption,
        extensions = 'Buttons',
        options = list(
            scrollX = TRUE,
            pageLength = pageLength,
            dom = 'Bfrtip',
            buttons = list(
                list(extend = 'csv', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data")),
                list(extend = 'excel', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data"))
            )
        ),
        rownames = FALSE
    )
    num_cols <- names(df)[vapply(df, is.numeric, logical(1))]
    pval_pat <- "^(p[._]?val|padj|adj[._]?p|fdr|q[._]?val|BH|bonferroni|p\\.value|adj\\.P\\.Val|PValue|FDR)"
    pval_cols <- grep(pval_pat, num_cols, ignore.case = TRUE, value = TRUE)
    other_cols <- setdiff(num_cols, pval_cols)
    if (length(other_cols) > 0) tbl <- DT::formatRound(tbl, columns = other_cols, digits = 2)
    if (length(pval_cols) > 0) {
        pval_idx <- which(names(df) %in% pval_cols) - 1L
        sci_render <- DT::JS(
            "function(data, type, row) {",
            "  if (type === 'display' && data !== null && data !== '') {",
            "    return parseFloat(data).toExponential(2);",
            "  }",
            "  return data;",
            "}"
        )
        col_defs <- lapply(pval_idx, function(i) list(targets = i, render = sci_render))
        tbl$x$options$columnDefs <- c(tbl$x$options$columnDefs, col_defs)
    }
    tbl
}

# Helper: render figure legend (paper-style caption below figure)
figure_legend <- function(text) {
    cat(sprintf("<p class='figure-legend' style='text-align: center; font-size: 0.9em; color: #555; margin: 4px auto 12px auto; max-width: 88%%; line-height: 1.45; font-style: italic;'>%s</p>\n", text))
}

# Helper: render commentary block for a figure_id (collapsible)
render_commentary <- function(figure_id) {
    if (!has_commentary) return(invisible(NULL))
    c <- commentary[[figure_id]]
    if (is.null(c)) return(invisible(NULL))

    cat("<details class='commentary-details' style='margin: 4px 0 18px 0;'>\n")
    cat("<summary style='cursor: pointer; color: #4a90d9; font-weight: 500; font-size: 0.93em; padding: 3px 0;'>Show AI Interpretation</summary>\n")

    html <- "<div class='commentary-block' style='background: #f8f9fa; border-left: 4px solid #4a90d9; padding: 12px 16px; margin: 6px 0 4px 0; border-radius: 0 4px 4px 0; font-size: 0.92em;'>\n"
    html <- paste0(html, "<p style='margin: 0 0 8px 0;'><strong>What this shows:</strong> ",
                   htmltools::htmlEscape(c$what_is_this %||% ""), "</p>\n")

    bio_interp <- c$biological_interpretation %||% ""
    if (nzchar(bio_interp)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Interpretation:</strong> ",
                       htmltools::htmlEscape(bio_interp), "</p>\n")
    }

    obs <- c$observations
    if (length(obs) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Key observations:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (o in obs) html <- paste0(html, "<li>", htmltools::htmlEscape(o), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    stat_assess <- c$statistical_assessment %||% ""
    if (nzchar(stat_assess)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Statistical note:</strong> ",
                       htmltools::htmlEscape(stat_assess), "</p>\n")
    }

    issues <- c$issues_checks
    if (length(issues) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Checks:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (issue in issues) html <- paste0(html, "<li>", htmltools::htmlEscape(issue), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    steps <- c$next_steps
    if (length(steps) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Next steps:</strong></p>\n<ul style='margin: 2px 0 4px 0; padding-left: 20px;'>\n")
        for (s in steps) html <- paste0(html, "<li>", htmltools::htmlEscape(s), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    backend <- c$backend %||% "unknown"
    confidence <- c$confidence %||% ""
    backend_label <- if (grepl("claude", backend, ignore.case = TRUE)) {
        "Claude (Anthropic) + K-Dense AI scientific-skills"
    } else if (grepl("openai", backend, ignore.case = TRUE)) {
        "OpenAI + K-Dense AI scientific-skills"
    } else if (backend == "data_driven") {
        "data-driven + K-Dense AI scientific-skills"
    } else {
        backend
    }
    if (nzchar(confidence) && confidence != "none") {
        backend_label <- paste0(backend_label, " | confidence: ", confidence)
    }
    html <- paste0(html, "<p style='margin: 4px 0 0 0; color: #888; font-size: 0.85em;'><em>Generated by: ",
                   htmltools::htmlEscape(backend_label), "</em></p>\n</div>\n")

    cat(html)
    cat("</details>\n")
}

# Paths
run_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
config_path <- file.path(run_dir, "execution_info", "config_used.yaml")

# Multiomics output subdirectories
integration_dir  <- file.path(run_dir, "integration")
diablo_dir       <- file.path(integration_dir, "diablo")
snf_dir          <- file.path(integration_dir, "snf")
mofa_dir         <- file.path(integration_dir, "mofa")
enrichment_dir   <- file.path(run_dir, "cross_enrichment")
rna_prot_dir     <- file.path(run_dir, "rna_protein_correlation")
multigsea_dir    <- file.path(run_dir, "multigsea")
commentary_dir   <- file.path(run_dir, "commentary")

# New analysis directories
feature_network_dir  <- file.path(run_dir, "feature_network")
concordance_dir      <- file.path(run_dir, "concordance")
consensus_base_dir   <- file.path(run_dir, "consensus")
consensus_dir        <- file.path(consensus_base_dir, "consensus")
stability_dir        <- file.path(consensus_base_dir, "stability")
foundational_base_dir <- file.path(run_dir, "foundational")
foundational_dir     <- file.path(foundational_base_dir, "foundational")
mechanistic_base_dir <- file.path(run_dir, "mechanistic")
mechanistic_dir      <- file.path(mechanistic_base_dir, "mechanistic")

# Load commentary if available
commentary <- NULL
has_commentary <- FALSE
commentary_json <- file.path(commentary_dir, "all_commentary.json")
if (file.exists(commentary_json)) {
    commentary <- tryCatch(
        jsonlite::read_json(commentary_json),
        error = function(e) NULL
    )
    has_commentary <- !is.null(commentary) && length(commentary) > 0
}

# Load config
cfg <- yaml::read_yaml(config_path)
multi_cfg <- cfg$modes$multiomics

# Detect which sections have data
has_diablo      <- dir.exists(diablo_dir) && length(list.files(diablo_dir, "\\.png$")) > 0
has_snf         <- dir.exists(snf_dir) && length(list.files(snf_dir, "\\.png$")) > 0
has_mofa        <- dir.exists(mofa_dir) && length(list.files(mofa_dir, "\\.png$")) > 0
has_integration <- has_diablo || has_snf || has_mofa
has_enrichment  <- dir.exists(enrichment_dir) && length(list.files(enrichment_dir, "\\.png$|\\.csv$")) > 0
has_rna_prot    <- dir.exists(file.path(rna_prot_dir, "plots")) && length(list.files(file.path(rna_prot_dir, "plots"), "\\.png$")) > 0
has_multigsea   <- dir.exists(multigsea_dir) && length(list.files(multigsea_dir, "\\.png$")) > 0
has_feature_network <- dir.exists(file.path(feature_network_dir, "plots")) && length(list.files(file.path(feature_network_dir, "plots"), "\\.png$")) > 0
has_concordance     <- dir.exists(concordance_dir) && length(list.files(concordance_dir, "\\.png$|\\.csv$")) > 0
has_consensus       <- dir.exists(consensus_dir) && (length(list.files(file.path(consensus_dir, "plots"), "\\.png$")) > 0 || length(list.files(consensus_dir, "\\.csv$")) > 0)
has_stability       <- dir.exists(stability_dir) && length(list.files(file.path(stability_dir, "plots"), "\\.png$")) > 0
has_foundational    <- dir.exists(foundational_dir) && (length(list.files(file.path(foundational_dir, "plots"), "\\.png$")) > 0 || length(list.files(foundational_dir, "\\.csv$")) > 0)
has_mechanistic     <- dir.exists(mechanistic_dir) && (length(list.files(file.path(mechanistic_dir, "plots"), "\\.png$")) > 0 || length(list.files(mechanistic_dir, "\\.csv$")) > 0)

# Load gene-protein mapping if available
mapping_file <- file.path(run_dir, "gene_protein_mapping.csv")
gene_protein_mapping <- NULL
if (file.exists(mapping_file)) {
    gene_protein_mapping <- tryCatch(
        read.csv(mapping_file, stringsAsFactors = FALSE),
        error = function(e) NULL
    )
}
```

```{r force-dependencies, include=FALSE}
# Force DT and Plotly dependencies to be included in the HTML
DT::datatable(data.frame(x = 1))
plotly::plot_ly(data.frame(x = 1, y = 1), x = ~x, y = ~y, type = "scatter", mode = "markers")
```

# Analysis Parameters {.tabset}

## Project

```{r params-project}
proj <- data.frame(
  Parameter = c("Project Name", "Analysis Round", "Project Directory",
                 "Raw Data Path", "Output Path"),
  Value = c(cfg$project$name %||% "N/A",
            cfg$project$analysis_round %||% "N/A",
            cfg$project$dir %||% "N/A",
            cfg$paths$raw %||% "N/A",
            cfg$paths$out %||% "N/A"),
  stringsAsFactors = FALSE
)
kable(proj, col.names = c("Parameter", "Value"))
```

## Integration Methods

```{r params-integration}
methods <- multi_cfg$integration$methods %||% c("DIABLO", "SNF")
feat_sel <- multi_cfg$feature_selection$method %||% "variance"
feat_n <- multi_cfg$feature_selection$top_n %||% 500
cond_col <- multi_cfg$condition_column %||% cfg$design$condition_column %||% "condition"

settings <- data.frame(
  Parameter = c(
    "Integration Methods",
    "Feature Selection Method",
    "Top N Features",
    "Condition Column",
    "Require 1:1 Gene-Protein Mapping"
  ),
  Value = c(
    paste(methods, collapse = ", "),
    feat_sel,
    feat_n,
    cond_col,
    ifelse(isTRUE(multi_cfg$require_one_to_one_mapping), "Yes", "No")
  ),
  stringsAsFactors = FALSE
)
kable(settings, col.names = c("Parameter", "Value"))
```

## Omics Layers

```{r params-omics}
omics_modes <- setdiff(names(cfg$modes), "multiomics")
# In output mode, omics modes may not be in config — use global$omics_present
if (length(omics_modes) == 0 && !is.null(cfg$global$omics_present)) {
    omics_modes <- unlist(cfg$global$omics_present)
}
if (length(omics_modes) > 0) {
    omics_tbl <- data.frame(
        Omics = omics_modes,
        Configured = rep("Yes", length(omics_modes)),
        stringsAsFactors = FALSE
    )
    kable(omics_tbl, col.names = c("Omics Layer", "Configured"))
} else {
    cat("*No individual omics layers found in config.*\n")
}
```

---

# RNA-Protein Correlation {.tabset}

```{r rna-prot-note, eval=!has_rna_prot, results='asis'}
cat("*No RNA-protein correlation results found. This analysis requires both transcriptomics and proteomics data.*\n")
```

## Concordance Distribution

```{r rna-prot-concordance, eval=has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "rna_protein_concordance.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r rna-prot-concordance-legend, eval=has_rna_prot, results='asis'}
figure_legend("Distribution of per-gene RNA-protein Pearson correlations across samples. The histogram shows how well transcript levels predict protein abundance for individual genes.")
render_commentary("rna_protein_concordance")
```

## DE Scatter (All Contrasts)

```{r rna-prot-de-scatter, eval=has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "rna_protein_de_scatter.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r rna-prot-de-legend, eval=has_rna_prot, results='asis'}
figure_legend("RNA vs protein differential expression scatter plot per contrast. Each point is a gene/protein pair. Diagonal concordance indicates matching fold-change direction; off-diagonal points suggest post-transcriptional regulation.")
render_commentary("rna_protein_de_scatter")
```

## Per-Contrast DE Details {.tabset}

```{r rna-prot-per-contrast, eval=has_rna_prot, results='asis'}
per_contrast_dir <- file.path(rna_prot_dir, "per_contrast")
if (dir.exists(per_contrast_dir)) {
    contrast_dirs <- list.dirs(per_contrast_dir, full.names = TRUE, recursive = FALSE)
    for (cdir in contrast_dirs) {
        cname <- basename(cdir)
        cat(sprintf("\n### %s\n\n", gsub("_", " ", cname)))

        de_scatter <- file.path(cdir, "plots", "rna_protein_de_scatter.png")
        if (file.exists(de_scatter)) {
            cat(knitr::knit_child(text = sprintf(
                '```{r rna-prot-scatter-%s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                gsub("[^a-zA-Z0-9]", "", cname), de_scatter
            ), quiet = TRUE))
        }

        te_hist <- file.path(cdir, "plots", "translation_efficiency_hist.png")
        if (file.exists(te_hist)) {
            cat(knitr::knit_child(text = sprintf(
                '```{r rna-prot-te-%s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                gsub("[^a-zA-Z0-9]", "", cname), te_hist
            ), quiet = TRUE))
        }

        te_scatter <- file.path(cdir, "plots", "translation_efficiency_scatter.png")
        if (file.exists(te_scatter)) {
            cat(knitr::knit_child(text = sprintf(
                '```{r rna-prot-tes-%s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
                gsub("[^a-zA-Z0-9]", "", cname), te_scatter
            ), quiet = TRUE))
        }

        # Summary table
        de_csv <- file.path(cdir, "tables", "rna_protein_de_concordance.csv")
        if (file.exists(de_csv)) {
            de_df <- read.csv(de_csv, stringsAsFactors = FALSE)
            n_total <- nrow(de_df)
            n_conc <- sum(de_df$concordant, na.rm = TRUE)
            r_val <- cor(de_df$rna_log2FC, de_df$protein_log2FC, use = "complete.obs")
            cat(sprintf("\n**%s:** %d gene-protein pairs, r = %.2f, %d/%d (%.2f%%) concordant\n\n",
                        gsub("_", " ", cname), n_total, r_val,
                        n_conc, n_total, 100 * n_conc / n_total))
        }
    }
}
```

## Translation Efficiency (All Contrasts)

```{r rna-prot-te-hist, eval=has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "translation_efficiency_hist.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r rna-prot-te-hist-legend, eval=has_rna_prot, results='asis'}
figure_legend("Distribution of translation efficiency changes per contrast (protein log2FC - RNA log2FC). Values near 0 indicate transcriptional control; large positive/negative values suggest post-transcriptional regulation.")
render_commentary("translation_efficiency_hist")
```

```{r rna-prot-te-scatter, eval=has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "translation_efficiency_scatter.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r rna-prot-te-scatter-legend, eval=has_rna_prot, results='asis'}
figure_legend("Translation efficiency scatter plot per contrast comparing RNA and protein fold-changes.")
render_commentary("translation_efficiency_scatter")
```

---

# MultiGSEA Plots

```{r multigsea-note, eval=!has_multigsea, results='asis'}
cat("*No MultiGSEA plots found.*\n")
```

```{r multigsea-plots, eval=has_multigsea, results='asis'}
gsea_pngs <- list.files(multigsea_dir, pattern = "\\.png$", full.names = TRUE)
for (gp in gsea_pngs) {
    pair_name <- gsub("multigsea_|\\.png$", "", basename(gp))
    display_name <- gsub("_", " ", pair_name)
    cat(sprintf("\n## %s {.unnumbered}\n\n", display_name))
    cat(sprintf("![%s](%s)\n\n", display_name, gp))
    figure_legend(sprintf("MultiGSEA enrichment correlation plot for %s. Each point is a pathway; axis values represent enrichment scores. Color indicates joint significance status.", display_name))
    render_commentary(paste0("multigsea_", pair_name))
}
```

---

# Cross-Omics Enrichment

```{r enrichment-note, eval=!has_enrichment, results='asis'}
cat("*No cross-omics enrichment results found. Enrichment analysis requires pathway results from at least 1 omics layer.*\n")
```

```{r enrichment-per-omics, eval=has_enrichment, results='asis'}
# Show per-omics enrichment barplots
omics_barplots <- list.files(enrichment_dir, pattern = "_top_pathways\\.png$", full.names = TRUE)
if (length(omics_barplots) > 0) {
    cat("\n## Per-Omics Pathway Enrichment {.unnumbered}\n\n")
    for (bp in omics_barplots) {
        omics_name <- gsub("_top_pathways\\.png$", "", basename(bp))
        cat(sprintf("\n### %s {.unnumbered}\n\n", tools::toTitleCase(omics_name)))
        cat(sprintf("![%s pathways](%s)\n\n", tools::toTitleCase(omics_name), bp))
    }
}
```

```{r enrichment-per-omics-legend, eval=has_enrichment, results='asis'}
figure_legend("Per-omics KEGG pathway enrichment. Bar length represents -log10(p-value). Red dashed line indicates p = 0.05 significance threshold.")
```

```{r enrichment-per-omics-table, eval=has_enrichment}
# Show enrichment tables for each omics
csv_files <- list.files(enrichment_dir, pattern = "_enriched_pathways\\.csv$", full.names = TRUE)
for (cf in csv_files) {
    omics_name <- gsub("_enriched_pathways\\.csv$", "", basename(cf))
    tbl <- read.csv(cf, stringsAsFactors = FALSE)
    if (nrow(tbl) > 0) {
        print(make_dt(
            head(tbl, 200),
            caption = paste0(tools::toTitleCase(omics_name), " Enriched KEGG Pathways"),
            pageLength = 15
        ))
    }
}
```

```{r enrichment-heatmap, eval=has_enrichment, results='asis'}
p <- file.path(enrichment_dir, "cross_omics_pathway_heatmap.png")
if (file.exists(p)) {
    cat("\n## Cross-Omics Pathway Comparison {.unnumbered}\n\n")
    cat(sprintf("![Cross-Omics Pathway Heatmap](%s)\n\n", p))
}
```

```{r enrichment-heatmap-legend, eval=has_enrichment, results='asis'}
p <- file.path(enrichment_dir, "cross_omics_pathway_heatmap.png")
if (file.exists(p)) {
    figure_legend("Cross-omics pathway enrichment heatmap. Color intensity represents -log10(p-value) for each pathway across omics layers. Pathways are hierarchically clustered to reveal shared biological themes.")
    render_commentary("cross_omics_pathway_heatmap")
}
```

```{r enrichment-table, eval=has_enrichment}
meta_file <- file.path(enrichment_dir, "cross_omics_pathways_meta_analysis.csv")
if (file.exists(meta_file)) {
    meta_tbl <- read.csv(meta_file, stringsAsFactors = FALSE)
    make_dt(
        head(meta_tbl, 200),
        caption = "Cross-Omics Pathway Meta-Analysis (Fisher combined p-values)",
        pageLength = 15
    )
}
```

```{r enrichment-table-legend, eval=has_enrichment, results='asis'}
meta_file <- file.path(enrichment_dir, "cross_omics_pathways_meta_analysis.csv")
if (file.exists(meta_file)) {
    figure_legend("Top pathways from cross-omics meta-analysis. P-values from individual omics layers were combined using Fisher's method. Pathways significant in multiple omics are most informative.")
    render_commentary("cross_omics_meta_analysis")
}
```

---

# Data Harmonization

```{r harmonization-summary, results='asis'}
# Look for harmonization summary files
harmony_summary_file <- file.path(run_dir, "harmonization_summary.csv")
if (file.exists(harmony_summary_file)) {
    harmony_summary <- read.csv(harmony_summary_file, stringsAsFactors = FALSE)
    cat("## Harmonization Overview\n\n")
    kable(harmony_summary)
} else {
    cat("Multi-omics data were harmonized into a unified **MultiAssayExperiment** object. ",
        "Each omics layer was aligned to a common set of samples.\n\n")

    # Show omics layers
    omics_dirs <- list.dirs(integration_dir, full.names = FALSE, recursive = FALSE)
    if (length(omics_dirs) > 0) {
        cat("**Omics layers in integration:** ", paste(omics_dirs, collapse = ", "), "\n\n")
    }
}
```

```{r gene-protein-mapping, eval=!is.null(gene_protein_mapping), results='asis'}
cat("## Gene-Protein Mapping\n\n")
cat(sprintf("A total of **%d gene-protein pairs** were mapped across transcriptomics and proteomics layers.\n\n",
            nrow(gene_protein_mapping)))
```

```{r gene-protein-table, eval=!is.null(gene_protein_mapping) && nrow(gene_protein_mapping) > 0}
make_dt(
    head(gene_protein_mapping, 500),
    caption = "Gene-Protein Mapping (first 500 rows)",
    pageLength = 10
)
```

```{r harmonization-legend, eval=!is.null(gene_protein_mapping), results='asis'}
figure_legend("Gene-protein mapping table showing matched features between transcriptomics and proteomics layers.")
render_commentary("gene_protein_mapping")
```

---

# Integration Results {.tabset}

```{r integration-note, eval=!has_integration, results='asis'}
cat("*No integration results found. Integration methods may not have been configured or may have failed.*\n")
```

## DIABLO

### Sample Scores {.tabset}

#### Static

```{r diablo-sample-plot, eval=has_diablo, results='asis'}
p <- file.path(diablo_dir, "diablo_sample_plot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r diablo-sample-legend, eval=has_diablo, results='asis'}
figure_legend("DIABLO sample scores plot. Samples are projected onto the first two components of the multi-omics discriminant analysis. Colors represent experimental groups.")
render_commentary("diablo_sample_plot")
```

#### With Sample Names

```{r diablo-sample-labeled, eval=has_diablo, results='asis'}
p_lab <- file.path(diablo_dir, "diablo_sample_plot_labeled.png")
if (file.exists(p_lab)) knitr::include_graphics(p_lab) else cat("*Labeled plot not available.*\n")
```

#### Interactive

```{r diablo-sample-interactive, eval=has_diablo, fig.width=8, fig.height=6}
scores_csv <- file.path(diablo_dir, "diablo_sample_scores.csv")
if (file.exists(scores_csv) && requireNamespace("plotly", quietly = TRUE)) {
    sc <- read.csv(scores_csv, stringsAsFactors = FALSE)
    sc$hover_text <- paste0("Sample: ", sc$sample, "<br>Condition: ", sc$condition)
    pl <- plotly::plot_ly(sc, x = ~comp1, y = ~comp2, color = ~condition,
                          text = ~hover_text, hoverinfo = "text",
                          type = "scatter", mode = "markers",
                          marker = list(size = 10)) |>
        plotly::layout(xaxis = list(title = "Component 1"),
                       yaxis = list(title = "Component 2"),
                       title = "DIABLO Sample Scores — Hover for sample names")
    pl
} else {
    cat("*Interactive plot not available.*\n")
}
```

### Variable Loadings {.tabset}

#### Static

```{r diablo-variable-plot, eval=has_diablo, results='asis'}
p <- file.path(diablo_dir, "diablo_variable_plot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r diablo-variable-legend, eval=has_diablo, results='asis'}
figure_legend("DIABLO variable loadings plot. Top contributing features from each omics layer are shown on the first two components. Labels are omitted for clarity; hover over points in the Interactive tab for feature names.")
render_commentary("diablo_variable_plot")
```

#### Interactive

```{r diablo-variable-interactive, eval=has_diablo, results='asis'}
loadings_csv <- file.path(diablo_dir, "diablo_variable_loadings.csv")
if (file.exists(loadings_csv)) {
    ld <- read.csv(loadings_csv, stringsAsFactors = FALSE)
    ld$text <- paste0("Feature: ", ld$feature, "\nOmics: ", ld$omics,
                      "\nComp1: ", round(ld$comp1, 3),
                      "\nComp2: ", round(ld$comp2, 3))
    p_var <- ggplot(ld, aes(x = comp1, y = comp2, color = omics, text = text)) +
        geom_point(alpha = 0.7, size = 2) +
        labs(x = "Component 1 loading", y = "Component 2 loading",
             title = "DIABLO Variable Loadings", color = "Omics") +
        theme_minimal()
    plotly::ggplotly(p_var, tooltip = "text")
} else {
    cat("*Loadings CSV not available for interactive plot.*\n")
}
```

### Circos Plot {.tabset}

#### Static

```{r diablo-circos-plot, eval=has_diablo, results='asis'}
p <- file.path(diablo_dir, "diablo_circos_plot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r diablo-circos-legend, eval=has_diablo, results='asis'}
figure_legend("DIABLO circos plot showing cross-omics feature correlations. Lines connect correlated features across omics layers. Labels are omitted for clarity; see the Interactive tab for a searchable heatmap of cross-omics correlations.")
render_commentary("diablo_circos_plot")
```

#### Interactive

```{r diablo-circos-interactive, eval=has_diablo, results='asis'}
circos_csv <- file.path(diablo_dir, "diablo_circos_correlations.csv")
if (file.exists(circos_csv)) {
    edges <- read.csv(circos_csv, stringsAsFactors = FALSE)
    edges <- edges[order(edges$weight, decreasing = TRUE), ]
    edges <- head(edges, 80)
    plotly::plot_ly(
        x = edges$from, y = edges$to, z = edges$weight,
        text = paste0(edges$from, " [", edges$from_omics, "] \u2014 ",
                      edges$to, " [", edges$to_omics, "]",
                      "\nWeight: ", round(edges$weight, 3)),
        type = "heatmap", hoverinfo = "text",
        colorscale = list(c(0, "#f7fbff"), c(1, "#08306b"))
    ) %>%
        plotly::layout(
            title = "Cross-Omics Feature Correlations (DIABLO)",
            xaxis = list(title = "", tickangle = 45, tickfont = list(size = 8)),
            yaxis = list(title = "", tickfont = list(size = 8)),
            margin = list(b = 120, l = 120)
        )
} else {
    cat("*Circos correlation CSV not available for interactive plot.*\n")
}
```

```{r diablo-feature-heatmap, eval=has_diablo, results='asis'}
hm <- file.path(diablo_dir, "diablo_feature_heatmap.png")
if (file.exists(hm)) {
    cat("\n### Multi-Omics Feature Heatmap {.unnumbered}\n\n")
    cat(sprintf("![DIABLO Feature Heatmap](%s)\n\n", hm))
    figure_legend("Heatmap of top DIABLO features across omics layers. Values are z-scored per feature. Rows are grouped by omics type; columns are samples annotated by condition.")
    render_commentary("diablo_feature_heatmap")
}
```

```{r diablo-top-features, eval=has_diablo}
# Load top feature tables
diablo_feat_files <- list.files(diablo_dir, pattern = "diablo_top_features_.*\\.csv$", full.names = TRUE)
widgets <- list()
for (ff in diablo_feat_files) {
    omics_name <- gsub("diablo_top_features_|\\.csv$", "", basename(ff))
    df <- tryCatch(read.csv(ff, stringsAsFactors = FALSE), error = function(e) NULL)
    if (!is.null(df) && nrow(df) > 0) {
        widgets <- c(widgets, list(
            htmltools::tags$h4(paste("Top DIABLO Features:", omics_name)),
            make_dt(df, caption = paste("DIABLO top features -", omics_name))
        ))
    }
}
htmltools::tagList(widgets)
```

## SNF

```{r snf-heatmap, eval=has_snf, results='asis'}
p <- file.path(snf_dir, "snf_network_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r snf-heatmap-legend, eval=has_snf, results='asis'}
figure_legend("SNF fused similarity network heatmap. Rows and columns are samples; color intensity represents similarity across all omics layers after network fusion.")
render_commentary("snf_network_heatmap")
```

```{r snf-silhouette, eval=has_snf, results='asis'}
p <- file.path(snf_dir, "snf_silhouette.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r snf-silhouette-legend, eval=has_snf, results='asis'}
figure_legend("SNF cluster silhouette plot. Silhouette width quantifies how well each sample fits its assigned cluster. Values near 1 indicate good clustering; values near 0 or negative suggest misclassification.")
render_commentary("snf_silhouette")
```

```{r snf-clusters-table, eval=has_snf}
clusters_file <- file.path(snf_dir, "snf_clusters.csv")
if (file.exists(clusters_file)) {
    clusters <- read.csv(clusters_file, stringsAsFactors = FALSE)
    make_dt(clusters, caption = "SNF Cluster Assignments")
}
```

## MOFA2

```{r mofa-variance, eval=has_mofa, results='asis'}
p <- file.path(mofa_dir, "mofa_variance_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mofa-variance-legend, eval=has_mofa, results='asis'}
figure_legend("MOFA2 variance explained heatmap. Each cell shows the proportion of variance in a given omics view (rows) explained by a latent factor (columns). Factors capturing shared variation across multiple views are most informative.")
render_commentary("mofa_variance_heatmap")
```

```{r mofa-factors-12, eval=has_mofa, results='asis'}
p <- file.path(mofa_dir, "mofa_factors_1_2.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mofa-factors-12-legend, eval=has_mofa, results='asis'}
figure_legend("MOFA2 Factor 1 vs Factor 2 scatter plot. Each point is a sample, colored by experimental condition. Separation along factors indicates biologically relevant latent dimensions.")
render_commentary("mofa_factors_1_2")
```

```{r mofa-factors-13, eval=has_mofa, results='asis'}
p <- file.path(mofa_dir, "mofa_factors_1_3.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mofa-factors-13-legend, eval=has_mofa, results='asis'}
figure_legend("MOFA2 Factor 1 vs Factor 3 scatter plot.")
render_commentary("mofa_factors_1_3")
```

```{r mofa-weights, eval=has_mofa, results='asis'}
weight_files <- list.files(mofa_dir, pattern = "mofa_top_weights_.*\\.png$", full.names = TRUE)
if (length(weight_files) > 0) {
    for (wf in weight_files) {
        view_name <- gsub("mofa_top_weights_|\\.png$", "", basename(wf))
        cat(sprintf("\n### Top Weights: %s\n\n", view_name))
        cat(sprintf("![MOFA Top Weights %s](%s)\n\n", view_name, wf))
        figure_legend(sprintf("Top 20 features by absolute weight for MOFA Factor in the %s view.", view_name))
        fig_id <- paste0("mofa_top_weights_", view_name)
        render_commentary(fig_id)
    }
}
```

---

# Foundational Cross-Omics Analysis {.tabset}

```{r foundational-note, eval=!has_foundational, results='asis'}
cat("*No foundational cross-omics analysis results found.*\n")
```

## Cross-Omics Correlation

```{r foundational-corr-boxplot, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "crossomics_correlation_boxplot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r foundational-corr-boxplot-legend, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "crossomics_correlation_boxplot.png")
if (file.exists(p)) {
    figure_legend("Cross-omics correlation boxplot. Distribution of pairwise sample correlations between omics layers. High median correlations indicate concordant omics profiles.")
    render_commentary("crossomics_correlation_boxplot")
}
```

```{r foundational-corr-density, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "crossomics_correlation_density.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r foundational-corr-density-legend, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "crossomics_correlation_density.png")
if (file.exists(p)) {
    figure_legend("Cross-omics correlation density plot. Smoothed distribution of per-sample correlations between omics layers.")
    render_commentary("crossomics_correlation_density")
}
```

## Sample Clustering

```{r foundational-clustering, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "sample_clustering_consistency.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r foundational-clustering-legend, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "sample_clustering_consistency.png")
if (file.exists(p)) {
    figure_legend("Sample clustering consistency across omics layers. Measures whether samples cluster similarly regardless of omics platform, indicating robust biological signal.")
    render_commentary("sample_clustering_consistency")
}
```

## Condition Separation

```{r foundational-condition, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "sample_condition_separation.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r foundational-condition-legend, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "sample_condition_separation.png")
if (file.exists(p)) {
    figure_legend("Sample condition separation across omics layers. Shows how well each omics layer separates experimental conditions, helping identify the most informative data types.")
    render_commentary("sample_condition_separation")
}
```

## Sample Discordance

```{r foundational-discordance, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "sample_discordance_distribution.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r foundational-discordance-legend, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "sample_discordance_distribution.png")
if (file.exists(p)) {
    figure_legend("Distribution of sample discordance scores across omics layers. Samples with high discordance may indicate technical issues or genuinely complex biological states.")
    render_commentary("sample_discordance_distribution")
}
```

## Summary Tables

```{r foundational-tables, eval=has_foundational}
widgets <- list()
for (info in list(
    list(file = "foundational_analysis_summary.csv", cap = "Foundational Analysis Summary"),
    list(file = "crossomics_correlation_summary.csv", cap = "Cross-Omics Correlation Summary"),
    list(file = "sample_concordance_summary.csv", cap = "Sample Concordance Summary"),
    list(file = "sample_discordance.csv", cap = "Sample Discordance Scores")
)) {
    csv_path <- file.path(foundational_dir, "tables", info$file)
    if (file.exists(csv_path)) {
        df <- tryCatch(read.csv(csv_path, stringsAsFactors = FALSE), error = function(e) NULL)
        if (!is.null(df) && nrow(df) > 0)
            widgets <- c(widgets, list(make_dt(head(df, 200), caption = info$cap)))
    }
}
htmltools::tagList(widgets)
```

---

# Cross-Omics Concordance

```{r concordance-note, eval=!has_concordance, results='asis'}
cat("*No cross-omics concordance results found.*\n")
```

```{r concordance-plot, eval=has_concordance, results='asis'}
conc_pngs <- list.files(concordance_dir, pattern = "\\.png$", full.names = TRUE)
for (cp in conc_pngs) {
    display_name <- gsub("_", " ", gsub("\\.png$", "", basename(cp)))
    cat(sprintf("\n## %s {.unnumbered}\n\n", tools::toTitleCase(display_name)))
    cat(sprintf("![%s](%s)\n\n", display_name, cp))
    figure_legend(sprintf("Cross-omics concordance analysis: %s. Shows the agreement in differential expression direction and magnitude between omics layers.", display_name))
    fig_id <- gsub("\\.png$", "", basename(cp))
    render_commentary(fig_id)
}
```

```{r concordance-tables, eval=has_concordance}
widgets <- list()
conc_csvs <- list.files(concordance_dir, pattern = "\\.csv$", full.names = TRUE)
for (cf in conc_csvs) {
    tbl_name <- gsub("_", " ", gsub("\\.csv$", "", basename(cf)))
    df <- tryCatch(read.csv(cf, stringsAsFactors = FALSE), error = function(e) NULL)
    if (!is.null(df) && nrow(df) > 0)
        widgets <- c(widgets, list(make_dt(head(df, 300), caption = tools::toTitleCase(tbl_name))))
}
htmltools::tagList(widgets)
```

---

# Feature Networks {.tabset}

```{r featnet-note, eval=!has_feature_network, results='asis'}
cat("*No feature network results found.*\n")
```

## Tri-Omics Network

```{r featnet-triomics, eval=has_feature_network, results='asis'}
p <- file.path(feature_network_dir, "plots", "triomics_feature_network.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r featnet-triomics-legend, eval=has_feature_network, results='asis'}
p <- file.path(feature_network_dir, "plots", "triomics_feature_network.png")
if (file.exists(p)) {
    figure_legend("Tri-omics feature network connecting top features across all omics layers. Node colors represent omics type; edge thickness reflects correlation strength. Hub nodes are key integrative features.")
    render_commentary("triomics_feature_network")
}
```

## Per-Factor Networks {.tabset}

```{r featnet-factors, eval=has_feature_network, results='asis'}
factor_pngs <- list.files(file.path(feature_network_dir, "plots"),
                          pattern = "factor_.*_feature_network\\.png$",
                          full.names = TRUE)
factor_pngs <- sort(factor_pngs)
for (fp in factor_pngs) {
    factor_name <- gsub("_feature_network\\.png$", "", basename(fp))
    factor_name <- gsub("factor_", "", factor_name)
    cat(sprintf("\n### %s {.unnumbered}\n\n", factor_name))
    cat(knitr::knit_child(text = sprintf(
        '```{r featnet-%s, echo=FALSE}\nknitr::include_graphics("%s")\n```\n',
        gsub("[^a-zA-Z0-9]", "", factor_name), fp
    ), quiet = TRUE))
    figure_legend(sprintf("Feature network for %s. Shows cross-omics feature connections driven by this integration factor.", factor_name))
    render_commentary(paste0("factor_", factor_name, "_feature_network"))
}
```

## Pathway Crosstalk

```{r featnet-crosstalk, eval=has_feature_network, results='asis'}
p <- file.path(feature_network_dir, "plots", "pathway_crosstalk_network.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r featnet-crosstalk-legend, eval=has_feature_network, results='asis'}
p <- file.path(feature_network_dir, "plots", "pathway_crosstalk_network.png")
if (file.exists(p)) {
    figure_legend("Pathway crosstalk network. Nodes are enriched pathways; edges connect pathways that share features across omics layers, revealing biological process interdependencies.")
    render_commentary("pathway_crosstalk_network")
}
```

## Network Tables

```{r featnet-tables, eval=has_feature_network}
widgets <- list()
tables_dir <- file.path(feature_network_dir, "tables")
if (dir.exists(tables_dir)) {
    net_csvs <- list.files(tables_dir, pattern = "\\.csv$", full.names = TRUE)
    for (cf in net_csvs) {
        tbl_name <- gsub("_", " ", gsub("\\.csv$", "", basename(cf)))
        df <- tryCatch(read.csv(cf, stringsAsFactors = FALSE), error = function(e) NULL)
        if (!is.null(df) && nrow(df) > 0)
            widgets <- c(widgets, list(make_dt(head(df, 300), caption = tools::toTitleCase(tbl_name))))
    }
}
htmltools::tagList(widgets)
```

---

# Consensus Analysis {.tabset}

```{r consensus-note, eval=!has_consensus && !has_stability, results='asis'}
cat("*No consensus analysis results found. Consensus analysis compares results across integration methods.*\n")
```

## Method Agreement

```{r consensus-agreement, eval=has_consensus, results='asis'}
p <- file.path(consensus_dir, "plots", "method_agreement_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r consensus-agreement-legend, eval=has_consensus, results='asis'}
p <- file.path(consensus_dir, "plots", "method_agreement_heatmap.png")
if (file.exists(p)) {
    figure_legend("Method agreement heatmap comparing integration results across DIABLO, SNF, and MOFA2. Higher values indicate stronger agreement between methods in sample grouping or feature selection.")
    render_commentary("method_agreement_heatmap")
}
```

## Robust Features

```{r consensus-robust, eval=has_consensus, results='asis'}
p <- file.path(consensus_dir, "plots", "robust_features.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r consensus-robust-legend, eval=has_consensus, results='asis'}
p <- file.path(consensus_dir, "plots", "robust_features.png")
if (file.exists(p)) {
    figure_legend("Robust features identified by multiple integration methods. Features consistently selected across methods are most likely to be biologically relevant.")
    render_commentary("robust_features")
}
```

## Feature Rank Correlation

```{r consensus-rank-corr, eval=has_consensus, results='asis'}
p <- file.path(consensus_dir, "plots", "feature_rank_correlation.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r consensus-rank-corr-legend, eval=has_consensus, results='asis'}
p <- file.path(consensus_dir, "plots", "feature_rank_correlation.png")
if (file.exists(p)) {
    figure_legend("Feature rank correlation between integration methods. High rank correlation indicates methods agree on which features are most important.")
    render_commentary("feature_rank_correlation")
}
```

## Meta-Integration PCA

```{r consensus-meta-pca, eval=has_consensus, results='asis'}
p <- file.path(consensus_dir, "plots", "meta_integration_pca.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r consensus-meta-pca-legend, eval=has_consensus, results='asis'}
p <- file.path(consensus_dir, "plots", "meta_integration_pca.png")
if (file.exists(p)) {
    figure_legend("Meta-integration PCA combining results from all integration methods. This unified view highlights consensus sample groupings.")
    render_commentary("meta_integration_pca")
}
```

## Feature Stability

```{r stability-plot, eval=has_stability, results='asis'}
p <- file.path(stability_dir, "plots", "feature_stability.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r stability-legend, eval=has_stability, results='asis'}
p <- file.path(stability_dir, "plots", "feature_stability.png")
if (file.exists(p)) {
    figure_legend("Feature selection stability across bootstrap iterations. Features with high stability scores are robustly selected regardless of sample composition.")
    render_commentary("feature_stability")
}
```

## Consensus Tables

```{r consensus-tables, eval=has_consensus}
widgets <- list()
for (info in list(
    list(file = "robust_features.csv", cap = "Robust Features (Multi-Method Consensus)", n = 300),
    list(file = "feature_importance_comparison.csv", cap = "Feature Importance Comparison Across Methods", n = 300),
    list(file = "meta_integration_results.csv", cap = "Meta-Integration Results", n = 200),
    list(file = "method_ari_matrix.csv", cap = "Method Agreement — Adjusted Rand Index", n = NULL),
    list(file = "method_nmi_matrix.csv", cap = "Method Agreement — Normalized Mutual Information", n = NULL)
)) {
    csv_path <- file.path(consensus_dir, info$file)
    if (file.exists(csv_path)) {
        df <- tryCatch(read.csv(csv_path, stringsAsFactors = FALSE), error = function(e) NULL)
        if (!is.null(df) && nrow(df) > 0) {
            if (!is.null(info$n)) df <- head(df, info$n)
            widgets <- c(widgets, list(make_dt(df, caption = info$cap)))
        }
    }
}
htmltools::tagList(widgets)
```

```{r stability-table, eval=has_stability}
stab_csv <- file.path(stability_dir, "feature_selection_stability.csv")
if (file.exists(stab_csv)) {
    df <- tryCatch(read.csv(stab_csv, stringsAsFactors = FALSE), error = function(e) NULL)
    if (!is.null(df) && nrow(df) > 0) make_dt(head(df, 300), caption = "Feature Selection Stability Scores")
}
```

---

# Mechanistic Insights {.tabset}

```{r mechanistic-note, eval=!has_mechanistic, results='asis'}
cat("*No mechanistic analysis results found.*\n")
```

## Regulatory Network

```{r mech-regulatory, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_regulatory_network.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mech-regulatory-legend, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_regulatory_network.png")
if (file.exists(p)) {
    figure_legend("Multi-omics regulatory network. Nodes represent genes/proteins; edges indicate regulatory relationships inferred from cross-omics correlations. Hub nodes with high connectivity are candidate master regulators.")
    render_commentary("mech_regulatory_network")
}
```

## Translation Efficiency

```{r mech-te-dist, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_translation_efficiency_dist.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mech-te-dist-legend, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_translation_efficiency_dist.png")
if (file.exists(p)) {
    figure_legend("Distribution of translation efficiency scores. Translation efficiency (protein log2FC - RNA log2FC) reveals post-transcriptional regulatory mechanisms.")
    render_commentary("mech_translation_efficiency_dist")
}
```

```{r mech-te-scatter, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_translation_efficiency_scatter.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mech-te-scatter-legend, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_translation_efficiency_scatter.png")
if (file.exists(p)) {
    figure_legend("Translation efficiency scatter plot. Each point is a gene/protein pair; deviation from the diagonal indicates post-transcriptional regulation. Color highlights genes with extreme translation efficiency.")
    render_commentary("mech_translation_efficiency_scatter")
}
```

## RNA-Protein Correlation

```{r mech-rnaprot-corr, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_rna_protein_correlation_dist.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mech-rnaprot-corr-legend, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_rna_protein_correlation_dist.png")
if (file.exists(p)) {
    figure_legend("Distribution of per-gene RNA-protein correlations from the mechanistic analysis. Genes with low correlation are candidates for post-transcriptional regulation.")
    render_commentary("mech_rna_protein_correlation_dist")
}
```

## Protein Stability

```{r mech-stability, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_protein_stability_dist.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mech-stability-legend, eval=has_mechanistic, results='asis'}
p <- file.path(mechanistic_dir, "plots", "mech_protein_stability_dist.png")
if (file.exists(p)) {
    figure_legend("Protein stability distribution inferred from RNA-protein dynamics. Proteins with high stability scores maintain abundance despite transcript changes, suggesting long half-lives or regulated degradation.")
    render_commentary("mech_protein_stability_dist")
}
```

## Mechanistic Tables

```{r mech-tables, eval=has_mechanistic}
widgets <- list()
for (info in list(
    list(file = "mech_hub_regulators.csv", cap = "Hub Regulators — Multi-Omics Regulatory Network"),
    list(file = "mech_post_transcriptional_regulation.csv", cap = "Post-Transcriptional Regulation Candidates"),
    list(file = "mech_translation_efficiency.csv", cap = "Translation Efficiency Scores"),
    list(file = "mech_protein_stability.csv", cap = "Protein Stability Estimates"),
    list(file = "mech_rna_protein_correlations.csv", cap = "RNA-Protein Correlations"),
    list(file = "mechanistic_analysis_summary.csv", cap = "Mechanistic Analysis Summary")
)) {
    csv_path <- file.path(mechanistic_dir, "tables", info$file)
    if (file.exists(csv_path)) {
        df <- tryCatch(read.csv(csv_path, stringsAsFactors = FALSE), error = function(e) NULL)
        if (!is.null(df) && nrow(df) > 0)
            widgets <- c(widgets, list(make_dt(head(df, 200), caption = info$cap)))
    }
}
htmltools::tagList(widgets)
```

---

# Output Files

```{r output-files, results='asis'}
cat(sprintf("All multi-omics integration results have been saved to:\n\n`%s`\n\n", run_dir))

cat("**Subdirectories:**\n\n")

subdirs <- list.dirs(run_dir, full.names = FALSE, recursive = FALSE)
subdirs <- subdirs[subdirs != "execution_info"]
if (length(subdirs) > 0) {
    for (sd in sort(subdirs)) {
        n_files <- length(list.files(file.path(run_dir, sd), recursive = TRUE))
        cat(sprintf("- **%s/** (%d files)\n", sd, n_files))
    }
}
```

---

# Methods

```{r methods, results='asis'}
cat("## Multi-Omics Integration Pipeline\n\n")

cat("This report was generated by the **multiomics-core** pipeline. ",
    "The following analytical steps were performed:\n\n")

cat("1. **Data Harmonization**: Individual omics layers (preprocessed by their respective sub-pipelines) ",
    "were aligned to a common set of samples and assembled into a `MultiAssayExperiment` object.\n\n")

methods <- multi_cfg$integration$methods %||% c("DIABLO", "SNF")
step_num <- 2
if ("DIABLO" %in% methods) {
    cat(sprintf("%d. **DIABLO (Data Integration Analysis for Biomarker discovery using Latent cOmponents)**: ", step_num),
        "A supervised multi-omics integration method from the mixOmics package that identifies ",
        "discriminant features across omics layers using sparse PLS-DA.\n\n")
    step_num <- step_num + 1
}
if ("SNF" %in% methods) {
    cat(sprintf("%d. **SNF (Similarity Network Fusion)**: ", step_num),
        "An unsupervised method that constructs sample similarity networks for each omics layer ",
        "and fuses them into a single network for clustering.\n\n")
    step_num <- step_num + 1
}
if ("MOFA2" %in% methods) {
    cat(sprintf("%d. **MOFA2 (Multi-Omics Factor Analysis v2)**: ", step_num),
        "An unsupervised Bayesian factor model that identifies latent factors capturing ",
        "shared and omics-specific sources of variation.\n\n")
    step_num <- step_num + 1
}

cat(sprintf("%d. **Cross-Omics Enrichment**: Pathway enrichment results were aggregated across omics ", step_num),
    "using Fisher's method for combined p-values.\n\n")

cat("## Software\n\n")
cat("```\n")
cat(capture.output(sessionInfo()), sep = "\n")
cat("\n```\n")
```
