---
title: "Multi-Omics Integration Report"
subtitle: "multiomics-core Pipeline"
author: "Bioinformatician: `r yaml::read_yaml(file.path(normalizePath(dirname(knitr::current_input())), 'execution_info', 'config_used.yaml'))$project$analyst`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    fig_width: 10
    fig_height: 7
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%"
)

library(yaml)
library(knitr)
library(DT)
library(plotly)
library(ggplot2)
library(htmltools)
library(jsonlite)

# Helper function for downloadable DT tables
make_dt <- function(df, caption = NULL, pageLength = 15) {
    DT::datatable(
        df,
        caption = caption,
        extensions = 'Buttons',
        options = list(
            scrollX = TRUE,
            pageLength = pageLength,
            dom = 'Bfrtip',
            buttons = list(
                list(extend = 'csv', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data")),
                list(extend = 'excel', filename = gsub("[^a-zA-Z0-9]", "_", caption %||% "data"))
            )
        ),
        rownames = FALSE
    )
}

# Helper: render figure legend (paper-style caption below figure)
figure_legend <- function(text) {
    cat(sprintf("<p class='figure-legend' style='text-align: center; font-size: 0.9em; color: #555; margin: 4px auto 12px auto; max-width: 88%%; line-height: 1.45; font-style: italic;'>%s</p>\n", text))
}

# Helper: render commentary block for a figure_id (collapsible)
render_commentary <- function(figure_id) {
    if (!has_commentary) return(invisible(NULL))
    c <- commentary[[figure_id]]
    if (is.null(c)) return(invisible(NULL))

    cat("<details class='commentary-details' style='margin: 4px 0 18px 0;'>\n")
    cat("<summary style='cursor: pointer; color: #4a90d9; font-weight: 500; font-size: 0.93em; padding: 3px 0;'>Show AI Interpretation</summary>\n")

    html <- "<div class='commentary-block' style='background: #f8f9fa; border-left: 4px solid #4a90d9; padding: 12px 16px; margin: 6px 0 4px 0; border-radius: 0 4px 4px 0; font-size: 0.92em;'>\n"
    html <- paste0(html, "<p style='margin: 0 0 8px 0;'><strong>What this shows:</strong> ",
                   htmltools::htmlEscape(c$what_is_this %||% ""), "</p>\n")

    bio_interp <- c$biological_interpretation %||% ""
    if (nzchar(bio_interp)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Interpretation:</strong> ",
                       htmltools::htmlEscape(bio_interp), "</p>\n")
    }

    obs <- c$observations
    if (length(obs) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Key observations:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (o in obs) html <- paste0(html, "<li>", htmltools::htmlEscape(o), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    stat_assess <- c$statistical_assessment %||% ""
    if (nzchar(stat_assess)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Statistical note:</strong> ",
                       htmltools::htmlEscape(stat_assess), "</p>\n")
    }

    issues <- c$issues_checks
    if (length(issues) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Checks:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (issue in issues) html <- paste0(html, "<li>", htmltools::htmlEscape(issue), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    steps <- c$next_steps
    if (length(steps) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0;'><strong>Next steps:</strong></p>\n<ul style='margin: 2px 0 4px 0; padding-left: 20px;'>\n")
        for (s in steps) html <- paste0(html, "<li>", htmltools::htmlEscape(s), "</li>\n")
        html <- paste0(html, "</ul>\n")
    }

    backend <- c$backend %||% "unknown"
    confidence <- c$confidence %||% ""
    backend_label <- if (grepl("claude", backend, ignore.case = TRUE)) {
        "Claude (Anthropic) + K-Dense AI scientific-skills"
    } else if (grepl("openai", backend, ignore.case = TRUE)) {
        "OpenAI + K-Dense AI scientific-skills"
    } else if (backend == "data_driven") {
        "data-driven + K-Dense AI scientific-skills"
    } else {
        backend
    }
    if (nzchar(confidence) && confidence != "none") {
        backend_label <- paste0(backend_label, " | confidence: ", confidence)
    }
    html <- paste0(html, "<p style='margin: 4px 0 0 0; color: #888; font-size: 0.85em;'><em>Generated by: ",
                   htmltools::htmlEscape(backend_label), "</em></p>\n</div>\n")

    cat(html)
    cat("</details>\n")
}

# Paths
run_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
config_path <- file.path(run_dir, "execution_info", "config_used.yaml")

# Multiomics output subdirectories
integration_dir  <- file.path(run_dir, "integration")
diablo_dir       <- file.path(integration_dir, "diablo")
snf_dir          <- file.path(integration_dir, "snf")
mofa_dir         <- file.path(integration_dir, "mofa")
concordance_dir  <- file.path(run_dir, "concordance")
enrichment_dir   <- file.path(run_dir, "cross_enrichment")
rna_prot_dir     <- file.path(run_dir, "rna_protein_correlation")
multigsea_dir    <- file.path(run_dir, "multigsea")
foundational_dir <- file.path(run_dir, "foundational", "foundational")
mechanistic_dir  <- file.path(run_dir, "mechanistic", "mechanistic")
consensus_dir    <- file.path(run_dir, "consensus", "consensus")
stability_dir    <- file.path(run_dir, "consensus", "stability")
commentary_dir   <- file.path(run_dir, "commentary")

# Load commentary if available
commentary <- NULL
has_commentary <- FALSE
commentary_json <- file.path(commentary_dir, "all_commentary.json")
if (file.exists(commentary_json)) {
    commentary <- tryCatch(
        jsonlite::read_json(commentary_json),
        error = function(e) NULL
    )
    has_commentary <- !is.null(commentary) && length(commentary) > 0
}

# Load config
cfg <- yaml::read_yaml(config_path)
multi_cfg <- cfg$modes$multiomics

# Detect which sections have data
has_diablo      <- dir.exists(diablo_dir) && length(list.files(diablo_dir, "\\.png$")) > 0
has_snf         <- dir.exists(snf_dir) && length(list.files(snf_dir, "\\.png$")) > 0
has_mofa        <- dir.exists(mofa_dir) && length(list.files(mofa_dir, "\\.png$")) > 0
has_integration <- has_diablo || has_snf || has_mofa
has_concordance <- dir.exists(concordance_dir) && length(list.files(concordance_dir, "\\.png$|\\.csv$")) > 0
has_enrichment  <- dir.exists(enrichment_dir) && length(list.files(enrichment_dir, "\\.png$|\\.csv$")) > 0
has_rna_prot    <- dir.exists(file.path(rna_prot_dir, "plots")) && length(list.files(file.path(rna_prot_dir, "plots"), "\\.png$")) > 0
has_multigsea   <- dir.exists(multigsea_dir) && length(list.files(multigsea_dir, "\\.png$")) > 0
has_foundational <- dir.exists(foundational_dir) && length(list.files(file.path(foundational_dir, "plots"), "\\.png$")) > 0
has_mechanistic <- dir.exists(mechanistic_dir) && length(list.files(file.path(mechanistic_dir, "plots"), "\\.png$")) > 0
has_consensus   <- dir.exists(consensus_dir) && length(list.files(file.path(consensus_dir, "plots"), "\\.png$")) > 0

# Load gene-protein mapping if available
mapping_file <- file.path(run_dir, "gene_protein_mapping.csv")
gene_protein_mapping <- NULL
if (file.exists(mapping_file)) {
    gene_protein_mapping <- tryCatch(
        read.csv(mapping_file, stringsAsFactors = FALSE),
        error = function(e) NULL
    )
}
```

```{r force-dependencies, include=FALSE}
# Force DT and Plotly dependencies to be included in the HTML
DT::datatable(data.frame(x = 1))
plotly::plot_ly(data.frame(x = 1, y = 1), x = ~x, y = ~y, type = "scatter", mode = "markers")
```

# Analysis Parameters {.tabset}

## Project

```{r params-project}
proj <- data.frame(
  Parameter = c("Project Name", "Analysis Round", "Project Directory",
                 "Raw Data Path", "Output Path"),
  Value = c(cfg$project$name %||% "N/A",
            cfg$project$analysis_round %||% "N/A",
            cfg$project$dir %||% "N/A",
            cfg$paths$raw %||% "N/A",
            cfg$paths$out %||% "N/A"),
  stringsAsFactors = FALSE
)
kable(proj, col.names = c("Parameter", "Value"))
```

## Integration Methods

```{r params-integration}
methods <- multi_cfg$integration$methods %||% c("DIABLO", "SNF")
feat_sel <- multi_cfg$feature_selection$method %||% "variance"
feat_n <- multi_cfg$feature_selection$top_n %||% 500
cond_col <- multi_cfg$condition_column %||% cfg$design$condition_column %||% "condition"

settings <- data.frame(
  Parameter = c(
    "Integration Methods",
    "Feature Selection Method",
    "Top N Features",
    "Condition Column",
    "Require 1:1 Gene-Protein Mapping"
  ),
  Value = c(
    paste(methods, collapse = ", "),
    feat_sel,
    feat_n,
    cond_col,
    ifelse(isTRUE(multi_cfg$require_one_to_one_mapping), "Yes", "No")
  ),
  stringsAsFactors = FALSE
)
kable(settings, col.names = c("Parameter", "Value"))
```

## Omics Layers

```{r params-omics}
omics_modes <- setdiff(names(cfg$modes), "multiomics")
omics_tbl <- data.frame(
    Omics = omics_modes,
    Configured = "Yes",
    stringsAsFactors = FALSE
)
kable(omics_tbl, col.names = c("Omics Layer", "Configured"))
```

---

# Data Harmonization

```{r harmonization-summary, results='asis'}
# Look for harmonization summary files
harmony_summary_file <- file.path(run_dir, "harmonization_summary.csv")
if (file.exists(harmony_summary_file)) {
    harmony_summary <- read.csv(harmony_summary_file, stringsAsFactors = FALSE)
    cat("## Harmonization Overview\n\n")
    kable(harmony_summary)
} else {
    cat("Multi-omics data were harmonized into a unified **MultiAssayExperiment** object. ",
        "Each omics layer was aligned to a common set of samples.\n\n")

    # Show omics layers
    omics_dirs <- list.dirs(integration_dir, full.names = FALSE, recursive = FALSE)
    if (length(omics_dirs) > 0) {
        cat("**Omics layers in integration:** ", paste(omics_dirs, collapse = ", "), "\n\n")
    }
}
```

```{r gene-protein-mapping, eval=!is.null(gene_protein_mapping), results='asis'}
cat("## Gene-Protein Mapping\n\n")
cat(sprintf("A total of **%d gene-protein pairs** were mapped across transcriptomics and proteomics layers.\n\n",
            nrow(gene_protein_mapping)))
```

```{r gene-protein-table, eval=!is.null(gene_protein_mapping) && nrow(gene_protein_mapping) > 0}
make_dt(
    head(gene_protein_mapping, 500),
    caption = "Gene-Protein Mapping (first 500 rows)",
    pageLength = 10
)
```

```{r harmonization-legend, eval=!is.null(gene_protein_mapping), results='asis'}
figure_legend("Gene-protein mapping table showing matched features between transcriptomics and proteomics layers.")
render_commentary("gene_protein_mapping")
```

---

# Integration Results {.tabset}

```{r integration-note, eval=!has_integration, results='asis'}
cat("*No integration results found. Integration methods may not have been configured or may have failed.*\n")
```

## DIABLO

```{r diablo-sample-plot, eval=has_diablo, results='asis'}
p <- file.path(diablo_dir, "diablo_sample_plot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r diablo-sample-legend, eval=has_diablo, results='asis'}
figure_legend("DIABLO sample scores plot. Samples are projected onto the first two components of the multi-omics discriminant analysis. Colors represent experimental groups.")
render_commentary("diablo_sample_plot")
```

```{r diablo-variable-plot, eval=has_diablo, results='asis'}
p <- file.path(diablo_dir, "diablo_variable_plot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r diablo-variable-legend, eval=has_diablo, results='asis'}
figure_legend("DIABLO variable loadings plot. Top contributing features from each omics layer are shown on the first two components.")
render_commentary("diablo_variable_plot")
```

```{r diablo-circos-plot, eval=has_diablo, results='asis'}
p <- file.path(diablo_dir, "diablo_circos_plot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r diablo-circos-legend, eval=has_diablo, results='asis'}
figure_legend("DIABLO circos plot showing cross-omics feature correlations. Lines connect correlated features across omics layers.")
render_commentary("diablo_circos_plot")
```

```{r diablo-top-features, eval=has_diablo}
# Load top feature tables
diablo_feat_files <- list.files(diablo_dir, pattern = "diablo_top_features_.*\\.csv$", full.names = TRUE)
if (length(diablo_feat_files) > 0) {
    for (ff in diablo_feat_files) {
        omics_name <- gsub("diablo_top_features_|\\.csv$", "", basename(ff))
        df <- tryCatch(read.csv(ff, stringsAsFactors = FALSE), error = function(e) NULL)
        if (!is.null(df) && nrow(df) > 0) {
            cat(sprintf("\n### Top DIABLO Features: %s\n\n", omics_name))
            print(make_dt(df, caption = paste("DIABLO top features -", omics_name)))
        }
    }
}
```

## SNF

```{r snf-heatmap, eval=has_snf, results='asis'}
p <- file.path(snf_dir, "snf_network_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r snf-heatmap-legend, eval=has_snf, results='asis'}
figure_legend("SNF fused similarity network heatmap. Rows and columns are samples; color intensity represents similarity across all omics layers after network fusion.")
render_commentary("snf_network_heatmap")
```

```{r snf-silhouette, eval=has_snf, results='asis'}
p <- file.path(snf_dir, "snf_silhouette.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r snf-silhouette-legend, eval=has_snf, results='asis'}
figure_legend("SNF cluster silhouette plot. Silhouette width quantifies how well each sample fits its assigned cluster. Values near 1 indicate good clustering; values near 0 or negative suggest misclassification.")
render_commentary("snf_silhouette")
```

```{r snf-clusters-table, eval=has_snf}
clusters_file <- file.path(snf_dir, "snf_clusters.csv")
if (file.exists(clusters_file)) {
    clusters <- read.csv(clusters_file, stringsAsFactors = FALSE)
    make_dt(clusters, caption = "SNF Cluster Assignments")
}
```

## MOFA2

```{r mofa-variance, eval=has_mofa, results='asis'}
p <- file.path(mofa_dir, "mofa_variance_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mofa-variance-legend, eval=has_mofa, results='asis'}
figure_legend("MOFA2 variance explained heatmap. Each cell shows the proportion of variance in a given omics view (rows) explained by a latent factor (columns). Factors capturing shared variation across multiple views are most informative.")
render_commentary("mofa_variance_heatmap")
```

```{r mofa-factors-12, eval=has_mofa, results='asis'}
p <- file.path(mofa_dir, "mofa_factors_1_2.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mofa-factors-12-legend, eval=has_mofa, results='asis'}
figure_legend("MOFA2 Factor 1 vs Factor 2 scatter plot. Each point is a sample, colored by experimental condition. Separation along factors indicates biologically relevant latent dimensions.")
render_commentary("mofa_factors_1_2")
```

```{r mofa-factors-13, eval=has_mofa, results='asis'}
p <- file.path(mofa_dir, "mofa_factors_1_3.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r mofa-factors-13-legend, eval=has_mofa, results='asis'}
figure_legend("MOFA2 Factor 1 vs Factor 3 scatter plot.")
render_commentary("mofa_factors_1_3")
```

```{r mofa-weights, eval=has_mofa, results='asis'}
weight_files <- list.files(mofa_dir, pattern = "mofa_top_weights_.*\\.png$", full.names = TRUE)
if (length(weight_files) > 0) {
    for (wf in weight_files) {
        view_name <- gsub("mofa_top_weights_|\\.png$", "", basename(wf))
        cat(sprintf("\n### Top Weights: %s\n\n", view_name))
        cat(knitr::include_graphics(wf))
        figure_legend(sprintf("Top 20 features by absolute weight for MOFA Factor in the %s view.", view_name))
        fig_id <- paste0("mofa_top_weights_", view_name)
        render_commentary(fig_id)
    }
}
```

---

# Concordance & Overlap {.tabset}

```{r concordance-note, eval=!has_concordance, results='asis'}
cat("*No concordance results found. Concordance analysis requires DE results from at least 2 omics layers.*\n")
```

```{r concordance-plots, eval=has_concordance, results='asis'}
conc_pngs <- list.files(concordance_dir, pattern = "\\.png$", full.names = TRUE)
for (cp in conc_pngs) {
    pair_name <- gsub("concordance_|\\.png$", "", basename(cp))
    cat(sprintf("\n## %s\n\n", gsub("_", " vs ", pair_name)))
    cat(knitr::include_graphics(cp))
    figure_legend(sprintf("Concordance scatter plot for %s. Each point represents a feature with DE results in both omics layers. Colors indicate concordance categories (concordant, discordant, single-omics significant, not significant).", gsub("_", " vs ", pair_name)))
    render_commentary(paste0("concordance_", pair_name))
}
```

```{r concordance-tables, eval=has_concordance}
conc_csvs <- list.files(concordance_dir, pattern = "\\.csv$", full.names = TRUE)
if (length(conc_csvs) > 0) {
    for (cf in conc_csvs) {
        df <- tryCatch(read.csv(cf, stringsAsFactors = FALSE), error = function(e) NULL)
        if (!is.null(df) && nrow(df) > 0) {
            tbl_name <- gsub("\\.csv$", "", basename(cf))
            cat(sprintf("\n### %s\n\n", tbl_name))
            print(make_dt(head(df, 500), caption = tbl_name, pageLength = 10))
        }
    }
}
```

---

# Cross-Omics Enrichment

```{r enrichment-note, eval=!has_enrichment, results='asis'}
cat("*No cross-omics enrichment results found. Enrichment analysis requires pathway results from at least 2 omics layers.*\n")
```

```{r enrichment-heatmap, eval=has_enrichment, results='asis'}
p <- file.path(enrichment_dir, "cross_omics_pathway_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r enrichment-heatmap-legend, eval=has_enrichment, results='asis'}
figure_legend("Cross-omics pathway enrichment heatmap. Color intensity represents -log10(p-value) for each pathway across omics layers. Pathways are hierarchically clustered to reveal shared biological themes.")
render_commentary("cross_omics_pathway_heatmap")
```

```{r enrichment-table, eval=has_enrichment}
meta_file <- file.path(enrichment_dir, "cross_omics_pathways_meta_analysis.csv")
if (file.exists(meta_file)) {
    meta_tbl <- read.csv(meta_file, stringsAsFactors = FALSE)
    make_dt(
        head(meta_tbl, 200),
        caption = "Cross-Omics Pathway Meta-Analysis (Fisher combined p-values)",
        pageLength = 15
    )
}
```

```{r enrichment-table-legend, eval=has_enrichment, results='asis'}
figure_legend("Top pathways from cross-omics meta-analysis. P-values from individual omics layers were combined using Fisher's method. Pathways significant in multiple omics are most informative.")
render_commentary("cross_omics_meta_analysis")
```

---

# MultiGSEA Plots

```{r multigsea-note, eval=!has_multigsea, results='asis'}
cat("*No MultiGSEA plots found.*\n")
```

```{r multigsea-plots, eval=has_multigsea, results='asis'}
gsea_pngs <- list.files(multigsea_dir, pattern = "\\.png$", full.names = TRUE)
for (gp in gsea_pngs) {
    pair_name <- gsub("multigsea_|\\.png$", "", basename(gp))
    cat(sprintf("\n## %s {.unnumbered}\n\n", gsub("_", " vs ", pair_name)))
    cat(knitr::include_graphics(gp))
    figure_legend(sprintf("MultiGSEA enrichment correlation plot for %s. Each point is a pathway; axis values represent enrichment scores. Color indicates joint significance status.", gsub("_", " vs ", pair_name)))
    render_commentary(paste0("multigsea_", pair_name))
}
```

---

# RNA-Protein Correlation {.tabset}

```{r rna-prot-note, eval=!has_rna_prot, results='asis'}
cat("*No RNA-protein correlation results found. This analysis requires both transcriptomics and proteomics data.*\n")
```

## Concordance Distribution

```{r rna-prot-concordance, eval=has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "rna_protein_concordance.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r rna-prot-concordance-legend, eval=has_rna_prot, results='asis'}
figure_legend("Distribution of per-gene RNA-protein Pearson correlations across samples. The histogram shows how well transcript levels predict protein abundance for individual genes.")
render_commentary("rna_protein_concordance")
```

## DE Scatter

```{r rna-prot-de-scatter, eval=has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "rna_protein_de_scatter.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r rna-prot-de-legend, eval=has_rna_prot, results='asis'}
figure_legend("RNA vs protein differential expression scatter plot. Each point is a gene/protein pair. Diagonal concordance indicates matching fold-change direction; off-diagonal points suggest post-transcriptional regulation.")
render_commentary("rna_protein_de_scatter")
```

## Translation Efficiency

```{r rna-prot-te-hist, eval=has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "translation_efficiency_hist.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r rna-prot-te-hist-legend, eval=has_rna_prot, results='asis'}
figure_legend("Distribution of translation efficiency changes (protein log2FC - RNA log2FC). Values near 0 indicate transcriptional control; large positive/negative values suggest post-transcriptional regulation.")
render_commentary("translation_efficiency_hist")
```

```{r rna-prot-te-scatter, eval=has_rna_prot, results='asis'}
p <- file.path(rna_prot_dir, "plots", "translation_efficiency_scatter.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r rna-prot-te-scatter-legend, eval=has_rna_prot, results='asis'}
figure_legend("Translation efficiency scatter plot comparing RNA and protein fold-changes.")
render_commentary("translation_efficiency_scatter")
```

---

# Foundational Cross-Omics Analysis {.tabset}

```{r foundational-note, eval=!has_foundational, results='asis'}
cat("*No foundational cross-omics analysis results found.*\n")
```

## Cross-Omics Correlations

```{r foundational-corr-density, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "crossomics_correlation_density.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r foundational-corr-density-legend, eval=has_foundational, results='asis'}
figure_legend("Density plot of pairwise cross-omics feature correlations. The distribution shape indicates the overall degree of concordance between omics layers.")
render_commentary("crossomics_correlation_density")
```

```{r foundational-corr-boxplot, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "crossomics_correlation_boxplot.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r foundational-corr-boxplot-legend, eval=has_foundational, results='asis'}
figure_legend("Boxplot of cross-omics feature correlations stratified by omics pair.")
render_commentary("crossomics_correlation_boxplot")
```

## Sample Clustering

```{r foundational-clustering, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "sample_clustering_consistency.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r foundational-clustering-legend, eval=has_foundational, results='asis'}
figure_legend("Sample clustering consistency across omics layers. Samples that cluster together in multiple omics layers provide stronger biological signals.")
render_commentary("sample_clustering_consistency")
```

```{r foundational-condition-sep, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "sample_condition_separation.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r foundational-condition-sep-legend, eval=has_foundational, results='asis'}
figure_legend("Sample condition separation across omics layers.")
render_commentary("sample_condition_separation")
```

## Pathway Overlap

```{r foundational-pathway-heatmap, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "pathway_overlap_heatmap.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r foundational-pathway-legend, eval=has_foundational, results='asis'}
figure_legend("Pathway overlap heatmap across omics layers. Shared enriched pathways indicate convergent biological processes detected by multiple omics technologies.")
render_commentary("pathway_overlap_heatmap")
```

## Cross-Omics Modules

```{r foundational-modules, eval=has_foundational, results='asis'}
p <- file.path(foundational_dir, "plots", "crossomics_module_composition.png")
if (file.exists(p)) {
    knitr::include_graphics(p)
}
```

```{r foundational-modules-legend, eval=has_foundational, results='asis'}
figure_legend("Cross-omics module composition showing the contribution of each omics layer to co-expression modules.")
render_commentary("crossomics_module_composition")
```

---

# Mechanistic Inference {.tabset}

```{r mechanistic-note, eval=!has_mechanistic, results='asis'}
cat("*No mechanistic inference results found. This analysis requires sufficient integration results and may not run for all datasets.*\n")
```

```{r mechanistic-plots, eval=has_mechanistic, results='asis'}
mech_plots_dir <- file.path(mechanistic_dir, "plots")
mech_pngs <- list.files(mech_plots_dir, pattern = "\\.png$", full.names = TRUE)

# Define plot titles and legends
mech_labels <- list(
    mech_rna_protein_correlation_dist = list(
        title = "RNA-Protein Correlation Distribution",
        legend = "Distribution of RNA-protein expression correlations used in mechanistic modeling."
    ),
    mech_translation_efficiency_scatter = list(
        title = "Translation Efficiency Scatter",
        legend = "Scatter plot of translation efficiency estimates from mechanistic inference."
    ),
    mech_translation_efficiency_dist = list(
        title = "Translation Efficiency Distribution",
        legend = "Distribution of translation efficiency changes across genes."
    ),
    mech_protein_stability_dist = list(
        title = "Protein Stability Distribution",
        legend = "Distribution of estimated protein stability changes."
    ),
    mech_tf_association_barplot = list(
        title = "TF Association Barplot",
        legend = "Top transcription factor associations identified from multi-omics data."
    ),
    mech_mediation_proportion_barplot = list(
        title = "Mediation Proportion",
        legend = "Proportion of the treatment effect mediated through different molecular layers."
    ),
    mech_mediation_effect_decomposition = list(
        title = "Mediation Effect Decomposition",
        legend = "Decomposition of direct and indirect effects from mediation analysis."
    )
)

for (mp in mech_pngs) {
    plot_id <- gsub("\\.png$", "", basename(mp))
    info <- mech_labels[[plot_id]]
    tab_title <- if (!is.null(info)) info$title else gsub("_", " ", gsub("^mech_", "", plot_id))
    legend_text <- if (!is.null(info)) info$legend else paste("Mechanistic inference result:", tab_title)

    cat(sprintf("\n## %s\n\n", tab_title))
    cat(knitr::include_graphics(mp))
    figure_legend(legend_text)
    render_commentary(plot_id)
}
```

---

# Consensus Analysis {.tabset}

```{r consensus-note, eval=!has_consensus, results='asis'}
cat("*No consensus analysis results found. This analysis requires multiple integration methods to have completed.*\n")
```

```{r consensus-plots, eval=has_consensus, results='asis'}
cons_pngs <- list.files(file.path(consensus_dir, "plots"), pattern = "\\.png$", full.names = TRUE)
for (cp in cons_pngs) {
    plot_id <- gsub("\\.png$", "", basename(cp))
    tab_title <- gsub("_", " ", plot_id)
    cat(sprintf("\n## %s\n\n", tab_title))
    cat(knitr::include_graphics(cp))
    figure_legend(paste("Consensus analysis:", tab_title))
    render_commentary(plot_id)
}
```

```{r stability-section, eval=dir.exists(stability_dir) && length(list.files(file.path(stability_dir, "plots"), "\\.png$")) > 0, results='asis'}
cat("\n## Stability Analysis\n\n")
stab_pngs <- list.files(file.path(stability_dir, "plots"), pattern = "\\.png$", full.names = TRUE)
for (sp in stab_pngs) {
    plot_id <- gsub("\\.png$", "", basename(sp))
    cat(knitr::include_graphics(sp))
    figure_legend(paste("Stability analysis:", gsub("_", " ", plot_id)))
    render_commentary(plot_id)
    cat("\n\n")
}
```

---

# Output Files

```{r output-files, results='asis'}
cat(sprintf("All multi-omics integration results have been saved to:\n\n`%s`\n\n", run_dir))

cat("**Subdirectories:**\n\n")

subdirs <- list.dirs(run_dir, full.names = FALSE, recursive = FALSE)
subdirs <- subdirs[subdirs != "execution_info"]
if (length(subdirs) > 0) {
    for (sd in sort(subdirs)) {
        n_files <- length(list.files(file.path(run_dir, sd), recursive = TRUE))
        cat(sprintf("- **%s/** (%d files)\n", sd, n_files))
    }
}
```

---

# Methods

```{r methods, results='asis'}
cat("## Multi-Omics Integration Pipeline\n\n")

cat("This report was generated by the **multiomics-core** pipeline. ",
    "The following analytical steps were performed:\n\n")

cat("1. **Data Harmonization**: Individual omics layers (preprocessed by their respective sub-pipelines) ",
    "were aligned to a common set of samples and assembled into a `MultiAssayExperiment` object.\n\n")

methods <- multi_cfg$integration$methods %||% c("DIABLO", "SNF")
if ("DIABLO" %in% methods) {
    cat("2. **DIABLO (Data Integration Analysis for Biomarker discovery using Latent cOmponents)**: ",
        "A supervised multi-omics integration method from the mixOmics package that identifies ",
        "discriminant features across omics layers using sparse PLS-DA.\n\n")
}
if ("SNF" %in% methods) {
    cat("3. **SNF (Similarity Network Fusion)**: ",
        "An unsupervised method that constructs sample similarity networks for each omics layer ",
        "and fuses them into a single network for clustering.\n\n")
}
if ("MOFA2" %in% methods) {
    cat("4. **MOFA2 (Multi-Omics Factor Analysis v2)**: ",
        "An unsupervised Bayesian factor model that identifies latent factors capturing ",
        "shared and omics-specific sources of variation.\n\n")
}

cat("5. **Concordance Analysis**: DE results from individual omics were compared to identify ",
    "concordant and discordant features across layers.\n\n")

cat("6. **Cross-Omics Enrichment**: Pathway enrichment results were aggregated across omics ",
    "using Fisher's method for combined p-values.\n\n")

cat("## Software\n\n")
cat("```\n")
cat(capture.output(sessionInfo()), sep = "\n")
cat("\n```\n")
```
