<!-- _qc_normalization_section.Rmd — shared child document -->
<!-- Included by both metabolomics_report.Rmd and qc_summary_report.Rmd.      -->
<!-- Expects the parent to have defined qc_files and the rendering helpers     -->
<!-- (embed_png, render_interactive_pca, render_dt, %||%).                    -->
<!-- params$qc_comparison_file must be set.                                    -->

```{r nqc-child-setup, include=FALSE}
# stage_files(): filter flat qc_suite_files paths by /qc/<stage>/<subset>/ segment.
# qc_files is defined in the parent environment; referenced directly below.
stage_files <- function(files, stage, subset = "with_qc") {
  pat <- paste0("/qc/", stage, "/", subset, "/")
  files[grepl(pat, files, fixed = TRUE)]
}
```

## Benchmark

```{r nqc-benchmark}

qc_file <- params$qc_comparison_file

if (!is.null(qc_file) &&
    length(qc_file) >= 1 &&
    !is.na(qc_file[1]) &&
    nzchar(qc_file[1]) &&
    file.exists(qc_file[1])) {

  bench <- utils::read.table(
    qc_file[1],
    sep = "\t",
    header = TRUE,
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  # ---- Column reordering ------------------------------------------------
  rsd_cols   <- grep("^median_rsd", colnames(bench), value = TRUE)
  pca_cols   <- grep("^pca_",       colnames(bench), value = TRUE)
  other_cols <- setdiff(colnames(bench), c(rsd_cols, pca_cols))

  nf_pos <- match("n_features", other_cols)

  if (!is.na(nf_pos) && length(rsd_cols) > 0) {
    ordered_cols <- c(
      other_cols[seq_len(nf_pos)],
      rsd_cols,
      other_cols[seq(nf_pos + 1L, length(other_cols))],
      pca_cols
    )
  } else {
    ordered_cols <- c(other_cols, rsd_cols, pca_cols)
  }

  bench <- bench[, intersect(ordered_cols, colnames(bench)), drop = FALSE]

  # ---- Round numeric columns --------------------------------------------
  num_cols <- vapply(bench, is.numeric, logical(1L))
  bench[num_cols] <- lapply(bench[num_cols], round, 4)

  render_dt(
    bench,
    caption = paste0(
      "Normalization QC Benchmark — ",
      "raw_linear (met_imputed) → log → norm_*. ",
      "Median RSD on linear scale; lower is better. ",
      "pca_pc1_delta = PC1 gain vs log stage."
    ),
    pageLength = 10
  )
}

```

## Density {.tabset}

*Log2-scale intensity distributions per sample. Tighter, more overlapping curves indicate better between-sample normalization.*

### Log

```{r nqc-density-log, out.width="100%", echo=FALSE, results='asis'}
path <- stage_files(qc_files, "log")
cat(embed_png(path, "density"))
```

### TSS

```{r nqc-density-tss, out.width="100%", echo=FALSE, results='asis'}
path <- stage_files(qc_files, "norm_tss")
cat(embed_png(path, "density"))
```

### Median

```{r nqc-density-median, out.width="100%", echo=FALSE, results='asis'}
path <- stage_files(qc_files, "norm_median")
cat(embed_png(path, "density"))
```

### PQN

```{r nqc-density-pqn, out.width="100%", echo=FALSE, results='asis'}
path <- stage_files(qc_files, "norm_pqn")
cat(embed_png(path, "density"))
```

### EigenMS

```{r nqc-density-eigenms, out.width="100%", echo=FALSE, results='asis'}
path <- stage_files(qc_files, "norm_eigenms")
cat(embed_png(path, "density"))
```

## Sample Distance {.tabset}

*Euclidean distance between samples on log2-normalised data. Tighter within-group distances and larger between-group distances indicate better separation.*

### Log

```{r nqc-dist-log, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "log"), "distance"))
```

### TSS

```{r nqc-dist-tss, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "norm_tss"), "distance"))
```

### Median

```{r nqc-dist-median, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "norm_median"), "distance"))
```

### PQN

```{r nqc-dist-pqn, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "norm_pqn"), "distance"))
```

### EigenMS

```{r nqc-dist-eigenms, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "norm_eigenms"), "distance"))
```

## PCA (PC1/2) {.tabset}

*PCA on Log2 values. Greater separation between biological groups and tighter within-group clustering indicates better normalization. Hover over points to see sample labels.*

### Log

```{r nqc-pca-log, echo=FALSE}
render_interactive_pca(png_files = stage_files(qc_files, "log"))
```

### TSS

```{r nqc-pca-tss, echo=FALSE}
render_interactive_pca(png_files = stage_files(qc_files, "norm_tss"))
```

### Median

```{r nqc-pca-median, echo=FALSE}
render_interactive_pca(png_files = stage_files(qc_files, "norm_median"))
```

### PQN

```{r nqc-pca-pqn, echo=FALSE}
render_interactive_pca(png_files = stage_files(qc_files, "norm_pqn"))
```

### EigenMS

```{r nqc-pca-eigenms, echo=FALSE}
render_interactive_pca(png_files = stage_files(qc_files, "norm_eigenms"))
```

## Boxplots {.tabset}

*Per-sample median and spread. Well-normalized data shows aligned medians and comparable IQRs across samples.*

### Log

```{r nqc-boxplot-log, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "log"), "boxplot"))
```

### TSS

```{r nqc-boxplot-tss, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "norm_tss"), "boxplot"))
```

### Median

```{r nqc-boxplot-median, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "norm_median"), "boxplot"))
```

### PQN

```{r nqc-boxplot-pqn, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "norm_pqn"), "boxplot"))
```

### EigenMS

```{r nqc-boxplot-eigenms, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "norm_eigenms"), "boxplot"))
```

## RLE {.tabset}

*Relative Log Expression: per-sample deviation from feature medians. Well-normalized data centers around zero with tight IQRs.*

### Log

```{r nqc-rle-log, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "log"), "rle_boxplot"))
```

### TSS

```{r nqc-rle-tss, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "norm_tss"), "rle_boxplot"))
```

### Median

```{r nqc-rle-median, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "norm_median"), "rle_boxplot"))
```

### PQN

```{r nqc-rle-pqn, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "norm_pqn"), "rle_boxplot"))
```

### EigenMS

```{r nqc-rle-eigenms, out.width="100%", echo=FALSE, results='asis'}
cat(embed_png(stage_files(qc_files, "norm_eigenms"), "rle_boxplot"))
```

## QC Noise Reduction {.tabset}

*Each pair compares PCA **with** injected QC samples (all samples) vs **without** them (biological samples only).
A marked reduction in PC1 variance or tighter clustering in the no-QC panel confirms that the QC
injections inflate technical variance and were successfully identified.*

### TSS — with QC

```{r nqc-noise-tss-wq, echo=FALSE}
render_interactive_pca(png_files = stage_files(qc_files, "norm_tss", "with_qc"))
```

### TSS — without QC

```{r nqc-noise-tss-nq, echo=FALSE}
render_interactive_pca(png_files = stage_files(qc_files, "norm_tss", "no_qc"))
```

### PQN — with QC

```{r nqc-noise-pqn-wq, echo=FALSE}
render_interactive_pca(png_files = stage_files(qc_files, "norm_pqn", "with_qc"))
```

### PQN — without QC

```{r nqc-noise-pqn-nq, echo=FALSE}
render_interactive_pca(png_files = stage_files(qc_files, "norm_pqn", "no_qc"))
```

### EigenMS — with QC

```{r nqc-noise-eigenms-wq, echo=FALSE}
render_interactive_pca(png_files = stage_files(qc_files, "norm_eigenms", "with_qc"))
```

### EigenMS — without QC

```{r nqc-noise-eigenms-nq, echo=FALSE}
render_interactive_pca(png_files = stage_files(qc_files, "norm_eigenms", "no_qc"))
```
