<!-- _qc_normalization_section.Rmd — shared child document -->
<!-- Included by both metabolomics_report.Rmd and qc_summary_report.Rmd.      -->
<!-- Expects the parent to have defined qc_files and the rendering helpers     -->
<!-- (embed_png, render_dt, %||%).  params$qc_comparison_file must be set.    -->

```{r nqc-child-setup, include=FALSE}
# stage_files(): filter flat qc_suite_files paths by /qc/<stage>/<subset>/ segment.
# qc_files is defined in the parent environment; referenced directly below.
stage_files <- function(files, stage, subset = "with_qc") {
  pat <- paste0("/qc/", stage, "/", subset, "/")
  files[grepl(pat, files, fixed = TRUE)]
}
```

## Benchmark

```{r nqc-benchmark}
if (!is.null(params$qc_comparison_file) &&
    length(params$qc_comparison_file) > 0 &&
    file.exists(params$qc_comparison_file[1])) {
  bench    <- utils::read.table(params$qc_comparison_file[1], sep = "\t",
                                header = TRUE, stringsAsFactors = FALSE,
                                check.names = FALSE)
  num_cols <- vapply(bench, is.numeric, logical(1L))
  bench[num_cols] <- lapply(bench[num_cols], round, 4)
  render_dt(bench,
            caption = paste0("Normalization QC Benchmark \u2014 ",
                             "raw_linear (met_imputed) \u2192 log \u2192 norm_*. ",
                             "Median RSD on linear scale; lower is better. ",
                             "pca_pc1_delta = PC1 gain vs log stage."),
            pageLength = 10)
}
```

## Density {.tabset}

*Log2-scale intensity distributions per sample. Tighter, more overlapping curves indicate better between-sample normalization.*

### Log

```{r nqc-density-log, out.width="100%"}
embed_png(stage_files(qc_files, "log"), "density\\.png")
```

### TSS

```{r nqc-density-tss, out.width="100%"}
embed_png(stage_files(qc_files, "norm_tss"), "density\\.png")
```

### Median

```{r nqc-density-median, out.width="100%"}
embed_png(stage_files(qc_files, "norm_median"), "density\\.png")
```

### PQN

```{r nqc-density-pqn, out.width="100%"}
embed_png(stage_files(qc_files, "norm_pqn"), "density\\.png")
```

## PCA (PC1/2) {.tabset}

*PCA on Log2 values. Greater separation between biological groups and tighter within-group clustering indicates better normalization.*

### Log

```{r nqc-pca-log, out.width="100%"}
embed_png(stage_files(qc_files, "log"), "pca_pc1_pc2\\.png")
```

### TSS

```{r nqc-pca-tss, out.width="100%"}
embed_png(stage_files(qc_files, "norm_tss"), "pca_pc1_pc2\\.png")
```

### Median

```{r nqc-pca-median, out.width="100%"}
embed_png(stage_files(qc_files, "norm_median"), "pca_pc1_pc2\\.png")
```

### PQN

```{r nqc-pca-pqn, out.width="100%"}
embed_png(stage_files(qc_files, "norm_pqn"), "pca_pc1_pc2\\.png")
```

## Boxplots {.tabset}

*Per-sample median and spread. Well-normalized data shows aligned medians and comparable IQRs across samples.*

### Log

```{r nqc-box-log, out.width="100%"}
embed_png(stage_files(qc_files, "log"), "boxplot\\.png")
```

### TSS

```{r nqc-box-tss, out.width="100%"}
embed_png(stage_files(qc_files, "norm_tss"), "boxplot\\.png")
```

### Median

```{r nqc-box-median, out.width="100%"}
embed_png(stage_files(qc_files, "norm_median"), "boxplot\\.png")
```

### PQN

```{r nqc-box-pqn, out.width="100%"}
embed_png(stage_files(qc_files, "norm_pqn"), "boxplot\\.png")
```

## QC Noise Reduction {.tabset}

*Each pair compares PCA **with** injected QC samples (all samples) vs **without** them (biological samples only).
A marked reduction in PC1 variance or tighter clustering in the no-QC panel confirms that the QC
injections inflate technical variance and were successfully identified.*

### TSS — with QC

```{r nqc-noise-tss-wq, out.width="100%"}
embed_png(stage_files(qc_files, "norm_tss", "with_qc"), "pca_pc1_pc2\\.png")
```

### TSS — without QC

```{r nqc-noise-tss-nq, out.width="100%"}
embed_png(stage_files(qc_files, "norm_tss", "no_qc"), "pca_pc1_pc2\\.png")
```

### PQN — with QC

```{r nqc-noise-pqn-wq, out.width="100%"}
embed_png(stage_files(qc_files, "norm_pqn", "with_qc"), "pca_pc1_pc2\\.png")
```

### PQN — without QC

```{r nqc-noise-pqn-nq, out.width="100%"}
embed_png(stage_files(qc_files, "norm_pqn", "no_qc"), "pca_pc1_pc2\\.png")
```
