---
title: "Metabolomics Analysis Report"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
    df_print: paged
params:
  pre: NULL
  qc_res: NULL
  de_res: NULL
  clustering_res: NULL
  rf_res: NULL
  plsda_res: NULL
  enrichment_res: NULL
  config: NULL
  qc_comparison_file: NULL
  qc_suite_files: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
library(ggplot2)
library(DT)

# Helper: safely print a ggplot or return invisible(NULL)
safe_print_plot <- function(p) {
  if (inherits(p, "gg") || inherits(p, "ggplot")) {
    print(p)
  } else {
    invisible(NULL)
  }
}

# Helper: embed a PNG from a file path vector by basename pattern
embed_png <- function(files, pattern, fig_width = "100%") {
  if (is.null(files) || length(files) == 0) return(invisible(NULL))
  hits <- files[grepl(pattern, basename(files), ignore.case = TRUE)]
  hits <- hits[file.exists(hits)]
  if (length(hits) == 0) return(invisible(NULL))
  knitr::include_graphics(hits[1])
}

# Helper: render DT datatable inside parameterized Rmd
render_dt <- function(df, caption = NULL, pageLength = 10) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  htmltools::tagList(
    DT::datatable(df, caption = caption,
                  rownames = FALSE,
                  extensions = "Buttons",
                  options = list(pageLength = pageLength,
                                 scrollX = TRUE,
                                 dom = "Bfrtip",
                                 buttons = list("csv", "excel")),
                  class = "compact stripe hover")
  )
}

# Helper: interactive PCA from pca_scores.tsv written by the QC suite.
# Falls back to static PNG via knitr::asis_output() if plotly is absent.
`%||%` <- function(a, b) if (!is.null(a)) a else b
render_interactive_pca <- function(pcs = c(1, 2), png_files = NULL) {
  if (!is.null(png_files) && requireNamespace("plotly", quietly = TRUE)) {
    # Primary: explicit .tsv path in file vector
    scores_hits <- png_files[grepl("pca_scores\\.tsv$", png_files, perl = TRUE)]
    # Fallback: derive from pca_pc1_pc2.png (same directory) in case the .tsv
    # was not tracked in the file vector due to a superassignment bug or caching.
    pca_pngs <- png_files[grepl("pca_pc1_pc2\\.png$", png_files, perl = TRUE)]
    if (length(pca_pngs) > 0L)
      scores_hits <- unique(c(scores_hits, file.path(dirname(pca_pngs), "pca_scores.tsv")))
    scores_hits <- scores_hits[file.exists(scores_hits)]
    if (length(scores_hits) > 0L) {
      scores <- tryCatch(
        utils::read.table(scores_hits[1L], sep = "\t", header = TRUE,
                          stringsAsFactors = FALSE, check.names = FALSE),
        error = function(e) NULL
      )
      if (!is.null(scores)) {
        pc_x <- paste0("PC", pcs[1]); pc_y <- paste0("PC", pcs[2])
        vx   <- scores[[paste0("var_PC", pcs[1])]][1L] %||% NA_real_
        vy   <- scores[[paste0("var_PC", pcs[2])]][1L] %||% NA_real_
        reserved   <- c("sample", grep("^PC|^var_PC", colnames(scores), value = TRUE))
        extra_cols <- setdiff(colnames(scores), reserved)
        color_col  <- if (length(extra_cols) > 0L) extra_cols[1L] else "sample"
        shape_col  <- if (length(extra_cols) > 1L) extra_cols[2L] else NULL
        if (all(c(pc_x, pc_y) %in% colnames(scores))) {
          tt    <- paste0("sample: ", scores[["sample"]])
          if (color_col %in% colnames(scores) && color_col != "sample")
            tt  <- paste0(tt, "<br>", color_col, ": ", scores[[color_col]])
          if (!is.null(shape_col) && shape_col %in% colnames(scores))
            tt  <- paste0(tt, "<br>", shape_col, ": ", scores[[shape_col]])
          x_lab <- if (!is.na(vx)) sprintf("%s: %.1f%% variance", pc_x, 100*vx) else pc_x
          y_lab <- if (!is.na(vy)) sprintf("%s: %.1f%% variance", pc_y, 100*vy) else pc_y
          p <- plotly::plot_ly(data=scores,
            x=stats::as.formula(paste0("~`",pc_x,"`")),
            y=stats::as.formula(paste0("~`",pc_y,"`")),
            color=if(color_col %in% colnames(scores))
                    stats::as.formula(paste0("~factor(`",color_col,"`)")) else NULL,
            symbol=if(!is.null(shape_col) && shape_col %in% colnames(scores))
                     stats::as.formula(paste0("~factor(`",shape_col,"`)")) else NULL,
            text=tt, hoverinfo="text", type="scatter", mode="markers",
            marker=list(size=9))
          return(plotly::layout(p,
            xaxis=list(title=x_lab), yaxis=list(title=y_lab),
            legend=list(title=list(text=color_col))))
        }
      }
    }
  }
  if (!is.null(png_files)) {
    html <- embed_png(png_files, "pca_pc1_pc2")
    if (nzchar(html)) return(knitr::asis_output(html))
  }
  invisible(NULL)
}

cfg <- params$config
metab_cfg <- cfg$modes$metabolomics
```

# Summary {.tabset}

```{r summary}
pre <- params$pre
qc_res <- params$qc_res
de_res <- params$de_res
clustering_res <- params$clustering_res
rf_res <- params$rf_res
plsda_res <- params$plsda_res
enrichment_res <- params$enrichment_res

n_features <- if (!is.null(pre)) nrow(pre$expr_work) else NA
n_samples  <- if (!is.null(pre)) ncol(pre$expr_work) else NA

norm_method <- "N/A"
if (!is.null(pre$info$normalization)) {
  ni <- pre$info$normalization
  parts <- character(0)
  if (!is.null(ni$sample_norm) && tolower(ni$sample_norm) != "none")
    parts <- c(parts, toupper(ni$sample_norm))
  if (!is.null(ni$transform) && tolower(ni$transform) != "none")
    parts <- c(parts, ni$transform)
  if (!is.null(ni$scaling) && tolower(ni$scaling) != "none")
    parts <- c(parts, ni$scaling)
  if (length(parts) > 0) norm_method <- paste(parts, collapse = " + ")
}

de_method <- if (!is.null(de_res$method)) de_res$method else "N/A"

module_status <- data.frame(
  Module = c("Preprocessing", "QC Diagnostics", "Differential Expression",
             "Random Forest", "PLS-DA", "Enrichment"),
  Status = c(
    if (!is.null(pre)) "Completed" else "Skipped",
    if (!is.null(qc_res)) "Completed" else "Skipped",
    if (!is.null(de_res) && !is.null(de_res$summary_df)) "Completed" else "Skipped",
    if (!is.null(rf_res)) "Completed" else "Skipped",
    if (!is.null(plsda_res)) "Completed" else "Skipped",
    if (!is.null(enrichment_res)) "Completed" else "Skipped"
  ),
  stringsAsFactors = FALSE
)
```

| Parameter | Value |
|-----------|-------|
| Features  | `r n_features` |
| Samples   | `r n_samples` |
| Normalization | `r norm_method` |
| DE method | `r de_method` |

```{r summary-table}
knitr::kable(module_status, align = "ll")
```

---

`r if (is.null(params$qc_comparison_file) || length(params$qc_comparison_file) == 0) "<!--"`

# Normalization QC {.tabset}

```{r nqc-setup, include=FALSE}
qc_files <- params$qc_suite_files %||% character(0)
```

```{r qc-norm-section, child='_qc_normalization_section.Rmd'}
```

`r if (is.null(params$qc_comparison_file) || length(params$qc_comparison_file) == 0) "-->"`

---

`r if (is.null(qc_res)) "<!--"`

# QC Diagnostics {.tabset}

## PCA

```{r qc-pca, fig.width=9, fig.height=7}
if (!is.null(qc_res$plots)) {
  # Find the primary PCA plot (no _noQC, no extra color suffix)
  pca_keys <- grep("^pca_1_2", names(qc_res$plots), value = TRUE)
  pca_main <- pca_keys[!grepl("_noQC$", pca_keys)][1]
  if (!is.na(pca_main)) safe_print_plot(qc_res$plots[[pca_main]])

  pca13_keys <- grep("^pca_1_3", names(qc_res$plots), value = TRUE)
  pca13_main <- pca13_keys[!grepl("_noQC$", pca13_keys)][1]
  if (!is.na(pca13_main)) safe_print_plot(qc_res$plots[[pca13_main]])
}
```

## Density

```{r qc-density, fig.width=9, fig.height=6}
if (!is.null(qc_res$plots)) {
  safe_print_plot(qc_res$plots[["density_raw"]])
  safe_print_plot(qc_res$plots[["density_log2raw"]])
  safe_print_plot(qc_res$plots[["density_norm"]])
}
```

## Boxplots

```{r qc-boxplots, fig.width=10, fig.height=6}
if (!is.null(qc_res$plots)) {
  safe_print_plot(qc_res$plots[["boxplot_raw"]])
  safe_print_plot(qc_res$plots[["boxplot"]])
}
```

## Heatmaps

```{r qc-heatmaps, out.width="100%"}
if (!is.null(qc_res$files)) {
  embed_png(qc_res$files, "sample_distance_heatmap\\.png$")
  embed_png(qc_res$files, "sample_correlation_heatmap\\.png$")
}
```

`r if (is.null(qc_res)) "-->"`

---

`r if (is.null(de_res) || is.null(de_res$summary_df)) "<!--"`

# Differential Expression {.tabset}

```{r de-setup}
de_contrasts <- NULL
if (!is.null(de_res) && !is.null(de_res$de_tables)) {
  de_cfg <- metab_cfg$de %||% list()
  # Contrast labels from de_tables names (set by run_metabolomics_de)
  de_contrasts <- names(de_res$de_tables)
}
```

```{r de-note, results='asis'}
if (!is.null(de_res) && !is.null(de_res$summary_df)) {
  n_sig_any <- sum(de_res$summary_df$pass_any_contrast == 1, na.rm = TRUE)
  if (n_sig_any == 0) {
    n_samp <- ncol(pre$expr_work)
    cat("\n> **Note:** No features reached significance (adjusted p < ",
        metab_cfg$de$p_cutoff %||% 0.05, " and |log2FC| >= ",
        round(log2(metab_cfg$de$linear_fc_cutoff %||% 1.5), 2),
        "). With only **", n_samp, " samples**, the t-test has limited power ",
        "to produce small p-values, resulting in shallow −log10(FDR) values ",
        "on the volcano plot y-axis. Consider increasing sample size, relaxing ",
        "thresholds, or using a rank-based method (e.g. GSEA) which does not ",
        "require a hard significance cutoff.\n\n", sep = "")
  }
}
```

```{r de-tabs, results='asis'}
if (!is.null(de_contrasts)) {
  for (ctr in de_contrasts) {
    # ctr is already a normalized label from de_res$de_tables names
    ctr_label <- ctr

    cat("\n## ", ctr, " {.tabset}\n\n", sep = "")

    # Volcano
    p_volcano <- de_res$plots[[paste0("volcano_", ctr_label)]]
    if (!is.null(p_volcano)) {
      cat("### Volcano\n\n")
      print(p_volcano)
      cat("\n\n")
    }

    # MA
    p_ma <- de_res$plots[[paste0("ma_", ctr_label)]]
    if (!is.null(p_ma)) {
      cat("### MA Plot\n\n")
      print(p_ma)
      cat("\n\n")
    }

    # P-value histogram
    p_phist <- de_res$plots[[paste0("pval_hist_", ctr_label)]]
    if (!is.null(p_phist)) {
      cat("### P-value Distribution\n\n")
      print(p_phist)
      cat("\n\n")
    }

    # Interactive volcano (plotly with searchable metabolite selector)
    ctr_tbl <- tryCatch(
      extract_contrast_table(de_res$summary_df, ctr_label),
      error = function(e) NULL
    )
    if (!is.null(ctr_tbl) &&
        requireNamespace("plotly", quietly = TRUE) &&
        requireNamespace("crosstalk", quietly = TRUE)) {

      # Attach compound names
      name_map_v <- NULL
      if (!is.null(pre$row_data) && "Name" %in% colnames(pre$row_data)) {
        name_map_v <- stats::setNames(
          as.character(pre$row_data$Name),
          as.character(pre$row_data$feature_id)
        )
      }

      vdf <- ctr_tbl
      vdf$Name <- if (!is.null(name_map_v)) {
        ifelse(is.na(name_map_v[vdf$feature_id]),
               vdf$feature_id, name_map_v[vdf$feature_id])
      } else {
        vdf$feature_id
      }
      vdf$neglog10_FDR <- -log10(pmax(
        ifelse(is.na(vdf$adj.P.Val), 1, vdf$adj.P.Val), 1e-300
      ))
      vdf$logFC_display <- round(vdf$logFC, 4)
      vdf$FDR_display   <- signif(vdf$adj.P.Val, 3)

      cat("### Interactive Volcano\n\n")
      cat("*Type metabolite names in the search box to highlight them. ",
          "Hover over any point to see details. Use lasso/box select on the plot.*\n\n")

      sd <- crosstalk::SharedData$new(vdf, key = ~Name, group = paste0("volcano_", ctr_label))

      vol_fig <- plotly::plot_ly(
        sd, x = ~logFC, y = ~neglog10_FDR,
        type = "scatter", mode = "markers",
        text = ~paste0(Name, "\nlog2FC: ", logFC_display, "\nFDR: ", FDR_display),
        hoverinfo = "text",
        marker = list(size = 7, opacity = 0.5, color = "grey30")
      )
      vol_fig <- plotly::layout(
        vol_fig,
        title = list(text = paste0("Volcano: ", ctr), font = list(size = 14)),
        xaxis = list(title = "log2 Fold Change"),
        yaxis = list(title = "-log10(FDR)"),
        dragmode = "lasso"
      )
      vol_fig <- plotly::highlight(
        vol_fig,
        on = "plotly_selected",
        off = "plotly_deselect",
        selectize = TRUE,
        persistent = TRUE,
        color = "dodgerblue",
        opacityDim = 0.15,
        selected = plotly::attrs_selected(
          marker = list(size = 10, opacity = 1),
          showlegend = FALSE
        )
      )

      print(htmltools::tagList(vol_fig))
      cat("\n\n")
    }

    # Top results table
    if (!is.null(ctr_tbl)) {
      ctr_tbl <- ctr_tbl[order(ctr_tbl$P.Value), ]
      top_tbl <- head(ctr_tbl, 50)
      cat("### Top Results\n\n")
      print(render_dt(top_tbl, caption = paste0("Top DE features: ", ctr)))
      cat("\n\n")
    }
  }
}
```

`r if (is.null(de_res) || is.null(de_res$summary_df)) "-->"`

---

`r if (is.null(clustering_res) || length(clustering_res$plots) == 0) "<!--"`

# Clustering {.tabset}

## Hierarchical Clustering

```{r clustering-hier, fig.width=10, fig.height=8}
if (!is.null(clustering_res) && !is.null(clustering_res$plots$p_cluster_hier)) {
  p <- clustering_res$plots$p_cluster_hier
  if (inherits(p, "pheatmap")) {
    grid::grid.newpage()
    grid::grid.draw(p$gtable)
  } else {
    safe_print_plot(p)
  }
}
```

```{r clustering-hier-info, results='asis'}
if (!is.null(clustering_res) && !is.null(clustering_res$objects$hm_hier_de)) {
  n_feat <- length(clustering_res$objects$hm_hier_de$feature_ids)
  cat(sprintf("\n\nHierarchical clustering on **%d DE features**.\n\n", n_feat))
}
```

`r if (is.null(clustering_res) || is.null(clustering_res$plots$partition_heatmap)) "<!--"`

## Partition Clustering

```{r clustering-partition, fig.width=10, fig.height=8}
if (!is.null(clustering_res) && !is.null(clustering_res$plots$partition_heatmap)) {
  p <- clustering_res$plots$partition_heatmap
  if (inherits(p, "pheatmap")) {
    grid::grid.newpage()
    grid::grid.draw(p$gtable)
  } else {
    safe_print_plot(p)
  }
}
```

```{r clustering-profiles, fig.width=10, fig.height=7}
if (!is.null(clustering_res) && !is.null(clustering_res$plots$cluster_profiles)) {
  safe_print_plot(clustering_res$plots$cluster_profiles)
}
```

`r if (is.null(clustering_res) || is.null(clustering_res$plots$partition_heatmap)) "-->"`

`r if (is.null(clustering_res) || length(clustering_res$plots) == 0) "-->"`

---

`r if (is.null(rf_res) && is.null(plsda_res)) "<!--"`

# Feature Selection {.tabset}

`r if (is.null(rf_res)) "<!--"`

## Random Forest {.tabset}

### Importance Plot

```{r rf-plot, fig.width=9, fig.height=7}
if (!is.null(rf_res$plots$rf_importance)) {
  safe_print_plot(rf_res$plots$rf_importance)
}
```

### Importance Table

```{r rf-table}
if (!is.null(rf_res$importance_df)) {
  imp_df <- rf_res$importance_df
  sort_col <- if ("MeanDecreaseAccuracy" %in% names(imp_df)) "MeanDecreaseAccuracy" else "importance"
  imp_df <- imp_df[order(-imp_df[[sort_col]]), ]
  render_dt(head(imp_df, 50), caption = "RF Feature Importance (top 50)")
}
```

`r if (is.null(rf_res)) "-->"`

`r if (is.null(plsda_res)) "<!--"`

## PLS-DA {.tabset}

### Scores Plot

```{r plsda-scores, fig.width=9, fig.height=7}
if (!is.null(plsda_res$plots$plsda_scores)) {
  safe_print_plot(plsda_res$plots$plsda_scores)
}
```

### VIP Plot

```{r plsda-vip-plot, fig.width=9, fig.height=7}
if (!is.null(plsda_res$plots$plsda_vip)) {
  safe_print_plot(plsda_res$plots$plsda_vip)
}
```

### VIP Table

```{r plsda-vip-table}
if (!is.null(plsda_res$vip_df)) {
  vip_df <- plsda_res$vip_df
  vip_col <- if ("VIP" %in% names(vip_df)) "VIP" else grep("^VIP", names(vip_df), value = TRUE)[1]
  vip_df <- vip_df[order(-vip_df[[vip_col]]), ]
  render_dt(head(vip_df, 50), caption = "PLS-DA VIP Scores (top 50)")
}
```

`r if (is.null(plsda_res)) "-->"`

`r if (is.null(rf_res) && is.null(plsda_res)) "-->"`

---

`r if (is.null(enrichment_res)) "<!--"`

# Enrichment {.tabset}

`r if (is.null(enrichment_res$qea)) "<!--"`

## QEA {.tabset}

### Lollipop Plot

```{r qea-lollipop, fig.width=12, fig.height=8}
if (!is.null(enrichment_res$plots$qea_lollipop)) {
  safe_print_plot(enrichment_res$plots$qea_lollipop)
}
```

### Barplot

```{r qea-barplot, fig.width=12, fig.height=8}
if (!is.null(enrichment_res$plots$qea_barplot)) {
  safe_print_plot(enrichment_res$plots$qea_barplot)
}
```

### Results Table

```{r qea-table}
if (!is.null(enrichment_res$qea$table)) {
  render_dt(enrichment_res$qea$table, caption = "QEA Pathway Enrichment")
}
```

`r if (is.null(enrichment_res$qea)) "-->"`

`r if (is.null(enrichment_res$ssgsea)) "<!--"`

## ssGSEA {.tabset}

### Lollipop Plot

```{r ssgsea-lollipop, fig.width=12, fig.height=8}
if (!is.null(enrichment_res$plots$ssgsea_lollipop)) {
  safe_print_plot(enrichment_res$plots$ssgsea_lollipop)
}
```

### Barplot

```{r ssgsea-barplot, fig.width=12, fig.height=8}
if (!is.null(enrichment_res$plots$ssgsea_barplot)) {
  safe_print_plot(enrichment_res$plots$ssgsea_barplot)
}
```

### Boxplots (static)

```{r ssgsea-boxplots, fig.width=12, fig.height=8}
if (!is.null(enrichment_res$plots$ssgsea_boxplots)) {
  safe_print_plot(enrichment_res$plots$ssgsea_boxplots)
}
```

### Pathway Explorer

```{r ssgsea-interactive, fig.width=10, fig.height=6}
# Interactive boxplot: select one or more pathways, compare conditions
if (!is.null(enrichment_res$ssgsea$scores) &&
    !is.null(enrichment_res$ssgsea$table)) {

  scores_mat <- enrichment_res$ssgsea$scores
  ssgsea_tbl <- enrichment_res$ssgsea$table

  # Resolve condition labels for each sample
  sample_col_i <- metab_cfg$effects$samples %||% "sample_id"
  cond_col_i   <- metab_cfg$enrichment$condition_column %||%
                   metab_cfg$de$condition_column %||%
                   metab_cfg$effects$color %||% "sample_type"
  meta_i       <- params$pre$meta
  conditions_i <- as.character(meta_i[[cond_col_i]][
    match(colnames(scores_mat), meta_i[[sample_col_i]])
  ])

  # Build long-format data for all pathways
  all_pathways <- rownames(scores_mat)
  pw_order     <- ssgsea_tbl$pathway[ssgsea_tbl$pathway %in% all_pathways]
  # Fall back to rowname order for pathways not in the table
  pw_order     <- c(pw_order, setdiff(all_pathways, pw_order))

  long_df <- do.call(rbind, lapply(pw_order, function(pw) {
    data.frame(
      pathway   = pw,
      sample    = colnames(scores_mat),
      score     = as.numeric(scores_mat[pw, ]),
      condition = conditions_i,
      stringsAsFactors = FALSE
    )
  }))

  # Use plotly with a dropdown to select pathways
  if (requireNamespace("plotly", quietly = TRUE)) {

    # Pre-select top 3 pathways by FDR
    default_pws <- utils::head(pw_order, 3)

    # Build one trace per pathway (visible for defaults, hidden for rest)
    fig <- plotly::plot_ly()

    for (pw in pw_order) {
      pw_data <- long_df[long_df$pathway == pw, ]
      vis <- pw %in% default_pws

      fig <- plotly::add_trace(
        fig,
        data = pw_data,
        x = ~condition,
        y = ~score,
        type = "box",
        boxpoints = "all",
        jitter = 0.4,
        pointpos = 0,
        text = ~sample,
        name = pw,
        visible = vis,
        marker = list(size = 6),
        showlegend = FALSE
      )
    }

    # Build dropdown buttons: each button shows one pathway
    buttons <- lapply(seq_along(pw_order), function(i) {
      vis_vec <- rep(FALSE, length(pw_order))
      vis_vec[i] <- TRUE
      pw_short <- if (nchar(pw_order[i]) > 60) {
        paste0(substr(pw_order[i], 1, 57), "...")
      } else {
        pw_order[i]
      }
      list(method = "restyle",
           args = list("visible", as.list(vis_vec)),
           label = pw_short)
    })

    # Add a "Top 3" button at the beginning
    top3_vis <- pw_order %in% default_pws
    buttons <- c(
      list(list(method = "restyle",
                args = list("visible", as.list(top3_vis)),
                label = paste0("Top 3 by FDR"))),
      buttons
    )

    fig <- plotly::layout(
      fig,
      title = list(text = "ssGSEA Scores by Condition",
                    font = list(size = 16)),
      xaxis = list(title = "Condition"),
      yaxis = list(title = "ssGSEA Score"),
      updatemenus = list(
        list(
          type = "dropdown",
          active = 0,
          x = 0.0, xanchor = "left",
          y = 1.15, yanchor = "top",
          buttons = buttons
        )
      )
    )

    fig
  } else {
    cat("*Install the `plotly` package for interactive pathway exploration.*\n")
  }
}
```

### Results Table

```{r ssgsea-table}
if (!is.null(enrichment_res$ssgsea$table)) {
  render_dt(enrichment_res$ssgsea$table, caption = "ssGSEA Pathway Enrichment")
}
```

`r if (is.null(enrichment_res$ssgsea)) "-->"`

`r if (is.null(enrichment_res$ora)) "<!--"`

## ORA {.tabset}

### Lollipop Plot

```{r ora-lollipop, fig.width=12, fig.height=8}
if (!is.null(enrichment_res$plots$ora_lollipop)) {
  safe_print_plot(enrichment_res$plots$ora_lollipop)
}
```

### Dot Plot

```{r ora-dotplot, fig.width=12, fig.height=8}
if (!is.null(enrichment_res$plots$ora_dotplot)) {
  safe_print_plot(enrichment_res$plots$ora_dotplot)
}
```

### Results Table

```{r ora-table}
if (!is.null(enrichment_res$ora$table)) {
  render_dt(enrichment_res$ora$table, caption = "ORA Pathway Enrichment (Fisher's Exact Test)")
}
```

`r if (is.null(enrichment_res$ora)) "-->"`

`r if (is.null(enrichment_res$gsea)) "<!--"`

## GSEA {.tabset}

### Lollipop Plot

```{r gsea-lollipop, fig.width=12, fig.height=8}
if (!is.null(enrichment_res$plots$gsea_lollipop)) {
  safe_print_plot(enrichment_res$plots$gsea_lollipop)
}
```

### NES Barplot

```{r gsea-nes-barplot, fig.width=12, fig.height=8}
if (!is.null(enrichment_res$plots$gsea_nes_barplot)) {
  safe_print_plot(enrichment_res$plots$gsea_nes_barplot)
}
```

### Results Table

```{r gsea-table}
if (!is.null(enrichment_res$gsea$table)) {
  render_dt(enrichment_res$gsea$table, caption = "GSEA Pathway Enrichment (fgsea, ranked by log2FC)")
}
```

`r if (is.null(enrichment_res$gsea)) "-->"`

`r if (is.null(enrichment_res)) "-->"`

---

# Session Info

<details>
<summary>Click to expand</summary>

```{r session-info}
sessionInfo()
```

</details>
