---
title: "Metabolomics Analysis Report"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    theme: united
    code_folding: hide
    df_print: paged
params:
  pre: NULL
  qc_res: NULL
  de_res: NULL
  rf_res: NULL
  plsda_res: NULL
  enrichment_res: NULL
  config: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
library(ggplot2)
library(DT)

# Helper: safely print a ggplot or return invisible(NULL)
safe_print_plot <- function(p) {
  if (inherits(p, "gg") || inherits(p, "ggplot")) {
    print(p)
  } else {
    invisible(NULL)
  }
}

# Helper: embed a PNG from a file path vector by basename pattern
embed_png <- function(files, pattern, fig_width = "100%") {
  if (is.null(files) || length(files) == 0) return(invisible(NULL))
  hits <- files[grepl(pattern, basename(files), ignore.case = TRUE)]
  hits <- hits[file.exists(hits)]
  if (length(hits) == 0) return(invisible(NULL))
  knitr::include_graphics(hits[1])
}

# Helper: render DT datatable inside parameterized Rmd
render_dt <- function(df, caption = NULL, pageLength = 10) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  htmltools::tagList(
    DT::datatable(df, caption = caption,
                  rownames = FALSE,
                  options = list(pageLength = pageLength,
                                 scrollX = TRUE),
                  class = "compact stripe hover")
  )
}

cfg <- params$config
metab_cfg <- cfg$modes$metabolomics
```

# Summary {.tabset}

```{r summary}
pre <- params$pre
qc_res <- params$qc_res
de_res <- params$de_res
rf_res <- params$rf_res
plsda_res <- params$plsda_res
enrichment_res <- params$enrichment_res

n_features <- if (!is.null(pre)) nrow(pre$expr_work) else NA
n_samples  <- if (!is.null(pre)) ncol(pre$expr_work) else NA

norm_method <- "N/A"
if (!is.null(pre$info$normalization)) {
  ni <- pre$info$normalization
  parts <- character(0)
  if (!is.null(ni$sample_norm) && tolower(ni$sample_norm) != "none")
    parts <- c(parts, toupper(ni$sample_norm))
  if (!is.null(ni$transform) && tolower(ni$transform) != "none")
    parts <- c(parts, ni$transform)
  if (!is.null(ni$scaling) && tolower(ni$scaling) != "none")
    parts <- c(parts, ni$scaling)
  if (length(parts) > 0) norm_method <- paste(parts, collapse = " + ")
}

de_method <- if (!is.null(de_res$method)) de_res$method else "N/A"

module_status <- data.frame(
  Module = c("Preprocessing", "QC Diagnostics", "Differential Expression",
             "Random Forest", "PLS-DA", "Enrichment"),
  Status = c(
    if (!is.null(pre)) "Completed" else "Skipped",
    if (!is.null(qc_res)) "Completed" else "Skipped",
    if (!is.null(de_res) && !is.null(de_res$summary_df)) "Completed" else "Skipped",
    if (!is.null(rf_res)) "Completed" else "Skipped",
    if (!is.null(plsda_res)) "Completed" else "Skipped",
    if (!is.null(enrichment_res)) "Completed" else "Skipped"
  ),
  stringsAsFactors = FALSE
)
```

| Parameter | Value |
|-----------|-------|
| Features  | `r n_features` |
| Samples   | `r n_samples` |
| Normalization | `r norm_method` |
| DE method | `r de_method` |

```{r summary-table}
knitr::kable(module_status, align = "ll")
```

---

`r if (is.null(qc_res)) "<!--"`

# QC Diagnostics {.tabset}

## PCA

```{r qc-pca, fig.width=9, fig.height=7}
if (!is.null(qc_res$plots)) {
  # Find the primary PCA plot (no _noQC, no extra color suffix)
  pca_keys <- grep("^pca_1_2", names(qc_res$plots), value = TRUE)
  pca_main <- pca_keys[!grepl("_noQC$", pca_keys)][1]
  if (!is.na(pca_main)) safe_print_plot(qc_res$plots[[pca_main]])

  pca13_keys <- grep("^pca_1_3", names(qc_res$plots), value = TRUE)
  pca13_main <- pca13_keys[!grepl("_noQC$", pca13_keys)][1]
  if (!is.na(pca13_main)) safe_print_plot(qc_res$plots[[pca13_main]])
}
```

## Density

```{r qc-density, fig.width=9, fig.height=6}
if (!is.null(qc_res$plots)) {
  safe_print_plot(qc_res$plots[["density_raw"]])
  safe_print_plot(qc_res$plots[["density_norm"]])
}
```

## Boxplots

```{r qc-boxplots, fig.width=10, fig.height=6}
if (!is.null(qc_res$plots)) {
  safe_print_plot(qc_res$plots[["boxplot_raw"]])
  safe_print_plot(qc_res$plots[["boxplot"]])
}
```

## Heatmaps

```{r qc-heatmaps, out.width="100%"}
if (!is.null(qc_res$files)) {
  embed_png(qc_res$files, "sample_distance_heatmap\\.png$")
  embed_png(qc_res$files, "sample_correlation_heatmap\\.png$")
}
```

`r if (is.null(qc_res)) "-->"`

---

`r if (is.null(de_res) || is.null(de_res$summary_df)) "<!--"`

# Differential Expression {.tabset}

```{r de-setup}
de_contrasts <- NULL
if (!is.null(de_res) && !is.null(de_res$summary_df)) {
  de_cfg <- metab_cfg$de %||% list()
  de_contrasts <- de_cfg$contrasts
  if (is.list(de_contrasts)) de_contrasts <- unlist(de_contrasts)
}
```

```{r de-tabs, results='asis'}
if (!is.null(de_contrasts)) {
  for (ctr in de_contrasts) {
    ctr_label <- make_contrast_label(ctr)

    cat("\n## ", ctr, " {.tabset}\n\n", sep = "")

    # Volcano
    p_volcano <- de_res$plots[[paste0("volcano_", ctr_label)]]
    if (!is.null(p_volcano)) {
      cat("### Volcano\n\n")
      print(p_volcano)
      cat("\n\n")
    }

    # MA
    p_ma <- de_res$plots[[paste0("ma_", ctr_label)]]
    if (!is.null(p_ma)) {
      cat("### MA Plot\n\n")
      print(p_ma)
      cat("\n\n")
    }

    # P-value histogram
    p_phist <- de_res$plots[[paste0("pval_hist_", ctr_label)]]
    if (!is.null(p_phist)) {
      cat("### P-value Distribution\n\n")
      print(p_phist)
      cat("\n\n")
    }

    # Top results table
    ctr_tbl <- tryCatch(
      extract_contrast_table(de_res$summary_df, ctr_label),
      error = function(e) NULL
    )
    if (!is.null(ctr_tbl)) {
      ctr_tbl <- ctr_tbl[order(ctr_tbl$P.Value), ]
      top_tbl <- head(ctr_tbl, 50)
      cat("### Top Results\n\n")
      print(
        htmltools::tagList(
          DT::datatable(top_tbl,
                        caption = paste0("Top DE features: ", ctr),
                        rownames = FALSE,
                        options = list(pageLength = 10, scrollX = TRUE),
                        class = "compact stripe hover")
        )
      )
      cat("\n\n")
    }
  }
}
```

`r if (is.null(de_res) || is.null(de_res$summary_df)) "-->"`

---

`r if (is.null(rf_res) && is.null(plsda_res)) "<!--"`

# Feature Selection {.tabset}

`r if (is.null(rf_res)) "<!--"`

## Random Forest {.tabset}

### Importance Plot

```{r rf-plot, fig.width=9, fig.height=7}
if (!is.null(rf_res$plots$rf_importance)) {
  safe_print_plot(rf_res$plots$rf_importance)
}
```

### Importance Table

```{r rf-table}
if (!is.null(rf_res$importance_df)) {
  imp_df <- rf_res$importance_df
  sort_col <- if ("MeanDecreaseAccuracy" %in% names(imp_df)) "MeanDecreaseAccuracy" else "importance"
  imp_df <- imp_df[order(-imp_df[[sort_col]]), ]
  render_dt(head(imp_df, 50), caption = "RF Feature Importance (top 50)")
}
```

`r if (is.null(rf_res)) "-->"`

`r if (is.null(plsda_res)) "<!--"`

## PLS-DA {.tabset}

### Scores Plot

```{r plsda-scores, fig.width=9, fig.height=7}
if (!is.null(plsda_res$plots$plsda_scores)) {
  safe_print_plot(plsda_res$plots$plsda_scores)
}
```

### VIP Plot

```{r plsda-vip-plot, fig.width=9, fig.height=7}
if (!is.null(plsda_res$plots$plsda_vip)) {
  safe_print_plot(plsda_res$plots$plsda_vip)
}
```

### VIP Table

```{r plsda-vip-table}
if (!is.null(plsda_res$vip_df)) {
  vip_df <- plsda_res$vip_df
  vip_col <- if ("VIP" %in% names(vip_df)) "VIP" else grep("^VIP", names(vip_df), value = TRUE)[1]
  vip_df <- vip_df[order(-vip_df[[vip_col]]), ]
  render_dt(head(vip_df, 50), caption = "PLS-DA VIP Scores (top 50)")
}
```

`r if (is.null(plsda_res)) "-->"`

`r if (is.null(rf_res) && is.null(plsda_res)) "-->"`

---

`r if (is.null(enrichment_res)) "<!--"`

# Enrichment {.tabset}

`r if (is.null(enrichment_res$qea)) "<!--"`

## QEA {.tabset}

### Barplot

```{r qea-barplot, fig.width=12, fig.height=8}
if (!is.null(enrichment_res$plots$qea_barplot)) {
  safe_print_plot(enrichment_res$plots$qea_barplot)
}
```

### Results Table

```{r qea-table}
if (!is.null(enrichment_res$qea$table)) {
  render_dt(enrichment_res$qea$table, caption = "QEA Pathway Enrichment")
}
```

`r if (is.null(enrichment_res$qea)) "-->"`

`r if (is.null(enrichment_res$ssgsea)) "<!--"`

## ssGSEA {.tabset}

### Barplot

```{r ssgsea-barplot, fig.width=12, fig.height=8}
if (!is.null(enrichment_res$plots$ssgsea_barplot)) {
  safe_print_plot(enrichment_res$plots$ssgsea_barplot)
}
```

### Boxplots

```{r ssgsea-boxplots, fig.width=12, fig.height=8}
if (!is.null(enrichment_res$plots$ssgsea_boxplots)) {
  safe_print_plot(enrichment_res$plots$ssgsea_boxplots)
}
```

### Results Table

```{r ssgsea-table}
if (!is.null(enrichment_res$ssgsea$table)) {
  render_dt(enrichment_res$ssgsea$table, caption = "ssGSEA Pathway Enrichment")
}
```

`r if (is.null(enrichment_res$ssgsea)) "-->"`

`r if (is.null(enrichment_res)) "-->"`

---

# Session Info

<details>
<summary>Click to expand</summary>

```{r session-info}
sessionInfo()
```

</details>
