---
title: "Metabolomics — Normalization QC Review"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    theme: lumen
    code_folding: hide
    df_print: paged
params:
  qc_comparison_file: NULL
  qc_suite_files: NULL
  config: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center"
)
library(DT)
library(base64enc)

# Null-coalescing operator (mirrors pipeline convention)
`%||%` <- function(a, b) if (!is.null(a)) a else b

embed_png <- function(files, pattern, fig_width = "100%") {
  if (is.null(files) || length(files) == 0) return("")
  hits <- files[grepl(pattern, basename(files), fixed = TRUE, ignore.case = TRUE)]
  hits <- hits[file.exists(hits)]

  if (length(hits) == 0) return("")
  
  img_64 <- base64enc::dataURI(file = hits[1], mime = "image/png")
  
  return(sprintf('<img src="%s" style="width:%s; display:block; margin:auto;" />', 
                 img_64, fig_width))

}

render_dt <- function(df, caption = NULL, pageLength = 10) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  htmltools::tagList(
    DT::datatable(df, caption = caption,
                  rownames = FALSE,
                  extensions = "Buttons",
                  options = list(pageLength = pageLength,
                                 scrollX = TRUE,
                                 dom = "Bfrtip",
                                 buttons = list("csv", "excel")),
                  class = "compact stripe hover")
  )
}

# qc_files is referenced by _qc_normalization_section.Rmd via parent environment.
qc_files <- params$qc_suite_files %||% character(0)
```

> **Purpose:** Review normalization method performance before interpreting
> differential expression and enrichment results.  The **Benchmark** tab shows
> RSD and PC1 variance for every normalization stage, starting from the raw
> linear baseline (`met_imputed`).  Choose the method with the lowest
> `median_rsd_overall` and the highest `pca_pc1_delta`.

# Normalization QC {.tabset}

```{r qc-section, child='_qc_normalization_section.Rmd', eval=!is.null(params$qc_comparison_file) && length(params$qc_comparison_file) > 0}
```

```{r no-data-msg, results='asis', echo=FALSE, eval=is.null(params$qc_comparison_file) || length(params$qc_comparison_file) == 0}
cat("\n> **No QC data available.**  Run the pipeline with `qc.enabled: true`",
    "to generate the normalization QC outputs before rendering this report.\n")
```

---

# Session Info

<details>
<summary>Click to expand</summary>

```{r session-info}
sessionInfo()
```

</details>
