---
title: "Metabolomics — Normalization QC Review"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    theme: lumen
    code_folding: hide
    df_print: paged
params:
  qc_comparison_file: NULL
  qc_suite_files: NULL
  config: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center"
)
library(DT)
library(base64enc)

# Null-coalescing operator (mirrors pipeline convention)
`%||%` <- function(a, b) if (!is.null(a)) a else b

# ---------------------------------------------------------------------------
# embed_png — Base64-encode a PNG file for self-contained HTML output.
# API is unchanged; only called when interactive PCA is unavailable.
# ---------------------------------------------------------------------------
embed_png <- function(files, pattern, fig_width = "100%") {
  if (is.null(files) || length(files) == 0) return("")
  hits <- files[grepl(pattern, basename(files), fixed = TRUE, ignore.case = TRUE)]
  hits <- hits[file.exists(hits)]
  if (length(hits) == 0) return("")
  img_64 <- base64enc::dataURI(file = hits[1], mime = "image/png")
  sprintf('<img src="%s" style="width:%s; display:block; margin:auto;" />',
          img_64, fig_width)
}

# ---------------------------------------------------------------------------
# render_interactive_pca — interactive plotly PCA scatter.
#
# Primary path  (mat + meta + cfg_mode supplied):
#   Calls qc_pca_scatter() without file I/O to obtain a ggplot, enriches
#   it with a text aesthetic (sample_id + group), then wraps in
#   plotly::ggplotly().  Requires the 'plotly' package.
#
# Scores-file path (png_files supplied, pca_scores.tsv present alongside PNG):
#   Reads the TSV saved by qc_full_metabolomics_suite() — columns:
#   sample, PC1, PC2[, PC3], <color_col>, var_PC1, var_PC2[, var_PC3].
#   Builds a plotly scatter with hover labels directly from the file.
#
# Final fallback (plotly absent or no scores file found):
#   Embeds the pre-rendered pca_pc1_pc2.png via embed_png().
#   Uses knitr::asis_output() so raw HTML renders without results='asis'.
#
# @param mat       features x samples numeric matrix  (or NULL)
# @param meta      sample metadata data.frame          (or NULL)
# @param cfg_mode  config$modes$metabolomics           (or NULL)
# @param pcs       length-2 integer vector of PCs, e.g. c(1, 2)
# @param png_files character vector of file paths — used for scores + PNG fallback
# @return plotly widget, knitr::asis_output (PNG), or NULL (silent)
# ---------------------------------------------------------------------------
render_interactive_pca <- function(mat      = NULL,
                                   meta     = NULL,
                                   cfg_mode = NULL,
                                   pcs      = c(1, 2),
                                   png_files = NULL) {

  # ---- Primary: live data path -------------------------------------------
  if (!is.null(mat) && !is.null(meta) && !is.null(cfg_mode)) {
    if (!requireNamespace("plotly", quietly = TRUE)) {
      message("render_interactive_pca: 'plotly' not installed; using static PNG.")
    } else {
      p <- tryCatch(
        qc_pca_scatter(mat, meta, cfg_mode, pcs = pcs, out_file = NULL),
        error = function(e) {
          message("render_interactive_pca: qc_pca_scatter failed: ", e$message)
          NULL
        }
      )

      if (!is.null(p)) {
        sample_col <- cfg_mode$effects$samples %||% "sample_id"
        color_col  <- cfg_mode$effects$color   %||% sample_col
        scores     <- p$data   # plot_pca_scatter uses scores as ggplot data

        if (!is.null(scores) && is.data.frame(scores)) {
          id_vals <- if ("sample" %in% colnames(scores)) scores[["sample"]] else
                         seq_len(nrow(scores))
          grp_col <- intersect(color_col, colnames(scores))
          tt <- paste0(sample_col, ": ", id_vals)
          if (length(grp_col) == 1 && grp_col != "sample")
            tt <- paste0(tt, "<br>", grp_col, ": ", scores[[grp_col]])
          p$data[[".tooltip"]] <- tt
          p <- p + ggplot2::aes(text = .data[[".tooltip"]])
        }

        return(plotly::ggplotly(p, tooltip = "text"))
      }
    }
  }

  # ---- Scores-file path: read pca_scores.tsv written by the QC suite ----
  if (!is.null(png_files) && requireNamespace("plotly", quietly = TRUE)) {
    # Primary: explicit .tsv path in file vector
    scores_hits <- png_files[grepl("pca_scores\\.tsv$", png_files, perl = TRUE)]
    # Fallback: derive from pca_pc1_pc2.png (same directory) in case the .tsv
    # was not tracked in the file vector due to a superassignment bug or caching.
    pca_pngs <- png_files[grepl("pca_pc1_pc2\\.png$", png_files, perl = TRUE)]
    if (length(pca_pngs) > 0)
      scores_hits <- unique(c(scores_hits, file.path(dirname(pca_pngs), "pca_scores.tsv")))
    scores_hits <- scores_hits[file.exists(scores_hits)]

    if (length(scores_hits) > 0) {
      scores <- tryCatch(
        utils::read.table(scores_hits[1], sep = "\t", header = TRUE,
                          stringsAsFactors = FALSE, check.names = FALSE),
        error = function(e) {
          message("render_interactive_pca: could not read pca_scores.tsv: ", e$message)
          NULL
        }
      )

      if (!is.null(scores)) {
        pc_x  <- paste0("PC", pcs[1])
        pc_y  <- paste0("PC", pcs[2])
        vx    <- scores[[paste0("var_PC", pcs[1])]][1] %||% NA_real_
        vy    <- scores[[paste0("var_PC", pcs[2])]][1] %||% NA_real_

        # Identify color and shape columns from non-PC, non-var, non-sample columns
        reserved <- c("sample", grep("^PC|^var_PC", colnames(scores), value = TRUE))
        extra_cols <- setdiff(colnames(scores), reserved)
        color_col <- if (length(extra_cols) > 0) extra_cols[1] else "sample"
        shape_col <- if (length(extra_cols) > 1) extra_cols[2] else NULL

        if (!all(c(pc_x, pc_y) %in% colnames(scores))) {
          message("render_interactive_pca: expected columns ", pc_x, "/", pc_y,
                  " not found in pca_scores.tsv; falling back to static PNG.")
        } else {
          tt    <- paste0("sample: ", scores[["sample"]])
          if (color_col %in% colnames(scores) && color_col != "sample")
            tt  <- paste0(tt, "<br>", color_col, ": ", scores[[color_col]])
          if (!is.null(shape_col) && shape_col %in% colnames(scores))
            tt  <- paste0(tt, "<br>", shape_col, ": ", scores[[shape_col]])
          x_lab <- if (!is.na(vx)) sprintf("%s: %.1f%% variance", pc_x, 100 * vx) else pc_x
          y_lab <- if (!is.na(vy)) sprintf("%s: %.1f%% variance", pc_y, 100 * vy) else pc_y

          p <- plotly::plot_ly(
            data      = scores,
            x         = stats::as.formula(paste0("~`", pc_x, "`")),
            y         = stats::as.formula(paste0("~`", pc_y, "`")),
            color     = if (color_col %in% colnames(scores))
                          stats::as.formula(paste0("~factor(`", color_col, "`)")) else NULL,
            symbol    = if (!is.null(shape_col) && shape_col %in% colnames(scores))
                          stats::as.formula(paste0("~factor(`", shape_col, "`)")) else NULL,
            text      = tt,
            hoverinfo = "text",
            type      = "scatter",
            mode      = "markers",
            marker    = list(size = 9)
          )
          p <- plotly::layout(p,
            xaxis  = list(title = x_lab),
            yaxis  = list(title = y_lab),
            legend = list(title = list(text = color_col))
          )
          return(p)
        }
      }
    }
  }

  # ---- Final fallback: embed pre-rendered PNG ----------------------------
  if (!is.null(png_files)) {
    html <- embed_png(png_files, "pca_pc1_pc2")
    if (nzchar(html)) return(knitr::asis_output(html))
  }

  invisible(NULL)
}

render_dt <- function(df, caption = NULL, pageLength = 10) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  htmltools::tagList(
    DT::datatable(df, caption = caption,
                  rownames = FALSE,
                  extensions = "Buttons",
                  options = list(pageLength = pageLength,
                                 scrollX = TRUE,
                                 dom = "Bfrtip",
                                 buttons = list("csv", "excel")),
                  class = "compact stripe hover")
  )
}

# qc_files is referenced by _qc_normalization_section.Rmd via parent environment.
qc_files <- params$qc_suite_files %||% character(0)
```

> **Purpose:** Review normalization method performance before interpreting
> differential expression and enrichment results.  The **Benchmark** tab shows
> RSD and PC1 variance for every normalization stage, starting from the raw
> linear baseline (`met_imputed`).  Choose the method with the lowest
> `median_rsd_overall` and the highest `pca_pc1_delta`.

# Normalization QC {.tabset}

```{r qc-section, child='_qc_normalization_section.Rmd', eval=!is.null(params$qc_comparison_file) && length(params$qc_comparison_file) > 0}
```

```{r no-data-msg, results='asis', echo=FALSE, eval=is.null(params$qc_comparison_file) || length(params$qc_comparison_file) == 0}
cat("\n> **No QC data available.**  Run the pipeline with `qc.enabled: true`",
    "to generate the normalization QC outputs before rendering this report.\n")
```

---

# Session Info

<details>
<summary>Click to expand</summary>

```{r session-info}
sessionInfo()
```

</details>
