---
title: "`r local({ cfg <- params$config; pname <- cfg$project$name %||% ''; analyst <- gsub('_', ' ', cfg$project$analyst %||% ''); parts <- c('Metabolomics Analysis Report'); if (nzchar(pname)) parts <- c(parts, pname); if (nzchar(analyst)) parts <- c(parts, analyst); paste(parts, collapse = ' - ') })`"
subtitle: "multiomics-core Pipeline"
author: "`r paste0('Bioinformatician: ', gsub('_', ' ', params$config$project$analyst %||% ''))`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    fig_width: 10
    fig_height: 7
    self_contained: true
    code_folding: hide
    df_print: paged
params:
  pre: NULL
  qc_res: NULL
  de_res: NULL
  rf_res: NULL
  plsda_res: NULL
  enrichment_res: NULL
  network_res: NULL
  clustering_res: NULL
  qc_comparison_file: NULL
  qc_suite_files: NULL
  commentary_file: NULL
  config: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%"
)

library(ggplot2)
library(DT)
library(htmltools)

if (requireNamespace("plotly", quietly = TRUE))   library(plotly)
if (requireNamespace("crosstalk", quietly = TRUE)) library(crosstalk)
if (requireNamespace("jsonlite", quietly = TRUE))  library(jsonlite)

# Source interactive cutoff panel helpers
tryCatch({
  .cp_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
  .cp_found <- FALSE
  for (.cp_i in 1:6) {
    .cp_try <- file.path(.cp_dir, "R", "core", "17_cutoff_panel.R")
    if (file.exists(.cp_try)) { source(.cp_try); .cp_found <- TRUE; break }
    .cp_dir <- dirname(.cp_dir)
  }
  if (!.cp_found) message("cutoff_panel helpers not found (walked 6 levels up)")
}, error = function(e) message("cutoff_panel helpers not found: ", e$message))

# The Rmd renders in a child R session that doesn't inherit functions sourced
# by the parent _targets.R run. Inline the small helpers the report uses
# (same definitions the lipidomics report Rmd uses) so the child has them.
make_contrast_label <- function(contrast_str) {
  parts <- strsplit(contrast_str, " - ", fixed = TRUE)[[1]]
  paste(trimws(parts), collapse = "_vs_")
}

extract_contrast_table <- function(summary_df, contrast) {
  linear_fc <- summary_df[[paste0("linearFC.", contrast)]]
  logfc <- ifelse(linear_fc >= 0, log2(linear_fc), -log2(abs(linear_fc)))
  # Pipeline emits pvalue.X / padj.X (DESeq2-style); the older versions emitted
  # P.Value.X / adj.P.Val.X (limma-style). Read whichever exists.
  pcol_raw <- if (paste0("pvalue.", contrast) %in% names(summary_df))
    paste0("pvalue.", contrast) else paste0("P.Value.", contrast)
  pcol_adj <- if (paste0("padj.", contrast) %in% names(summary_df))
    paste0("padj.", contrast) else paste0("adj.P.Val.", contrast)
  data.frame(
    feature_id = summary_df$feature_id,
    logFC      = logfc,
    AveExpr    = summary_df[[paste0("AveExpr.", contrast)]],
    P.Value    = summary_df[[pcol_raw]],
    adj.P.Val  = summary_df[[pcol_adj]],
    stringsAsFactors = FALSE
  )
}

# ---------------------------------------------------------------------------
# Helper: downloadable DT table with CSV/Excel export (proteomics style)
# ---------------------------------------------------------------------------
make_dt <- function(df, caption = NULL, pageLength = 15) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  safe_caption <- gsub("[^a-zA-Z0-9]", "_", caption %||% "data")
  tbl <- DT::datatable(
    df,
    caption  = caption,
    extensions = "Buttons",
    options = list(
      scrollX    = TRUE,
      pageLength = pageLength,
      dom        = "Bfrtip",
      buttons    = list(
        list(extend = "csv",   filename = safe_caption),
        list(extend = "excel", filename = safe_caption)
      )
    ),
    rownames = FALSE,
    class    = "compact stripe hover"
  )
  # Format numeric columns: scientific notation for p-values, 2 decimals for the rest
  num_cols <- names(df)[vapply(df, is.numeric, logical(1))]
  pval_pattern <- "^(p[._]?val|p[._]?adj|padj|adj[._]?p|fdr|q[._]?val|BH|bonferroni|p\\.value|adj\\.P\\.Val|PValue|FDR)"
  pval_cols <- grep(pval_pattern, num_cols, ignore.case = TRUE, value = TRUE)
  other_cols <- setdiff(num_cols, pval_cols)
  if (length(other_cols) > 0) {
    tbl <- DT::formatRound(tbl, columns = other_cols, digits = 2)
  }
  if (length(pval_cols) > 0) {
    pval_idx <- which(names(df) %in% pval_cols) - 1L
    sci_render <- DT::JS(
      "function(data, type, row) {",
      "  if (type === 'display' && data !== null && data !== '') {",
      "    return parseFloat(data).toExponential(2);",
      "  }",
      "  return data;",
      "}"
    )
    col_defs <- lapply(pval_idx, function(i) list(targets = i, render = sci_render))
    tbl$x$options$columnDefs <- c(tbl$x$options$columnDefs, col_defs)
  }
  htmltools::tagList(tbl)
}

# ---------------------------------------------------------------------------
# Helper: paper-style figure legend below plots (matching proteomics)
# ---------------------------------------------------------------------------
figure_legend <- function(text) {
  cat(sprintf("<p class='figure-legend' style='text-align: center; font-size: 0.9em; color: #555; margin: 4px auto 12px auto; max-width: 88%%; line-height: 1.45; font-style: italic;'>%s</p>\n", text))
}

# ---------------------------------------------------------------------------
# Helper: safely print a ggplot or return invisible(NULL)
# ---------------------------------------------------------------------------
safe_print_plot <- function(p) {
  if (inherits(p, "gg") || inherits(p, "ggplot")) {
    print(p)
  } else {
    invisible(NULL)
  }
}

# ---------------------------------------------------------------------------
# Helper: embed a PNG from a file path vector by basename pattern
# ---------------------------------------------------------------------------
embed_png <- function(files, pattern, fig_width = "100%") {
  if (is.null(files) || length(files) == 0) return(invisible(NULL))
  hits <- files[grepl(pattern, basename(files), ignore.case = TRUE)]
  hits <- hits[file.exists(hits)]
  if (length(hits) == 0) return(invisible(NULL))
  knitr::include_graphics(hits[1])
}

# ---------------------------------------------------------------------------
# Unpack config and module results
# ---------------------------------------------------------------------------
cfg       <- params$config
metab_cfg <- cfg$modes$metabolomics

# ---------------------------------------------------------------------------
# Report section visibility (all TRUE by default; override via config)
#
# Each section can be:
#   true/false          — toggle entire section
#   map with sub-keys   — toggle individual plots (section shown if any sub-key is true)
#
# Example config:
#   report_sections:
#     qc:
#       pca: true
#       density: false        # hide density plots
#       boxplots: true
#       heatmaps: true
#     de:
#       volcano: true
#       tables: false         # hide DE result tables
#       explorer: true
# ---------------------------------------------------------------------------
rs <- metab_cfg$report_sections %||% list()

# Helper: resolve a section entry that can be TRUE/FALSE or a list of sub-keys
resolve_section <- function(entry, sub_defaults) {
    if (is.null(entry) || isTRUE(entry)) return(sub_defaults)
    if (identical(entry, FALSE)) return(lapply(sub_defaults, function(x) FALSE))
    # entry is a list of sub-keys
    out <- sub_defaults
    for (nm in names(entry)) {
        if (nm %in% names(out)) out[[nm]] <- isTRUE(entry[[nm]])
    }
    out
}

# Top-level sections (no sub-keys)
show_executive_summary <- isTRUE(rs$executive_summary %||% TRUE)
show_parameters        <- isTRUE(rs$parameters        %||% TRUE)
show_module_status     <- isTRUE(rs$module_status      %||% TRUE)
show_de_heatmaps       <- isTRUE(rs$de_heatmaps        %||% TRUE)
show_network           <- isTRUE(rs$network             %||% TRUE)
show_mummichog         <- isTRUE(rs$mummichog           %||% TRUE)
show_output_files      <- isTRUE(rs$output_files        %||% TRUE)
show_methods           <- isTRUE(rs$methods             %||% TRUE)
show_session_info      <- isTRUE(rs$session_info        %||% TRUE)

# QC sub-keys
qc_subs <- resolve_section(rs$qc, list(pca = TRUE, density = TRUE, boxplots = TRUE, heatmaps = TRUE))
show_qc          <- any(unlist(qc_subs))
show_qc_pca      <- qc_subs$pca
show_qc_density  <- qc_subs$density
show_qc_boxplots <- qc_subs$boxplots
show_qc_heatmaps <- qc_subs$heatmaps

# DE sub-keys
# `ma` and `pval_hist` default OFF (Apr 2026 reviewer feedback — keep volcano only).
de_subs <- resolve_section(
  rs$de,
  list(volcano = TRUE, ma = FALSE, pval_hist = FALSE, tables = TRUE, explorer = TRUE)
)
show_de          <- any(unlist(de_subs))
show_de_volcano  <- de_subs$volcano
show_de_ma       <- de_subs$ma
show_de_pvalhist <- de_subs$pval_hist
show_de_tables   <- de_subs$tables
show_de_explorer <- de_subs$explorer

# Feature Selection sub-keys
fs_subs <- resolve_section(rs$feature_selection, list(rf = TRUE, plsda = TRUE))
show_feature_selection <- any(unlist(fs_subs))
show_fs_rf             <- fs_subs$rf
show_fs_plsda          <- fs_subs$plsda

# Enrichment sub-keys
enr_subs <- resolve_section(rs$enrichment, list(qea = TRUE, ssgsea = TRUE, ora = TRUE, gsea = TRUE))
show_enrichment     <- any(unlist(enr_subs))
show_enr_qea        <- enr_subs$qea
show_enr_ssgsea     <- enr_subs$ssgsea
show_enr_ora        <- enr_subs$ora
show_enr_gsea       <- enr_subs$gsea

# ---------------------------------------------------------------------------
# Load AI commentary (if available)
# ---------------------------------------------------------------------------
commentary <- NULL
has_commentary <- FALSE
if (!is.null(params$commentary_file) && file.exists(params$commentary_file)) {
    commentary <- tryCatch(
        jsonlite::read_json(params$commentary_file),
        error = function(e) NULL
    )
    has_commentary <- !is.null(commentary) && length(commentary) > 0
}

# Helper: render commentary block for a figure_id (collapsible)
render_commentary <- function(figure_id) {
    if (!has_commentary) return(invisible(NULL))
    c <- commentary[[figure_id]]
    if (is.null(c)) return(invisible(NULL))

    cat("<details class='commentary-details' style='margin: 4px 0 18px 0;'>\n")
    cat("<summary style='cursor: pointer; color: #4a90d9; font-weight: 500; font-size: 0.93em; padding: 3px 0;'>Show AI Interpretation</summary>\n")

    html <- "<div class='commentary-block' style='background: #f8f9fa; border-left: 4px solid #4a90d9; padding: 12px 16px; margin: 6px 0 4px 0; border-radius: 0 4px 4px 0; font-size: 0.92em;'>\n"
    html <- paste0(html, "<p style='margin: 0 0 8px 0;'><strong>What this shows:</strong> ",
                   htmltools::htmlEscape(c$what_is_this %||% ""), "</p>\n")

    bio_interp <- c$biological_interpretation %||% ""
    if (nzchar(bio_interp)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Interpretation:</strong> ",
                       htmltools::htmlEscape(bio_interp), "</p>\n")
    }

    obs <- c$key_observations %||% c$observations
    if (!is.null(obs) && length(obs) > 0) {
        html <- paste0(html, "<p style='margin: 4px 0 4px 0;'><strong>Key observations:</strong></p>\n<ul style='margin: 2px 0 8px 0; padding-left: 20px;'>\n")
        for (ob in obs) {
            html <- paste0(html, "<li>", htmltools::htmlEscape(ob), "</li>\n")
        }
        html <- paste0(html, "</ul>\n")
    }

    stat_note <- c$statistical_note %||% c$statistical_assessment %||% ""
    if (nzchar(stat_note)) {
        html <- paste0(html, "<p style='margin: 4px 0 8px 0;'><strong>Statistical note:</strong> ",
                       htmltools::htmlEscape(stat_note), "</p>\n")
    }

    # Backend attribution with K-Dense credit
    backend <- c$backend %||% "unknown"
    confidence <- c$confidence %||% ""
    backend_label <- if (grepl("claude", backend, ignore.case = TRUE)) {
        "Claude (Anthropic) + K-Dense AI scientific-skills"
    } else if (grepl("openai", backend, ignore.case = TRUE)) {
        "OpenAI + K-Dense AI scientific-skills"
    } else if (backend == "data_driven") {
        "data-driven + K-Dense AI scientific-skills"
    } else {
        paste0(backend, " + K-Dense AI scientific-skills")
    }
    if (nzchar(confidence) && confidence != "none") {
        backend_label <- paste0(backend_label, " | confidence: ", confidence)
    }
    html <- paste0(html, "<p style='margin: 4px 0 0 0; color: #888; font-size: 0.85em;'><em>Generated by: ",
                   htmltools::htmlEscape(backend_label), "</em></p>\n")

    html <- paste0(html, "</div>\n")
    cat(html)
    cat("</details>\n")
}
```

```{r force-dependencies, include=FALSE}
# Force DT and Plotly JS dependencies to be included in the self-contained HTML
# Note: filter="top" forces noUiSlider dependency (needed for numeric range filters)
DT::datatable(data.frame(x = 1), filter = "top")
if (requireNamespace("plotly", quietly = TRUE)) {
  plotly::plot_ly(data.frame(x = 1, y = 1), x = ~x, y = ~y,
                  type = "scatter", mode = "markers")
}
```

```{r unpack-results, include=FALSE}
pre            <- params$pre
qc_res         <- params$qc_res
de_res         <- params$de_res
rf_res         <- params$rf_res
plsda_res      <- params$plsda_res
enrichment_res <- params$enrichment_res

n_features <- if (!is.null(pre)) nrow(pre$expr_work) else NA
n_samples  <- if (!is.null(pre)) ncol(pre$expr_work) else NA

# Resolve group / condition column
cond_col <- metab_cfg$de$condition_column %||%
            metab_cfg$effects$color %||% "sample_type"
n_groups <- NA
if (!is.null(pre$meta) && cond_col %in% colnames(pre$meta)) {
  n_groups <- length(unique(pre$meta[[cond_col]]))
}

# Normalization summary
norm_method <- "N/A"
if (!is.null(pre$info$normalization)) {
  ni    <- pre$info$normalization
  parts <- character(0)
  if (!is.null(ni$sample_norm) && tolower(ni$sample_norm) != "none")
    parts <- c(parts, toupper(ni$sample_norm))
  if (!is.null(ni$transform) && tolower(ni$transform) != "none")
    parts <- c(parts, ni$transform)
  if (!is.null(ni$scaling) && tolower(ni$scaling) != "none")
    parts <- c(parts, ni$scaling)
  if (length(parts) > 0) norm_method <- paste(parts, collapse = " + ")
}

de_method <- if (!is.null(de_res$method)) de_res$method else "N/A"

# DE contrasts
de_contrasts <- NULL
if (!is.null(de_res) && !is.null(de_res$summary_df)) {
  de_cfg       <- metab_cfg$de %||% list()
  de_contrasts <- de_cfg$contrasts
  if (is.list(de_contrasts)) de_contrasts <- unlist(de_contrasts)
}
```

`r if (!show_executive_summary) "<!--"`

# Executive Summary

```{r executive-summary, results='asis'}
# ---- Gradient panel (matching proteomics style) --------------------------
cat("<div style='background: linear-gradient(135deg, #f8f9fa 0%, #e9ecef 100%); border: 1px solid #dee2e6; border-radius: 8px; padding: 20px 24px; margin: 10px 0 20px 0;'>\n")
cat("<div style='display: flex; flex-wrap: wrap; gap: 16px; margin-bottom: 16px;'>\n")

# Metric card helper
metric_card <- function(value, label, color = "#2c3e50") {
  cat(sprintf(
    "<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: %s;'>%s</div><div style='font-size: 0.85em; color: #666;'>%s</div></div>\n",
    color, as.character(value), label
  ))
}

# Samples
if (!is.na(n_samples)) metric_card(n_samples, "Samples")

# Groups
if (!is.na(n_groups)) metric_card(n_groups, "Groups")

# Metabolites
if (!is.na(n_features)) metric_card(format(n_features, big.mark = ","), "Metabolites")

# Contrasts
n_contrasts <- length(de_contrasts)
if (n_contrasts > 0) metric_card(n_contrasts, "Contrasts")

# Total DE metabolites
n_de_total <- NA
if (!is.null(de_res$summary_df) && "pass_any_contrast" %in% colnames(de_res$summary_df)) {
  n_de_total <- sum(de_res$summary_df$pass_any_contrast == 1, na.rm = TRUE)
  metric_card(format(n_de_total, big.mark = ","), "DE Metabolites", color = "#c0392b")
}

cat("</div>\n")

# ---- DE breakdown per contrast -------------------------------------------
if (!is.null(de_contrasts) && length(de_contrasts) > 0 && !is.null(de_res$summary_df)) {
  padj_cut <- metab_cfg$de$p_cutoff %||% 0.05
  if (!is.null(metab_cfg$de$logfc_cutoff)) {
    log2_lfc <- metab_cfg$de$logfc_cutoff
  } else {
    lfc_cut  <- metab_cfg$de$linear_fc_cutoff %||% 1.5
    log2_lfc <- log2(lfc_cut)
  }

  cat("<div style='margin: 12px 0;'>\n")
  cat("<strong>Differential Expression per Contrast:</strong>\n<ul style='margin: 6px 0; padding-left: 20px;'>\n")

  for (ctr in de_contrasts) {
    ctr_label <- make_contrast_label(ctr)
    ctr_tbl   <- tryCatch(extract_contrast_table(de_res$summary_df, ctr_label),
                           error = function(e) NULL)
    if (!is.null(ctr_tbl)) {
      sig   <- !is.na(ctr_tbl$P.Value) & ctr_tbl$P.Value <= padj_cut &
               !is.na(ctr_tbl$logFC) & abs(ctr_tbl$logFC) >= log2_lfc
      n_up   <- sum(sig & ctr_tbl$logFC > 0, na.rm = TRUE)
      n_down <- sum(sig & ctr_tbl$logFC < 0, na.rm = TRUE)
      if (exists("cutoff_summary_span")) {
        cat(sprintf("<li><strong>%s</strong>: %s</li>\n",
                    ctr, cutoff_summary_span(ctr, n_up, n_down, "metabolites")))
      } else {
        cat(sprintf("<li><strong>%s</strong>: %d DE metabolites (%d up, %d down)</li>\n",
                    ctr, n_up + n_down, n_up, n_down))
      }
    }
  }
  cat("</ul>\n</div>\n")
}

cat("</div>\n")
```

```{r technical-metadata-table-metab, results='asis'}
# Technical metadata table — what ran, and feature counts before/after filtering.
.tm_n_before <- {
    if (!is.null(pre$expr_raw)) nrow(pre$expr_raw)
    else if (!is.null(pre$info$n_features_before)) pre$info$n_features_before
    else NA_integer_
}
.tm_n_after <- {
    if (!is.null(pre$expr_work)) nrow(pre$expr_work)
    else if (!is.null(pre$info$n_features_after)) pre$info$n_features_after
    else NA_integer_
}

.tm_steps <- list(
    list(label = "Filtering",
         ran   = !is.na(.tm_n_before) && !is.na(.tm_n_after) && .tm_n_before != .tm_n_after),
    list(label = "Normalization",
         ran   = (tolower(pre$info$normalization$method %||% metab_cfg$preprocessing$chosen_norm %||% "none") != "none")),
    list(label = "Differential expression",
         ran   = !is.null(de_res) && !is.null(de_res$summary_df)),
    list(label = "Feature selection (RF / PLS-DA)",
         ran   = !is.null(rf_res) || !is.null(plsda_res)),
    list(label = "Enrichment (QEA / ORA / GSEA / Mummichog)",
         ran   = !is.null(enrichment_res) && (length(enrichment_res) > 0)),
    list(label = "Clustering",
         ran   = isTRUE(metab_cfg$clustering$enabled))
)
.tm_steps_str <- paste(vapply(.tm_steps, function(s) {
    sprintf("%s: %s", s$label, if (isTRUE(s$ran)) "yes" else "skipped")
}, character(1)), collapse = "<br>")

.tm_fmt_n <- function(n) if (is.na(n)) "n/a" else format(n, big.mark = ",")

cat('<table style="margin: 16px 0 24px 0; border-collapse: collapse; width: 100%; border: 1px solid #d0d7de; border-radius: 6px; overflow: hidden; font-size: 14px;">\n')
cat('<thead><tr style="background: #f6f8fa; color: #24292f;">\n')
cat('<th style="text-align:left; padding: 10px 14px; border-bottom: 1px solid #d0d7de;">Technical metadata</th>\n')
cat('<th style="text-align:left; padding: 10px 14px; border-bottom: 1px solid #d0d7de;">Value</th>\n')
cat('</tr></thead><tbody>\n')

.tm_row <- function(k, v) cat(sprintf(
    '<tr><td style="padding: 8px 14px; border-bottom: 1px solid #eaecef; font-weight: 600; color: #24292f;">%s</td><td style="padding: 8px 14px; border-bottom: 1px solid #eaecef; color: #57606a;">%s</td></tr>\n',
    k, v))

.tm_row("Analysis steps performed", .tm_steps_str)
.tm_row("Metabolites before filtering", .tm_fmt_n(.tm_n_before))
.tm_row("Metabolites after filtering",  .tm_fmt_n(.tm_n_after))
if (!is.na(.tm_n_before) && !is.na(.tm_n_after) && .tm_n_before > 0) {
    .tm_row("Retained", sprintf("%.1f%%", 100 * .tm_n_after / .tm_n_before))
}
.tm_row("DE method",
        sprintf("%s (p &le; %s, |linear FC| &ge; %s)",
                toupper(metab_cfg$de$method %||% "limma"),
                signif(metab_cfg$de$p_cutoff %||% 0.05, 3),
                metab_cfg$de$linear_fc_cutoff %||% 1.5))
cat('</tbody></table>\n')
```

`r if (!show_executive_summary) "-->"`

---

`r if (!show_parameters) "<!--"`

# Analysis Parameters {.tabset}

## Project

```{r params-project}
proj <- data.frame(
  Parameter = c("Project Name", "Analyst", "Project Directory",
                "Raw Data Path", "Output Path"),
  Value = c(
    cfg$project$name %||% "N/A",
    gsub("_", " ", cfg$project$analyst %||% "N/A"),
    cfg$project$dir %||% "N/A",
    cfg$paths$raw %||% "N/A",
    cfg$paths$out %||% "N/A"
  ),
  stringsAsFactors = FALSE
)
make_dt(proj, "Project Parameters")
```

## Experimental Design

```{r params-design}
if (!is.null(pre$meta) && cond_col %in% colnames(pre$meta)) {
  design_tbl <- as.data.frame(table(pre$meta[[cond_col]]))
  colnames(design_tbl) <- c("Group", "n (replicates)")
  make_dt(design_tbl, "Experimental Design")
}
```

## Pipeline Settings

```{r params-pipeline}
settings <- data.frame(
  Parameter = c(
    "Normalization",
    "DE Method",
    "Adjusted p-value Cutoff",
    "Linear FC Cutoff",
    "Random Forest",
    "PLS-DA",
    "Enrichment"
  ),
  Value = c(
    norm_method,
    de_method,
    as.character(metab_cfg$de$p_cutoff %||% 0.05),
    as.character(metab_cfg$de$linear_fc_cutoff %||% 1.5),
    if (!is.null(rf_res)) "Enabled" else "Disabled",
    if (!is.null(plsda_res)) "Enabled" else "Disabled",
    if (!is.null(enrichment_res)) "Enabled" else "Disabled"
  ),
  stringsAsFactors = FALSE
)
make_dt(settings, "Pipeline Settings")
```

## Contrasts

```{r params-contrasts}
if (!is.null(de_contrasts) && length(de_contrasts) > 0) {
  ctr_df <- data.frame(
    `#` = seq_along(de_contrasts),
    Contrast = de_contrasts,
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
  make_dt(ctr_df, "Contrasts")
} else {
  cat("No contrasts defined.\n")
}
```

`r if (!show_parameters) "-->"`

---

`r if (!show_module_status) "<!--"`

# Module Status

```{r module-status}
module_status <- data.frame(
  Module = c("Preprocessing", "QC Diagnostics", "Differential Expression",
             "Random Forest", "PLS-DA", "Enrichment"),
  Status = c(
    if (!is.null(pre)) "Completed" else "Skipped",
    if (!is.null(qc_res)) "Completed" else "Skipped",
    if (!is.null(de_res) && !is.null(de_res$summary_df)) "Completed" else "Skipped",
    if (!is.null(rf_res)) "Completed" else "Skipped",
    if (!is.null(plsda_res)) "Completed" else "Skipped",
    if (!is.null(enrichment_res)) "Completed" else "Skipped"
  ),
  stringsAsFactors = FALSE
)
make_dt(module_status, "Module Status", pageLength = 10)
```

`r if (!show_module_status) "-->"`

---

`r if (is.null(qc_res) || !show_qc) "<!--"`

# QC Diagnostics {.tabset}

`r if (!show_qc_pca) "<!--"`

## PCA {.tabset}

```{r qc-pca-helpers, include=FALSE}
# Helper: pick the primary PCA key for a given subset (with-QC vs without-QC)
.pick_pca_key <- function(plots, axes_prefix, subset = c("with_qc", "no_qc")) {
  subset <- match.arg(subset)
  keys <- grep(paste0("^", axes_prefix), names(plots), value = TRUE)
  if (subset == "no_qc") {
    keys <- keys[grepl("_noQC$", keys)]
  } else {
    keys <- keys[!grepl("_noQC$", keys)]
  }
  if (length(keys) == 0) NA_character_ else keys[1]
}

# Decide which subsets to expose as tabs: only show "without QC" if the
# pipeline produced _noQC plots (i.e. QC samples actually exist).
.pca_has_noqc <- any(grepl("_noQC$", names(qc_res$plots %||% list())))
.pca_subsets <- if (.pca_has_noqc) {
  list(list(tab = "With QC samples",    sub = "with_qc"),
       list(tab = "Without QC samples", sub = "no_qc"))
} else {
  list(list(tab = "All samples", sub = "with_qc"))
}
```

### Static {.tabset}

```{r qc-pca-static, fig.width=9, fig.height=7, results='asis'}
if (!is.null(qc_res$plots)) {
  for (sub in .pca_subsets) {
    cat(sprintf("\n#### %s\n\n", sub$tab))
    pca_main <- .pick_pca_key(qc_res$plots, "pca_1_2", sub$sub)
    if (!is.na(pca_main)) {
      safe_print_plot(qc_res$plots[[pca_main]])
      figure_legend(sprintf(
        "PCA of normalized metabolomics data (PC1 vs PC2) — %s. Each point represents one sample, colored by experimental group.",
        tolower(sub$tab)))
      render_commentary(paste0("pca_pc1_vs_pc2", if (sub$sub == "no_qc") "_noQC" else ""))
    }
    pca13_main <- .pick_pca_key(qc_res$plots, "pca_1_3", sub$sub)
    if (!is.na(pca13_main)) {
      safe_print_plot(qc_res$plots[[pca13_main]])
      figure_legend(sprintf(
        "PCA projection onto PC1 vs PC3 — %s. Captures additional variance beyond the first two components.",
        tolower(sub$tab)))
      render_commentary(paste0("pca_pc1_vs_pc3", if (sub$sub == "no_qc") "_noQC" else ""))
    }
  }
}
```

### With Sample Names {.tabset}

```{r qc-pca-labeled, fig.width=9, fig.height=7, results='asis'}
if (!is.null(qc_res$plots)) {
  for (sub in .pca_subsets) {
    cat(sprintf("\n#### %s\n\n", sub$tab))
    pca_main <- .pick_pca_key(qc_res$plots, "pca_1_2", sub$sub)
    if (!is.na(pca_main)) {
      p <- qc_res$plots[[pca_main]]
      if ("sample" %in% colnames(p$data)) {
        p_lab <- p + ggplot2::geom_text(ggplot2::aes(label = sample), size = 2.5, vjust = -0.8, show.legend = FALSE)
        safe_print_plot(p_lab)
      }
    }
    pca13_main <- .pick_pca_key(qc_res$plots, "pca_1_3", sub$sub)
    if (!is.na(pca13_main)) {
      p13 <- qc_res$plots[[pca13_main]]
      if ("sample" %in% colnames(p13$data)) {
        p13_lab <- p13 + ggplot2::geom_text(ggplot2::aes(label = sample), size = 2.5, vjust = -0.8, show.legend = FALSE)
        safe_print_plot(p13_lab)
      }
    }
  }
}
```

### Interactive

```{r qc-pca-interactive, results='asis'}
if (!is.null(qc_res$plots) && requireNamespace("plotly", quietly = TRUE)) {
  pca_keys <- grep("^pca_1_2", names(qc_res$plots), value = TRUE)
  pca_main <- pca_keys[!grepl("_noQC$", pca_keys)][1]
  if (!is.na(pca_main)) {
    p <- qc_res$plots[[pca_main]]
    sc <- p$data
    if ("sample" %in% colnames(sc)) {
      non_pc <- setdiff(colnames(sc), c("sample", "sample_id", grep("^PC", colnames(sc), value = TRUE)))
      color_var <- if (length(non_pc) > 0) non_pc[1] else NA
      sc$hover_text <- paste0("Sample: ", sc$sample,
                              if (!is.na(color_var)) paste0("<br>", color_var, ": ", sc[[color_var]]) else "")
      pl <- plotly::plot_ly(sc, x = ~PC1, y = ~PC2,
                            color = if (!is.na(color_var)) as.formula(paste0("~`", color_var, "`")) else NULL,
                            text = ~hover_text, hoverinfo = "text",
                            type = "scatter", mode = "markers",
                            marker = list(size = 10)) |>
          plotly::layout(title = "PCA: PC1 vs PC2 — Hover for sample names")
      print(htmltools::tagList(pl))
    }
  }
} else {
  cat("*Interactive PCA not available (plotly package missing).*\n")
}
```

`r if (!show_qc_pca) "-->"`

`r if (!show_qc_density) "<!--"`

## Density

```{r qc-density, fig.width=9, fig.height=6, results='asis'}
if (!is.null(qc_res$plots)) {
  if (!is.null(qc_res$plots[["density_raw"]])) {
    safe_print_plot(qc_res$plots[["density_raw"]])
    figure_legend("Per-sample density of log2(raw intensities) before any normalization. Well-overlapping curves indicate consistent sample loading and acquisition.")
    render_commentary("density_raw")
  }
  ni_dens <- pre$info$normalization
  has_scaling_dens <- !is.null(ni_dens$scaling) && tolower(ni_dens$scaling) != "none"
  if (has_scaling_dens && !is.null(qc_res$plots[["density_transformed"]])) {
    safe_print_plot(qc_res$plots[["density_transformed"]])
    figure_legend("Per-sample density after the transform step. This is the matrix used for differential expression testing. Well-overlapping curves confirm the transform brought samples onto a comparable scale.")
  }
  if (!is.null(qc_res$plots[["density_norm"]])) {
    safe_print_plot(qc_res$plots[["density_norm"]])
    if (has_scaling_dens) {
      figure_legend(sprintf(
        "Per-sample density after full normalization (%s + %s). %s scaling standardizes each metabolite independently (mean=0, SD=1), so divergent per-sample curves reflect biological differences between groups. This matrix is used for PCA, heatmaps, and multivariate analyses (RF, PLS-DA).",
        ni_dens$transform %||% "", ni_dens$scaling %||% "",
        tools::toTitleCase(ni_dens$scaling %||% "")
      ))
    } else {
      figure_legend("Per-sample density after normalization. Well-aligned curves confirm effective normalization across samples.")
      render_commentary("density_normalized")
    }
  }
}
```

`r if (!show_qc_density) "-->"`

`r if (!show_qc_boxplots) "<!--"`

## Boxplots

```{r qc-boxplots, fig.width=10, fig.height=6, results='asis'}
if (!is.null(qc_res$plots)) {
  if (!is.null(qc_res$plots[["boxplot_raw"]])) {
    safe_print_plot(qc_res$plots[["boxplot_raw"]])
    figure_legend("Per-sample boxplots of log2-transformed raw metabolite intensities before normalization. Log2 transformation is applied for visualization to make intensity distributions comparable across the dynamic range.")
    render_commentary("boxplot_raw")
  }
  # Show transform-only boxplot when scaling is applied; otherwise show
  # the fully normalized boxplot (they are identical when no scaling).
  ni_box <- pre$info$normalization
  has_scaling_box <- !is.null(ni_box$scaling) && tolower(ni_box$scaling) != "none"
  if (has_scaling_box && !is.null(qc_res$plots[["boxplot_transformed"]])) {
    safe_print_plot(qc_res$plots[["boxplot_transformed"]])
    figure_legend("Per-sample boxplots after the transform step. This is the matrix used for differential expression testing. Aligned medians confirm the transform effectively standardized sample distributions.")
  }
  if (!is.null(qc_res$plots[["boxplot"]])) {
    safe_print_plot(qc_res$plots[["boxplot"]])
    if (has_scaling_box) {
      figure_legend(sprintf(
        "Per-sample boxplots after full normalization (%s + %s). %s scaling standardizes each metabolite independently (mean=0, SD=1), so shifted medians between groups reflect biological differences. This matrix is used for PCA, heatmaps, and multivariate analyses (RF, PLS-DA).",
        ni_box$transform %||% "", ni_box$scaling %||% "",
        tools::toTitleCase(ni_box$scaling %||% "")
      ))
    } else {
      figure_legend("Per-sample boxplots after normalization. Comparable medians and spreads indicate successful normalization.")
      render_commentary("boxplot_normalized")
    }
  }
}
```

`r if (!show_qc_boxplots) "-->"`

`r if (!show_qc_heatmaps) "<!--"`

## Heatmaps {.tabset}

```{r qc-heatmaps, out.width="100%", results='asis'}
# Distance heatmap only (correlation heatmap intentionally omitted from the
# report after Apr 2026 reviewer feedback — pipeline still writes it to disk).
# Tabs split by sample subset when QC samples exist.
if (!is.null(qc_res$files)) {
  fnames <- basename(qc_res$files)
  dist_with <- qc_res$files[file.exists(qc_res$files) &
                            fnames == "sample_distance_heatmap.png"]
  dist_no   <- qc_res$files[file.exists(qc_res$files) &
                            fnames == "sample_distance_heatmap_noQC.png"]

  hm_subsets <- list()
  if (length(dist_with) > 0) {
    hm_subsets[[length(hm_subsets) + 1]] <- list(
      tab  = if (length(dist_no) > 0) "With QC samples" else "All samples",
      path = dist_with[1],
      cid  = "sample_distance_heatmap"
    )
  }
  if (length(dist_no) > 0) {
    hm_subsets[[length(hm_subsets) + 1]] <- list(
      tab  = "Without QC samples",
      path = dist_no[1],
      cid  = "sample_distance_heatmap_noQC"
    )
  }

  if (length(hm_subsets) == 0) {
    cat("*No sample distance heatmap available.*\n\n")
  } else {
    for (sub in hm_subsets) {
      cat(sprintf("\n### %s\n\n", sub$tab))
      cat(sprintf("![Sample Distance Heatmap -- %s](%s){width=100%%}\n\n",
                  sub$tab, sub$path))
      figure_legend(sprintf(
        "Sample distance heatmap based on Euclidean distances between normalized metabolite profiles (%s). Darker colors indicate larger distances (less similar samples).",
        tolower(sub$tab)))
      render_commentary(sub$cid)
    }
  }
}
```

`r if (!show_qc_heatmaps) "-->"`

`r if (is.null(qc_res) || !show_qc) "-->"`

---

`r if (is.null(de_res) || is.null(de_res$summary_df) || !show_de) "<!--"`

# Differential Expression

```{r cutoff-panel, results='asis'}
if (exists("cutoff_panel_html")) {
    .lfc_default <- if (!is.null(metab_cfg$de$logfc_cutoff)) {
        metab_cfg$de$logfc_cutoff
    } else {
        log2(metab_cfg$de$linear_fc_cutoff %||% 1.5)
    }
    .p_default <- metab_cfg$de$p_cutoff %||% 0.05
    cat(cutoff_panel_html(.lfc_default, .p_default, default_ptype = "pval"))
}
```

```{r de-prep, include=FALSE}
# Prepare shared data used across DE sub-sections
padj_cut_de <- metab_cfg$de$p_cutoff %||% 0.05
# logfc_cutoff is applied directly to |logFC| (in the transform's scale).
# linear_fc_cutoff is a backwards-compat alias converted via log2().
if (!is.null(metab_cfg$de$logfc_cutoff)) {
  log2_lfc_de <- metab_cfg$de$logfc_cutoff
} else {
  lfc_cut_de  <- metab_cfg$de$linear_fc_cutoff %||% 1.5
  log2_lfc_de <- log2(lfc_cut_de)
}

# Build feature_id -> Name lookup
name_map_v <- NULL
name_col_report <- intersect(c("Name", "original_id"), colnames(pre$row_data))[1]
if (!is.null(pre$row_data) && !is.na(name_col_report)) {
  name_vals <- as.character(pre$row_data[[name_col_report]])
  fid_vals <- as.character(pre$row_data$feature_id)
  # Only use names that differ from feature_id (actually resolved)
  resolved <- name_vals != fid_vals & !is.na(name_vals) & name_vals != ""
  if (any(resolved)) {
    name_map_v <- stats::setNames(name_vals, fid_vals)
  }
}

resolve_name <- function(fid) {
  if (!is.null(name_map_v) && fid %in% names(name_map_v) && !is.na(name_map_v[fid])) {
    name_map_v[fid]
  } else {
    fid
  }
}

# Truncate long labels for plot axes / heatmap rows so a single huge IUPAC
# name does not stretch the plot area. Table/hover values keep the full name.
truncate_label <- function(x, max_len = 40) {
  x <- as.character(x)
  long <- !is.na(x) & nchar(x) > max_len
  x[long] <- paste0(substr(x[long], 1, max_len - 1), "\u2026")
  x
}

# Build norm intensity y-label
norm_y_label_report <- "Normalized intensity"
ni_r <- pre$info$normalization
if (!is.null(ni_r)) {
  np <- character(0)
  if (!is.null(ni_r$transform) && tolower(ni_r$transform) != "none") np <- c(np, ni_r$transform)
  if (!is.null(ni_r$sample_norm) && tolower(ni_r$sample_norm) != "none") np <- c(np, toupper(ni_r$sample_norm))
  if (!is.null(ni_r$scaling) && tolower(ni_r$scaling) != "none") np <- c(np, ni_r$scaling)
  if (length(np) > 0) norm_y_label_report <- paste0("Normalized intensity (", paste(np, collapse = " + "), ")")
}
```

```{r de-note, results='asis'}
if (!is.null(de_res) && !is.null(de_res$summary_df)) {
  n_sig_any <- sum(de_res$summary_df$pass_any_contrast == 1, na.rm = TRUE)
  if (n_sig_any == 0) {
    n_samp <- ncol(pre$expr_work)
    cat("\n> **Note:** No features reached significance (p-value < ",
        padj_cut_de, " and |logFC| >= ", round(log2_lfc_de, 2),
        "). With only **", n_samp, " samples**, the test has limited power. ",
        "Consider increasing sample size, relaxing thresholds, or using a ",
        "rank-based method (e.g. GSEA).\n\n", sep = "")
  }
}
```

```{r de-sections, results='asis', fig.width=10, fig.height=7}
if (!is.null(de_contrasts)) {
  for (ctr in de_contrasts) {
    ctr_label <- make_contrast_label(ctr)

    ctr_tbl <- tryCatch(
      extract_contrast_table(de_res$summary_df, ctr_label),
      error = function(e) NULL
    )
    if (is.null(ctr_tbl)) next

    # Add Name column
    ctr_tbl$Name <- vapply(ctr_tbl$feature_id, resolve_name, character(1), USE.NAMES = FALSE)

    # Direction column for coloring (uses raw P.Value)
    ctr_tbl$Direction <- ifelse(
      !is.na(ctr_tbl$P.Value) & ctr_tbl$P.Value <= padj_cut_de &
        !is.na(ctr_tbl$logFC) & abs(ctr_tbl$logFC) >= log2_lfc_de,
      ifelse(ctr_tbl$logFC > 0, "Up", "Down"), "NS"
    )
    ctr_tbl$Direction <- factor(ctr_tbl$Direction, levels = c("NS", "Down", "Up"))
    ctr_tbl$neglog10_pval <- pmin(-log10(pmax(
      ifelse(is.na(ctr_tbl$P.Value), 1, ctr_tbl$P.Value), 1e-300
    )), 50)

    # ================================================================
    # VOLCANO
    # ================================================================
    if (show_de_volcano) {
    cat("\n## Volcano: ", ctr, " {.tabset}\n\n", sep = "")

    # -- Static Volcano --
    # Single static volcano. Default is raw P-value (matches default_ptype="pval"
    # in the cutoff panel); padj can be switched to in the interactive section.
    vol_pval_png <- de_res$files[grepl(paste0("volcano_", ctr_label, "_pval\\.png$"),
                                       basename(de_res$files))]
    vol_pval_png <- vol_pval_png[file.exists(vol_pval_png)]
    if (length(vol_pval_png) == 0) {
      vol_pval_png <- de_res$files[grepl(paste0("volcano_", ctr_label, "(_padj)?\\.png$"),
                                         basename(de_res$files))]
      vol_pval_png <- vol_pval_png[file.exists(vol_pval_png)]
    }
    if (length(vol_pval_png) > 0) {
      cat("### Static\n\n")
      cat(sprintf("\n![Volcano: %s](%s){width=100%%}\n\n", ctr, vol_pval_png[1]))
      figure_legend(paste0("Volcano plot for contrast ", ctr,
        ". Each point is a metabolite; x-axis shows log2(fold change) and y-axis shows -log10(raw P-value). Coloured points exceed both fold-change and significance thresholds."))
      render_commentary(paste0("volcano_", ctr_label))
      cat("\n\n")
    }

    # -- Interactive Volcano --
    if (requireNamespace("plotly", quietly = TRUE)) {
      cat("### Interactive\n\n")

      # Unique ID for this contrast's JS namespace
      vol_cid <- gsub("[^a-zA-Z0-9]", "", ctr_label)

      vol_p <- ggplot2::ggplot(
        ctr_tbl,
        ggplot2::aes(x = logFC, y = neglog10_pval, color = Direction,
                     text = paste0("Metabolite: ", Name,
                                   "<br>log2(FC): ", round(logFC, 3),
                                   "<br>P.Value: ", signif(P.Value, 3),
                                   "<br>adj.P.Val: ", signif(adj.P.Val, 3)))
      ) +
        ggplot2::geom_point(alpha = 0.6, size = 1.5) +
        ggplot2::scale_color_manual(values = c("NS" = "grey70", "Down" = "blue", "Up" = "red")) +
        ggplot2::geom_vline(xintercept = c(-log2_lfc_de, log2_lfc_de), linetype = "dashed", alpha = 0.5) +
        ggplot2::geom_hline(yintercept = -log10(padj_cut_de), linetype = "dashed", alpha = 0.5) +
        ggplot2::labs(x = "log2(FC)", y = "-log10(P.Value)") +
        ggplot2::theme_minimal()

      vol_plotly <- plotly::ggplotly(vol_p, tooltip = "text")

      # Emit search box + highlight JS (matching RNA/prot pattern)
      cat(sprintf('
<div style="margin-bottom: 15px; position: relative;">
<label for="volcano_genes_%s"><strong>Highlight metabolites:</strong></label>
<div style="display: inline-block; position: relative; margin-left: 10px;">
<input type="text" id="volcano_genes_%s" placeholder="Start typing a metabolite name..."
  autocomplete="off"
  style="padding: 8px; border-radius: 4px; border: 1px solid #ccc; width: 350px;"
  oninput="volcanoAutocomplete_%s(this)" onkeydown="volcanoAutoKey_%s(event)">
<div id="volcano_ac_%s" style="position: absolute; top: 100%%; left: 0; width: 350px; max-height: 200px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 4px 4px; z-index: 1000; display: none; box-shadow: 0 4px 6px rgba(0,0,0,0.1);"></div>
</div>
<button onclick="highlightVolcanoGenes_%s()" style="padding: 8px 15px; margin-left: 5px; background: #007bff; color: white; border: none; border-radius: 4px; cursor: pointer;">Highlight</button>
<button onclick="clearVolcanoHighlight_%s()" style="padding: 8px 15px; margin-left: 5px; background: #6c757d; color: white; border: none; border-radius: 4px; cursor: pointer;">Clear</button>
<div id="volcano_tags_%s" style="margin-top: 5px; display: flex; flex-wrap: wrap; gap: 4px;"></div>
</div>
', vol_cid, vol_cid, vol_cid, vol_cid, vol_cid, vol_cid, vol_cid, vol_cid))

      cat(sprintf('<div id="volcano_plot_%s">\n', vol_cid))
      print(htmltools::tagList(vol_plotly))
      cat('</div>\n')

      # Register volcano plot for cutoff panel
      if (exists("cutoff_register_plot")) {
        vol_reg_df <- data.frame(
          name = as.character(ctr_tbl$Name),
          logFC = ctr_tbl$logFC,
          pval = ctr_tbl$P.Value,
          adjPval = ctr_tbl$adj.P.Val,
          stringsAsFactors = FALSE
        )
        # Pipeline pass flag — single source of truth for DE counts
        pass_col_m <- paste0("pass.", ctr_label)
        if (!is.null(de_res$summary_df) && pass_col_m %in% names(de_res$summary_df)) {
          pf <- as.numeric(de_res$summary_df[[pass_col_m]])
          # Align to ctr_tbl rows by feature_id
          pf_aligned <- pf[match(ctr_tbl$feature_id, de_res$summary_df$feature_id)]
          vol_reg_df$passFlag <- ifelse(is.na(pf_aligned), 0, pf_aligned)
        }
        # The plotly widget is inside the wrapper div; JS uses querySelector fallback
        cat(cutoff_register_plot(
          paste0("volcano_plot_", vol_cid),
          vol_reg_df, "volcano", "Metabolite", ctr
        ))
      }

      # JS for autocomplete + highlight
      cat(sprintf('<script>
(function() {
  var cid = "%s";
  var selectedGenes = [];
  var acIndex = -1;
  function getMetaboliteNames() {
    var plotDiv = document.getElementById("volcano_plot_" + cid);
    if (!plotDiv) plotDiv = document.querySelector("#volcano_plot_" + cid + " .plotly");
    var el = plotDiv ? plotDiv.querySelector(".js-plotly-plot") || plotDiv : null;
    if (!el || !el.data) return [];
    var names = [];
    for (var t = 0; t < el.data.length; t++) {
      var d = el.data[t];
      if (!d.text) continue;
      for (var i = 0; i < d.text.length; i++) {
        var txt = typeof d.text[i] === "string" ? d.text[i] : "";
        var m = txt.match(/Metabolite:\\s*(.+?)(?:<br>|$)/);
        if (m && m[1] && names.indexOf(m[1]) === -1) names.push(m[1]);
      }
    }
    return names.sort();
  }
  function renderTags() {
    var c = document.getElementById("volcano_tags_" + cid);
    c.innerHTML = "";
    selectedGenes.forEach(function(g) {
      var tag = document.createElement("span");
      tag.style.cssText = "display:inline-flex;align-items:center;background:#007bff;color:white;padding:2px 8px;border-radius:12px;font-size:12px;gap:4px;";
      var txt = document.createTextNode(g + " ");
      tag.appendChild(txt);
      var btn = document.createElement("span");
      btn.textContent = String.fromCharCode(215);
      btn.style.cssText = "cursor:pointer;font-weight:bold;";
      btn.addEventListener("click", (function(gene){return function(){window["removeVolcanoGene_"+cid](gene);};})(g));
      tag.appendChild(btn);
      c.appendChild(tag);
    });
  }
  function applyHighlight() {
    var plotDiv = document.getElementById("volcano_plot_" + cid);
    var el = plotDiv ? plotDiv.querySelector(".js-plotly-plot") || plotDiv : null;
    if (!el || !el.data) return;
    if (selectedGenes.length === 0) { Plotly.relayout(el, {annotations:[]}); return; }
    var gl = selectedGenes.map(function(g){return g.toLowerCase();});
    var ann = [];
    for (var t = 0; t < el.data.length; t++) {
      var d = el.data[t];
      if (!d.text) continue;
      for (var i = 0; i < d.text.length; i++) {
        var txt = typeof d.text[i] === "string" ? d.text[i] : "";
        var m2 = txt.match(/Metabolite:\\s*(.+?)(?:<br>|$)/);
        if (m2 && gl.indexOf(m2[1].toLowerCase()) !== -1) {
          ann.push({x:d.x[i],y:d.y[i],text:m2[1],showarrow:true,arrowhead:2,ax:20,ay:-30,
            font:{size:11,color:"black"},bgcolor:"rgba(255,255,255,0.9)",bordercolor:"black",borderwidth:1});
        }
      }
    }
    Plotly.relayout(el, {annotations:ann});
  }
  window["volcanoAutocomplete_" + cid] = function(el) {
    var ac = document.getElementById("volcano_ac_" + cid);
    var val = el.value.trim().toLowerCase(); ac.innerHTML = ""; acIndex = -1;
    if (val.length < 1) {ac.style.display="none"; return;}
    var all = getMetaboliteNames();
    var matches = all.filter(function(g){return g.toLowerCase().indexOf(val)!==-1 && selectedGenes.indexOf(g)===-1;}).slice(0,15);
    if (matches.length===0){ac.style.display="none"; return;}
    matches.forEach(function(g){
      var item=document.createElement("div"); item.textContent=g;
      item.style.cssText="padding:6px 10px;cursor:pointer;font-size:13px;";
      item.onmouseenter=function(){this.style.background="#e9ecef";};
      item.onmouseleave=function(){this.style.background="white";};
      item.onclick=function(){selectedGenes.push(g);el.value="";ac.style.display="none";renderTags();applyHighlight();};
      ac.appendChild(item);
    });
    ac.style.display="block";
  };
  window["volcanoAutoKey_" + cid] = function(e) {
    var ac = document.getElementById("volcano_ac_" + cid);
    var items = ac.children;
    if (e.key==="ArrowDown"){acIndex=Math.min(acIndex+1,items.length-1);for(var i=0;i<items.length;i++)items[i].style.background=i===acIndex?"#e9ecef":"white";}
    else if(e.key==="ArrowUp"){acIndex=Math.max(acIndex-1,0);for(var i=0;i<items.length;i++)items[i].style.background=i===acIndex?"#e9ecef":"white";}
    else if(e.key==="Enter"&&acIndex>=0&&items[acIndex]){items[acIndex].click();e.preventDefault();}
  };
  window["highlightVolcanoGenes_" + cid] = function() {
    var el = document.getElementById("volcano_genes_" + cid);
    var val = el.value.trim();
    if (val && selectedGenes.indexOf(val)===-1) { selectedGenes.push(val); el.value=""; renderTags(); applyHighlight(); }
  };
  window["clearVolcanoHighlight_" + cid] = function() { selectedGenes=[]; renderTags(); applyHighlight(); };
  window["removeVolcanoGene_" + cid] = function(g) { selectedGenes=selectedGenes.filter(function(x){return x!==g;}); renderTags(); applyHighlight(); };
})();
</script>\n', vol_cid))

      figure_legend(paste0("Interactive volcano plot for ", ctr, ". Type metabolite names in the search box to annotate them on the plot. Hover over points for details."))
      cat("\n\n")
    }

    } # end show_de_volcano

    # ================================================================
    # MA PLOT  (default-off after Apr 2026 reviewer feedback; toggle via
    #           report_sections$de$ma: true to restore)
    # ================================================================
    if (show_de_ma) {
    ma_png <- de_res$files[grepl(paste0("MA_", ctr_label, "\\.png$"), basename(de_res$files))]
    ma_png <- ma_png[file.exists(ma_png)]
    if (length(ma_png) > 0) {
      cat("\n## MA Plot: ", ctr, " {.tabset}\n\n", sep = "")

      cat("### Static\n\n")
      cat(sprintf("\n![MA Plot: %s](%s){width=100%%}\n\n", ctr, ma_png[1]))
      figure_legend(paste0("MA plot for contrast ", ctr, ". The x-axis shows average expression and the y-axis shows log2 fold change. Significant features are highlighted."))
      cat("\n\n")

      # Interactive MA
      if (requireNamespace("plotly", quietly = TRUE)) {
        cat("### Interactive\n\n")

        ma_cid <- paste0("ma_", gsub("[^a-zA-Z0-9]", "", ctr_label))

        ma_p <- ggplot2::ggplot(
          ctr_tbl,
          ggplot2::aes(x = AveExpr, y = logFC, color = Direction,
                       text = paste0("Metabolite: ", Name,
                                     "<br>log2(FC): ", round(logFC, 3),
                                     "<br>AveExpr: ", round(AveExpr, 3),
                                     "<br>P.Value: ", signif(P.Value, 3),
                                     "<br>adj.P.Val: ", signif(adj.P.Val, 3)))
        ) +
          ggplot2::geom_point(alpha = 0.6, size = 1.5) +
          ggplot2::scale_color_manual(values = c("NS" = "grey70", "Down" = "blue", "Up" = "red")) +
          ggplot2::geom_hline(yintercept = c(-log2_lfc_de, 0, log2_lfc_de),
                              linetype = c("dashed", "solid", "dashed"),
                              alpha = c(0.5, 0.3, 0.5)) +
          ggplot2::labs(x = "Average Expression", y = "log2(FC)") +
          ggplot2::theme_minimal()

        ma_plotly <- plotly::ggplotly(ma_p, tooltip = "text")

        cat(sprintf('
<div style="margin-bottom: 15px; position: relative;">
<label for="ma_genes_%s"><strong>Highlight metabolites:</strong></label>
<div style="display: inline-block; position: relative; margin-left: 10px;">
<input type="text" id="ma_genes_%s" placeholder="Start typing a metabolite name..."
  autocomplete="off"
  style="padding: 8px; border-radius: 4px; border: 1px solid #ccc; width: 350px;"
  oninput="maAutocomplete_%s(this)" onkeydown="maAutoKey_%s(event)">
<div id="ma_ac_%s" style="position: absolute; top: 100%%; left: 0; width: 350px; max-height: 200px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 4px 4px; z-index: 1000; display: none; box-shadow: 0 4px 6px rgba(0,0,0,0.1);"></div>
</div>
<button onclick="highlightMAGenes_%s()" style="padding: 8px 15px; margin-left: 5px; background: #007bff; color: white; border: none; border-radius: 4px; cursor: pointer;">Highlight</button>
<button onclick="clearMAHighlight_%s()" style="padding: 8px 15px; margin-left: 5px; background: #6c757d; color: white; border: none; border-radius: 4px; cursor: pointer;">Clear</button>
<div id="ma_tags_%s" style="margin-top: 5px; display: flex; flex-wrap: wrap; gap: 4px;"></div>
</div>
', ma_cid, ma_cid, ma_cid, ma_cid, ma_cid, ma_cid, ma_cid, ma_cid))

        cat(sprintf('<div id="ma_plot_%s">\n', ma_cid))
        print(htmltools::tagList(ma_plotly))
        cat('</div>\n')

        # Register MA plot for cutoff panel
        if (exists("cutoff_register_plot")) {
          ma_reg_df <- data.frame(
            name = as.character(ctr_tbl$Name),
            logFC = ctr_tbl$logFC,
            pval = ctr_tbl$P.Value,
            avgExpr = ctr_tbl$AveExpr,
            adjPval = ctr_tbl$adj.P.Val,
            stringsAsFactors = FALSE
          )
          pass_col_m <- paste0("pass.", ctr_label)
          if (!is.null(de_res$summary_df) && pass_col_m %in% names(de_res$summary_df)) {
            pf <- as.numeric(de_res$summary_df[[pass_col_m]])
            pf_aligned <- pf[match(ctr_tbl$feature_id, de_res$summary_df$feature_id)]
            ma_reg_df$passFlag <- ifelse(is.na(pf_aligned), 0, pf_aligned)
          }
          cat(cutoff_register_plot(
            paste0("ma_plot_", ma_cid),
            ma_reg_df, "ma", "Metabolite", ctr
          ))
        }

        cat(sprintf('<script>
(function() {
  var cid = "%s"; var selectedGenes = []; var acIndex = -1;
  function getNames() {
    var plotDiv = document.getElementById("ma_plot_" + cid);
    var el = plotDiv ? plotDiv.querySelector(".js-plotly-plot") || plotDiv : null;
    if (!el || !el.data) return [];
    var names = [];
    for (var t = 0; t < el.data.length; t++) {
      var d = el.data[t]; if (!d.text) continue;
      for (var i = 0; i < d.text.length; i++) {
        var m = (typeof d.text[i]==="string"?d.text[i]:"").match(/Metabolite:\\s*(.+?)(?:<br>|$)/);
        if (m && m[1] && names.indexOf(m[1])===-1) names.push(m[1]);
      }
    } return names.sort();
  }
  function renderTags() {
    var c = document.getElementById("ma_tags_" + cid); c.innerHTML = "";
    selectedGenes.forEach(function(g) {
      var tag = document.createElement("span");
      tag.style.cssText = "display:inline-flex;align-items:center;background:#007bff;color:white;padding:2px 8px;border-radius:12px;font-size:12px;gap:4px;";
      var txt = document.createTextNode(g + " ");
      tag.appendChild(txt);
      var btn = document.createElement("span");
      btn.textContent = String.fromCharCode(215);
      btn.style.cssText = "cursor:pointer;font-weight:bold;";
      btn.addEventListener("click",(function(gene){return function(){window["removeMAGene_"+cid](gene);};})(g));
      tag.appendChild(btn);
      c.appendChild(tag);
    });
  }
  function applyHighlight() {
    var plotDiv = document.getElementById("ma_plot_" + cid);
    var el = plotDiv ? plotDiv.querySelector(".js-plotly-plot") || plotDiv : null;
    if (!el || !el.data) return;
    if (selectedGenes.length===0){Plotly.relayout(el,{annotations:[]});return;}
    var gl = selectedGenes.map(function(g){return g.toLowerCase();});
    var ann = [];
    for (var t=0;t<el.data.length;t++){var d=el.data[t];if(!d.text)continue;
      for(var i=0;i<d.text.length;i++){var m2=(typeof d.text[i]==="string"?d.text[i]:"").match(/Metabolite:\\s*(.+?)(?:<br>|$)/);
        if(m2&&gl.indexOf(m2[1].toLowerCase())!==-1){ann.push({x:d.x[i],y:d.y[i],text:m2[1],showarrow:true,arrowhead:2,ax:20,ay:-30,font:{size:11,color:"black"},bgcolor:"rgba(255,255,255,0.9)",bordercolor:"black",borderwidth:1});}
    }} Plotly.relayout(el,{annotations:ann});
  }
  window["maAutocomplete_"+cid]=function(el){var ac=document.getElementById("ma_ac_"+cid);var val=el.value.trim().toLowerCase();ac.innerHTML="";acIndex=-1;if(val.length<1){ac.style.display="none";return;}var all=getNames();var matches=all.filter(function(g){return g.toLowerCase().indexOf(val)!==-1&&selectedGenes.indexOf(g)===-1;}).slice(0,15);if(matches.length===0){ac.style.display="none";return;}matches.forEach(function(g){var item=document.createElement("div");item.textContent=g;item.style.cssText="padding:6px 10px;cursor:pointer;font-size:13px;";item.onmouseenter=function(){this.style.background="#e9ecef";};item.onmouseleave=function(){this.style.background="white";};item.onclick=function(){selectedGenes.push(g);el.value="";ac.style.display="none";renderTags();applyHighlight();};ac.appendChild(item);});ac.style.display="block";};
  window["maAutoKey_"+cid]=function(e){var ac=document.getElementById("ma_ac_"+cid);var items=ac.children;if(e.key==="ArrowDown"){acIndex=Math.min(acIndex+1,items.length-1);for(var i=0;i<items.length;i++)items[i].style.background=i===acIndex?"#e9ecef":"white";}else if(e.key==="ArrowUp"){acIndex=Math.max(acIndex-1,0);for(var i=0;i<items.length;i++)items[i].style.background=i===acIndex?"#e9ecef":"white";}else if(e.key==="Enter"&&acIndex>=0&&items[acIndex]){items[acIndex].click();e.preventDefault();}};
  window["highlightMAGenes_"+cid]=function(){var el=document.getElementById("ma_genes_"+cid);var val=el.value.trim();if(val&&selectedGenes.indexOf(val)===-1){selectedGenes.push(val);el.value="";renderTags();applyHighlight();}};
  window["clearMAHighlight_"+cid]=function(){selectedGenes=[];renderTags();applyHighlight();};
  window["removeMAGene_"+cid]=function(g){selectedGenes=selectedGenes.filter(function(x){return x!==g;});renderTags();applyHighlight();};
})();
</script>\n', ma_cid))
        figure_legend(paste0("Interactive MA plot for ", ctr, ". Search and highlight specific metabolites."))
        cat("\n\n")
      }
    }

    } # end show_de_ma

    # ================================================================
    # P-VALUE DISTRIBUTION  (default-off after Apr 2026 reviewer feedback;
    #                        toggle via report_sections$de$pval_hist: true)
    # ================================================================
    if (show_de_pvalhist) {
    phist_png <- de_res$files[grepl(paste0("pvalue_hist_", ctr_label, "\\.png$"), basename(de_res$files))]
    phist_png <- phist_png[file.exists(phist_png)]
    if (length(phist_png) > 0) {
      cat("\n## P-value Distribution: ", ctr, "\n\n", sep = "")
      cat(sprintf("\n![P-value Distribution: %s](%s){width=85%%}\n\n", ctr, phist_png[1]))
      figure_legend(paste0("Distribution of raw p-values for contrast ", ctr, ". A uniform distribution suggests no differential signal; enrichment near zero indicates true positives."))
      cat("\n\n")
    }

    } # end show_de_pvalhist

    # ================================================================
    # DE RESULTS TABLE (statistics + normalized intensity)
    # ================================================================
    if (show_de_tables) {
    cat("\n## DE Results: ", ctr, " {.tabset}\n\n", sep = "")

    # Top 50 table
    top_tbl <- ctr_tbl[order(ctr_tbl$P.Value), ]
    top_tbl <- head(top_tbl, 50)
    top_display <- top_tbl[, c("feature_id", "Name", "logFC", "AveExpr", "P.Value", "adj.P.Val", "Direction")]
    cat("### Top 50 DE Metabolites\n\n")
    top_dt <- make_dt(top_display, caption = paste0("Top 50 DE metabolites: ", ctr))
    cat(knitr::knit_print(top_dt))
    cat("\n\n")

    # Full table with normalized intensities
    cat("### Full Results + Expression\n\n")
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>",
        "Full DE results merged with normalized expression values. ",
        "Use the search box to find specific metabolites and the buttons to export.</em></p>\n")

    full_tbl <- ctr_tbl[order(ctr_tbl$P.Value), ,drop=FALSE]
    full_display <- full_tbl[, c("feature_id", "Name", "logFC", "AveExpr", "P.Value", "adj.P.Val", "Direction")]

    # Merge transformed expression (glog10, before scaling) for meaningful values
    de_tbl_expr <- pre$expr_log %||% pre$expr_work
    if (!is.null(de_tbl_expr)) {
      expr_df2 <- as.data.frame(de_tbl_expr, check.names = FALSE)
      expr_df2$feature_id <- rownames(de_tbl_expr)
      merged <- merge(full_display, expr_df2, by = "feature_id", all.x = TRUE)
      if (nrow(merged) > 0) full_display <- merged
    }

    full_dt <- DT::datatable(
      full_display,
      caption = paste0("DE Results + Normalized Expression: ", ctr),
      extensions = "Buttons",
      options = list(
        scrollX = TRUE, pageLength = 10, dom = "Bfrtip",
        buttons = list(
          list(extend = "csv", filename = paste0("DE_expression_", ctr_label)),
          list(extend = "excel", filename = paste0("DE_expression_", ctr_label))
        ),
        search = list(regex = TRUE, caseInsensitive = TRUE)
      ),
      rownames = FALSE, filter = "top",
      class = "compact stripe hover"
    ) |> DT::formatSignif(
      columns = which(vapply(full_display, is.numeric, logical(1))), digits = 4
    )
    cat(knitr::knit_print(full_dt))
    cat("\n\n")

    } # end show_de_tables

  } # end contrast loop
}
```

```{r metabolite-explorer-prep, include=FALSE}
# Prepare data for the metabolite expression explorer
explorer_ready <- FALSE
explorer_json_str <- "{}"

# Use expr_log (transform-only, e.g. glog10) for the explorer so that means
# reflect actual intensity levels rather than z-scored zeros.
# Fall back to expr_work if expr_log is not available.
explorer_expr <- pre$expr_log %||% pre$expr_work

if (!is.null(explorer_expr) && !is.null(pre$meta)) {
  # Build metabolite display names
  metab_ids <- rownames(explorer_expr)
  metab_display <- vapply(metab_ids, resolve_name, character(1), USE.NAMES = FALSE)

  # Build sample -> group mapping
  sample_col_exp <- metab_cfg$effects$samples %||% "sample_id"
  group_col_exp  <- cond_col
  meta_exp       <- pre$meta

  if (sample_col_exp %in% names(meta_exp) && group_col_exp %in% names(meta_exp)) {
    group_map_exp <- stats::setNames(as.character(meta_exp[[group_col_exp]]),
                                      meta_exp[[sample_col_exp]])
    sample_groups_exp <- as.character(group_map_exp[colnames(explorer_expr)])
    sample_groups_exp[is.na(sample_groups_exp)] <- "Unknown"
  } else {
    sample_groups_exp <- rep("All", ncol(explorer_expr))
  }

  # Build expression matrix as list of numeric vectors
  expr_matrix_list <- lapply(seq_len(nrow(explorer_expr)), function(i) {
    round(as.numeric(explorer_expr[i, ]), 3)
  })

  explorer_data <- list(
    genes   = metab_display,
    samples = colnames(explorer_expr),
    groups  = sample_groups_exp,
    expr    = expr_matrix_list
  )

  explorer_json_str <- jsonlite::toJSON(explorer_data, auto_unbox = FALSE)
  explorer_ready <- TRUE
}
```

`r if (explorer_ready && show_de_explorer) "" else "<!--"`

## Metabolite Explorer

Search for metabolites and visualize their transformed expression (glog10, before scaling) across groups.

```{r metabolite-explorer, results='asis'}
if (explorer_ready) {
  cat(sprintf('
<div style="margin: 15px 0 20px 0;">
<div style="display: flex; align-items: center; gap: 10px; flex-wrap: wrap;">
<label for="boxplot_gene_input" style="font-weight: bold; white-space: nowrap;">Search metabolite:</label>
<div style="position: relative; flex: 1; max-width: 400px;">
<input type="text" id="boxplot_gene_input" placeholder="Start typing a metabolite name..."
  autocomplete="off"
  style="padding: 10px 12px; border-radius: 6px; border: 1px solid #ccc; width: 100%%; font-size: 14px; box-sizing: border-box;"
  oninput="boxplotAutocomplete(this)" onkeydown="boxplotAutoKey(event)">
<div id="boxplot_ac" style="position: absolute; top: 100%%; left: 0; width: 100%%; max-height: 220px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 6px 6px; z-index: 1000; display: none; box-shadow: 0 4px 8px rgba(0,0,0,0.12);"></div>
</div>
<button onclick="addGeneToPlot()" style="padding: 10px 20px; background: #007bff; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Add</button>
<button onclick="clearBoxplot()" style="padding: 10px 20px; background: #6c757d; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Clear All</button>
</div>
<div id="boxplot_gene_tags" style="margin-top: 8px; display: flex; flex-wrap: wrap; gap: 6px;"></div>
<div id="boxplot_gene_info" style="margin-top: 8px; font-size: 0.9em; color: #555;"></div>
</div>
<div id="boxplot_container" style="width: 100%%; min-height: 450px; margin-bottom: 10px;"></div>
<div id="boxplot_export_bar" style="display: none; margin-bottom: 20px; gap: 8px; align-items: center;">
  <span style="font-size: 0.85em; color: #555; margin-right: 4px;">Export:</span>
  <button onclick="exportBoxplot(\'svg\')" style="padding: 5px 14px; background: #28a745; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">SVG</button>
  <button onclick="exportBoxplot(\'png\')" style="padding: 5px 14px; background: #17a2b8; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">PNG</button>
</div>
'))

  # Inject JS data and rendering logic
  cat(sprintf('<script>
var geneExprData = %s;
var bpGroupLabel = "%s";
var bpYLabel = "%s";
var boxplotAcIndex = -1;
var selectedGenes = [];

function boxplotAutocomplete(el) {
  var ac = document.getElementById("boxplot_ac");
  var val = el.value.trim().toLowerCase(); ac.innerHTML = ""; boxplotAcIndex = -1;
  if (val.length < 1) { ac.style.display = "none"; return; }
  var matches = [];
  for (var i = 0; i < geneExprData.genes.length; i++) {
    if (geneExprData.genes[i].toLowerCase().indexOf(val) !== -1) {
      if (selectedGenes.some(function(g){return g.idx===i;})) continue;
      matches.push({name: geneExprData.genes[i], idx: i});
      if (matches.length >= 20) break;
    }
  }
  if (matches.length === 0) { ac.style.display = "none"; return; }
  matches.forEach(function(m) {
    var item = document.createElement("div"); item.textContent = m.name;
    item.style.cssText = "padding: 8px 12px; cursor: pointer; font-size: 13px; border-bottom: 1px solid #f0f0f0;";
    item.onmouseenter = function() { this.style.background = "#e9ecef"; };
    item.onmouseleave = function() { this.style.background = "white"; };
    item.onclick = function() { document.getElementById("boxplot_gene_input").value = ""; ac.style.display = "none"; addGeneByIdx(m.idx, m.name); };
    ac.appendChild(item);
  });
  ac.style.display = "block";
}
function boxplotAutoKey(e) {
  var ac = document.getElementById("boxplot_ac"); var items = ac.children;
  if (e.key==="ArrowDown"){boxplotAcIndex=Math.min(boxplotAcIndex+1,items.length-1);for(var i=0;i<items.length;i++)items[i].style.background=i===boxplotAcIndex?"#e9ecef":"white";}
  else if(e.key==="ArrowUp"){boxplotAcIndex=Math.max(boxplotAcIndex-1,0);for(var i=0;i<items.length;i++)items[i].style.background=i===boxplotAcIndex?"#e9ecef":"white";}
  else if(e.key==="Enter"){if(boxplotAcIndex>=0&&items[boxplotAcIndex])items[boxplotAcIndex].click(); else addGeneToPlot(); e.preventDefault();}
}
function addGeneToPlot() {
  var el = document.getElementById("boxplot_gene_input");
  var val = el.value.trim(); if (!val) return;
  var vl = val.toLowerCase();
  for (var i = 0; i < geneExprData.genes.length; i++) {
    if (geneExprData.genes[i].toLowerCase() === vl) { addGeneByIdx(i, geneExprData.genes[i]); el.value = ""; return; }
  }
  // Partial match
  for (var i = 0; i < geneExprData.genes.length; i++) {
    if (geneExprData.genes[i].toLowerCase().indexOf(vl) !== -1) { addGeneByIdx(i, geneExprData.genes[i]); el.value = ""; return; }
  }
}
function addGeneByIdx(idx, name) {
  if (selectedGenes.some(function(g){return g.idx===idx;})) return;
  selectedGenes.push({name:name, idx:idx}); renderGeneTags(); renderMultiBoxplot();
}
function removeGene(idx) {
  selectedGenes = selectedGenes.filter(function(g){return g.idx!==idx;});
  renderGeneTags(); if (selectedGenes.length>0) renderMultiBoxplot();
  else { document.getElementById("boxplot_container").innerHTML=""; document.getElementById("boxplot_gene_info").innerHTML=""; document.getElementById("boxplot_export_bar").style.display="none"; }
}
function clearBoxplot() {
  selectedGenes = []; renderGeneTags();
  document.getElementById("boxplot_container").innerHTML = "";
  document.getElementById("boxplot_gene_info").innerHTML = "";
  document.getElementById("boxplot_export_bar").style.display = "none";
}
function renderGeneTags() {
  var c = document.getElementById("boxplot_gene_tags"); c.innerHTML = "";
  selectedGenes.forEach(function(g) {
    var tag = document.createElement("span");
    tag.style.cssText = "display:inline-flex;align-items:center;background:#007bff;color:white;padding:3px 10px;border-radius:14px;font-size:12px;gap:4px;";
    var txt = document.createTextNode(g.name + " ");
    tag.appendChild(txt);
    var btn = document.createElement("span");
    btn.textContent = String.fromCharCode(215);
    btn.style.cssText = "cursor:pointer;font-weight:bold;margin-left:3px;";
    btn.addEventListener("click", (function(idx){return function(){removeGene(idx);};})(g.idx));
    tag.appendChild(btn);
    c.appendChild(tag);
  });
}
function renderMultiBoxplot() {
  if (selectedGenes.length === 0) return;
  var groups = geneExprData.groups; var samples = geneExprData.samples;
  var uGroups = []; groups.forEach(function(g){if(uGroups.indexOf(g)===-1)uGroups.push(g);});
  var colors = ["#e41a1c","#377eb8","#4daf4a","#984ea3","#ff7f00","#a65628","#f781bf","#999999"];
  var nGenes = selectedGenes.length; var traces = [];
  if (nGenes === 1) {
    var gIdx = selectedGenes[0].idx; var values = geneExprData.expr[gIdx];
    uGroups.forEach(function(g, gi) {
      var gv = [], gs = [];
      for (var i=0;i<values.length;i++) { if(groups[i]===g){gv.push(values[i]);gs.push(samples[i]);} }
      traces.push({y:gv,type:"box",name:g,boxpoints:"all",jitter:0.4,pointpos:0,text:gs,
        hovertemplate:"Sample: %%{text}<br>Expression: %%{y:.3f}<extra>"+g+"</extra>",
        marker:{color:colors[gi%%colors.length],size:6,opacity:0.7},
        line:{color:colors[gi%%colors.length]},fillcolor:colors[gi%%colors.length]+"33"});
    });
    var layout = {title:{text:selectedGenes[0].name+" Expression",font:{size:16}},yaxis:{title:bpYLabel},xaxis:{title:bpGroupLabel},showlegend:false,margin:{t:50,b:80,l:70,r:30},boxgap:0.3};
  } else {
    uGroups.forEach(function(g, gi) {
      var xv=[],yv=[],st=[];
      selectedGenes.forEach(function(sg){var vals=geneExprData.expr[sg.idx];for(var i=0;i<vals.length;i++){if(groups[i]===g){xv.push(sg.name);yv.push(vals[i]);st.push(samples[i]);}}});
      traces.push({x:xv,y:yv,type:"box",name:g,boxpoints:"all",jitter:0.4,pointpos:0,text:st,
        hovertemplate:"%%{x}<br>Sample: %%{text}<br>Expression: %%{y:.3f}<extra>"+g+"</extra>",
        marker:{color:colors[gi%%colors.length],size:5,opacity:0.7},
        line:{color:colors[gi%%colors.length]},fillcolor:colors[gi%%colors.length]+"33"});
    });
    var gNames = selectedGenes.map(function(sg){return sg.name;});
    var layout = {title:{text:"Metabolite Expression Comparison",font:{size:16}},yaxis:{title:bpYLabel},xaxis:{title:"Metabolite",categoryorder:"array",categoryarray:gNames},showlegend:true,legend:{title:{text:bpGroupLabel}},boxmode:"group",margin:{t:50,b:100,l:70,r:30},boxgap:0.15,boxgroupgap:0.1};
  }
  var ch = Math.max(450, 350 + nGenes * 30);
  document.getElementById("boxplot_container").style.height = ch + "px";
  Plotly.newPlot("boxplot_container", traces, layout, {responsive:true});
  var info = selectedGenes.map(function(sg){var vals=geneExprData.expr[sg.idx].filter(function(v){return !isNaN(v);});var mean=vals.reduce(function(a,b){return a+b;},0)/vals.length;return "<strong>"+sg.name+"</strong>: mean "+mean.toFixed(3)+" ("+vals.length+" samples)";});
  document.getElementById("boxplot_gene_info").innerHTML = info.join(" &nbsp;|&nbsp; ");
  document.getElementById("boxplot_export_bar").style.display = "flex";
}
function exportBoxplot(fmt) {
  var c = document.getElementById("boxplot_container");
  if (!c||!c.data){alert("No plot to export.");return;}
  var gNames = selectedGenes.map(function(g){return g.name;}).join("_");
  var w = fmt==="svg"?1200:2400, h = fmt==="svg"?800:1600, s = fmt==="svg"?1:3;
  Plotly.downloadImage(c,{format:fmt,width:w,height:h,scale:s,filename:"metabolite_expression_"+gNames});
}
</script>\n', explorer_json_str, group_col_exp, norm_y_label_report))

  figure_legend("Metabolite Expression Explorer. Search for metabolites by name, add multiple for side-by-side comparison. Each box shows the distribution of normalized expression values per group. Individual sample points are overlaid. Use the export buttons to save as SVG or PNG.")
}
```

`r if (explorer_ready && show_de_explorer) "" else "-->"`

`r if (is.null(de_res) || is.null(de_res$summary_df) || !show_de) "-->"`

---

`r if (is.null(de_res) || is.null(de_res$summary_df) || !show_de_heatmaps) "<!--"`

# Top DE Heatmaps {.tabset}

Heatmaps of differentially expressed metabolites (adjusted p-value < `r padj_cut_de`, z-score scaled) with hierarchical clustering.

```{r de-heatmaps, results='asis', fig.width=10, fig.height=8}
if (!is.null(de_res$summary_df) && requireNamespace("pheatmap", quietly = TRUE)) {
  # Find the Diagnostic_plots directory for heatmap images
  hm_out_dir <- NULL
  if (length(de_res$files) > 0) {
    png_files <- de_res$files[grepl("\\.png$", de_res$files)]
    if (length(png_files) > 0) {
      diag_dirs <- unique(dirname(png_files))
      diag_dirs <- diag_dirs[dir.exists(diag_dirs)]
      if (length(diag_dirs) > 0) hm_out_dir <- diag_dirs[1]
    }
  }
  if (is.null(hm_out_dir)) hm_out_dir <- file.path(dirname(knitr::current_input(dir = TRUE)), "metabolomics", "Diagnostic_plots")
  dir.create(hm_out_dir, recursive = TRUE, showWarnings = FALSE)

  # Identify DE features for heatmaps using raw P.Value (ignoring FC cutoff).
  # Heatmaps benefit from showing all significant metabolites; the clustering
  # pattern reveals biology even for features with modest fold changes.
  sum_df <- de_res$summary_df
  pval_cols <- grep("^P\\.Value", colnames(sum_df), value = TRUE)
  if (length(pval_cols) > 0) {
    pval_mat <- sum_df[, pval_cols, drop = FALSE]
    min_pval_per_feat <- apply(pval_mat, 1, function(x) min(x, na.rm = TRUE))
    de_feats <- sum_df$feature_id[min_pval_per_feat <= padj_cut_de]
  } else {
    de_feats <- character(0)
  }

  # Sort by minimum raw p-value
  if (length(de_feats) > 0) {
    feat_pval <- min_pval_per_feat[sum_df$feature_id %in% de_feats]
    de_feats <- de_feats[order(feat_pval)]
  }

  # Use expr_log (transform-only) for heatmaps — z-scoring removes scale anyway,
  # but expr_log preserves natural variance structure better than auto-scaled data.
  hm_expr <- pre$expr_log %||% pre$expr_work

  if (length(de_feats) >= 2 && !is.null(hm_expr)) {
    # Build annotation
    anno_df <- NULL
    if (!is.null(pre$meta) && cond_col %in% colnames(pre$meta)) {
      sample_col_hm <- metab_cfg$effects$samples %||% "sample_id"
      anno_df <- data.frame(
        row.names = pre$meta[[sample_col_hm]],
        Group = pre$meta[[cond_col]],
        stringsAsFactors = FALSE
      )
    }

    # Build display labels (use Name if available)
    hm_label_map <- if (!is.null(name_map_v)) name_map_v else stats::setNames(de_feats, de_feats)

    for (n_top in c(10, 25, 50, 100)) {
      if (n_top > length(de_feats)) next
      top_ids <- utils::head(de_feats, n_top)

      mat_hm <- hm_expr[top_ids, , drop = FALSE]
      # Z-score scale rows
      mat_z <- t(scale(t(mat_hm)))
      mat_z[is.na(mat_z)] <- 0

      # Use display names for rows (truncated so long IUPAC names don't shift the plot)
      row_labels <- vapply(rownames(mat_z), function(fid) {
        nm <- hm_label_map[fid]
        if (!is.na(nm) && nzchar(nm)) nm else fid
      }, character(1), USE.NAMES = FALSE)
      rownames(mat_z) <- truncate_label(row_labels, max_len = 40)

      cat(sprintf("\n## Top %d\n\n", n_top))

      # Adjust figure height based on number of features
      fig_h <- max(6, min(20, 4 + n_top * 0.15))
      show_rownames <- n_top <= 60

      hm_file <- file.path(hm_out_dir, sprintf("heatmap_top%d_DE.png", n_top))
      hm_ok <- tryCatch({
        p_hm <- pheatmap::pheatmap(
          mat_z,
          annotation_col = anno_df,
          cluster_rows = TRUE,
          cluster_cols = TRUE,
          show_rownames = show_rownames,
          show_colnames = TRUE,
          fontsize_row = max(6, min(10, 160 / n_top)),
          fontsize_col = 9,
          main = sprintf("Top %d DE Metabolites (z-score)", n_top),
          color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
          border_color = NA,
          silent = TRUE
        )
        ggplot2::ggsave(hm_file, p_hm, width = 10, height = fig_h, dpi = 150)
        TRUE
      }, error = function(e) { message("Heatmap top ", n_top, " failed: ", e$message); FALSE })
      if (hm_ok && file.exists(hm_file)) {
        cat(sprintf("\n![Top %d DE Heatmap](%s){width=100%%}\n\n", n_top, hm_file))
        figure_legend(sprintf(
          "Heatmap of the top %d differentially expressed metabolites (ranked by adjusted p-value). Values are z-score normalized per metabolite. Rows and columns are hierarchically clustered.",
          n_top
        ))
        cat("\n\n")
      }
    }

    # All DE heatmap (if reasonable size and different from last top-N)
    if (length(de_feats) >= 2 && length(de_feats) <= 500) {
      n_all <- length(de_feats)
      mat_all <- hm_expr[de_feats, , drop = FALSE]
      mat_z_all <- t(scale(t(mat_all)))
      mat_z_all[is.na(mat_z_all)] <- 0

      row_labels_all <- vapply(rownames(mat_z_all), function(fid) {
        nm <- hm_label_map[fid]
        if (!is.na(nm) && nzchar(nm)) nm else fid
      }, character(1), USE.NAMES = FALSE)
      rownames(mat_z_all) <- truncate_label(row_labels_all, max_len = 40)

      cat(sprintf("\n## All DE (%d)\n\n", n_all))

      fig_h_all <- max(8, min(30, 4 + n_all * 0.06))
      hm_all_file <- file.path(hm_out_dir, sprintf("heatmap_all%d_DE.png", n_all))
      hm_all_ok <- tryCatch({
        p_hm_all <- pheatmap::pheatmap(
          mat_z_all,
          annotation_col = anno_df,
          cluster_rows = TRUE,
          cluster_cols = TRUE,
          show_rownames = n_all <= 60,
          show_colnames = TRUE,
          fontsize_row = max(4, min(8, 100 / n_all)),
          fontsize_col = 8,
          main = sprintf("All %d DE Metabolites (z-score)", n_all),
          color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
          border_color = NA,
          silent = TRUE
        )
        ggplot2::ggsave(hm_all_file, p_hm_all, width = 10, height = fig_h_all, dpi = 150)
        TRUE
      }, error = function(e) { message("Heatmap all DE failed: ", e$message); FALSE })
      if (hm_all_ok && file.exists(hm_all_file)) {
        cat(sprintf("\n![All DE Heatmap](%s){width=100%%}\n\n", hm_all_file))
      }
      figure_legend(sprintf(
        "Heatmap of all %d differentially expressed metabolites. Values are z-score normalized per metabolite.",
        n_all
      ))
      cat("\n\n")
    }
  } else if (length(de_feats) < 2) {
    cat("\n*Fewer than 2 DE metabolites — heatmap not generated.*\n\n")
  }
}
```

`r if (is.null(de_res) || is.null(de_res$summary_df) || !show_de_heatmaps) "-->"`

---

```{r clustering-discover, include=FALSE}
# ---- Discover clustering outputs on disk for metabolomics ----
clustering_root_metab <- file.path(dirname(knitr::current_input(dir = TRUE)),
                                   "metabolomics", "Clustering")
hier_dir_m   <- file.path(clustering_root_metab, "Hierarchical")
part_root_m  <- file.path(clustering_root_metab, "Partition_clustering")
binary_dir_m <- file.path(clustering_root_metab, "Binary_patterns")

hier_hm_m <- file.path(hier_dir_m, "Hierarchical_DE_heatmap.png")

part_subdirs_m <- if (dir.exists(part_root_m)) {
  list.dirs(part_root_m, recursive = FALSE)
} else character(0)
part_subdir_m <- if (length(part_subdirs_m) > 0) part_subdirs_m[1] else NA_character_
part_hm_m     <- if (!is.na(part_subdir_m)) file.path(part_subdir_m, "Partition_clustering_heatmap.png") else NA_character_

part_profile_pngs_m <- if (!is.na(part_subdir_m) && dir.exists(part_subdir_m)) {
  list.files(part_subdir_m, pattern = "^cluster_profiles_cluster[0-9]+\\.png$",
             full.names = TRUE)
} else character(0)
if (length(part_profile_pngs_m) > 0) {
  ord <- order(as.integer(sub(".*cluster([0-9]+)\\.png$", "\\1",
                              basename(part_profile_pngs_m))))
  part_profile_pngs_m <- part_profile_pngs_m[ord]
}

binary_pngs_m <- if (dir.exists(binary_dir_m)) {
  list.files(binary_dir_m, pattern = "^Heatmap_[01]+\\.png$", full.names = TRUE)
} else character(0)

clustering_enabled_m <- isTRUE(metab_cfg$clustering$enabled)
has_hier_m   <- file.exists(hier_hm_m)
has_part_m   <- !is.na(part_hm_m) && file.exists(part_hm_m)
has_binary_m <- length(binary_pngs_m) > 0
clustering_show_m <- clustering_enabled_m && (has_hier_m || has_part_m || has_binary_m)
```

```{r clustering-header-m, eval=clustering_show_m, results='asis'}
cat("# Clustering {.tabset}\n\n")
cat("Clustering was performed on the differentially expressed metabolites. Each method below answers a different question:\n\n")
cat("- **Hierarchical** — bottom-up tree built from feature-by-feature distances; reveals nested groupings.\n")
cat("- **Partition** — splits features into a chosen number of clusters with similar trajectories across groups.\n")
cat("- **Binary patterns** — assigns each feature to an on/off (1/0) pattern across groups based on correlation.\n\n")
```

```{r clustering-hier-m, eval=clustering_show_m && has_hier_m, results='asis'}
cat("## Hierarchical\n\n")
cat(sprintf("![Hierarchical clustering heatmap](%s){width=100%%}\n\n", hier_hm_m))
figure_legend("Hierarchical clustering heatmap of differentially expressed metabolites (rows z-scored). Row dendrogram reflects feature similarity; samples are annotated by experimental group.")
render_commentary("clustering_hierarchical")
```

```{r clustering-partition-m, eval=clustering_show_m && has_part_m, results='asis'}
cat("## Partition {.tabset}\n\n")
cat("### Heatmap\n\n")
cat(sprintf("![Partition clustering heatmap](%s){width=100%%}\n\n", part_hm_m))
figure_legend("Partition clustering heatmap. Rows are grouped by cluster (gaps separate clusters); within-cluster ordering follows the feature similarity.")

if (length(part_profile_pngs_m) > 0) {
  cat("\n### Cluster Profiles {.tabset}\n\n")
  for (pf in part_profile_pngs_m) {
    k <- sub(".*cluster([0-9]+)\\.png$", "\\1", basename(pf))
    cat(sprintf("\n#### Cluster %s\n\n", k))
    cat(sprintf("![Cluster %s expression profile](%s){width=80%%}\n\n", k, pf))
    figure_legend(sprintf(
      "Mean (±SE) expression profile across experimental groups for metabolites in cluster %s.", k))
  }
}
render_commentary("clustering_partition")
```

```{r clustering-binary-m, eval=clustering_show_m && has_binary_m, results='asis'}
cat("## Binary Patterns {.tabset}\n\n")
for (bf in binary_pngs_m) {
  pat <- sub("^Heatmap_([01]+)\\.png$", "\\1", basename(bf))
  cat(sprintf("\n### Pattern %s\n\n", pat))
  cat(sprintf("![Binary pattern %s](%s){width=100%%}\n\n", pat, bf))
  figure_legend(sprintf(
    "Heatmap of metabolites assigned to binary on/off pattern '%s' across experimental groups.", pat))
}
render_commentary("clustering_binary")
```

```{r clustering-divider-m, eval=clustering_show_m, results='asis'}
cat("\n---\n\n")
```

`r if ((is.null(rf_res) && is.null(plsda_res)) || !show_feature_selection) "<!--"`

# Feature Selection {.tabset}

`r if (is.null(rf_res) || !show_fs_rf) "<!--"`

## Random Forest {.tabset}

### Importance Plot

```{r rf-plot, fig.width=9, fig.height=7, results='asis'}
if (!is.null(rf_res$plots$rf_importance)) {
  safe_print_plot(rf_res$plots$rf_importance)
  figure_legend("Random Forest variable importance plot showing the top features ranked by their contribution to classification accuracy. Higher Mean Decrease Accuracy indicates greater importance.")
  render_commentary("rf_importance")
}
```

### Importance Table

```{r rf-table}
if (!is.null(rf_res$importance_df)) {
  imp_df <- rf_res$importance_df
  sort_col <- if ("MeanDecreaseAccuracy" %in% names(imp_df)) "MeanDecreaseAccuracy" else "importance"
  imp_df <- imp_df[order(-imp_df[[sort_col]]), ]
  make_dt(head(imp_df, 50), caption = "RF Feature Importance (top 50)")
}
```

`r if (is.null(rf_res) || !show_fs_rf) "-->"`

`r if (is.null(plsda_res) || !show_fs_plsda) "<!--"`

## PLS-DA {.tabset}

### Scores Plot

```{r plsda-scores, fig.width=9, fig.height=7, results='asis'}
if (!is.null(plsda_res$plots$plsda_scores)) {
  safe_print_plot(plsda_res$plots$plsda_scores)
  figure_legend("PLS-DA scores plot showing sample separation along the first two latent variables. Clustering by group indicates good discriminatory power of the metabolomic profiles.")
  render_commentary("plsda_scores")
}
```

### VIP Plot

```{r plsda-vip-plot, fig.width=9, fig.height=7, results='asis'}
if (!is.null(plsda_res$plots$plsda_vip)) {
  safe_print_plot(plsda_res$plots$plsda_vip)
  figure_legend("PLS-DA Variable Importance in Projection (VIP) scores. Features with VIP > 1 are generally considered important for group discrimination.")
}
```

### VIP Table

```{r plsda-vip-table}
if (!is.null(plsda_res$vip_df)) {
  vip_df <- plsda_res$vip_df
  vip_col <- if ("VIP" %in% names(vip_df)) "VIP" else grep("^VIP", names(vip_df), value = TRUE)[1]
  vip_df <- vip_df[order(-vip_df[[vip_col]]), ]
  make_dt(head(vip_df, 50), caption = "PLS-DA VIP Scores (top 50)")
}
```

`r if (is.null(plsda_res) || !show_fs_plsda) "-->"`

`r if ((is.null(rf_res) && is.null(plsda_res)) || !show_feature_selection) "-->"`

---

`r if (is.null(enrichment_res) || !show_enrichment) "<!--"`

# Enrichment {.tabset}

`r if (is.null(enrichment_res$qea) || !show_enr_qea) "<!--"`

## QEA {.tabset}

### Lollipop Plot

```{r qea-lollipop, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$qea_lollipop)) {
  safe_print_plot(enrichment_res$plots$qea_lollipop)
  figure_legend("Quantitative Enrichment Analysis (QEA) lollipop plot showing the top enriched metabolic pathways. Bar length represents statistical significance; color indicates enrichment score direction.")
}
```

### Barplot

```{r qea-barplot, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$qea_barplot)) {
  safe_print_plot(enrichment_res$plots$qea_barplot)
  figure_legend("Barplot of QEA enriched pathways ranked by significance. Pathways are ordered from most to least significant.")
}
```

### Results Table

```{r qea-table}
if (!is.null(enrichment_res$qea$table)) {
  make_dt(enrichment_res$qea$table, caption = "QEA Pathway Enrichment")
}
```

`r if (is.null(enrichment_res$qea) || !show_enr_qea) "-->"`

`r if (is.null(enrichment_res$ssgsea) || !show_enr_ssgsea) "<!--"`

## ssGSEA {.tabset}

### Lollipop Plot

```{r ssgsea-lollipop, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$ssgsea_lollipop)) {
  safe_print_plot(enrichment_res$plots$ssgsea_lollipop)
  figure_legend("ssGSEA lollipop plot showing the top enriched metabolic pathways based on single-sample Gene Set Enrichment Analysis scores.")
}
```

### Barplot

```{r ssgsea-barplot, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$ssgsea_barplot)) {
  safe_print_plot(enrichment_res$plots$ssgsea_barplot)
  figure_legend("Barplot of ssGSEA enrichment results, showing pathway enrichment scores ranked by significance.")
}
```

### Boxplots (static)

```{r ssgsea-boxplots, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$ssgsea_boxplots)) {
  safe_print_plot(enrichment_res$plots$ssgsea_boxplots)
  figure_legend("Static boxplots of ssGSEA pathway scores across conditions. Each box represents the distribution of enrichment scores per condition for the top pathways.")
}
```

### Pathway Explorer

```{r ssgsea-interactive, fig.width=10, fig.height=6, results='asis'}
# Interactive boxplot: select one or more pathways, compare conditions
if (!is.null(enrichment_res$ssgsea$scores) &&
    !is.null(enrichment_res$ssgsea$table)) {

  scores_mat <- enrichment_res$ssgsea$scores
  ssgsea_tbl <- enrichment_res$ssgsea$table

  # Resolve condition labels for each sample
  sample_col_i <- metab_cfg$effects$samples %||% "sample_id"
  cond_col_i   <- metab_cfg$enrichment$condition_column %||%
                   metab_cfg$de$condition_column %||%
                   metab_cfg$effects$color %||% "sample_type"
  meta_i       <- params$pre$meta
  conditions_i <- as.character(meta_i[[cond_col_i]][
    match(colnames(scores_mat), meta_i[[sample_col_i]])
  ])

  # Build long-format data for all pathways
  all_pathways <- rownames(scores_mat)
  pw_order     <- ssgsea_tbl$pathway[ssgsea_tbl$pathway %in% all_pathways]
  # Fall back to rowname order for pathways not in the table
  pw_order     <- c(pw_order, setdiff(all_pathways, pw_order))

  long_df <- do.call(rbind, lapply(pw_order, function(pw) {
    data.frame(
      pathway   = pw,
      sample    = colnames(scores_mat),
      score     = as.numeric(scores_mat[pw, ]),
      condition = conditions_i,
      stringsAsFactors = FALSE
    )
  }))

  # Use plotly with a dropdown to select pathways
  if (requireNamespace("plotly", quietly = TRUE)) {

    # Pre-select top 3 pathways by FDR
    default_pws <- utils::head(pw_order, 3)

    # Build one trace per pathway (visible for defaults, hidden for rest)
    fig <- plotly::plot_ly()

    for (pw in pw_order) {
      pw_data <- long_df[long_df$pathway == pw, ]
      vis <- pw %in% default_pws

      fig <- plotly::add_trace(
        fig,
        data = pw_data,
        x = ~condition,
        y = ~score,
        type = "box",
        boxpoints = "all",
        jitter = 0.4,
        pointpos = 0,
        text = ~sample,
        name = pw,
        visible = vis,
        marker = list(size = 6),
        showlegend = FALSE
      )
    }

    # Build dropdown buttons: each button shows one pathway
    buttons <- lapply(seq_along(pw_order), function(i) {
      vis_vec <- rep(FALSE, length(pw_order))
      vis_vec[i] <- TRUE
      pw_short <- if (nchar(pw_order[i]) > 60) {
        paste0(substr(pw_order[i], 1, 57), "...")
      } else {
        pw_order[i]
      }
      list(method = "restyle",
           args = list("visible", as.list(vis_vec)),
           label = pw_short)
    })

    # Add a "Top 3" button at the beginning
    top3_vis <- pw_order %in% default_pws
    buttons <- c(
      list(list(method = "restyle",
                args = list("visible", as.list(top3_vis)),
                label = paste0("Top 3 by FDR"))),
      buttons
    )

    fig <- plotly::layout(
      fig,
      title = list(text = "ssGSEA Scores by Condition",
                    font = list(size = 16)),
      xaxis = list(title = "Condition"),
      yaxis = list(title = "ssGSEA Score"),
      updatemenus = list(
        list(
          type = "dropdown",
          active = 0,
          x = 0.0, xanchor = "left",
          y = 1.15, yanchor = "top",
          buttons = buttons
        )
      )
    )

    print(htmltools::tagList(fig))
    figure_legend("Interactive ssGSEA pathway explorer. Use the dropdown to select individual pathways and compare their enrichment scores across conditions. Points represent individual samples.")
    cat("\n\n")
  } else {
    cat("*Install the `plotly` package for interactive pathway exploration.*\n")
  }
}
```

### Results Table

```{r ssgsea-table}
if (!is.null(enrichment_res$ssgsea$table)) {
  make_dt(enrichment_res$ssgsea$table, caption = "ssGSEA Pathway Enrichment")
}
```

`r if (is.null(enrichment_res$ssgsea) || !show_enr_ssgsea) "-->"`

`r if (is.null(enrichment_res$ora) || !show_enr_ora) "<!--"`

## ORA {.tabset}

### Lollipop Plot

```{r ora-lollipop, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$ora_lollipop)) {
  safe_print_plot(enrichment_res$plots$ora_lollipop)
  figure_legend("Over-Representation Analysis (ORA) lollipop plot showing pathways significantly enriched among DE metabolites using Fisher's exact test.")
}
```

### Dot Plot

```{r ora-dotplot, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$ora_dotplot)) {
  safe_print_plot(enrichment_res$plots$ora_dotplot)
  figure_legend("ORA dot plot displaying enriched pathways. Point size indicates the number of DE metabolites in the pathway; color represents significance.")
}
```

### Results Table

```{r ora-table}
if (!is.null(enrichment_res$ora$table)) {
  make_dt(enrichment_res$ora$table, caption = "ORA Pathway Enrichment (Fisher's Exact Test)")
}
```

`r if (is.null(enrichment_res$ora) || !show_enr_ora) "-->"`

`r if (is.null(enrichment_res$gsea) || !show_enr_gsea) "<!--"`

## GSEA {.tabset}

### Lollipop Plot

```{r gsea-lollipop, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$gsea_lollipop)) {
  safe_print_plot(enrichment_res$plots$gsea_lollipop)
  figure_legend("GSEA lollipop plot showing pathways ranked by Normalized Enrichment Score (NES). Positive NES indicates up-regulation; negative indicates down-regulation of the pathway.")
}
```

### NES Barplot

```{r gsea-nes-barplot, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$gsea_nes_barplot)) {
  safe_print_plot(enrichment_res$plots$gsea_nes_barplot)
  figure_legend("GSEA Normalized Enrichment Score (NES) barplot. Metabolites were pre-ranked by log2 fold change; pathways are ordered by NES magnitude.")
}
```

### Results Table

```{r gsea-table}
if (!is.null(enrichment_res$gsea$table)) {
  make_dt(enrichment_res$gsea$table, caption = "GSEA Pathway Enrichment (fgsea, ranked by log2FC)")
}
```

`r if (is.null(enrichment_res$gsea) || !show_enr_gsea) "-->"`

`r if (is.null(enrichment_res) || !show_enrichment) "-->"`

---

`r if (is.null(params$network_res) || !show_network) "<!--"`

# DE Metabolite Network

```{r network-description, results='asis'}
net <- params$network_res
if (!is.null(net)) {
  cat(sprintf(
    "Interactive network of differentially expressed metabolites connected by **known KEGG reaction pairs** (substrate-product relationships). Nodes are colored by regulation direction (red = up-regulated, blue = down-regulated) and sized by statistical significance (-log10 p-value). Gray edges represent known biochemical reactions from the KEGG database.\n\n"))
  cat(sprintf(
    "**Summary:** %d total DE metabolites | %d with valid KEGG compound IDs | %d connected by reaction pairs | %d edges | %d isolated nodes removed\n\n",
    net$n_total_de, net$n_with_kegg, net$n_connected,
    nrow(net$edges), net$n_isolated))
}
```

```{r network-plotly, fig.width=12, fig.height=10, results='asis'}
if (!is.null(net) && !is.null(net$graph)) {
  # Rebuild plotly inline from graph data
  g <- net$graph
  nodes_df <- net$nodes

  set.seed(42)
  layout_mat <- igraph::layout_with_fr(g)

  node_x <- layout_mat[, 1]
  node_y <- layout_mat[, 2]

  # Edge traces
  edge_x <- c(); edge_y <- c()
  el <- igraph::as_edgelist(g)
  if (nrow(el) > 0) {
    for (i in seq_len(nrow(el))) {
      src <- which(igraph::V(g)$name == el[i, 1])
      tgt <- which(igraph::V(g)$name == el[i, 2])
      edge_x <- c(edge_x, node_x[src], node_x[tgt], NA)
      edge_y <- c(edge_y, node_y[src], node_y[tgt], NA)
    }
  }

  idx <- match(igraph::V(g)$name, nodes_df$name)
  hover_text <- paste0(
    "<b>", nodes_df$name[idx], "</b><br>",
    "KEGG: ", nodes_df$kegg_id[idx], "<br>",
    "log2FC: ", round(nodes_df$logFC[idx], 2), "<br>",
    "p-value: ", signif(nodes_df$pvalue[idx], 3), "<br>",
    "Direction: ", nodes_df$direction[idx])

  fig <- plotly::plot_ly()
  if (length(edge_x) > 0) {
    fig <- plotly::add_trace(fig, x = edge_x, y = edge_y, type = "scatter",
      mode = "lines", line = list(color = "#CCCCCC", width = 1),
      hoverinfo = "none", showlegend = FALSE)
  }
  for (dir in c("Up", "Down")) {
    mask <- nodes_df$direction[idx] == dir
    if (!any(mask)) next
    fig <- plotly::add_trace(fig, x = node_x[mask], y = node_y[mask],
      type = "scatter", mode = "markers+text",
      marker = list(size = nodes_df$size[idx][mask],
                    color = nodes_df$color[idx][mask],
                    line = list(color = "#333333", width = 1)),
      text = nodes_df$name[idx][mask], textposition = "top center",
      textfont = list(size = 8), hovertext = hover_text[mask],
      hoverinfo = "text", name = paste0(dir, "-regulated"))
  }
  fig <- plotly::layout(fig,
    title = list(text = "DE Metabolite Network (KEGG Reaction Pairs)", font = list(size = 16)),
    xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE, title = ""),
    yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE, title = ""),
    plot_bgcolor = "white", paper_bgcolor = "white",
    legend = list(x = 0.02, y = 0.98))
  print(htmltools::tagList(fig))
  figure_legend("Interactive DE metabolite network. Nodes represent differentially expressed metabolites with valid KEGG compound IDs. Edges (gray lines) represent known KEGG reaction pairs where metabolites are substrates and products in the same biochemical reaction. Isolated metabolites (no known reaction links) are removed for clarity.")
}
```

`r if (is.null(params$network_res) || !show_network) "-->"`

---

`r if (is.null(enrichment_res$mummichog) || !show_mummichog) "<!--"`

# Mummichog Pathway Analysis (m/z-based) {.tabset}

```{r mummichog-description, results='asis'}
mummi_cfg <- metab_cfg$enrichment$mummichog %||% list()
mummi_res <- enrichment_res$mummichog

# Organism labels
org_labels <- list(
  ame     = "Apis mellifera (honey bee)",
  eco     = "Escherichia coli K-12",
  ame_eco = "Combined: Apis mellifera + E. coli",
  hsa     = "Homo sapiens (human)",
  dme     = "Drosophila melanogaster (fruit fly)",
  cel     = "Caenorhabditis elegans (worm)",
  sce     = "Saccharomyces cerevisiae (yeast)"
)

organisms <- mummi_cfg$organisms
if (is.list(organisms)) organisms <- unlist(organisms)
org_display <- vapply(gsub("\\+", "_", organisms), function(o) {
  org_labels[[o]] %||% o
}, character(1))

p_cut   <- mummi_cfg$p_cutoff %||% 0.05
ppm     <- mummi_cfg$tolerance_ppm %||% 10
n_perm  <- mummi_cfg$n_permutations %||% 100

cat("## Methodology\n\n")
cat("[**Mummichog**](https://shuzhao-li.github.io/mummichog.org/) (Li et al., 2013) performs pathway enrichment analysis directly from mass spectrometry features (m/z values) without requiring prior metabolite identification. ")
cat("This approach leverages the fact that metabolic pathways are defined by sets of metabolites with known exact masses; by testing whether significant m/z features map to pathway members more often than expected by chance, mummichog bypasses the identification bottleneck of traditional approaches.\n\n")

cat("**Algorithm:**\n\n")
cat("1. All input m/z features are matched to candidate KEGG compounds using mass tolerance and common adduct patterns\n")
cat("2. Significant features (p < ", p_cut, ") are separated from the full reference list\n")
cat(sprintf("3. For each pathway, the overlap between significant features and pathway members is calculated\n"))
cat(sprintf("4. Empirical p-values are estimated by %d permutations of the significant feature list\n\n", n_perm))

cat("**Parameters:**\n\n")
cat(sprintf("- Significance cutoff: p < %s\n", p_cut))
cat(sprintf("- Mass tolerance: %d ppm\n", ppm))
cat(sprintf("- Permutations: %d\n\n", n_perm))

cat("**Reference organism models:**\n\n")
cat("Since no genome-scale metabolic model exists for the study organism, ")
cat("pathway analysis was performed using the closest available KEGG organism models as proxies:\n\n")
for (i in seq_along(organisms)) {
  org_clean <- gsub("\\+", "_", organisms[i])
  cat(sprintf("- **%s** (`%s`)\n", org_display[i], organisms[i]))
}
cat("\n")
if (any(grepl("\\+", organisms))) {
  cat("The combined model merges pathways from both organisms, prefixed by organism code, to enable cross-kingdom pathway analysis in a single run.\n\n")
}

cat("*Reference: Li S, Park Y, Duraisingham S, et al. Predicting network activity from high throughput metabolomics. PLoS Comput Biol. 2013;9(7):e1003123.*\n\n")
```

```{r mummi-dynamic-organisms, results='asis', fig.width=12, fig.height=8}
# KEGG organism code -> display name mapping
org_display <- c(
  hsa = "Human (*Homo sapiens*)", mmu = "Mouse (*Mus musculus*)",
  rno = "Rat (*Rattus norvegicus*)", dre = "Zebrafish (*Danio rerio*)",
  cel = "*C. elegans*", dme = "*Drosophila melanogaster*",
  sce = "*S. cerevisiae*", ame = "Honey Bee (*Apis mellifera*)",
  eco = "*E. coli*"
)

mummi_results <- enrichment_res$mummichog$results
mummi_orgs <- names(mummi_results)
mummi_orgs <- mummi_orgs[!vapply(mummi_results, is.null, logical(1))]

for (org in mummi_orgs) {
  org_label <- org_display[org] %||% org
  cat(sprintf("\n## %s {.tabset}\n\n", org_label))

  # Lollipop
  loll_key <- paste0("mummichog_", org, "_lollipop")
  if (!is.null(enrichment_res$plots[[loll_key]])) {
    cat("### Lollipop Plot\n\n")
    print(enrichment_res$plots[[loll_key]])
    cat(sprintf('\n<p class="figure-legend">Mummichog lollipop plot for %s KEGG metabolic model. Point size indicates the number of significant EmpiricalCompounds overlapping each pathway.</p>\n\n', org))
  }

  # Barplot
  bar_key <- paste0("mummichog_", org, "_barplot")
  if (!is.null(enrichment_res$plots[[bar_key]])) {
    cat("### Barplot\n\n")
    print(enrichment_res$plots[[bar_key]])
    cat(sprintf('\n<p class="figure-legend">Mummichog barplot for %s pathways ranked by significance.</p>\n\n', org))
  }

  # Table
  if (!is.null(mummi_results[[org]]$table)) {
    cat("### Results Table\n\n")
    dt <- make_dt(mummi_results[[org]]$table,
                  caption = sprintf("Mummichog Pathway Enrichment - %s", org))
    # Force DT widget to render in loop
    print(htmltools::tagList(dt))
    cat("\n\n")
  }
}

# Cross-organism heatmap (only if multiple organisms)
if (length(mummi_orgs) > 1 && !is.null(enrichment_res$plots$mummichog_organism_heatmap)) {
  cat("\n## Cross-Organism Comparison\n\n")
  print(enrichment_res$plots$mummichog_organism_heatmap)
  cat('\n<p class="figure-legend">Cross-organism comparison heatmap showing pathway enrichment significance across different metabolic models.</p>\n\n')
}
```

`r if (is.null(enrichment_res$mummichog) || !show_mummichog) "-->"`

---

`r if (!show_output_files) "<!--"`

# Output Files

```{r output-files}
files_tbl <- data.frame(
  File = c(
    "metabolomics_report.html",
    "de_summary_all.csv",
    "expression_matrix_normalized.csv"
  ),
  Description = c(
    "This HTML report with all QC, DE, and enrichment results",
    "Full differential expression summary table across all contrasts",
    "Normalized metabolite expression matrix used for downstream analysis"
  ),
  stringsAsFactors = FALSE
)

# Add enrichment output rows if enrichment was run
if (!is.null(enrichment_res)) {
  enrich_rows <- data.frame(
    File = character(0), Description = character(0), stringsAsFactors = FALSE
  )
  if (!is.null(enrichment_res$qea))
    enrich_rows <- rbind(enrich_rows, data.frame(
      File = "enrichment_qea_results.csv",
      Description = "QEA pathway enrichment results table",
      stringsAsFactors = FALSE))
  if (!is.null(enrichment_res$ssgsea))
    enrich_rows <- rbind(enrich_rows, data.frame(
      File = "enrichment_ssgsea_results.csv",
      Description = "ssGSEA pathway enrichment results and scores",
      stringsAsFactors = FALSE))
  if (!is.null(enrichment_res$ora))
    enrich_rows <- rbind(enrich_rows, data.frame(
      File = "enrichment_ora_results.csv",
      Description = "ORA pathway enrichment results (Fisher's exact test)",
      stringsAsFactors = FALSE))
  if (!is.null(enrichment_res$gsea))
    enrich_rows <- rbind(enrich_rows, data.frame(
      File = "enrichment_gsea_results.csv",
      Description = "GSEA pathway enrichment results (fgsea)",
      stringsAsFactors = FALSE))
  if (nrow(enrich_rows) > 0) files_tbl <- rbind(files_tbl, enrich_rows)
}

# Add feature selection outputs
if (!is.null(rf_res))
  files_tbl <- rbind(files_tbl, data.frame(
    File = "rf_importance.csv",
    Description = "Random Forest feature importance scores",
    stringsAsFactors = FALSE))
if (!is.null(plsda_res))
  files_tbl <- rbind(files_tbl, data.frame(
    File = "plsda_vip_scores.csv",
    Description = "PLS-DA VIP scores for all features",
    stringsAsFactors = FALSE))

# Add network outputs
if (!is.null(params$network_res)) {
  files_tbl <- rbind(files_tbl, data.frame(
    File = c("de_metabolite_network.html", "network_nodes.tsv", "network_edges.tsv"),
    Description = c(
      "Interactive DE metabolite network (KEGG reaction pairs)",
      "Network node table with KEGG IDs, fold changes, and p-values",
      "Network edge table (KEGG reaction pairs)"),
    stringsAsFactors = FALSE))
}

# Add mummichog outputs
if (!is.null(enrichment_res$mummichog)) {
  mummi_orgs <- names(enrichment_res$mummichog$results)
  mummi_orgs <- mummi_orgs[!vapply(enrichment_res$mummichog$results, is.null, logical(1))]
  for (org in mummi_orgs) {
    files_tbl <- rbind(files_tbl, data.frame(
      File = paste0("mummichog_", org, "_results.tsv"),
      Description = paste0("Mummichog pathway enrichment results (", org, " model)"),
      stringsAsFactors = FALSE))
  }
}

make_dt(files_tbl, "Key Output Files")
```

`r if (!show_output_files) "-->"`

---

`r if (!show_methods) "<!--"`

# Methods

```{r methods-text, results='asis'}
padj_cut_m <- metab_cfg$de$p_cutoff %||% 0.05
lfc_cut_m  <- metab_cfg$de$linear_fc_cutoff %||% 1.5

n_metab <- if (!is.null(pre)) nrow(pre$expr_work) else "N"
n_de_m  <- "N"
if (!is.null(de_res$summary_df) && "pass_any_contrast" %in% colnames(de_res$summary_df)) {
  n_de_m <- sum(de_res$summary_df$pass_any_contrast == 1, na.rm = TRUE)
}

cat("**Metabolomics Analysis Methods**\n\n")

cat("Metabolomics data were processed using the multiomics-core pipeline. ")

if (norm_method != "N/A") {
  cat(sprintf("Data were normalized using %s. ", norm_method))
}

cat(sprintf(
  "Differential expression analysis was performed using %s. ", de_method
))

cat(sprintf(
  "Metabolites were considered differentially expressed at adjusted p-value <= %.2f and |fold change| >= %.1f. ",
  padj_cut_m, lfc_cut_m
))

cat(sprintf(
  "A total of %s metabolites were quantified, with %s identified as differentially expressed in at least one contrast.",
  format(as.numeric(n_metab), big.mark = ","),
  format(as.numeric(n_de_m), big.mark = ",")
))

if (!is.null(enrichment_res)) {
  enrich_methods <- character(0)
  if (!is.null(enrichment_res$qea))    enrich_methods <- c(enrich_methods, "QEA")
  if (!is.null(enrichment_res$ssgsea)) enrich_methods <- c(enrich_methods, "ssGSEA")
  if (!is.null(enrichment_res$ora))    enrich_methods <- c(enrich_methods, "ORA")
  if (!is.null(enrichment_res$gsea))   enrich_methods <- c(enrich_methods, "GSEA")
  if (length(enrich_methods) > 0) {
    cat(sprintf(" Pathway enrichment analysis was performed using %s.",
                paste(enrich_methods, collapse = ", ")))
  }
  if (!is.null(enrichment_res$mummichog)) {
    mummi_cfg_m <- metab_cfg$enrichment$mummichog %||% list()
    orgs_m <- mummi_cfg_m$organisms
    if (is.list(orgs_m)) orgs_m <- unlist(orgs_m)
    cat(sprintf(
      " Additionally, mummichog (Li et al., 2013) m/z-based pathway analysis was performed using KEGG metabolic models for %s.",
      paste(orgs_m, collapse = ", ")))
  }
}

if (!is.null(rf_res) || !is.null(plsda_res)) {
  fs_methods <- character(0)
  if (!is.null(rf_res))    fs_methods <- c(fs_methods, "Random Forest")
  if (!is.null(plsda_res)) fs_methods <- c(fs_methods, "PLS-DA")
  cat(sprintf(" Feature selection was performed using %s.",
              paste(fs_methods, collapse = " and ")))
}

cat("\n\n")

cat("**Copy-ready version:**\n\n")
cat("<div style='background-color: #f5f5f5; padding: 15px; border-radius: 5px; font-size: 0.9em;'>\n")

cat(sprintf(
  "Metabolomics analysis was performed using the multiomics-core pipeline (v1.0). %sDifferential expression was assessed with %s (adjusted p <= %.2f, |FC| >= %.1f). A total of %s metabolites were quantified, with %s differentially expressed.",
  if (norm_method != "N/A") sprintf("Data were normalized using %s. ", norm_method) else "",
  de_method, padj_cut_m, lfc_cut_m,
  format(as.numeric(n_metab), big.mark = ","),
  format(as.numeric(n_de_m), big.mark = ",")
))

cat("\n</div>\n")
```

## Statistical Analysis

```{r methods-stats, results='asis'}
de_method_s <- de_method

test_desc <- switch(tolower(de_method_s),
  "t_test"       = "Welch's t-tests (unequal variance) were performed for each metabolite.",
  "t_test_equal" = "Student's t-tests (equal variance assumed, MetaboAnalyst convention) were performed for each metabolite.",
  "wilcoxon"     = "Wilcoxon rank-sum tests were used as a non-parametric alternative for differential expression testing.",
  "limma"        = "Moderated t-tests were performed using the *limma* R/Bioconductor package, which applies empirical Bayes shrinkage of variance estimates to improve statistical power.",
  sprintf("Differential expression was assessed using the *%s* method.", de_method_s)
)

cat(test_desc, "\n\n")

cat(sprintf(
  "Multiple testing correction was applied using the Benjamini-Hochberg procedure to control the false discovery rate (FDR). Metabolites were considered differentially expressed at adjusted p-value $\\leq$ %s and |linear fold change| $\\geq$ %s.\n\n",
  padj_cut_m, lfc_cut_m
))

cat(sprintf(
  "All analyses were performed in R (version %s) using the multiomics-core pipeline.\n\n",
  paste(R.version$major, R.version$minor, sep = ".")
))
```

`r if (!show_methods) "-->"`

---

`r if (!show_session_info) "<!--"`

# Session Info

<details>
<summary>Click to expand</summary>

```{r session-info}
sessionInfo()
```

</details>

`r if (!show_session_info) "-->"`

---

<p style="text-align:center; color:grey; font-size:0.9em;">
Report generated by multiomics-core pipeline | `r format(Sys.time(), '%Y-%m-%d %H:%M')`
</p>

```{r dt-tab-fix, results='asis', echo=FALSE}
# Fix DT tables inside hidden tabs: redraw when tab becomes visible
cat('
<script>
$(document).on("shown.bs.tab", function(e) {
  $($(e.target).attr("href")).find(".datatables.html-widget").each(function() {
    var widget = HTMLWidgets.find("#" + this.id);
    if (widget && widget.table) {
      widget.table.columns.adjust().draw(false);
    }
  });
});
$(document).ready(function() {
  setTimeout(function() {
    $(".datatables.html-widget").each(function() {
      var widget = HTMLWidgets.find("#" + this.id);
      if (widget && widget.table) {
        widget.table.columns.adjust().draw(false);
      }
    });
  }, 500);
});
</script>
')
```
