---
title: "`r local({ cfg <- params$config; pname <- cfg$project$name %||% ''; analyst <- gsub('_', ' ', cfg$project$analyst %||% ''); parts <- c('Metabolomics Analysis Report'); if (nzchar(pname)) parts <- c(parts, pname); if (nzchar(analyst)) parts <- c(parts, analyst); paste(parts, collapse = ' - ') })`"
subtitle: "multiomics-core Pipeline"
author: "`r paste0('Bioinformatician: ', gsub('_', ' ', params$config$project$analyst %||% ''))`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    fig_width: 10
    fig_height: 7
    self_contained: true
    code_folding: hide
    df_print: paged
params:
  pre: NULL
  qc_res: NULL
  de_res: NULL
  rf_res: NULL
  plsda_res: NULL
  enrichment_res: NULL
  config: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%"
)

library(ggplot2)
library(DT)
library(htmltools)

if (requireNamespace("plotly", quietly = TRUE))   library(plotly)
if (requireNamespace("crosstalk", quietly = TRUE)) library(crosstalk)
if (requireNamespace("jsonlite", quietly = TRUE))  library(jsonlite)

# Source interactive cutoff panel helpers
tryCatch({
  .cp_dir <- normalizePath(dirname(knitr::current_input()), mustWork = TRUE)
  .cp_found <- FALSE
  for (.cp_i in 1:6) {
    .cp_try <- file.path(.cp_dir, "R", "core", "17_cutoff_panel.R")
    if (file.exists(.cp_try)) { source(.cp_try); .cp_found <- TRUE; break }
    .cp_dir <- dirname(.cp_dir)
  }
  if (!.cp_found) message("cutoff_panel helpers not found (walked 6 levels up)")
}, error = function(e) message("cutoff_panel helpers not found: ", e$message))

# ---------------------------------------------------------------------------
# Helper: downloadable DT table with CSV/Excel export (proteomics style)
# ---------------------------------------------------------------------------
make_dt <- function(df, caption = NULL, pageLength = 15) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  safe_caption <- gsub("[^a-zA-Z0-9]", "_", caption %||% "data")
  tbl <- DT::datatable(
    df,
    caption  = caption,
    extensions = "Buttons",
    options = list(
      scrollX    = TRUE,
      pageLength = pageLength,
      dom        = "Bfrtip",
      buttons    = list(
        list(extend = "csv",   filename = safe_caption),
        list(extend = "excel", filename = safe_caption)
      )
    ),
    rownames = FALSE,
    class    = "compact stripe hover"
  )
  # Format numeric columns: scientific notation for p-values, 2 decimals for the rest
  num_cols <- names(df)[vapply(df, is.numeric, logical(1))]
  pval_pattern <- "^(p[._]?val|padj|adj[._]?p|fdr|q[._]?val|BH|bonferroni|p\\.value|adj\\.P\\.Val|PValue|FDR)"
  pval_cols <- grep(pval_pattern, num_cols, ignore.case = TRUE, value = TRUE)
  other_cols <- setdiff(num_cols, pval_cols)
  if (length(other_cols) > 0) {
    tbl <- DT::formatRound(tbl, columns = other_cols, digits = 2)
  }
  if (length(pval_cols) > 0) {
    pval_idx <- which(names(df) %in% pval_cols) - 1L
    sci_render <- DT::JS(
      "function(data, type, row) {",
      "  if (type === 'display' && data !== null && data !== '') {",
      "    return parseFloat(data).toExponential(2);",
      "  }",
      "  return data;",
      "}"
    )
    col_defs <- lapply(pval_idx, function(i) list(targets = i, render = sci_render))
    tbl$x$options$columnDefs <- c(tbl$x$options$columnDefs, col_defs)
  }
  htmltools::tagList(tbl)
}

# ---------------------------------------------------------------------------
# Helper: paper-style figure legend below plots (matching proteomics)
# ---------------------------------------------------------------------------
figure_legend <- function(text) {
  cat(sprintf("<p class='figure-legend' style='text-align: center; font-size: 0.9em; color: #555; margin: 4px auto 12px auto; max-width: 88%%; line-height: 1.45; font-style: italic;'>%s</p>\n", text))
}

# ---------------------------------------------------------------------------
# Helper: safely print a ggplot or return invisible(NULL)
# ---------------------------------------------------------------------------
safe_print_plot <- function(p) {
  if (inherits(p, "gg") || inherits(p, "ggplot")) {
    print(p)
  } else {
    invisible(NULL)
  }
}

# ---------------------------------------------------------------------------
# Helper: embed a PNG from a file path vector by basename pattern
# ---------------------------------------------------------------------------
embed_png <- function(files, pattern, fig_width = "100%") {
  if (is.null(files) || length(files) == 0) return(invisible(NULL))
  hits <- files[grepl(pattern, basename(files), ignore.case = TRUE)]
  hits <- hits[file.exists(hits)]
  if (length(hits) == 0) return(invisible(NULL))
  knitr::include_graphics(hits[1])
}

# ---------------------------------------------------------------------------
# Unpack config and module results
# ---------------------------------------------------------------------------
cfg       <- params$config
metab_cfg <- cfg$modes$metabolomics
```

```{r force-dependencies, include=FALSE}
# Force DT and Plotly JS dependencies to be included in the self-contained HTML
DT::datatable(data.frame(x = 1))
if (requireNamespace("plotly", quietly = TRUE)) {
  plotly::plot_ly(data.frame(x = 1, y = 1), x = ~x, y = ~y,
                  type = "scatter", mode = "markers")
}
```

```{r unpack-results, include=FALSE}
pre            <- params$pre
qc_res         <- params$qc_res
de_res         <- params$de_res
rf_res         <- params$rf_res
plsda_res      <- params$plsda_res
enrichment_res <- params$enrichment_res

n_features <- if (!is.null(pre)) nrow(pre$expr_work) else NA
n_samples  <- if (!is.null(pre)) ncol(pre$expr_work) else NA

# Resolve group / condition column
cond_col <- metab_cfg$de$condition_column %||%
            metab_cfg$effects$color %||% "sample_type"
n_groups <- NA
if (!is.null(pre$meta) && cond_col %in% colnames(pre$meta)) {
  n_groups <- length(unique(pre$meta[[cond_col]]))
}

# Normalization summary
norm_method <- "N/A"
if (!is.null(pre$info$normalization)) {
  ni    <- pre$info$normalization
  parts <- character(0)
  if (!is.null(ni$sample_norm) && tolower(ni$sample_norm) != "none")
    parts <- c(parts, toupper(ni$sample_norm))
  if (!is.null(ni$transform) && tolower(ni$transform) != "none")
    parts <- c(parts, ni$transform)
  if (!is.null(ni$scaling) && tolower(ni$scaling) != "none")
    parts <- c(parts, ni$scaling)
  if (length(parts) > 0) norm_method <- paste(parts, collapse = " + ")
}

de_method <- if (!is.null(de_res$method)) de_res$method else "N/A"

# DE contrasts
de_contrasts <- NULL
if (!is.null(de_res) && !is.null(de_res$summary_df)) {
  de_cfg       <- metab_cfg$de %||% list()
  de_contrasts <- de_cfg$contrasts
  if (is.list(de_contrasts)) de_contrasts <- unlist(de_contrasts)
}
```

# Executive Summary

```{r executive-summary, results='asis'}
# ---- Gradient panel (matching proteomics style) --------------------------
cat("<div style='background: linear-gradient(135deg, #f8f9fa 0%, #e9ecef 100%); border: 1px solid #dee2e6; border-radius: 8px; padding: 20px 24px; margin: 10px 0 20px 0;'>\n")
cat("<div style='display: flex; flex-wrap: wrap; gap: 16px; margin-bottom: 16px;'>\n")

# Metric card helper
metric_card <- function(value, label, color = "#2c3e50") {
  cat(sprintf(
    "<div style='flex: 1; min-width: 120px; background: white; border-radius: 6px; padding: 12px 16px; text-align: center; box-shadow: 0 1px 3px rgba(0,0,0,0.08);'><div style='font-size: 1.8em; font-weight: bold; color: %s;'>%s</div><div style='font-size: 0.85em; color: #666;'>%s</div></div>\n",
    color, as.character(value), label
  ))
}

# Samples
if (!is.na(n_samples)) metric_card(n_samples, "Samples")

# Groups
if (!is.na(n_groups)) metric_card(n_groups, "Groups")

# Metabolites
if (!is.na(n_features)) metric_card(format(n_features, big.mark = ","), "Metabolites")

# Contrasts
n_contrasts <- length(de_contrasts)
if (n_contrasts > 0) metric_card(n_contrasts, "Contrasts")

# Total DE metabolites
n_de_total <- NA
if (!is.null(de_res$summary_df) && "pass_any_contrast" %in% colnames(de_res$summary_df)) {
  n_de_total <- sum(de_res$summary_df$pass_any_contrast == 1, na.rm = TRUE)
  metric_card(format(n_de_total, big.mark = ","), "DE Metabolites", color = "#c0392b")
}

cat("</div>\n")

# ---- DE breakdown per contrast -------------------------------------------
if (!is.null(de_contrasts) && length(de_contrasts) > 0 && !is.null(de_res$summary_df)) {
  padj_cut <- metab_cfg$de$p_cutoff %||% 0.05
  if (!is.null(metab_cfg$de$logfc_cutoff)) {
    log2_lfc <- metab_cfg$de$logfc_cutoff
  } else {
    lfc_cut  <- metab_cfg$de$linear_fc_cutoff %||% 1.5
    log2_lfc <- log2(lfc_cut)
  }

  cat("<div style='margin: 12px 0;'>\n")
  cat("<strong>Differential Expression per Contrast:</strong>\n<ul style='margin: 6px 0; padding-left: 20px;'>\n")

  for (ctr in de_contrasts) {
    ctr_label <- make_contrast_label(ctr)
    ctr_tbl   <- tryCatch(extract_contrast_table(de_res$summary_df, ctr_label),
                           error = function(e) NULL)
    if (!is.null(ctr_tbl)) {
      sig   <- !is.na(ctr_tbl$P.Value) & ctr_tbl$P.Value <= padj_cut &
               !is.na(ctr_tbl$logFC) & abs(ctr_tbl$logFC) >= log2_lfc
      n_up   <- sum(sig & ctr_tbl$logFC > 0, na.rm = TRUE)
      n_down <- sum(sig & ctr_tbl$logFC < 0, na.rm = TRUE)
      if (exists("cutoff_summary_span")) {
        cat(sprintf("<li><strong>%s</strong>: %s</li>\n",
                    ctr, cutoff_summary_span(ctr, n_up, n_down, "metabolites")))
      } else {
        cat(sprintf("<li><strong>%s</strong>: %d DE metabolites (%d up, %d down)</li>\n",
                    ctr, n_up + n_down, n_up, n_down))
      }
    }
  }
  cat("</ul>\n</div>\n")
}

cat("</div>\n")
```

---

# Analysis Parameters {.tabset}

## Project

```{r params-project}
proj <- data.frame(
  Parameter = c("Project Name", "Analyst", "Project Directory",
                "Raw Data Path", "Output Path"),
  Value = c(
    cfg$project$name %||% "N/A",
    gsub("_", " ", cfg$project$analyst %||% "N/A"),
    cfg$project$dir %||% "N/A",
    cfg$paths$raw %||% "N/A",
    cfg$paths$out %||% "N/A"
  ),
  stringsAsFactors = FALSE
)
make_dt(proj, "Project Parameters")
```

## Experimental Design

```{r params-design}
if (!is.null(pre$meta) && cond_col %in% colnames(pre$meta)) {
  design_tbl <- as.data.frame(table(pre$meta[[cond_col]]))
  colnames(design_tbl) <- c("Group", "n (replicates)")
  make_dt(design_tbl, "Experimental Design")
}
```

## Pipeline Settings

```{r params-pipeline}
settings <- data.frame(
  Parameter = c(
    "Normalization",
    "DE Method",
    "Adjusted p-value Cutoff",
    "Linear FC Cutoff",
    "Random Forest",
    "PLS-DA",
    "Enrichment"
  ),
  Value = c(
    norm_method,
    de_method,
    as.character(metab_cfg$de$p_cutoff %||% 0.05),
    as.character(metab_cfg$de$linear_fc_cutoff %||% 1.5),
    if (!is.null(rf_res)) "Enabled" else "Disabled",
    if (!is.null(plsda_res)) "Enabled" else "Disabled",
    if (!is.null(enrichment_res)) "Enabled" else "Disabled"
  ),
  stringsAsFactors = FALSE
)
make_dt(settings, "Pipeline Settings")
```

## Contrasts

```{r params-contrasts}
if (!is.null(de_contrasts) && length(de_contrasts) > 0) {
  ctr_df <- data.frame(
    `#` = seq_along(de_contrasts),
    Contrast = de_contrasts,
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
  make_dt(ctr_df, "Contrasts")
} else {
  cat("No contrasts defined.\n")
}
```

---

# Module Status

```{r module-status}
module_status <- data.frame(
  Module = c("Preprocessing", "QC Diagnostics", "Differential Expression",
             "Random Forest", "PLS-DA", "Enrichment"),
  Status = c(
    if (!is.null(pre)) "Completed" else "Skipped",
    if (!is.null(qc_res)) "Completed" else "Skipped",
    if (!is.null(de_res) && !is.null(de_res$summary_df)) "Completed" else "Skipped",
    if (!is.null(rf_res)) "Completed" else "Skipped",
    if (!is.null(plsda_res)) "Completed" else "Skipped",
    if (!is.null(enrichment_res)) "Completed" else "Skipped"
  ),
  stringsAsFactors = FALSE
)
make_dt(module_status, "Module Status", pageLength = 10)
```

---

`r if (is.null(qc_res)) "<!--"`

# QC Diagnostics {.tabset}

## PCA {.tabset}

### Static

```{r qc-pca, fig.width=9, fig.height=7, results='asis'}
if (!is.null(qc_res$plots)) {
  # Find the primary PCA plot (no _noQC, no extra color suffix)
  pca_keys <- grep("^pca_1_2", names(qc_res$plots), value = TRUE)
  pca_main <- pca_keys[!grepl("_noQC$", pca_keys)][1]
  if (!is.na(pca_main)) {
    safe_print_plot(qc_res$plots[[pca_main]])
    figure_legend("Principal Component Analysis (PCA) of normalized metabolomics data, showing sample separation along PC1 and PC2. Each point represents one sample, colored by experimental group.")
  }

  pca13_keys <- grep("^pca_1_3", names(qc_res$plots), value = TRUE)
  pca13_main <- pca13_keys[!grepl("_noQC$", pca13_keys)][1]
  if (!is.na(pca13_main)) {
    safe_print_plot(qc_res$plots[[pca13_main]])
    figure_legend("PCA projection onto PC1 vs PC3, capturing additional variance beyond the first two components.")
  }
}
```

### With Sample Names

```{r qc-pca-labeled, fig.width=9, fig.height=7, results='asis'}
if (!is.null(qc_res$plots)) {
  pca_keys <- grep("^pca_1_2", names(qc_res$plots), value = TRUE)
  pca_main <- pca_keys[!grepl("_noQC$", pca_keys)][1]
  if (!is.na(pca_main)) {
    p <- qc_res$plots[[pca_main]]
    if ("sample" %in% colnames(p$data)) {
      p_lab <- p + ggplot2::geom_text(ggplot2::aes(label = sample), size = 2.5, vjust = -0.8, show.legend = FALSE)
      safe_print_plot(p_lab)
    }
  }
  pca13_keys <- grep("^pca_1_3", names(qc_res$plots), value = TRUE)
  pca13_main <- pca13_keys[!grepl("_noQC$", pca13_keys)][1]
  if (!is.na(pca13_main)) {
    p13 <- qc_res$plots[[pca13_main]]
    if ("sample" %in% colnames(p13$data)) {
      p13_lab <- p13 + ggplot2::geom_text(ggplot2::aes(label = sample), size = 2.5, vjust = -0.8, show.legend = FALSE)
      safe_print_plot(p13_lab)
    }
  }
}
```

### Interactive

```{r qc-pca-interactive, fig.width=9, fig.height=7}
if (!is.null(qc_res$plots) && requireNamespace("plotly", quietly = TRUE)) {
  pca_keys <- grep("^pca_1_2", names(qc_res$plots), value = TRUE)
  pca_main <- pca_keys[!grepl("_noQC$", pca_keys)][1]
  if (!is.na(pca_main)) {
    p <- qc_res$plots[[pca_main]]
    sc <- p$data
    if ("sample" %in% colnames(sc)) {
      color_var <- setdiff(colnames(sc), c("sample", grep("^PC", colnames(sc), value = TRUE)))[1]
      sc$hover_text <- paste0("Sample: ", sc$sample, if (!is.na(color_var)) paste0("<br>", color_var, ": ", sc[[color_var]]) else "")
      pl <- plotly::plot_ly(sc, x = ~PC1, y = ~PC2,
                            color = if (!is.na(color_var)) as.formula(paste0("~`", color_var, "`")) else NULL,
                            text = ~hover_text, hoverinfo = "text",
                            type = "scatter", mode = "markers",
                            marker = list(size = 10)) |>
          plotly::layout(title = "PCA: PC1 vs PC2 — Hover for sample names")
      print(pl)
    }
  }
} else {
  cat("*Interactive PCA not available (plotly package missing).*\n")
}
```

## Density

```{r qc-density, fig.width=9, fig.height=6, results='asis'}
if (!is.null(qc_res$plots)) {
  if (!is.null(qc_res$plots[["density_raw"]])) {
    safe_print_plot(qc_res$plots[["density_raw"]])
    figure_legend("Per-sample density of log2(raw intensities) before any normalization. Well-overlapping curves indicate consistent sample loading and acquisition.")
  }
  ni_dens <- pre$info$normalization
  has_scaling_dens <- !is.null(ni_dens$scaling) && tolower(ni_dens$scaling) != "none"
  if (has_scaling_dens && !is.null(qc_res$plots[["density_transformed"]])) {
    safe_print_plot(qc_res$plots[["density_transformed"]])
    figure_legend("Per-sample density after the transform step. This is the matrix used for differential expression testing. Well-overlapping curves confirm the transform brought samples onto a comparable scale.")
  }
  if (!is.null(qc_res$plots[["density_norm"]])) {
    safe_print_plot(qc_res$plots[["density_norm"]])
    if (has_scaling_dens) {
      figure_legend(sprintf(
        "Per-sample density after full normalization (%s + %s). %s scaling standardizes each metabolite independently (mean=0, SD=1), so divergent per-sample curves reflect biological differences between groups. This matrix is used for PCA, heatmaps, and multivariate analyses (RF, PLS-DA).",
        ni_dens$transform %||% "", ni_dens$scaling %||% "",
        tools::toTitleCase(ni_dens$scaling %||% "")
      ))
    } else {
      figure_legend("Per-sample density after normalization. Well-aligned curves confirm effective normalization across samples.")
    }
  }
}
```

## Boxplots

```{r qc-boxplots, fig.width=10, fig.height=6, results='asis'}
if (!is.null(qc_res$plots)) {
  if (!is.null(qc_res$plots[["boxplot_raw"]])) {
    safe_print_plot(qc_res$plots[["boxplot_raw"]])
    figure_legend("Per-sample boxplots of raw metabolite intensities before normalization.")
  }
  # Show transform-only boxplot when scaling is applied; otherwise show
  # the fully normalized boxplot (they are identical when no scaling).
  ni_box <- pre$info$normalization
  has_scaling_box <- !is.null(ni_box$scaling) && tolower(ni_box$scaling) != "none"
  if (has_scaling_box && !is.null(qc_res$plots[["boxplot_transformed"]])) {
    safe_print_plot(qc_res$plots[["boxplot_transformed"]])
    figure_legend("Per-sample boxplots after the transform step. This is the matrix used for differential expression testing. Aligned medians confirm the transform effectively standardized sample distributions.")
  }
  if (!is.null(qc_res$plots[["boxplot"]])) {
    safe_print_plot(qc_res$plots[["boxplot"]])
    if (has_scaling_box) {
      figure_legend(sprintf(
        "Per-sample boxplots after full normalization (%s + %s). %s scaling standardizes each metabolite independently (mean=0, SD=1), so shifted medians between groups reflect biological differences. This matrix is used for PCA, heatmaps, and multivariate analyses (RF, PLS-DA).",
        ni_box$transform %||% "", ni_box$scaling %||% "",
        tools::toTitleCase(ni_box$scaling %||% "")
      ))
    } else {
      figure_legend("Per-sample boxplots after normalization. Comparable medians and spreads indicate successful normalization.")
    }
  }
}
```

## Heatmaps

```{r qc-heatmaps, out.width="100%", results='asis'}
if (!is.null(qc_res$files)) {
  # Find heatmap PNGs from file list
  dist_hits <- qc_res$files[grepl("sample_distance_heatmap\\.png$", basename(qc_res$files), ignore.case = TRUE)]
  dist_hits <- dist_hits[file.exists(dist_hits)]
  if (length(dist_hits) > 0) {
    cat(sprintf("![Sample Distance Heatmap](%s){width=100%%}\n\n", dist_hits[1]))
    figure_legend("Sample distance heatmap based on Euclidean distances between normalized metabolite profiles. Darker colors indicate more similar samples.")
  }

  cor_hits <- qc_res$files[grepl("sample_correlation_heatmap\\.png$", basename(qc_res$files), ignore.case = TRUE)]
  cor_hits <- cor_hits[file.exists(cor_hits)]
  if (length(cor_hits) > 0) {
    cat(sprintf("\n![Sample Correlation Heatmap](%s){width=100%%}\n\n", cor_hits[1]))
    figure_legend("Pearson correlation heatmap across all samples. High within-group correlation (warm colors) indicates good replicate consistency.")
  }
}
```

`r if (is.null(qc_res)) "-->"`

---

`r if (is.null(de_res) || is.null(de_res$summary_df)) "<!--"`

# Differential Expression

```{r cutoff-panel, results='asis'}
if (exists("cutoff_panel_html")) {
    .lfc_default <- if (!is.null(metab_cfg$de$logfc_cutoff)) {
        metab_cfg$de$logfc_cutoff
    } else {
        log2(metab_cfg$de$linear_fc_cutoff %||% 1.5)
    }
    .p_default <- metab_cfg$de$p_cutoff %||% 0.05
    cat(cutoff_panel_html(.lfc_default, .p_default, default_ptype = "pval"))
}
```

```{r de-prep, include=FALSE}
# Prepare shared data used across DE sub-sections
padj_cut_de <- metab_cfg$de$p_cutoff %||% 0.05
# logfc_cutoff is applied directly to |logFC| (in the transform's scale).
# linear_fc_cutoff is a backwards-compat alias converted via log2().
if (!is.null(metab_cfg$de$logfc_cutoff)) {
  log2_lfc_de <- metab_cfg$de$logfc_cutoff
} else {
  lfc_cut_de  <- metab_cfg$de$linear_fc_cutoff %||% 1.5
  log2_lfc_de <- log2(lfc_cut_de)
}

# Build feature_id -> Name lookup
name_map_v <- NULL
if (!is.null(pre$row_data) && "Name" %in% colnames(pre$row_data)) {
  name_map_v <- stats::setNames(
    as.character(pre$row_data$Name),
    as.character(pre$row_data$feature_id)
  )
}

resolve_name <- function(fid) {
  if (!is.null(name_map_v) && fid %in% names(name_map_v) && !is.na(name_map_v[fid])) {
    name_map_v[fid]
  } else {
    fid
  }
}

# Build norm intensity y-label
norm_y_label_report <- "Normalized intensity"
ni_r <- pre$info$normalization
if (!is.null(ni_r)) {
  np <- character(0)
  if (!is.null(ni_r$transform) && tolower(ni_r$transform) != "none") np <- c(np, ni_r$transform)
  if (!is.null(ni_r$sample_norm) && tolower(ni_r$sample_norm) != "none") np <- c(np, toupper(ni_r$sample_norm))
  if (!is.null(ni_r$scaling) && tolower(ni_r$scaling) != "none") np <- c(np, ni_r$scaling)
  if (length(np) > 0) norm_y_label_report <- paste0("Normalized intensity (", paste(np, collapse = " + "), ")")
}
```

```{r de-note, results='asis'}
if (!is.null(de_res) && !is.null(de_res$summary_df)) {
  n_sig_any <- sum(de_res$summary_df$pass_any_contrast == 1, na.rm = TRUE)
  if (n_sig_any == 0) {
    n_samp <- ncol(pre$expr_work)
    cat("\n> **Note:** No features reached significance (p-value < ",
        padj_cut_de, " and |logFC| >= ", round(log2_lfc_de, 2),
        "). With only **", n_samp, " samples**, the test has limited power. ",
        "Consider increasing sample size, relaxing thresholds, or using a ",
        "rank-based method (e.g. GSEA).\n\n", sep = "")
  }
}
```

```{r de-sections, results='asis', fig.width=10, fig.height=7}
if (!is.null(de_contrasts)) {
  for (ctr in de_contrasts) {
    ctr_label <- make_contrast_label(ctr)

    ctr_tbl <- tryCatch(
      extract_contrast_table(de_res$summary_df, ctr_label),
      error = function(e) NULL
    )
    if (is.null(ctr_tbl)) next

    # Add Name column
    ctr_tbl$Name <- vapply(ctr_tbl$feature_id, resolve_name, character(1), USE.NAMES = FALSE)

    # Direction column for coloring (uses raw P.Value)
    ctr_tbl$Direction <- ifelse(
      !is.na(ctr_tbl$P.Value) & ctr_tbl$P.Value <= padj_cut_de &
        !is.na(ctr_tbl$logFC) & abs(ctr_tbl$logFC) >= log2_lfc_de,
      ifelse(ctr_tbl$logFC > 0, "Up", "Down"), "NS"
    )
    ctr_tbl$Direction <- factor(ctr_tbl$Direction, levels = c("NS", "Down", "Up"))
    ctr_tbl$neglog10_pval <- pmin(-log10(pmax(
      ifelse(is.na(ctr_tbl$P.Value), 1, ctr_tbl$P.Value), 1e-300
    )), 50)

    # ================================================================
    # VOLCANO
    # ================================================================
    cat("\n## Volcano: ", ctr, " {.tabset}\n\n", sep = "")

    # -- Static Volcano --
    vol_png <- de_res$files[grepl(paste0("volcano_", ctr_label, "\\.png$"), basename(de_res$files))]
    vol_png <- vol_png[file.exists(vol_png)]
    if (length(vol_png) > 0) {
      cat("### Static\n\n")
      cat(sprintf("\n![Volcano: %s](%s){width=100%%}\n\n", ctr, vol_png[1]))
      figure_legend(paste0("Volcano plot for contrast ", ctr, ". Each point is a metabolite; x-axis shows log2(fold change) and y-axis shows -log10(p-value). Colored points exceed both fold-change and significance thresholds."))
      cat("\n\n")
    }

    # -- Interactive Volcano --
    if (requireNamespace("plotly", quietly = TRUE)) {
      cat("### Interactive\n\n")

      # Unique ID for this contrast's JS namespace
      vol_cid <- gsub("[^a-zA-Z0-9]", "", ctr_label)

      vol_p <- ggplot2::ggplot(
        ctr_tbl,
        ggplot2::aes(x = logFC, y = neglog10_pval, color = Direction,
                     text = paste0("Metabolite: ", Name,
                                   "<br>log2(FC): ", round(logFC, 2),
                                   "<br>P.Value: ", sprintf("%.2e", P.Value),
                                   "<br>adj.P.Val: ", sprintf("%.2e", adj.P.Val)))
      ) +
        ggplot2::geom_point(alpha = 0.6, size = 1.5) +
        ggplot2::scale_color_manual(values = c("NS" = "grey70", "Down" = "blue", "Up" = "red")) +
        ggplot2::geom_vline(xintercept = c(-log2_lfc_de, log2_lfc_de), linetype = "dashed", alpha = 0.5) +
        ggplot2::geom_hline(yintercept = -log10(padj_cut_de), linetype = "dashed", alpha = 0.5) +
        ggplot2::labs(x = "log2(FC)", y = "-log10(P.Value)") +
        ggplot2::theme_minimal()

      vol_plotly <- plotly::ggplotly(vol_p, tooltip = "text")

      # Emit search box + highlight JS (matching RNA/prot pattern)
      cat(sprintf('
<div style="margin-bottom: 15px; position: relative;">
<label for="volcano_genes_%s"><strong>Highlight metabolites:</strong></label>
<div style="display: inline-block; position: relative; margin-left: 10px;">
<input type="text" id="volcano_genes_%s" placeholder="Start typing a metabolite name..."
  autocomplete="off"
  style="padding: 8px; border-radius: 4px; border: 1px solid #ccc; width: 350px;"
  oninput="volcanoAutocomplete_%s(this)" onkeydown="volcanoAutoKey_%s(event)">
<div id="volcano_ac_%s" style="position: absolute; top: 100%%; left: 0; width: 350px; max-height: 200px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 4px 4px; z-index: 1000; display: none; box-shadow: 0 4px 6px rgba(0,0,0,0.1);"></div>
</div>
<button onclick="highlightVolcanoGenes_%s()" style="padding: 8px 15px; margin-left: 5px; background: #007bff; color: white; border: none; border-radius: 4px; cursor: pointer;">Highlight</button>
<button onclick="clearVolcanoHighlight_%s()" style="padding: 8px 15px; margin-left: 5px; background: #6c757d; color: white; border: none; border-radius: 4px; cursor: pointer;">Clear</button>
<div id="volcano_tags_%s" style="margin-top: 5px; display: flex; flex-wrap: wrap; gap: 4px;"></div>
</div>
', vol_cid, vol_cid, vol_cid, vol_cid, vol_cid, vol_cid, vol_cid, vol_cid))

      cat(sprintf('<div id="volcano_plot_%s">\n', vol_cid))
      print(htmltools::tagList(vol_plotly))
      cat('</div>\n')

      # Register volcano plot for cutoff panel
      if (exists("cutoff_register_plot")) {
        vol_reg_df <- data.frame(
          name = as.character(ctr_tbl$Name),
          logFC = ctr_tbl$logFC,
          pval = ctr_tbl$P.Value,
          adjPval = ctr_tbl$adj.P.Val,
          stringsAsFactors = FALSE
        )
        # The plotly widget is inside the wrapper div; JS uses querySelector fallback
        cat(cutoff_register_plot(
          paste0("volcano_plot_", vol_cid),
          vol_reg_df, "volcano", "Metabolite", ctr
        ))
      }

      # JS for autocomplete + highlight
      cat(sprintf('<script>
(function() {
  var cid = "%s";
  var selectedGenes = [];
  var acIndex = -1;
  function getMetaboliteNames() {
    var plotDiv = document.getElementById("volcano_plot_" + cid);
    if (!plotDiv) plotDiv = document.querySelector("#volcano_plot_" + cid + " .plotly");
    var el = plotDiv ? plotDiv.querySelector(".js-plotly-plot") || plotDiv : null;
    if (!el || !el.data) return [];
    var names = [];
    for (var t = 0; t < el.data.length; t++) {
      var d = el.data[t];
      if (!d.text) continue;
      for (var i = 0; i < d.text.length; i++) {
        var txt = typeof d.text[i] === "string" ? d.text[i] : "";
        var m = txt.match(/Metabolite:\\s*(.+?)(?:<br>|$)/);
        if (m && m[1] && names.indexOf(m[1]) === -1) names.push(m[1]);
      }
    }
    return names.sort();
  }
  function renderTags() {
    var c = document.getElementById("volcano_tags_" + cid);
    c.innerHTML = "";
    selectedGenes.forEach(function(g) {
      var tag = document.createElement("span");
      tag.style.cssText = "display:inline-flex;align-items:center;background:#007bff;color:white;padding:2px 8px;border-radius:12px;font-size:12px;gap:4px;";
      var txt = document.createTextNode(g + " ");
      tag.appendChild(txt);
      var btn = document.createElement("span");
      btn.textContent = String.fromCharCode(215);
      btn.style.cssText = "cursor:pointer;font-weight:bold;";
      btn.addEventListener("click", (function(gene){return function(){window["removeVolcanoGene_"+cid](gene);};})(g));
      tag.appendChild(btn);
      c.appendChild(tag);
    });
  }
  function applyHighlight() {
    var plotDiv = document.getElementById("volcano_plot_" + cid);
    var el = plotDiv ? plotDiv.querySelector(".js-plotly-plot") || plotDiv : null;
    if (!el || !el.data) return;
    if (selectedGenes.length === 0) { Plotly.relayout(el, {annotations:[]}); return; }
    var gl = selectedGenes.map(function(g){return g.toLowerCase();});
    var ann = [];
    for (var t = 0; t < el.data.length; t++) {
      var d = el.data[t];
      if (!d.text) continue;
      for (var i = 0; i < d.text.length; i++) {
        var txt = typeof d.text[i] === "string" ? d.text[i] : "";
        var m2 = txt.match(/Metabolite:\\s*(.+?)(?:<br>|$)/);
        if (m2 && gl.indexOf(m2[1].toLowerCase()) !== -1) {
          ann.push({x:d.x[i],y:d.y[i],text:m2[1],showarrow:true,arrowhead:2,ax:20,ay:-30,
            font:{size:11,color:"black"},bgcolor:"rgba(255,255,255,0.9)",bordercolor:"black",borderwidth:1});
        }
      }
    }
    Plotly.relayout(el, {annotations:ann});
  }
  window["volcanoAutocomplete_" + cid] = function(el) {
    var ac = document.getElementById("volcano_ac_" + cid);
    var val = el.value.trim().toLowerCase(); ac.innerHTML = ""; acIndex = -1;
    if (val.length < 1) {ac.style.display="none"; return;}
    var all = getMetaboliteNames();
    var matches = all.filter(function(g){return g.toLowerCase().indexOf(val)!==-1 && selectedGenes.indexOf(g)===-1;}).slice(0,15);
    if (matches.length===0){ac.style.display="none"; return;}
    matches.forEach(function(g){
      var item=document.createElement("div"); item.textContent=g;
      item.style.cssText="padding:6px 10px;cursor:pointer;font-size:13px;";
      item.onmouseenter=function(){this.style.background="#e9ecef";};
      item.onmouseleave=function(){this.style.background="white";};
      item.onclick=function(){selectedGenes.push(g);el.value="";ac.style.display="none";renderTags();applyHighlight();};
      ac.appendChild(item);
    });
    ac.style.display="block";
  };
  window["volcanoAutoKey_" + cid] = function(e) {
    var ac = document.getElementById("volcano_ac_" + cid);
    var items = ac.children;
    if (e.key==="ArrowDown"){acIndex=Math.min(acIndex+1,items.length-1);for(var i=0;i<items.length;i++)items[i].style.background=i===acIndex?"#e9ecef":"white";}
    else if(e.key==="ArrowUp"){acIndex=Math.max(acIndex-1,0);for(var i=0;i<items.length;i++)items[i].style.background=i===acIndex?"#e9ecef":"white";}
    else if(e.key==="Enter"&&acIndex>=0&&items[acIndex]){items[acIndex].click();e.preventDefault();}
  };
  window["highlightVolcanoGenes_" + cid] = function() {
    var el = document.getElementById("volcano_genes_" + cid);
    var val = el.value.trim();
    if (val && selectedGenes.indexOf(val)===-1) { selectedGenes.push(val); el.value=""; renderTags(); applyHighlight(); }
  };
  window["clearVolcanoHighlight_" + cid] = function() { selectedGenes=[]; renderTags(); applyHighlight(); };
  window["removeVolcanoGene_" + cid] = function(g) { selectedGenes=selectedGenes.filter(function(x){return x!==g;}); renderTags(); applyHighlight(); };
})();
</script>\n', vol_cid))

      figure_legend(paste0("Interactive volcano plot for ", ctr, ". Type metabolite names in the search box to annotate them on the plot. Hover over points for details."))
      cat("\n\n")
    }

    # ================================================================
    # MA PLOT
    # ================================================================
    ma_png <- de_res$files[grepl(paste0("MA_", ctr_label, "\\.png$"), basename(de_res$files))]
    ma_png <- ma_png[file.exists(ma_png)]
    if (length(ma_png) > 0) {
      cat("\n## MA Plot: ", ctr, " {.tabset}\n\n", sep = "")

      cat("### Static\n\n")
      cat(sprintf("\n![MA Plot: %s](%s){width=100%%}\n\n", ctr, ma_png[1]))
      figure_legend(paste0("MA plot for contrast ", ctr, ". The x-axis shows average expression and the y-axis shows log2 fold change. Significant features are highlighted."))
      cat("\n\n")

      # Interactive MA
      if (requireNamespace("plotly", quietly = TRUE)) {
        cat("### Interactive\n\n")

        ma_cid <- paste0("ma_", gsub("[^a-zA-Z0-9]", "", ctr_label))

        ma_p <- ggplot2::ggplot(
          ctr_tbl,
          ggplot2::aes(x = AveExpr, y = logFC, color = Direction,
                       text = paste0("Metabolite: ", Name,
                                     "<br>log2(FC): ", round(logFC, 2),
                                     "<br>AveExpr: ", round(AveExpr, 2),
                                     "<br>P.Value: ", sprintf("%.2e", P.Value),
                                     "<br>adj.P.Val: ", sprintf("%.2e", adj.P.Val)))
        ) +
          ggplot2::geom_point(alpha = 0.6, size = 1.5) +
          ggplot2::scale_color_manual(values = c("NS" = "grey70", "Down" = "blue", "Up" = "red")) +
          ggplot2::geom_hline(yintercept = c(-log2_lfc_de, 0, log2_lfc_de),
                              linetype = c("dashed", "solid", "dashed"),
                              alpha = c(0.5, 0.3, 0.5)) +
          ggplot2::labs(x = "Average Expression", y = "log2(FC)") +
          ggplot2::theme_minimal()

        ma_plotly <- plotly::ggplotly(ma_p, tooltip = "text")

        cat(sprintf('
<div style="margin-bottom: 15px; position: relative;">
<label for="ma_genes_%s"><strong>Highlight metabolites:</strong></label>
<div style="display: inline-block; position: relative; margin-left: 10px;">
<input type="text" id="ma_genes_%s" placeholder="Start typing a metabolite name..."
  autocomplete="off"
  style="padding: 8px; border-radius: 4px; border: 1px solid #ccc; width: 350px;"
  oninput="maAutocomplete_%s(this)" onkeydown="maAutoKey_%s(event)">
<div id="ma_ac_%s" style="position: absolute; top: 100%%; left: 0; width: 350px; max-height: 200px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 4px 4px; z-index: 1000; display: none; box-shadow: 0 4px 6px rgba(0,0,0,0.1);"></div>
</div>
<button onclick="highlightMAGenes_%s()" style="padding: 8px 15px; margin-left: 5px; background: #007bff; color: white; border: none; border-radius: 4px; cursor: pointer;">Highlight</button>
<button onclick="clearMAHighlight_%s()" style="padding: 8px 15px; margin-left: 5px; background: #6c757d; color: white; border: none; border-radius: 4px; cursor: pointer;">Clear</button>
<div id="ma_tags_%s" style="margin-top: 5px; display: flex; flex-wrap: wrap; gap: 4px;"></div>
</div>
', ma_cid, ma_cid, ma_cid, ma_cid, ma_cid, ma_cid, ma_cid, ma_cid))

        cat(sprintf('<div id="ma_plot_%s">\n', ma_cid))
        print(htmltools::tagList(ma_plotly))
        cat('</div>\n')

        # Register MA plot for cutoff panel
        if (exists("cutoff_register_plot")) {
          ma_reg_df <- data.frame(
            name = as.character(ctr_tbl$Name),
            logFC = ctr_tbl$logFC,
            pval = ctr_tbl$P.Value,
            avgExpr = ctr_tbl$AveExpr,
            adjPval = ctr_tbl$adj.P.Val,
            stringsAsFactors = FALSE
          )
          cat(cutoff_register_plot(
            paste0("ma_plot_", ma_cid),
            ma_reg_df, "ma", "Metabolite", ctr
          ))
        }

        cat(sprintf('<script>
(function() {
  var cid = "%s"; var selectedGenes = []; var acIndex = -1;
  function getNames() {
    var plotDiv = document.getElementById("ma_plot_" + cid);
    var el = plotDiv ? plotDiv.querySelector(".js-plotly-plot") || plotDiv : null;
    if (!el || !el.data) return [];
    var names = [];
    for (var t = 0; t < el.data.length; t++) {
      var d = el.data[t]; if (!d.text) continue;
      for (var i = 0; i < d.text.length; i++) {
        var m = (typeof d.text[i]==="string"?d.text[i]:"").match(/Metabolite:\\s*(.+?)(?:<br>|$)/);
        if (m && m[1] && names.indexOf(m[1])===-1) names.push(m[1]);
      }
    } return names.sort();
  }
  function renderTags() {
    var c = document.getElementById("ma_tags_" + cid); c.innerHTML = "";
    selectedGenes.forEach(function(g) {
      var tag = document.createElement("span");
      tag.style.cssText = "display:inline-flex;align-items:center;background:#007bff;color:white;padding:2px 8px;border-radius:12px;font-size:12px;gap:4px;";
      var txt = document.createTextNode(g + " ");
      tag.appendChild(txt);
      var btn = document.createElement("span");
      btn.textContent = String.fromCharCode(215);
      btn.style.cssText = "cursor:pointer;font-weight:bold;";
      btn.addEventListener("click",(function(gene){return function(){window["removeMAGene_"+cid](gene);};})(g));
      tag.appendChild(btn);
      c.appendChild(tag);
    });
  }
  function applyHighlight() {
    var plotDiv = document.getElementById("ma_plot_" + cid);
    var el = plotDiv ? plotDiv.querySelector(".js-plotly-plot") || plotDiv : null;
    if (!el || !el.data) return;
    if (selectedGenes.length===0){Plotly.relayout(el,{annotations:[]});return;}
    var gl = selectedGenes.map(function(g){return g.toLowerCase();});
    var ann = [];
    for (var t=0;t<el.data.length;t++){var d=el.data[t];if(!d.text)continue;
      for(var i=0;i<d.text.length;i++){var m2=(typeof d.text[i]==="string"?d.text[i]:"").match(/Metabolite:\\s*(.+?)(?:<br>|$)/);
        if(m2&&gl.indexOf(m2[1].toLowerCase())!==-1){ann.push({x:d.x[i],y:d.y[i],text:m2[1],showarrow:true,arrowhead:2,ax:20,ay:-30,font:{size:11,color:"black"},bgcolor:"rgba(255,255,255,0.9)",bordercolor:"black",borderwidth:1});}
    }} Plotly.relayout(el,{annotations:ann});
  }
  window["maAutocomplete_"+cid]=function(el){var ac=document.getElementById("ma_ac_"+cid);var val=el.value.trim().toLowerCase();ac.innerHTML="";acIndex=-1;if(val.length<1){ac.style.display="none";return;}var all=getNames();var matches=all.filter(function(g){return g.toLowerCase().indexOf(val)!==-1&&selectedGenes.indexOf(g)===-1;}).slice(0,15);if(matches.length===0){ac.style.display="none";return;}matches.forEach(function(g){var item=document.createElement("div");item.textContent=g;item.style.cssText="padding:6px 10px;cursor:pointer;font-size:13px;";item.onmouseenter=function(){this.style.background="#e9ecef";};item.onmouseleave=function(){this.style.background="white";};item.onclick=function(){selectedGenes.push(g);el.value="";ac.style.display="none";renderTags();applyHighlight();};ac.appendChild(item);});ac.style.display="block";};
  window["maAutoKey_"+cid]=function(e){var ac=document.getElementById("ma_ac_"+cid);var items=ac.children;if(e.key==="ArrowDown"){acIndex=Math.min(acIndex+1,items.length-1);for(var i=0;i<items.length;i++)items[i].style.background=i===acIndex?"#e9ecef":"white";}else if(e.key==="ArrowUp"){acIndex=Math.max(acIndex-1,0);for(var i=0;i<items.length;i++)items[i].style.background=i===acIndex?"#e9ecef":"white";}else if(e.key==="Enter"&&acIndex>=0&&items[acIndex]){items[acIndex].click();e.preventDefault();}};
  window["highlightMAGenes_"+cid]=function(){var el=document.getElementById("ma_genes_"+cid);var val=el.value.trim();if(val&&selectedGenes.indexOf(val)===-1){selectedGenes.push(val);el.value="";renderTags();applyHighlight();}};
  window["clearMAHighlight_"+cid]=function(){selectedGenes=[];renderTags();applyHighlight();};
  window["removeMAGene_"+cid]=function(g){selectedGenes=selectedGenes.filter(function(x){return x!==g;});renderTags();applyHighlight();};
})();
</script>\n', ma_cid))
        figure_legend(paste0("Interactive MA plot for ", ctr, ". Search and highlight specific metabolites."))
        cat("\n\n")
      }
    }

    # ================================================================
    # P-VALUE DISTRIBUTION
    # ================================================================
    phist_png <- de_res$files[grepl(paste0("pvalue_hist_", ctr_label, "\\.png$"), basename(de_res$files))]
    phist_png <- phist_png[file.exists(phist_png)]
    if (length(phist_png) > 0) {
      cat("\n## P-value Distribution: ", ctr, "\n\n", sep = "")
      cat(sprintf("\n![P-value Distribution: %s](%s){width=85%%}\n\n", ctr, phist_png[1]))
      figure_legend(paste0("Distribution of raw p-values for contrast ", ctr, ". A uniform distribution suggests no differential signal; enrichment near zero indicates true positives."))
      cat("\n\n")
    }

    # ================================================================
    # DE RESULTS TABLE (statistics + normalized intensity)
    # ================================================================
    cat("\n## DE Results: ", ctr, " {.tabset}\n\n", sep = "")

    # Top 50 table
    top_tbl <- ctr_tbl[order(ctr_tbl$P.Value), ]
    top_tbl <- head(top_tbl, 50)
    top_display <- top_tbl[, c("feature_id", "Name", "logFC", "AveExpr", "P.Value", "adj.P.Val", "Direction")]
    cat("### Top 50 DE Metabolites\n\n")
    print(make_dt(top_display, caption = paste0("Top 50 DE metabolites: ", ctr)))
    cat("\n\n")

    # Full table with normalized intensities
    cat("### Full Results + Expression\n\n")
    cat("<p style='font-size: 0.9em; color: #555; margin-bottom: 8px;'><em>",
        "Full DE results merged with normalized expression values. ",
        "Use the search box to find specific metabolites and the buttons to export.</em></p>\n")

    full_tbl <- ctr_tbl[order(ctr_tbl$P.Value), ,drop=FALSE]
    full_display <- full_tbl[, c("feature_id", "Name", "logFC", "AveExpr", "P.Value", "adj.P.Val", "Direction")]

    # Merge transformed expression (glog10, before scaling) for meaningful values
    de_tbl_expr <- pre$expr_log %||% pre$expr_work
    if (!is.null(de_tbl_expr)) {
      expr_df2 <- as.data.frame(de_tbl_expr, check.names = FALSE)
      expr_df2$feature_id <- rownames(de_tbl_expr)
      merged <- merge(full_display, expr_df2, by = "feature_id", all.x = TRUE)
      if (nrow(merged) > 0) full_display <- merged
    }

    # Format p-value columns as scientific notation strings before DT
    pval_pat <- "^(p[._]?val|padj|adj[._]?p|fdr|q[._]?val|BH|bonferroni|p\\.value|adj\\.P\\.Val|PValue|FDR)"
    pval_cols_de <- grep(pval_pat, names(full_display), ignore.case = TRUE, value = TRUE)
    for (pc in pval_cols_de) {
      if (is.numeric(full_display[[pc]])) {
        full_display[[pc]] <- ifelse(is.na(full_display[[pc]]), NA_character_,
                                     sprintf("%.2e", full_display[[pc]]))
      }
    }
    num_cols_de <- names(full_display)[vapply(full_display, is.numeric, logical(1))]
    tbl_de <- DT::datatable(
      full_display,
      caption = paste0("DE Results + Normalized Expression: ", ctr),
      extensions = "Buttons",
      options = list(
        scrollX = TRUE, pageLength = 10, dom = "Bfrtip",
        buttons = list(
          list(extend = "csv", filename = paste0("DE_expression_", ctr_label)),
          list(extend = "excel", filename = paste0("DE_expression_", ctr_label))
        ),
        search = list(regex = TRUE, caseInsensitive = TRUE)
      ),
      rownames = FALSE, filter = "top",
      class = "compact stripe hover"
    )
    if (length(num_cols_de) > 0) tbl_de <- DT::formatRound(tbl_de, columns = num_cols_de, digits = 2)
    print(htmltools::tagList(tbl_de))
    cat("\n\n")

  } # end contrast loop
}
```

```{r metabolite-explorer-prep, include=FALSE}
# Prepare data for the metabolite expression explorer
explorer_ready <- FALSE
explorer_json_str <- "{}"

# Use expr_log (transform-only, e.g. glog10) for the explorer so that means
# reflect actual intensity levels rather than z-scored zeros.
# Fall back to expr_work if expr_log is not available.
explorer_expr <- pre$expr_log %||% pre$expr_work

if (!is.null(explorer_expr) && !is.null(pre$meta)) {
  # Build metabolite display names
  metab_ids <- rownames(explorer_expr)
  metab_display <- vapply(metab_ids, resolve_name, character(1), USE.NAMES = FALSE)

  # Build sample -> group mapping
  sample_col_exp <- metab_cfg$effects$samples %||% "sample_id"
  group_col_exp  <- cond_col
  meta_exp       <- pre$meta

  if (sample_col_exp %in% names(meta_exp) && group_col_exp %in% names(meta_exp)) {
    group_map_exp <- stats::setNames(as.character(meta_exp[[group_col_exp]]),
                                      meta_exp[[sample_col_exp]])
    sample_groups_exp <- as.character(group_map_exp[colnames(explorer_expr)])
    sample_groups_exp[is.na(sample_groups_exp)] <- "Unknown"
  } else {
    sample_groups_exp <- rep("All", ncol(explorer_expr))
  }

  # Build expression matrix as list of numeric vectors
  expr_matrix_list <- lapply(seq_len(nrow(explorer_expr)), function(i) {
    round(as.numeric(explorer_expr[i, ]), 2)
  })

  explorer_data <- list(
    genes   = metab_display,
    samples = colnames(explorer_expr),
    groups  = sample_groups_exp,
    expr    = expr_matrix_list
  )

  explorer_json_str <- jsonlite::toJSON(explorer_data, auto_unbox = FALSE)
  explorer_ready <- TRUE
}
```

`r if (explorer_ready) "" else "<!--"`

## Metabolite Explorer

Search for metabolites and visualize their transformed expression (glog10, before scaling) across groups.

```{r metabolite-explorer, results='asis'}
if (explorer_ready) {
  cat(sprintf('
<div style="margin: 15px 0 20px 0;">
<div style="display: flex; align-items: center; gap: 10px; flex-wrap: wrap;">
<label for="boxplot_gene_input" style="font-weight: bold; white-space: nowrap;">Search metabolite:</label>
<div style="position: relative; flex: 1; max-width: 400px;">
<input type="text" id="boxplot_gene_input" placeholder="Start typing a metabolite name..."
  autocomplete="off"
  style="padding: 10px 12px; border-radius: 6px; border: 1px solid #ccc; width: 100%%; font-size: 14px; box-sizing: border-box;"
  oninput="boxplotAutocomplete(this)" onkeydown="boxplotAutoKey(event)">
<div id="boxplot_ac" style="position: absolute; top: 100%%; left: 0; width: 100%%; max-height: 220px; overflow-y: auto; background: white; border: 1px solid #ccc; border-top: none; border-radius: 0 0 6px 6px; z-index: 1000; display: none; box-shadow: 0 4px 8px rgba(0,0,0,0.12);"></div>
</div>
<button onclick="addGeneToPlot()" style="padding: 10px 20px; background: #007bff; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Add</button>
<button onclick="clearBoxplot()" style="padding: 10px 20px; background: #6c757d; color: white; border: none; border-radius: 6px; cursor: pointer; font-size: 14px;">Clear All</button>
</div>
<div id="boxplot_gene_tags" style="margin-top: 8px; display: flex; flex-wrap: wrap; gap: 6px;"></div>
<div id="boxplot_gene_info" style="margin-top: 8px; font-size: 0.9em; color: #555;"></div>
</div>
<div id="boxplot_container" style="width: 100%%; min-height: 450px; margin-bottom: 10px;"></div>
<div id="boxplot_export_bar" style="display: none; margin-bottom: 20px; gap: 8px; align-items: center;">
  <span style="font-size: 0.85em; color: #555; margin-right: 4px;">Export:</span>
  <button onclick="exportBoxplot(\'svg\')" style="padding: 5px 14px; background: #28a745; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">SVG</button>
  <button onclick="exportBoxplot(\'png\')" style="padding: 5px 14px; background: #17a2b8; color: white; border: none; border-radius: 4px; cursor: pointer; font-size: 13px;">PNG</button>
</div>
'))

  # Inject JS data and rendering logic
  cat(sprintf('<script>
var geneExprData = %s;
var bpGroupLabel = "%s";
var bpYLabel = "%s";
var boxplotAcIndex = -1;
var selectedGenes = [];

function boxplotAutocomplete(el) {
  var ac = document.getElementById("boxplot_ac");
  var val = el.value.trim().toLowerCase(); ac.innerHTML = ""; boxplotAcIndex = -1;
  if (val.length < 1) { ac.style.display = "none"; return; }
  var matches = [];
  for (var i = 0; i < geneExprData.genes.length; i++) {
    if (geneExprData.genes[i].toLowerCase().indexOf(val) !== -1) {
      if (selectedGenes.some(function(g){return g.idx===i;})) continue;
      matches.push({name: geneExprData.genes[i], idx: i});
      if (matches.length >= 20) break;
    }
  }
  if (matches.length === 0) { ac.style.display = "none"; return; }
  matches.forEach(function(m) {
    var item = document.createElement("div"); item.textContent = m.name;
    item.style.cssText = "padding: 8px 12px; cursor: pointer; font-size: 13px; border-bottom: 1px solid #f0f0f0;";
    item.onmouseenter = function() { this.style.background = "#e9ecef"; };
    item.onmouseleave = function() { this.style.background = "white"; };
    item.onclick = function() { document.getElementById("boxplot_gene_input").value = ""; ac.style.display = "none"; addGeneByIdx(m.idx, m.name); };
    ac.appendChild(item);
  });
  ac.style.display = "block";
}
function boxplotAutoKey(e) {
  var ac = document.getElementById("boxplot_ac"); var items = ac.children;
  if (e.key==="ArrowDown"){boxplotAcIndex=Math.min(boxplotAcIndex+1,items.length-1);for(var i=0;i<items.length;i++)items[i].style.background=i===boxplotAcIndex?"#e9ecef":"white";}
  else if(e.key==="ArrowUp"){boxplotAcIndex=Math.max(boxplotAcIndex-1,0);for(var i=0;i<items.length;i++)items[i].style.background=i===boxplotAcIndex?"#e9ecef":"white";}
  else if(e.key==="Enter"){if(boxplotAcIndex>=0&&items[boxplotAcIndex])items[boxplotAcIndex].click(); else addGeneToPlot(); e.preventDefault();}
}
function addGeneToPlot() {
  var el = document.getElementById("boxplot_gene_input");
  var val = el.value.trim(); if (!val) return;
  var vl = val.toLowerCase();
  for (var i = 0; i < geneExprData.genes.length; i++) {
    if (geneExprData.genes[i].toLowerCase() === vl) { addGeneByIdx(i, geneExprData.genes[i]); el.value = ""; return; }
  }
  // Partial match
  for (var i = 0; i < geneExprData.genes.length; i++) {
    if (geneExprData.genes[i].toLowerCase().indexOf(vl) !== -1) { addGeneByIdx(i, geneExprData.genes[i]); el.value = ""; return; }
  }
}
function addGeneByIdx(idx, name) {
  if (selectedGenes.some(function(g){return g.idx===idx;})) return;
  selectedGenes.push({name:name, idx:idx}); renderGeneTags(); renderMultiBoxplot();
}
function removeGene(idx) {
  selectedGenes = selectedGenes.filter(function(g){return g.idx!==idx;});
  renderGeneTags(); if (selectedGenes.length>0) renderMultiBoxplot();
  else { document.getElementById("boxplot_container").innerHTML=""; document.getElementById("boxplot_gene_info").innerHTML=""; document.getElementById("boxplot_export_bar").style.display="none"; }
}
function clearBoxplot() {
  selectedGenes = []; renderGeneTags();
  document.getElementById("boxplot_container").innerHTML = "";
  document.getElementById("boxplot_gene_info").innerHTML = "";
  document.getElementById("boxplot_export_bar").style.display = "none";
}
function renderGeneTags() {
  var c = document.getElementById("boxplot_gene_tags"); c.innerHTML = "";
  selectedGenes.forEach(function(g) {
    var tag = document.createElement("span");
    tag.style.cssText = "display:inline-flex;align-items:center;background:#007bff;color:white;padding:3px 10px;border-radius:14px;font-size:12px;gap:4px;";
    var txt = document.createTextNode(g.name + " ");
    tag.appendChild(txt);
    var btn = document.createElement("span");
    btn.textContent = String.fromCharCode(215);
    btn.style.cssText = "cursor:pointer;font-weight:bold;margin-left:3px;";
    btn.addEventListener("click", (function(idx){return function(){removeGene(idx);};})(g.idx));
    tag.appendChild(btn);
    c.appendChild(tag);
  });
}
function renderMultiBoxplot() {
  if (selectedGenes.length === 0) return;
  var groups = geneExprData.groups; var samples = geneExprData.samples;
  var uGroups = []; groups.forEach(function(g){if(uGroups.indexOf(g)===-1)uGroups.push(g);});
  var colors = ["#e41a1c","#377eb8","#4daf4a","#984ea3","#ff7f00","#a65628","#f781bf","#999999"];
  var nGenes = selectedGenes.length; var traces = [];
  if (nGenes === 1) {
    var gIdx = selectedGenes[0].idx; var values = geneExprData.expr[gIdx];
    uGroups.forEach(function(g, gi) {
      var gv = [], gs = [];
      for (var i=0;i<values.length;i++) { if(groups[i]===g){gv.push(values[i]);gs.push(samples[i]);} }
      traces.push({y:gv,type:"box",name:g,boxpoints:"all",jitter:0.4,pointpos:0,text:gs,
        hovertemplate:"Sample: %%{text}<br>Expression: %%{y:.3f}<extra>"+g+"</extra>",
        marker:{color:colors[gi%%colors.length],size:6,opacity:0.7},
        line:{color:colors[gi%%colors.length]},fillcolor:colors[gi%%colors.length]+"33"});
    });
    var layout = {title:{text:selectedGenes[0].name+" Expression",font:{size:16}},yaxis:{title:bpYLabel},xaxis:{title:bpGroupLabel},showlegend:false,margin:{t:50,b:80,l:70,r:30},boxgap:0.3};
  } else {
    uGroups.forEach(function(g, gi) {
      var xv=[],yv=[],st=[];
      selectedGenes.forEach(function(sg){var vals=geneExprData.expr[sg.idx];for(var i=0;i<vals.length;i++){if(groups[i]===g){xv.push(sg.name);yv.push(vals[i]);st.push(samples[i]);}}});
      traces.push({x:xv,y:yv,type:"box",name:g,boxpoints:"all",jitter:0.4,pointpos:0,text:st,
        hovertemplate:"%%{x}<br>Sample: %%{text}<br>Expression: %%{y:.3f}<extra>"+g+"</extra>",
        marker:{color:colors[gi%%colors.length],size:5,opacity:0.7},
        line:{color:colors[gi%%colors.length]},fillcolor:colors[gi%%colors.length]+"33"});
    });
    var gNames = selectedGenes.map(function(sg){return sg.name;});
    var layout = {title:{text:"Metabolite Expression Comparison",font:{size:16}},yaxis:{title:bpYLabel},xaxis:{title:"Metabolite",categoryorder:"array",categoryarray:gNames},showlegend:true,legend:{title:{text:bpGroupLabel}},boxmode:"group",margin:{t:50,b:100,l:70,r:30},boxgap:0.15,boxgroupgap:0.1};
  }
  var ch = Math.max(450, 350 + nGenes * 30);
  document.getElementById("boxplot_container").style.height = ch + "px";
  Plotly.newPlot("boxplot_container", traces, layout, {responsive:true});
  var info = selectedGenes.map(function(sg){var vals=geneExprData.expr[sg.idx].filter(function(v){return !isNaN(v);});var mean=vals.reduce(function(a,b){return a+b;},0)/vals.length;return "<strong>"+sg.name+"</strong>: mean "+mean.toFixed(3)+" ("+vals.length+" samples)";});
  document.getElementById("boxplot_gene_info").innerHTML = info.join(" &nbsp;|&nbsp; ");
  document.getElementById("boxplot_export_bar").style.display = "flex";
}
function exportBoxplot(fmt) {
  var c = document.getElementById("boxplot_container");
  if (!c||!c.data){alert("No plot to export.");return;}
  var gNames = selectedGenes.map(function(g){return g.name;}).join("_");
  var w = fmt==="svg"?1200:2400, h = fmt==="svg"?800:1600, s = fmt==="svg"?1:3;
  Plotly.downloadImage(c,{format:fmt,width:w,height:h,scale:s,filename:"metabolite_expression_"+gNames});
}
</script>\n', explorer_json_str, group_col_exp, norm_y_label_report))

  figure_legend("Metabolite Expression Explorer. Search for metabolites by name, add multiple for side-by-side comparison. Each box shows the distribution of normalized expression values per group. Individual sample points are overlaid. Use the export buttons to save as SVG or PNG.")
}
```

`r if (explorer_ready) "" else "-->"`

`r if (is.null(de_res) || is.null(de_res$summary_df)) "-->"`

---

`r if (is.null(de_res) || is.null(de_res$summary_df)) "<!--"`

# Top DE Heatmaps {.tabset}

Heatmaps of differentially expressed metabolites (adjusted p-value < `r padj_cut_de`, z-score scaled) with hierarchical clustering.

```{r de-heatmaps, results='asis', fig.width=10, fig.height=8}
if (!is.null(de_res$summary_df) && requireNamespace("pheatmap", quietly = TRUE)) {
  # Find the Diagnostic_plots directory for heatmap images
  hm_out_dir <- NULL
  if (length(de_res$files) > 0) {
    png_files <- de_res$files[grepl("\\.png$", de_res$files)]
    if (length(png_files) > 0) {
      diag_dirs <- unique(dirname(png_files))
      diag_dirs <- diag_dirs[dir.exists(diag_dirs)]
      if (length(diag_dirs) > 0) hm_out_dir <- diag_dirs[1]
    }
  }
  if (is.null(hm_out_dir)) hm_out_dir <- file.path(dirname(knitr::current_input(dir = TRUE)), "metabolomics", "Diagnostic_plots")
  dir.create(hm_out_dir, recursive = TRUE, showWarnings = FALSE)

  # Identify DE features for heatmaps using raw P.Value (ignoring FC cutoff).
  # Heatmaps benefit from showing all significant metabolites; the clustering
  # pattern reveals biology even for features with modest fold changes.
  sum_df <- de_res$summary_df
  pval_cols <- grep("^P\\.Value", colnames(sum_df), value = TRUE)
  if (length(pval_cols) > 0) {
    pval_mat <- sum_df[, pval_cols, drop = FALSE]
    min_pval_per_feat <- apply(pval_mat, 1, function(x) min(x, na.rm = TRUE))
    de_feats <- sum_df$feature_id[min_pval_per_feat <= padj_cut_de]
  } else {
    de_feats <- character(0)
  }

  # Sort by minimum raw p-value
  if (length(de_feats) > 0) {
    feat_pval <- min_pval_per_feat[sum_df$feature_id %in% de_feats]
    de_feats <- de_feats[order(feat_pval)]
  }

  # Use expr_log (transform-only) for heatmaps — z-scoring removes scale anyway,
  # but expr_log preserves natural variance structure better than auto-scaled data.
  hm_expr <- pre$expr_log %||% pre$expr_work

  if (length(de_feats) >= 2 && !is.null(hm_expr)) {
    # Build annotation
    anno_df <- NULL
    if (!is.null(pre$meta) && cond_col %in% colnames(pre$meta)) {
      sample_col_hm <- metab_cfg$effects$samples %||% "sample_id"
      anno_df <- data.frame(
        row.names = pre$meta[[sample_col_hm]],
        Group = pre$meta[[cond_col]],
        stringsAsFactors = FALSE
      )
    }

    # Build display labels (use Name if available)
    hm_label_map <- if (!is.null(name_map_v)) name_map_v else stats::setNames(de_feats, de_feats)

    for (n_top in c(10, 25, 50, 100)) {
      if (n_top > length(de_feats)) next
      top_ids <- utils::head(de_feats, n_top)

      mat_hm <- hm_expr[top_ids, , drop = FALSE]
      # Z-score scale rows
      mat_z <- t(scale(t(mat_hm)))
      mat_z[is.na(mat_z)] <- 0

      # Use display names for rows
      row_labels <- vapply(rownames(mat_z), function(fid) {
        nm <- hm_label_map[fid]
        if (!is.na(nm) && nzchar(nm)) nm else fid
      }, character(1), USE.NAMES = FALSE)
      rownames(mat_z) <- row_labels

      cat(sprintf("\n## Top %d\n\n", n_top))

      # Adjust figure height based on number of features
      fig_h <- max(6, min(20, 4 + n_top * 0.15))
      show_rownames <- n_top <= 60

      hm_file <- file.path(hm_out_dir, sprintf("heatmap_top%d_DE.png", n_top))
      hm_ok <- tryCatch({
        p_hm <- pheatmap::pheatmap(
          mat_z,
          annotation_col = anno_df,
          cluster_rows = TRUE,
          cluster_cols = TRUE,
          show_rownames = show_rownames,
          show_colnames = TRUE,
          fontsize_row = max(5, min(9, 120 / n_top)),
          fontsize_col = 8,
          main = sprintf("Top %d DE Metabolites (z-score)", n_top),
          color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
          border_color = NA,
          silent = TRUE
        )
        ggplot2::ggsave(hm_file, p_hm, width = 10, height = fig_h, dpi = 150)
        TRUE
      }, error = function(e) { message("Heatmap top ", n_top, " failed: ", e$message); FALSE })
      if (hm_ok && file.exists(hm_file)) {
        cat(sprintf("\n![Top %d DE Heatmap](%s){width=100%%}\n\n", n_top, hm_file))
        figure_legend(sprintf(
          "Heatmap of the top %d differentially expressed metabolites (ranked by adjusted p-value). Values are z-score normalized per metabolite. Rows and columns are hierarchically clustered.",
          n_top
        ))
        cat("\n\n")
      }
    }

    # All DE heatmap (if reasonable size and different from last top-N)
    if (length(de_feats) >= 2 && length(de_feats) <= 500) {
      n_all <- length(de_feats)
      mat_all <- hm_expr[de_feats, , drop = FALSE]
      mat_z_all <- t(scale(t(mat_all)))
      mat_z_all[is.na(mat_z_all)] <- 0

      row_labels_all <- vapply(rownames(mat_z_all), function(fid) {
        nm <- hm_label_map[fid]
        if (!is.na(nm) && nzchar(nm)) nm else fid
      }, character(1), USE.NAMES = FALSE)
      rownames(mat_z_all) <- row_labels_all

      cat(sprintf("\n## All DE (%d)\n\n", n_all))

      fig_h_all <- max(8, min(30, 4 + n_all * 0.06))
      hm_all_file <- file.path(hm_out_dir, sprintf("heatmap_all%d_DE.png", n_all))
      hm_all_ok <- tryCatch({
        p_hm_all <- pheatmap::pheatmap(
          mat_z_all,
          annotation_col = anno_df,
          cluster_rows = TRUE,
          cluster_cols = TRUE,
          show_rownames = n_all <= 60,
          show_colnames = TRUE,
          fontsize_row = max(4, min(8, 100 / n_all)),
          fontsize_col = 8,
          main = sprintf("All %d DE Metabolites (z-score)", n_all),
          color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
          border_color = NA,
          silent = TRUE
        )
        ggplot2::ggsave(hm_all_file, p_hm_all, width = 10, height = fig_h_all, dpi = 150)
        TRUE
      }, error = function(e) { message("Heatmap all DE failed: ", e$message); FALSE })
      if (hm_all_ok && file.exists(hm_all_file)) {
        cat(sprintf("\n![All DE Heatmap](%s){width=100%%}\n\n", hm_all_file))
      }
      figure_legend(sprintf(
        "Heatmap of all %d differentially expressed metabolites. Values are z-score normalized per metabolite.",
        n_all
      ))
      cat("\n\n")
    }
  } else if (length(de_feats) < 2) {
    cat("\n*Fewer than 2 DE metabolites — heatmap not generated.*\n\n")
  }
}
```

`r if (is.null(de_res) || is.null(de_res$summary_df)) "-->"`

---

`r if (is.null(rf_res) && is.null(plsda_res)) "<!--"`

# Feature Selection {.tabset}

`r if (is.null(rf_res)) "<!--"`

## Random Forest {.tabset}

### Importance Plot

```{r rf-plot, fig.width=9, fig.height=7, results='asis'}
if (!is.null(rf_res$plots$rf_importance)) {
  safe_print_plot(rf_res$plots$rf_importance)
  figure_legend("Random Forest variable importance plot showing the top features ranked by their contribution to classification accuracy. Higher Mean Decrease Accuracy indicates greater importance.")
}
```

### Importance Table

```{r rf-table}
if (!is.null(rf_res$importance_df)) {
  imp_df <- rf_res$importance_df
  sort_col <- if ("MeanDecreaseAccuracy" %in% names(imp_df)) "MeanDecreaseAccuracy" else "importance"
  imp_df <- imp_df[order(-imp_df[[sort_col]]), ]
  make_dt(head(imp_df, 50), caption = "RF Feature Importance (top 50)")
}
```

`r if (is.null(rf_res)) "-->"`

`r if (is.null(plsda_res)) "<!--"`

## PLS-DA {.tabset}

### Scores Plot

```{r plsda-scores, fig.width=9, fig.height=7, results='asis'}
if (!is.null(plsda_res$plots$plsda_scores)) {
  safe_print_plot(plsda_res$plots$plsda_scores)
  figure_legend("PLS-DA scores plot showing sample separation along the first two latent variables. Clustering by group indicates good discriminatory power of the metabolomic profiles.")
}
```

### VIP Plot

```{r plsda-vip-plot, fig.width=9, fig.height=7, results='asis'}
if (!is.null(plsda_res$plots$plsda_vip)) {
  safe_print_plot(plsda_res$plots$plsda_vip)
  figure_legend("PLS-DA Variable Importance in Projection (VIP) scores. Features with VIP > 1 are generally considered important for group discrimination.")
}
```

### VIP Table

```{r plsda-vip-table}
if (!is.null(plsda_res$vip_df)) {
  vip_df <- plsda_res$vip_df
  vip_col <- if ("VIP" %in% names(vip_df)) "VIP" else grep("^VIP", names(vip_df), value = TRUE)[1]
  vip_df <- vip_df[order(-vip_df[[vip_col]]), ]
  make_dt(head(vip_df, 50), caption = "PLS-DA VIP Scores (top 50)")
}
```

`r if (is.null(plsda_res)) "-->"`

`r if (is.null(rf_res) && is.null(plsda_res)) "-->"`

---

`r if (is.null(enrichment_res)) "<!--"`

# Enrichment {.tabset}

`r if (is.null(enrichment_res$qea)) "<!--"`

## QEA {.tabset}

### Lollipop Plot

```{r qea-lollipop, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$qea_lollipop)) {
  safe_print_plot(enrichment_res$plots$qea_lollipop)
  figure_legend("Quantitative Enrichment Analysis (QEA) lollipop plot showing the top enriched metabolic pathways. Bar length represents statistical significance; color indicates enrichment score direction.")
}
```

### Barplot

```{r qea-barplot, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$qea_barplot)) {
  safe_print_plot(enrichment_res$plots$qea_barplot)
  figure_legend("Barplot of QEA enriched pathways ranked by significance. Pathways are ordered from most to least significant.")
}
```

### Results Table

```{r qea-table}
if (!is.null(enrichment_res$qea$table)) {
  make_dt(enrichment_res$qea$table, caption = "QEA Pathway Enrichment")
}
```

`r if (is.null(enrichment_res$qea)) "-->"`

`r if (is.null(enrichment_res$ssgsea)) "<!--"`

## ssGSEA {.tabset}

### Lollipop Plot

```{r ssgsea-lollipop, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$ssgsea_lollipop)) {
  safe_print_plot(enrichment_res$plots$ssgsea_lollipop)
  figure_legend("ssGSEA lollipop plot showing the top enriched metabolic pathways based on single-sample Gene Set Enrichment Analysis scores.")
}
```

### Barplot

```{r ssgsea-barplot, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$ssgsea_barplot)) {
  safe_print_plot(enrichment_res$plots$ssgsea_barplot)
  figure_legend("Barplot of ssGSEA enrichment results, showing pathway enrichment scores ranked by significance.")
}
```

### Boxplots (static)

```{r ssgsea-boxplots, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$ssgsea_boxplots)) {
  safe_print_plot(enrichment_res$plots$ssgsea_boxplots)
  figure_legend("Static boxplots of ssGSEA pathway scores across conditions. Each box represents the distribution of enrichment scores per condition for the top pathways.")
}
```

### Pathway Explorer

```{r ssgsea-interactive, fig.width=10, fig.height=6, results='asis'}
# Interactive boxplot: select one or more pathways, compare conditions
if (!is.null(enrichment_res$ssgsea$scores) &&
    !is.null(enrichment_res$ssgsea$table)) {

  scores_mat <- enrichment_res$ssgsea$scores
  ssgsea_tbl <- enrichment_res$ssgsea$table

  # Resolve condition labels for each sample
  sample_col_i <- metab_cfg$effects$samples %||% "sample_id"
  cond_col_i   <- metab_cfg$enrichment$condition_column %||%
                   metab_cfg$de$condition_column %||%
                   metab_cfg$effects$color %||% "sample_type"
  meta_i       <- params$pre$meta
  conditions_i <- as.character(meta_i[[cond_col_i]][
    match(colnames(scores_mat), meta_i[[sample_col_i]])
  ])

  # Build long-format data for all pathways
  all_pathways <- rownames(scores_mat)
  pw_order     <- ssgsea_tbl$pathway[ssgsea_tbl$pathway %in% all_pathways]
  # Fall back to rowname order for pathways not in the table
  pw_order     <- c(pw_order, setdiff(all_pathways, pw_order))

  long_df <- do.call(rbind, lapply(pw_order, function(pw) {
    data.frame(
      pathway   = pw,
      sample    = colnames(scores_mat),
      score     = as.numeric(scores_mat[pw, ]),
      condition = conditions_i,
      stringsAsFactors = FALSE
    )
  }))

  # Use plotly with a dropdown to select pathways
  if (requireNamespace("plotly", quietly = TRUE)) {

    # Pre-select top 3 pathways by FDR
    default_pws <- utils::head(pw_order, 3)

    # Build one trace per pathway (visible for defaults, hidden for rest)
    fig <- plotly::plot_ly()

    for (pw in pw_order) {
      pw_data <- long_df[long_df$pathway == pw, ]
      vis <- pw %in% default_pws

      fig <- plotly::add_trace(
        fig,
        data = pw_data,
        x = ~condition,
        y = ~score,
        type = "box",
        boxpoints = "all",
        jitter = 0.4,
        pointpos = 0,
        text = ~sample,
        name = pw,
        visible = vis,
        marker = list(size = 6),
        showlegend = FALSE
      )
    }

    # Build dropdown buttons: each button shows one pathway
    buttons <- lapply(seq_along(pw_order), function(i) {
      vis_vec <- rep(FALSE, length(pw_order))
      vis_vec[i] <- TRUE
      pw_short <- if (nchar(pw_order[i]) > 60) {
        paste0(substr(pw_order[i], 1, 57), "...")
      } else {
        pw_order[i]
      }
      list(method = "restyle",
           args = list("visible", as.list(vis_vec)),
           label = pw_short)
    })

    # Add a "Top 3" button at the beginning
    top3_vis <- pw_order %in% default_pws
    buttons <- c(
      list(list(method = "restyle",
                args = list("visible", as.list(top3_vis)),
                label = paste0("Top 3 by FDR"))),
      buttons
    )

    fig <- plotly::layout(
      fig,
      title = list(text = "ssGSEA Scores by Condition",
                    font = list(size = 16)),
      xaxis = list(title = "Condition"),
      yaxis = list(title = "ssGSEA Score"),
      updatemenus = list(
        list(
          type = "dropdown",
          active = 0,
          x = 0.0, xanchor = "left",
          y = 1.15, yanchor = "top",
          buttons = buttons
        )
      )
    )

    print(htmltools::tagList(fig))
    figure_legend("Interactive ssGSEA pathway explorer. Use the dropdown to select individual pathways and compare their enrichment scores across conditions. Points represent individual samples.")
    cat("\n\n")
  } else {
    cat("*Install the `plotly` package for interactive pathway exploration.*\n")
  }
}
```

### Results Table

```{r ssgsea-table}
if (!is.null(enrichment_res$ssgsea$table)) {
  make_dt(enrichment_res$ssgsea$table, caption = "ssGSEA Pathway Enrichment")
}
```

`r if (is.null(enrichment_res$ssgsea)) "-->"`

`r if (is.null(enrichment_res$ora)) "<!--"`

## ORA {.tabset}

### Lollipop Plot

```{r ora-lollipop, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$ora_lollipop)) {
  safe_print_plot(enrichment_res$plots$ora_lollipop)
  figure_legend("Over-Representation Analysis (ORA) lollipop plot showing pathways significantly enriched among DE metabolites using Fisher's exact test.")
}
```

### Dot Plot

```{r ora-dotplot, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$ora_dotplot)) {
  safe_print_plot(enrichment_res$plots$ora_dotplot)
  figure_legend("ORA dot plot displaying enriched pathways. Point size indicates the number of DE metabolites in the pathway; color represents significance.")
}
```

### Results Table

```{r ora-table}
if (!is.null(enrichment_res$ora$table)) {
  make_dt(enrichment_res$ora$table, caption = "ORA Pathway Enrichment (Fisher's Exact Test)")
}
```

`r if (is.null(enrichment_res$ora)) "-->"`

`r if (is.null(enrichment_res$gsea)) "<!--"`

## GSEA {.tabset}

### Lollipop Plot

```{r gsea-lollipop, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$gsea_lollipop)) {
  safe_print_plot(enrichment_res$plots$gsea_lollipop)
  figure_legend("GSEA lollipop plot showing pathways ranked by Normalized Enrichment Score (NES). Positive NES indicates up-regulation; negative indicates down-regulation of the pathway.")
}
```

### NES Barplot

```{r gsea-nes-barplot, fig.width=12, fig.height=8, results='asis'}
if (!is.null(enrichment_res$plots$gsea_nes_barplot)) {
  safe_print_plot(enrichment_res$plots$gsea_nes_barplot)
  figure_legend("GSEA Normalized Enrichment Score (NES) barplot. Metabolites were pre-ranked by log2 fold change; pathways are ordered by NES magnitude.")
}
```

### Results Table

```{r gsea-table}
if (!is.null(enrichment_res$gsea$table)) {
  make_dt(enrichment_res$gsea$table, caption = "GSEA Pathway Enrichment (fgsea, ranked by log2FC)")
}
```

`r if (is.null(enrichment_res$gsea)) "-->"`

`r if (is.null(enrichment_res)) "-->"`

---

# Output Files

```{r output-files}
files_tbl <- data.frame(
  File = c(
    "metabolomics_report.html",
    "de_summary_all.csv",
    "expression_matrix_normalized.csv"
  ),
  Description = c(
    "This HTML report with all QC, DE, and enrichment results",
    "Full differential expression summary table across all contrasts",
    "Normalized metabolite expression matrix used for downstream analysis"
  ),
  stringsAsFactors = FALSE
)

# Add enrichment output rows if enrichment was run
if (!is.null(enrichment_res)) {
  enrich_rows <- data.frame(
    File = character(0), Description = character(0), stringsAsFactors = FALSE
  )
  if (!is.null(enrichment_res$qea))
    enrich_rows <- rbind(enrich_rows, data.frame(
      File = "enrichment_qea_results.csv",
      Description = "QEA pathway enrichment results table",
      stringsAsFactors = FALSE))
  if (!is.null(enrichment_res$ssgsea))
    enrich_rows <- rbind(enrich_rows, data.frame(
      File = "enrichment_ssgsea_results.csv",
      Description = "ssGSEA pathway enrichment results and scores",
      stringsAsFactors = FALSE))
  if (!is.null(enrichment_res$ora))
    enrich_rows <- rbind(enrich_rows, data.frame(
      File = "enrichment_ora_results.csv",
      Description = "ORA pathway enrichment results (Fisher's exact test)",
      stringsAsFactors = FALSE))
  if (!is.null(enrichment_res$gsea))
    enrich_rows <- rbind(enrich_rows, data.frame(
      File = "enrichment_gsea_results.csv",
      Description = "GSEA pathway enrichment results (fgsea)",
      stringsAsFactors = FALSE))
  if (nrow(enrich_rows) > 0) files_tbl <- rbind(files_tbl, enrich_rows)
}

# Add feature selection outputs
if (!is.null(rf_res))
  files_tbl <- rbind(files_tbl, data.frame(
    File = "rf_importance.csv",
    Description = "Random Forest feature importance scores",
    stringsAsFactors = FALSE))
if (!is.null(plsda_res))
  files_tbl <- rbind(files_tbl, data.frame(
    File = "plsda_vip_scores.csv",
    Description = "PLS-DA VIP scores for all features",
    stringsAsFactors = FALSE))

make_dt(files_tbl, "Key Output Files")
```

---

# Methods

```{r methods-text, results='asis'}
padj_cut_m <- metab_cfg$de$p_cutoff %||% 0.05
lfc_cut_m  <- metab_cfg$de$linear_fc_cutoff %||% 1.5

n_metab <- if (!is.null(pre)) nrow(pre$expr_work) else "N"
n_de_m  <- "N"
if (!is.null(de_res$summary_df) && "pass_any_contrast" %in% colnames(de_res$summary_df)) {
  n_de_m <- sum(de_res$summary_df$pass_any_contrast == 1, na.rm = TRUE)
}

cat("**Metabolomics Analysis Methods**\n\n")

cat("Metabolomics data were processed using the multiomics-core pipeline. ")

if (norm_method != "N/A") {
  cat(sprintf("Data were normalized using %s. ", norm_method))
}

cat(sprintf(
  "Differential expression analysis was performed using %s. ", de_method
))

cat(sprintf(
  "Metabolites were considered differentially expressed at adjusted p-value <= %.2f and |fold change| >= %.2f. ",
  padj_cut_m, lfc_cut_m
))

cat(sprintf(
  "A total of %s metabolites were quantified, with %s identified as differentially expressed in at least one contrast.",
  format(as.numeric(n_metab), big.mark = ","),
  format(as.numeric(n_de_m), big.mark = ",")
))

if (!is.null(enrichment_res)) {
  enrich_methods <- character(0)
  if (!is.null(enrichment_res$qea))    enrich_methods <- c(enrich_methods, "QEA")
  if (!is.null(enrichment_res$ssgsea)) enrich_methods <- c(enrich_methods, "ssGSEA")
  if (!is.null(enrichment_res$ora))    enrich_methods <- c(enrich_methods, "ORA")
  if (!is.null(enrichment_res$gsea))   enrich_methods <- c(enrich_methods, "GSEA")
  if (length(enrich_methods) > 0) {
    cat(sprintf(" Pathway enrichment analysis was performed using %s.",
                paste(enrich_methods, collapse = ", ")))
  }
}

if (!is.null(rf_res) || !is.null(plsda_res)) {
  fs_methods <- character(0)
  if (!is.null(rf_res))    fs_methods <- c(fs_methods, "Random Forest")
  if (!is.null(plsda_res)) fs_methods <- c(fs_methods, "PLS-DA")
  cat(sprintf(" Feature selection was performed using %s.",
              paste(fs_methods, collapse = " and ")))
}

cat("\n\n")

cat("**Copy-ready version:**\n\n")
cat("<div style='background-color: #f5f5f5; padding: 15px; border-radius: 5px; font-size: 0.9em;'>\n")

cat(sprintf(
  "Metabolomics analysis was performed using the multiomics-core pipeline (v1.0). %sDifferential expression was assessed with %s (adjusted p <= %.2f, |FC| >= %.2f). A total of %s metabolites were quantified, with %s differentially expressed.",
  if (norm_method != "N/A") sprintf("Data were normalized using %s. ", norm_method) else "",
  de_method, padj_cut_m, lfc_cut_m,
  format(as.numeric(n_metab), big.mark = ","),
  format(as.numeric(n_de_m), big.mark = ",")
))

cat("\n</div>\n")
```

## Statistical Analysis

```{r methods-stats, results='asis'}
de_method_s <- de_method

test_desc <- switch(tolower(de_method_s),
  "t_test"       = "Welch's t-tests (unequal variance) were performed for each metabolite.",
  "t_test_equal" = "Student's t-tests (equal variance assumed, MetaboAnalyst convention) were performed for each metabolite.",
  "wilcoxon"     = "Wilcoxon rank-sum tests were used as a non-parametric alternative for differential expression testing.",
  "limma"        = "Moderated t-tests were performed using the *limma* R/Bioconductor package, which applies empirical Bayes shrinkage of variance estimates to improve statistical power.",
  sprintf("Differential expression was assessed using the *%s* method.", de_method_s)
)

cat(test_desc, "\n\n")

cat(sprintf(
  "Multiple testing correction was applied using the Benjamini-Hochberg procedure to control the false discovery rate (FDR). Metabolites were considered differentially expressed at adjusted p-value $\\leq$ %s and |linear fold change| $\\geq$ %s.\n\n",
  padj_cut_m, lfc_cut_m
))

cat(sprintf(
  "All analyses were performed in R (version %s) using the multiomics-core pipeline.\n\n",
  paste(R.version$major, R.version$minor, sep = ".")
))
```

---

# Session Info

<details>
<summary>Click to expand</summary>

```{r session-info}
sessionInfo()
```

</details>

---

<p style="text-align:center; color:grey; font-size:0.9em;">
Report generated by multiomics-core pipeline | `r format(Sys.time(), '%Y-%m-%d %H:%M')`
</p>
